### 2009-04-18 response-to-reviews.txt
Import from CVS.

> ------------------------------------------------------------------------
> 
> Reviewer #1 (Remarks for the Author):

Both reviews were extremely constructive and a genuine pleasure to
read.

> There are very interesting and potentially quite influential ideas in
> this paper, but too many of them at once are breathlessly muddled up
> together in a fog of words, without sufficiently careful description
> and logical analysis.

We are excited about the possibilities of this project, which we began
3 years ago with the goal of building algorithms to reconstruct
ancient ribosomal RNA.

The main products of this work, and the points of the paper, are:

 1. the mathematics of the new algorithms (tree transducer composition
 and "loopy DP");

 2. implementations of these algorithms as working code (the
 "indiegram" program).

The investigations of various other aspects of RNA reconstruction
(such as error rates, alignment quality & benchmarks of our other
software programs) are secondary to these main algorithmic points.

We agree that the inclusion of these secondary aspects in the main
paper was distracting. Accordingly we have moved these sections to the
supplementary information and trimmed them substantially.

(See below for an explanation of why they were included in the first
place, and what their relevance is; essentially, they are precursors to
future benchmarks we anticipate conducting pending implementation of
more sophisticated MCMC kernel algorithms.)

> The paper does not convey a clear, thorough, reproducible piece of
> science to the reader. It lacks focus, discipline, and rigor. The
> authors need to choose one story they want to tell, then tell it
> completely and thoroughly.

The story we are trying to tell is the systematic extension of
Pair-SCFG-like models to multiple sequences, with a working (albeit
memory-hungry) implementation on a three-branch tree.

We have attempted to bring this story into much sharper focus with our
revised Introduction and structure.

> The paper opens with some grand statements about a research program to
> develop "paleogenetic" algorithms to reconstruct the RNA world. I
> think the authors are using this story to dress up what they're doing,
> rather than really taking this research program seriously and
> analyzing it logically. For starters, in order to phylogenetically
> reconstruct a "primordial ribozyme" such as the "all-RNA primordial
> ribosome" from the "RNA world" from extant sequences, the authors need
> to be able to argue that the last common ancestor of rRNAs lived in
> the RNA world, not the modern DNA/RNA/protein world. It seems clear,
> however, that the last universal common ancestor (LUCA) of all known
> extant life had a reasonably complicated and more or less standard
> DNA/RNA/protein biochemistry, because (even taking abundant lateral
> transfer into account) of how many proteins and biochemical pathways
> can be traced back to LUCA. A serious attempt to reconstruct any
> aspect of the RNA world by phylogenetic analysis must deal with the
> probability that LUCA is almost certainly a bottleneck that occurs
> after the RNA world had passed (if the RNA world existed at all). This
> is not to say that ancestral reconstruction of structural RNAs isn't
> interesting, even if they reconstruct RNAs that lived in less fabulous
> creatures than ribosaurs -- it's only to say that the authors' stated
> goals seem under-analyzed and over-grandiose, and therefore are
> offputting to the reader. Much of the introduction's verbiage about
> the RNA world could be deleted, and the aims of the paper could be
> focused more realistically, concisely, and clearly.

Point taken.

Reconstructing early ribosomes has been a goal of this work for
several years, and Crick's all-RNA ribosome is often cited as a
convincing argument (or at least an inspiration) for the RNA world
hypothesis. So we did not think that, by quoting the RNA world
literature, we were being grandiose or trying to "dress up" our work.

However, we see the reviewer's point that "LUCA != RNA world" and we
agree it would be clearer and more consistent to focus on LUCA.  (Our
algorithms can only reconstruct RNA sequence from periods during which
a "more or less standard biochemistry" was established, and the
reviewers' points about bottlenecks at LUCA are good ones.)

Accordingly, we have rewritten the introduction with more realistic
expectations for the epochs we may actually be able to reach (in fact,
only the first paragraphs of the Introduction and Abstract ever
mentioned the RNA World; the newly revised manuscript barely mentions
it).

We enjoyed the word "ribosaur" -- is it published, or otherwise in
current usage? (Not according to Google...)

We think that the word "paleogenetics" is reasonably well-established
(by the work of, e.g. Pauling & Zuckerkandl, Svante Paabo, Steve
Benner, etc., which we think this naturally relates to); however, if
the reviewer feels that it is confusing or inappropriate, then we can
change the wording accordingly.

> At the same time, the introduction lacks a clear problem definition
> statement. What is the input; what is the desired output; what is the
> main idea behind the approach for transforming input to output?

The central problem, as we now state more clearly in the Introduction,
is this:

 Given a conditionally normalized Pair SCFG, modeling P(Y|X) where X
 and Y are foldback RNA structures, along with a phylogenetic tree,
 what is the corresponding multiple-sequence SCFG; and can we run the
 Cocke-Younger-Kasami algorithm on it (suitably constrained) ?

The solution to this problem is given by the transducer composition
algorithm, which is the core result of this paper.

> Throughout the paper, terms and programs are thrown around in a
> muddled soup, as if everything is a general purpose magic inference
> tool. _xrate_ estimates substitution rate parameters (p.2) and folds
> RNA (p.5); _evoldoer_ generates pair SCFGs (p.3) and "implements"
> TKFST (p.5), whatever "implements" may mean (any given probabilistic
> model may be used for a wide range of different simulations or
> inference problems). Do they slice and dice tomatoes too? At far too
> many points, the authors recite lists of possible applications that
> might be tackled someday, or lists of other approaches that can be
> viewed as special cases of their approach -- without ever particularly
> clearly defining what their approach is. I'm all in favor of
> generalizable approaches and powerful probabilistic models, but I want
> to see at least one specific, well-described problem solved before one
> expounds on all the other things that might possibly be done. That's
> what the Discussion is for.

We have emphasized the central problem of constructing a model of
multiple sequences given a model of the evolution of two sequences and
a phylogenetic tree (the "transducer composition" algorithm) in the
revised manuscript, including a full statement of inputs and outputs.

We have also removed excessive references to program names as
requested by both reviewers.

We have attempted to clarify the possible applications of these
methodologies and place them in the context of prior work.  Some of
these applications/questions are:

-- Given a conditional distribution P(Y | X), can one invert this and
compute (or sample from) the distribution of the ancestral sequence,
i.e. P(X | Y)?

-- Can one repeatedly apply this model and e.g. sample Z from P(Z|Y)
where Z is also a sequence/structure? Can one sum over all values of Y
and compute P(Z|X) directly, or sample Z from it?  (This is indirectly
related to the problem of time-dependence, since the branch from X-->Z
represents twice as much evolutionary time as the branch from X-->Y.)

-- How well does the model handle multiple sequences and missing data?
E.g. can we sample from the distribution of unobserved "grandparent"
sequences, given a set of observed "grandchildren" (and unobserved
"parents")?

-- What about if you have an unobserved parent with three observed
children? Can you sample from P(parent|children) ?  (This is closest
to the task we actually implement; doing this is sufficient to address
all of the other questions using an MCMC approach, as shown by our
2001 paper and several other researchers.)

-- Can we estimate the multiple alignment and phylogeny of a given
set of sequences, starting only with the model P(Y | X) ?

All of these questions can be addressed by the multi-sequence SCFGs we
describe, using standard SCFG algorithms (Inside-Outside, CYK, etc.)
and constrained MCMC variants on those algorithms.

> The authors could do a much better job of recognizing that ancestral
> sequence reconstruction has been done for decades: both parsimony
> methods and maximum likelihood methods for phylogenetic inference
> explicitly infer sequences at ancestral nodes in a tree.

Thank you for pointing this out; we did not intend to slight previous
authors.  In our rewritten Discussion, we have cited more of the
basic papers that established the computational and mathematical
foundations of the field.  We also cite the 2007 book edited by David
Liberles, an excellent survey of current uses of substitution models
for ancestral reconstruction, as well as the origins of the field
(e.g. Pauling & Zuckerkandl's paper, parsimony & ML methods, and so on).

> What's of interest here is whether that can be done not only for
> residue substitution events, but for insertion/deletion events as
> well,

We completely agree.
Insertions and deletions in unstructured sequences can be well-handled by
conditionally-normalized Pair HMMs, or "string transducer" state machines,
for which several implementations of the transducer composition
algorithm for building multi-sequence models from pairwise models now exist.
One of the central goals of this paper was to extend this theory to
the case of RNA secondary structures (or parse-tree transducers), in
order to model effects like indels that conserve RNA structure.

> and whether the inferences are sufficiently accurate that we can
> take them seriously and actually synthesize the inferred ancestor(s)
> and believe that they tell us something about the structure and
> function of extinct biomolecules.
> 
> The Methods are unnecessarily opaque and confusing, focusing too much
> on peripheral terminology and too little on an actual problem at
> hand. For example, for all this talk of "transducer theory", it is
> never clear how a "transducer" is meaningfully different from any
> other conditional probability model used in sequence analysis -- such
> as profile HMMs and profile SCFGs, which model a conditional
> distribution P(Y | X) via a model built from sequence X, P(Y | X) ~=
> P(Y | model) P(model | X) using a point estimate of an appropriate
> model derived from X, where X could be either a query
> sequence/alignment in a similarity search,

We apologize for the lack of clarity.  We have attempted to rectify
this in the rewritten manuscript by first discussing
jointly-normalized models (Pair HMMs) and then introducing
conditionally-normalized models (transducers).

We believe that the transducer is a meaningful improvement on
HMMs/SCFGs because of the transducer composition algorithm, which
offers a natural way to construct models of multiple sequences; for
example, given a model P(Y|X), and a suitable phylogeny, we can build
a model like this

   P(A, B, C, D, E | X, TREE)

where X is the root sequence of the tree, and {A,B,C,D,E} are leaves.
This model correctly accounts for all combinations of overlapping
indel events that can possibly occur on the branches of the tree --
for example, an insertion on an ancient branch can be interrupted by
an insertion on an intermediate branch, then both insertions can be
[partially] deleted on an even younger branch. The model is a
multi-sequence HMM, so that multiple sequence alignment is simply an
application of the Viterbi algorithm (or posterior decoding), and MCMC
Gibbs samplers can be constructed by applying suitable constraints to
the Forward algorithm.

Furthermore, dynamic programming (and MCMC-sampling) algorithms exist
to invert this model

   P(X | A, B, C, D, E, TREE)

Ancestral reconstruction involves sampling from and/or maximizing this
inverted distribution. Many other Bayesian manipulations can similarly
be performed on this likelihood function.

This ability to generalize systematically to multiple sequences is
missing from HMMs and SCFGs. While it is possible to convert a
single-sequence HMM into a phylo-HMM (by having it emit alignment
columns instead of single symbols), and even to simulate gaps to some
extent (by modeling them as an additional character), there is no
simple recipe (other than transducer composition) for extending a Pair
HMM to a corresponding multi-sequence HMM that includes affine gap
penalties, overlapping indels, nested indels, etc.

> model derived from X, where X could be [...]
> an ancestor on a tree in a phylogenetic inference problem.

We appreciate the point that there are similarities between Profile
SCFGs and the transducers which we describe here. For example, the
paper by Klein & Eddy (2003), describing the RSEARCH program,
introduces a profile SCFG which may be regarded as modeling a
conditional distribution P(Y|X). However, to our knowledge, existing
literature on Profile SCFGs does not describe the formalisms which we
need to reconstruct structured sequences on trees.  In particular, we
need ways to find:

-- the corresponding joint Pair-SCFG, modeling P(X, Y);
-- the Bayesian inverse Profile-SCFG, modeling P(X | Y);
-- situations involving more than two sequences, e.g. the
   Profile-SCFG for P(X | Y, Z), where X is the last common ancestor
   of Y and Z; or the corresponding Pair-SCFG for P(Y, Z); or the
   corresponding Triplet-SCFG for P(X, Y, Z).

Transducers provide a generic, systematic way to build such
multi-sequence, phylogenetically-structured SCFGs. In this paper we
attempt to describe how to perform these procedures.

> Seems to me that the critical thing,
> which the authors actually _underemphasize_, is how one makes this
> conditional probability model P(Y | X) be dependent on evolutionary
> time t, P(Y | X,t), via an underlying continuous-time Markov model of
> evolution in terms of rates of individual events including
> substitutions, insertions, and deletions.

This is a good point.  Time-dependent parameterization is indeed
critical, but it can be handled as a separate problem from the one we
address here.  Our goal is to provide a modular solution to the
ancestral reconstruction problem, wherein the problem of creating
accurate (two-sequence) models can be treated as disjoint from the
problem of extending those models to multiple sequences.  In this
paper we have focused on the second problem: Given a model of the
evolution of two sequences, how do we construct a model of many
sequences?

For this reason, while the time parametrization problem is certainly
central to the ancestral reconstruction problem, we believe that it
can be treated as a distinct problem from the multi-sequence model
construction problem which is the focus of this paper.  As the
reviewer points out, we largely gloss over how to choose an
appropriate time parametrization, and instead focus on:
Suppose you do have such a time-dependent model. How do you then
extend it from two sequences to multiple sequences?

More generally, how do you sample from some generic distribution

P(X_1, X_2 ... X_M  |  Y_1, Y_2 ... Y_M,   TREE)

where {X_1 ... X_M} are all ancestors, and {Y_1 ... Y_M} are all
descendants, related by a given TREE?

For substitution models, the answer to this question involves
Felsenstein's pruning algorithm.

For indel/RNA models, our answer is that you construct a
multi-sequence SCFG and then sample paths through it (appropriately
conditioned on the data {Y_n}).  Transducers allow you to build this
Multi-SCFG, assuming you already know the Pair-SCFG (i.e. the
parse-tree transducer).

Note that neither of these answers is the same as the problem of
getting time-dependent parameters! For substitution processes, you get
time-dependent parameters by matrix exponentiation (i.e. solving a
continuous-time Markov chain).

For RNA/indel processes, the TKFST model described in earlier work
(Holmes 2004) represents one attempt to get to these time-dependent
parameters for point substitutions, covariant substitutions,
insertions, deletions and structural changes.

TKFST is flawed, and certainly needs work to make it more realistic;
however, we believe that it's a reasonable simple model to test our
solution to the second half of the reconstruction problem.

TKFST is not the main focus of the present paper -- it's just the
particular pairwise model we're using as an example.  We could have
used another pairwise model to illustrate transducer composition, and
we have attempted to make sure that the composition algorithm itself
is not presented in a TKFST-specific way.

> This time-dependent evolutionary model has essentially nothing to do
> with (is separable from) "transducer theory" as far as I understand;

We agree; with the exception that, if you take the time-dependent model as
representing a single branch, and consider extending it to an
arbitrary phylogenetic tree using the sum-product algorithm, then this
line of reasoning will lead you to an algorithm which is essentially
the same as the Inside-Outside algorithm on a Multi-SCFG constructed
by transducer composition.

For example, the multidimensional DP algorithm of Hein (PSB, 2001) for
computing the TKF model likelihood on a binary phylogenetic tree is
exactly equivalent to the a Forward algorithm for a composite
phylo-transducer. Though not originally expressed in transducer
formalism, Hein's algorithm is (arguably) easier to explain that way.

So there are conceptual relationships. It's also true that serial
composition of transducers amounts to "adding" two branch lengths, so
this is another relationship to the time-dependent theory (indeed, it
can be considered a "test" of the additive property of time-dependent
models).

But these are subtle connections. We agree with the reviewer that these
two questions (branch length and tree topology) can for most intents &
purposes be considered to be separable -- just as, with substitution
models, matrix exponentiation is separable from pruning.

> "transducer theory" is just a different set of words for conditional
> probability models, with the evolutionary continuous-time Markov
> model providing the particular _parameters_ that get plugged into
> the conditional probability model.
> (That is, I believe that I can interchange a
> "string transducer" with an HMM and "parse tree transducer" with an
> SCFG with no loss of information. If this is true, the exact names of
> things aren't important so long as they're clearly defined and
> explained for the particular purpose at hand, and one shouldn't be
> trying so hard to brand "transducer theory" as new. If I'm wrong, I
> would like to see the authors' clear explanation.)

While they are very similar, the essential difference is that there is
no general-purpose recipe to convert a Pair-HMM or Pair-SCFG into a
multiple-sequence HMM or SCFG. In order to do this you would need

(a) to label one sequence as the ancestor and the other as the
descendant;
(b) to ensure that the HMM/SCFG was then conditionally normalized;
(c) to devise an appropriate algorithm for constructing the state
space and the rule space of the composite machine (which essentially
amounts to co-ordinating transitions between the component pairwise
machines).

Steps (a) and (b) are trivial, but step (c) is nontrivial and is the
point of this paper. (The analogous algorithm for string transducers
is described in (Holmes, 2003).)

The reviewer is correct that, e.g., Pair HMMs and "string transducers"
are very similar, so that a conditionally-normalized Pair HMM is
essentially the same as a string transducer. This just amounts to a
restatement of steps (a) and (b); the innovation is in step (c).

We did not aim arbitrarily to brand transducers as new; rather, we are
attempting to highlight connections to existing literature.  We have
attempted to make this clearer in the revised manuscript by
emphasizing the term's origin in the computational linguistics
literature ("parse-tree transducer" comes from the same literature).
We have also attempted to make clearer the tight connections between
existing HMM theory in bioinformatics and the "transducer theory"
described in computational linguistics (c.f. Mehryar Mohri et al.)

The idea of transducer composition, above and beyond the
straightforward idea of a conditional probability model, has been used
in several other recent papers. For example:

(1) Paten B, Herrero J, Fitzgerald S, Beal K, Flicek P, Holmes I,
Birney E.  Genome-wide nucleotide-level mammalian ancestor
reconstruction.  Genome Res. 2008.

(2) Satija R, Pachter L, Hein J.  Combining statistical alignment and
phylogenetic footprinting to detect regulatory elements.
Bioinformatics. 2008 May 15;24(10):1236-42.

(3) Varadarajan A, Bradley RK, Holmes IH.  Tools for simulating
evolution of aligned genomic regions with integrated parameter
estimation.  Genome Biol. 2008 9(10):R147.

We think that the transducer formalism makes these papers considerably
simpler to read (and provides a connecting thread between them).

> As an example of
> what I mean by underemphasizing the evolutionary model, consider the
> very term "statistical alignment" that the authors use to describe for
> their subfield: the heart of the approach is neither statistical (a
> "statistic" being an arbitrary measure of something, as opposed to a
> probability model) nor really about "alignment" (the problem at hand
> is phylogenetic inference, not alignment; alignment is necessary, but
> only as a nuisance variable in the inferences that the authors are
> actually interested in making) -- not to mention that there is a large
> body of other work on probabilistic alignment approaches that the
> authors seem to be trying to marginalize by distinguishing only their
> particular favorite flavor as "statistical alignment".

We intended to use the term "statistical alignment" in the context
introduced by Jotun Hein's 2000 investigation of the TKF91 model.  It
has since been used to describe probabilistic alignment
approaches that attempt to optimize, and/or sample from, a
phylogenetic likelihood function based on an evolutionary model of
indels.

This phylogenetic likelihood function is our understanding of the
difference between "statistical alignment" and other probabilistic
approaches to multiple alignment -- such as aligning everything to a
consensus HMM (c.f. 'hmmalign' in the HMMER package), or combining
Pair-HMMs with a non-phylogenetic objective function for
posterior-decoding (c.f. PROBCONS and AMAP).

However, we agree with the reviewer that the term is ambiguous, and
after some consideration we have de-emphasized the term in the revised
manuscript, leaving only the following sentence:

  The literature on phylogenetic alignment using probabilistic indel
  models has been dubbed ``Statistical Alignment'' by Hein et al
  (2000).  However, we have avoided this term in the present
  discussion, since almost every structural RNA alignment program
  makes some use of statistics.

Other than our (now reduced) use of this phrase, we hope there is
nothing about this paper that implies we are trying to marginalize
other alignment approaches. In fact if there is, we would greatly
appreciate any specifics the reviewer can provide, as this is
something we'd seek to correct.

> I struggled, section by section, to decipher in the Results what the
> take home message of the paper -- or even any individual section -- is
> supposed to be. Each section of Results has similar problems; I'll do
> a fairly detailed analysis of just the first results section, then
> I'll be brief with the rest.
> 
> Results 2.1 attempts to describe an experiment in which the authors
> ask whether it is necessary to use "covariant" rate models, as opposed
> to point substitution rate models, to accurately reconstruct ancestral
> RNA sequences. There are immediately a number of problems just in the
> conception of this experiment, including:
> 
> 1. How is it relevant? The introduction claimed that "the central
> result of this paper is the development ["of a theory for statistical
> alignment and reconstruction of RNA"] building on the transducer
> framework for protein and DNA reconstruction". But rate matrices for
> single residues or even for base pairs are standard models in the
> field [...] and have nothing really to do with "transducer theory"
> nor with inference of indel structure, supposedly the main point of
> the paper.

We agree that this section distracts from the theoretical presentation
of transducer composition, and we have moved it to the supplementary
info.

To answer the question: "How is it relevant [to the central result]?"
We hope that this is primarily answered by the revisions we have made
to the paper; some further discussion follows here.

A phylo-SCFG such as the one we are benchmarking here can be viewed as
a special case of a multi-sequence SCFG, and can be similarly derived
using transducer composition. (If the parse-tree transducer does not
allow indels, i.e. if the indel rate is zero, and if its substitutions
use a time-dependent parameterization based on a continuous-time
Markov chain, then you end up with a phylo-SCFG like EVOFOLD or
PFOLD.)

So, in addition to being a framework for constructing multi-SCFGs with
indels and covariant substitutions, this is (at the same time) a
framework for constructing phylo-SCFGs. In case this seems like
excessive machinery for such a simple task, we note that there is
precedent for viewing phylo-HMMs as constrained or trivial special
cases of multi-sequence transducer-based HMMs (c.f. Siepel et al;
Diallo, Blanchette et al) and that such a view can be a useful guide
as to what approximations to make. So we think it is useful to make
this connection.

Multi-sequence SCFGs are memory-hungry. Pending development and
implementation of more efficient constraints or sampling algorithms
(such as the Redelings-Suchard MCMC kernel cited in the paper, which
involves doing two composite-transducer DP steps followed by a
Metropolis-Hastings accept/reject step), we are not yet able to do
systematic high-throughput tests of the more general & complex
multi-SCFGs (such as TKFST-based multi-SCFGs) which allow indels as
well as substitutions.

To recap, so far we only have a general-purpose transducer compositor
(that's this paper); this is a step towards building a
Redelings-Suchard sampler for RNA, but only a step; building the R-S
sampler (or some other efficient MCMC sampler) would be a significant
further step.

Efficient general-purpose implementations of phylo-SCFGs do currently
exist (conditional on an alignment) and xrate is one such general
implementation.

Our goal, in including the xrate reconstruction benchmark, was to
provide signposts to later multi-SCFG tests, by beginning with a
benchmark of the special-case phylo-SCFG.

However, we thank the reviewer for pointing out that these results
detract from the central message, and so we have accordingly moved
them to the supplementary material.

> (incidentally, the authors cited zero prior work on using base
> pair rate matrices was cited, but a fair amount exists)

The original paper did cite two papers by Knudsen and Hein
that use basepair rate matrices and the revised paper adds a citation
to RSEARCH (see above).  We can certainly expand on this with
further citations if the reviewer feels it appropriate.
We intended no slight to previous work, and have attempted to
emphasize prior art in the revised manuscript.

> 2. Why do the experiment at all, when the answer is already known and
> trivial? Obviously a point substitution model will not constrain
> ancestral base pairs to be valid base pairs; obviously modeling with
> base pairing correlations makes for a better model of a structural
> RNA. The reason to do a *quantitative* experiment to address a
> question whose qualitative answer is trivially obvious would be if the
> quantitative error rate was meaningful in some sense that had been set
> up in a problem definition statement. This would require deeper
> thought and discussion about how accurate an ancestral RNA
> reconstruction actually needs to be, for an intended purpose. For
> example, one might imagine a cost-benefit analysis of how many
> sequences need to be synthesized to find a functional ancestral
> sequence for a given error rate, if we're taking this research program
> seriously; then there might be a threshold prediction error rate that
> one needs to drop below, in order to make synthesis and testing of the
> predictions reasonable.

These are intriguing suggestions. The proposed Operations Research
study of the entire problem (including synthesis) is an especially
interesting new direction.

Our less-ambitious approach to the research program was to set up a
competitive benchmark of different reconstruction methods, using some
metric of reconstruction quality.

We believe that "percentage nucleotides correctly reconstructed" is a
reasonable benchmark metric for reconstruction quality.  It has been
used previously in other work. (See, for example, Paten et al,
Genome-wide nucleotide-level mammalian ancestor reconstruction. Genome
Res. 2008.)

We acknowledge that our current results in this area are preliminary,
which is why we have now relegated this section to the supplementary
info.

We are certainly trying to take this research program seriously.
However, these methods are challenging -- it took us three years to
get this far (i.e., implementing RNA structure on a three-branch tree)
-- and even our tightly-constrained algorithms push the limits of current
computational power, except on very simple RNAs.

Nevertheless, we think the transducer composition and loopy DP
algorithms, combined with our working implementations, are noteworthy
enough to publish. (For example, we think this is a useful step
towards a Redelings-Suchard kernel for RNA.)

> But instead, the authors do go ahead and draw
> only qualitative conclusions from their results in Fig 1, that
> modeling base pairs is better than not modeling base pairs: "we
> conclude that deep phylogenetic reconstructions of ribosomes, and
> other RNAs, will require covariant substitution models that take
> account of RNA secondary structure". But this, anyone working on RNA
> already "knew". Wait, the authors might say, maybe we shouldn't trust
> our intuition? Fine -- but ironically, in the very next sentence, they
> show how willing they are to make essentially the same obvious
> qualitative leap, on the most relevant point to the theory they want
> to express, without any quantitative analysis: "We may reasonably
> deduce that indel reconstructions will similarly need to take account
> of RNA structure". This "deduction" does not follow from the residue
> substitution results in Figure 1 any more than it already followed
> from what we already knew about RNA structure and evolution.

We agree, these particular qualitative conclusions were not 
surprising, and the logical chain of reasoning (placing these
benchmarks before the development of the model, when in fact they are
limited benchmarks OF the model) was muddled. We have removed these
parts.

We have every intention, eventually, to investigate the indel
reconstruction accuracy in a systematic high-throughput way (as we
have begun to investigate the accuracy of substitution
reconstruction). The memory complexity of the 3-sequence algorithm
means that most structures in RFAM are just out of range on current
hardware, so we restricted ourselves to tests of a simplified model.

Our revised manuscript discusses these limitations more clearly.

> Even beyond the conception and rationale of the experiment shown in
> section 2.1, the design and execution of the experiment are
> inadequately explained. What base pair rate models did you estimate --
> a completely unconstrained 16x16 base pair model? a general-reversible
> model? or a model with more constraints? What "75-taxon phylogeny" did
> you use (Rfam does not provide phylogenies); with what branch lengths?
> When you simulate 5000 "random alignments" of descendant sequences on
> this tree, how do you choose the initial ancestral sequence for each
> simulated alignment (it must have a base-paired structure for the
> experiment to be reasonable; there's many ways you could have done
> this, but it's unclear which, if any, were used).

In our revised manuscript, we have attempted to explain these points
more clearly for ease of reproducibility.

Most of the explanations that the reviewer is seeking were available
in the original manuscript, but as the reviewer points out, they were
frequently obscured.  We have attempted to correct this.

Specifically:
-- The actual distribution of ancestral structures is
unimportant (since we're not attempting to infer structure in this
particular benchmark, but are assuming that we already know the
correct structure) but we could have stated this more clearly.
-- Although we did say that the model was "reversible", we didn't
explicitly state that the 16x16 rate matrix was otherwise fully
general and unconstrained. Nor did we explicitly say that the
ancestral sequence was sampled from the equilibrium of this reversible
model.
-- The parametric structure of the 16x16 rate matrix and the initial
distribution of ancestral structures were both based on the PFOLD
phylo-grammar, which was referenced in the following benchmark
section, but not in this section.
-- The actual grammar over structures should not affect the final
result for this section, because we are assuming that we *know* the
true structure. The only thing that really matters is that the
structure includes both basepairs and single nucleotides, which we
then handle separately. Again, it would have been better to err on the
side of excessive experimental detail, rather than leave this out.

We hope that all of these explanations are now clearer in the text,
and we thank the reviewer for the feedback.

> Why did you strip only the ancestral (root) sequence out of the
> alignment? shouldn't _all_ internal node sequences be stripped out,
> if we're trying to simulate the ability to reconstruct ancestral
> seqs from extant seqs?

Yes, this _is_ what we did -- we took out all the internal node
sequences and the ancestral root sequence. By "ancestral sequences"
we didn't mean only the root sequence, but all the ancestors
(i.e. internal nodes) in the tree. We have attempted to clear this up
in the text.

> Results 2.2 follows a similar pattern. The authors use bold face to
> "emphasize that this [TKFST] is not just a method for simultaneous
> alignment and folding of RNA" -- even before they make a concrete
> problem definition of what TKFST _is_! A reproducible description of
> TKFST isn't given here; it is "reviewed in Section 3.1". No, it isn't
> given there either; Section 3.1 meanders in an unclear fashion, still
> without problem definition, hitting a variety of topics connected to
> TKFST, while saying at various points things like "full details... are
> given in the Supplementary Material". So much for this "theory" for
> "computational reconstruction of the RNA World" being the "central
> result" of the paper! What the heck is it doing lying around in pieces
> in the Results, Methods, and Supplementary Material, as opposed to one
> clear central exposition?

The reason for this apparent omission of a full description of TKFST
is that is not the central result of this paper.
Rather, the "theory" we were referring to was the
composition algorithm for extending any pairwise SCFG to multiple
sequences related by a tree. TKFST was merely an example pairwise
model that we were using to illustrate this algorithm.

(At the risk of excessively repeating ourselves: TKFST is a prototype
solution to the "time-dependent parameterization" problem, a.k.a. the
"branch length" problem; whereas the problem we're really trying to
solve here is the "multiple sequences" problem, a.k.a. the "tree
topology" problem.)

Hopefully the increased emphasis on the transducer composition
algorithm helps clear this up.  We have tried to give a slightly more
concrete description of the TKFST model; we can beef this up if the
reviewer feels it appropriate.

In order to emphasize that the transducer composition algorithm is our
main result, we have moved the results which the reviewer refers to to
the supplementary material.  These results did not use transducer
composition; they just used TKFST as a Pair-SCFG to do progressive
alignment (or, strictly speaking, multiple alignment by single-linkage
clustering).

We acknowledge that including this benchmark in a paper about
phylogenetic transducer composition was, therefore, rather confusing.
The reason we included it at all was, again, that this alignment
quality benchmark does resemble one of the benchmarks we intend to
repeat with transducer-based algorithms as soon as (a) RAM becomes
cheap enough or (b) we develop more efficient/constrained algorithms
by extending the techniques presented here.

Furthermore, this benchmark does speak to the usefulness of the TKFST
model, which (while not the central result) *is* relevant to this
paper, since we have used TKFST as our "test model" for transducer
composition.

As with the other benchmark, we have moved this section to the
supplementary info, and we hope the rewrite makes things clearer.

> My comments on the rest of the paper would continue repetitively in
> the same vein but since this review is already long, I'll stop. In
> summary, I think the authors are working on very interesting issues
> that they're explaining atrociously. Writing a clear paper requires
> more discipline. The authors need to go back to the drawing board for
> this paper, pick one point they want to make, and write a clear,
> logical paper that conveys one nice story rigorously.

Thanks again for the constructive criticism. We have tried to follow
these suggestions.

> some small points:
> 
> - group II introns are not "76-91 nt" (Table 3 and elsewhere); they
>   are much larger than this. The authors must be using the Rfam
>   Intron_gpII alignment, which only deals with two of the six
>   consensus structural domains (domains V and VI).

Thank you for this correction. We have added text noting this at
several places in the manuscript and supplementary information. (The
sequences were taken from BraliBaseII and thence from Rfam 5.0.)

> - it's a lot more useful to a reviewer to get a manuscript in
>   manuscript form (i.e. at least 1.5x linespaced if not doublespaced),
>   not a tightly spaced journal format, so there's ample room for
>   annotation.
>
> - the fonts in Figures 1 and 2 are too small.

We have implemented these suggestions.

> Reviewer #2 (Remarks for the Author):
> 
> I. The manuscript describes an method for simultaneously inferring the
> alignment and secondary structure of three modern RNA sequences along
> with that of the internal node on a 3-taxon tree - a previously
> unsolved problem to the extent of my knowledge. The manuscript also
> assesses the TKF Structure Tree (TKFST) model of RNA structure
> evolution in terms of its ability to help reconstruct ancestral
> alignments and structures. Both of these results (new method, model
> assessment) are important steps towards the goal of reconstructing the
> sequence and structure of ancestral RNA molecules, which is a very
> interesting biological question.
> 
> This topic is by nature complex and involved, and I am unfortunately
> not an expert in RNA structure. Nevertheless, I think that the authors
> display a deep and intuitive grasp of this field that is quite
> impressive, while being moderately understandable to the outsider. In
> most cases, I was able to follow the description of rather complex
> procedures on my first pass through the paper even though I was not
> previously acquainted with SCFGs in any detail. I think that the
> readability for readers that are less familiar with the background can
> be improved by adding some background in a few places, which I will
> try to indicate below.
> 
> II. My major advice on improving the readability of the paper is to
> decrease or remove emphasis on the names of programs. In general, an
> inference or estimate should be described in terms of (i) the model
> used, (ii) the type of estimate (e.g. maximum likelihood estimate,
> posterior mean, etc.) and (iii) the algorithms employed. It should not
> be described in terms of the name of the program that is run to
> perform the inference. For example, in the beginning of section 2.1,
> the authors write "We first used xrate to estimate parameters..." when
> they should say something like "We used an EM algorithm to compute
> maximum likelihood estimates of parameters for the substitution
> model. We held the alignment fixed, and summed over the unknown
> secondary structure and ancestral sequences."

This is an excellent point.  We have modified the text accordingly.

> (I recognize that the authors wish to point out their work in
> extending existing programs to provide infrastructure for estimating
> ancestral sequences.)
> 
> Secondly, I think that the authors downplay the very real value of
> their analyses by focusing too heavily on the exact protocol of the
> analyses and playing down the questions and results in addition to the
> methods. For example, in Section 2.1 they write "We implemented ..." ,
> "We first used xrate ..." , "Next we used the companion program
> SIMGRAM ...", "We stripped the ancestral sequence..." It would be
> helpful if the authors began section 2.1 with a description of the
> tests that would be performed in that section, and what questions they
> answer, instead of beginning by describing xrate. Similarly, in
> section 2.2 the authors make an important point in bold font: "We
> emphasize that this is not just a method for simultaneous alignment
> and folding of RNA. Our intention here is a critical assessment of the
> realism of TKFST as an evolutionary model, not simply an
> implementation of the algorithm of Sankoff."  Nevertheless, this
> section is titled "Multiple Alignment and RNA folding". Hopefully
> these examples clarify my general point about how the authors can
> improve readability.

Thank you for the suggestion.  We have moved what were Sections 2.1
and 2.2 to the supplementary material in order to make clear that our
principal result is the transducer composition algorithm for building
models of multiple RNAs on a tree.

> III. Specific suggestions
> 
> 1. [ Abstract ] I presume that xrate estimates maximum likelihood
> parameters estimates using the EM algorithm. I presume that xrate sums
> over all structures when estimating parameters? When it is used to
> estimate structures, I presume that structure estimates are
> conditioned on maximum likelihood rate estimates?
> 
> These points should be clarified somewhere, though perhaps not in the
> abstract.

This is correct: structure estimates are conditioned on the "model,"
which is parametrized with ML rate estimates.  We have moved the
"xrate" results to the supplementary material (so they aren't
mentioned in the abstract).  We have also clarified this point in the
supplementary material.

> 2. [Abstract] The abstract is written as if the main topic of interest
> is programs, not algorithms and biological questions. However, the
> authors underestimate the importance of their work by focusing on
> programs.

We have rewritten the abstract accordingly to place more focus on the
algorithms rather than the programs which implement them.

> 3. [ pg 2. / Section 2.1 ]
> 
> While implementing programs is an important part of biological
> research, I think that the authors are downplaying their true
> contribution in this section by describing the program names that were
> used instead of focusing on the methods and algorithms.

We agree and have revised the text accordingly.

> [pg 2 / Section 2.1 ] "We implemented ancestral sequence
> reconstruction in xrate, including posterior probabilities that the
> reconstructions are correct."
> 
> This single sentence should be expanded to at least one or two
> paragraphs describing the type of estimation that is being performed
> as well as the type of algorithm that implements this estimation. I
> presume that (i) xrate estimates parameter values of the substitution
> model using maximum likelihood.  (ii) I presume that in the definition
> of the likelihood the alignment is held constant, but the secondary
> structures and ancestral sequences are summed over (as latent
> variables).  (iii) The maximization over parameters is carried out
> using an EM approach.
> 
> If this is the case, then what kind of estimates of the structure, and
> of ancestral sequences, are being performed? I presume that xrate
> conditions on the maximum likelihood estimates of substitution
> parameter values and then selects the ancestral sequence (or
> structure) that has the highest posterior probability? If so, then
> these points might be a good starting place for a description of xrate
> and the extension that you performed.

We have expanded this text in the Supplementary Material.

> 4. [ pg 3 / Section 2.2 ] The authors should retitle this section to
> indicate that it is primarily about the critical assessment of the
> TKFST model, and focus less on the exact protocol. Perhaps the exact
> steps could be described in a separate "Methods" section that
> describes protocol details. These details must be provided so that the
> reader fully understands what the authors are doing, but they are not
> usually the main point.

This is an excellent suggestion.  We have attempted to make clear that
these results form a critical assessment of TKFST's appropriateness as
a model of RNA structural evolution when we reference these results in
the main text.

> The authors explicitly say "We emphasize that this is not just a
> method for simultaneous alignment and folding of RNA. Our intention
> here is a critical assessment of the realism of the TKFST as an
> evolutionary model." This is a good point - but the title and logical
> structure of this section focuses more on the protocol details for
> running programs.
> 
> 5. [pg 3 / "pair SCFG" ] This has not been defined yet. Many readers
> will be unfamiliar enough with SCFGs that it is important to define
> pair-SCFG's before saying too much about them.

The first use of "Pair SCFG" now appears after SCFGs have been
defined, and in the context of a discussion of pairwise models, while
citing previous work. Hopefully these pointers are now enough.  We can
add more information if the reviewer feels that it would be helpful.

> 6. [ pg. 3 ] It would be helpful to add one or two more sentences of
> explanation about sensitivity and PPV mean, and how they are applied
> to alignments and RNA structures.

We have added explanatory text to the table caption, defining and
discussing sensitivity and PPV for alignments and structures.

> 7. [ pg. 4 ] It is curious that the stemloc native grammar doesn't
> seem to cause much improvement in alignment sensitivity and PPV even
> though it has an affine gap penalty. Can you say why this is the case?
> What is the average gap length - is it near 1?
>
> 8. It would be useful to give parameter estimates from PFOLD or the
> stemloc-native grammar to indicate what kind of properties these RNA
> molecules have, and whether or not a model like TKFST that has fewer
> parameters should be expected to do a good job modeling the evolution
> of such sequences. For example, if the average gap length is near
> zero, then that parameter is not helping the PFOLD model much, but if
> it is large, then we should expect the fewer parameters of the TKFST
> model to result in worse model fit and inferior prediction quality.

The reviewer's insight into the affine gap penalty appears to be very
accurate. We have added the following sentences to the discussion of
these results:

One possible (partial) explanation [for the close results] is that the
affine gap penalty is not helping all that much: the mean gap length
in the stemloc grammar ranges from 1.3 to 3.5 bases (in loops) and 1.4
to 1.7 basepairs (in stems), so the improvement over a linear gap
penalty is small. (The mean gap lengths are specified as ranges
because stemloc's native grammar set includes four distance settings;
mean gap length is a function of distance.)

> 9. [pg. 5 ] "To test this, we tried re-folding the multiple alignments
> with xrate..."  This sentence is unclear because it uses the program
> name xrate instead of describing the procedure that was carried out,
> and because the word "re-folding" is a bit too terse. You might say,
> "To test this, we re-estimated rate parameters and secondary structure
> using the PFOLD grammar while fixing the multiple sequence alignment
> to the value estimated in the previous step." You might then
> optionally add that this was accomplished using xrate if you wish to
> emphasize which program was used.

We have replaced this sentence with the one suggested by the reviewer.

> 10a. [pg. 5] "The structure prediction accuracy of the TKFST and
> stemloc-native grammars was much better when combined with xrate, and
> both grammars were comparable..."
> 
> The authors should not say "when combined with xrate" but should
> indicate what they are using xrate to do.

We have changed this to "...when combined with a post-processing
structure prediction step".

> 10b. When the authors re-estimate the structure using xrate, they use
> a different model (PFOLD) to do so. Why, then, are these new structure
> estimates taken to represent "the structure prediction accuracy of the
> TKFST and stemloc-native grammars"? Don't they represent the
> structure-prediction accuracy of the PFOLD grammar, conditioned on
> alignments inferred using the TKFST and stemloc-native grammars?

Yes, this is true, so we have deleted the words "of the TKFST and
stemloc-native grammars", and added the reviewer's suggested phrasing
to the table caption.

> I don't see a good reason why the PFOLD model should be used during
> re-estimation of the secondary structure. If the problem is early
> commitment to incorrect second structures during an algorithm like
> progressive alignment, then re-estimating the structure with the TKFST
> and stemloc-native models (respectively) should eliminate the problem.

This is true, but we currently have no implemented software that can
predict structures from multiple alignments using parse-tree
transducer models or multi-SCFGs. We only have software (xrate) that
can do this for phylo-SCFGs (like PFOLD). Ideally we would use the
same model throughout the analysis; however the point here is to
demonstrate that the alignments are at least sufficiently accurate to
get reasonable structure predictions from a method that takes an
alignment as input (PFOLD).

> 10c. [Table 2] I don't think the last two lines should be
> characterized by "+xrate" since xrate is just a software
> package. Instead, you should choose a label that indicates what you
> are doing.  (I) I am not sure, but it looks like you are using PFOLD?
> (II) If so, then perhaps the last two rows should be written as
> "+PFOLD?"

True, we have now changed this from "+xrate" to "+PFOLD".

> 10d. [pg. 5] "We parametrized the singlet and branch transducers of
> the TKFST model using estimates reported by PFOLD, a phylo-grammar for
> RNA secondary structure prediction, and evoldoer, an implementation of
> the Structure Tree model for two sequences."
> 
> (i) How can you obtain estimates for the parameters in the TKFST model
> by using a different model - the PFOLD model?  (ii) This sentance
> again describes the names of program used instead of the methods and
> algorithms by which the estimates were constructed.  (iii) I would
> suggest rephrasing the last phrase "an implementation of the Structure
> Tree model for two sequences". This is because technically you don't
> implement models, you implement inference algorithms that use
> models. So, here, evoldoer implements inference of the pairwise
> alignment between two sequences under the TKFST model (I think). If
> one algorithm (DP) implements the hard part of a wide variety of
> inference methods then you could say so, I suppose.

(i,ii)
Thank you for pointing this out, and we apologize for the oversight.
We have described the origin of the so-called "PFOLD" parameters,
which we used because they have proven effective for RNA structure
prediction, in much more detail.  We used the substitution rate
estimates reported by Knudsen & Hein directly for our model.

(iii)
We have changed the text as recommended.

> 13. "Pair-SCFG" Can the authors describe a pair-SCFG in more detail?
> Although the manuscript compares the pair-SCFG to a pair-HMM, the
> authors should still point out (for example) that a pair-SCFG is an
> SCFG that emits pairs of letters, not a pair of SCFGs. It might also
> be worth explicitly pointing out that the transformation rule "S1 → x
> S2" in the SCFG corresponds to (in a pair-HMM) (I)state S1 emits
> letter "x" (II)state S1 then transitions to state S2.  These two
> things are probably worth pointing out, given that they do not add
> much length to the paper.

We have included this suggestion.  The introduction to the Methods
section describes the phenomena which Pair SCFGs can model and the
caption to Table 2 includes the example recommended by the reviewer.

> 14. [pg. 8] Define "X-terminals" and "Y-terminals." This is important
> terminology that is re-used, and it would be worth making sure the
> reader understands it here in order to clarify what follows. Also, you
> might clarify that X here represents the ancestor and Y the descendent
> along a branch of the tree, since this isn't always the case (I
> think).

We have replaced this terminology with just "terminals" for clarity and
we describe terminal symbols as analogous to emitted characters.
The text now tries to make clear that X is the ancestor and Y the descendant.

> 15. [pg. 10 ] "In practice, we find the size of the state space is
> approximately linear in the number of branches..."  Can you say
> whether you find this to be true for several different grammars, or
> just for (e.g.) the TKFST grammar? It would seem that the number of
> states should grow exponentially in the number of wait states for each
> branch transducer, or faster. However, because the TKFST model does
> not have affine gap penalties, it may have only one wait state,
> leading to the linear growth that you observed. Regardless of whether
> this guess is correct or incorrect, a comment about how affine gap
> penalties affect the number of states would be helpful here.

These are excellent points.

We do not have a proof of this complexity behavior, but we have observed this
behavior for several models.  However, as we do not have theoretical
upper-bound results, we have toned down these statements in the
text to clarify that they are merely empirical observations and that
we are unsure how they will generalize.

For technical reasons, TKFST actually has several wait states.
We believe that the growth will be sub-exponential in the number of
wait states because we are able to eliminate many of them using the
procedure described for eliminating "windback" states in the section
"Reduction of computational complexity."  We have noted this in the text.

> If this is the case, then could you say something like "In the case of
> the TKFST model" instead of "In practice..."? Also, can you clarify
> "realistic biological models never reach this bound"? What aspect of
> realistic models prevents the bound from being reached?

This bound required un-biological state configurations such as ones
which align un-related sequence.  We now state this clearly in the text.

> 16. [pg. 11] "... is sufficient for a likelihood-sampling algorithm
> over a tree".  The phrase "likelihood-sampling" is vague - technically
> the likelihood itself isn't being sampled. You could say something
> like "is sufficient for a calculating the likelihood of a full tree,
> which would be used in likelihood-based algorithms for sampling branch
> lengths, substitution rates, and other model parameters". Or perhaps
> you mean, sufficient for sampling the alignment and the RNA structure
> at all nodes?

We have changed this as recommended.

> [pg. 11] "Evaluation of the likelihood function of a
> node conditioned on its neighborhood..."  Can you rewrite this for
> accuracy and clarity?  (I) "... the likelihood function of a node ..."
> - this is vague, and focuses more on commonly used algorithms than on
> what exactly those algorithms calculate.

We have changed this as recommended.

> (II) "... conditioned on its
> neighborhood..." - technically, you are not statistically conditioning
> on the sequence at every node in the neighborhood, but just on
> (perhaps) the sequence at the ancestor node. The sequences at the
> other nodes are given, but you are acutally computing the probability
> of the sequences at these nodes, not conditioning on them.

We may not fully understand the distinction which the reviewer is
making, but if the sequence at a neighboring node of the node in
question is observed, then considering the distribution which is joint
with the observed neighboring node and the distribution which is
conditional w.r.t. the same node is equivalent.  The node is observed,
so they differ only by a multiplicative constant, ie, the marginal
evaluated at the observation.

We can state this point more clearly in the text if the reviewer feels
it appropriate.

> [pg 6., Fig 3] "Maximum likelihood reconstruction of ancestral
> structures" When you say "ancestral structures", do you mean the
> structure at the internal node on an unrooted 3-taxon tree? This is
> not the ancestor of the three leaf sequences, but the common ancestor
> of two of the sequences, using the third as an
> outgroup. Reconstructing the ancestor of all three would require
> knowing where the root of the tree was, if I understand correctly.

Thank you; this is an excellent point.  We have clarified the text
accordingly (in the caption of Figure 2 showing the corresponding
trees and inferred alignments).

> In summary, I hope that these suggestions are useful to the authors in
> improving readability of the paper. The only point that may be a real
> error is point 10a,b,c,d. Overall the manuscript was well-written and
> makes important contributions to the field of RNA structure
> reconstruction.
> 
> -Benjamin Redelings
>

### 2009-04-18 response-to-reviews2.txt
Import from CVS.

> ------------------------------------------------------------------------
>
> Reviewer #1 (Remarks for the Author):
>
> This new draft is vastly improved. (One comment I penned into the 
> margin at the end of the introduction is "lightyears better"). The 
> abstract and introduction now clearly state the problem and its 
> background. The methods are now clear enough up to about page 10 of 
> the manuscript. After that, however, the manuscript breaks down. It is 
> still too vague for me.
>
> It's now much clearer that the manuscript promises two main results: a 
> "transducer composition algorithm" for building a multisequence SCFG 
> from pairwise transducers, given a tree; and a dynamic programming 
> algorithm for calculating maximum likelihood parses under this 
> grammar, given N input sequences. The results consist of implementing 
> the DP algorithm for a particular RNA evolutionary model (called 
> "TKFST") and trying it out on a couple of small toy examples of three 
> short RNA sequences (nanos TCEs, and tRNAs). Though one can get the 
> rough gist of how the authors' approach works, many important details 
> necessary for an acceptable level of understanding and reproducibility 
> still seem to be missing.
>
> The problem statement (p.5) is ok but still needs some clarification, 
> and perhaps some discussion from the perspective of where the 
> rate-limiting problems in "paleogenetic reconstruction" really lie. 
> What exactly is given and what exactly is going to be inferred? Does 
> "Given a phylogenetic tree" mean that both tree topology and its 
> branch lengths are assumed known? (The examples in Figures 1 and 2 
> seem to indicate so.) Are individual secondary structures assumed to 
> be known for the observed sequences? (Again, Figures 1 and 2 suggest 
> so.) And from this are we inferring the alignment of the observed 
> sequences, and the alignment, secondary structure, and sequence of the 
> unobserved ancestors? Aside from clarification, one reason that it 
> would be good to nail this down is to bring out that the authors 
> apparently believe that it is reasonable to consider individual 
> secondary structures to be "known" in the input, whereas the alignment 
> is an "unknown" that must be inferred. To my mind,
> this would get the rate-limiting inferences backwards. If I were 
> really trying to reconstruct the evolution of an important RNA, I 
> would have a lovingly handcrafted multiple alignment already, much 
> better than any automated inference program would be likely to produce 
> (c.f. Gutell rRNA alignments, or catalytic intron alignments) as well 
> as a highly accurate consensus secondary structure, but individual 
> secondary structures (where they differed from consensus) would remain 
> in doubt, and would need to be inferred. Not that structural alignment 
> isn't of interest; just that it's not obvious that it's the most 
> important task if paleogenetic reconstruction is the real
> motivation.

Thank you for the detailed recommendations.  We now state clearly that
we take the phylogeny as given, although we also note in the
Discussion that in principle we could construct a basic sampler
over trees using our method (though we have not constructed or
evaluated such a sampler).

We have added more discussion of the potential rate-limiting steps and
corresponding constraints to the manuscript ("Overview").  As the
reviewer points out, perhaps the most realistic pieces of outside
information to have are a consensus structure for sequences and/or a
curated multiple alignment.  In principle these can all be
incorporated into our inference algorithms with the alignment
constraints used by e.g. QRNA and PFOLD, or the simultaneous fold and
alignment constraints used by e.g. Stemloc and CONSAN.

For this paper we used constraints on individual structures because
these are simple to implement.  We state this clearly in the revised
text and note that while experimentally-determined structures of
individual RNAs are rare, curated multiple alignments are
more-commonly available and can be used to constrain the algorithms
which we describe.

In principle we think that the reconstruction might benefit as well
from partial alignment constraints as from partial fold constraints;
that is, as well as being able to deviate slightly from the consensus
fold (as the reviewer suggests), the reconstruction algorithm would
probably need to deviate slightly from the supplied alignment as
well. (This will be necessary because not every evolutionarily related
basepair is likely to be perfectly aligned.) This sort of flexible
fold and alignment constraint is, of course, implemented in CONSAN and
Stemloc. However, we note that the Pair SCFGs used in both those
algorithms were first implemented using much simpler constraints:
specifically, the fold-only constraints of (Holmes 2002), and the
alignment-only constraints of (Rivas & Eddy 2001).

We do hope that the sophistication of the constraint schemes used with
these "tree transducer" models will increase over time. Eventually we
hope to develop alignment constraints, simultaneous alignment-and-fold
constraints, and other constraints (and/or sampling schemes) to enable
efficient analysis of multiple phylogenetically-related sequences. The
present work represents a first attempt at using these sorts of models
for more than two sequences, but we hope it will not be the last
attempt, and we acknowledge room for improvement.

> It is still not clear eenough what the authors' model is, though it is 
> supposed to be a main result. Page 7 states "a prescriptive formalism 
> for transducers is given in Text S1". Relegating the deliverables of a 
> manuscript to supplementary material is not an appropriate use of 
> supplementary material; a reader should not have to refer to a second 
> manuscript to understand the main points. The authors need to find a 
> way to explain their main results in the actual manuscript. There is 
> still much vagueness and redundancy in this draft, which could be 
> removed and replaced with more relevant explanation. For example, in 
> this same section where results are relegated to supplementary 
> material, there's a whole redundant paragraph (p.8) on how conditional 
> models represent directed edges; but this concept has already been 
> explained sufficiently. This is followed (bottom p. 8, "Consider a 
> phylogenetic tree...") by yet another statement of the same concept of 
> using conditional models, still in
> terms too vague and conceptual to enable reproduction of the work.

Thank you for pointing this out.  We have removed the reference to the
supplementary material in this section and have attempted to make the
main text a self-contained description of the composition algorithm.
We have attempted to remove redundant text, including the two
paragraphs which the reviewer notes.

> It is still not clear enough what the authors mean by a "composition 
> algorithm". The description of the generative model on pp 9-10 seems 
> redundant with the "composition algorithm" on pp 11-12; pp 11-12 don't 
> seem to add anything that I didn't already understand from pp 9-10, 
> and is still not precise and concrete enough for me to really 
> understand the actual model in play here. My understanding is that a 
> "composition algorithm" is just how one chains conditional models 
> together to calculate the desired joint probability of the observed 
> data. Conceptually, this chaining seems obvious enough to me when a 
> phylogenetic tree is given, and the use of so much space to describe 
> the concept is puzzling; but perhaps I'm missing something. In any 
> case, what the reader really wants is a precise explanation of the 
> exact model in use here, not a lot of text about the well-understood 
> concept of conditional likelihoods on branches of a tree.

We have reorganized the presentation of the "composition algorithm" to
attempt to clarify both what it is and why it is important.  The
reviewer is completely correct that it is, indeed, a prescription
for chaining together conditional models.  However, while the concept
of so chaining together models is intuitive, in practice it can be a
daunting task for SCFGs thanks to the presence of bifurcations.  For
example, while both the model generating a prior distribution over ancestral 
sequences (the "singlet transducer," with 7 states) and the model describing
evolution along a single branch (the "branch transducer," with 21 states)
are simple and compact, the three-taxon ensemble model has 230 states
and 1,789 transitions.

In summary, while we completely agree that the model construction procedure
is no more than a formalization of an intuitive concept, we believe
that it is nonetheless substantial, due to the sheer complexity
of the ensemble models so created.  The "TKFST" model which we use
here is relatively simple; more realistic models will give rise to even
more complex ensemble models.

> Perhaps most glaringly, the section entitled "dynamic programming 
> inference algorithms" never actually states an algorithm! It should. 
> The "loopy DP" algorithm is claimed in the abstract (and in the 
> authors' response to reviews) to be a key result of the paper, but it 
> is only referred to ("See Text S1 for more [sic] details of this 
> algorithm"). Again this is a misuse of supplementary material. A paper 
> is not supposed to be a mere advertisement of cool things we're 
> working on, and cooler things we're thinking about working on. Too 
> much of this manuscript still reads this way (though it's greatly 
> improved from the previous version). A paper should deliver crisp, 
> hard-nosed, reproducible results to a reader. In the case of an 
> algorithms paper, the reader should be able to understand and 
> implement the key ideas from a careful and complete reading of the 
> paper. Generally speaking, the authors are still spending way too much 
> time talking about ideas for their future, and too little
> teaching the reader something concrete in the present. Relegating 
> important stuff to supplementary material while spending so much room 
> on vague conceptual stuff in the text is aggravating (I would be 
> sympathetic if the manuscript were closely on point, and stuff was 
> being moved to supp material solely to keep length down and maintain 
> clarity of the essential results description). Supplementary material 
> should be used for stuff that doesn't fit (large data sets, source 
> code), is better distributed in electronic form (video, formatted 
> data) or is truly parenthetical to the main results (details that the 
> reader doesn't need to see to understand the paper). Only in a last 
> resort (in space-limited journals, which does not include PLoS 
> Computational Biology) should it contain any actual results of the paper.

Thank you for the feedback; we have attempted to revise the paper
accordingly such that it is completely self-contained and
reproducible.  We have given complete specifications of the CYK,
Inside, Outside, and CYK traceback algorithms for three sequences,
whether completely unconstrained, or constrained by a pre-determined
set of folds for each sequence ("fold envelopes").  They are written
in pseudocode which corresponds closely to our actual implementation
in the program (INDIEGRAM) which we describe.

We have entirely replaced the "loopy DP" algorithm with an algorithm
for exactly eliminating null cycles from general SCFGs.  Per the
reviewer's recommendation, the full description of this algorithm is
described in the main text ("Eliminating null cycles in SCFGs") of the
revised manuscript.  This new algorithm avoids the increased time
complexity associated with the iterative inference of "loopy DP."

> It's possible that I would come to adequately understand these 
> algorithms if I spent a lot more time thinking about them and studying 
> the "Supplementary" material. As always when refereeing, I feel I need 
> to allow for the possibility that I'm slower than the average reader. 
> So I'd actually be willing to forgo much of the above criticism if I 
> could skip ahead to the application and results and be impressed with 
> the utility of the approach, so that I'd feel motivated to go back and 
> work even harder to understand it. But the figures and results here 
> really do not inspire much confidence either.
>
> The key results figure is Figure 2, which shows the "reconstruction" 
> of "ancestral" sequences for three nanos TCEs and three tRNAs on star 
> topologies. The problems I have with this figure are manyfold, including:
>
> - In both cases the "ancestor" is in fact related by negligible branch 
> length to one of the observed sequences. But of course this 
> immediately makes the inference problem trivial, completely 
> uninteresting, and definitively *not* "ancient reconstruction". In 
> this situation, one can simply infer that the "ancestral" sequence is 
> identical to this closely related observed sequence. As noted in the 
> figure legend, that's almost what happens in the tRNA example (the 
> "ancestor" is nearly identical to AC008670.6). The figure legend fails 
> to note it also happens with the nanos TCEs, where the "ancestor" is 
> in fact completely identical to M72421.1. Even for a toy problem, come 
> on; this is way too unchallenging and toylike.
>
> - Even so, it looks like the inference algorithm is actually screwing 
> up the inference of the ancestor, moving in the wrong direction from 
> simply assuming identity with the near-zero-branch-length related 
> sequence. How does the algorithm possibly infer an ancestral G in a 
> column where it sees two gaps and an A (3' most base of the first stem 
> loop of tRNA)? How does it infer an ancestral gap in a column where it 
> sees two G's and an A and no gaps (the multifurcated base between the 
> D arm and the anticodon arm)? The authors need to explain how these 
> could possibly be right; they seem like red flags that the algorithm 
> isn't working properly.
>
> - This isn't an experiment. Showing that an algorithm produces a 
> result without crashing doesn't say much about whether the result 
> should be relied upon. If the purpose of these algorithms is to 
> accurately reconstruct ancestral RNAs, the authors need to design and 
> perform experiments to evaluate the accuracy of their reconstructions, 
> even if (for computational complexity reasons) these experiments can 
> only be done on a small, noncomprehensive set of anecdotes. I would 
> even settle for the normal positive control "sanity check" experiment 
> of simulating data under exactly their model, then verifying that 
> their model can reconstruct the appropriate likelihoods and inferences.

We agree that these reconstructions were not very informative.
Unfortunately these were the only sequences that we could find which
were short enough to be tractable on the hardware we had available. We
now report simulation results instead, as suggested.

We thank the reviewer for pointing out errors in the original
figures. These arose during manual conversion of indiegram's output
parse trees into Stockholm-format alignments, and due to a bug in
xrate's reconstruction code. We have now automated this process
(i.e. indiegram now outputs Stockholm format) and fixed the bug in
xrate, verifying that this has eliminated the errors (e.g. see the
simulation results).

We removed the tRNA and nanos reconstructions and posted (corrected)
versions online, at the following URL (which is linked from the
indiegram webpage):

http://biowiki.org/IndiegramExamples

We hope that these may still be useful for illustrative purposes for
anyone trying to use the software, but we have removed references from
the paper (since, as the reviewer notes, the short branch lengths make
the answers somewhat obvious).

The paper still includes the (experimentally-based) runtime estimates for
different length classes of RNA genes.

> More minor comments follow.
>
> title: A paper should deliver something akin to what its title seems 
> to promise. There is no result in this paper that bears meaningfully 
> on "primordial ribozymes" or "ancient RNA". The main results are 
> instead a promising but rough and very preliminary theoretical 
> framework for modeling the evolution of structural RNAs, with some 
> small toy examples of applications to three small RNA sequences. The 
> grand idea of paleogenetics motivates the work, and certainly should 
> be discussed, but it doesn't belong in the title unless you're 
> actually doing (and testing!) paleogenetics. A less grand title, more 
> consistent with results on toy three-sequence data sets, is appropriate.

We have accordingly changed the title to "Evolutionary models of
structured RNA".

> p.3: as far as I'm aware, the Felsenstein "pruning algorithm" and 
> "peeling algorithm" are just alternative names for the same algorithm 
> for recursively computing the likelihood of observed sequences related 
> by a tree, so I don't understand the distinctions being drawn in the 
> introduction between using the "pruning algorithm" to solve the 
> likelihood (problem "A") versus the "peeling algorithm" to sample from 
> the posterior (problem "C"). All four problems A-D are solved by the 
> same algorithm (or trivial postprocessing of its results) in the 
> ungapped, substitution-only case. See for example Felsenstein's book 
> "Inferring Phylogenies", p. 253, where he states the equivalence 
> between his pruning algorithm and Elston and Stewart's peeling algorithm.

We certainly agree that the peeling and pruning algorithms are
equivalent for the case of reversible independent-sites models, in the
following sense.  For a reversible model, the posterior distribution
of the state at any ancestral node can be obtained by re-rooting the
tree at that node and then performing the pruning algorithm,
traversing the tree post-order (terminating at the chosen root).
However, to then compute the posterior distribution of the state at a
*different* ancestral node would require re-rooting the tree at that
node and recomputing the entire DP matrix (or at least the part
corresponding to the branches whose direction had been reversed by the
re-rooting operation).  In contrast, the peeling algorithm yields all
such posterior probability distributions in a single traversal, so it
is more efficient (and distinct from pruning).

More generally, the two algorithms are distinct for the case of
irreversible models. More precisely, pruning is a restricted case of
peeling which only gives probabilities at the root node; since
irreversible models do not permit re-rooting of the tree, pruning
cannot (by itself) be used to provide posterior distributions for
internal nodes (only the root node).

We drew the distinction primarily to illustrate the different
applications for which the algorithms are frequently used (e.g., "A"
vs. "C"), but we can certainly revise this text further if the
reviewer feels that it is appropriate to do so.

> p.6 bottom: pair HMMs are not really "standard"; they are not widely 
> used or known. (There is still too much of a general tendency in the 
> manuscript to use technically narrow jargon in place of actual 
> explanation, and assume the reader will be able to fill in the gaps. 
> Sometimes, though, the revision gets it right: for example, other than 
> this "standard pair HMM" reference, the authors do an excellent job on 
> pp6-7 introducing the concept and motivation of conditional 
> probability models first, and *then* connecting to the linguistics 
> literature and the jargon term "parse-tree transducer".)

Thank you for pointing out this oversight.  We briefly describe what a
Pair HMM is in the revised manuscript.

> p.7 why is the model that generates P(X) at the root a "singlet 
> transducer"? It's not transducing anything (it's not conditionally 
> dependent on an input parse tree).

We chose to call it a "singlet" transducer so that we used similar
terminology for the model generating the marginal and conditional
distributions.  A singlet transducer can be viewed as a special case
of a transducer, where the input being transduced is empty.  Indeed,
by "singlet transducer" we mean the same thing as is often referred to
as a "single-sequence SCFG".  We apologize if it is poor terminology;
we can change it if the reviewer feels that doing so would help
clarify the methods.

> p.13 The DP inference algorithm section says "the standard CYK ... 
> algorithms fail". But then it says "Maximum likelihood inference is 
> performed with a CYK algorithm", and even "maximum likelihood 
> inference can therefore be performed with a standard three sequence 
> CYK algorithm". Which is it?

Thank you for pointing out this inconsistency, which we have clarified
in the revised text.  Using the standard CYK algorithm on the composed
model fails; however, the model can be simply modified as described in
the revised text to permit standard CYK to work properly.

> p.14 It seems possible that the problem with "null cycles" (residues 
> in ancestors that leave no homologous residues in observed sequences) 
> may simply go away if you view the problem as a missing data problem, 
> and that the likelihood being inferred is the likelihood of alignment 
> columns needed to account for the observed data (marginalized over 
> missing columns needed to account for null cycles) rather than the 
> joint likelihood of the "observed" columns and the "null cycle" 
> columns. This is what dropped out of the equations in the phylogenetic 
> inference described by Rivas and Eddy (2008) PLoS CB 4:e1000172, for 
> example.

We completely agree and have included an extended discussion of this in
the revised text.  The primary difficulty is that while this marginalization
is reasonably straightforward for independent-sites models and HMMs,
the possible presence of null bifurcations complicates matters for SCFGs.

> p.15-18 The whole section on "terminological asides" contributes 
> little to the paper. It would be vastly preferable to spend this space 
> describing what you're actually doing, as opposed to making vague 
> conceptual connections. Save the conceptual connections for a review 
> article someday.

We have accordingly stricken this section from the paper.

> p.19 bottom: "The results of these results, reported in Text S1...": 
> again, results in the paper should not be relegated to supplementary 
> material. A test that we use here is that you should be able to make 
> no references at all to Supplementary Material in the text of a paper, 
> but just list it at the end, and the reader would still be happy. Good 
> is "oh look, the source code and computer-readable data files are 
> available too", and bad is "dammit, you mean I have to go to another 
> manuscript to see the result of the experiment they just described?"

Thank you for this suggestion.  We have moved those results back into
the main text of the paper and made them into a section discussing
evaluating the TKFST model by making it part of a multiple RNA
alignment program.

> p.21 "A TKFST-based reconstruction could in principle falsify 
> Hopfield's hypothesis": how, exactly? How could you falsify a 
> biological hypothesis with a model whose realism and accuracy you've 
> called into question?

[we have deleted the references to Hopfield's hypothesis]

> p.21 in general the whole "reconstructing RNAs with a simple model" 
> section is too chatty and imprecise. "Both predicted ancestral 
> structures closely match the RFAM-annotated consensus structures", for 
> example, is not a quantitative result, and was not set up to address 
> any particular focused question (like, for instance, does our method 
> do better than some simpler approach of addressing the same problem?) 
> Again too much space is devoted to vague forward-looking conceptual 
> claims ("Though it's beyond the scope of the paper, a revised pairwise 
> model could readily be designed...") and too little attention is given 
> to the actual model and problems at hand.

> Table 2: it's unclear why we need separate states L and I_L, or S and 
> I_S; these nonterminals have identical production rules.

[distinction between clearest representation of the grammar for didactic purposes, versus the best representation for implementation purposes; for implementation we used a Moore machine representation]

> Similarly, I 
> don't see why we need B_e; S and I_S can produce L directly; and L and 
> I_L can produce an LS bifurcation directly, rather than invoking 
> nonterminal B. The L and I_L productions don't appear to be correctly 
> normalized, unless the (undefined) term p_S is 0.0.

> A production I_L 
> \rightarrow x I_L does not correspond to a pair-HMM rule, as the 
> caption states; this is just a standard regular grammar production.

Thank you for pointing this out.  We have corrected this in the 
revised text.

> Table 4: perhaps I'm slow, but this doesn't seem like the clearest 
> explanation of a branch transducer. I'm expecting to see rules 
> corresponding to, for example, the ancestor produced G-C from a 
> basepair state; the branch transducer chooses a basepair->basepair 
> substitution transduction as its state path with some probability, 
> then eats the GC and emits an xy. The production rules in Table 4 look 
> more like a single-sequence SCFG to me (x,y do appear in the wait 
> state conditionals, but where are x,y coming from on the left hand 
> side -- where are they being absorbed from?)

> The productions also 
> don't seem to correspond entirely to the state definitions in Table 5; 
> for example, if M_S is a "match" state describing "base pair 
> substitution" that "absorbs (x,y)" and "emits (u,v)" according to 
> Table 5, then why does the Table 4 give no production rule for 
> emitting (u,v) given (x,y)? Instead it's the wait state W_S in Table 4 
> that has the expected (x,y) -> (u,v) production rule. Maybe the
> rationale would become clear to me if I spend more time staring at 
> this, but it seems to need more explanation and/or clarification and 
> cleanup.

We wrote the rules here in the Mealy-machine framework, where inputs
and outputs are attached to transitions (c.f., the Moore-machine framework,
where ouputs depend only on the current state of the machine).  The
rule
W_S -> u M_s v ;       alpha_2 (t) * p (u, v| x, y)
is intended to correspond to a transition in a Mealy machine
from state W_S to M_S, where the transition has an input of (x, y)
and an output of (u, v).  We have attempted to clarify this in the
revised text.

We apologize for the confusion; we have attempted to follow the
standard SCFG notation as closely as possible in an attempt to be
clear, but we certainly welcome further suggestions to clarify
the model.

> Figure 1: 5' and 3' ends of RNA structures should be labeled.

> Figure 3: It is very unclear what this figure is supposed to be 
> representing (and I had the same problem with it when it appeared in 
> reference [28]). It appears to simply be a representation of a parse 
> tree of an individual RNA structure. It does not seem to show any kind 
> of evolutionary model from one sequence/structure/parse tree to 
> another, as the caption and text would seem to imply.

[we have included stree-mutations.pdf, which was also in the cited paper, and illustrates the dynamic behavior of this model]

> Figure 4 and Figure 5 seem unnecessary. Both are too schematic to 
> convey useful information. The concept of conditional likelihood along 
> branches of a tree is obvious and well understood to your readers in 
> the substitution-only framework of the Felsenstein algorithm. The 
> authors should focus these figures like this on their *particular* 
> model for the transducers, and exactly how they accomodate insertion 
> and deletion, rather than the obvious concept of "P(X | Y,t)".

We felt that it was better to include information at the risk of
stating the obvious, rather than risk omitting necessary
information. The figure illustrating "P(X | Y,t)" may be obvious but
the figure illustrating the following distribution...

P(W | root_distribution) P(X | W,T_X) P(Y | W,T_Y) P(Z | W,T_Z)

...is, we feel, not as obvious; and its simplicity may help clarify
the more complicated concepts about transducer composition that appear
in the text. Our hope is that since this is listed as a minor
criticism, we may accept it as a stylistic difference.

> The figures are referred to out of order; I try not to succumb to the 
> hobgoblins of foolish consistencies, but it's pretty egregious here (I 
> think Figure 1 isn't referred to until p.21; whereas I think Figure 4 
> is mentioned first, on p.9.) Similarly, the bibliography seems 
> excessively typo-laden. (Ref 25 lists the author as "J P". Ref 28 
> gives no page numbers. Ref 33 shows no volume nor page numbers. And so 
> on.)

[we have corrected this..... I hope]

### 2009-04-18 cover-letter-resubmission-1.tex
Import from CVS.

\documentclass[10pt]{letter}
\pagestyle{plain}
%\usepackage[ansinew]{inputenc}

%\addtolength{\topmargin}{-0.5in}

%\sloppy
%\nonfrenchspacing

\renewcommand{\baselinestretch}{1.0}

\address{
Ian Holmes\\
Department of Bioengineering\\
University of California\\
Berkeley, CA 94720\\
}

\signature{}
\begin{document}
\begin{letter}{}

\opening{Dear Prof. Stormo:}

\vspace{\baselineskip}

Thank you very much for your attention to our earlier manuscript and for forwarding the extremely encouraging reviews.

Based on the insightful comments of the reviewers, we have made sweeping changes to the structure and presentation of the manuscript.
We believe that it is greatly improved as a result.

We would be most grateful if you would consider the accompanying revised manuscript, along with the attached point-by-point response to the reviewers' earlier criticisms,
describing the changes we have made.

All authors have read and approved of this manuscript.

%As editor, we suggest David Haussler.

%We suggest the following reviewers:
%David Haussler (if not editor),
%Jeff Thorne,
%Marc Suchard,
%Jotun Hein,
%Sean Eddy.
%We have provided contact details for these reviewers in the online
%submission process.

\closing{Sincerely,\vskip 20pt Robert K. Bradley\\Ian Holmes}

%% If you have things to enclose, specify here
%\encl{Encl. 1}
%\encl{Encl. 2}

\end{letter}

\end{document}

### 2009-04-18 cover-letter.tex
Import from CVS.

\documentclass[10pt]{letter}
\pagestyle{plain}
%\usepackage[ansinew]{inputenc}

%\addtolength{\topmargin}{-0.5in}

%\sloppy
%\nonfrenchspacing

\renewcommand{\baselinestretch}{1.0}

\address{
Ian Holmes\\
Department of Bioengineering\\
University of California\\
Berkeley, CA 94720\\
}

\signature{}
\begin{document}
\begin{letter}{}

\opening{Dear Editors:}

\vspace{\baselineskip}

The new science of {\bf Ancestral Sequence Reconstruction} is bringing evolutionary origins within experimental reach.
This field has come of age in the past five years.
Many ancient protein sequences have been reconstructed and experimentally tested
(Benner {\em et al}, Nature 2003;
Thornton {\em et al}, Science 2007;
Liberles (ed.), Oxford University Press 2007).
There have also been major attempts at the
computational reconstruction of genomes
(Blanchette, Green, Miller \& Haussler, Genome Research 2004;
Paten {\em et al}, Genome Research, in press).

Here, we present initial steps towards a theory, tool-set and benchmark for phylogenetic reconstruction of structured RNA,
such as ribosomal RNA.
This year is the fortieth anniversary of the {\bf RNA world hypothesis}
(Woese, Harper \& Row 1968).
It's also the 40th anniversary of the all-RNA primordial ribosome
(Crick, JMB 1968),
a structural model of which was recently proposed
(Smith, Gutell {\em et al}, Biology Direct, 2008).
Yet to our knowledge, the attached manuscript represents the first broad pilot study of concrete algorithms for reconstructing RNA.

Our paper describes several new software tools, all backed by simulation and/or benchmarking results.
However, we feel that the key contribution is a complete, self-consistent theoretical framework for modeling (and reconstructing) RNA evolution on a phylogenetic tree.
This theory is the RNA analogue of the {\bf transducer theory} for proteins and DNA
(Bradley and Holmes, Bioinformatics 2007),
which was recently used in the first whole-genome mammalian-ancestor reconstruction
(Paten, Herrero, Birney, Holmes \& Flicek, Genome Research, in press).
The theory is closely related to that of ``Statistical Alignment''
(Wong, Suchard and Huelsenbeck, Science 2008)
and sequence automata for reconstructing indel histories
(L\"{o}ytynoja \& Goldman, Science 2008).
Unlike those previous theories (which model only indels and point substitutions),
our model includes covariant base-pair substitutions and changes in RNA structure.

\newpage
All authors have read and approved of this manuscript.

%As editor, we suggest David Haussler.

%We suggest the following reviewers:
%David Haussler (if not editor),
%Jeff Thorne,
%Marc Suchard,
%Jotun Hein,
%Sean Eddy.
%We have provided contact details for these reviewers in the online
%submission process.

\closing{Sincerely,\vskip 20pt Robert K. Bradley\\Ian Holmes}

%% If you have things to enclose, specify here
%\encl{Encl. 1}
%\encl{Encl. 2}

\end{letter}

\end{document}

### 2009-04-18 defs.tex
Import from CVS.

\usepackage{setspace}
\doublespacing

\input{parsetree.sty}

\usepackage{array}

\usepackage[boxed]{algorithm2e}
\usepackage{mathrsfs}

% Labels & references for sections, figures and tables
% Commented out \secref definition & replaced \seclabel with a dummy, because PLoS doesn't like numbered section references -- IH, 11/11/08
\newcommand{\secref}[1]{Section~\ref{sec:#1}}
\newcommand{\seclabel}[1]{\label{sec:#1}}

\newcommand{\appref}[1]{Appendix~\ref{app:#1}}
\newcommand{\applabel}[1]{\label{app:#1}}

\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
\newcommand{\figlabel}[1]{\label{fig:#1}}

\newcommand{\tabnum}[1]{\ref{tab:#1}}
\newcommand{\tabref}[1]{Table~\tabnum{#1}}
\newcommand{\tablabel}[1]{\label{tab:#1}}

\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
\newcommand{\alglabel}[1]{\label{alg:#1}}

\newcommand{\eqnref}[1]{Equation~\ref{eqn:#1}}
\newcommand{\eqnlabel}[1]{\label{eqn:#1}}

% Frequently used proper nouns that should have consistent typeface, capitalization, etc.
\newcommand{\pfold}{PFOLD}
\newcommand\darturl{{\tt http://biowiki.org/dart}}
\newcommand\indiegramurl{{\tt http://biowiki.org/IndieGram}}
\newcommand{\dart}{DART}

\newcommand{\bralibaseII}{BRalibaseII}
\newcommand{\balibase}{BAliBASE}

\newcommand{\stemloc}{STEMLOC}
\newcommand{\xrate}{XRATE}
\newcommand{\evoldoer}{EVOLDOER}
\newcommand{\indiegram}{INDIEGRAM}

\newcommand{\contrafold}{CONTRAFOLD}
\newcommand{\stemlocama}{STEMLOC-AMA}
\newcommand{\consan}{CONSAN}
\newcommand{\clustalw}{ClustalW}
\newcommand{\foldalign}{Foldalign}
\newcommand{\foldalignm}{FoldalignM}
\newcommand{\mastr}{MASTR}
\newcommand{\rnasampler}{RNASampler}
\newcommand{\murlet}{Murlet}
\newcommand{\rnaalifold}{RNAAlifold}
\newcommand{\viennarna}{ViennaRNA}

\newcommand{\RFAM}{RFAM}

\newcommand{\denovo}{{\em de novo}}

\newcommand{\nth}{\mathrm{n}^{\mathrm{th}}}
\newcommand{\mth}{\mathrm{m}^{\mathrm{th}}}

\newcommand{\type}[0]{\operatorname{type}}
\newcommand{\emit}[0]{\operatorname{emit}}
\newcommand{\absorb}[0]{\operatorname{absorb}}
\newcommand{\argmax}[0]{\operatorname{argmax}}
\newcommand{\argmin}[0]{\operatorname{argmin}}
\newcommand{\parent}[0]{\operatorname{parent}}

\newcommand{\weight}[0]{\operatorname{Weight}}
\newcommand{\transweight}[3]{t\left(#2,#3|\theta^{(#1)}\right)}  % {m}{from}{to}
\newcommand{\emissionweight}[3]{e_{#2}\left(#3|\theta^{(#1)}\right)}  % {m}{emitstate}{emission}
\newcommand{\totalemissionweight}[2]{\mathbb{e}_{#1}(#2)}

%\newcommand{\cyk}[7]{\gamma_{#1}(#2,#3,#4,#5,#6,#7)}
\newcommand{\cyk}[7]{\gamma_{#1}\left(#2,#3,#4\right)}
%\newcommand{\cyk}[7]{\gamma_{#1}(#2,#3,#4)}
%\newcommand{\inside}[7]{\alpha_{#1}(#2,#3,#4,#5,#6,#7)}
\newcommand{\inside}[7]{\alpha_{#1}\left(#2,#3,#4\right)}
%\newcommand{\outside}[7]{\beta_{#1}(#2,#3,#4,#5,#6,#7)}
\newcommand{\outside}[7]{\beta_{#1}\left(#2,#3\right,#4\right)}

\newcommand{\Tnull}{\mathrm{null}}
\newcommand{\Tgap}{\mathrm{GAP}}

\newcommand{\statetype}[1]{\mathtt{#1}}
\newcommand{\state}[3]{{}^{#2}\mathtt{#1}^{#3}}
% Note that this will only work for left-bifurcations!
%\newcommand{\bifurc}[3]{{\phantom{]}}^{#2}\mathtt{B[#3\,#1]}}}
\newcommand{\bifurc}[3]{\mathtt{#1[#2\,#3]}}

% state types for branch transducers
\newcommand{\Sstart}[0]{\statetype{Start}}
\newcommand{\Smatch}[0]{\statetype{Match}}
\newcommand{\Sinsert}[0]{\statetype{Insert}}
\newcommand{\Swait}[0]{\statetype{Wait}}
\newcommand{\Send}[0]{\statetype{End}}

% state types for evolutionary model
\newcommand{\Snull}[0]{\statetype{Null}}
\newcommand{\Sbifurc}[0]{\statetype{Bifurcation}}
\newcommand{\Semit}[0]{\statetype{Emit}}

\newcommand{\statevec}[1]{\left(\begin{array}{c}#1_1\\\vdots\\#1_N\end{array}\right)}
\newcommand{\fourvec}[4]{\left(\begin{array}{c}#1\\#2\\#3\\#4\end{array}\right)}
\newcommand{\nakedfourvec}[4]{\begin{array}{c}#1\\#2\\#3\\#4\end{array}}
\newcommand{\threevec}[3]{\left(\begin{array}{c}#1\\#2\\#3\end{array}\right)}
\newcommand{\nakedthreevec}[3]{\begin{array}{c}#1\\#2\\#3\end{array}}
\newcommand{\bvec}[1]{\boldsymbol{#1}}
\newcommand{\activenode}[1]{n(\bvec{#1})}

\newcommand{\foldenv}{\mathscr{F}}

\newcommand{\mDn}[0]{m \vartriangleright n}
\newcommand{\mNDn}[0]{m \ntriangleright n}
\newcommand{\mDEn}[0]{m \trianglerighteq n}
\newcommand{\mNDEn}[0]{m \ntrianglerighteq n}

\newcommand{\beqn}{\begin{equation}}
\newcommand{\eeqn}{\end{equation}}

\newcommand\gnons{{\cal N}}
\newcommand\gnulls{{\bf N}}
\newcommand\gbifs{{\bf B}}
\newcommand\gemits{{\bf E}}
\newcommand\gterms{\Omega}
\newcommand\grules{{\bf R}}
\newcommand\grhs{( \gnons \cup \Omega )^\ast}
\newcommand\gtrees{{\cal T}}

\newcommand\leftside[1]{{\cal L}(#1)}
\newcommand\rightside[1]{{\cal R}(#1)}
\newcommand\allrules[1]{{\cal A}(#1)}

\newcommand\revisionnotes
{\noindent \hrulefill{} \\ {\bf Revision Notes:}
\begin{enumerate}
\item These comments are not a complete summary of the review, just a partial strategy for addressing some of the points.
The review is the more authoritative guide...
He is giving us the benefit of the doubt but acceptance is still very much in the balance, I think, so we need to stick at it,
and take it that extra step beyond minimal compliance with his suggestions.
%
\item Obviously we have to do simulation tests and hopefully EvolSayer will be of some use there.
One thing to be careful of is that there are certain regimes of rate/branch-length parameter space where the ``trivial'' outcomes identified by the reviewer
(ancestral sequence is always identical to one of the descendants) are the norm.
Oscar and I have been looking at this quite a bit recently (see also 2003 paper by Mossel).
The regime we are probably most interested in is where the branch lengths are of comparable length
and so the two slightly-longer branches can ``out-vote'' the shorter branch under certain conditions.
This should probably be commented on in the paper: the reviewer says he would be happy with ``a small, noncomprehensive set of anecdotes''
but if we can rationalize why some parameter settings are more interesting than others, then that's all to the better.
%
\item I would actually be very wary of taking that ``small noncomprehensive set of anecdotes'' at face value.
This is one area where I sense we're being given enough rope to hang ourselves... an unsatisfactory response to the simulation issue will be unequivocal grounds for rejection.
Obviously there are limits to the number of simulations we can run, but with EvolSayer's output statistics (allowing us to filter out ``interesting'' cases), we can automate it somewhat.
%
\item I wrote out the TKF structure tree for the case $a \to b \to c$,
which can be used to estimate the intermediate sequence $b$
(e.g. for Redelings-Suchard).
I think this could be useful in explaining tree transducer composition more concretely, and why is important:
by inspecting these tables, the reader can see that the structure of this grammar is quite complicated,
and should then be prepared for the idea that the ``composition algorithm'' is simply generating this table
(or other tables for other phylogenies).
The latex for this is in the file ``tkfst.tex'', and is currently include'd at the start of the supplementary information (it should go in the main paper though, I think).
%
\item Can we dump out the grammar for the three-branch TKFST and include it in the Supplement?
(Or the main paper?) I realize it's huge, but this very hugeness does motivate the transducer composition algorithm,
in the same way that the two-branch model (which is now included in a more human-readable form) does.
%
\item If we are moving the description of loopy DP to the main paper, I think it needs to be more precise.
Currently it just says ``we repeatedly iterate over strongly-connected components of the state sub graph'',
which falls short of the level of precision/detail with which the other DP algorithms are presented.
The description should encompass alternate strategies (i.e. orders of visiting the states) that you have considered:
randomized, Tarjan's algorithm, etc.
The reader should be left understanding the pros-and-cons of each,
 e.g. worst-case performance or difficulty of implementation.
No need to agonize about this, but be helpful to potential future developers.
  (Example reader: a grad student in our group who might try to implement loopy DP themselves.)
%
\item The descriptions of the DP algorithms do not currently include any traceback algorithms (CYK or Inside);
therefore, we haven't technically described how to do ancestral reconstruction!
Obviously these algorithms are almost trivial, but they're quite important.
Implementing stochastic loopy-Inside traceback in tripletscfgdp.\{h,cc\} is a key step in transforming it into a useful tool
(for example, with this extension, Indiegram would be pretty much ready-to-go for Redelings-Suchard sampling).
The deterministic CYK traceback algorithm is {\em even more} important to present here, because implementing this algorithm is currently our main result!
The paper should describe {\em both} these traceback algorithms.
This shouldn't be that hard to do in a comprehensible, rigorous style
 (loopy traceback is much simpler than the algorithms for restoring eliminated states to traceback paths, which are painfully described in my 2003 paper).
%
\item All algorithms should be described in full, rather than giving recipes for deriving some algorithms from others.
In particular, the CYK algorithm is currently summarized by statements along the lines of
``replace sums over paths through the model (or equivalently, parses) with the $\max()$ operation (e.g., in equation X, sum-over-complicated-expression will be replaced with max-over-complicated-expression)''.
It shouldn't be too hard to write a latex command/macro that can be used (with ``$\sum$'' or ``$\max$'' as an argument) to generate complete versions of {\em both} algorithms.
%
\item The paper the reviewer cites as an example of the simplicity of missing data estimation (Rivas and Eddy (2008) PLoS CB 4:e1000172) is an IID-columns model for a multiple alignment.
When he says that this issue ``may simply go away if you view the problem as a missing data problem'',
that is precisely what we are talking about when we discuss null cycle elimination.
I've tried to explain in ``tkfst.tex'' why this is harder than it superficially appears:
\begin{enumerate}
\item eliminating missing ``empty'' columns from an IID-columns model (such as in Rivas and Eddy, 2008) is easy.
\item eliminating missing emissions from an HMM is slightly harder (c.f. ``wing retraction'' algorithm in HMMer source code)
but general algorithms do exist.
\item eliminating null states from an SCFG is much harder, due to null bifurcations.
\end{enumerate}
%
\item The reviewer asks for early discussion of constraints in the context of the ``rate-limiting steps of paleogenetics''.
We should take this opportunity to clarify that our approach is (1) formulate the model,  (2) apply appropriate constraints.
Our emphasis here is on (1).
Cite the literature: QRNA represents an alignment constraint on a pair SCFG, Holmes-Rubin 2002 uses fold constraints on pair SCFGs,
stemloc uses fold+alignment constraints, Ortheus uses an alignment constraint, etc, but none of these things can be done
(in a probabilistic modeling framework at least) until you can formulate the model.
For this paper, we chose to condition on structures essentially because they are easy to constrain
(fold envelopes can be applied independently to each sequence).
But we could equally easily have conditioned on alignments. Constraints are easy to tweak, once you have a model.
%
\item The reviewer also comments on the phylogeny being a given.
We can note that even though we constrain the phylogeny, more realistic phylogenetic applications are possible.
Given a structural phylo-alignment (tree+alignment+structure), we can calculate the likelihood; therefore, given two such structural phylo-alignments,
we can say which one is more probable (describe their relative posterior probabilities, etc.).
From this it is possible to make a crude MCMC sampler, just by proposing/accepting/rejecting alignment/tree/structure changes.
An {\em efficient} MCMC sampler needs to be able to make local topology changes and to resample the multiple alignment around a topology change;
we can do this too, in principle, but it's beyond this paper.
%
\item As I understand it, pruning \& peeling are not exactly equivalent as the reviewer claims, {\em except} in the special case of a reversible model.
Specifically, pruning is a {\em restricted case} of peeling, giving you the posterior probabilities at the root node, but not any other node.
Of course, if you have a reversible model, you can re-root the tree anywhere, and so the two are equivalent for reversible models only.
When Felsenstein says the two approaches are the same (on p253 of his book), it is quite a vague statement:
e.g. later in the same paragraph, he says that both peeling and pruning are variants of Horner's rule, and particular cases of dynamic programming.
All of this is somewhat academic; basically I'm thinking the pruning/peeling distinction might yet have some pedagogic value, and we might not need to abandon it entirely.
%
\item We should address all the reviewer's comments about the figures if possible.
I have added xfig files ``stree.fig'' and ``stree-mutations.fig'' to the repository; the latter shows an evolutionary trajectory of the TKFST, as requested by the reviewer.
%
\item Finally, a very minor/trivial point: would you consider a name-change for IndieGram to something beginning with ``Evol''?
The EvolDeeds package is EvolDoer (previously released as ``tkfstalign''), EvolSayer and IndieGram.
I put a few random suggestions for Evol-names at {\bf biowiki.org/EvolDeeds}.
Don't take this too seriously, of course!
I'll understand if you want to stick with IndieGram, and this is obviously negligible compared to the other revisions.
%
\end{enumerate}
\hrulefill{}}

### 2009-04-18 nullcycle.tex
Import from CVS.

\section{Null cycle elimination from SCFGs}

\newcommand\gnons{{\cal N}}
\newcommand\gnulls{{\bf N}}
\newcommand\gbifs{{\bf B}}
\newcommand\gemits{{\bf E}}
\newcommand\gterms{\Omega}
\newcommand\grules{{\bf R}}
\newcommand\grhs{( \gnons \cup \Omega )^\ast}
\newcommand\gtrees{{\cal T}}

\newcommand\leftside[1]{{\cal L}(#1)}
\newcommand\rightside[1]{{\cal R}(#1)}
\newcommand\allrules[1]{{\cal A}(#1)}

Null cycles can be removed from SCFGs by (i) separating bifurcations into
those which have one or more empty children (and can therefore be represented using transition or termination rules)
and those that have two nonempty children; and then (ii) encapsulating all paths that do not emit any terminal symbols
(and may therefore cause cyclic dependencies in the DP matrix) as effective direct transitions, with probabilities obtained by geometric series summation of the grammar's transition matrix.

This section formally describes the null-cycle elimination procedure, along with procedures for probabilistically ``restoring'' null cycles to sampled parse trees
and Inside-Outside expectation counts.

\subsection{Definitions}

Let $G = ( \Omega, \gnons, \grules, P )$ be a stochastic context-free grammar (SCFG)
consisting of
a set of terminal symbols $\Omega$,
a set of nonterminal symbols (a.k.a. ``states'') $\gnons$,
a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
and a non-negative weight function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.

(The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
The interpretation is that $P$ is a probabilistic weight function, and $\grules$ is informally the set of rules with non-zero weight.)

Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e. rules in which $A$ appears on the left),
and let $\rightside{B}$ be the set of rules that can generate nonterminal $A$ (i.e. rules in which $A$ appears on the right).
Let $\allrules{B} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
Define these also on sets of nonterminals, e.g. $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.

Let $\gtrees$ be the set of all parse trees for $G$.
Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
then define the parse tree weight $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.

If $\sum_{T \in \gtrees} P(T) = 1$, we say that $G$ is {\em probabilistically normalized by parse tree}.
If $\sum_{\rho \in \leftside{A}} P(\rho) = 1$ for all $A \in \gnons$, we say that $G$ is {\em probabilistically normalized by production rule}.
Note that normalization by production rule $\Rightarrow$ normalization by parse tree.

Let $S \in \Omega^\ast$ denote a terminal sequence.
Let $\mathrm{seq}:\gtrees \to \Omega^\ast$ be the function mapping a parse tree to its terminal sequence.
and let $\mathrm{root}:\gtrees \to \gnons$ be the function returning the root nonterminal of a parse tree.
The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
Of particular relevance to null state elimination are the probabilities $P(\epsilon|A)$,
where $\epsilon$ is the empty sequence.
These are the probabilities that a nonterminal will expire without generating any sequence.

Following earlier formalisms \citep{Durbin98,HolmesRubin2002a,Holmes2005}
we say that the grammar $G$ has {\em RNA normal-form rules} if
each rule in $\grules$ takes one of the following four forms

\noindent \hrulefill
\begin{description}
\item[Termination rules] with one nonterminal on the left and the empty string on the right
\begin{eqnarray*}
A & \to & \epsilon
\end{eqnarray*}
\item[Transition rules] with one nonterminal on the left and the right
\begin{eqnarray*}
A & \to & B
\end{eqnarray*}
\item[Bifurcation rules] with one nonterminal on the left and two on the right
\begin{eqnarray*}
A & \to & B\ C
\end{eqnarray*}
\item[Emission rules] with one nonterminal on the left and right, and at least one terminal on the right
\begin{eqnarray*}
A & \to & x\ B \\
A & \to & B\ y \\
A & \to & x\ B\ y
\end{eqnarray*}
\end{description}
\noindent \hrulefill

Here $A,B,C \in \gnons$ and $x,y \in \Omega$.

Furthermore, we say that the grammar $G$ has {\em RNA normal-form states} if
each nonterminal (state) $A \in \gnons$ takes one of the following forms

\noindent \hrulefill
\begin{description}
\item[Null states]: $\leftside{A}$ contains only transition and termination rules.
\item[Bifurcation states]: $\leftside{A}$ contains exactly one bifurcation rule, $A \to X\ Y$, where $X \neq Y$ and $X,Y$ are both null states.
\item[Emit states]: $\leftside{A}$ contains only emission rules. Further, $\leftside{A} = \rightside{B}$ for some null state $B$.
This null state $B$ is called $A$'s {\em post-emit} state.
\end{description}
\noindent \hrulefill

Define a {\em null cycle} to be a nonterminal $A$
and a sequence of rules $\rho_1, \rho_2 \ldots \rho_k$ that, when applied consecutively to $A$, leave $A$ unchanged;
that is, $\rho_k(\rho_{k-1} \ldots \rho_2(\rho_1(A))) = A$.
Define a {\em null subtree} to be a null cycle using at least one bifurcation rule.

Suppose that $G = ( \Omega, \gnons, \grules, P )$ and $G' = ( \Omega, \gnons', \grules', P' )$ are two grammars.
We say that $G$ and $G'$ are {\em equivalent in sequence} if there is a mapping between nonterminals, $f:\gnons \to \gnons'$,
such that $P(S|A) = P(S|f(A))$.
We say that $G$ and $G'$ are {\em equivalent in parse} if there is a mapping between parse trees, $g:\gtrees \to \gtrees'$,
such that $P'(T') = \sum_{T:T'=g(T)} P(T)$ and we can define
\[
P(T|T') = P(T|g(T)=T') = \frac{P(T)}{P'(T')}
\]

Note that equivalence in parse $\Rightarrow$ equivalence in sequence.

Suppose that a grammar $G$ contains null cycles.
We seek to transform $G$ into a grammar $G''$ that is equivalent in parse and sequence, but has no null cycles.
If $G$ is normalized by parse tree, then $G''$ will be too; but $G''$ is not (necessarily) normalized by production rule.

We do the transformation in two steps, $G \to G' \to G''$.
We also provide a stochastic ``null cycle restoration'' algorithm for sampling from $P(T|T'')$.

\subsection{Eliminate null bifurcations}

Note first that any SCFG can be transformed into an equivalent one with RNA normal-form states by adding null states.
Without loss of generality, we therefore consider SCFGs with RNA normal-form states.

Let $G = ( \Omega, \gnons, \grules, P )$ be such an SCFG.
Let $\gnulls \subseteq \gnons$ be the set of null states in $\gnons$, {\bf excluding} post-emit states.
Let $\gemits \subset \gnons$ be the set of post-emit states.
Let $\gbifs \subseteq \gnons$ be the set of bifurcation states.

We here define the grammar $G'$,
\[
G' = \left( \Omega, \gnons', \grules', P' \right)
\]
which is equivalent to $G$ in sequence but has no null subtrees.

The new states $\gnons'$ are defined as follows.
Start with $\gnons$; remove bifurcation states;
then introduce a new state $N'$ for every null $N \in \gnulls$ and three new states $B^0, B', B^2$ for every bifurcation $B \in \gbifs$.

\[
\gnons' = 
\left( \bigcup_{N \in \gnulls} \{ N' \} \right) \cup \left( \bigcup_{B \in \gbifs} \{ B^0, B', B^2 \} \right) \cup \gnons \setminus \gbifs
\]

The idea is that $N'$ is a null state that is equivalent to null state $N$ for non-empty sequence.
That is, for any terminal sequence $S$,

\[
P'(S|N') = \left\{ \begin{array}{ll} P(S|N) & \mbox{if $S \neq \epsilon$} \\ 0 & \mbox{if $S = \epsilon$} \end{array} \right.
\]

Similarly, $B'$ is a null state that is equivalent to bifurcation state $B$ for non-empty sequence,
while $B^2$ is a null state that is is equivalent to $B$ for parse trees where both children of $B$ are non-empty.

In contrast, $B^0$ is {\bf exactly} equivalent to $B$ in sequence.
However, null subtrees in $B^0$ are explicitly accounted for,
their probabilities factored into the transitions $B' \to L'$ and $B' \to R'$ and the termination $B^0 \to \epsilon$,
using the inside probabilities for empty sequences, $P(\epsilon|L)$ and $P(\epsilon|R)$.

The probabilities $P(\epsilon|X)$ are related by the following system of equations

\[
P(\epsilon|X) = P(X \to \epsilon) + \sum_Y P(X \to Y) P(\epsilon|Y) + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R)
\]

which are nonlinearly coupled but may be solved numerically, e.g. by the Newton-Raphson method,
or by iterated approximation starting from a lower bound $P(\epsilon|X) \geq 0$.

The new rules and their probabilities are
\[
\begin{array}{rrcl}
\forall \rho \in (\grules \setminus \allrules{\gbifs}): & P'(\rho)             & = & P(\rho) \\
                                \forall N \in \gnulls: & P'(N' \to \epsilon)  & = & 0  \\
                          \ldots \forall B \in \gbifs: & P'(N' \to B')        & = & P(N \to B) \\
                         \ldots \forall M \in \gnulls: & P'(N' \to M')        & = & P(N \to M) \\
                         \ldots \forall E \in \gemits: & P'(N' \to E)         & = & P(N \to E) \\
               \forall (A \to B) \in \rightside{\gbifs}: & P'(A \to B^0)        & = & P(A \to B) \\
              \forall (B \to LR) \in \leftside{\gbifs}: & P'(B^0 \to \epsilon) & = & P(B \to LR)\ P(\epsilon | L)\ P(\epsilon | R) \\
                                                       & P'(B^0 \to B')       & = & 1 \\
                                                       & P'(B' \to \epsilon)  & = & 0 \\
                                                       & P'(B' \to L')        & = & P(B \to LR)\ P(\epsilon | R) \\
                                                       & P'(B' \to R')        & = & P(B \to LR)\ P(\epsilon | L) \\
                                                       & P'(B' \to B^2)       & = & 1 \\
                                                       & P'(B^2 \to L'\ R')   & = & P(B \to LR) \\
\end{array}
\]

Note that grammar $G'$, like $G$, has RNA normal-form states.
Note also that $G'$ is not, in general, normalized by production rule;
however, $G'$ is equivalent to $G$ in parse, and is therefore normalized by parse tree (if $G$ is).

\subsection{Eliminate null states}

We now proceed to eliminate null cycles from $G'$.
Since null subtrees have been eliminated, the remaining null cycles use only transition rules.

We create the grammar
\[
G'' = \left( \Omega, \gnons', \grules'', P'' \right)
\]
which has the same nonterminals as $G'$, but different rules and rule probabilities.

Let $\gnulls' \subseteq \gnons'$ be the set of null states in $\gnons'$, {\bf including} post-emit states.

Define ${\bf t}$, the {\em transition matrix} of $G'$, as $t_{XY} = P'(X \to Y)$ for all $X,Y \in \gnons'$.
The effective transition probability $q_{XY}$ between two states $X,Y$
sums over all paths through null states (in the summand, $n$ is the length of the null path):
\[
{\bf q} = \sum_{n=0}^\infty {\bf t}^n = ({\bf 1}-{\bf t})^{-1}
\]

Here ${\bf 1}$ is the identity matrix.
The matrix inverse may be computed in the usual ways (Gauss-Jordan elimination, LU decomposition, etc.)

Since the bifurcation states of $G'$ explicitly generate non-empty sequence,
the system of equations relating the probabilities $P'(\epsilon|X)$ is completely linear
and may be solved in the same way.
Let $u_X = P'(\epsilon|X)$ and $v_X = P'(X \to \epsilon)$.
Then

\[
{\bf u} = {\bf v} + {\bf tu}
\]

whose solution is ${\bf u} = {\bf qv}$.
Thus $P'(\epsilon|X) = \sum_Y q_{XY} P'(Y \to \epsilon)$.

We now define $P''$ as follows.
\begin{itemize}
\item For all rules $\rho \in \grules'$ that are {\bf not} transitions or terminations, set $P''(\rho) = P'(\rho)$.
\item For terminations from null states $X \in \gnulls'$, set $P''(X \to \epsilon) = u_X$.
(NB this may create some terminations $X \to \epsilon$ that were not present in $G'$.)
\item For transitions from null states $X \in \gnulls'$ to non-null states $Y \notin \gnulls'$, set $P''(X \to Y) = q_{XY}$.
(NB this may create some transitions $X \to Y$ that were not present in $G'$.)
\item For transitions {\bf to} null states $X \in \gnulls'$, set $P''(A \to X) = 0$.
\end{itemize}

Although we have left null states in $G''$, they now have no incoming transitions
and are inaccessible unless they are post-bifurcation states (i.e. states which appear on the right-hand side of bifurcation rules), post-emit states, or the start (root) state.
All other null states can therefore be dropped from $\gnons''$.

\subsection{Restore null states}

Suppose that $\rho''$ is a rule in $G''$.
Algorithm~\ref{alg:restoreTransitions} samples from the distribution of equivalent parse subtrees in $G'$ (possibly containing null cycles).

In order to sample from $P(T'|T'')$ we need simply map Algorithm~\ref{alg:restoreTransitions} to each rule in $T''$.

For parameter estimation by Expectation Maximization, and some other applications, it is useful to know the expected number
of times that a transition was used, summed over the posterior distribution of parse trees (including those with null cycles).
If $n''(\rho'')$ is the expected number of times that transition $\rho''$ was used according to an Inside-Outside computation on $G''$,
then the corresponding expectations $n'(\rho')$ are given by

\begin{eqnarray*}
       n'(X \to Y) & = & n''(X \to Y) \frac{P''(X \to Y)}{q_{XY}} + \sum_Z n''(X \to Z) \frac{P''(X \to Y) q_{YZ}}{q_{XZ}} \\
n'(X \to \epsilon) & = & n''(X \to \epsilon) \frac{P''(X \to \epsilon)}{P''(\epsilon | X)}
 + \sum_W n''(W \to \epsilon) \frac{q_{WX} P''(X \to \epsilon)}{P''(\epsilon | W)}
\end{eqnarray*}

Expectations for other rules (bifurcations and emissions) are the same for $G'$ as $G''$.

\begin{algorithm}[!ht]
  \SetLine
  \SetKwFunction{rnd}{Random[0,1]}
  \KwIn{Rule $\rho'' \in \grules''$}
  \KwOut{Parse (sub)tree $T' \in \gtrees'$}

  \Switch{$\rho''$}{
    \Case{$X \to Y$}{
      \eIf{\rnd $< P(X \to Y) / q_{XY}$}{
        \KwRet{$(X \to Y)$} \;
      }{
        Let $z = \sum_W q_{XW} P'(W \to Y)$ \;
        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to Y) / z$ \;
        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to Y)$} \;
      }
    }
    \Case{$X \to \epsilon$}{
      \eIf{\rnd $< P'(X \to \epsilon) / P'(\epsilon | X)$}{
        \KwRet{$(X \to \epsilon)$} \;
      }{
        Let $z = \sum_W q_{XW} P'(W \to \epsilon)$ \;
        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to \epsilon) / z$ \;
        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to \epsilon)$} \;
      }
    }
    \Other{
      \KwRet{$(\rho'')$} \;
    }
  }
  \caption{\label{alg:restoreTransitions}
    Subroutine $\mathrm{restoreTransitions}(\rho'')$.
  }
\end{algorithm}

\subsection{Restore null bifurcations}

Suppose that $\rho'$ is a rule in $G'$.
Algorithm~\ref{alg:restoreBifurcations} samples from the distribution of equivalent parse subtrees in $G$ (possibly containing null subtrees).
This algorithm also calls Algorithm~\ref{alg:sampleNullSubtree}, which samples from the distribution of empty subtrees rooted at a particular nonterminal.

In order to sample from $P(T|T')$ we need simply map Algorithm~\ref{alg:restoreBifurcations} to each rule in $T'$.

If $n'(\rho')$ is the expected number of times that rule $\rho'$ was used by $G'$,
then the corresponding expectations $n(\rho)$ are given by

\begin{eqnarray*}
    n(B \to L\ R) & = & n'(B' \to L') + n'(B' \to R') + n'(B^0 \to L'\ R') + d(B \to L\ R) \\
       n(X \to Y) & = & n'(X \to Y) + d(X \to Y) \\
n(X \to \epsilon) & = & n'(X \to \epsilon) + d(X \to \epsilon)
\end{eqnarray*}

Expectations for emissions are the same for $G$ as $G'$.

In these expressions $d(\rho)$ is the expected usage of rule $\rho$ by null subtrees:

\[
d(\rho) = \sum_{(B \to LR) \in \grules} \left[
n'(B' \to L') c_R(\rho)
+ n'(B' \to R') c_L(\rho)
+ n'(B^0 \to \epsilon) (c_L(\rho) + c_R(\rho))
\right]
\]

where $c_W(\rho)$ is the expected usage of rule $\rho$ by an empty parse tree rooted at $W$,
given by

\begin{eqnarray*}
c_X(\rho) P(\epsilon | X) & = & P(X \to \epsilon) \delta_{\rho=(X \to \epsilon)}
 + \sum_Y P(X \to Y) P(\epsilon|Y) \left( c_Y(\rho) + \delta_{\rho=(X \to Y)} \right)
\\ & &
 + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R) \left( c_L(\rho) + c_R(\rho) + \delta_{\rho=(X \to LR)} \right)
\end{eqnarray*}

where $\delta_U$ is the Kronecker delta (1 if condition $U$ is true, 0 if it is false).
Note that in contrast to the system of equations for $P(\epsilon|X)$, this is a linear system of equations
of the form ${\bf c} = {\bf Mc} + {\bf k}$, which may be solved by matrix inversion:
${\bf c} = ({\bf 1}-{\bf M})^{-1} {\bf k}$.

\begin{algorithm}[!ht]
  \SetLine
  \SetKwFunction{rnd}{Random[0,1]}
  \KwIn{Rule $\rho' \in \grules'$}
  \KwOut{Parse (sub)tree $T \in \gtrees$}

  \Switch{$\rho'$}{
    \Case{$B^0 \to \epsilon$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
      \KwRet{$(B \to T_L\ T_R)$} \;
    }
    \Case{$B' \to L'$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
      \KwRet{$(B \to L\ T_R)$} \;
    }
    \Case{$B' \to R'$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
      \KwRet{$(B \to T_L\ R)$} \;
    }
    \Case{$B^2 \to L'\ R'$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      \KwRet{$(B \to L\ R)$} \;
    }
    \lCase{$B^0 \to B'$}{
      \KwRet{$()$} \;
    }
    \lCase{$B' \to B^2$}{
      \KwRet{$()$} \;
    }
    \Other{
      Let $\rho$ be the original rule in $\grules$ \;
      \KwRet{$(\rho)$} \;
    }
  }

  \caption{\label{alg:restoreBifurcations}
    Subroutine $\mathrm{restoreBifurcations}(\rho')$.
  }
\end{algorithm}

\begin{algorithm}[!ht]
  \SetLine
  \SetKwFunction{rnd}{Random[0,1]}
  \KwIn{Nonterminal $A \in \gnons$}
  \KwOut{Parse tree $T \in \gtrees: \mathrm{seq}(T) = \epsilon, \mathrm{root}(T) = A$}

  \eIf{$A$ is a bifurcation state, $A \to X\ Y$,}{
    $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
    $T_Y \leftarrow \mathrm{sampleNullSubtree}(Y)$ \;
    \KwRet{$(A \rightarrow T_X\ T_Y)$} \;
  }{
    \eIf{\rnd $< P(A \to \epsilon) / P(\epsilon|A)$}{
      \KwRet{$(A \to \epsilon)$} \;
    }{
      Let $z = \sum_Y P(A \to Y) P(\epsilon | Y)$ \;
      Sample state $X$ from probability distribution $P(X) = P(A \to X) P(\epsilon | X) / z$ \;
      $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
      \KwRet{$(A \rightarrow T_X)$} \;
    }
  }

  \caption{\label{alg:sampleNullSubtree}
    Subroutine $\mathrm{sampleNullSubtree}(A)$.
  }
\end{algorithm}

### 2009-04-18 paper.tex
Import from CVS.

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{PLoS2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **
%% Please insert a running head of 30 characters or less.  
%% Include it twice, once between each set of braces
\markboth{Modeling structured RNA}{Modeling structured RNA}

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW
\input{defs.tex}
%% END MACROS SECTION

\begin{document}

% Title must be 150 words or less
\begin{flushleft}
  {\Large
    \textbf{Evolutionary triplet models of structured RNA}
  }
\\
Robert K. Bradley$^{1}$, 
Ian Holmes$^{1,2,\ast}$
\\
\textbf{1} Biophysics Graduate Group, University of California, Berkeley, CA, USA
\\
\textbf{2} Department of Bioengineering, University of California, Berkeley, CA, USA
\\
$\ast$ E-mail: indiegram@postbox.biowiki.org
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\newpage
\section*{Abstract}
The existence of an all-RNA primordial ribosome was proposed by
Francis Crick in 1968.  Recently, a domain-level model for the
original ribosome was proposed by Smith et al.  Given the abundance of
ribosomal RNA sequence data and the many successful reconstructions of
ancestral proteins, the reconstruction and synthesis of ancestral
ribosomes (and other RNAs) is a feasible goal for paleogenetics.  This
will require new bioinformatics tools and methods, in particular a
robust theoretical framework for reconstructing histories of
substitutions, indels and changes in RNA structure.

We describe a ``transducer composition'' algorithm for extending
pairwise statistical models of RNA structural evolution to models of
multiple sequences, related by a phylogenetic tree.  This algorithm
draws on formal models of computational linguistics as well as the
1985 ``protosequence'' algorithm of David Sankoff.  The output of the
composition algorithm is a multiple-sequence stochastic context-free
grammar.  We describe dynamic programming algorithms, which are robust
to null cycles and empty bifurcations, for parsing this grammar.
Example applications include structural alignment of non-coding RNAs,
propagation of structural information from an
experimentally-characterized sequence to its homologs, and inference
of the ancestral structure of a set of diverged RNAs.

We implemented these algorithms for a three-taxon phylogeny.  Using
our implementation on a simple model of pairwise RNA structural
evolution, we simulated data under the model and tested the dependence
of alignment inference and reconstruction accuracy on the evolutionary
distance of the outgroup of the phylogeny.  Our results suggest that
we may be able to accurately reconstruct ancestral structures despite
sequence and structural divergence if the evolutionary process is
well-understood.
We discuss the implications of our results for RNA algorithm design
and phylogenetic reconstruction of RNA evolutionary history.

The programs described are available
as part of the \dart\ software package for sequence analysis,
released under the GNU Public License at
\darturl.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\newpage
\section*{Author Summary}

\newpage
\section*{Introduction}
% Biological motivation
In 1968, Francis Crick hypothesized that the first ribosome consisted entirely of RNA, without any protein cofactors \cite{Crick68}.
A domain structure for this primeval ribosome was recently proposed \cite{SmithEtAl2008}.
To synthesize such a reconstructed ribosome, or reconstructions of other evolutionarily significant RNAs
such as group II introns \cite{LehmannSchmidt2003} or telomerase \cite{AntalEtAl2002},
it will be necessary to develop methods that can predict the sequences of ancient RNAs
based on the divergent sequences of their many descendants.

% SCFGs for RNA
Inspection of RNA alignments, for example in the \RFAM\ database
\cite{GriffithsJonesEtAl2003}, suggests that an evolutionary model for
RNA structure must eventually include multiple layers of detail: point
substitutions, covariant substitutions of base-pairs
\cite{HancockTautzDover88,LeontisEtAl2002}, indels
\cite{TakeshiTsutomu2008}, local changes in secondary structure such
as helix slippage \cite{HancockDover90}, and changes in domain
structure \cite{SmithEtAl2008}.  Stochastic context-free grammars
(SCFGs), which can efficiently model the long-range correlations
generated by RNA secondary structures, are natural probabilistic models of such
phenomena and have been used for ncRNA homology detection
\cite{Eddy94,Durbin98,KleinEddy2003,NawrockiEddy2007}, gene prediction
\cite{RivasEddy99,PedersenEtAl2006}, folding
\cite{KnudsenHein2003,DowellEddy2004} and alignment
\cite{Holmes2005,DowellEddy2006,BradleyPachterHolmes2008}.

% Two parts: pairwise & multiple
By analogy with substitution processes, which are well-understood \cite{Felsenstein2003},
we may split the ancestral reconstruction problem into two halves.
The first half is the development of a {\bf pairwise model},
describing the probability distribution $P(Y|X)$ of a descendant ($Y$) conditional on its immediate ancestor ($X$).
In substitution processes, the pairwise model is a conditional substitution matrix.
Often (but not always) the pairwise model is derived from an instantaneous model of change,
i.e., a continuous-time Markov chain (parametrized by a rate matrix);
obtaining the transition probabilities of this chain (via exponentiation of the rate matrix)
yields a pairwise model whose parameters are smoothly-varying functions of a finite time parameter $T$.
A pairwise model represents an individual branch of a phylogenetic tree, with $T$ representing the length of that branch.

% (second part: multiple)
The second half of the problem involves extending the model (and related inference algorithms) from a single branch to a complete phylogeny,
i.e., from a pairwise model of two sequences to a {\bf multiple-sequence model} of many sequences.
In a typical situation, the sequences at the leaves of the tree are observed, but those at internal nodes are unobserved.
Questions of interest then include:
\begin{description}
\item[A.] What is the likelihood for the observed sequence data?
\item[B.] Can we sample from (or exactly compute) the posterior distribution of the unobserved sequence at the root node?
\item[C.] Can we sample from the posterior of the unobserved sequences at the other internal nodes?
\item[D.] Can we estimate summaries of the evolutionary history, such as
  the number of substitution events on each branch (for a substitution
  model), the alignment (for a model which includes indels), or
  changes in the underlying structure (for a model of RNA structure)?
\end{description}

For substitution models, there has been extensive work focused on answering each of these questions.
Given a pairwise substitution model, questions A and B can be
answered exactly by Felsenstein's pruning algorithm
\cite{Felsenstein81} and question C can be answered by the peeling
algorithm (first presented for pedigree analysis by Elston and Stewart \cite{ElstonStewart71}).
The estimation of evolutionary histories (question D) has been
addressed by exact summarization \cite{HolmesRubin2002b} and sampling
\cite{Nielsen2001} approaches.
Many of these questions and answers can be unified by an
appropriate framework;
for example, another representation of answers A-C
is that the ``pruning'' \& ``peeling'' algorithms (combined) are just
the sum-product algorithm on a directed graphical model \cite{Pearl82},
yielding exact marginal distributions for unobserved variables.
Graphical models also suggest general-purpose sampling approaches
in addition to the exact sum-product algorithm.

% We want a multiple SCFG. Not a pairwise SCFG...
The two halves of the reconstruction problem --- developing a pairwise model and then extending it to multiple sequences --- are largely independent.
Felsenstein's pruning algorithm, for example, is essentially blind to the parametric form of the pairwise substitution model;
it just assumes that a substitution model is provided for every branch.
Subsequent models developed by other researchers can be plugged into the pruning algorithm without modification \cite{Yang94b,WhelanGoldman2001}.

We therefore addressed the problem of modeling the indel-evolution of
multiple structured RNAs in a similarly-modular fashion by separating
the creation of pairwise and multiple-sequence models.  In previous
work, we addressed the first (pairwise) part of the RNA reconstruction
problem by describing a simple continuous-time model of RNA structural
evolution \cite{Holmes2004}.  This model corresponded to a Pair SCFG
with a time-dependent parametrization which we used to
simultaneously align and predict the structure of
related RNAs.  The focus of the present work is to solve the second
(multiple-sequence) part of the RNA reconstruction problem by giving a
general procedure for extending a pairwise model to multiple
sequences related by a phylogenetic tree.  This process
yields a multiple-sequence SCFG, a natural model of the evolutionary
relationships between multiple structured RNAs.
% (just as a Pair SCFG is the natural model of the relationship between two RNAs).

The main contributions of this paper are
(1) an algorithm that transforms a phylogenetic ensemble of pair grammars
 into a coherent, multiple-sequence SCFG,
(2) dynamic programming (DP) algorithms for performing inference under
this multiple-sequence SCFG, and
% (which appear to converge even if the SCFG contains null bifurcations and null cycles);
(3) freely-available software implementing algorithms (1) and (2) for
the simplified case of a three-branch star-topology tree.
While the idea of composing conditionally-normalized models on trees
is intuitive, the resulting models can be very complex, even for
simple models of RNA evolution, making (1) necessary.
Studies of related indel models have suggested that an implementation
of dynamic programming (DP) algorithms on the three-branch tree topology is sufficient to
draw samples from the posterior distribution of ancestral sequences on
more complex tree topologies, using Markov Chain Monte Carlo or MCMC
\cite{HolmesBruno2001,JensenHein2002,RedelingsSuchard2005},
suggesting that (2) and (3) are, in principle, sufficient for analyzing trees
relating many sequences.

% Benchmarking
We show that our algorithm produces a multiple-sequence grammar which is
much more compact than suggested by naive approaches to model
construction.  We provide analyses of the asymptotic complexities of
models constructed using our procedure and provide estimates of the
time and memory required to reconstruct the structures of several RNA
families for the case of a 3-taxon phylogeny, which we have
implemented in the program \indiegram. While by these estimates only
the smallest sequences currently fit into affordable memory, thereby
preventing us from applying our method to many problems of interest,
we assess the conduct a simulation study to investigate the accuracy
with which we can hope to reconstruct ancestral structures over long
evolutionary time in the presence of structural divergence.

In the Discussion, we speculate on algorithmic extensions that
may reduce memory requirements, inspired by related work in
reconstructing DNA and protein sequences.

% Furthermore, we report benchmark results for special cases that can fit into memory:
% (1) the accuracy of reconstruction for the case where indels are prohibited and alignment is constrained, and
% (2) the case where we build a progressive multiple sequence alignment using pairwise comparisons.

\newpage
\section*{Methods}
We describe below a general method for constructing a multiple-sequence
stochastic grammar for alignment, folding and ancestral reconstruction of RNA,
given a phylogenetic tree and a description of the evolutionary
process acting along each branch.

% Functional constraints on structured RNA act largely at the level of
% secondary structure, so a realistic model of RNA evolution must model
% phenomena such as covariant
% base-pair substitutions or indels which preserve stem structures but
% modify stem length.
% Pair SCFGs---generative formal grammars or stochastic
% models of two sequences---can model these phenomena and serve as efficient models
% for RNA alignment and structure-prediction algorithms
% \cite{Holmes2005,DowellEddy2006,KiryuEtAl2007,BradleyPachterHolmes2008}.
% 
% Our method for generating a multiple-sequence model is qualitatively
% similar to the method presented in \cite{Holmes2003} for HMMs, but
% modified to permit bifurcations and covariant substitutions as modeled
% by SCFGs. Grammars constructed by this algorithm can be used to infer
% probable structures of ancestral RNAs.

\subsection*{Overview}

Our problem statement is this:
%\vspace{.5\baselineskip}
%\fbox{
%\parbox{.8\textwidth}
\textbf{Given a phylogenetic tree relating several structured RNAs and a
  description of the evolution of a structured RNA
  along a single branch of the tree (in the form of a Pair SCFG),
  (1) find the corresponding
  phylogenetic multiple-sequence grammar and (2) use that grammar to
  reconstruct, {\em a posteriori}, the evolutionary histories of the
  RNAs.}
%}
%\vspace{.5\baselineskip}
We assume that the phylogeny, including both the tree topology and
branch lengths, is known.

This paper focuses on model construction and inference algorithms
rather than the heuristics and constraints which will be necessary to make these
algorithms fast enough for analysis of many biological datasets.
As discussed below, the complexity of our inference algorithms is
prohibitively high for many problems of interest.
However, this complexity can be significantly
reduced by incorporating outside knowledge.  For example, if we know
the consensus structure of several sequences or their individual
structures, then we can constrain our algorithms accordingly.
Similarly, we might consider only ancestral structures which are
compatible with a curated multiple alignment.  Such constraints are
commonly used by programs for RNA sequence analysis \cite{RivasEddy2001,Holmes2005}.

The following sections introduce more precise definitions for two-sequence
models of RNA structure and proceed to outline our algorithms for
(1) combining these two-sequence models on a phylogenetic tree and
(2) using the composite phylogenetic grammars for inference.

% To generalize from two sequences to many sequences,
% we place a parse-tree transducer on every branch of a phylogenetic tree.

\subsection*{Two-sequence models} \seclabel{twoseq}

We discuss the general problem of creating state-space models of the
evolution of related sequences, beginning with models of substitution processes acting at independent
sites (as studied in likelihood phylogenetics) and generalizing to models of
indels first in primary sequences, then in sequences with conserved secondary structure.

A stochastic model for the evolution of one sequence (the ancestor, $X$) into
another (the descendant, $Y$) over an interval of time ($T$) can be described by a
joint distribution, $P (X, Y | T)$.
This joint distribution can be factored as $P (X) \cdot P (Y | X, T)$, 
where $P(X)$ is the marginal distribution over ancestral sequences
and $P(Y | X, T)$ is the conditional distribution over descendant sequences given an ancestral sequence.
In terms of phylogenetics, the conditional distribution $P(Y | X, T)$ describes the evolution $X \stackrel{T}{\to} Y$
along a branch of length $T$.

It is possible to ``multiply'' two such models together.
More precisely, one multiplies two conditional distributions and sums out the intermediate sequence.
Thus, successive evolution along two branches $X \stackrel{T_1}{\to} Y \stackrel{T_2}{\to} Z$
is modeled by the distribution
\[
P(Y,Z|X,T_1,T_2) = P(Y | X, T_1) P(Z | Y, T_2)
\]

and we can sum sequence $Y$ out of this, obtaining the distribution
\[
P(Z|X,T_1+T_2) = \sum_Y P(Y | X, T_1) P(Z | Y, T_2)
\]
for the composite branch
$X \stackrel{T_1 + T_2}{\longrightarrow} Z$.

This formalism underlies likelihood phylogenetics.
Working under the independent-sites assumption, $P (X, Y | T)$ is the $(X,Y)$'th element of the joint substitution matrix for a single site
and $P(Y | X, T)$ is the corresponding element of the conditional matrix.
The conditional matrix is in fact the matrix exponential
 $\exp({\bf R}T)$, where ${\bf R}$ is the substitution rate matrix \cite{HolmesRubin2002b}.
Composition of two branches just amounts to a matrix multiplication.

A similar formalism can be used to describe the evolution of whole
sequences with indels.
Suppose that the joint distribution $P (X, Y | T)$ is the distribution
modeled by a pair hidden Markov model (Pair HMM) \cite{Durbin98},
a probabilistic model of the evolution of two sequences under the
approximation that only adjacent characters are directly correlated,
and the marginal $P(X)$ is the distribution of a single-sequence HMM,
a probabilistic model of single sequences under the same approximation.
The conditional distribution $P(Y | X, T)$ then corresponds to a conditional Pair HMM:
a finite-state machine which transforms one sequence (the input, $X$) into another (the output, $Y$).
Following computational linguists, we call this conditionally-normalized state machine a
 {\bf string transducer} or simply a {\bf transducer}
\cite{BradleyHolmes2007}.
Because of its conditional normalization, this state machine is
distinct from a standard Pair HMM.
A Pair HMM has two outputs $X$ and $Y$ and emits symbols to both of those outputs,
while a transducer absorbs symbols from the input $X$ and emits
symbols to the output $Y$.
Despite this distinction, Pair HMMs and transducers share very similar inference algorithms;
for example, $P(Y | X, T)$ is computed using a direct analogue of the Forward algorithm \cite{Durbin98}.

We extend this formalism to the case of structured RNA as follows.
Let $X$ and $Y$ now represent structured RNA sequences, or more
precisely, parse trees.
A single-sequence SCFG models the marginal $P(X)$;
the joint distribution $P (X, Y | T)$ is modeled by a jointly-normalized Pair SCFG \cite{Durbin98}.
The conditional distribution $P(Y | X, T)$ is modeled by a conditionally-normalized Pair SCFG.
Following terminology from computational linguistics
\cite{ComonEtAl2007-TreeTransducers}, we call this
conditionally-normalized grammar a {\bf parse-tree transducer}.

String transducers are special cases of parse-tree transducers, just
as HMMs are special cases of SCFGs.
Henceforth, we will drop the distinction between strings and parse trees.
We will also refer interchangeably to ``states'' (in the state-machine representation)
and ``nonterminals'' (in the grammar representation).
Likewise, we will refer interchangeably to ``state paths'' (machines) and ``parse trees'' (grammars).

\paragraph{Terminology and normalization.}

The stochastic state machine which generates parse-trees from the marginal distribution $P(X)$ is a {\bf singlet transducer}.
The machine that absorbs parse-trees $X$ and generates modified parse-trees $Y$ from the conditional distribution $P(Y | X, T)$ is a {\bf branch transducer}.
Singlet transducers only emit symbols to their output sequence;
branch transducers, in contrast, can both emit symbols to their output
and absorb symbols from their input sequence.

Much as the states of Pair HMMs with no null states, other than the $\Sstart$
and $\Send$ states, can be classified as $\Sinsert$ or $\Smatch$,
transducers can have states of type $\Sstart$, $\Send$,
$\Swait$, $\Sinsert$ and $\Smatch$.
The first three state types, $\Sstart$, $\Send$ and $\Swait$, are null:
They do not emit or absorb any symbols and are required solely for
``organizational'' purposes (see following section).
Two types of states can emit and/or absorb symbols, $\Sinsert$ and $\Smatch$.
An $\Sinsert$ state emits a symbol to the output, without absorbing anything.
A $\Smatch$ state absorbs a symbol on the input and either emits
the same symbol to the output, substitutes a different output symbol,
or emits no output symbol at all (the latter corresponds to a deletion).

As stated above, the Pair SCFG must be ``conditionally normalized''
so that models can be chained together,
extending the pairwise model to multiple sequences.
The transformation rules are partitioned into co-normalized groups:
within each group, the rule probabilities must sum to one.
In a jointly-normalized Pair SCFG, each group corresponds to the set of all rules that can be applied to a given nonterminal
(i.e., all outgoing transitions from a particular state).
In a conditionally-normalized Pair SCFG, the groups are finer-grained:
each co-normalized group includes all rules that can be applied to a given nonterminal, {\em for a given set of absorbed symbols}.
%(In practice it is often expedient to relax these normalization guidelines somewhat,
%while preserving the overall probabilistic normalization over output parse trees.
%For example, the normalization of states $B$ and $B_p$ in \tabref{tkfstbranch} occurs on transitions into, rather than out of, those states.)

% Conditional normalization is required so that models can be chained together,
% extending the pairwise model to multiple sequences.
% In the language of graphical models, the jointly-normalized Pair SCFG represents an undirected
% edge in the model, whereas the conditionally-normalized SCFG represents a
% directed edge.  The rooted phylogeny which is input to our algorithm
% is a directed graphical model and so we require the conditional normalization.

\subsection*{Multiple-sequence models}

We can use the concepts of factoring probability distributions introduced in the two-sequence
framework to model the common descent of many homologous sequences.
Given a phylogenetic tree and two-sequence model, we wish to obtain a
multiple-sequence SCFG describing the common descent of the observed sequences.

% We begin with some technical definitions and rules for combining models,
% then proceed to a more colloquial description of the collective behavior
% of a phylogenetic ensemble of transducers co-ordinated by these rules.

% Consider a phylogenetic tree relating $n$ structured RNAs
% $\{ X_1 , ... , X_n \}$ lying on the leaves of the tree,
% and a further $m$ ancestral RNAs
% $\{ X_{n+1} , ... , X_{n+m} \}$
% lying at the internal nodes of the tree.
% Because any node of the tree is independent of its sibling node conditioned on their common parent,
% we can factor the probability distribution $P(X_1 , ... , X_{n+m})$ into terms of the form
% $P(X_i | X_j, T)$, precisely the conditional distribution which is generated by a branch transducer.
% This has a ready biological interpretation:  A branch transducer
% evolves a sequence (and structure) along a branch of a phylogenetic tree,
% so we extend the two-sequence models just described to evolutionary models of many sequences
% by composing branch transducers on the branches of the tree.

A singlet transducer (which emits, but does not absorb, symbols) lies at the root of the phylogeny and 
serves as a generative model of the ancestral sequence.
To represent the evolution of an ancestral sequence into many descendant sequences,
we place a branch transducer on each branch of the phylogeny.
Figures \ref{fig:parent} and \ref{fig:threeway} show simple examples
of building a model to represent such an evolution.

Throughout this paper we make frequently refer to two and
three-taxon (star) phylogenies.  In all cases, the sequence $W$ is
assumed to be the (unobserved) ancestral sequence and the sequences
$X$, $Y$, and $Z$ the (observed) extant sequences (as in Figures
\ref{fig:parent} and \ref{fig:threeway}).

\subsubsection*{The composition algorithm}

While this composition of conditionally-normalized models on a
phylogenetic tree is intuitive, in practice building such an ensemble
model is challenging due to the sheer number of possible states and
transitions of the ensemble model.
The maximum possible state space of the ensemble is the Cartesian product of the
individual transducer state spaces. If the singlet transducer has $a$
states, each branch transducer has $b$ states, and there are $N$
branches in the phylogeny, then an upper bound on the number of
ensemble states is $O(a \cdot b^N)$.
However, in practice there are many fewer states than suggested by
this bound; many state configurations are not reachable.  For example,
for the tree with two extant sequences and a single parent,
the branch transducers above leaves $X$ and $Y$ cannot simultaneously
be in $\Sinsert$ states, as this would correspond to aligning
non-homologous (inserted) characters.
Similarly, while an upper bound on the number of possible transitions
in the transition matrix of the ensemble model is $O((a \cdot b^N)^2)$,
in practice models never reach this bound,
due both to inaccessible configurations (such as the one described above)
and the sparseness of transitions between the remaining, accessible configurations.

While the accessible state space of the ensemble is smaller than that
given by the exponential upper bound, it is generally nonetheless too
complex to deal with by hand.
For example, the simple model of RNA structural evolution described in
Results yields an ensemble model of three sequences with 230 states
and 1,789 transitions.  More realistic models of RNA give rise to even
larger ensemble models.

We therefore need an algorithm to efficiently construct the ``state graph'' of the
ensemble model, consisting of a list of accessible states and the
possible transitions between them. 
By analogy with algorithms for uninformed graph search in artificial
intelligence, the transition graph of the ensemble can be constructed
by an uninformed depth-first search, where at each step of the search
we obtain the next possible ensemble states by changing the state of one or
more of the singlet or branch transducers. Beginning with the entire
ensemble in state $\Sstart$, the depth-first search of states
continues until all nodes are in state $\Send$.

The allowed transitions of the ensemble can be categorized as follows:
\begin{enumerate}
\item {\bf Null Transition}:
  A branch transducer makes a transition into a $\Swait$ state, with no terminal emission or bifurcation.
\item {\bf Terminal Emission}:
  A singlet or branch transducer makes a transition into a state of type $\Sinsert$,
  emitting left and/or right terminal symbols (e.g., a single base or base-pair).
  These symbols are absorbed by the immediately-descended transducers,
  which are pushed into states of type $\Smatch$
  and may themselves emit terminal symbols
  that will be absorbed by {\em their} descendant transducers.
  This continues down the tree: The terminal symbols are passed from parents, to children, to grandchildren
  (albeit possibly being replaced by other terminal symbols as they're propagated down)
  and they propel branch transducers into $\Smatch$ states as they go.
  Eventually, the cascade of emitted terminal symbols stops when all the symbols have been deleted,
  or when the cascade reaches the leaves of the tree.
\item {\bf Bifurcation}:
  A singlet or branch transducer makes a transition into a state of type $\Sinsert$ that spawns
  left and/or right nonterminal states.
  These nonterminals are processed recursively down the tree, just as in a
  terminal emission (conceptually, a bifurcation is a ``nonterminal emission'').
  As with terminal emissions, absorption of nonterminal emissions propel
  descendant transducers into $\Smatch$ states,
  making transitions which may themselves propagate nonterminals
  further down the tree.
  A biologically-relevant example of a bifurcation is the insertion of a stem
  into an ancestral RNA structure,
  which then may be conserved or deleted in the descendant structures.
\item {\bf End Transition}:
  The singlet transducer at the root makes a transition to the $\Send$ state,
  pushing all the descendant branch transducers into $\Send$ states and
  terminating the current branch of the parse tree.
\end{enumerate}

Co-ordination between the various branch machines is achieved by
specifying an ordering on the nodes (preorder with respect to the
phylogenetic tree) and by having branch transducers ``pause'' in $\Swait$
states while waiting to absorb a symbol from the node above.  
Only one transducer is allowed to make a spontaneous transition at a
time. If this transition corresponds to a terminal
emission or a bifurcation, then this may force descendant transducers
into making reactive transitions.

The four types of allowed transitions listed above can be formalized simply.
Let the total order on the nodes correspond to a preorder traversal of
the tree, from which it follows that $m \prec n$ if $m$ is ancestral to $n$.
Let ${\cal T}_m$ denote the singlet or branch transducer which emits symbols to node $m$.
Transducer ${\cal T}_m$ changes state if and only if one of the
following three mutually-exclusive conditions holds:
\begin{description}
\item{Type 1:} Transducer ${\cal T}_m$ is not in a $\Swait$ state,
  while all its successor transducers ${\cal T}_n$ are in $\Swait$
  states (where $m \prec n$).  ${\cal T}_m$ is free to make any transition.
\item{Type 2:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer emits a symbol, forcing ${\cal T}_m$ into a $\Smatch$ state so it can absorb that symbol.
\item{Type 3:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer enters the $\Send$ state, forcing ${\cal T}_m$ into the $\Send$ state as well.
\end{description}
These possible transitions of the ensemble generate multiple alignments as follows:
\begin{enumerate}
\item
  The singlet transducer and all branch transducers begin in their respective $\Sstart$ states.
\item \label{item:InitialWindback}
  Before any residues can appear at the root, the branch transducers all ``wind back'' into $\Swait$ states,
  via type-1 transitions. This occurs in reverse order (i.e.,a postorder traversal of the tree).
\item During this initial windback, clade-specific insertions can occur.
  This process is described in detail at step \ref{item:CladeInsertion}.
\item \label{item:SingletStart}
  With all the branch transducers wound back into $\Swait$ states,
  the singlet transducer makes a (type-1) transition into an $\Sinsert$ state, emitting a symbol to the sequence at the root node.
\item The transducers on outgoing branches from the root then make (type-2) transitions into $\Smatch$ states,
  either copying the root symbol to their own outputs,
  substituting it for a different symbol or staying silent (this silence corresponds to a clade-specific deletion;
  in our formalism, both substitutions and deletions are handled by $\Smatch$ states.)
\item The transducers on branches one step away from the root then process the symbols which reached them
  (if any did), followed by transducers on branches two steps away from the root, then three steps, and so on
  (these can all be regarded as occurring simultaneously, in a single ``wave'' or ``cascade'' of emissions).
\item \label{item:Windback}
  Eventually the emitted symbols are propagated, via type-2 transitions,
  all the way to the tips of the tree (if they survived)
  or to the nodes where they were deleted (if they did not survive).
  The wave of type-2 transitions has left a lot of branch transducers in $\Sinsert$ and $\Smatch$ states.
\item The branch transducers then, in postorder, each wind back into $\Swait$ states, just as at step \ref{item:InitialWindback}.
  (These windback transitions can be collapsed into a single ensemble transition, as with the emission cascade;
  however, the windback may be interrupted by clade-specific insertions; see below.)
\item \label{item:CladeInsertion}
  During the postorder windback, each branch transducer gets an opportunity to generate a new symbol (via type-1 transitions to $\Sinsert$ states).
  (If such a transition to $\Sinsert$ occurs, it corresponds to a clade-specific insertion.
  This insertion is propagated down the tree via a wave of type-2 transitions, as above,
  then we go back to step \ref{item:Windback}.)
\item Eventually, the entire ensemble has wound back, so that every transducer is in a $\Swait$ state except the singlet transducer at the root,
  which is still in an $\Sinsert$ state.
  At this point, all clade-specific insertions have been processed.
\item The singlet transducer now makes another type-1 transition.
  If this transition is to an $\Sinsert$ state, the entire cycle begins again: the singlet transducer emits the next symbol at the root,
  and we go back to step \ref{item:SingletStart}.
\item If, on the other hand, the singlet transducer enters its $\Send$ state,
  then a wave of type-3 transitions drives all the branch transducers into their respective $\Send$ states too,
  bringing the entire ensemble to a halt.
\end{enumerate}

The total sizes of both the state space and the transition matrix are,
in general, dramatically smaller than implied by the exponential upper
bounds of $O(a \cdot b^N)$ and $O((a \cdot b^N)^2)$.  While we do not
have provable bounds on the size of the state space, we have observed
that the size of the state space is roughly linear in the number of
branches, $O(a \cdot b \cdot N)$, and the number of transitions is
approximately linear in the number of states, for several pairwise models,
including the pairwise model we use here.
However, these empirical observations are based on a
limited class of pairwise models and we do not have theoretical results for
how they will generalize to other pairwise models.  We do believe, however,
that the worst-case exponential bound will be avoided by (1)
omitting inaccessible state configurations and (2) eliminating
``windback'' states as described in the following section (which we
believe will prevent affine gap penalties from generating exponential
growth in the number of states).

Therefore, for the models which we have characterized, the search
algorithm given below for enumerating all allowed transitions of the
ensemble model typically generates $O(b)$ transitions from any given
state, thereby creating a very sparse transition matrix of size $O(a \cdot b^2 \cdot N)$.

\subsection*{Inference algorithms for multiple-sequence models}

In this section, we describe dynamic programming (DP) algorithms for
inferring the alignment, structure and evolutionary history
of multiple related RNAs, using the multiple-sequence SCFG we have derived.
The model construction algorithm described above creates an evolutionary model
which generates a joint distribution over both ancestral and extant
sequences.  If we regard the ancestral sequences as hidden variables
to be marginalized over, then this model corresponds to a
jointly-normalized multiple-sequence SCFG (modeling the extant sequences).
We can then use the Cocke-Younger-Kasami (CYK) algorithm to impute the
maximum-likelihood alignment, secondary structure and common ancestor
of the extant sequences, or the Inside algorithm to sample from
the joint posterior distribution of these unknowns.

The possibility of fragments of ancestral sequence that are unobserved
in any of the descendant sequences, which can be unbounded in length,
leads to complications when performing inference on the multiple-sequence
SCFG. The SCFG can contain null cycles, corresponding to generation of
these unobserved fragments, the existence of which complicates the
inference algorithms.

Due to this possibility of unobserved ancestral sequence,
the standard CYK and Inside algorithms fail on the ensemble model.
Residues in the ancestor can be absent in all extant sequences;
the potential number of such unobserved ancestral residues is unbounded.
This manifests as cycles of $\Snull$ states within the multiple-sequence SCFG,
states which emit to the ancestral sequence but not to any extant sequence.
The presence of bifurcations with one or more possibly-empty child branches,
corresponding to entire substructures in the ancestral sequence that
are absent in any extant sequence, further complicates inference.
While there exists a straightforward fix for maximum-likelihood
inference with the CYK algorithm, the Inside algorithm requires
further computational machinery.

In order to correctly perform maximum-likelihood inference, we only
require that the lowest-probability parse corresponds to the parse
returned by the CYK algorithm; the entire
probability distribution over parses need not be preserved.
Because a state path which includes a cycle of $\Snull$ states
necessarily has lower probability than an equivalent one without, we
can simply ignore such $\Snull$ cycles during CYK.
Bifurcations with empty child branches cannot be similarly ignored,
but this problem can be addressed by creating equivalent effective
direct transitions between the parents of these bifurcation states and
their non-empty children.  Once these transformations have been performed,
maximum-likelihood inference can be performed using a standard
three-sequence CYK algorithm on the ensemble model.
(Similarly, the null cycle elimination discussed in computational
linguistics requires only that the CYK parse be preserved \cite{GecsegSteinby97}.)
However, posterior-decoding inference, which uses the Inside and
Outside algorithms, required a modified DP algorithm which preserves
the entire probability distribution over parses; the transformations
just described to make CYK work properly are insufficient for the
Inside and Outside algorithms.

There are two possible solutions to this difficulty: We can either
create new versions of the standard inference algorithms for SCFGs
which can cope with SCFGs with null cycles, or we can modify the
structure of the SCFGs generated by our model-construction algorithm
such that they are free of null cycles.  The first approach of
creating new algorithms is simple in principle---we can simply
iteratively re-fill entries of the DP matrix, corresponding to summing
the probabilities for unobserved null cycles, until the approximation
is sufficiently good---but it is difficult to implement efficiently,
and necessarily has worse time complexity than the standard inference
algorithms due to the need to iterate over parts of the (large) DP
matrix.  The second approach, of modifying the structure of the SCFGs
themselves, has a closed-form solution for the case of HMMs, but to our
knowledge, efficient exact null-cycle elimination for SCFGs is an open
mathematical problem (complicated by the presence of null cycles
involving bifurcations).

We therefore developed an iterative approximation to remove null cycles from
the ensemble model in order to use the standard inference algorithms
for SCFGs without incurring the further computational complexity
required by modifying the DP algorithms themselves.

\subsubsection*{Eliminating null cycles in SCFGs}

Null cycles in SCFGs constitute a ``missing data'' problem, wherein
the unobserved values of variables contribute to probabilities of
observed events. Such missing data problems are common in
probabilistic models of sequences and can
frequently be dealt with by exactly marginalizing over the
probabilities of the unobserved variables.  For example, eliminating
missing ``empty'' columns from an independent-sites model of sequence,
wherein alignment columns are independent, is straightforward
\cite{RivasEddy2008} and missing emissions in a HMM can be dealt with by matrix inversion
\cite{Holmes2003,Holmes2007,Lunter2007} or the ``wing retraction''
techniques used in HMMER \cite{hmmer}.
Algorithms for exact elimination of HMM null cycles by
matrix inversion are well-known \cite{Holmes2003} and are implemented
by programs such as PhyloComposer \cite{Holmes2007} and
HMMoC \cite{Lunter2007}. These algorithms effectively replace the HMM
with an equivalent HMM that contains no null cycles but generates the
same probability distribution over sequences.
However, these solutions do not easily generalize to SCFGs, due to the
possible presence of null bifurcations.

Null cycles can be removed from SCFGs by (i) separating bifurcations
into those which have one or more empty children (and can therefore be
represented using transition or termination rules) and those that have
two nonempty children and then (ii) encapsulating all paths that do
not emit any terminal symbols (and may therefore cause cyclic
dependencies in the DP matrix) as effective direct transitions, with
probabilities obtained by geometric series summation of the grammar's
transition matrix.

Both (i) empty bifurcations and (ii) paths without terminal symbols
generically appear in the ensemble model.
An empty bifurcation occurs when a child branch of a bifurcation state
transitions to the $\Send$ state without emitting any symbols and can
be removed from the model by creating an effective direct transition
encapsulating the empty bifurcation.  For
example, we can create an effective direct transition $N_1 \rightarrow
N_3$ between null states $N_1$ and $N_3$ in place of the empty
bifurcation $N_1 \rightarrow B \rightarrow (N_2 \,\, N_3) \rightarrow
(End \,\, N_3)$, where $B$ is a bifurcation state with children $(N_2
\,\, N_3)$. Bifurcation states are the most computationally-costly
part of our models, so it is important to eliminate as many as
possible without reducing model expressiveness.

Similarly, our model construction procedure generically produces many
``windback'' null transitions to non-emitting placeholder
states. If $\Semit$ states $E_1$ and $E_2$ are connected in the state
graph by $\Snull$ states $N_1$ and $N_2$, then the transition $E_1
\rightarrow N_1 \rightarrow N_2 \rightarrow E_2$ with  probability
$P(E_1 \rightarrow N_1) \cdot P(N_1 \rightarrow N_2) \cdot P(N_2
\rightarrow E_2)$ can be replaced with a transition $E_1 \rightarrow
E_2$ with an identical probability.

% In practice, eliminating windback
% states reduces the size of the state space by $\simeq 20\%$, resulting
% in a significant reduction in memory needed. However, eliminating
% $\Snull$ states can increase the number of transitions in the model,
% so this reduction in memory usage comes at the cost of a small
% increase in runtime.

We give a formal description of the null-cycle elimination procedure,
along with procedures for probabilistically ``restoring'' null cycles
to sampled parse trees and Inside-Outside expectation counts, below.

\paragraph{Definitions.}

Let $G = ( \Omega, \gnons, \grules, P )$ be a stochastic context-free grammar (SCFG)
consisting of
a set of terminal symbols $\Omega$,
a set of nonterminal symbols (a.k.a. ``states'') $\gnons$,
a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
and a non-negative weight function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.

(The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)

Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e., rules in which $A$ appears on the left),
and let $\rightside{B}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right).
Let $\allrules{B} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
Define these also on sets of nonterminals, e.g., $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.

Let $\gtrees$ be the set of all parse trees for $G$.
Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
then define the parse tree weight $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.

If $\sum_{T \in \gtrees} P(T) = 1$, we say that $G$ is {\em probabilistically normalized by parse tree}.
If $\sum_{\rho \in \leftside{A}} P(\rho) = 1$ for all $A \in \gnons$, we say that $G$ is {\em probabilistically normalized by production rule}.
Note that normalization by production rule $\Rightarrow$ normalization by parse tree.

Let $S \in \Omega^\ast$ denote a terminal sequence.
Let $\mathrm{seq}:\gtrees \to \Omega^\ast$ be the function mapping a parse tree to its terminal sequence.
and let $\mathrm{root}:\gtrees \to \gnons$ be the function returning the root nonterminal of a parse tree.
The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
Of particular relevance to null state elimination are the probabilities $P(\epsilon|A)$,
where $\epsilon$ is the empty sequence.
These are the probabilities that a nonterminal will expire without generating any sequence.

Following earlier formalisms \cite{Durbin98,HolmesRubin2002a,Holmes2005},
we say that the grammar $G$ has {\em RNA normal-form rules} if
each rule in $\grules$ takes one of the following four forms:

\noindent \hrulefill
\begin{description}
\item[Termination rules] with one nonterminal on the left and the empty string on the right
\begin{eqnarray*}
A & \to & \epsilon
\end{eqnarray*}
\item[Transition rules] with one nonterminal on the left and the right
\begin{eqnarray*}
A & \to & B
\end{eqnarray*}
\item[Bifurcation rules] with one nonterminal on the left and two on the right
\begin{eqnarray*}
A & \to & B\ C
\end{eqnarray*}
\item[Emission rules] with one nonterminal on the left and right and at least one terminal on the right
\begin{eqnarray*}
A & \to & x\ B \\
A & \to & B\ y \\
A & \to & x\ B\ y
\end{eqnarray*}
\end{description}
\noindent \hrulefill

\noindent
Here $A,B,C \in \gnons$ and $x,y \in \Omega$.

Furthermore, we say that the grammar $G$ has {\em RNA normal-form states} if
each nonterminal (state) $A \in \gnons$ takes one of the following forms:

\noindent \hrulefill
\begin{description}
\item[Null states]: $\leftside{A}$ contains only transition and termination rules.
\item[Bifurcation states]: $\leftside{A}$ contains exactly one bifurcation rule, $A \to X\ Y$, where $X \neq Y$ and $X,Y$ are both null states.
\item[Emit states]: $\leftside{A}$ contains only emission rules. Further, $\leftside{A} = \rightside{B}$ for some null state $B$.
This null state $B$ is called $A$'s {\em post-emit} state.
\end{description}
\noindent \hrulefill

Define a {\em null cycle} to be a nonterminal $A$
and a sequence of rules $\rho_1, \rho_2 \ldots \rho_k$ that, when applied consecutively to $A$, leave $A$ unchanged;
that is, $\rho_k(\rho_{k-1} \ldots \rho_2(\rho_1(A))) = A$.
Define a {\em null subtree} to be a null cycle using at least one bifurcation rule.

Suppose that $G = ( \Omega, \gnons, \grules, P )$ and $G' = ( \Omega, \gnons', \grules', P' )$ are two grammars.
We say that $G$ and $G'$ are {\em equivalent in sequence} if there is a mapping between nonterminals, $f:\gnons \to \gnons'$,
such that $P(S|A) = P(S|f(A))$.
We say that $G$ and $G'$ are {\em equivalent in parse} if there is a mapping between parse trees, $g:\gtrees \to \gtrees'$,
such that $P'(T') = \sum_{T:T'=g(T)} P(T)$ and we can define
\[
P(T|T') = P(T|g(T)=T') = \frac{P(T)}{P'(T')}
\]

Note that equivalence in parse $\Rightarrow$ equivalence in sequence.

Suppose that a grammar $G$ contains null cycles.
We seek to transform $G$ into a grammar $G''$ that is equivalent in parse and sequence, but has no null cycles.
If $G$ is normalized by parse tree, then $G''$ will be too; but $G''$ is not (necessarily) normalized by production rule.

We do the transformation in two steps, $G \to G' \to G''$.
We also provide a stochastic ``null cycle restoration'' algorithm for sampling from $P(T|T'')$.

\paragraph{Eliminating null bifurcations.}

Note first that any SCFG can be transformed into an equivalent one with RNA normal-form states by adding null states.
Without loss of generality, we therefore consider SCFGs with RNA normal-form states.

Let $G = ( \Omega, \gnons, \grules, P )$ be such an SCFG.
Let $\gnulls \subseteq \gnons$ be the set of null states in $\gnons$, {\bf excluding} post-emit states.
Let $\gemits \subset \gnons$ be the set of post-emit states.
Let $\gbifs \subseteq \gnons$ be the set of bifurcation states.

We here define the grammar $G'$,
\[
G' = \left( \Omega, \gnons', \grules', P' \right)
\]
which is equivalent to $G$ in sequence but has no null subtrees.

The new states $\gnons'$ are defined as follows.
Start with $\gnons$; remove bifurcation states;
then introduce a new state $N'$ for every null $N \in \gnulls$ and three new states $B^0, B', B^2$ for every bifurcation $B \in \gbifs$.

\[
\gnons' = 
\left( \bigcup_{N \in \gnulls} \{ N' \} \right) \cup \left( \bigcup_{B \in \gbifs} \{ B^0, B', B^2 \} \right) \cup \gnons \setminus \gbifs
\]

The idea is that $N'$ is a null state that is equivalent to null state $N$ for non-empty sequence.
That is, for any terminal sequence $S$,

\[
P'(S|N') = \left\{ \begin{array}{ll} P(S|N) & \mbox{if $S \neq \epsilon$} \\ 0 & \mbox{if $S = \epsilon$} \end{array} \right.
\]

Similarly, $B'$ is a null state that is equivalent to bifurcation state $B$ for non-empty sequence,
while $B^2$ is a null state that is is equivalent to $B$ for parse trees where both children of $B$ are non-empty.

In contrast, $B^0$ is {\bf exactly} equivalent to $B$ in sequence.
However, null subtrees in $B^0$ are explicitly accounted for,
their probabilities factored into the transitions $B' \to L'$ and $B' \to R'$ and the termination $B^0 \to \epsilon$,
using the inside probabilities for empty sequences, $P(\epsilon|L)$ and $P(\epsilon|R)$.

The probabilities $P(\epsilon|X)$ are related by the following system of equations

\[
P(\epsilon|X) = P(X \to \epsilon) + \sum_Y P(X \to Y) P(\epsilon|Y) + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R)
\]

which are nonlinearly coupled but may be solved numerically, e.g., by the Newton-Raphson method,
or by iterated approximation starting from a lower bound $P(\epsilon|X) \geq 0$.

The new rules and their probabilities are
\[
\begin{array}{rrcl}
\forall \rho \in (\grules \setminus \allrules{\gbifs}): & P'(\rho)             & = & P(\rho) \\
                                \forall N \in \gnulls: & P'(N' \to \epsilon)  & = & 0  \\
                          \ldots \forall B \in \gbifs: & P'(N' \to B')        & = & P(N \to B) \\
                         \ldots \forall M \in \gnulls: & P'(N' \to M')        & = & P(N \to M) \\
                         \ldots \forall E \in \gemits: & P'(N' \to E)         & = & P(N \to E) \\
               \forall (A \to B) \in \rightside{\gbifs}: & P'(A \to B^0)        & = & P(A \to B) \\
              \forall (B \to LR) \in \leftside{\gbifs}: & P'(B^0 \to \epsilon) & = & P(B \to LR)\ P(\epsilon | L)\ P(\epsilon | R) \\
                                                       & P'(B^0 \to B')       & = & 1 \\
                                                       & P'(B' \to \epsilon)  & = & 0 \\
                                                       & P'(B' \to L')        & = & P(B \to LR)\ P(\epsilon | R) \\
                                                       & P'(B' \to R')        & = & P(B \to LR)\ P(\epsilon | L) \\
                                                       & P'(B' \to B^2)       & = & 1 \\
                                                       & P'(B^2 \to L'\ R')   & = & P(B \to LR) \\
\end{array}
\]

Note that grammar $G'$, like $G$, has RNA normal-form states.
Note also that $G'$ is not, in general, normalized by production rule;
however, $G'$ is equivalent to $G$ in parse and is therefore normalized by parse tree (if $G$ is).

\paragraph{Eliminating null states.}

We now proceed to eliminate null cycles from $G'$.
Since null subtrees have been eliminated, the remaining null cycles use only transition rules.

We create the grammar
\[
G'' = \left( \Omega, \gnons', \grules'', P'' \right)
\]
which has the same nonterminals as $G'$, but different rules and rule probabilities.

Let $\gnulls' \subseteq \gnons'$ be the set of null states in $\gnons'$, {\bf including} post-emit states.

Define ${\bf t}$, the {\em transition matrix} of $G'$, as $t_{XY} = P'(X \to Y)$ for all $X,Y \in \gnons'$.
The effective transition probability $q_{XY}$ between two states $X,Y$
sums over all paths through null states (in the summand, $n$ is the length of the null path):
\[
{\bf q} = \sum_{n=0}^\infty {\bf t}^n = ({\bf 1}-{\bf t})^{-1}
\]
Here ${\bf 1}$ is the identity matrix.
The matrix inverse may be computed in the usual ways (Gauss-Jordan elimination, LU decomposition, etc.)

Since the bifurcation states of $G'$ explicitly generate non-empty sequence,
the system of equations relating the probabilities $P'(\epsilon|X)$ is completely linear
and may be solved in the same way.
Let $u_X = P'(\epsilon|X)$ and $v_X = P'(X \to \epsilon)$.
Then

\[
{\bf u} = {\bf v} + {\bf tu}
\]

whose solution is ${\bf u} = {\bf qv}$.
Thus $P'(\epsilon|X) = \sum_Y q_{XY} P'(Y \to \epsilon)$.

We now define $P''$ as follows.
\begin{itemize}
\item For all rules $\rho \in \grules'$ that are {\bf not} transitions or terminations, set $P''(\rho) = P'(\rho)$.
\item For terminations from null states $X \in \gnulls'$, set $P''(X \to \epsilon) = u_X$.
(NB this may create some terminations $X \to \epsilon$ that were not present in $G'$.)
\item For transitions from null states $X \in \gnulls'$ to non-null states $Y \notin \gnulls'$, set $P''(X \to Y) = q_{XY}$.
(NB this may create some transitions $X \to Y$ that were not present in $G'$.)
\item For transitions {\bf to} null states $X \in \gnulls'$, set $P''(A \to X) = 0$.
\end{itemize}

Although we have left null states in $G''$, they now have no incoming transitions
and are inaccessible unless they are post-bifurcation states (i.e.,states which appear on the right-hand side of bifurcation rules), post-emit states, or the start (root) state.
All other null states can therefore be dropped from $\gnons''$.

\paragraph{Restoring null states.}

Suppose that $\rho''$ is a rule in $G''$.
Algorithm~\ref{alg:restoreTransitions} samples from the distribution of equivalent parse subtrees in $G'$ (possibly containing null cycles).
In order to sample from $P(T'|T'')$ we need simply map Algorithm~\ref{alg:restoreTransitions} to each rule in $T''$.

For parameter estimation by Expectation Maximization and some other applications, it is useful to know the expected number
of times that a transition was used, summed over the posterior distribution of parse trees (including those with null cycles).
If $n''(\rho'')$ is the expected number of times that transition $\rho''$ was used according to an Inside-Outside computation on $G''$,
then the corresponding expectations $n'(\rho')$ are given by

\begin{eqnarray*}
       n'(X \to Y) & = & n''(X \to Y) \frac{P''(X \to Y)}{q_{XY}} + \sum_Z n''(X \to Z) \frac{P''(X \to Y) q_{YZ}}{q_{XZ}} \\
n'(X \to \epsilon) & = & n''(X \to \epsilon) \frac{P''(X \to \epsilon)}{P''(\epsilon | X)}
 + \sum_W n''(W \to \epsilon) \frac{q_{WX} P''(X \to \epsilon)}{P''(\epsilon | W)}
\end{eqnarray*}

Expectations for other rules (bifurcations and emissions) are the same for $G'$ as $G''$.

\paragraph{Restoring null bifurcations.}

Suppose that $\rho'$ is a rule in $G'$.
Algorithm~\ref{alg:restoreBifurcations} samples from the distribution of equivalent parse subtrees in $G$ (possibly containing null subtrees).
This algorithm also calls Algorithm~\ref{alg:sampleNullSubtree}, which samples from the distribution of empty subtrees rooted at a particular nonterminal.
In order to sample from $P(T|T')$ we need simply map Algorithm~\ref{alg:restoreBifurcations} to each rule in $T'$.

If $n'(\rho')$ is the expected number of times that rule $\rho'$ was used by $G'$,
then the corresponding expectations $n(\rho)$ are given by

\begin{eqnarray*}
    n(B \to L\ R) & = & n'(B' \to L') + n'(B' \to R') + n'(B^0 \to L'\ R') + d(B \to L\ R) \\
       n(X \to Y) & = & n'(X \to Y) + d(X \to Y) \\
n(X \to \epsilon) & = & n'(X \to \epsilon) + d(X \to \epsilon)
\end{eqnarray*}
Expectations for emissions are the same for $G$ as $G'$.
In the above expressions $d(\rho)$ is the expected usage of rule $\rho$ by null subtrees:

\[
d(\rho) = \sum_{(B \to LR) \in \grules} \left[
n'(B' \to L') c_R(\rho)
+ n'(B' \to R') c_L(\rho)
+ n'(B^0 \to \epsilon) (c_L(\rho) + c_R(\rho))
\right]
\]

where $c_W(\rho)$ is the expected usage of rule $\rho$ by an empty parse tree rooted at $W$,
given by

\begin{eqnarray*}
c_X(\rho) P(\epsilon | X) & = & P(X \to \epsilon) \delta_{\rho=(X \to \epsilon)}
 + \sum_Y P(X \to Y) P(\epsilon|Y) \left( c_Y(\rho) + \delta_{\rho=(X \to Y)} \right)
\\ & &
 + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R) \left( c_L(\rho) + c_R(\rho) + \delta_{\rho=(X \to LR)} \right)
\end{eqnarray*}

where $\delta_U$ is the Kronecker delta (1 if condition $U$ is true, 0 if it is false).
Note that in contrast to the system of equations for $P(\epsilon|X)$, this is a linear system of equations
of the form ${\bf c} = {\bf Mc} + {\bf k}$, which may be solved by matrix inversion:
${\bf c} = ({\bf 1}-{\bf M})^{-1} {\bf k}$.

\subsubsection*{Dynamic programming algorithms for inference}

Once we have performed the transformations described above to remove
null cycles from the multiple-sequence SCFGs generated by our
model-construction algorithm, we can use the standard CYK, Inside and
Outside algorithms for multiple-sequence SCFGs given by Sankoff \cite{Sankoff85}.

The asymptotic time and memory complexities of our inference algorithms
are essentially the same as for the Sankoff algorithm \cite{Sankoff85}:
the DP algorithms take memory $O(A \cdot L^{2N})$ and time $O(B \cdot L^{3N})$
for $N$ sequences of length $L$, where $A$ is the number of
(accessible) states in the multiple-SCFG and $B$ is the number of
bifurcations.
Note that $A$ and $B$ are also dependent on $N$ (see ``The TKFST model
on a three-taxon phylogeny'').

Exact inference on a star phylogeny with $N$ extant sequences therefore
has complexities $O(A \cdot L^{2N})$ and $O(B \cdot L^{3N})$ in space and
time for a multiple-SCFG with $A$ states and $B$ bifurcations. As
described earlier, in practice we frequently have expert knowledge
about the structures or evolutionary histories of the sequences of
interest, such as a curated multiple alignment.  We can use these
constraints to reduce the accessible volume of the DP matrix \cite{Holmes2005}.
Algorithms \ref{alg:inside}--\ref{alg:cyktraceback} give the Inside, CYK,
Outside and CYK traceback algorithms for a three-taxon star phylogeny,
constrained using the ``fold envelopes'' described below.

\paragraph{Fold Envelopes.}
We use the fold envelope concept \cite{HolmesRubin2002a,Holmes2004}
to constrain the set of structures which our algorithms consider.
A fold envelope $\foldenv^{(X)}$ for a sequence $X$ is a set of coordinate pairs satisfying
\beqn \eqnlabel{foldenv}
\foldenv^{(X)} \subset \left\{ (i,j):0 \le i \le j \le L^{(X)} \right\} \, ;
\quad L^{(X)} = |X| \, .
\eeqn
We consider a subsequence $x_{i+1} \dots x_{j}$ only if the corresponding coordinate pair $(i,j) \in \foldenv^{(X)}$.
The unconstrained fold envelope has set equality in \eqnref{foldenv}.

We use two orderings for sequences in the fold envelopes.  An inside-outside ordering is used for the iteration 
in the Inside algorithm: Subsequences are ordered such that each successive subsequence 
contains all previous subsequences in the fold envelope.  More precisely, subsequences in $\foldenv^{(X)}$ are sorted  
in the same order as coordinate pairs $(i,j)$ are generated by the
iteration $\{\,\mathrm{for}\, i = L^{(X)} \,\mathrm{to}\, 0 \,\,\{
\mathrm{for}\, j = i \,\mathrm{to}\, L^{(X)} \}\,\}$.

The Outside algorithm uses an outside-inside ordering, where subsequences starting at a fixed $i$ are reverse-sorted 
with respect to the inside-outside ordering described above.  Subsequences in $\foldenv^{(X)}$ are then sorted 
in the same order as coordinate pairs $(i,j)$ are generated by the 
iteration $\{\,\mathrm{for}\, i = 0 \,\mathrm{to}\, L^{(X)} \,\,\{ \mathrm{for}\, j = L^{(X)}\, \mathrm{to}\, i \}\,\}$.
We frequently refer to subsequences by their index in the fold envelope.
The $\mth$ subsequence in $\foldenv^{(X)}$ is labeled $m^{(X)}$ and corresponds to the coordinate pair $(i_{m}, j_{m})$.  The 
index of a pair $(i,j) \in \foldenv^{(X)}$ is $n^{(X)}[i,j]$.

In order to take full advantage of the reduction in computational complexity offered by restricting our inference algorithms to subsequences contained in the fold envelopes, we must avoid iterating over unreachable combinations of subsequences (unreachable because they are not permitted by the fold envelope constraints).
An efficient implementation relies on iterators over subsequences in the fold envelope which are connected by production rules of the ensemble grammar.
Inward and outward emission connections for a sequence $X$, specifying which subsequence is reachable from 
a given subsequence $m^{(X)}$ and ensemble state $\bvec{b}$, are defined as
\begin{align*}
c_{in}\left(\bvec{b}; m^{(X)}\right) &= n^{(X)} \left[ i_{m} + \Delta^{(X)}_L(\bvec{b}), j_{m} - \Delta^{(X)}_R(\bvec{b}) \right] \\
c_{out}\left(\bvec{b}; m^{(X)}\right) &= n^{(X)} \left[ i_{m} - \Delta^{(X)}_L(\bvec{b}), j_{m} + \Delta^{(X)}_R(\bvec{b}) \right] \, ,
\end{align*}
where the quantities $\Delta^{(X)}_L(\bvec{b})$ and $\Delta^{(X)}_R(\bvec{b})$ are the lengths of the left and right emissions of the ensemble state $b$ to the sequence $X$.
(Recall that the $\mth$ subsequence in $\foldenv^{(X)}$ is labeled $m^{(X)}$ and corresponds to the coordinate pair $(i_{m}, j_{m})$.)
The emission connection is undefined if the corresponding subsequence is not in the fold envelope.
Inward, outward-left and outward-right bifurcation connections, specifying which subsequences are connected by bifurcation production rules of the ensemble SCFG,
are defined for a subsequence $n^{(X)}$ as
\begin{align}
b_{in}(n^{(X)}) &= \left\{ \left( n^{(X)}_L,n^{(X)}_R \right): i_{n^{(X)}_L} = i_{n^{(X)}},\, j_{n^{(X)}_L} = i_{n^{(X)}_R},\, j_{n^{(X)}_R} = j_{n^{(X)}} \right\} \\
b_{out,L}(n^{(X)}) &= \left\{ \left( n^{(X)}_O,n^{(X)}_L \right): i_{n^{(X)}_L} = i_{n^{(X)}_O},\, j_{n^{(X)}_L} = i_{n^{(X)}},\, j_{n^{(X)}} = j_{n^{(X)}_O} \right\} \\
b_{out,R}(n^{(X)}) &= \left\{ \left( n^{(X)}_O,n^{(X)}_R \right): i_{n^{(X)}} = i_{n^{(X)}_O},\, j_{n^{(X)}} = i_{n^{(X)}_R},\, j_{n^{(X)}_R} = j_{n^{(X)}_O} \right\}
\end{align}
We generally write out explicit subsequence coordinate pairs $(i,j)$ when their usage will make mathematical formulae clearer and fold-envelope labels $n^{(X)}$ when writing pseudocode.

\paragraph{The Inside algorithm.}
The Inside algorithm is used to calculate the likelihood of sequences under an ensemble model.  It is analogous to the Forward algorithm for HMMs.

The inside probability $\alpha_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)}\right)$ is the summed probability of the triplet of subsequences 
$\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ for sequences $X, Y, Z$.
under all paths through the model which are rooted in state $\bvec{a}$.  
\algref{inside} gives pseudocode for the fold-envelope version of the Inside algorithm.  The subroutines 
$\mathrm{calcTransEmitProb}$, $\mathrm{calcLBifurcProb}$ and $\mathrm{calcRBifurcProb}$
used in the Inside algorithm are defined below.

The transition and emission probability $\mathrm{calcTransEmitProb}(\bvec{a}; \cdot)$ can be calculated 
by iterating over ensemble states $\bvec{b}$ which connect the subsequence triplet $\left(n^{(X)}, n^{(Y)}, n^{(Z)}\right)$ to others in the fold envelopes.
Pseudocode for the constrained calculation is given in \algref{insidetransemit}.

The left-bifurcation probability for an ensemble state $\bvec{a}$
bifurcating to two ensemble states,
\newline
$\mathrm{calcLBifurcProb}\left( \bvec{a}; n^{(X)}_L,n^{(X)}_R,n^{(Y)}_L,n^{(Y)}_R,n^{(Z)}_L,n^{(Z)}_R \right)$, is
\[ \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} P (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \alpha_{\bvec{c}}\left( n^{(X)}_L,n^{(Y)}_L,n^{(Z)}_L\right) \, \alpha_{\bvec{b}}\left( n^{(X)}_R,n^{(Y)}_R,n^{(Z)}_R\right) \]
and the right-bifurcation probability for an ensemble state $\bvec{a}$
bifurcating to two ensemble states,
\newline
$\mathrm{calcRBifurcProb}\left( \bvec{a}; n^{(X)}_L,n^{(X)}_R,n^{(Y)}_L,n^{(Y)}_R,n^{(Z)}_L,n^{(Z)}_R \right)$, is
\[ \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} P (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \alpha_{\bvec{b}}\left( n^{(X)}_L,n^{(Y)}_L,n^{(Z)}_L\right) \, \alpha_{\bvec{d}}\left( n^{(X)}_R,n^{(Y)}_R,n^{(Z)}_R\right) . \]
The boundary condition of the probability of 0-length subsequences is determined by the probability of transitions to $\Send$.
The termination condition is 
\[ P \left( X, Y, Z \right) = \alpha_{\Sstart}\left(N^{(X)}, N^{(Y)}, N^{(Z)}\right) \, , \]
where $\Sstart$ is the unique start state of the ensemble grammar and $N^{(X)}$ is the outermost subsequence for sequence $X$, etc.

Note that we are assuming that the transformations described in ``Eliminating null states'' have been performed, such that there are no cycles of $\Snull$ states as well as no empty bifurcations.

\paragraph{The CYK algorithm.}
The CYK algorithm is used to calculate the probability of the
most-likely state path (or parse) capable of generating the input
sequences.  It is analogous to the Viterbi algorithm for HMMs.

The CYK algorithm can be obtained from the Inside algorithm by replacing sums over paths through the ensemble model
with the $\max$ operation.
The CYK probabilities for indices $\gamma_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)}\right)$ then represent 
the probability of the most likely path through the model generating 
the triplet of subsequences $\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$.

The resulting CYK algorithm is shown in \algref{cyk}.
The subroutine $\mathrm{calcTransEmitProb}$ is defined in \algref{cyktransemit}.  The subroutines $\mathrm{calcLBifurcProb}$ and $\mathrm{calcRBifurcProb}$
used in the CYK algorithm are defined as 
\[ \max_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} P (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \gamma_{\bvec{c}}\left( n^{(X)}_L,n^{(Y)}_L,n^{(Z)}_L\right) \, \gamma_{\bvec{b}}\left( n^{(X)}_R,n^{(Y)}_R,n^{(Z)}_R\right) \]
and
\[ \max_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} P (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \gamma_{\bvec{b}}\left( n^{(X)}_L,n^{(Y)}_L,n^{(Z)}_L\right) \, \alpha_{\bvec{d}}\left( n^{(X)}_R,n^{(Y)}_R,n^{(Z)}_R\right) . \]

\paragraph{The Outside algorithm.}
The Outside algorithm is primarily used an an intermediary for posterior decoding on the model.  It is analogous to the Backward algorithm for HMMs.

The outside probability $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)}\right)$ for an ensemble state $\bvec{b}$
is the summed probability of the sequences $X, Y, Z$
under all paths through the ensemble model which are rooted in the start state of the model, excluding all paths for the triplet of subsequences $\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$
which are rooted in the ensemble state $\bvec{b}$.
\algref{outside} gives pseudocode for the fold-envelope version of the Outside algorithm.
The subroutines $\mathrm{calcTransEmitProb}$, $\mathrm{calcLBifurcProb}$ and $\mathrm{calcRBifurcProb}$
used in the Outside algorithm are defined below.

As with the Inside and CYK algorithms, the transition and emission probability $\mathrm{calcTransEmitProb}$ can be calculated efficiently using the subsequence connections defined earlier (\algref{outsidetransemit}).
The left-bifurcation probability $\mathrm{calcLBifurcProb}\left( \bvec{b}; n^{(X)}_O,n^{(X)}_L,n^{(Y)}_O,n^{(Y)}_L,n^{(Z)}_O,n^{(Z)}_L \right)$ is
\[ \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} P (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \beta_{\bvec{a}}\left( n^{(X)}_O,n^{(Y)}_O,n^{(Z)}_O\right) \, \alpha_{\bvec{c}}\left( n^{(X)}_L,n^{(Y)}_L,n^{(Z)}_L\right) \]
and the right-bifurcation probability $\mathrm{calcRBifurcProb}\left( \bvec{b}; n^{(X)}_O,n^{(X)}_R,n^{(Y)}_O,n^{(Y)}_R,n^{(Z)}_O,n^{(Z)}_R \right)$ is
\[ \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} P (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \beta_{\bvec{a}}\left( n^{(X)}_O,n^{(Y)}_O,n^{(Z)}_O\right) \, \alpha_{\bvec{d}}\left( n^{(X)}_R,n^{(Y)}_R,n^{(Z)}_R\right) \]
The boundary condition is just
\[
\beta_{\Sstart} \left(N^{(X)}, N^{(Y)}, N^{(Z)}\right) = 1\, ,
\]
where $N^{(X)}$ is the outermost subsequence for sequence $X$, etc.

\paragraph{The CYK traceback algorithm.}
The CYK traceback algorithm, in combination with the CYK algorithm, is
used to find the most-likely state path generating the extant sequences
(in other words, the maximum-likelihood parse generating the observed data).
It is analogous to the Viterbi traceback algorithm for HMMs.
\algref{cyktraceback} gives the constrained CYK traceback algorithm.

% Results and Discussion can be combined.
\newpage
\section*{Results}

\subsection*{Automated grammar construction}

We implemented our model construction algorithm on the three-taxon
star phylogeny.
Given a singlet transducer modeling ancestral structures and a branch
transducer modeling structural evolution, our Perl modules generate
C++ code for the corresponding jointly-normalized three-sequence
(Triplet) SCFG.
Any model of structural evolution which can be represented as a Pair
SCFG and factored into singlet and branch transducers is permitted as input to
the packages, allowing for flexible, automated model design.

% Simultaneous alignment and folding of three RNA sequences is highly-demanding computationally.
% The completely-unconstrained algorithm requires time
% $O(L^{3N})$ and memory $O(L^{2N})$ and so, for $N \geq 3$, becomes very expensive.
% Even constrained versions are expensive:
% e.g., inferring the common ancestor of three RNA sequences {\em whose secondary structures have been experimentally determined}
% still takes at least $O(L^3)$ time and memory.

% Despite this expense, construction and analysis of the three-sequence model and algorithm is a useful step towards developing more practical inference algorithms.
% This point is illustrated by several papers in the molecular evolution literature,
% which describe algorithms for the practical reconstruction of indel histories in DNA and protein sequences
% \cite{HolmesBruno2001,JensenHein2002,RedelingsSuchard2005,PatenHolmesBirney2008}.
% In these papers, the generalization of a model from two sequences to three (or more)
% is followed by the observation that the resulting multiple-sequence problem is very expensive in memory;
% the memory complexity is then somehow reduced to the complexity of the two- or (in the case of \cite{JensenHein2002}) the three-sequence problem,
% with an accompanying tradeoff in time or accuracy.

% In particular, the work of Redelings and Suchard \cite{RedelingsSuchard2005}
% reduces a three-sequence inference problem
% into two Metropolis-Hastings steps (a {\em Propose} step and an {\em Accept} step),
% each of which reduces to a two-sequence inference problem.
% Thus, one expensive three-way alignment is decomposed into two less-expensive pairwise alignments.
% In the {\em Propose} step, two sequences are aligned and their common ancestor is sampled.
% In the {\em Accept} step, the proposed ancestor is aligned to the remaining sequence,
% and the move is accepted with a probability given by a Hastings ratio.

% The Redelings-Suchard approach is motivated by the fact that the
% two-sequence case is often tractable where the three-sequence case is not.
% This is especially true of the memory-intensive analsis of RNA secondary structure.
% While the three-sequence inference algorithms which we report here are computationally expensive,
% we believe that these results can be used to develop reduced-complexity algorithms
% (e.g., similar to those of Redelings and Suchard),
% allowing memory complexities similar to the pairwise case.
% The work described here may hopefully be regarded as one step towards the construction of such approaches
% (Metropolis-Hastings or otherwise).

\subsection*{A simple model of RNA structural evolution}

We illustrated our method for building models of structured sequences
using a simplified model which was introduced in previous work,
the TKF Structure Tree model \cite{Holmes2004}, a simple probabilistic
model of the evolution of RNA structure.

The TKF Structure Tree (TKFST) model is based on the
Thorne-Kishino-Felsenstein (TKF) model of the stochastic evolution of
primary sequences via indel events \cite{ThorneEtal91}.  In the
original TKF model, sequence evolves under a time-homogeneous linear
birth-death-immigration process \cite{Kendall1948}.  Single characters
are inserted with rate $\lambda$ and deleted with rate $\mu$.  At
equilibrium, sequences obey a geometric length distribution with
parameter $\kappa$.  Although this model has flaws (e.g., it lacks
affine gap penalties, rate heterogeneity and context-dependent
mutation rates), it illustrates many of the key ideas used by more
sophisticated indel models, notably the possibility for systematic
derivation of pairwise alignment automata from first principles via
analysis of birth-death processes \cite{Feller71,ThorneEtal91}.

The TKF Structure Tree model is an extension of the TKF model to RNA
structure.  In this model, loop and stem regions are
mutually nested (\figref{tkfst}).  Both loops and stems have geometric length
distributions with parameters $\kappa_1$ and $\kappa_2$.  Single bases
are inserted and deleted in loops with rates $\lambda_1$ and $\mu_1$;
similarly, base-pairs are inserted and deleted in stems with rates
$\lambda_2$ and $\mu_2$.  Insertions of a new stem into an existing
loop sequence (or deletions of an existing stem) occur at the same
rate as single-base insertions (or deletions) and can model
large-scale structural changes (\figref{tkfst-mutations}).

We parametrized the singlet and branch transducers of the TKFST model
using estimates reported by a phylo-grammar for RNA secondary
structure prediction, \pfold\ \cite{KnudsenHein2003}, and an
implementation of pairwise alignment for the TKF Structure Tree model,
\evoldoer\ \cite{Holmes2004}.
The equilibrium distributions of unpaired and paired nucleotides of
the singlet and branch transducers, as well as the substitution models
of unpaired and paired nucleotides of the branch transducers, were
derived from the substitution rate matrices of the \pfold\ program.
These rate matrices, which have proven useful for RNA structure
prediction \cite{KnudsenHein2003,DowellEddy2004,BradleyEtAl2008},
were derived from the Bayreuth tRNA database \cite{SprinzlEtAl1998}
and the European large subunit rRNA database \cite{DeRijkEtAl1998}.

This continuous-time model corresponds to a Pair SCFG and as such
fits neatly into our modeling framework once the probability
distribution is appropriately factored into marginal and conditional
distributions (generated by singlet and branch transducers).
Tables \ref{tab:tkfstsinglet-types} and \ref{tab:tkfstsinglet} show the states 
and transitions of the singlet transducer (single-sequence SCFG) 
which generates ancestral sequence under the Structure Tree model.
Tables \ref{tab:tkfstbranch-types} and \ref{tab:tkfstbranch} show the states
and transitions of the branch transducer (conditionally-normalized Pair SCFG)
which evolves a sequence and structure along a branch of the phylogenetic tree.

The equilibrium distribution and transition probabilities between states of the TKFST model can be 
expressed in terms of functions of the evolutionary time along a branch
and the insertion and deletion rates $\lambda_1$ and $\mu_1$ of the model.
The length of ancestral sequences is geometric in $\kappa_1$ (\tabref{tkfstsinglet}), defined as
\[
\kappa_1 = \lambda_1 / \mu_1 \, .
\]
The three functions $\alpha(t)$, $\beta(t)$ and $\gamma(t)$ which govern 
the transition probabilities in \tabref{tkfstbranch} are defined
for loop sequences as
\begin{align}
  \alpha_1 (t) &= \exp \left( -\mu_1 t \right) \nonumber \\
  \beta_1 (t) &= \frac{\lambda_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) } \nonumber \\
  \gamma_1 (t) &= 1 - \frac{\mu_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\left( 1 - \exp (- \mu_1 t) \right) \left(\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) \right) } \nonumber
\end{align}
and similarly for stem sequences \cite{Holmes2004}.

\subsection*{Assessing TKFST as a model of RNA structure}

The TKFST model, like the original TKF model, probably needs refinements in order to be useful.
For example, it fails to model certain phenomena observed in natural
RNA structures (such as base-stacking or tetraloops) and in alignments
of those structures (such as helix slippage).
We assessed its appropriateness as a model of RNA structural evolution
by conducting benchmarks of its capabilities for (1) multiple sequence
alignment of structured RNAs, summing over all possible structures,
and (2) structure prediction of homologous structured RNAs and
comparing its performance to one of the best-performing pairwise
SCFGs for RNA multiple alignment \cite{BradleyPachterHolmes2008}.  The
results of these benchmarks, reported in \tabref{bralibaseII-align}
and \tabref{bralibaseII-struct}, suggest that TKFST is a
useful guide for deriving more complicated models of RNA evolution:
while it has relatively poor sensitivity (but high positive predictive
value) as a base-pairing predictor, it is competitive with one of the
most accurate RNA multiple sequence alignment programs \cite{BradleyPachterHolmes2008}.

TKFST's poorer performance at base-pairing prediction is likely
due to its much-simpler model of RNA structure.
The richer grammar, as described in \cite{Holmes2005},
is richer than TKFST:
excluding the substitution model, it has 14 free parameters
(compared to TKFST's 4),
uses an affine gap penalty
(compared to TKFST's linear gap penalty),
and explicitly models structural features
such as multiple-branched loops, symmetric/asymmetric bulges, and minimum loop lengths.
Unlike TKFST, the richer grammar is structurally unambiguous:
a one-to-one mapping exists from structures to parse trees.
Although we use the TKFST model as an illustrative example of a Pair
SCFG that can be extended with our method, the model is not
fundamental to our approach and can be replaced by a different
and more realistic pairwise model, such as the pairwise SCFG used in 
these comparisons \cite{BradleyPachterHolmes2008}.
We anticipate that further improvements should be possible by reviewing
other comparisons of SCFGs at structure prediction,
such as the study by \cite{DowellEddy2004}.

\subsection*{The TKFST model on a three-taxon phylogeny}

We used our model-construction algorithm to build the grammar
corresponding to the TKFST model acting on a star phylogeny with three
(extant) leaf sequences and a single (unobserved) ancestral
sequence. We chose this phylogeny for two reasons: (1) it is the
simplest extension of the well-studied, standard two-sequence (Pair
SCFG) model and (2) algorithms on a phylogeny with three leaves
should be sufficient for ergodic sampling of reconstructions on any larger phylogeny,
using an MCMC kernel similar to that of \cite{JensenHein2002}.

The statistics of the TKFST model on the three-taxon phylogeny
illustrate the advantages of our procedure for model
construction.  While the singlet and branch transducers are relatively simple---the singlet transducer,
shown in \tabref{tkfstsinglet}, has 7 total states and 2 bifurcation states and
the branch transducer, shown in \tabref{tkfstbranch}, has 21 total states
and 6 bifurcation states---the ensemble model of three extant sequences 
is very complex.
The naive exponential upper-bound gives a maximal state space of size $O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.
Using our uninformed search algorithm, we determine that there are 287 accessible states
and 686 possible transitions between these states (compare with the $287^2 \simeq 8 \cdot 10^4$ transitions
estimated with the exponential calculation).  After performing the
transformations described in ``Eliminating null cycles in SCFGs'' to
eliminate useless windback states,
the ensemble model has a reduced state space with 230 states,
albeit at the cost of extra transitions, bring the total to 1,789 transitions
(here we are trading reduced memory complexity, which is linear in the number
of states, for increased time complexity, which is linear in the number of transitions).
Note that both before and after the reduction in complexity,
the total number of states and transitions are less than the approximate bounds
of $O(a \cdot b \cdot N) = O(7 \cdot 21 \cdot 3) \Rightarrow 441$ states
and
$O(a \cdot b^2 \cdot N) = O(7 \cdot 21^2 \cdot 3) \Rightarrow 9,261$ transitions
given in ``The composition algorithm.''

% to do: dump output of tripletscfg.pl --tkfst into supplement
Text S? contains the complete model specification, including states and transitions,
produced by our model composition algorithm.  The extreme complexity of the
ensemble model, despite the simplicity of the underlying model of RNA
structure, makes clear the necessity for automated procedures for
model construction.

We implemented constrained maximum-likelihood inference of the structural alignment 
and ancestral structure of three extant sequences (Algorithms \ref{alg:cyk}-\ref{alg:cyktraceback})
in a C++ program (\indiegram).
For tractability, \indiegram\ uses the concept of fold envelopes
described earlier to limit the fold space considered by the CYK algorithm,
permitting structural information for the three extant sequences to be
(optionally) supplied as input.
If no structural information is supplied, then \indiegram\
uses a single-sequence SCFG to estimate a set of plausible folds \cite{Holmes2005},
which are used to constrain the CYK algorithm.

The inference algorithms in \indiegram\ could be further constrained
to enforce, for example, a fixed multiple alignment or consensus
structure of extant sequences.  While experimentally-determined
structures of individual RNAs are rare, curated deep sequence
alignments, such as those constructed for ribosomal RNAs
\cite{Gutell93}, are more frequently available.  By constraining the
inference algorithms with such sequence alignments, the memory and
time complexity of the algorithms could be dramatically reduced.  Such
constraints can be naturally expressed with ``alignment envelopes,''
the sequence-level counterpart of fold envelopes \cite{Holmes2005}.
However, in this paper we focus on model construction and inference
algorithms and postpone exploration of heuristics and constraints of
these algorithms for future work.

\subsection*{Reconstructing small RNAs with the TKFST model}

While reconstructing large RNAs such as ribosomal subunits is
currently computationally-inaccessible without further heuristics to
constrain our algorithms, reconstructing small RNAs of biological
interest will soon be feasible.
We used the TKF Structure Tree model on a 3-taxon phylogeny to 
reconstruct most-probable ancestral structures of three \emph{nanos} 3'
translational control elements (TCEs) and three tRNAs, two small RNAs whose
structures are well-known and which are accessible given
current computational limitations. While the phylogenetic trees
relating the tested sequences have short branch lengths, making the
reconstruction problem easy by forcing the reconstructed structures to
be essentially-identical to those of one of the extant RNAs
(reconstructed structures not shown; see \indiegramurl), these
are nonetheless useful proof-of-concept datasets.

\tabref{complexities} shows estimates of the memory and time required
to reconstruct biologically-interesting subunits of the two RNAs
tested, the \emph{nanos} 3' translational control element and tRNAs,
as well as two small RNAs which show significant structural
divergence, the Y RNAs and Group II introns, and therefore promise to be
interesting candidates for ancestral reconstruction.

\subsection*{The dependence of reconstruction accuracy on outgroup
  branch length}

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length.
We simulated the evolution of RNAs under the TKFST model along
3-taxon phylogenies, where we kept the branch lengths of the two
sibling species constant and varied the branch length of the
outgroup.  We fixed the tree topology as $((X:1,Y:1):1,Z:z)$, where
the branch length $z$ of the outgroup was varied between 0.5 and 3.0
and a branch length of 1 corresponds to one expected substitution per site
in loop sequence.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$, loops of length $\in [3, 10]$ and
stems of lengths $\in [1, 20]$.

We expected to find a ``transition point'' in our reconstruction
accuracy near the point $z \simeq 1$, corresponding to an outgroup
distance equal to the distance from either of the sibling species to
the most recent common ancestor.  Our results, shown in
\figref{simulations}, indicate such a transition point for alignment
accuracy but not for ancestral reconstruction accuracy.
% to do: here now

\cite{Mossel2003}

While these results correspond to a best-case scenario in which we
have a perfectly-parametrized evolutionary model, recent results
demonstrating that inference of sequence alignment is surprisingly
robust to mis-parametrized models \cite{Lunter2007} suggest that the same
may be true for inference of ancestral structures.

\newpage
\section*{Discussion}

Following the conception of paleogenetics \cite{PaulingZuckerkandl63},
a large number of synthetic reconstructions of ancient proteins have been
reported in the literature
\cite{MalcolmEtAl90,StackhouseEtAl90,ZhangRosenberg2002,GaucherEtAl2003,ThomsonEtAl2005,ChangEtAl2002,SunEtAl2002,OrtlundEtAl2007}.
There are also the beginnings of a movement to reconstruct ancient DNA
\cite{IvicsEtAl91,AdeyEtAl94,BlanchetteEtAl2004,NoonanEtAl2006,Gaucher2007b,EliasTuller2007,PatenHolmesBirney2008}.
Given the importance of the RNA world hypothesis to
current discussions of the origin of life \cite{MarintchevWagner2004,Muller2005,CavalierSmith2006,WilliamKoonin2006,Forterre2006,Yakhnin2007,DanchinEtAl2007},
the many modern-day relics of this world
 \cite{LeeEtAl93,LoweEddy99,Eddy01,MandalEtAl2003}
and the recent proposal of a structural model for the primordial ribosome
 \cite{SmithEtAl2008},
we believe that phylogenetic reconstruction of ancient RNA is a significant
problem, deserving of strong bioinformatics support.

% Prior art
The work reported in this paper builds on substantial prior art in the areas of evolutionary modeling and ancestral reconstruction.
The reconstruction of ancient sequences was first proposed in 1963 by Pauling and Zuckerkandl \cite{PaulingZuckerkandl63};
current applications of this idea, mostly using substitution models, are surveyed in the book edited by Liberles \cite{Liberles2007}.
Many algorithms in phylogenetics implicitly reconstruct substitution histories,
whether by parsimony \cite{Edwards63,HendyPenny82} or likelihood \cite{Felsenstein81}.
There has been much effort to model the evolution of sequences related
by a tree with models permitting indels
\cite{ThorneEtal91,HeinEtal2000,Hein2001,HolmesBruno2001,SteelHein2001,Holmes2003,KnudsenMiyamoto2003,LunterSongMiklosHein2003,MiklosLunterHolmes2004,LunterEtAl2004,BradleyHolmes2007,SatijaEtAl2008}.
Recent work has extended these ideas to the reconstruction of indel
histories \cite{KimSinha2007,DialloEtAl2007},
particularly at the genomic scale \cite{MillerHausslerEtAl2006,PatenHolmesBirney2008}.
There is substantial prior work in computational linguistics
on the theory of transducers for sequences \cite{MohriPereiraRiley2000}
and parse trees \cite{Rounds70,Thatcher70,GecsegSteinby97,ComonEtAl2007-TreeTransducers}
(from which we take the terms ``string transducer'' and ``parse-tree transducer'').
We draw on the bioinformatics literature for SCFGs
\cite{Eddy94,Sakakibara94c,Durbin98},
especially Pair SCFGs \cite{RivasEddy99,Holmes2005,DowellEddy2006}
and phylogenetic SCFGs \cite{KnudsenHein2003}.
In particular, an early example of a pairwise conditional model $P(Y|X)$ for structure-dependent RNA evolution
may be found in \cite{KleinEddy2003}.
A conditional framework similar to ours in some respects is described by Sakakibara {\em et al} \cite{Sakakibara2003}.
The statistical inference algorithms for multiple-sequence SCFGs are closely related to the ``protosequence algorithm'' of Sankoff \cite{Sankoff85},
a seminal work in both RNA structural alignment and ancestral reconstruction.

% In this paper we have reported innovations to two programs, \xrate\ and \stemloc,
% rendering them suitable for the challenge of reconstructing
% the ancestral RNA world from its modern descendants.
% The \xrate\ program can now reconstruct ancestral RNA substitution histories,
% while \stemloc\ can align and predict RNA structures
% using externally-supplied grammars
% (including grammars derived from evolutionary models).
% We have tested these programs using simulations, cross-validations, alignment benchmarks and structure prediction benchmarks.

% Building on these results, we have presented
% a general algorithm for constructing a multiple-sequence stochastic grammar
% modeling the common phylogenetic descent of a group of RNA sequences,
% along with a specific implementation of this algorithm---\indiegram---suitable
% for three-way RNA alignment and reconstruction.
% Using the TKF Structure Tree model, we demonstrated
% the capability of such composite grammars
% to infer ancestral structures of structured RNAs.

While we have focused on the TKF Structure Tree model in our Results,
our model-construction algorithm is applicable to any model of the evolution of secondary
structure which can be expressed as a parse-tree transducer.
Realistic structural and thermodynamic effects---such as base-stacking
or loop length distributions---can, in principle, be incorporated.
%Furthermore, while an evolutionary model of structural evolution
%is necessarily complex,
%our model-construction algorithm enforces a compact parametrization,
%allowing fast training on structural alignments of pairs of sequences.

% to do: discuss the possibility of progressive reconstruction a la Ortheus

An implementation of inference algorithms for models on the three-taxon phylogeny
is sufficient to construct an MCMC sampling algorithm
over many sequences on an arbitrary phylogeny.
A sketch of such a sampling algorithm is as follows:
at each step of the sampling algorithm, we re-sample the sequence
and structure of the ancestral node $W$,
conditioned on the sequences and structures of $X$, $Y$ and $Z$.
The structural alignment of all four sequences can change at each step,
providing for fast mixing and guaranteeing ergodicity.
This move is similar to the sampler proposed by \cite{JensenHein2002} for 
models with a HMM structure.
Note that this, in principle, permits construction of a crude sampler
to simultaneously infer phylogeny as well, by proposing and accepting
or rejecting changes to the underlying tree as well as the implied
structural alignment.

% With reference to the motivating problem of reconstructing ribosomal RNA,
% the sequence-level reconstructions with \xrate\ 
% (described in \secref{subst-reconstruction})
% are certainly feasible on computing resources available now or in the near future.

Reconstructing structural changes of large RNAs
using the three-way sampling kernel which we have described
would require resources far in excess of what are currently available;
barring the availability of supercomputers with terabytes of memory,
such algorithms will only be feasible for short RNAs (\tabref{complexities}).
Thus, alternate strategies will have to be explored.
One promising direction is to consider variations on the three-way sampling kernel,
such as the importance-sampling approach described for the TKF model
by \cite{RedelingsSuchard2005}.
This approach first proposes an ancestor $W$ by aligning extant
sequence $X$ to $Y$ (ignoring $Z$); then, in a second step, the proposed $W$ is
independently aligned to $Z$. The proposed three-way alignment and
reconstruction is then randomly accepted (or rejected) using a
Hastings ratio based on the three-way transducer composition. The
complexity of this kernel is the same as the pairwise case;
with suitable constraints, this is feasible for RNA
grammars on present hardware, at least for ribosomal domains (if not
yet whole subunits---although pairwise alignment of those should also
be possible soon). The approach of Redelings and Suchard therefore
merits future consideration in the context of modeling the evolution
of RNAs on a tree.

The evolutionary models we have described form a theoretical
foundation for studying the structural evolution of RNA gene families
in detail. By opening the door to ancestral sequence reconstruction
and paleogenomics, these algorithms will allow us to test our
knowledge of RNA evolution by direct experimental investigation
of early ribonucleic machines.

%All programs described here are released under the GNU General Public License and
%can be download from \darturl\ as part of the \dart\ software package for sequence analysis.

% Do NOT remove this, even if you are not including acknowledgments
\newpage
\section*{Acknowledgments}

We thank
Jeff Thorne,
Jamie Cate,
Jennifer Doudna,
Jotun Hein,
David Haussler,
Sean Eddy,
Elena Rivas and
Eric Westhof
for inspirational discussions.
We are grateful to Ben Redelings and one anonymous referee for illuminating criticism.

%\section*{References}
% The bibtex filename
\bibliography{../latex-inputs/alignment,../latex-inputs/reconstruction,../latex-inputs/duplication,../latex-inputs/genomics,../latex-inputs/ncrna,../latex-inputs/url}

\clearpage
\section*{Figure Legends}

\begin{figure}[!ht]
  \centering
%  \includegraphics [scale=0.4] {figs/parent.pdf}
   \caption{
     \textbf{A simple example of transducer composition to build an
       evolutionary model of two extant sequences.}
    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
     A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
     and two branch transducers (the gray boxes labeled $T_X$ and $T_Y$)
     mutate it according to the specified evolutionary model.
     Gray nodes correspond to observed data and white nodes unobserved data.
   }
   \figlabel{parent}
 \end{figure}

 \begin{figure}[!ht]
   \centering
 %  \includegraphics [scale=0.4] {figs/threeway.pdf}
   \caption{
     \textbf{Transducer composition on a star phylogeny with three
       (extant) leaf sequences.}
     An ancestral sequence $W$ evolves into three descendant sequences $X$, $Y$ and $Z$.
     A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
     and three branch transducers (the gray boxes labeled $T_X$, $T_Y$ and $T_Z$)
     mutate it according to the specified evolutionary model.
     Gray nodes correspond to observed data and white nodes unobserved data.
     If the branch transducers are time-reversible, then this star phylogeny with three leaves is
     the neighborhood of any interior node in a (binary) phylogenetic tree, 
     from which it follows that evaluating the likelihood function
     on this star phylogeny is sufficient for a sampling algorithm on an arbitrary phylogeny.
   }
\figlabel{threeway}
 \end{figure}

\begin{figure}[!ht]
  \centering
%  \includegraphics [width=0.5\textwidth] {figs_plos/stree-with-structure.pdf}
  \caption{
    \textbf{The TKF Structure Tree model represents the
      evolution of RNA structure as nested stem and loop sequences.}
    The model consists of recursively nested loop sequences (gray,
    horizontal) and stem sequences (black, vertical). The loops are
    sequences of unpaired bases and the stems are sequences of
    covarying base-pairs. Both loop and stem sequences evolve
    according to the Thorne-Kishino-Felsenstein (TKF) model
    \cite{ThorneEtal91} of molecular evolution.  Figure is extended
    from a similar version in \cite{Holmes2004}.
  }
  \figlabel{tkfst}
\end{figure}

\begin{figure}[!ht]
  \centering
%  \includegraphics [width=0.5\textwidth] {figs_plos/stree-mutations-with-structures.pdf}
  \caption{
    \textbf{Evolution of a RNA structure under the TKF Structure Tree
      model.}
    The TKF Structure Tree model includes phenomena such as point
    mutations in loop sequences ($1 \rightarrow 2$ and $4 \rightarrow 5$), covariant
    mutations in stem sequences ($2 \rightarrow 3$), insertions in
    loop sequences ($3 \rightarrow 4$), insertions in stem sequences
    ($5 \rightarrow 6$), structural insertions ($6 \rightarrow 7$),
    and structural deletions ($7 \rightarrow 8$).
   Figure is extended from a similar version in \cite{Holmes2004}.
  }
  \figlabel{tkfst-mutations}
\end{figure}

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/simulations.pdf}
  \caption{
    \textbf{Dependence of reconstruction accuracy on outgroup branch
      length.}
    We allowed the outgroup branch length $z$ to vary within $[0.5,
    3.0]$, where a branch length of 1 corresponds to one expected
    substitution per site in loop sequence.  We divided this interval
    into 25 bins and simulated 25 alignments for each bin.
    Parameters used were:
  }
  \figlabel{simulations}
\end{figure}

\clearpage
\section*{Tables}

\begin{table}[!ht]
  \caption{
    \textbf{State types of the singlet transducer (single-sequence
      SCFG) of the TKF Structure Tree model.}}
  \begin{tabular}{|l|rrrl|}
    \hline
    State & $\type$ & $\absorb$ & $\emit$ & description \\ \hline
    $L$ & $\Sstart$ & & & Start of a loop \\ \hline
    $I_L$ & $\Sinsert$ & & $(x,\Tnull)$ & Single-base emission \\ \hline \hline
    $S$ & $\Sstart$ & & & Start of a stem \\ \hline
    $I_S$ & $\Sinsert$ & & $(x,y)$ & Base-pair emission \\ \hline \hline
    $B$ & $\Sinsert$ & & $(L\,S)$ & Bifurcation \\ \hline
  \end{tabular}
  \begin{flushleft}
    Singlet transducers can only have states of type $\Sstart$ or
    $\Sinsert$.
  \end{flushleft}
 \tablabel{tkfstsinglet-types}
\end{table}

% Use table to allow the table to span the width of the entire page.
\begin{table}[!ht]
  \caption{
    \textbf{Singlet transducer (single-sequence SCFG) of the TKF Structure Tree model.}}
  \begin{tabular}{|rcl|r||rcl|r|}
    \hline
    Source & $\rightarrow$ & Destination & probability & Source & $\rightarrow$ & Destination & probability \\ \hline
    $L$ & $\rightarrow$ & $u \,\, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $S$ & $\rightarrow$ & $u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$ \\ \hline
    & $|$ & $B$ & $\kappa_l \cdot \pi_l(S)$ & & $|$ & $B_e$ & $1-\kappa_s$ \\ \hline
    & $|$ & $\Send$ & $1-\kappa_l$ & & & & \\ \hline
    $I_L$ & $\rightarrow$ & $u \, \, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $I_S$ & $\rightarrow$ & $u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$\\ \hline
    & $|$ & $B$ & $\kappa_l \cdot \pi_l(S)$ & & $|$ & $B_e$ & $1-\kappa_s$\\ \hline
    & $|$ & $\Send$ & $1-\kappa_l$ & & & & \\ \hline \hline
    $B$ & $\rightarrow$ & $(L \,\, S)$ & 1 & & & & \\ \hline
    $B_e$ & $\rightarrow$ & $(L \,\, \Send)$ & 1 & & & & \\ \hline
  \end{tabular}
  \begin{flushleft}
    The state types for this model are shown in \tabref{tkfstsinglet-types}.
    The singlet transducer generates ancestral RNA sequences and
    structures.
    We use the notation of formal grammars to represent state
    transformation rules; for example, the rule $I_L \rightarrow
    u\,\,I_L$ corresponds to (in a Pair HMM) an $\Sinsert$ state $I_L$
    emitting a nucleotide $u$ and then making a self-transition.
    Both loop ($L$ and $I_L$) and stem ($S$ and $I_S$) sequence evolve as TKF sequences
    with length parameters $\kappa_l$ and $\kappa_s$ (defined in ``A simple model of RNA structural evolution'').
    $\pi_l(u)$ and $\pi_s(uv)$ are the equilibrium distributions of unpaired nucleotides $x$ and
    paired nucleotides $(u,v)$.
    The bifurcation state $B_e$ is used to end stem sequences
    (only loop sequences are allowed to transition to the empty string).
  \end{flushleft}
  \tablabel{tkfstsinglet}
\end{table}

\begin{table}[!ht]
  \caption{
    \textbf{State types of the branch transducer
      (conditionally-normalized Pair SCFG) of the TKF Structure Tree
      model.}}
  \begin{tabular}{|l|rrrl|} \hline
    State & $\type$ & $\absorb$ & $\emit$ & description \\ \hline
    $L$ & $\Sstart$ & & & Start of a loop \\ \hline
    $I_L$ & $\Sinsert$ & & $(u,\Tnull)$ & Single-base insertion \\ \hline
    $M_L$ & $\Smatch$ & $(x,\Tnull)$ & $(u,\Tnull)$ & Single-base substitution \\ \hline
    $D_L$ & $\Smatch$ & $(x,\Tnull)$ & & Single-base deletion \\ \hline
    $W_L$ & $\Swait$ & & & Wait for next base \\ \hline \hline
    $S$ & $\Sstart$ & & & Start of a stem \\ \hline
    $I_S$ & $\Sinsert$ & & $(u,v)$ & Base-pair insertion \\ \hline
    $M_S$ & $\Smatch$ & $(x,y)$ & $(u,v)$ & Base-pair substitution \\ \hline
    $D_S$ & $\Smatch$ & $(x,y)$ & & Base-pair deletion \\ \hline
    $W_S$ & $\Swait$ & & & Wait for next base-pair \\ \hline \hline
    $B_i$ & $\Sinsert$ & & $(L_i\,S_i)$ & Stem insertion \\ \hline
    $B$ & $\Smatch$ & $(L\,S)$ & $(L\,S)$ & Stem conservation \\ \hline
    $B_p$ & $\Smatch$ & $(L\,S)$ & $(L\,\Send)$ & Stem deletion \\ \hline
    $B_e$ & $\Smatch$ & $(L\,\Send)$ & $(L\,\Send)$ & Stem extinction \\ \hline
  \end{tabular}
  \begin{flushleft}
    States which have the same names as states of the singlet transducer in \tabref{tkfstsinglet-types}
    are the branch-transducer equivalents of the corresponding singlet-transducer states
    (e.g., a $\Smatch$ state might be the branch equivalent of an $\Sinsert$ state).
    States $L_i$ and $S_i$ are the $\Sstart$ states of a sub-model (not shown) identical
    in structure to the singlet transducer.  They are used to insert a
    new stem-loop structure.
  \end{flushleft}
  \tablabel{tkfstbranch-types}
\end{table}

% Use table to allow the table to span the width of the entire page.
\begin{table}[!ht]
  \caption{
    \textbf{Branch transducer (conditionally-normalized Pair SCFG) of
      the TKF Structure Tree model.}}
  \begin{tabular}{|rcl|r||rcl|r|} \hline
    Source & $\rightarrow$ & Destination & probability & Source & $\rightarrow$ & Destination & probability \\ \hline
    $L$ & $\rightarrow$ & $w \,\, I_L$ & $\beta_l(t) \cdot \pi_l(w)$ & $S$ & $\rightarrow$ & $w \,\, I_S \,\, x$ & $\beta_s(t) \cdot \pi_s(wx)$ \\ \hline
    & $|$ & $B_i$ & $\beta_l(t) \cdot \pi_l(S)$ & & $|$ & $W_S$ & $1-\beta_s(t)$ \\ \hline
    & $|$ & $W_L$ & $1-\beta_l(t)$ & & & &  \\ \hline
    $I_L$ & $\rightarrow$ & $w \,\, I_L$ & $\beta_l(t) \cdot \pi_l(w)$ & $I_S$ & $\rightarrow$ & $w \,\, I_S \,\, x$ & $\beta_s(t) \cdot \pi_s(wx)$ \\ \hline
    & $|$ & $B_i$ & $\beta_l(t) \cdot \pi_l(S)$ & & $|$ & $W_S$ & $1-\beta_s(t)$ \\ \hline
    & $|$ & $W_L$ & $1-\beta_l(t)$ & & & & \\ \hline
    $M_L$ & $\rightarrow$ & $w \,\, I_L$ & $\beta_l(t) \cdot \pi_l(w)$ & $M_S$ & $\rightarrow$ & $w \,\, I_S \,\, x$ & $\beta_s(t) \cdot \pi_s(wx)$ \\ \hline
    & $|$ & $B_i$ & $\beta_l(t) \cdot \pi_l(S)$ & & $|$ & $W_S$ & $1-\beta_s(t)$ \\ \hline
    & $|$ & $W_L$ & $1-\beta_l(t)$ & & & &  \\ \hline
    $D_L$ & $\rightarrow$ & $w \,\, I_L$ & $\gamma_l(t) \cdot \pi_l(w)$ & $D_S$ & $\rightarrow$ & $w \,\, I_S \,\, x$ & $\gamma_s(t) \cdot \pi_s(wx)$ \\ \hline
    & $|$ & $B_i$ & $\gamma_l(t) \cdot \pi_l(S)$ & & $|$ & $W_S$ & $1-\gamma_s(t)$ \\ \hline
    & $|$ & $W_L$ & $1-\gamma_l(t)$ & & & &  \\ \hline
    $W_L$ & $\rightarrow$ & $u \,\, w \,\, M_L$ & $\alpha_l(t) \cdot M_l(u,w)$ & $W_S$ & $\rightarrow$ & $u \,\, w \,\, M_S \,\, x \,\, v$ & $\alpha_s(t) \cdot M_s(uv,wx)$ \\ \hline
    & $|$ & $u \,\, D_L$ & $1-\alpha_l(t)$ & & $|$ & $u \,\, D_S \,\, v$ & $1-\alpha_s(t)$ \\ \hline
    & $|$ & $B$ & $\alpha_l(t)$ & & $|$ & $B_e$ & $1$ \\ \hline
    & $|$ & $B_p$ & $1-\alpha_l(t)$ & & & &  \\ \hline
    & $|$ & $\Send$ & $1$ & & & &  \\ \hline \hline
    $B$ & $\rightarrow$ & $L \,\, S$ & 1 & $B_p$ & $\rightarrow$ & $L \, \, \Send$ & 1 \\ \hline
    $B_i$ & $\rightarrow$ & $L_i \,\, S_i$ & 1 & $B_e$ & $\rightarrow$ & $L \, \, \Send$ & 1 \\ \hline \hline
  \end{tabular}
  \begin{flushleft}
    The state types for this model are shown in
    \tabref{tkfstbranch-types}. The branch transducer evolves a
    sequence and structure along a branch of the phylogenetic
    tree. States $L_i$ and $S_i$ are the $\Sstart$ states for a
    sub-model corresponding to an insertion of a new stem in the
    descendant sequences; the sub-model (not shown) is identical in
    structure to the singlet transducer shown in
    \tabref{tkfstsinglet}. $\pi_l(w)$ and $\pi_s(wx)$ are the
    equilibrium distributions of, respectively, descendant unpaired
    nucleotide $w$ and descendant paired nucleotides $(w,x)$;
    $M_l(u,w)$ and $M_l(uv,wx)$ are the conditional distributions
    (i.e. match probabilities) of, respectively, descendant unpaired
    nucleotide $w$ given an ancestral unpaired nucleotide $u$ and
    descendant paired nucleotides $(w,x)$ given ancestral nucleotides
    $(u,v)$. The functions $\alpha_n(t)$, $\beta_n(t)$ and
    $\gamma_n(t)$ are parametrized by the insertion and deletion rates
    of the TKFST model and are defined in ``A simple model of RNA
    structural evolution.''
  \end{flushleft}
  \tablabel{tkfstbranch}
\end{table}

% Tables created as follows:
% sort -k 1,2 bralibaseII.results | perl -pe 's/0\.(\d\d)(\d)\d+/$1.$2/g;s/ +/ & /g' | cols
% ...then manual editing

% Method                          & Metric      & U5   & g2intron & rRNA & tRNA & TPP_riboswitch
% stemloc.progressive             & Align_Sens  & 82.6 & 74.2     & 92.6 & 93.2 & 76.8          
% stemloc.progressive             & Align_ppv   & 83.7 & 74.8     & 92.8 & 93.9 & 80.7          
% stemloc.progressive             & Struct_Sens & 74.9 & 64.3     & 51.0 & 74.0 & 57.1          
% stemloc.progressive             & Struct_ppv  & 73.9 & 56.7     & 59.0 & 76.4 & 63.3          
% stemloc.progressive.xrate       & Align_Sens  & 82.6 & 74.2     & 92.6 & 93.2 & 76.8          
% stemloc.progressive.xrate       & Align_ppv   & 83.7 & 74.8     & 92.8 & 93.9 & 80.7          
% stemloc.progressive.xrate       & Struct_Sens & 90.5 & 88.2     & 77.2 & 88.5 & 72.3          
% stemloc.progressive.xrate       & Struct_ppv  & 66.2 & 63.2     & 64.4 & 70.6 & 74.7          
% stemloc.tkfst.progressive       & Align_Sens  & 81.6 & 75.4     & 91.4 & 94.6 & 62.4          
% stemloc.tkfst.progressive       & Align_ppv   & 81.7 & 75.0     & 92.6 & 94.4 & 73.7          
% stemloc.tkfst.progressive       & Struct_Sens & 37.9 & 42.1     & 37.4 & 70.9 & 36.2          
% stemloc.tkfst.progressive       & Struct_ppv  & 68.0 & 63.8     & 66.5 & 88.3 & 77.2          
% stemloc.tkfst.progressive.xrate & Align_Sens  & 81.6 & 75.4     & 91.4 & 94.6 & 62.4          
% stemloc.tkfst.progressive.xrate & Align_ppv   & 81.7 & 75.0     & 92.6 & 94.4 & 73.7          
% stemloc.tkfst.progressive.xrate & Struct_Sens & 80.8 & 84.5     & 80.6 & 88.6 & 59.8          
% stemloc.tkfst.progressive.xrate & Struct_ppv  & 60.7 & 64.1     & 66.9 & 71.1 & 63.9

% We first modified \stemloc\ to allow the user to specify the Pair SCFG that it uses for progressive alignment.
% We then used our program \evoldoer, an implementation of TKFST for pairwise alignment,
% to generate Pair SCFGs corresponding to the TKFST model at timepoints $\{0.01,0.1,0.2,0.3,0.4\}$,
% using loop and stem substitution rate matrices estimated with \xrate\ from
% ribosomal RNA alignments \cite{DeRijkEtAl2001,DowellEddy2006}
% and TKFST parameters $\{\lambda_n,\mu_n\}$ estimated previously \cite{Holmes2004}.

\begin{table}[ht]
  \caption{
    \textbf{Percentage sensitivity and positive predictive value (Sensitivity / PPV) for pairwise nucleotide-level alignments in the \bralibaseII\ benchmark.}}
  \begin{tabular}{|r|rrrr|}
    \hline
    & U5                      & g2intron                & rRNA                    & tRNA \\
    \hline
    TKFST grammar     &      81.6  /      81.7  & {\bf 75.4} / {\bf
      75.0} &      91.4  /      92.6  & {\bf 94.6} / {\bf 94.4} \\ \hline
    \stemloc\ grammar & {\bf 82.6} / {\bf 83.7} &      74.2  /
    74.8  & {\bf 92.6} / {\bf 92.8} &      93.2  /      93.9 \\ \hline
    \end{tabular}
\begin{flushleft}
  We compared the performance of the TKFST model for progressive multiple alignment
  of RNAs against the performance of a grammar with a richer model of RNA structure (``\stemloc''; \cite{Holmes2005}).
  Sensitivity is defined as $\mbox{TP}/(\mbox{TP} + \mbox{FN})$ and PPV is defined as $\mbox{TP}/(\mbox{TP} + \mbox{FP})$,
  where TP is the number of true positives (correctly aligned residue pairs),
  FN is the number of false negatives (residue pairs that should have been aligned but were not) and
  FP is the number of false positives (residue pairs that were incorrectly aligned).
  These statistics are summed over all pairs of sequences in the multiple alignment;
  therefore, ``Sensitivity'' for pairwise residue alignments is equivalent to the Sum of Pairs Score, or SPS \cite{ThompsonEtal99}.
  ``g2intron'' is the RFAM entry {\tt Intron\_gpII}, containing domains V and VI of the Group II intron.
  \end{flushleft}
  \tablabel{bralibaseII-align}
\end{table}

\begin{table}[ht]
  \caption{
    \textbf{Percentage sensitivity and positive predictive value (Sensitivity / PPV) for predicted base-pairs in the \bralibaseII\ benchmark.}}
  \begin{tabular}{|r|rrrr|}
    \hline
    & U5                      & g2intron                & rRNA                    & tRNA \\
    \hline
    TKFST grammar      &      37.9  /      68.0   &      42.1  /
    63.8  &      37.4  /      66.5  &      70.9  / {\bf 88.3} \\ \hline
    \stemloc\ grammar  &      74.9  / {\bf 73.9}  &      64.3  /
    56.7  &      51.0  /      59.0  &      74.0  /      76.4  \\ \hline
  \end{tabular}
  \begin{flushleft}
    We compared the performance of the TKFST model for structure-prediction accuracy during
    progressive multiple alignment of RNAs against the performance of a grammar with 
    a richer model of RNA structure (``\stemloc''; \cite{Holmes2005}).
    ``g2intron'' is the RFAM entry {\tt Intron\_gpII}, containing domains V and VI of the Group II intron.
  \end{flushleft}
  \tablabel{bralibaseII-struct}
\end{table}

% nanos: Counted 3672701 subseqs and 48323860 bifurcations.
% tRNA: Counted 12573792 subseqs and 404730480 bifurcations.
% Y: Counted 34474759 subseqs and 1429713088 bifurcations.
% g2intron: Counted 139037184 subseqs and 1895528744 bifurcations.
% Scaling constant is approximately
% A * (404730480 / 48323860) = (19 / 3) => A = 0.76
% time = A * (# bifs / # bifs for nanos) * time for nanos
% Y time = 0.76 * (1429713088 / 48323860) * 3 min = 67 min
% g2intron time = 0.76 * (1895528744 / 48323860) * 3 min = 89 min
\begin{table}[!ht]
  \caption{
   \textbf{Estimates of the memory and time required to reconstruct ancestral
     structures of three RNAs from several families of biological interest
     (as reported by \indiegram).}}
 \begin{tabular}{|l|rrr|} \hline
    Family & Sequence lengths & Memory & Time \\ \hline
    \emph{nanos} 3' TCE & 61-64 nt & 3 Gb & 3 min \\ \hline
    tRNA & 69-73 nt & 11 Gb & 19 min \\ \hline
    Y RNA & 47-81 nt & 33 Gb & 70 min \\ \hline
    Group II intron (domains V and VI) & 76-91 nt & 122 Gb & 90 min \\ \hline
  \end{tabular}
  \begin{flushleft}
    The \emph{nanos} 3' translational control element (TCE) sequences
    are the seed sequences of the
    corresponding RFAM family \cite{GriffithsJonesEtAl2003} and the
    three tRNA sequences are from the \bralibaseII\ database
    \cite{GardnerEtAl2005} (identifiers AB042432.1-14140\_14072,
    Z82044.1-16031\_16103 and AC008670.6-83725\_83795) The Group II
    intron sequences (identifiers Z00044.1-87253\_87177,
    X57546.1-2817\_2907 and X04465.1-2700\_2775) are from BralibaseII
    \cite{GardnerEtAl2005}.  The Y RNAs are hY1, hY4, and hY5 from
    \cite{TeunissenEtAl2000}; sequence lengths exclude the conserved
    stem S1.  The time estimates are for a 2.2GHz AMD Opteron 848 CPU.
  \end{flushleft}
  \tablabel{complexities}
\end{table}

\clearpage
\section*{Algorithms}

\begin{algorithm}[!ht]
  \SetLine
  \SetKwFunction{rnd}{Random[0,1]}
  \KwIn{Rule $\rho'' \in \grules''$}
  \KwOut{Parse (sub)tree $T' \in \gtrees'$}

  \Switch{$\rho''$}{
    \Case{$X \to Y$}{
      \eIf{\rnd $< P(X \to Y) / q_{XY}$}{
        \KwRet{$(X \to Y)$} \;
      }{
        Let $z = \sum_W q_{XW} P'(W \to Y)$ \;
        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to Y) / z$ \;
        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to Y)$} \;
      }
    }
    \Case{$X \to \epsilon$}{
      \eIf{\rnd $< P'(X \to \epsilon) / P'(\epsilon | X)$}{
        \KwRet{$(X \to \epsilon)$} \;
      }{
        Let $z = \sum_W q_{XW} P'(W \to \epsilon)$ \;
        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to \epsilon) / z$ \;
        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to \epsilon)$} \;
      }
    }
    \Other{
      \KwRet{$(\rho'')$} \;
    }
  }
  \caption{\label{alg:restoreTransitions}
    Subroutine $\mathrm{restoreTransitions}(\rho'')$ for the null cycle elimination procedure.
  }
\end{algorithm}

\begin{algorithm}[!ht]
  \SetLine
  \SetKwFunction{rnd}{Random[0,1]}
  \KwIn{Rule $\rho' \in \grules'$}
  \KwOut{Parse (sub)tree $T \in \gtrees$}

  \Switch{$\rho'$}{
    \Case{$B^0 \to \epsilon$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
      \KwRet{$(B \to T_L\ T_R)$} \;
    }
    \Case{$B' \to L'$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
      \KwRet{$(B \to L\ T_R)$} \;
    }
    \Case{$B' \to R'$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
      \KwRet{$(B \to T_L\ R)$} \;
    }
    \Case{$B^2 \to L'\ R'$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      \KwRet{$(B \to L\ R)$} \;
    }
    \lCase{$B^0 \to B'$}{
      \KwRet{$()$} \;
    }
    \lCase{$B' \to B^2$}{
      \KwRet{$()$} \;
    }
    \Other{
      Let $\rho$ be the original rule in $\grules$ \;
      \KwRet{$(\rho)$} \;
    }
  }

  \caption{\label{alg:restoreBifurcations}
    Subroutine $\mathrm{restoreBifurcations}(\rho')$ for the null cycle elimination procedure.
  }
\end{algorithm}

\begin{algorithm}[!ht]
  \SetLine
  \SetKwFunction{rnd}{Random[0,1]}
  \KwIn{Nonterminal $A \in \gnons$}
  \KwOut{Parse tree $T \in \gtrees: \mathrm{seq}(T) = \epsilon, \mathrm{root}(T) = A$}

  \eIf{$A$ is a bifurcation state, $A \to X\ Y$,}{
    $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
    $T_Y \leftarrow \mathrm{sampleNullSubtree}(Y)$ \;
    \KwRet{$(A \rightarrow T_X\ T_Y)$} \;
  }{
    \eIf{\rnd $< P(A \to \epsilon) / P(\epsilon|A)$}{
      \KwRet{$(A \to \epsilon)$} \;
    }{
      Let $z = \sum_Y P(A \to Y) P(\epsilon | Y)$ \;
      Sample state $X$ from probability distribution $P(X) = P(A \to X) P(\epsilon | X) / z$ \;
      $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
      \KwRet{$(A \rightarrow T_X)$} \;
    }
  }

  \caption{\label{alg:sampleNullSubtree}
    Subroutine $\mathrm{sampleNullSubtree}(A)$ for the null cycle elimination procedure.
  }
\end{algorithm}

\begin{algorithm}[!ht]

  \SetKwData{bifurcProb}{bifurcProb}
  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
  \SetLine
  \KwIn{sequences $X, Y, Z$}

  \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
    \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
      \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
        \BlankLine
        \ForEach{state $\bvec{a}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ $\leftarrow$ \calcTransEmitProb$\left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
          store $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \KwRet{$\alpha_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;

  \caption{\alglabel{inside}
    The constrained Inside algorithm for three sequences $X, Y, Z$.
    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
  }
\end{algorithm}

\begin{algorithm}[!ht]
  \SetKwData{emitProb}{emitProb}
  \SetLine
  \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Inside matrix $\alpha$}

  \emitProb $\leftarrow 0$\;
  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
    \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \alpha_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
  }
  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
    \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
    \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \alpha_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
  }
  \KwRet{\emitProb}\;
  \caption{\alglabel{insidetransemit}
    Subroutine $\mathrm{calcTransEmitProb}()$ for the Inside algorithm.
    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
  }
\end{algorithm}

\begin{algorithm}[!ht]

  \SetKwData{maxProb}{maxProb}
  \SetKwData{bifurcProb}{bifurcProb}
  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
  \SetLine
  \KwIn{sequences $X, Y, Z$}

  \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
    \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
      \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
        \BlankLine
        \ForEach{state $\bvec{a}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ \\
          $\qquad \leftarrow \max \left( \calcTransEmitProb \left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right), \bifurcProb \right)$ \;
          store $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \KwRet{$\gamma_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;

  \caption{\alglabel{cyk}
    The constrained CYK algorithm for three sequences $X, Y, Z$.
    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
  }
\end{algorithm}

\begin{algorithm}[!ht]
  \SetKwData{emitProb}{emitProb}
  \SetLine
  \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate CYK matrix $\gamma$}

  \emitProb $\leftarrow 0$\;
  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
    \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{b} \right)\, \gamma_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \right)$\;
  }
  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
    \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
    \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \gamma_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right) \right)$\;
  }
  \KwRet{\emitProb}\;
  \caption{\alglabel{cyktransemit}
    Subroutine $\mathrm{calcTransEmitProb}()$ for the CYK algorithm.
    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
  }
\end{algorithm}

\begin{algorithm}[!ht]
  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
  \SetKwData{bifurcProb}{bifurcProb}
  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
  \SetLine
  \KwIn{sequences $X, Y, Z$, Inside matrix $\alpha$}

  \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
    \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
      \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
        \BlankLine
        \ForEach{state $\bvec{b}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(X)}_O,n^{(X)}_L \right) \in b_{out,L}\left( n^{(X)} \right)$}{
            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_L \right) \in b_{out,L}\left( n^{(Y)} \right)$}{
              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_L \right) \in b_{out,L}\left( n^{(Z)} \right)$}{
                \bifurcProb += \calcLBifurcProb($\bvec{b}; \cdot$)\;
              }
            }
          }
          \ForEach{$\left( n^{(X)}_O,n^{(X)}_R \right) \in b_{out,R}\left( n^{(X)} \right)$}{
            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_R \right) \in b_{out,R}\left( n^{(Y)} \right)$}{
              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_R \right) \in b_{out,R}\left( n^{(Z)} \right)$}{
                \bifurcProb += \calcRBifurcProb($\bvec{b}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \leftarrow$ \calcTransEmitProb$\left( \bvec{b}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
          store $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \caption{\alglabel{outside}
    The constrained Outside algorithm for three sequences $X, Y, Z$.
    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Outside fill order
    with $\Semit$ states first, then $\Snull$ states in topological order.
  }
\end{algorithm}

\begin{algorithm}[!ht]
  \SetKwData{emitProb}{emitProb}
  \SetLine
  \KwIn{state $\bvec{b}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Outside matrix $\beta$}

  \emitProb $\leftarrow 0$\;
  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{b}$} {
    \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \beta_{\bvec{a}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
  }
  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
    \lIf{$c_{out}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{out}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{out}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
    \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \beta_{\bvec{a}}\left( c_{out}\left(\bvec{b}; n^{(X)}\right), c_{out}\left(\bvec{b}; n^{(Y)}\right), c_{out}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
  }
  \KwRet{\emitProb}\;
  \caption{\alglabel{outsidetransemit}
    Subroutine $\mathrm{calcTransEmitProb}()$ for the Outside algorithm.
    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
  }
\end{algorithm}

\begin{algorithm}[!ht]

  \SetKw{Select}{select}
  \SetKw{Pop}{pop}
  \SetKw{Push}{push}
  \SetKw{Goto}{goto}
  \SetKw{Mainloop}{main loop}
  \SetKwData{coordinateStack}{coordinateStack}
  \SetKwFunction{goto}{goto}
  \SetKwBlock{Begin}{begin main loop:}{end}
  \SetLine
  \KwIn{sequences $X, Y, Z$}

  pushdown stack coordinateStack; \tcc*[f]{hold (state, subsequence triplet) pairs} \\
  $\bvec{a} \leftarrow \Sstart$; \tcc*[f]{current ensemble state} \\
  $n^{(X)} \leftarrow N^{(X)}$; \tcc*[f]{current $X$ subsequence; $N^{(X)}$ is the outermost subsequence} \\
  $n^{(Y)} \leftarrow N^{(Y)}$; \tcc*[f]{current $Y$ subsequence; $N^{(Y)}$ is the outermost subsequence} \\
  $n^{(Z)} \leftarrow N^{(Z)}$; \tcc*[f]{current $Z$ subsequence; $N^{(Z)}$ is the outermost subsequence} \\
  clear \coordinateStack;\

  \Begin{
    output current state $\bvec{a}$ and subsequence triplet $\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
    \SetLine
    \uIf(\tcc*[f]{end of a parse subtree}){$\bvec{a}$ is the $\Send$ state}{
      \lIf(\tcc*[f]{end of the parse tree}){\coordinateStack is empty}{\Return}
      \Pop $\left( \bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ from \coordinateStack\;
      \Goto \Mainloop\;
    }
    \uElseIf(\tcc*[f]{bifurcation $\bvec{a} \to \bvec{c} \bvec{b}$}){$\bvec{a}$ is a bifurcation state}{
      \Select $\left( n^{(X)}_L, n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$,
      $\left( n^{(Y)}_L, n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$,
      $\left( n^{(Z)}_L, n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$ \\
      such that \\
      $\gamma_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)} \right) = \gamma_{\bvec{a}} \left(n^{(X)}_L, n^{(Y)}_L, n^{(Z)}_L \right)  \gamma_{\bvec{a}} \left(n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ \;
      \Push $\left( \bvec{c}, n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ onto \coordinateStack\;
      $\bvec{a} \leftarrow \bvec{b}$\;
      $n^{(X)} \leftarrow n^{(X)}_L$\;
      $n^{(Y)} \leftarrow n^{(Y)}_L$\;
      $n^{(Z)} \leftarrow n^{(Z)}_L$\;
      \Goto \Mainloop\;
    }
    \Else(\tcc*[f]{$\Semit$ or $\Snull$ state}){
      $m^{(X)} \leftarrow c_{in}\left(\bvec{b}; n^{(X)}\right)$ \;
      $m^{(Y)} \leftarrow c_{in}\left(\bvec{b}; n^{(Y)}\right)$ \;
      $m^{(Z)} \leftarrow c_{in}\left(\bvec{b}; n^{(Z)}\right)$ \;
      \Select $\bvec{b} \in \left\{ b : \exists \,\bvec{a} \to \bvec{l}\bvec{b}\bvec{r} \right\}$ \\
      such that \\
      $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) = P(\bvec{a} \to \bvec{l}\bvec{b}\bvec{r}) \gamma_{\bvec{b}} \left( m^{(X)}, m^{(Y)}, m^{(Z)} \right)$\;
      $\bvec{a} \leftarrow \bvec{b}$\;
      $n^{(X)} \leftarrow m^{(X)}$\;
      $n^{(Y)} \leftarrow m^{(Y)}$\;
      $n^{(Z)} \leftarrow m^{(Z)}$\;
      \Goto \Mainloop\;
    }
  }
  \caption{\alglabel{cyktraceback}
    The constrained CYK traceback algorithm for three sequences $X, Y, Z$.
  }
\end{algorithm}

\end{document}

### 2009-04-18 retired.tex
Import from CVS.

\ref{tab:stemissions} and state transitions in Table \ref{tab:sttransitions}.

\begin{table}[ht]
\centering
\begin{tabular}{lcccc}
state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{ST})$\\ \hline
$\state{L'}{}{}$ & $\Sstart$ \\
$\state{I'_L}{}{y}$ & $\Sinsert$ & & $(y\,\Tnull)$ & $p_L(y)$ \\
$\state{S'}{}{}$ & $\Sstart$ \\
$\state{I'_S}{}{xy}$ & $\Sinsert$ & & $(x,y)$ & $p_S(x,y)$ \\
$\bifurc{B'}{S'}{L'}$ & $\Sinsert$ & & & 1 \\
$\bifurc{B'}{\Send}{L'}$ & $\Sinsert$ & & & 1 \\
\\
$\state{L}{}{}$ & $\Sstart$ \\
$\state{I_L}{}{y}$ & $\Sinsert$ & & $(y\,\Tnull)$ & $p_L(y)$ \\
$\state{M_L}{x}{y}$ & $\Smatch$ & $(x\,\Tnull)$ & $(y\,\Tnull)$ & $e_L(y|x)$ \\
$\state{D_L}{x}{}$ & $\Smatch$ & $(x\,\Tnull)$ &  & 1 \\
$\state{W_L}{}{}$ & $\Swait$ \\
$\bifurc{B}{S}{L}$ & $\Smatch$ & & & 1 \\
$\bifurc{Bp}{\Send}{L}$ & $\Smatch$ & & & 1 \\
$\bifurc{B'}{S'}{L}$ & $\Sinsert$ & & & 1 \\
\\
$\state{S}{}{}$ & $\Sstart$ \\
$\state{I_S}{}{xy}$ & $\Sinsert$ & & $(x,y)$ & $p_S(x,y)$ \\
$\state{M_S}{uv}{xy}$ & $\Smatch$ & $(u,v)$ & $(x,y)$ & $e_S(x,y|u,v)$ \\
$\state{D_S}{uv}{}$ & $\Smatch$ & $(u,v)$ &  & 1 \\
$\state{W_S}{}{}$ & $\Swait$ \\
$\bifurc{B}{\Send}{L}$ & $\Smatch$ & & & 1 \\
\\
\end{tabular}
\caption{\label{tab:stemissions}Absorptions and emissions of states of the TKF91 Structure Tree transducer.}
\end{table}

The primed states $\{L',\,I'_L,\,S',\,I'_S,\,\bifurc{L'}{}{},\,\bifurc{\Send}{}{}\}$ are those of the singlet machine 
which lies at the highest ancestral node $\hat{n} = \argmin_m \{\type(a_m) \in \{\Sstart,\,\Sinsert\}\}$
(as defined in \secref{transitions}).  $x,\,y,\,u,\,v$ are any terminal symbols.

\begin{table}[ht]
\centering
\begin{tabular}{rclcrclcrclc}
& & & $t(\,\bullet\,|\mathrm{ST})$ & & & & $t(\,\bullet\,|\mathrm{ST})$ & & & & $t(\,\bullet\,|\mathrm{ST})$ \\ \cline{4-4} \cline{8-8} \cline{12-12}
$\state{L'}{}{}$ & $\to$ & $\state{I'_L}{}{y}$ & $\kappa_L$ & $\state{I'_L}{}{y}$ & $\to$ & $\state{I'_L}{}{y}$ & $\kappa_L$ \\
& & $\bifurc{B'}{S'}{L'}$ & 777 & & & $\bifurc{B'}{S'}{L'}$ & 777 \\
& & $\Send{}{}$ & $(1-\kappa_L)$ & & & $\Send{}{}$ & $(1-\kappa_L)$ \\
$\state{S'}{}{}$ & $\to$ & $\state{I'_S}{}{xy}$ & $\kappa_S$ & $\state{I'_S}{}{xy}$ & $\to$ & $\state{I'_S}{}{xy}$ & $\kappa_S$ \\
& & $\bifurc{B'}{\Send}{L'}$ & $(1-\kappa_S)$ & & & $\bifurc{B'}{\Send}{L'}$ & $(1-\kappa_S)$ \\
\\
$\state{L}{}{}$ & $\to$ & $\state{I_L}{}{y}$ & $\beta_L$ & $\state{I_L}{}{y}$ & $\to$ & $\state{I_L}{}{y}$ & $\beta_L$ & $\state{W_L}{}{}$ & $\to$ & $\state{M_L}{x}{y}$ & $\alpha_L$ \\
& & $\bifurc{B'}{S'}{L}$ & 777 & & & $\bifurc{B'}{S'}{L}$ & 777  & & & $\state{D_L}{x}{}$ & $(1-\alpha_L)$ \\
& & $\state{W_L}{}{}$ & $(1-\beta_L)$ & & & $\state{W_L}{}{}$ & $(1-\beta_L)$  & & & $\bifurc{B}{S}{L}$ & $\alpha_L$ \\
& & & & $\state{M_L}{x}{y}$ & $\to$ & $\state{I_L}{}{y}$ & $\beta_L$  & & & $\bifurc{Bp}{\Send}{L}$ & $(1-\alpha_L)$ \\
& & & & & & $\bifurc{B'}{S'}{L}$ & 777 & & & $\Send$ & 1 \\
& & & & & & $\state{W_L}{}{}$ & $(1-\beta_L)$ \\
& & & & $\state{D_L}{x}{}$ & $\to$ & $\state{I_L}{}{y}$ & $\gamma_L$ \\
& & & & & & $\bifurc{B'}{S'}{L}$ & 777 \\
& & & & & & $\state{W_L}{}{}$ & $(1-\gamma_L)$ \\
\\
$\state{S}{}{}$ & $\to$ & $\state{I_S}{}{xy}$ & $\beta_S$ & $\state{I_S}{}{xy}$ & $\to$ & $\state{I_S}{}{xy}$ & $\beta_S$ & $\state{W_S}{}{}$ & $\to$ & $\state{M_S}{uv}{xy}$ & $\alpha_S$ \\
& & $\state{W_S}{}{}$ & $(1-\beta_S)$ & & & $\state{W_S}{}{}$ & $(1-\beta_S)$  & & & $\state{D_S}{uv}{}$ & $(1-\alpha_S)$ \\
& & & & $\state{M_S}{uv}{xy}$ & $\to$ & $\state{I_S}{}{xy}$ & $\beta_S$  & & & $\bifurc{B}{\Send}{L}$ & 1 \\
& & & & & & $\state{W_S}{}{}$ & $(1-\beta_S)$ \\
& & & & $\state{D_S}{uv}{}$ & $\to$ & $\state{I_S}{}{xy}$ & $\gamma_S$ \\
& & & & & & $\state{W_S}{}{}$ & $(1-\gamma_S)$ \\
\end{tabular}
\caption{\label{tab:sttransitions}Allowed transitions between states of the TKF91 Structure Tree transducer.}
\end{table}

***Need to fix weights currently marked with '777'.

***Add a note that this model makes it too easy to prune subtrees

\section{Old Stuff}

The inside algorithm includes a summation over the possible ``children'' of a state $\bvec{a}$.  We separate
these states into sets
\begin{align}
  \mathscr{C}_{\bvec{a}}^{\mathrm{null}} &= \left\{ \bvec{b}\,|\,\exists\,\, \bvec{a} \to \bvec{b}, \type(b_{\activenode{a}}) \in \{ \Sstart,\,\Swait \} \right\} \\
  \mathscr{C}_{\bvec{a}}^{\mathrm{emit}}(\bvec{y},\,\bvec{z}) &= \left\{ \bvec{b}\,|\,\exists\,\, \bvec{a} \to \bvec{y'}\,\bvec{b}\,\bvec{z'}, \type(b_{\activenode{a}}) \in \{\Sinsert,\,\Smatch\} \right\} \\
  \mathscr{C}_{\bvec{a}}^{\mathrm{bifur,L}} &= \left\{ \bvec{b}\,|\,\exists\,\, \bvec{a} \to \bvec{c}\,\bvec{b}, \type\left(\emit(b_{\activenode{a}})\right) \, \mathrm{defined} \right\} \\
  \mathscr{C}_{\bvec{a}}^{\mathrm{bifur,R}} &= \left\{ \bvec{b}\,|\,\exists\,\, \bvec{a} \to \bvec{b}\,\bvec{d}, \type\left(\emit(b_{\activenode{a}})\right) \, \mathrm{defined} \right\} ,
\end{align}
where the requirement ``$\type\left(\emit(b_{\activenode{a}})\right) \, \mathrm{defined}$'' is equivalent to requiring that the bifurcation state $b_{\activenode{a}}$
emit a nonterminal (the function $\type()$ is defined only on the domain of nonterminals).
Recall that bifurcations are considered as emissions of nonterminals, in our view.
We use the shorthand notation
\beqn
\mathscr{C}_{\bvec{a}}(\bvec{y},\,\bvec{z}) = \left\{ \begin{array}{ll}\mathscr{C}_{\bvec{a}}^{\mathrm{null}} & \bvec{y} = \bvec{z} = \Tnull \\ \mathscr{C}_{\bvec{a}}^{\mathrm{emit}}(\bvec{y},\,\bvec{z}) & \bvec{y} \ne \Tnull \,\mathrm{or}\, \bvec{z} \ne \Tnull \end{array} \right.
\eeqn

Similarly, the outside algorithm includes a summation over the possible ``parents'' of a state $\bvec{b}$.  We 
separate these states into sets
\begin{align}
  \mathscr{P}_{\bvec{b}}^{\mathrm{null}} &= \left\{ \bvec{a}\,|\,\exists\,\, \bvec{a} \to \bvec{b}, \type(b_{\activenode{a}}) \in \{ \Sstart,\,\Swait \} \right\} \\
  \mathscr{P}_{\bvec{b}}^{\mathrm{emit}}(\bvec{y},\,\bvec{z}) &= \left\{ \bvec{a}\,|\,\exists\,\, \bvec{a} \to \bvec{y'}\,\bvec{b}\,\bvec{z'}, \type(b_{\activenode{a}}) \in \{\Sinsert,\,\Smatch\} \right\} \\
  \mathscr{P}_{\bvec{b}}^{\mathrm{bifur,L}} &= \left\{ \bvec{a}\,|\,\exists\,\, \bvec{a} \to \bvec{c}\,\bvec{b}, \type\left(\emit(b_{\activenode{a}})\right) \, \mathrm{defined} \right\} \\
  \mathscr{P}_{\bvec{b}}^{\mathrm{bifur,R}} &= \left\{ \bvec{a}\,|\,\exists\,\, \bvec{a} \to \bvec{b}\,\bvec{d}, \type\left(\emit(b_{\activenode{a}})\right) \, \mathrm{defined} \right\} ,
\end{align}
We use the shorthand notation
\beqn
\mathscr{P}_{\bvec{b}}(\bvec{y},\,\bvec{z}) = \left\{ \begin{array}{ll}\mathscr{P}_{\bvec{b}}^{\mathrm{null}} & \bvec{y} = \bvec{z} = \Tnull \\ \mathscr{P}_{\bvec{b}}^{\mathrm{emit}}(\bvec{y},\,\bvec{z}) & \bvec{y} \ne \Tnull \,\mathrm{or}\, \bvec{z} \ne \Tnull \end{array} \right.
\eeqn

The superscripts indicate the nature of the states contained in the ``parent'' or ``child'' sets.  
For example, a transition from $\bvec{a}$ to a state in $\mathscr{C}_{\bvec{a}}^{\mathrm{null}}$
results in no emissions or absorptions.

(Below is from before I switched to fold-envelope index notation instead of explicit subsequence coordinate pairs.)

The left- and right-bifurcation probabilities are calculated in the Inside algorithm as
\begin{align}
  \mathrm{calcLBifurcProb}(\bvec{a}; \cdot) &= \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} \weight (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \inside{\bvec{c}}{i_1}{k_1}{i_2}{k_2}{i_3}{k_3} \, \inside{\bvec{b}}{k_1}{j_1}{k_2}{j_2}{k_3}{j_3} \nonumber \\
  \mathrm{calcRBifurcProb}(\bvec{a}; \cdot) &= \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} \weight (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \inside{\bvec{b}}{i_1}{k_1}{i_2}{k_2}{i_3}{k_3} \, \inside{\bvec{d}}{k_1}{j_1}{k_2}{j_2}{k_3}{j_3} \, . \nonumber
\end{align}

The left- and right-bifurcation probabilities are calculated in the Outside algorithm as
\begin{align}
  \mathrm{calcLBifurcProb}(\bvec{b}; \cdot) &=  \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} \weight (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \outside{\bvec{a}}{k_1}{j_1}{k_2}{j_2}{k_3}{j_3} \, \inside{\bvec{c}}{k_1}{i_1}{k_2}{i_2}{k_3}{i_3} \nonumber \\
  \mathrm{calcRBifurcProb}(\bvec{b}; \cdot) &=  \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} \weight (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \outside{\bvec{a}}{i_1}{k_1}{i_2}{k_2}{i_3}{k_3} \, \inside{\bvec{d}}{j_1}{k_1}{j_2}{k_2}{j_3}{k_3} \, .\nonumber
\end{align}

\subsection{TKF91}
We can cast the TKF91 model \citep{ThorneEtal91} as a transducer as required by our formalism for building a composed machine.

The TKF91 grammar generates a regular string language and so can be described by an HMM.
After introducing a state of type $\Swait$, the corresponding transducer has states
$\Phi \in \{\state{S}{}{},\,\state{I}{}{y},\,\state{M}{x}{y},\,\state{D}{x}{},\,\state{W}{}{}\}$.  Terminal absorptions and emissions are
shown in Table \ref{tab:tkf91emissions} and state transitions in Table \ref{tab:tkf91transitions}.

\begin{table}[ht]
\centering
\begin{tabular}{lcccc}
state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKF91})$\\ \hline
$\state{S'}{}{}$ & $\Sstart$ \\
$\state{I'}{}{y}$ & $\Sinsert$ & & $(y\,\Tnull)$ & $p(y)$ \\
\\
$\state{S}{}{}$ & $\Sstart$ \\
$\state{I}{}{y}$ & $\Sinsert$ & & $(y\,\Tnull)$ & $p(y)$ \\
$\state{M}{x}{y}$ & $\Smatch$ & $(x\,\Tnull)$ & $(y\,\Tnull)$ & $e(y|x)$ \\
$\state{D}{x}{}$ & $\Smatch$ & $(x\,\Tnull)$ &  & 1 \\
$\state{W}{}{}$ & $\Swait$ \\
\end{tabular}
\caption{\label{tab:tkf91emissions}Absorptions and emissions of states of the TKF91 transducer.}
\end{table}

$x,\,y$ are any terminal symbols.  The primed states $S',\,I'$ are those of the singlet machine which lies at the root node of the guide tree.

\begin{table}[ht]
\centering
\begin{tabular}{rclcrclcrclc}
& & & $t(\,\bullet\,|\mathrm{TKF91})$ & & & & $t(\,\bullet\,|\mathrm{TKF91})$ & & & & $t(\,\bullet\,|\mathrm{TKF91})$ \\ \cline{4-4} \cline{8-8} \cline{12-12}
$\state{S'}{}{}$ & $\to$ & $\state{I'}{}{y}$ & $\kappa$ & $\state{I'}{}{y}$ & $\to$ & $\state{I'}{}{y}$ & $\kappa$ \\
& & $\state{\Send}{}{}$ & $(1-\kappa)$ & & & $\state{\Send}{}{}$ & $(1-\kappa)$ \\
\\
$\state{S}{}{}$ & $\to$ & $\state{I}{}{y}$ & $\beta$ & $\state{I}{}{y}$ & $\to$ & $\state{I}{}{y}$ & $\beta$ & $\state{W}{}{}$ & $\to$ & $\state{M}{x}{y}$ & $\alpha$ \\
& & $\state{W}{}{}$ & $(1-\beta)$ & & & $\state{W}{}{}$ & $(1-\beta)$ & & & $\state{D}{x}{}$ & $(1-\alpha)$ \\
& & & & $\state{M}{x}{y}$ & $\to$ & $\state{I}{}{y}$ & $\beta$  & & & $\Send$ & 1 \\
& & & & & & $\state{W}{}{}$ & $(1-\beta)$ \\
& & & & $\state{D}{x}{}$ & $\to$ & $\state{I}{}{y}$ & $\gamma$ \\
& & & & & & $\state{W}{}{}$ & $(1-\gamma)$ \\
\end{tabular}
\caption{\label{tab:tkf91transitions}Allowed transitions between states of the TKF91 transducer.}
\end{table}

(Below is from before I switched to fold-envelope index notation instead of explicit subsequence coordinate pairs.)

The left- and right-bifurcation probabilities are calculated in the Inside algorithm as
\begin{align}
  \mathrm{calcLBifurcProb}(\bvec{a}; \cdot) &= \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} \weight (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \inside{\bvec{c}}{i_1}{k_1}{i_2}{k_2}{i_3}{k_3} \, \inside{\bvec{b}}{k_1}{j_1}{k_2}{j_2}{k_3}{j_3} \nonumber \\
  \mathrm{calcRBifurcProb}(\bvec{a}; \cdot) &= \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} \weight (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \inside{\bvec{b}}{i_1}{k_1}{i_2}{k_2}{i_3}{k_3} \, \inside{\bvec{d}}{k_1}{j_1}{k_2}{j_2}{k_3}{j_3} \, . \nonumber
\end{align}

The left- and right-bifurcation probabilities are calculated in the Outside algorithm as
\begin{align}
  \mathrm{calcLBifurcProb}(\bvec{b}; \cdot) &=  \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} \weight (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \outside{\bvec{a}}{k_1}{j_1}{k_2}{j_2}{k_3}{j_3} \, \inside{\bvec{c}}{k_1}{i_1}{k_2}{i_2}{k_3}{i_3} \nonumber \\
  \mathrm{calcRBifurcProb}(\bvec{b}; \cdot) &=  \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} \weight (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \outside{\bvec{a}}{i_1}{k_1}{i_2}{k_2}{i_3}{k_3} \, \inside{\bvec{d}}{j_1}{k_1}{j_2}{k_2}{j_3}{k_3} \, .\nonumber
\end{align}

### 2009-04-18 supplementary.tex
Import from CVS.

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{PLoS2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **
%% Please insert a running head of 30 characters or less.  
%% Include it twice, once between each set of braces
\markboth{Evolutionary modeling}{Evolutionary modeling}

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW
\input{defs.tex}
%% END MACROS SECTION

\begin{document}

% Title must be 150 words or less
\begin{flushleft}
  {\Large
    \textbf{Evolutionary models of structured RNA: Text S1}
  }
\\
Robert K. Bradley$^{1}$, 
Ian Holmes$^{1,2,\ast}$
\\
\textbf{1} Biophysics Graduate Group, University of California, Berkeley, CA, USA
\\
\textbf{2} Department of Bioengineering, University of California, Berkeley, CA, USA
\\
$\ast$ E-mail: indiegram@postbox.biowiki.org
\end{flushleft}

\newpage
\tableofcontents

\newpage
\section{Two-sequence transducer models}

We here give formal definitions of two-sequence models and the state machines which
generate the factored probability distribution $P (X, Y | \Delta T) = P (X) \cdot P (Y | X, \Delta T)$,
where the marginal $P(X)$ is generated by a {\bf singlet transducer}
and the conditional $P (Y | X, \Delta T)$ by a {\bf branch transducer}.
We refer to the branch transducer as $\theta$, so the conditional distribution
is more precisely $P (Y | X, \Delta T, \theta)$.

\subsection{Formal Grammars} \seclabel{grammars}

There is a close relationship between formal grammars and the singlet and branch transducer abstract state machines.
By labeling the nonterminals of a regular or stochastic context-free grammar (SCFG) as states of an abstract machine, allowing these states to absorb and emit 
appropriate terminal symbols,\footnote{The relationship between formal grammars and abstract state machines is more transparent
in the Mealy machine view, but the Moore machine view turns out to be more useful for our purposes.},
and carefully assigning transition and emission weights,
we can create a state machine which generates the same language, with the same distribution of weights, as that produced by the original grammar.\footnote{We speak of
weights associated with a transition rather than probabilities in order to allow for more general models.}
We therefore use the terms ``nonterminal'' and ``state'' interchangeably
and refer to the ``parse tree'' generated by singlet and branch transducers.

Phrased more precisely, there is an isomorphism between the singlet and branch transducers of two-sequence models
and Pair SCFGs.  Practically speaking, this means that for every joint distribution $P(X, Y | \Delta T)$ 
generated by a singlet and branch transducers, there exists a Pair SCFG which generates the same distribution.

\subsection{States, transitions and emissions}
A singlet transducer has states of type $\Sstart$ and $\Sinsert$.
Each state $\phi\in\Phi$ of the branch transducer has type
\[ \type(\phi) \in \lbrace\Sstart,\,\Sinsert,\,\Smatch,\,\Swait,\,\Send\rbrace \, . \]
State typing is as follows:
\begin{itemize}
  \item A $\Sstart$ state begins a branch of the parse tree.
  \item An $\Sinsert$ state emits, but does not absorb, symbols.
  \item A $\Smatch$ state absorbs and (possibly) emits symbols.
  \item A $\Swait$ state is a null state which allows a branch transducer to ``pause''
    while it waits for an input symbol from another transducer.
    In the two-sequence case, the input symbols is emitted by the singlet transducer generating
    the ancestral sequence.
  \item An $\Send$ state ends a branch of the parse tree.
\end{itemize}
The names of state types are similar to the $\{\Sstart,\,\Sinsert,\,\Smatch,\,\Send\}$ states of the familiar Pair HMM.
Deletions are handled as a special case by states of type $\Smatch$ which absorb but do not emit symbols pairs.
Only states of type $\Smatch$ can absorb symbol pairs $(x,y)$.

Bifurcations in the grammar, when considered as emission of nonterminals, can be handled analogously
to terminal emission.  States $\phi: \type(\phi) \in \{\Sinsert,\,\Smatch\}$ can emit nonterminal pairs $(c\,d)$,
where $c$ or $d$ can be the null symbol, and the emitted pairs $(c\,d)$ can be absorbed by states of type $\Smatch$.

The emission weight of a nonterminal pair $(c\,d)$ from the state $b: \type(b) = \Smatch$, 
conditional on absorption of a nonterminal pair $(l\,m)$, is 
\[ e_b ( c\,d|l\,m,\theta ) \, . \]

The functions $\emit()$ and $\absorb()$ are defined to return the emission or absorption of a particular state,
\begin{align*}
  \emit(\phi) &:= \left\{\begin{array}{ll} (x,y) & x,y = \mathrm{terminal\,\,or\,\,null} \\(c\,d) & c,d = \mathrm{nonterminal\,\,or\,\,null}. \end{array} \right. \\
  \absorb(\phi) &:= \left\{\begin{array}{ll} (x,y) & x,y = \mathrm{terminal\,\,or\,\,null} \\(c\,d) & c,d = \mathrm{nonterminal\,\,or\,\,null}. \end{array} \right.
\end{align*}

We frequently use the notation ${}^{uv}\phi^{xy}$ to indicate that a state of type $\Smatch$ absorbs a symbol pair $(u,v)$
and emits a pair $(x,y)$.
The notation for bifurcations is slightly different:
A bifurcation state which left-emits a nonterminal $d$ and makes a transition to a state $\phi$ with weight 1 is written as $\bifurc{B}{d}{\phi}$.

A transition between states $a$ and $b$ of the branch transducer $\theta$ has a weight
\[ t(a,b|\theta) = t(a \to b|\theta). \]
Terminal emission is handled by states $\phi: \type(\phi) \in \{\Smatch,\,\Sinsert\}$, which emit symbol pairs $(x,y)$,
where $x$ or $y$ can be the null symbol.
In this paired-emission perspective, left single-terminal emissions $x$ are represented as $(x\,\Tnull)$;
right single-terminal emissions are handled similarly.
The weight of an emission of a terminal pair $(x,y)$ from a state $b: \type(b) = \Smatch$, 
conditioned on absorption of a terminal pair $(u,v)$, is 
\[ e_b ( x,y|u,v,\theta ) \, . \]
Recall that the emission weights of states of type $\Smatch$ are defined conditioned upon the absorbed symbols.

\newpage
\section{Multiple-sequence transducer models} \seclabel{evolutionarymodels}

We can use our two-sequence models to construct a model of many sequence related by a
guide phylogenetic tree.
The guide tree specifies the (conjectured) phylogenetic relationship of all sequences.
A singlet transducer, which emits, but does not absorb, symbols, lies at the root of the guide tree and 
serves as a generative model of the ancestral sequence.
A branch transducer represents the evolution of an ancestral sequence into a single descendant 
sequence, that is, the action of evolution along the single-branch tree (Ancestor $\to$ Descendant).
To represent the evolution of an ancestral sequence into many descendant sequences
(whose phylogenetic relationship is specified by the guide tree),
we place a branch transducer on each branch of the guide tree to create a multiple-sequence model.

If the branch grammar has no bifurcations and only left or right emissions are allowed, then
the language generated is a regular string language and the corresponding jointly normalized abstract state machine is 
an HMM.  In this simplest case our formalism for creating a multiple-sequence model reduces to that given 
by \cite{Holmes2003} for combining HMMs on a guide tree.

\subsection{The guide tree}
The nodes of the tree are labeled $1,\dots,N$ in the order reached by any preorder depth-first traversal of the tree.
The length of each branch $(\parent(m) \to m)$ is given by the evolutionary time $t_m$.
To specify ancestor-descendant relationships, we introduce notation:
$\mDn$ ($\mNDn$) means node $m$ is descended from (not descended from) node $n$, and
$\mDEn$ ($\mNDEn$) means node $m$ is descended from or identical to (not descended from and not identical to) node $n$.

\subsection{States, transitions and emissions}
The multiple-sequence model is formed by the composition/intersection of $(N-1)$ branch transducers 
such that there is a branch transducer on each branch $(\parent(m) \to m);\, m = 2,\dots,N$ of the guide tree and a singlet transducer at the root node.
Our framework allows for the placement of different branch transducers, with a unique set of nonterminals (state space) $\Phi$ and
allowed transitions between states and corresponding weights, on each branch.  $\theta^{(m)}$ denotes 
the branch transducer governing evolution along the branch $(\parent(m) \to m)$ of the guide tree.

States of the multiple-sequence model are represented by as N-dimensional vectors $\bvec{a}$,
\[ \bvec{a} = \statevec{a} \, . \]
These states are typed as
\[ \type(\bvec{a}) \in \lbrace\Sstart,\,\Semit,\,\Sbifurc,\,\Snull,\,\Send\rbrace \, . \]
State typing is as follows:
\begin{itemize}
  \item A $\Sstart$ state begins a branch of the parse tree of one or more of the $N$ sequences on the phylogenetic tree.
  \item An $\Semit$ state emits symbols (terminals) to one or more of the $N$ sequences.
  \item A $\Sbifurc$ state corresponds to a bifurcation in the parse tree of
    one or more of the $N$ sequences.
  \item A $\Snull$ state corresponds to any non-$\Send$ state which represents neither
    an emission or bifurcation.
  \item An $\Send$ state ends a branch of the parse tree.
\end{itemize}
States are typed according to the transition by which they are reachable
(details of the typing are given in \secref{transitions}).

Transitions and emissions of the multiple-sequence model are defined in terms of the transitions and emissions 
of the branch transducers at each node as well as the singlet transducer at the root node.
The weight of a transition $\bvec{a} \to \bvec{b}$ is therefore
\beqn
t(\bvec{a},\bvec{b}) = \prod_{m|a_m \ne b_m} \transweight{m}{a_m}{b_m} \, ,
\eeqn
and the weight of an emission $(\bvec{x}\,\bvec{y})$ from a state $\bvec{b}$ is
\beqn
\totalemissionweight{\bvec{b}}{\bvec{x}\,\bvec{y}} = \prod_m e_{b_m}\left(x_m\,y_m|x_{\parent(m)}\,y_{\parent(m)},\theta^{(m)}\right) \, .
\eeqn
We frequently will not explicitly write out the conditional dependence on the absorbed terminals $(x_{\parent(m)}\,y_{\parent(m)})$,
but the reader should keep in mind that in general emission weights will depend on the absorbed symbols.

\subsection{Formal grammars} \seclabel{multigrammars}

Analogously to the case with two-sequence models (\secref{grammars}),
there exists a one-to-one mapping between the multiple-sequence models generated by our model-construction algorithm 
and multi-sequence SCFGs.
In other words, given a singlet and branch transducers of a two-sequence model, as well
as a guide tree relating the extant sequences, there exists a corresponding multi-sequence SCFG
which generates the same joint probability distribution $P(X_1 , ... , X_n)$.

\subsection{Constructing the state graph} \seclabel{stategraph}

As described in the paper,
we need a way to efficiently construct the state graph of the multi-sequence model,
where the state graph consists of a list of accessible states and the possible transitions between them.
This state graph can be constructed by an uninformed depth-first search,
where at each step of the search we obtain the possible child nodes 
by applying one of the following possible transitions of the multi-sequence model:
\begin{enumerate}
\item {\bf Null transition}:
  The state of a single branch transducer is updated, with no terminal emission or bifurcation.
\item {\bf Terminal emission}:
  A state makes a transition to a $\Sinsert$ state.  The emitted symbol is passed down the guide
  tree to all descendant branch transducers, which transition to states of type $\Smatch$.
\item {\bf Bifurcation}:
  A state makes a transition to a special $\Sinsert$ state which emits a new branch of the parse tree.
  The emitted symbol is passed down the guide tree to all descendant branch transducers, 
  which transition to states of type $\Smatch$.
\item {\bf End transition}:
  The singlet transducer associated with the root sequence can transition to the $\Send$ state,
  signaling that this parse tree is finished.
\end{enumerate}

Each of these transitions is explained in detail in the following section.

\subsection{Allowed transitions} \seclabel{transitions}

Following \cite{Holmes2003},  we let transitions of the multi-sequence model begin at the active node and cascade down the guide tree
as appropriate.  Unless defined otherwise, node $n$ is the active node of the multi-sequence model with state $\bvec{a}$, 
\begin{align}
  n &= \activenode{a} \\
  &= \mathrm{argmax}_m \left\{\type(a_m) \notin \{\Swait,\,\Send\}\right\} \, . \label{eqn:activenode} 
\end{align}
Each possible allowed transition is obtained by making a valid change (updating) the 
state of the singlet or one or more of the branch transducers of the multi-sequence model.

\subsubsection*{Null Transitions}

\[ \statevec{a} \to \statevec{b} \]

\begin{align} \label{eqn:nulltransition}
  \type(\bvec{b}) &= \Snull  \\
  \weight(\bvec{a}\to\bvec{b}) &= t(\bvec{a},\bvec{b}) \\
  &= \transweight{n}{a_n}{b_n} .
\end{align}

This transition updates the state of the branch transducer at the active node $n$ of the guide tree, leaving 
the rest unchanged, with no corresponding terminal emission or bifurcation in the grammar.
Nodes other than the active node do not change state, $a_m = b_m \,\forall\, m \ne n$, and the only 
allowable transitions of this form are to states $b_n:\,\type(b_n) \in \{\Sstart,\,\Swait\}$.  Transitions
to states of type $\Sinsert$ or $\Smatch$ result in emissions, and transitions to the end state $\Send$
are handled as a special case.

\subsubsection*{Terminal Emission} \seclabel{terminalemission}
\[ \statevec{a} \to \statevec{x}\,\statevec{b}\,\statevec{y} \]

\begin{align} \label{eqn:terminalemission}
  \type(\bvec{b}) &= \Semit \\
  \weight(\bvec{a}\to\bvec{x}\,\bvec{b}\,\bvec{y}) &=  t(\bvec{a},\bvec{b}) \cdot \totalemissionweight{\bvec{b}}{\bvec{x}\,\bvec{y}} \\
  &= \left[ \prod_{m|a_m \ne b_m} \transweight{m}{a_m}{b_m} \right] \cdot \left[ \prod_m \emissionweight{m}{b_m}{x_m\,y_m} \right] ,
\end{align}
where we are defining states with no emissions to emit $(\Tnull\,\Tnull)$ with weight 1, 
$\emissionweight{m}{b_m}{x_m\,y_m} = 1$ if $(x_m\,y_m) = (\Tnull\,\Tnull)$ and $\emit(b_m) = \Tnull$.

The active node $n$ makes a transition to a state $b_n:\,\type(b_n) = \Sinsert$, emitting
a terminal symbol pair $\emit(b_n) = (x_n\,y_n)$.  This symbol pair is passed down the guide tree to descendant nodes $\{m|\mDn, \type(a_m) \ne \Send\}$, 
forcing them to make a transition from states of type $\Swait$ to states of type $\Smatch$, which can absorb
terminal pairs $(x,y)$.  Qualitatively, this transition and emission could represent the evolution
of two paired nucleotides along the subtree rooted at node $n$ of the complete guide tree.
If one of $\bvec{x}$ or $\bvec{y}$ is null, then \eqref{eqn:terminalemission} could represent the evolution of
a single unpaired nucleotide along the subtree rooted at node $n$.

\paragraph{Left emission.}
All terminals $\bvec{y}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission} are null.

\begin{tabular}{rl}
  Nodes $\mNDEn$: & $a_m = b_m$. \\
  & $(x_m\,y_m) = (\Tgap\,\Tnull)$. \\
  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
  & $(x_n\,y_n) = \emit(b_n)$. \\
  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (x_{\parent(m)}\,\Tnull)$. \\
  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tgap\,\Tnull) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull \end{array} \right. $
\end{tabular}

\paragraph{Right emission.}
All terminals $\bvec{x}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission} are null.

\begin{tabular}{rl}
  Nodes $\mNDEn$: & $a_m = b_m$. \\
  & $(x_m\,y_m) = (\Tnull\,\Tgap)$. \\
  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
  & $(x_n\,y_n) = \emit(b_n)$. \\
  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (\Tnull\,y_{\parent(m)})$. \\
  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tnull\,\Tgap) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull . \end{array} \right. $
\end{tabular}

\paragraph{Paired emission.}

There is at least one non-null terminal in both $\bvec{x}$ and $\bvec{y}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission}.

\begin{tabular}{rl}
  Nodes $\mNDEn$: & $a_m = b_m$. \\
  & $(x_m\,y_m) = (\Tgap\,\Tgap)$. \\
  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
  & $(x_n\,y_n) = \emit(b_n)$. \\
  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (x_{\parent(m)}\,y_{\parent(m)})$. \\
  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tgap\,\Tgap) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull . \end{array} \right. $
\end{tabular}

\subsubsection*{Bifurcations}
\[ \statevec{a} \to \statevec{c}\,\statevec{b}\,\statevec{d} \]

\begin{align} \label{eqn:nonterminalemission}
  \type(\bvec{b}) &= \Sbifurc \\
  \weight(\bvec{a}\to\bvec{c}\,\bvec{b}\,\bvec{d}) &=  t(\bvec{a},\bvec{b}) \cdot \totalemissionweight{\bvec{b}}{\bvec{c}\,\bvec{d}} \\
  &= \left[ \prod_{m|\,a_m \ne b_m} \transweight{m}{a_m}{b_m} \right] \cdot \left[ \prod_m \emissionweight{m}{b_m}{c_m\,d_m} \right] .
\end{align}
where we are defining the emission weight of the $\Send$ nonterminal to be 1, $\emissionweight{m}{b_m}{c_m\,d_m} = 1$ if $c_m = \Send$ or $d_m = \Send$.

Bifurcations are handled similarly to terminal emission.  The active node $n$ can undergo a bifurcation by making a transition 
$a_n \to b_n,\, b_n:\,\type(b_n) = \Sinsert$, emitting a pair of nonterminals $\emit(b_n) = (c_n\,d_n)$.  Descendant nodes $\{m|\mDn,\type(a_m) \ne \Send\}$
are forced to make a transition from states of type $\Swait$ to states of type $\Smatch$ which can absorb nonterminal pairs $(c\,d)$.
All emissions are pairwise, so left and right bifurcations are represented as pairs $(\bvec{c}\,\bvec{d})$ where either $\bvec{c}$ or $\bvec{d}$ is null.
If $\bvec{d}$ is null, then \eqref{eqn:nonterminalemission} could represent the insertion and subsequent evolution of a new RNA stem-loop structure.

\paragraph{Left bifurcation.}
All nonterminals $\bvec{d}$ in the transition $\bvec{a} \to \bvec{c}\,\bvec{b}\,\bvec{d}$ \eqref{eqn:nonterminalemission} are null.
The nonterminals $\bvec{c}$ are the ``new'' states (for example, corresponding to a newly formed stem);
the nonterminals $\bvec{b}$ are the states which will generate the (evolved) ancestral sequence.

\begin{tabular}{rl}
  Nodes $\mNDEn$: & $a_m = b_m$. \\
  & $(c_m\,d_m) = (\Send\,\Tnull)$. \\
  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
  & $(c_n\,d_n) = \emit(b_n)$. \\
  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m)$ \\
  & $(c_m\,d_m) = \emit(b_m) . $
\end{tabular}

\paragraph{Right bifurcation.}
All nonterminals $\bvec{c}$ in the transition $\bvec{a} \to \bvec{c}\,\bvec{b}\,\bvec{d}$ \eqref{eqn:nonterminalemission} are null.
The nonterminals $\bvec{d}$ are the ``new'' states;
the nonterminals $\bvec{b}$ are the states which will generate the (evolved) ancestral sequence.

\begin{tabular}{rl}
  Nodes $\mNDEn$: & $a_m = b_m$. \\
  & $(c_m\,d_m) = (\Tnull\,\Send)$. \\
  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
  & $(c_n\,d_n) = \emit(b_n)$. \\
  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m)$ \\
  & $(c_m\,d_m) = \emit(b_m) . $
\end{tabular}

\paragraph{Paired bifurcation}
There is at least one non-null nonterminal in both $\bvec{c}$ and $\bvec{d}$ in the transition $\bvec{a} \to \bvec{c}\,\bvec{b}\,\bvec{d}$ \eqref{eqn:nonterminalemission}.
The nonterminals $\bvec{c}$ and $\bvec{d}$ are both ``new'' states;
the nonterminals $\bvec{b}$ are the states which will generate the (evolved) ancestral sequence.

\begin{tabular}{rl}
  Nodes $\mNDEn$: & $a_m = b_m$. \\
  & $(c_m\,d_m) = (\Send\,\Send)$. \\
  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
  & $(c_n\,d_n) = \emit(b_n)$. \\
  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m)$ \\
  & $(c_m\,d_m) = \emit(b_m) . $
\end{tabular}

This paired-bifurcation is included for completeness--for example, it could be used to model symmetric loops--but it
increases the complexity of grammar parsing.

\subsubsection*{Transition to $\Send$}
\[ \statevec{a} \to \Send \]

\begin{align} \label{eqn:transtoend}
  \weight(\bvec{a}\to\Send) &=  t(\bvec{a},\Send) \\
  &= \prod_{m|\type(a_m)\ne\Send} \transweight{m}{a_m}{\Send} .
\end{align}

The singlet transducer associated with the highest active ancestral node,
\beqn
\hat{n} = \argmin_m \left\{ \type(a_m) \in \{\Sstart,\,\Sinsert\} \right\}
\eeqn
can make a transition to the state $\Send$,
forcing the entire multi-sequence model to transition to $\Send$.\footnote{The node $\hat{n}$ which initiates the transition to $\Send$ 
is the root of the greatest active subtree of the whole guide tree (called such because $a_m = \Send \, \forall \, m \ntrianglerighteq \hat{n}$).}
If the branch transducer does not permit inserted bifurcations then $\hat{n} = 1$ always,
but this is generically not true for a more general grammar
(for example, see the TKF91 Structure Tree model).

We require:

\begin{tabular}{rl}
  Nodes $m \ntrianglerighteq \hat{n}$: & $a_m = \Send$. \\
  Node $\hat{n}$: & $\type(a_{\hat{n}}) \in \{\Sstart,\,\Sinsert\}$ \\
  & $\exists$ a transition $a_{\hat{n}} \to \Send$.  \\
  Nodes $m \vartriangleright \hat{n}$: & $\type(a_m) = \Swait$ \\
  & $\exists$ a transition $a_m \to \Send$. \\
\end{tabular}

In many probabilistic models, the transition from a state of type $\Swait$ to the $\Send$ state has weight 1 conditional on absorbing 
the $\Send$ symbol (called $\epsilon$ in formal grammar theory),
but we here allow for a more general contribution to the total weight $\mathbb{F}$ of the transition.

$\hat{n}$ and $\weight(\bvec{a}\to\Send)$ are so defined in order to ensure that there exists a direct path
along which the end symbol can be passed down the tree.
The grammar should be designed such that if the singlet transducer at node $\hat{n}$ can make a transition to $\Send$,
then so can all machines at $\{m|m \trianglerighteq \hat{n}\}$.
The TKF Structure Tree model (\secref{tkfst}) satisfies this condition.
A more general approach is probably possible, but it involves summing over paths $a': a_m \to a' \to \Send$,
handling possible bifurcations in these paths, etc.

\newpage
\section{Inference with multi-sequence SCFGs}

The model construction algorithm described above creates a composed state machine
which generates the joint distribution $P(X, Y, Z, W)$.
As described in \secref{multigrammars}, there exists a corresponding multi-sequence SCFG
which generates the same probability distribution.
It follows that we can perform inference on our composed model with the well-known Sankoff
algorithms.

As described in the paper, we can use the CYK algorithm directly, but
the Inside and Outside algorithms fail for generic multi-sequence models
due to the presence of looping $\Snull$ transitions.  We here give algorithms for inference 
in the case of no looping $\Snull$ transitions and explain how to modify them to 
create {\bf loopy DP} iterative algorithms for our models.

We here give versions for the model of three extant sequences
show in \figref{threeway}; the generalization to $N$ sequences is straightforward.

\begin{figure}[!htb]
  \centering
  \includegraphics [scale=0.4] {figs/threeway.pdf}
  \caption{\figlabel{threeway}
    A star phylogeny with three (extant) leaf sequences.
    An ancestral sequence $W$ evolves into three descendant sequences $X$, $Y$ and $Z$.
    A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
    and three branch transducers (the gray boxes labeled $\Delta T_X$, $\Delta T_Y$ and $\Delta T_Z$)
    mutate it according to the specified multi-sequence model.
    Gray nodes correspond to observed data and white nodes unobserved data.
    If the branch transducers are time-reversible, then this star phylogeny with three leaves is
    the neighborhood of any interior node in a (binary) phylogenetic tree, 
    from which it follows that evaluating the likelihood function
    on this star phylogeny is sufficient for a sampling algorithm on an arbitrary phylogeny.
  }
\end{figure}

\subsection{Notation}
The 3 (ungapped) sequences to be aligned are $x^{(1)}$, $x^{(2)}$ and $x^{(3)}$
(we refer to them by an index rather than as $X$, $Y$ and $Z$ for notational convenience).
The $\mathrm{i}^{\mathrm{th}}$ symbol of sequence $x^{(1)}$ is written $x^{(1)}_i$ 
and the length of the same sequence is $L_1 = |x^{(1)}|$.

\subsection{Fold Envelopes} \seclabel{foldenv}
We use the fold envelope concept \cite{HolmesRubin2002a,Holmes2004} to constrain the set of structures our algorithms consider.
A fold envelope $\foldenv$ for a sequence $x$ is a set of coordinate pairs satisfying
\beqn \label{eqn:cabbage}
\foldenv \subset \left\{ (i,j):0 \le i \le j \le L \right\} \, .
\eeqn
We consider a subsequence $x_{i+1} \dots x_{j}$ only if the corresponding coordinate pair $(i,j) \in \foldenv$.
The unconstrained fold envelope has set equality in \eqref{eqn:cabbage}.

We use two orderings for sequences in the fold envelopes.  An inside-outside ordering is used for the iteration 
in the Inside algorithm: Subsequences are ordered such that each successive subsequence 
contains all previous subsequences in the fold envelope.  More precisely, subsequences in $\mathscr{F}$ are sorted  
in the same order as coordinate pairs $(i,j)$ are generated by the iteration $\{\,\mathrm{for}\, i = L \,\mathrm{to}\, 0 \,\,\{ \mathrm{for}\, j = i \,\mathrm{to}\, L \}\,\}$.

The Outside algorithm uses an outside-inside ordering, where subsequences starting at a fixed $i$ are reverse-sorted 
with respect to the inside-outside ordering described above.  Subsequences in $\mathscr{F}$ are then sorted 
in the same order as coordinate pairs $(i,j)$ are generated by the 
iteration $\{\,\mathrm{for}\, i = 0 \,\mathrm{to}\, L \,\,\{ \mathrm{for}\, j = L\, \mathrm{to}\, i \}\,\}$.

We frequently refer to subsequences by their index in the fold envelope.
The $\mth$ subsequence in $\foldenv$ is labeled $m$ and corresponds to the coordinate pair $(i_{m}, j_{m})$.  The 
index of a pair $(i,j) \in \foldenv$ is $n[i,j]$.

Inward and outward emission connections, specifying which subsequences are reachable from 
a given subsequence and state, are defined as
\begin{align}
c_{in}\left(\bvec{b}; m\right) &= n\left[ i_{m} + \Delta_{\bvec{b}}^{L}, j_{m} - \Delta_{\bvec{b}}^{R} \right] \\
c_{out}\left(\bvec{b}; m\right) &= n\left[ i_{m} - \Delta_{\bvec{b}}^{L}, j_{m} + \Delta_{\bvec{b}}^{R} \right]
\end{align}
The emission connection is undefined if the corresponding subsequence is not in the fold envelope.\footnote{We defined emission connections slightly differently than do \cite{HolmesRubin2002a}.}  Inward, outward-left and outward-right bifurcation connections are defined as
\begin{align}
b_{in}(n) &= \left\{ \left( n_L,n_R \right): i_{n_L} = i_{n},\, j_{n_L} = i_{n_R},\, j_{n_R} = j_{n} \right\} \\
b_{out,L}(n) &= \left\{ \left( n_O,n_L \right): i_{n_L} = i_{n_O},\, j_{n_L} = i_{n},\, j_{n} = j_{n_O} \right\} \\
b_{out,R}(n) &= \left\{ \left( n_O,n_R \right): i_{n} = i_{n_O},\, j_{n} = i_{n_R},\, j_{n_R} = j_{n_O} \right\}
\end{align}

We generally write out explicit subsequence coordinate pairs $(i,j)$ when their usage will make mathematical formulae clearer and 
fold-envelope labels $n$ when writing pseudocode.  An implementation will rely on such an iterator over subsequences in the fold envelope.

\subsection{Inside}
The inside probability $\inside{\bvec{a}}{i_1}{j_1}{i_2}{j_2}{i_3}{j_3}$ is the summed probability of subsequences 
$x^{(1)}_{i_1+1} \dots x^{(1)}_{j_1}$, $x^{(2)}_{i_2+1} \dots x^{(2)}_{j_2}$, $x^{(3)}_{i_3+1} \dots x^{(3)}_{j_3}$
under all paths through the model which are rooted in state $\bvec{a}$.  
\algref{inside} gives pseudocode for the fold-envelope version of the Inside algorithm.  The subroutines 
$\mathrm{calcTransEmitProb}()$, $\mathrm{calcLBifurcProb}()$ and $\mathrm{calcRBifurcProb}()$
used in the Inside algorithm are defined below.

\begin{algorithm}[!ht]

  \SetKwData{bifurcProb}{bifurcProb}
  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
  \SetLine
  \KwIn{sequences $x^{(1)},x^{(2)},x^{(3)}$}

  initialization\;
  \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(1)} \in \foldenv^{(1)}$} {
    \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(2)} \in \foldenv^{(2)}$} {
      \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(3)} \in \foldenv^{(3)}$} {
        \BlankLine
        \ForEach{state $\bvec{a}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(1)}_L,n^{(1)}_R \right) \in b_{in}\left( n^{(1)} \right)$} {
            \ForEach{$\left( n^{(2)}_L,n^{(2)}_R \right) \in b_{in}\left( n^{(2)} \right)$} {
              \ForEach{$\left( n^{(3)}_L,n^{(3)}_R \right) \in b_{in}\left( n^{(3)} \right)$} {
                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\alpha_{\bvec{a}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$ $\leftarrow$ \calcTransEmitProb$\left( \bvec{a}; n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
          $\alpha_{\bvec{a}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$ += \bifurcProb\;
          store $\alpha_{\bvec{a}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \KwRet{$\alpha_{\bvec{a}}\left( n[0,L_1],n[0,L_2],n[0,L_3] \right)$}\;

  \caption{\alglabel{inside}
    The constrained Inside algorithm for three sequences $x^{(1)},x^{(2)},x^{(3)}$.
    States $\bvec{a}$ in the iteration over states are sorted in Inside fill order
    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
  }
\end{algorithm}

The transition and emission probability $\mathrm{calcTransEmitProb}(\bvec{a}; \cdot)$ can be calculated recursively 
in the unconstrained case as
\begin{align} \label{eqn:insidetransemit}
  & \sum_{y_1\in\{x^{(1)}_{i_1+1},\Tnull\}}\sum_{z_1\in\{x^{(1)}_{j_1},\Tnull\}} \sum_{y_2\in\{x^{(2)}_{i_2+1},\Tnull\}}\sum_{z_2\in\{x^{(2)}_{j_2},\Tnull\}} \sum_{y_3\in\{x^{(3)}_{i_3+1},\Tnull\}}\sum_{z_3\in\{x^{(3)}_{j_3},\Tnull\}} \nonumber \\
  &\qquad \left[ \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{y}\bvec{b}\bvec{z}} \weight (\bvec{a}\to\bvec{y}\,\bvec{b}\,\bvec{z}) \inside{\bvec{b}}{i_1+|y_1|}{j_1-|z_1|}{i_2+|y_2|}{j_2-|z_2|}{i_3+|y_3|}{j_3-|z_3|} \right] . \nonumber
\end{align}
Null transitions are caught in the sum when the terminals $\bvec{y} = \bvec{z} = \Tnull$.
The constrained case is handled differently: The multiple sums are replaced by a single iteration over states $\bvec{b}$
which connect subsequences $n^{(1)}, n^{(2)}, n^{(3)}$ to others in the fold envelopes.
Pseudocode for the constrained calculation is given in \algref{insidetransemit}.

\begin{algorithm}[!ht]
  \SetKwData{emitProb}{emitProb}
  \SetKwFunction{Weight}{Weight}
  \SetLine
  \KwIn{state $\bvec{a}, n^{(1)}, n^{(2)}, n^{(3)}$, intermediate Inside matrix $\alpha$}

  \emitProb $\leftarrow 0$\;
  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
    \emitProb += \Weight$\left( \bvec{a} \to \bvec{b} \right)\, \alpha_{\bvec{b}}\left( n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
  }
  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{y}\,\bvec{b}\,\bvec{z}$} {
    \lIf{$c_{in}\left(\bvec{b}; n^{(1)}\right) \notin \foldenv^{(1)}$ or $c_{in}\left(\bvec{b}; n^{(2)}\right) \notin \foldenv^{(2)}$ or $c_{in}\left(\bvec{b}; n^{(3)}\right) \notin \foldenv^{(3)}$}{next\;}
    \emitProb += \Weight$\left( \bvec{a} \to \bvec{y}\,\bvec{b}\,\bvec{z} \right)\, \alpha_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(1)}\right), c_{in}\left(\bvec{b}; n^{(2)}\right), c_{in}\left(\bvec{b}; n^{(3)}\right) \right)$\;
  }
  \KwRet{\emitProb}\;
  \caption{\alglabel{insidetransemit}
    Subroutine $\mathrm{calcTransEmitProb}()$ for the Inside algorithm.
  }
\end{algorithm}

The left-bifurcation probability $\mathrm{calcLBifurcProb}\left( \bvec{a}; n^{(1)}_L,n^{(1)}_R,n^{(2)}_L,n^{(2)}_R,n^{(3)}_L,n^{(3)}_R \right)$ is
\[ \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} \weight (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \alpha_{\bvec{c}}\left( n^{(1)}_L,n^{(2)}_L,n^{(3)}_L\right) \, \alpha_{\bvec{b}}\left( n^{(1)}_R,n^{(2)}_R,n^{(3)}_R\right) \]
and the right-bifurcation probability $\mathrm{calcRBifurcProb}\left( \bvec{a}; n^{(1)}_L,n^{(1)}_R,n^{(2)}_L,n^{(2)}_R,n^{(3)}_L,n^{(3)}_R \right)$ is
\[ \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} \weight (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \alpha_{\bvec{b}}\left( n^{(1)}_L,n^{(2)}_L,n^{(3)}_L\right) \, \alpha_{\bvec{d}}\left( n^{(1)}_R,n^{(2)}_R,n^{(3)}_R\right) . \]

The boundary condition gives the probability of a subsequence of length 0,
\beqn \label{eqn:insidetoend}
\inside{\bvec{a}}{j_1}{j_1}{j_2}{j_2}{j_3}{j_3} = t(\bvec{a},\Send) + \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}} \weight (\bvec{a}\to\bvec{b}) \, \inside{\bvec{a}}{j_1}{j_1}{j_2}{j_2}{j_3}{j_3} \, ,
\eeqn
for $0 \le j_1 \le L_1, 0 \le j_2 \le L_2, 0 \le j_3 \le L_3$.
We are assuming that there are no cycles of $\Snull$ states as well as no bifurcations which can result in no emissions.
The termination condition is 
\[ P\left(x^{(1)},x^{(2)},x^{(3)}\right) = \alpha_{\mathrm{Start}}(0,L_1,0,L_2,0,L_3) \, , \]
where $\mathrm{Start}$ is the unique start state of the model.

We write the DP algorithms using the $\weight (\cdot)$ notation in order to preserve generality: In our formalism, a transition $a \to x\,b$ has $\emit(b) = x$, but the more common convention is to have $\emit(a) = x$.
The difference will show up only in the value assigned to $\weight (a \to x\,b)$, so our DP algorithms can be used in both cases.

\subsection{CYK}
The CYK algorithm can be obtained from the Inside algorithm by replacing sums over paths through the model (or equivalently, parses)
with the $\max()$ operation (e.g., in \eqref{eqn:insidetoend}, $\sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}}$ will be replaced by $ \max_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}}$).
The CYK probabilities for indices $(i_1,j_1,i_2,j_2,i_3,j_3)$ then represent 
the probability of the most likely path through the model for 
subsequences $x^{(1)}_{i_1+1} \dots x^{(1)}_{j_1}$, $x^{(2)}_{i_2+1} \dots x^{(2)}_{j_2}$, $x^{(3)}_{i_3+1} \dots x^{(3)}_{j_3}$.

\subsection{Outside}
The outside probability $\outside{\bvec{b}}{i_1}{j_1}{i_2}{j_2}{i_3}{j_3}$ is the summed probability of the sequences $x^{(1)}$, $x^{(2)}$, $x^{(3)}$ 
under all paths through the model which are rooted in the start state of the model, excluding all paths for the subsequences 
$x^{(1)}_{i_1+1} \dots x^{(1)}_{j_1}$, $x^{(2)}_{i_2+1} \dots x^{(2)}_{j_2}$, $x^{(3)}_{i_3+1} \dots x^{(3)}_{j_3}$ 
which are rooted in the state $\bvec{b}$.
\algref{outside} gives pseudocode for the fold-envelope version of the Outside algorithm.
The subroutines $\mathrm{calcTransEmitProb}()$, $\mathrm{calcLBifurcProb}()$ and $\mathrm{calcRBifurcProb}()$
used in the Outside algorithm are defined below.

\begin{algorithm}[!ht]
  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
  \SetKwData{bifurcProb}{bifurcProb}
  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
  \SetLine
  \KwIn{sequences $x^{(1)},x^{(2)},x^{(3)}$, CYK matrix $\alpha$}

  initialization\;
  \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(1)} \in \foldenv^{(1)}$} {
    \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(2)} \in \foldenv^{(2)}$} {
      \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(3)} \in \foldenv^{(3)}$} {
        \BlankLine
        \ForEach{state $\bvec{b}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(1)}_O,n^{(1)}_L \right) \in b_{out,L}\left( n^{(1)} \right)$}{
            \ForEach{$\left( n^{(2)}_O,n^{(2)}_L \right) \in b_{out,L}\left( n^{(2)} \right)$}{
              \ForEach{$\left( n^{(3)}_O,n^{(3)}_L \right) \in b_{out,L}\left( n^{(3)} \right)$}{
                \bifurcProb += \calcLBifurcProb($\bvec{b}; \cdot$)\;
              }
            }
          }
          \ForEach{$\left( n^{(1)}_O,n^{(1)}_R \right) \in b_{out,R}\left( n^{(1)} \right)$}{
            \ForEach{$\left( n^{(2)}_O,n^{(2)}_R \right) \in b_{out,R}\left( n^{(2)} \right)$}{
              \ForEach{$\left( n^{(3)}_O,n^{(3)}_R \right) \in b_{out,R}\left( n^{(3)} \right)$}{
                \bifurcProb += \calcRBifurcProb($\bvec{b}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\beta_{\bvec{b}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right) \leftarrow$ \calcTransEmitProb$\left( \bvec{b}; n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
          $\beta_{\bvec{b}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$ += \bifurcProb\;
          store $\beta_{\bvec{b}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \caption{\alglabel{outside}
    The constrained Outside algorithm for three sequences $x^{(1)},x^{(2)},x^{(3)}$.
    States $\bvec{a}$ in the iteration over states are sorted in Outside fill order
    with $\Semit$ states first, then $\Snull$ states in topological order.
  }
\end{algorithm}

The transition and emission probability $\mathrm{calcTransEmitProb}()$ can be calculated recursively in the unconstrained case as
\begin{align}
  & \sum_{y_1\in\{x^{(1)}_{i_1},\Tnull\}}\sum_{z_1\in\{x^{(1)}_{j_1+1},\Tnull\}} \sum_{y_2\in\{x^{(2)}_{i_2},\Tnull\}}\sum_{z_2\in\{x^{(2)}_{j_2+1},\Tnull\}} \sum_{y_3\in\{x^{(3)}_{i_3},\Tnull\}}\sum_{z_3\in\{x^{(3)}_{j_3+1},\Tnull\}} \nonumber \\
  & \qquad \left[ \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{y}\bvec{b}\bvec{z}} \weight (\bvec{a}\to\bvec{y}\,\bvec{b}\,\bvec{z}) \, \outside{\bvec{a}}{i_1-|y_1|}{j_1+|z_1|}{i_2-|y_2|}{j_2+|z_2|}{i_3-|y_3|}{j_3+|z_3|}  \right] . \nonumber
\end{align}
As with the Inside algorithm, an efficient implementation of the fold-envelope constraints requires a different treatment, given in \algref{outsidetransemit}.

The left-bifurcation probability $\mathrm{calcLBifurcProb}\left( \bvec{b}; n^{(1)}_O,n^{(1)}_L,n^{(2)}_O,n^{(2)}_L,n^{(3)}_O,n^{(3)}_L \right)$ is
\[ \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} \weight (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \beta_{\bvec{a}}\left( n^{(1)}_O,n^{(2)}_O,n^{(3)}_O\right) \, \alpha_{\bvec{c}}\left( n^{(1)}_L,n^{(2)}_L,n^{(3)}_L\right) \]
and the right-bifurcation probability $\mathrm{calcRBifurcProb}\left( \bvec{b}; n^{(1)}_O,n^{(1)}_R,n^{(2)}_O,n^{(2)}_R,n^{(3)}_O,n^{(3)}_R \right)$ is
\[ \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} \weight (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \beta_{\bvec{a}}\left( n^{(1)}_O,n^{(2)}_O,n^{(3)}_O\right) \, \alpha_{\bvec{d}}\left( n^{(1)}_R,n^{(2)}_R,n^{(3)}_R\right) \]

\begin{algorithm}[!ht]
  \SetKwData{emitProb}{emitProb}
  \SetKwFunction{Weight}{Weight}
  \SetLine
  \KwIn{state $\bvec{b}, n^{(1)}, n^{(2)}, n^{(3)}$, intermediate Outside matrix $\beta$}

  \emitProb $\leftarrow 0$\;
  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{b}$} {
    \emitProb += \Weight$\left( \bvec{a} \to \bvec{b} \right)\, \beta_{\bvec{a}}\left( n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
  }
  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{y}\,\bvec{b}\,\bvec{z}$} {
    \lIf{$c_{out}\left(\bvec{b}; n^{(1)}\right) \notin \foldenv^{(1)}$ or $c_{out}\left(\bvec{b}; n^{(2)}\right) \notin \foldenv^{(2)}$ or $c_{out}\left(\bvec{b}; n^{(3)}\right) \notin \foldenv^{(3)}$}{next\;}
    \emitProb += \Weight$\left( \bvec{a} \to \bvec{y}\,\bvec{b}\,\bvec{z} \right)\, \beta_{\bvec{a}}\left( c_{out}\left(\bvec{b}; n^{(1)}\right), c_{out}\left(\bvec{b}; n^{(2)}\right), c_{out}\left(\bvec{b}; n^{(3)}\right) \right)$\;
  }
  \KwRet{\emitProb}\;
  \caption{\alglabel{outsidetransemit}
    Subroutine $\mathrm{calcTransEmitProb}()$ for the Outside algorithm.
  }
\end{algorithm}

The boundary condition is just
\beqn
\outside{\mathrm{Start}}{0}{L_1}{0}{L_2}{0}{L_3} = 1 \, .
\eeqn

\subsection{Loopy DP}

The standard Inside and Outside algorithms fail on grammars with 1) cycles of $\Snull$ states, 
or 2) bifurcations with empty children
because they are incapable of performing a probabilistic summation over $\Snull$ cycles
(empty children of bifurcations can give rise to effective cycles of $\Snull$ states).
Phrased qualitatively, the standard Inside and Outside fail because the iterations over
states $\bvec{a}$ in \algref{inside} and \algref{outside}
can correspond to, at most, single $\Snull$ cycles rather than the infinitely many which are possible.

While we believe that an efficient analytic summation is impossible,
we can use an iterative approach to approximate an exact result to high precision.
A brute-force approach is to just repeatedly iterate over all states $\bvec{a}$ until
convergence is reached, but this is computationally-costly.
A much more efficient algorithm exists.
The key insight is that only $\Snull$ states which are strongly connected
(reachable from each other in the state graph by either other $\Snull$ states
or bifurcations with empty children) can contribute to any possible $\Snull$ cycle.
It follows that instead of repeatedly iterating over all states $\bvec{a}$,
it suffices to decompose the state (sub) graph of $\Snull$ states and bifurcation
states with possibly-empty children into its strongly-connected components
and repeatedly iterate over these connected components rather than the entire graph.
Decomposition of a graph into its strongly-connected components can
be done in linear time \cite{Tarjan1972},
so the overall complexity of inference is still determined by the DP iterations themselves.
If we iterate over each connected component 5 times, then upper bounds on 
the complexity of inference are
$O(5 \cdot A \cdot L^{2n})$ and $O(5 \cdot A \cdot L^{3n})$ in space and time for a model 
with $A$ states of $n$ extant sequences.
For the star phylogeny of \figref{threeway},

\newpage
\section{The TKF Structure Tree model as a transducer} \seclabel{tkfst}

We can cast the TKF91 Structure Tree model as a transducer.
{\bf Why is this important/useful? e.g. ``This means we can automatically deduce rules like those shown in Tables~\tabnum{abc} through~\tabref{agapbc}...''}

\subsection{The single-sequence TKFST model as a singlet transducer}

The states of the singlet transducer are shown in \tabref{tkfstsinglet}.
Allowed transitions between these states are shown in the paper.

\begin{table}[!ht]
  \centering
  \begin{tabular}{lrrrr}
    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
    $L$ & $\Sstart$ \\
    $I_L$ & $\Sinsert$ & & $(x,\Tnull)$ & $p(x)$ \\
    \\
    $S$ & $\Sstart$ \\
    $I_S$ & $\Sinsert$ & & $(x,y)$ & $p(x,y)$ \\
    \\
    $B$ & $\Sinsert$ & & $(L\,S)$ & 1 \\
  \end{tabular}
  \caption{
    \tablabel{tkfstsinglet}
    States of the singlet transducer of the TKF Structure Tree model \cite{Holmes2004}.
    Singlet transducers can only have states of type $\Sstart$ or $\Sinsert$.
    {\bf This is the indiegram-style transducer equivalent of the SCFG in \tabref{a} ...}
  }
\end{table}

\subsection{The two-sequence TKFST model as a branch transducer}

The states of the branch transducer are shown in \tabref{tkfstbranch}.
Allowed transitions between these states are shown in the paper.

\begin{table}[!ht]
  \centering
  \begin{tabular}{lrrrr}
    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
    $L$ & $\Sstart$ \\
    $I_L$ & $\Sinsert$ & & $(u,\Tnull)$ & $p(u)$ \\
    $M_L$ & $\Smatch$ & $(x,\Tnull)$ & $(u,\Tnull)$ & $p(u|x)$ \\
    $D_L$ & $\Smatch$ & $(x,\Tnull)$ & $(\Tgap,\Tnull)$ & 1 \\
    $W_L$ & $\Swait$ \\
    \\
    $S$ & $\Sstart$ \\
    $I_S$ & $\Sinsert$ & & $(u,v)$ & $p(u,v)$ \\
    $M_S$ & $\Smatch$ & $(x,y)$ & $(u,v)$ & $p(u,v|x,y)$ \\
    $D_S$ & $\Smatch$ & $(x,y)$ & $(\Tgap,\Tgap)$ & 1 \\
    $W_S$ & $\Swait$ \\
    \\
    $B_i$ & $\Sinsert$ & & $(L_i\,S_i)$ & 1 \\
    $B$ & $\Smatch$ & $(L\,S)$ & $(L\,S)$ & 1 \\
    $B_p$ & $\Smatch$ & $(L\,S)$ & $(L\,\Send)$ & 1 \\
    $B_e$ & $\Smatch$ & $(L\,\Send)$ & $(L\,\Send)$ & 1 \\
  \end{tabular}
  \caption{\tablabel{tkfstbranch}
    States of the branch transducer of the TKF Structure Tree model \cite{Holmes2004}.
    States which have the same names as states of the singlet transducer in \tabref{tkfstsinglet}
    are the branch equivalent of the corresponding singlet states
    (e.g. a $\Smatch$ state might be the branch equivalent of an $\Sinsert$ state).
    States $L_i$ and $S_i$ are the $\Sstart$ states of a sub-model (not shown) identical
    in structure to the singlet transducer.  They are used to insert a new stem-loop structure.
    {\bf This is the indiegram-style transducer equivalent of the SCFG in \tabref{ab} ...}
  }
\end{table}

\subsection{The multi-sequence TKFST model as a composite transducer}

We can use the state graph construction algorithm described 
in the paper and detailed in \secref{stategraph} to create a model of the simultaneous evolution
of several sequences.

\begin{figure}[!htb]
  \centering
  \includegraphics [scale=0.4] {figs/parent.pdf}
  \caption{\figlabel{parent}
    A simple example of transducer composition to build a multi-sequence model of two extant sequences.
    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
    A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
    and two branch transducers (the gray boxes labeled $\Delta T_X$ and $\Delta T_Y$)
    mutate it according to the specified multi-sequence model.
    Gray nodes correspond to observed data and white nodes unobserved data.
  }
\end{figure}

Consider the simple model shown in \figref{parent}.
The state of the multi-sequence model describing this model is a 3-vector $\bvec{a} = \left( a_1,\,a_2,\,a_3 \right)$,
where $a_1$ is the state of the singlet transducer generating the ancestral sequence $W$
and $a_2$ and $a_3$ are the states of the branch transducers evolving $W$ into
extant sequences $X$ and $Y$.

We can show some of the allowed transitions of this multi-sequence model.
The state of the branch transducer associated with the active node $n$ is shown in bold.

\paragraph{Stem creation.}

A stem is created at the root node $W$ (corresponding to a bifurcation in the singlet transducer $a_1$)
and survives in the sequence $X$ at node 2 but is deleted in the sequence $Y$ at node 3.

\begin{align}
  \nakedthreevec{1:}{2:}{3:} \quad
  \threevec{L}{L}{\mathbf{L}} &
  \to \threevec{\mathbf{L}}{W_L}{W_L}
  \to \threevec{B}{B}{B_p}
  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{S}{\mathbf{S}}{\Send}
  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\mathbf{S}}{W_S}{\Send} \\
  & \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\mathbf{S}}{W_S}{\Send}
\end{align}

\paragraph{Stem insertion.}

A stem sequence is inserted in sequence $X$ at node 2.

\begin{align}
  \nakedthreevec{1:}{2:}{3:} \quad
  \threevec{L}{L}{\mathbf{L}} \to \threevec{L}{\mathbf{L}}{W_L}
  \to \threevec{L}{B_i}{B}
  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\Send}{S}{\mathbf{S}}
  \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\Send}{\mathbf{S}}{W_S}
\end{align}

\paragraph{Stem termination.}

All stem sequences are ended by (possibly empty) loop sequences.

\begin{align}
  \nakedthreevec{1:}{2:}{3:} \quad
  \threevec{\mathbf{S}}{W_S}{W_S}
  \to \threevec{B_e}{B_e}{B_e}
  \to \threevec{\Send}{\Send}{\Send} \, \threevec{L}{L}{\mathbf{L}}
  \to \threevec{\Send}{\Send}{\Send} \, \threevec{\Send}{\Send}{\Send}
\end{align}

The functions $\alpha(t)$, $\beta(t)$ and $\gamma(t)$ are parametrized by the insertion and deletion rates of the Structure Tree model.
They are defined for loop sequences as
\begin{align}
  \kappa_1 &= \lambda_1 / \mu_1 \\
  \alpha_1 &= \exp \left( -\mu_1 t \right) \\
  \beta_1 &= \frac{\lambda_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) } \\
  \gamma_1 &= 1 - \frac{\mu_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\left( 1 - \exp (- \mu_1 t) \right) \left(\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) \right) }
\end{align}
and similarly for stem sequences \cite{Holmes2004}.

\newpage
\input{tkfst.tex}

\newpage
\section{Software}

We have written software tools in Perl and C++ implementing much of the theory described here.
All tools are available from \darturl\ as part of the \dart\ software package for sequence analysis.

\subsection{Automated grammar construction}
We implemented our model construction algorithm on the star phylogeny (\figref{threeway}) 
in a set of Perl scripts.
Given a singlet transducer modeling ancestral sequences and structures and
a branch transducer modeling structural evolution,
the scripts generate C++ code to create the corresponding jointly-normalized SCFG.
All possible models of structural evolution which can be represented by a Pair SCFG
are permitted as input to the scripts,
allowing for flexible and automated model design.

Given files describing the singlet and branch transducers,
including weights of all transitions which may be functions of evolutionary distance,
the package ComposedTreeTransducer::FourWayComposedTT can automatically generate
the state graph and transition matrix of a multi-sequence model of three extant sequences
(\figref{threeway}).
It removes the useless windback $\Snull$ states described in the paper and
introduces effective direct transitions caused by bifurcations with possibly-empty children.
The package ComposedTreeTransducer::TripletSCFG transforms a multi-sequence
model created by ComposedTreeTransducer::FourWayComposedTT into the 
corresponding jointly-normalized three-sequence SCFG (\secref{multigrammars})
and generates C++ code to build the model.

Example singlet and branch transducers files are provided for a simple 
Pair HMM model, a simple Pair SCFG model and the full TKF Structure Tree model.

\subsection{Reconstruction of ancestral structures}
The program \indiegram\ can perform maximum-likelihood inference on the
three-sequence SCFGs automatically generated by the ComposedTreeTransducer::TripletSCFG package.
Complete or no structural information for the three extant sequences can be supplied
as input.

\newpage
\section{Reconstruction of covariant substitution histories}

Our \xrate\ program is a multiple-alignment analysis tool
that estimates rate parameters for a broad class of models, including covariant RNA substitution models \cite{KlostermanEtAl2006}.
It can also be used to predict consensus secondary structures for alignments.
We implemented ancestral sequence reconstruction in \xrate, including posterior probabilities that the reconstructions are correct.

Given a multiple sequence alignment and a phylogenetic tree,
\xrate\ estimates maximum-likelihood values of rate and probability parameters by Expectation Maximization.
During parameter estimation, any unspecified ancestral sequences are summed out using Felsenstein's peeling algorithm.
The new feature is that the program can now find the ancestral sequence that has the highest posterior probability,
contingent on the maximum-likelihood estimates of the model parameters.
During the parameter estimation and ancestral reconstruction steps, the secondary structure may be specified, or it may be summed out as a latent variable.

For the ancestral reconstruction experiments described here,
we assumed that alignment, phylogeny and secondary structure were known,
but that ancestral sequences were unknown.

A simple computational experiment demonstrates the need for covariant substitution models.
We first used \xrate\ to estimate maximum-likelihood parameters for a covariant model of RNA base-pair substitution.
These rates were estimated from ribosomal RNA alignments
derived by \cite{DowellEddy2006} from the European rRNA database \cite{WuytsEtAl2004}.
We also fit a ``naive'' general reversible point-substitution model to these alignments, again using \xrate.
All subsequent results are implicitly conditioned on these maximum-likelihood rate estimates.
Next, we used the companion program SIMGRAM to generate 5000 random alignments (including ancestral sequence),
simulating on a 75-taxon phylogeny from an RNA gene family in the \RFAM\ database \cite{GriffithsJonesEtAl2003}.
(Specifically, we chose the type-I Hammerhead ribozyme, one of the 5\% most populous \RFAM\ families.
 We repeated this experiment with other \RFAM\ phylogenies, but the general trends reported here were not dependent on the choice of tree.)
We stripped the ancestral sequence out of the simulated alignments (leaving the true structure annotation intact),
then re-estimated the ancestral sequence with \xrate, using both the (correct) covariant substitution model and the (naive) non-covariant point-substitution model.
The imputed ancestral sequences were compared to the true sequences (known from the simulation),
and the model-derived posterior probabilities were compared to the empirical accuracy of the corresponding reconstructed sequence.

\begin{figure}[ht]
  \begin{center}
     \includegraphics [width=0.45\textwidth] {figs/hammer_paired.pdf}
     \hspace*{\fill}
     \includegraphics [width=0.45\textwidth] {figs/xval_paired.pdf}
  \end{center}
  \caption{
    \figlabel{simulation}
    Accuracy of confidence estimates for base-pair reconstruction in simulations (left) and cross-validations on real data (right).
    Reconstructions were performed using \xrate\ \cite{KlostermanEtAl2006},
    using both covariant base-pair substitution models
    and naive, non-covariant, single-base point-substitution models.
    The plots show the relationship between the posterior probability that a model-based base-pair reconstruction is correct (vertical axis),
    and the empirical probability that reconstructions with that posterior probability are actually correct (horizontal axis).
    If the model's posterior probability estimates are correct, these points should lie on a straight diagonal line.
    Left: simulation test on the hammerhead ribozyme phylogeny.
    A covariant model (estimated from rRNA alignments) was used to generate the simulated alignments on the given phylogeny; the same model was then used to reconstruct ancestral sequence and obtain confidence estimates.
    As expected, the model accurately estimates the probability that the reconstructed base-pair is correct.
    A naive, non-covariant model (the general reversible model, estimated from the same alignments) was also used for reconstruction and confidence estimates.
    This model {\em under}-estimates the probability of correct reconstructions,
    since its equilibrium frequency has a higher entropy than the covariant model:
    i.e. it assigns probability more evenly across the sixteen possible base-pairs (as opposed to the covariant model, which strongly favors the six canonical base-pairs).
    Right: cross-validation (``hold-one-out'') test using \RFAM\ alignments with published (verified) structures.
    Here, the covariant model slightly {\em over}-estimates the probability of correctness.
    The rRNA alignments (from which the model rates were estimated) have fewer non-canonical base-pairs than the \RFAM\ alignments,
    so that the covariant model entropy is slightly {\em lower} than the empirical base-pair distribution;
    this may explain the discrepancy.
    The naive model's confidence estimates are correct
    for high-accuracy reconstructions,
    over-estimates for medium-accuracy reconstructions,
    and under-estimates for low-accuracy reconstructions.
  }
\end{figure}

\figref{simulation} (left) shows the results of these comparisons.
When using the same (covariant) model for simulation and reconstruction, the posterior probabilities are an excellent unbiased estimate of the frequency with which the model reconstructions are correct.
However, the naive point-substitution model, which does not include covariant substitutions at base-paired sites, systematically under-estimates these probabilities.
% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | wc -l` / `wc -l $f`,"\n"}' Hammerhead_1.bptest
In our simulation, the ancestral error rate for the naive model (4.9\%) was significantly higher than that of the covariant model (1.7\%).
% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Hammerhead_1.bptest
Furthermore, 74\% of incorrect base-pairs predicted by the naive model were non-canonical (i.e. not AU, CG, GC, UA, GU or UG),
compared to only 2\% of incorrect predictions by the covariant model.

\begin{figure}[p]
  \begin{center}
     \includegraphics [width=0.45\textwidth] {figs/hammer_hist.pdf}
     \hspace*{\fill}
     \includegraphics [width=0.45\textwidth] {figs/xval_hist.pdf}
  \end{center}
  \caption{
    \figlabel{histograms}
    Proportions of reconstructed base-pairs that are incorrect (and the subset of those that are non-canonical)
    in simulations (left) and cross-validations on real data (right).
    Base-pair reconstructions that are incorrect AND non-canonical
    might potentially disrupt a synthetic reconstructed structure.
    While the covariant model does not always get more base-pairs correct than the naive model,
    it predicts fewer non-canonical base-pairs.
    The sharpest difference between the models are in the lowest posterior-probability categories,
    corresponding to low-confidence predictions (e.g. sequences on deep branches, or base-pairs with few homologues).
    In this category, the covariant model also makes far fewer errors than the naive model.
    See main text and caption to \figref{simulation} for details of simulation and cross-validation procedures.
    \RFAM\ families with verified structures tended to be less phylogenetically diverse,
    with roughly half the total branch length (on average), than the tree used for our simulations;
    this may explain the relative paucity of low posterior-probability errors in the right-hand plot, compared to the left.
% perl -e 'use Newick;use Stockholm::Database;$db=Stockholm::Database->from_file(shift);for$a(@$db){next if$a->columns>500 || @{$a->seqname}>400;$tree=Newick->from_string(@{$a->gf_NH});$b=0;for$l(@{$tree->branch_length}){$b+=$l}print"$b\n"}' ~/rfam/Training/Published.stock | meansd.pl
% perl -e 'use Newick;use Stockholm::Database;$db=Stockholm::Database->from_file(shift);for$a(@$db){$tree=Newick->from_string(@{$a->gf_NH});$b=0;for$l(@{$tree->branch_length}){$b+=$l}print"$b\n"}' ~/rfam/Training/Hammerhead_1.stock
  }
\end{figure}

We also performed cross-validation (hold-one-out) experiments. Starting from \RFAM\ alignments whose secondary structure is labeled ``Published'' (i.e. experimentally validated),
we estimated phylogenetic trees using the neighbor-joining algorithm
with the Jukes-Cantor model,
then removed individual sequences
 and attempted to reconstruct them using \xrate.
(This is possible because the substitution models are time-reversible, so we can treat any node as ancestral.)
In this experiment, the error rates for base-pair reconstructions again differed by a few percent, though both error rates were higher than in the simulation test
% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | wc -l` / `wc -l $f`,"\n"}' Published.xval-bptest
(covariant model: 11\%, non-covariant model: 16\%).
% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Published.xval-bptest
Of the incorrectly-reconstructed base-pairs, the non-covariant model again predicted more non-canonical pairs (40\%) than the covariant model (17\%),
especially at lower confidence levels (\figref{histograms}, right).

Posterior probability estimates for both models were
systematically higher in the cross-validation test than they were in the simulation (\figref{simulation}, right).
The presence of this trend in both models
suggests that it may be due to differences between the European rRNA alignments
(from which the rates were estimated)
and the \RFAM\ alignments (which were used in the cross-validation test).
This is confirmed by our empirical observations of
covariant rate matrices estimated from the two alignment datasets:
base-pairs in the \RFAM\ alignments are more frequently
non-canonical, and appear to evolve faster (relative to single-stranded regions),
than base-pairs in the rRNA alignments
(for details, see {\tt biowiki.org/RnaModels}).
Repeating the cross-validation experiment with rates estimated from the \RFAM\ alignments,
we find the covariant model predicted slightly fewer non-canonical base-pairs
% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Published.pubxval-bptest
(14\% rather than 17\%)
and produced slightly more accurate posterior probability estimates (data not shown).
Although this is no longer a strict cross-validation (since parameters were estimated from the test dataset),
it nevertheless emphasizes the value of correct models.
%Another possible source of errors in the cross-validation (compared to the simulation)
%is inaccurately estimated phylogenetic topologies and branch lengths;
%this might be improvable by using a covariant model in the phylogenetic estimation step (rather than Jukes-Cantor),
%and/or by using Bayesian phylogenetic estimation tools (rather than neighbor-joining).

Accurate estimates of posterior probabilities are likely to be important:
sub-optimal predictions are a significant piece of the reconstruction puzzle \cite{WilliamsEtAl2006,PollockChang2007,ChangEtAl2007}.
Ancestral reconstruction by probabilistic modeling does not yield a single definitive sequence, but rather a probability distribution over sequences.
Combinatorial synthesis by degenerate primer assembly has been advocated as a means of sampling from this distribution \cite{Gaucher2007}.
Nucleotide-perfect reconstructions may not always be possible:
it may be more productive to aim for accurate confidence estimates
(including sub-optimal predictions)
than to focus exclusively on getting everything right the first time.

Based on these tests, we conclude that deep phylogenetic reconstructions of ribosomes, and other RNAs, will require covariant substitution models that take account of RNA secondary structure.
We may reasonably deduce that indel reconstructions will, similarly, need to take account of RNA structure.
These results, therefore, strongly motivate the development of multi-sequence models for RNA structure, such as the TKF Structure Tree.

\section{Terminological asides}

This subsection contains some asides about terminology,
and can be omitted on a first reading of this paper.
The intent of this subsection is to highlight connections to related fields
in molecular evolution and computer science.

\subsection{Phylogenetic grammars}

We here describe some relationships to grammar models in current use in molecular evolution.

% Removing this sentence since I do not think that describing the relationship between transducers and SCFGs as an "injective mapping" will make it any clearer for most readers -- it is a bit like describing a string as a "concatenative monoid" -- IH
%There exists an injective mapping from the evolutionary models generated by our model-construction algorithm to multi-sequence SCFGs. In other words, 
Given a singlet transducer, a phylogenetic tree, and a mapping from tree to branch transducers,
the transducer composition algorithm returns a corresponding multi-sequence SCFG
which models the joint probability distribution $P(X_1 , ... , X_{n+m})$.

Several related bioinformatics grammars can be represented within this multi-SCFG framework,
allowing for substantial re-use of prior modeling theory:
\begin{itemize}
\item String transducers (which are similar to HMMs) can be viewed as special cases of parse-tree transducers,
without the facility to model long-distance correlations (base-pairs) or bifurcated structures.
\item Phylo-HMMs and phylo-SCFGs \cite{KlostermanEtAl2006}
can be viewed as special cases of (respectively) string transducers and parse-tree transducers.
(Phylo-HMMs and phylo-SCFGs typically model substitutions but not indels,
and so can be viewed as compositions of a richly-featured singleton transducer with a restricted class of branch transducers.)
\item Sakakibara \cite{Sakakibara2003} has described ``Pair HMMs on tree structures'' that are closely related to our parse-tree transducers
(though without the composition algorithms for scoring multiple alignments).
\item Sakakibara has also described pair stochastic tree-adjoining grammars, or TAGs \cite{Sakakibara2004}.
TAGs are a formal generalization of SCFGs to
mildly context-sensitive grammars \cite{MamitsukaAbe94,JoshiSchabes97,RivasEddy2000}.
We speculate that the conditionally-normalized transducer analogues of such Pair-TAGs may encompass the
phylogenetic model of covariant RNA substitutions and indels on pseudoknotted structures
described by \cite{MeyerMiklos2007},
just as parse-tree transducers encompass phylo-SCFGs.
\end{itemize}

\subsection{Statistical Alignment}

The term {\bf Statistical Alignment} was introduced by Hein \cite{HeinEtal2000}
to describe DP algorithms for phylogenetic grammars.
Although this term has been adopted by others in molecular evolution \cite{Miklos2002,Metzler2003,DrummondRambaut2007,SatijaEtAl2008},
we've avoided it (except in this section) for the following four reasons:
(i) Several RNA alignment methods already use another probability distribution
that's completely different to (though equally as valid as) the phylogenetic probability distribution over alignments:
namely, the Boltzmann probability distribution over secondary structures
\cite{GorodkinEtAl97,MathewsTurner02,HofackerEtAl2004}.
(ii) Many alignment tools make use of prior and posterior distributions over pairwise alignments
\cite{Miyazawa94,BucherHofmann96,Holmes98,BrayPachter2004,DoEtAl2004,Holmes2005,LoytynojaGoldman2005,SchwartzPachter2007,BradleyPachterHolmes2008};
all of these are arguably ``statistical'' alignment tools as well.
(iii) Some of the above-cited methods use sum-over-pairs alignment metrics, taking the expectation of such metrics over pairwise alignment posteriors,
and finding the multiple alignment that maximizes this expectation \cite{DoEtAl2004,SchwartzPachter2007,BradleyPachterHolmes2008};
others do progressive multiple alignment using pair grammars \cite{BrayPachter2004,Holmes2005,LoytynojaGoldman2005}.
Speculatively, such pair-based heuristics may (in some sense) approximate probabilistic alignment inference with true phylogenetic likelihoods;
similar convergences have, at least, been demonstrated between other pair-based heuristics \cite{EickmeyerEtAl2008}.
Thus, even using Hein's strict definition,
it's possible that a broad range of methods systematically {\em approximate} ``Statistical Alignment'';
this cautions against using the term too exclusively.
(iv) Even more broadly, many alignment tools make use of statistics to estimate alignment ``significance''.
In this frequentist approach, probability distributions for alignment scores (particularly Gumbel distributions)
are estimated for random sequences.
The tail integrals of these distributions then yield P-values for high-scoring database search results
\cite{KarlinAltschul93,Mott2000,Eddy2008}.
Thus, while the term ``Statistical Alignment'' has a quite specific technical meaning in molecular evolution
(a meaning which certainly encompasses the algorithms we describe here),
the broader bioinformatics literature may have competing claims to the term.
In any case, the goal of this work is statistical {\em reconstruction} of RNA protosequences,
rather than alignment {\em per se}.

\subsection{Parse trees and stacks}

Grammars are abstract models for languages;
finite-state automata are the corresponding abstract models for those languages' parsers.
Most of this paper uses the grammatical view of RNA structure evolution;
in this short section we digress briefly to mention the automata-theoretic view of these models.

The two non-local features that set SCFGs apart from HMMs are bifurcations and paired emissions \cite{Durbin98}.
In RNA structure analysis, these features are used to model (respectively) branched-loops and base-pairs.

A stochastic pushdown automaton can also model these features.
Such an automaton is essentially an HMM with a LIFO stack (LIFO = ``Last In, First Out'').

The pushdown automaton can parse two sequences simultaneously, aligning them and their parse trees.
It handles bifurcations and paired emissions by pushing symbols onto its stack:
a bifurcation by pushing a nonterminal, and a paired emission by pushing a terminal.
The pushed terminals and nonterminals are popped off the stack when the machine reaches its $\Send$ state.

The conditionally-normalized version of a pushdown automaton is a pushdown transducer:
a machine that absorbs {\em and} parses an input sequence (or absorbs an already-parsed input sequence)
and emits a (parsed) output sequence.
This is the machine that we describe as a parse-tree transducer.

\bibliography{../latex-inputs/alignment,../latex-inputs/duplication,../latex-inputs/genomics,../latex-inputs/ncrna}

\end{document}

### 2009-04-18 make-stemloc-pairhmm-training-sets.pl

@@ -1,0 +1,34 @@
+#!/usr/bin/perl -w
+
+use Stockholm::Database;
+use Newick;
+
+my ($dbfile) = @ARGV;
+defined($dbfile) or die "Usage: $0 <Stockholm database>\n";
+
+my $db = Stockholm::Database->from_file ($dbfile);
+
+my $binsize = .1;
+my @bin;
+
+for my $stock (@$db) {
+    my $tree = Newick->from_string (@{$stock->gf_NH});
+    my $total_len = 0;
+    map ($total_len += $_, grep (defined, @{$tree->branch_length}));
+    my $bin = int ($total_len / $binsize + .5);
+
+    $bin[$bin] = Stockholm::Database->new unless defined $bin[$bin];
+    push @{$bin[$bin]}, $stock;
+
+    warn "total_len=$total_len bin=$bin entries=",@{$bin[$bin]}+0,"\n";
+}
+
+my $suffix = $dbfile;
+$suffix =~ s/^.*?([^\/]+)$/$1/;
+
+for my $bin (0..$#bin) {
+    if (defined $bin[$bin]) {
+	my $outfile = "bin$bin.$suffix";
+	$bin[$bin]->to_file ($outfile);
+    }
+}

### 2009-04-18 rfamrectest.pl

@@ -1,0 +1,198 @@
+#!/usr/bin/perl -w
+
+
+# The following one-liner plots postprob against probability of correctness
+# (the third column is the error in the estimate of prob-of-correctness;
+#  actually it's the standard deviation of the beta distribution)
+
+# Replace "fields 4 8" by "fields 4 10" to get the postprob estimates for the naive grammar.
+# Replace "cat" by "grep pfoldF" to get stats for paired columns only.
+
+#  cat Published.xval-results | fields 4 8 | perl -e '$mul=shift;while(<>){($c,$p)=split;$bin=int($p/$mul+.5);++$c{$bin}if$c;++$t{$bin}}for$bin(sort{$a<=>$b}keys%t){$a=$c{$bin};$b=$t{$bin}-$c{$bin};$p=$a/($a+$b);$perr=sqrt($a*$b/(($a+$b)*($a+$b)*($a+$b+1)));print$bin*$mul," $p $perr\n"}' .1
+
+
+
+# The following one-liners break down incorrect basepair reconstructions
+# The first number printed is the number of incorrect reconstructions; the second, the number of these which were reconstructed as noncanonical basepairs
+
+# For the naive grammar:
+# grep "n= 0" Published.xval-results | perl -e 'while(<>){if(/pfoldF ..\/..\/(..)/){++$c{lc$1};++$t}}print$t," ",$t-$c{ua}-$c{au}-$c{cg}-$c{gc}-$c{gu}-$c{ug},"\n"'
+
+# For the main grammar:
+# grep "g= 0" Published.xval-results | perl -e 'while(<>){if(/pfoldF ..\/(..)\/../){++$c{lc$1};++$t}}print$t," ",$t-$c{ua}-$c{au}-$c{cg}-$c{gc}-$c{gu}-$c{ug},"\n"'
+
+
+# The following one-liners break down incorrect basepair reconstructions by postprob
+
[…]

### 2009-04-18 rnarectest.pl

@@ -1,0 +1,134 @@
+#!/usr/bin/perl -w
+
+
+# The following one-liner plots postprob against probability of correctness
+# (the third column is the error in the estimate of prob-of-correctness;
+#  actually it's the standard deviation of the beta distribution)
+
+# Replace "fields 4 8" by "fields 4 10" to get the postprob estimates for the naive grammar.
+# Replace "cat" by "grep pfoldF" to get stats for paired columns only.
+
+#  cat Hammerhead_1.results | fields 4 8 | perl -e '$mul=shift;while(<>){($c,$p)=split;$bin=int($p/$mul+.5);++$c{$bin}if$c;++$t{$bin}}for$bin(sort{$a<=>$b}keys%t){$a=$c{$bin};$b=$t{$bin}-$c{$bin};$p=$a/($a+$b);$perr=sqrt($a*$b/(($a+$b)*($a+$b)*($a+$b+1)));print$bin*$mul," $p $perr\n"}' .1
+
+
+
+# The following one-liners break down incorrect basepair reconstructions
+# The first number printed is the number of incorrect reconstructions; the second, the number of these which were reconstructed as noncanonical basepairs
+
+# For the naive grammar:
+# grep "n= 0" Hammerhead_1.results | perl -e 'while(<>){if(/pfoldF ..\/..\/(..)/){++$c{$1};++$t}}print$t," ",$t-$c{ua}-$c{au}-$c{cg}-$c{gc}-$c{gu}-$c{ug},"\n"'
+
+# For the main grammar:
+# grep "g= 0" Hammerhead_1.results | perl -e 'while(<>){if(/pfoldF ..\/(..)\/../){++$c{$1};++$t}}print$t," ",$t-$c{ua}-$c{au}-$c{cg}-$c{gc}-$c{gu}-$c{ug},"\n"'
+
+
+use Stockholm;
+use Newick;
+
+my $maxcols = 500;  # simulated alignments with more columns than this are junked
+
+my $gram = $ENV{DARTDIR} . "/grammars/pfold.eg";
+my $naive = $ENV{DARTDIR} . "/grammars/rev.eg";
+my $trials = 1;
+my $usage = "Usage: $0 [-g <covariant RNA grammar>] [-n <naive point-substitution grammar>] [-t trials] [-m maxcols] <alignment>\n";
+
+my @argv;
+while (@ARGV) {
+    my $arg = shift;
[…]

### 2009-04-18 simulations/compare.pl

@@ -1,0 +1,97 @@
+#!/usr/bin/perl -w
+
+use strict;
+use Stockholm;
+
+my $ref = Stockholm->from_file (shift);
+my $cmp = Stockholm->from_file (shift);
+
+@{$ref->seqname} = ("X", "Y", "Z");
+@{$cmp->seqname} = ("X", "Y", "Z");
+
+$ref->gc->{"ancestral_SS"} = $ref->gr->{"SS"}->{"root"};
+delete $ref->gr->{"SS"}->{"root"};
+
+while (my ($name, $data) = each %{$cmp->seqdata}) {
+  if ($name =~ /([XYZ])\/.+/) {
+    delete $cmp->seqdata->{$name};
+    $cmp->seqdata->{$1} = $data;
+    $cmp->gr->{"SS"}->{$1} = $cmp->gr->{"SS"}->{$name};
+    delete $cmp->gr->{"SS"}->{$name};
+  }
+}
+
+$ref->to_file ("/tmp/ref.stock");
+$cmp->to_file ("/tmp/cmp.stock");
+
+my $cmd = "cmpalign.pl /tmp/ref.stock /tmp/cmp.stock >/tmp/cmpalign.out";
+system ($cmd) == 0 or die "Couldn't run command: $cmd\n";
+
+open CMPALIGN, "</tmp/cmpalign.out" or die "Couldn't open '/tmp/cmpalign.out'\n";
+my ($acc, $sn, $ppv);
+while (<CMPALIGN>) {
+  if (/^Acc\s+(\S+)$/) { $acc = $1; }
+  if (/^Sn\s+(\S+)$/) { $sn = $1; }
+  if (/^PPV\s+(\S+)$/) { $ppv = $1; }  
+}
+close CMPALIGN;
+
+my $refanc = $ref->gc->{"ancestral_SS"};
+my $refseq = $ref->seqdata->{"root"};
+for (my $i = 0; $i < length $refseq; ++$i) {
+  if (substr ($refseq, $i, 1) ne '-') {
+    if (substr ($refanc, $i, 1) eq '-') {
+      substr ($refanc, $i, 1) = '.';
+    }
+    if (substr ($refanc, $i, 1) eq '<') {
+      substr ($refseq, $i, 1) = "C";
+    } elsif (substr ($refanc, $i, 1) eq '>') {
+      substr ($refseq, $i, 1) = "G";
+    } else  {
[…]

### 2009-04-18 simulations/generate-indiegram-subst.pl

@@ -1,0 +1,42 @@
+#!/usr/bin/perl -w
+
+use strict;
+use PhyloGram;
+
+my $g = PhyloGram->from_file ("/Users/rbradley/projects/dart/grammars/pfold.eg");
+
+my $chain = $g->find_chain ("LNUC");
+
+print $chain->terminal->to_string,"\n";
+my $mutateHash = $chain->mutate_hash;
+my @states = map (join ("",@{$_->state->value}), $chain->find_all("initial"));
+
+
+my %rates;
+foreach my $i (@states) {
+  my $total = 0;
+  foreach my $j (@states) {
+    my $m = $chain->mutate($i,$j,$mutateHash);
+    if (defined $m) {
+      $rates{$i}->{$j} = $m->rate->value;
+      $total += $rates{$i}->{$j};
+    } else {
+      warn "Setting $i -> $j = 0.\n";
+      $rates{$i}->{$j} = 0;
+    }
+  }
+  $rates{$i}->{$i} = -$total;
+}
+
+my @sortedstates = qw(aa ca ga ua ac cc gc uc ag cg gg ug au cu gu uu);
+
+print "bkstem << \"1 16\\n\"\n";
+print "<< \"a c g u A C G U b d h v B D H V\\n\"\n";
+foreach my $i (@sortedstates) {
+  print "<< \"";
+  foreach my $j (@sortedstates) {
+    print $rates{$i}->{$j}, " ";
+  }
+  print "\\n\"\n";
+}
+print "\"0\\n0\\n0\\n0\\n0\\n0\\n0\\n0\\n0\\n0\\n0\\n0\\n0\\n0\\n0\\n0\\n\"\n";

### 2009-04-18 simulations/plots.r

@@ -1,0 +1,76 @@
+
+require (graphics)
+require (CarbonEL)
+require (stats)
+
+toPDF = TRUE
+
+alignColor = "blue"
+structureColor = "red"
+
+alignShape = 16
+structureShape = 17
+
+## default plot filename
+plotfile = "simulations.pdf"
+plotdim = c (6, 4)
+
+## open PDF
+if (toPDF) {
+  pdf (file = plotfile, width = plotdim[1], height = plotdim[2])
+} else {
+  quartz (width = plotdim[1], height = plotdim[2])
+}
+
+## set up plot
+par (family = "serif")
+par (cex = 1)
+
+## read data
+data = read.table ("inferred/indiegram.analysis", header = TRUE, sep = "\t")
+
+## average datapoints for each outgroup length
+align = array()
+structure = array()
+bins = unique (data$z)
+for (bin in bins) {
+  align = rbind (
+    align,
+    c (bin, mean (subset (data$AlignAcc,
+                          data$z == bin))))
+  structure = rbind (
+    structure,
+    c (bin, mean (subset (data$StructureReconstruction,
+                          data$z == bin))))
+}
+
+## remove NA rows
+align = align[-1,]
+structure = structure[-1,]
+
+plot.new()
+plot.window (xlim = c (0.5, 3),
+             ylim = c (0.5, 1))
+axis (1,
+      c (0.5, 1, 1.5, 2, 2.5, 3))
+axis (2)
+title (xlab = "Outgroup branch length",
+       ylab = "Inference accuracy",
+       cex.lab = 1.2)
+
+points (align,
+        col = alignColor, pch = alignShape, cex = 1)
+lines (lowess (align), col = alignColor, lwd = 1.5)
+points (structure,
+        col = structureColor, pch = structureShape, cex = 1)
+lines (lowess (structure), col = structureColor, lwd = 1.5)
+
+legend ("bottomleft", legend = c ("Structural reconstruction accuracy", "Alignment accuracy"),
+        col = c (structureColor, alignColor),
+        pch = c (structureShape, alignShape),
+        xjust = 1,
+        inset = 0.02)
+
+if (toPDF) {
+  dev.off()
+}

### 2009-04-18 simulations/run-evolsayer.pl

@@ -1,0 +1,185 @@
+#!/usr/bin/perl -w
+
+use strict;
+use Stockholm;
+
+# params: chosen to agree with those hard-coded into indiegram/tkfst.cc
+my $looplen = 5;
+my $stemlen = 7/3.;
+my $pstem = 0.1;
+
+# minimum number of stems in ancestral structure
+my $ancstems_min = 2;
+my $looplength_max = 10;
+my $looplength_min = 3;
+my $stemlength_min = 0;
+my $seqlength_min = 30;
+my $seqlength_max = 70;
+
+# programs
+my $evolsayer = "$ENV{'DARTDIR'}/perl/evolsayer.pl";
+my $evolsayer_args = "-sub $ENV{'DARTDIR'}/grammars/pfold.eg -sroot -looplen $looplen -stemlen $stemlen -pstem $pstem";
+
+# tree
+my ($x, $y) = (1.0, 1.0);
+my ($z_min, $z_max) = (0, 2.5);
+my $num_bins = 25;
+my $z_increment = ($z_max - $z_min) / $num_bins;
+
+my $num_per_bin = 25;
+
+
+
+# output directory
+my $outdir = shift;
+die "Output directory not defined.\n" unless defined $outdir;
+
+# create simulations
+my $maxtries = 50000;
+my $bin = 1;
+for (my $z = $z_min; $z < $z_max; $z += $z_increment) {
+
+  # create directory to hold files for this outgroup branch length
+  my $cmd = "mkdir -p $outdir/$z";
+  if (! -d "$outdir/$z/") {
+    system ($cmd) == 0 or die "Couldn't create directory '$outdir/$z'.\n";
+  }
+
+  my $tries = 0;
+  my $cnt = 1;
+  while ($cnt <= $num_per_bin) {
+
+    # create simulated alignment
+    if (! -r "$outdir/$z/$cnt.reference.stock") {
+
+      if ($tries++ > $maxtries) {
+	die "Exceeded maximum number of tries.\n";
+      }
+      
+      my $tree = get_tree ($x, $y, $z);
+      open TREE, ">/tmp/evolsayer_tree.newick";
+      print TREE "$tree\n";
+      close TREE;
+      
+      # trick from http://perldoc.perl.org/functions/srand.html to get a constantly-changing seed
+      my $rndseed = time ^ $$ ^ unpack "%L*", `ps axww | gzip -f`;
[…]

### 2009-04-19 paper.tex
fixed typo in null cycle elimination section

@@
 Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e., rules in which $A$ appears on the left),
-and let $\rightside{B}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right).
-Let $\allrules{B} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
+and let $\rightside{A}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right, including bifurcations which also generate another nonterminal as well as $A$).
+Let $\allrules{A} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
 Define these also on sets of nonterminals, e.g., $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.

### 2009-04-20 paper.tex
further fixed up null cycle section

@@
 
-Null cycles in SCFGs constitute a ``missing data'' problem, wherein
-the unobserved values of variables contribute to probabilities of
-observed events. Such missing data problems are common in
-probabilistic models of sequences and can
-frequently be dealt with by exactly marginalizing over the
-probabilities of the unobserved variables.  For example, eliminating
-missing ``empty'' columns from an independent-sites model of sequence,
-wherein alignment columns are independent, is straightforward
-\cite{RivasEddy2008} and missing emissions in a HMM can be dealt with by matrix inversion
-\cite{Holmes2003,Holmes2007,Lunter2007} or the ``wing retraction''
-techniques used in HMMER \cite{hmmer}.
-Algorithms for exact elimination of HMM null cycles by
-matrix inversion are well-known \cite{Holmes2003} and are implemented
-by programs such as PhyloComposer \cite{Holmes2007} and
-HMMoC \cite{Lunter2007}. These algorithms effectively replace the HMM
-with an equivalent HMM that contains no null cycles but generates the
-same probability distribution over sequences.
-However, these solutions do not easily generalize to SCFGs, due to the
-possible presence of null bifurcations.
-
-Null cycles can be removed from SCFGs by (i) separating bifurcations
-into those which have one or more empty children (and can therefore be
-represented using transition or termination rules) and those that have
-two nonempty children and then (ii) encapsulating all paths that do
-not emit any terminal symbols (and may therefore cause cyclic
-dependencies in the DP matrix) as effective direct transitions, with
-probabilities obtained by geometric series summation of the grammar's
-transition matrix.
-
-Both (i) empty bifurcations and (ii) paths without terminal symbols
-generically appear in the ensemble model.
-An empty bifurcation occurs when a child branch of a bifurcation state
+The ensemble model contains empty bifurcations {\em and} empty paths.
+An {\bf empty bifurcation} occurs when a child branch of a bifurcation state
 transitions to the $\Send$ state without emitting any symbols and can
@@
 Similarly, our model construction procedure generically produces many
-``windback'' null transitions to non-emitting placeholder
-states. If $\Semit$ states $E_1$ and $E_2$ are connected in the state
+{\bf empty paths} which do not emit any terminal symbols.
+Most common are paths of ``windback'' null transitions via non-emitting placeholder states.
+If $\Semit$ states $E_1$ and $E_2$ are connected in the state
 graph by $\Snull$ states $N_1$ and $N_2$, then the transition $E_1
@@
 % increase in runtime.
+
+We refer to empty paths and empty bifurcations as {\bf null cycles}.
+They occur (in simpler form) in Hidden Markov Models (which are special cases of SCFGs)
+and independent-sites models (which can be viewed as special cases of HMMs).
+In all such models, null cycles represent events which are allowed by the model, but could never be observed.
+They can be accounted for by exact marginalization.
+For example, eliminating missing ``empty'' columns from an independent-sites model of sequence involves a straightforward application of a correction factor
+to account for the proportion of columns that are unobserved \cite{RivasEddy2008}.
+The slightly more complicated situation of missing emissions in a HMM can be dealt with by
+summing over all null paths, yielding a geometric series that is solvable by matrix inversion
+\cite{Holmes2003,Holmes2007,Lunter2007}; the ``wing retraction'' algorithms used in HMMER can be viewed as a special case of matrix inversion by iterated elimination \cite{hmmer}.
+
+Algorithms for exact elimination of HMM null cycles by matrix inversion are implemented
+by programs such as PhyloComposer \cite{Holmes2003,Holmes2007} and
+HMMoC \cite{Lunter2007}. These algorithms effectively replace the HMM
+with an equivalent HMM that contains no null cycles but generates the
+same probability distribution over sequences.
+However, these solutions do not easily generalize to SCFGs, due to the
+possible presence of null bifurcations.
+
+Here, we show how null cycles can be removed systematically from SCFGs by two steps:
+\begin{description}
+\item[(i)] separating bifurcations
+into those which have one or more empty children (and can therefore be
+represented using transition or termination rules) and those that have
+two nonempty children;
+\item[(ii)] replacing all empty paths through null states with effective direct transitions between non-null states,
+obtaining sum-over-paths probabilities by inverting the grammar's transition matrix.
+\end{description}
 

### 2009-04-20 paper.tex
moved Ortheus placeholder comment
further fixed up null cycle section

@@
 and independent-sites models (which can be viewed as special cases of HMMs).
-In all such models, null cycles represent events which are allowed by the model, but could never be observed.
-They can be accounted for by exact marginalization.
+In all such models, null cycles represent histories that are valid according to the model, but cannot be resolved by direct observation.
+Such null events can be real (e.g. ancestral residues that have been deleted in all extant lineages)
+or they can be artefactual (e.g. transitions between placeholder null states of an HMM).
+Often, one can account for the likelihood of such events by exact marginalization.
 For example, eliminating missing ``empty'' columns from an independent-sites model of sequence involves a straightforward application of a correction factor
@@
 
-% to do: discuss the possibility of progressive reconstruction a la Ortheus
-
 An implementation of inference algorithms for models on the three-taxon phylogeny
@@
 of RNAs on a tree.
+
+% to do: discuss the possibility of progressive reconstruction a la Ortheus
+% "Another heuristic that could be tried..."
 

### 2009-05-19 paper.tex
Finalizing edits...

@@
 \end{flushleft}
+
+% figures in figs_plos:
+% 1: parent.pdf
+% 2: threeway.pdf
+% 3: stree-with-structure.pdf
+% 4: stree-mutations-with-structures.pdf
+% 5: simulations.pdf
 
@@
 sibling species constant and varied the branch length of the
-outgroup.  We fixed the tree topology as $((X:1,Y:1):1,Z:z)$, where
-the branch length $z$ of the outgroup was varied between 0.5 and 3.0
-and a branch length of 1 corresponds to one expected substitution per site
-in loop sequence.  We chose the parameters of the TKFST model to
+outgroup.  We chose the parameters of the TKFST model to
 generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
@@
 
-We expected to find a ``transition point'' in our reconstruction
-accuracy near the point $z \simeq 1$, corresponding to an outgroup
-distance equal to the distance from either of the sibling species to
-the most recent common ancestor.  Our results, shown in
-\figref{simulations}, indicate such a transition point for alignment
-accuracy but not for ancestral reconstruction accuracy.
-% to do: here now
-
-\cite{Mossel2003}
+The results of our simulations, illustrated in \figref{simulations},
+demonstrate that the accuracy of both structural
+reconstruction and alignment accuracy depend on the outgroup
+branch length.  However, while the difficulty of alignment inference
+increases in proportion to the outgroup branch length, becoming
+near-impossible for very long outgroup branch lengths, the same need not
+true for structural reconstruction accuracy.  As the outgroup branch
+length increases, its structure becomes increasingly decoupled from
+that of its ancestor, allowing the algorithm to ``ignore'' the
+information which it contains.
+Similarly, while we can perfectly reconstruct the ancestral
+structure given an outgroup branch length of 0 (the trivial case), we
+cannot perfectly reconstruct the alignment.
 
@@
 demonstrating that inference of sequence alignment is surprisingly
-robust to mis-parametrized models \cite{Lunter2007} suggest that the same
-may be true for inference of ancestral structures.
+robust to mis-parametrized probabilistic models \cite{Lunter2007}
+suggest that the same may be true for probabilistic inference of
+ancestral structures.
 
@@
 such algorithms will only be feasible for short RNAs (\tabref{complexities}).
-Thus, alternate strategies will have to be explored.
-One promising direction is to consider variations on the three-way sampling kernel,
+Thus, alternate strategies will have to be explored, such as the
+progressive reconstruction used by the \ortheus\ program for
+reconstructing ancestral genomes \cite{PatenEtAl2008}.
+Another promising direction is to consider variations on the three-way sampling kernel,
 such as the importance-sampling approach described for the TKF model
@@
       length.}
-    We allowed the outgroup branch length $z$ to vary within $[0.5,
-    3.0]$, where a branch length of 1 corresponds to one expected
-    substitution per site in loop sequence.  We divided this interval
+    We investigated the dependence of structural reconstruction and
+    alignment accuracy on outgroup branch length by simulating the
+    evolution of three structural RNAs under our model.  The two
+    sister species were equidistant from the ancestral sequence, with
+    a branch length of 1, corresponding to one expected
+    substitution per site in loop sequence.  We allowed the outgroup
+    branch length to vary between $[0, 2.5]$.  We divided this interval
     into 25 bins and simulated 25 alignments for each bin.
-    Parameters used were:
+    Parameters used were $\lambda_1 = 0.025$ and $\mu_1 = 0.03$ for
+    loop sequence and $\lambda_2 = 0.007$ and $\mu_2 = 0.01$ for stem
+    sequence; the probability of a stem insertion was $0.1$.  These
+    yielded a mean loop length of 5 and a mean stem length of 2.33,
+    with $\sim 0.3$ substructure indels per alignment.
+    We selected challenging alignments to reconstruct by requiring
+    that there be at least two ancestral stems, loops of length $\in [3,
+    10]$ and stems of length $\in [1, 20]$; to reduce the complexity of
+    our algorithms we additionally required that the sequences have
+    lengths $\in [30, 70]$.
  }

### 2009-05-19 response-to-reviews2.txt
responses

@@
 
-We have accordingly changed the title to "Evolutionary models of
+We have accordingly changed the title to "Evolutionary triplet models of
 structured RNA".
@@
 
-[we have deleted the references to Hopfield's hypothesis]
+We have removed the references to Hopfield's hypothesis from the revised manuscript.
 
@@
 > to the actual model and problems at hand.
+
+We have reworded this section per our new focus on the simulation results.
 
@@
 
+5' and 3' ends of structures are now labeled (e.g., Figures 3 and 4 of
+the revised manuscript).
+
 > Figure 3: It is very unclear what this figure is supposed to be 
@@
 
-[we have included stree-mutations.pdf, which was also in the cited paper, and illustrates the dynamic behavior of this model]
+We have updated this figure and included an additional figure (also
+extended from a figure in [28]) which illustrates the requested
+mutations.  Both figures now include the structures for clarity.
 
@@
 
-[we have corrected this..... I hope]
+We have attempted to correct these mistakes; we apologize in advance
+for any which remain in the revised manuscript.

### 2009-05-20 paper.tex
Edits.

@@
 pairwise statistical models of RNA structural evolution to models of
-multiple sequences, related by a phylogenetic tree.  This algorithm
+multiple sequences related by a phylogenetic tree.  This algorithm
 draws on formal models of computational linguistics as well as the
@@
 evolution, we simulated data under the model and tested the dependence
-of alignment inference and reconstruction accuracy on the evolutionary
+of alignment inference and ancestral structure reconstruction accuracy on the evolutionary
 distance of the outgroup of the phylogeny.  Our results suggest that
@@
 \section*{Author Summary}
+% to do: write this
 
@@
 A domain structure for this primeval ribosome was recently proposed \cite{SmithEtAl2008}.
-To synthesize such a reconstructed ribosome, or reconstructions of other evolutionarily significant RNAs
+To synthesize such a reconstructed ribosome or reconstructions of other evolutionarily significant RNAs
 such as group II introns \cite{LehmannSchmidt2003} or telomerase \cite{AntalEtAl2002},
-it will be necessary to develop methods that can predict the sequences of ancient RNAs
+it will be necessary to develop methods that can predict the sequences
+and structures of ancient RNAs
 based on the divergent sequences of their many descendants.
@@
 % SCFGs for RNA
-Inspection of RNA alignments, for example in the \RFAM\ database
+An inspection of RNA alignments, such as those in the \RFAM\ database
 \cite{GriffithsJonesEtAl2003}, suggests that an evolutionary model for
@@
 for example, another representation of answers A-C
-is that the ``pruning'' \& ``peeling'' algorithms (combined) are just
+is that the ``pruning'' and ``peeling'' algorithms (combined) are just
 the sum-product algorithm on a directed graphical model \cite{Pearl82},
@@
 simultaneously align and predict the structure of
-related RNAs.  The focus of the present work is to solve the second
+pairs of related RNAs.  The focus of the present work is to solve the second
 (multiple-sequence) part of the RNA reconstruction problem by giving a
@@
 The main contributions of this paper are
-(1) an algorithm that transforms a phylogenetic ensemble of pair grammars
+(1) an algorithm that transforms a phylogenetic ensemble of pair
+grammars, representing models on branches of a phylogenetic tree,
 into a coherent, multiple-sequence SCFG,
@@
 preventing us from applying our method to many problems of interest,
-we assess the conduct a simulation study to investigate the accuracy
-with which we can hope to reconstruct ancestral structures over long
-evolutionary time in the presence of structural divergence.
+a simulation study suggests that we can hope to accurately reconstruct
+ancestral structures over long evolutionary time, even in the presence
+of structural divergence.
 
@@
 Similarly, we might consider only ancestral structures which are
-compatible with a curated multiple alignment.  Such constraints are
+compatible with a curated multiple sequence alignment.  Such constraints are
 commonly used by programs for RNA sequence analysis \cite{RivasEddy2001,Holmes2005}.
 
-The following sections introduce more precise definitions for two-sequence
-models of RNA structure and proceed to outline our algorithms for
+In the following sections we introduce more precise definitions for two-sequence
+models of RNA structure and outline our algorithms for
 (1) combining these two-sequence models on a phylogenetic tree and
@@
 sites (as studied in likelihood phylogenetics) and generalizing to models of
-indels first in primary sequences, then in sequences with conserved secondary structure.
+indels, first in primary sequences and then in sequences with
+conserved secondary structure.
 
@@
 a probabilistic model of single sequences under the same approximation.
-The conditional distribution $P(Y | X, T)$ then corresponds to a conditional Pair HMM:
+The conditional distribution $P(Y | X, T)$ then corresponds to a conditional Pair HMM,
 a finite-state machine which transforms one sequence (the input, $X$) into another (the output, $Y$).
@@
 We extend this formalism to the case of structured RNA as follows.
-Let $X$ and $Y$ now represent structured RNA sequences, or more
+Let $X$ and $Y$ now represent structured RNA sequences or, more
 precisely, parse trees.
 A single-sequence SCFG models the marginal $P(X)$;
-the joint distribution $P (X, Y | T)$ is modeled by a jointly-normalized Pair SCFG \cite{Durbin98}.
+a jointly-normalized Pair SCFG \cite{Durbin98} models the the joint distribution $P (X, Y | T)$.
 The conditional distribution $P(Y | X, T)$ is modeled by a conditionally-normalized Pair SCFG.
@@
 
-The stochastic state machine which generates parse-trees from the marginal distribution $P(X)$ is a {\bf singlet transducer}.
-The machine that absorbs parse-trees $X$ and generates modified parse-trees $Y$ from the conditional distribution $P(Y | X, T)$ is a {\bf branch transducer}.
+The stochastic state machine which generates parse trees from the marginal distribution $P(X)$ is a {\bf singlet transducer}.
+The machine that absorbs parse trees $X$ and generates modified parse trees $Y$ from the conditional distribution $P(Y | X, T)$ is a {\bf branch transducer}.
 Singlet transducers only emit symbols to their output sequence;
@@
 Two types of states can emit and/or absorb symbols, $\Sinsert$ and $\Smatch$.
-An $\Sinsert$ state emits a symbol to the output, without absorbing anything.
+An $\Sinsert$ state emits a symbol to the output without absorbing anything.
 A $\Smatch$ state absorbs a symbol on the input and either emits
 the same symbol to the output, substitutes a different output symbol,
-or emits no output symbol at all (the latter corresponds to a deletion).
+or emits no output symbol at all, the last corresponding to a deletion.
 
@@
 extending the pairwise model to multiple sequences.
-The transformation rules are partitioned into co-normalized groups:
+The transformation rules are partitioned into co-normalized groups;
 within each group, the rule probabilities must sum to one.
@@
 (i.e., all outgoing transitions from a particular state).
-In a conditionally-normalized Pair SCFG, the groups are finer-grained:
-each co-normalized group includes all rules that can be applied to a given nonterminal, {\em for a given set of absorbed symbols}.
+In a conditionally-normalized Pair SCFG, in contrast,
+each co-normalized group includes all rules that can be applied to a
+given nonterminal {\em for a given set of absorbed symbols}.
 %(In practice it is often expedient to relax these normalization guidelines somewhat,
@@
 framework to model the common descent of many homologous sequences.
-Given a phylogenetic tree and two-sequence model, we wish to obtain a
+Given a phylogenetic tree and a two-sequence model, we wish to obtain a
 multiple-sequence SCFG describing the common descent of the observed sequences.
@@
 
-Throughout this paper we make frequently refer to two and
+Throughout this paper we frequently refer to two and
 three-taxon (star) phylogenies.  In all cases, the sequence $W$ is
@@
 individual transducer state spaces. If the singlet transducer has $a$
-states, each branch transducer has $b$ states, and there are $N$
-branches in the phylogeny, then an upper bound on the number of
+states, each branch transducer has $b$ states, and the phylogeny has $N$
+branches, then an upper bound on the number of
 ensemble states is $O(a \cdot b^N)$.
@@
 in practice models never reach this bound,
-due both to inaccessible configurations (such as the one described above)
+due both to inaccessible configurations, such as the one described above,
 and the sparseness of transitions between the remaining, accessible configurations.
@@
   that will be absorbed by {\em their} descendant transducers.
-  This continues down the tree: The terminal symbols are passed from parents, to children, to grandchildren
-  (albeit possibly being replaced by other terminal symbols as they're propagated down)
+  This continues down the tree: The terminal symbols are passed from parents to children to grandchildren
+  (albeit possibly being replaced by other terminal symbols as they
+  are propagated down)
   and they propel branch transducers into $\Smatch$ states as they go.
-  Eventually, the cascade of emitted terminal symbols stops when all the symbols have been deleted,
+  Eventually, the cascade of emitted terminal symbols stops when all the symbols have been deleted
   or when the cascade reaches the leaves of the tree.
@@
   terminal emission (conceptually, a bifurcation is a ``nonterminal emission'').
-  As with terminal emissions, absorption of nonterminal emissions propel
+  As with terminal emissions, absorption of nonterminal emissions propels
   descendant transducers into $\Smatch$ states,
@@
   into an ancestral RNA structure,
-  which then may be conserved or deleted in the descendant structures.
+  which may then be conserved or deleted in the descendant structures.
 \item {\bf End Transition}:
@@
 Co-ordination between the various branch machines is achieved by
-specifying an ordering on the nodes (preorder with respect to the
-phylogenetic tree) and by having branch transducers ``pause'' in $\Swait$
+specifying an ordering on the nodes and by having branch transducers ``pause'' in $\Swait$
 states while waiting to absorb a symbol from the node above.  
@@
 
-The four types of allowed transitions listed above can be formalized simply.
+The four types of allowed transitions listed above can be formalized
+as follows.
 Let the total order on the nodes correspond to a preorder traversal of
@@
 branches, $O(a \cdot b \cdot N)$, and the number of transitions is
-approximately linear in the number of states, for several pairwise models,
-including the pairwise model we use here.
+approximately linear in the number of states for several pairwise models,
+including the pairwise model which we use here.
 However, these empirical observations are based on a
@@
 Therefore, for the models which we have characterized, the search
-algorithm given below for enumerating all allowed transitions of the
+algorithm given above for enumerating all allowed transitions of the
 ensemble model typically generates $O(b)$ transitions from any given
@@
 jointly-normalized multiple-sequence SCFG (modeling the extant sequences).
-We can then use the Cocke-Younger-Kasami (CYK) algorithm to impute the
+We can then use the Cocke-Younger-Kasami (CYK) algorithm to infer the
 maximum-likelihood alignment, secondary structure and common ancestor
-of the extant sequences, or the Inside algorithm to sample from
+of the extant sequences or the Inside algorithm to sample from
 the joint posterior distribution of these unknowns.
@@
 SCFG. The SCFG can contain null cycles, corresponding to generation of
-these unobserved fragments, the existence of which complicates the
-inference algorithms.
+these unobserved fragments, which complicates inference.
 
@@
 the standard CYK and Inside algorithms fail on the ensemble model.
-Residues in the ancestor can be absent in all extant sequences;
-the potential number of such unobserved ancestral residues is unbounded.
+A potentially-unbounded number of residues in the ancestor can be
+absent in all extant sequences.
 This manifests as cycles of $\Snull$ states within the multiple-sequence SCFG,
@@
 In order to correctly perform maximum-likelihood inference, we only
-require that the lowest-probability parse corresponds to the parse
+require that the maximum-likelihood parse tree corresponds to the parse
 returned by the CYK algorithm; the entire
@@
 create new versions of the standard inference algorithms for SCFGs
-which can cope with SCFGs with null cycles, or we can modify the
+which can cope with SCFGs with null cycles or we can modify the
 structure of the SCFGs generated by our model-construction algorithm
@@
 the probabilities for unobserved null cycles, until the approximation
-is sufficiently good---but it is difficult to implement efficiently,
+is sufficiently good---but it is difficult to implement efficiently
 and necessarily has worse time complexity than the standard inference
@@
 
-We therefore developed an iterative approximation to remove null cycles from
-the ensemble model in order to use the standard inference algorithms
+We therefore developed a method to remove null cycles from
+the ensemble model, described below, in order to use the standard inference algorithms
 for SCFGs without incurring the further computational complexity
@@
 and independent-sites models (which can be viewed as special cases of HMMs).
-In all such models, null cycles represent histories that are valid according to the model, but cannot be resolved by direct observation.
-Such null events can be real (e.g. ancestral residues that have been deleted in all extant lineages)
-or they can be artefactual (e.g. transitions between placeholder null states of an HMM).
+In all such models, null cycles represent histories that are valid
+according to the model but cannot be resolved by direct observation.
+Such null events can be real (e.g., ancestral residues that have been deleted in all extant lineages)
+or they can be artefactual (e.g., transitions between placeholder null states of an HMM).
 Often, one can account for the likelihood of such events by exact marginalization.
@@
 constraints to reduce the accessible volume of the DP matrix \cite{Holmes2005}.
-Algorithms \ref{alg:inside}--\ref{alg:cyktraceback} give the Inside, CYK,
+Algorithms \ref{alg:inside}--\ref{alg:cyktraceback}, described below, give the Inside, CYK,
 Outside and CYK traceback algorithms for a three-taxon star phylogeny,
-constrained using the ``fold envelopes'' described below.
+constrained using the ``fold envelope'' concept.
 
@@
 \end{align*}
-where the quantities $\Delta^{(X)}_L(\bvec{b})$ and $\Delta^{(X)}_R(\bvec{b})$ are the lengths of the left and right emissions of the ensemble state $b$ to the sequence $X$.
+where the quantities $\Delta^{(X)}_L(\bvec{b})$ and $\Delta^{(X)}_R(\bvec{b})$ are the lengths of the left and right emissions of the ensemble state $\bvec{b}$ to the sequence $X$.
 (Recall that the $\mth$ subsequence in $\foldenv^{(X)}$ is labeled $m^{(X)}$ and corresponds to the coordinate pair $(i_{m}, j_{m})$.)
@@
 \end{align}
-We generally write out explicit subsequence coordinate pairs $(i,j)$ when their usage will make mathematical formulae clearer and fold-envelope labels $n^{(X)}$ when writing pseudocode.
+We generally write out explicit subsequence coordinate pairs $(i,j)$
+when their usage will make mathematical formulas clearer and
+fold-envelope labels $n^{(X)}$ when writing pseudocode.
 
@@
 The inside probability $\alpha_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)}\right)$ is the summed probability of the triplet of subsequences 
-$\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ for sequences $X, Y, Z$.
+$\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ for sequences $X, Y, Z$
 under all paths through the model which are rooted in state $\bvec{a}$.  
@@
 with the $\max$ operation.
-The CYK probabilities for indices $\gamma_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)}\right)$ then represent 
+The CYK probability for indices $\gamma_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)}\right)$ then represents
 the probability of the most likely path through the model generating 
@@
 mutually nested (\figref{tkfst}).  Both loops and stems have geometric length
-distributions with parameters $\kappa_1$ and $\kappa_2$.  Single bases
-are inserted and deleted in loops with rates $\lambda_1$ and $\mu_1$;
+distributions with parameters $\kappa_l$ and $\kappa_s$.  Single bases
+are inserted and deleted in loops with rates $\lambda_l$ and $\mu_l$;
 similarly, base-pairs are inserted and deleted in stems with rates
-$\lambda_2$ and $\mu_2$.  Insertions of a new stem into an existing
+$\lambda_s$ and $\mu_s$.  Insertions of a new stem into an existing
 loop sequence (or deletions of an existing stem) occur at the same
@@
 using estimates reported by a phylo-grammar for RNA secondary
-structure prediction, \pfold\ \cite{KnudsenHein2003}, and an
+structure prediction, \pfold\ \cite{KnudsenHein2003} and an
 implementation of pairwise alignment for the TKF Structure Tree model,
@@
 expressed in terms of functions of the evolutionary time along a branch
-and the insertion and deletion rates $\lambda_1$ and $\mu_1$ of the model.
-The length of ancestral sequences is geometric in $\kappa_1$ (\tabref{tkfstsinglet}), defined as
-\[
-\kappa_1 = \lambda_1 / \mu_1 \, .
-\]
+and the insertion and deletion rates $\lambda_l$ and $\mu_l$ of the model.
+The length of ancestral sequences is geometric in $\kappa_l$ (\tabref{tkfstsinglet}), defined as $\kappa_l = \lambda_l / \mu_l$.
 The three functions $\alpha(t)$, $\beta(t)$ and $\gamma(t)$ which govern 
@@
 \begin{align}
-  \alpha_1 (t) &= \exp \left( -\mu_1 t \right) \nonumber \\
-  \beta_1 (t) &= \frac{\lambda_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) } \nonumber \\
-  \gamma_1 (t) &= 1 - \frac{\mu_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\left( 1 - \exp (- \mu_1 t) \right) \left(\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) \right) } \nonumber
+  \alpha_l (t) &= \exp \left( -\mu_l t \right) \nonumber \\
+  \beta_l (t) &= \frac{\lambda_l \left( 1 - \exp \left((\lambda_l - \mu_l) t \right) \right)}{\mu_l - \lambda_l \exp \left( (\lambda_l - \mu_l) t \right) } \nonumber \\
+  \gamma_l (t) &= 1 - \frac{\mu_l \left( 1 - \exp \left((\lambda_l - \mu_l) t \right) \right)}{\left( 1 - \exp (- \mu_l t) \right) \left(\mu_l - \lambda_l \exp \left( (\lambda_l - \mu_l) t \right) \right) } \nonumber
 \end{align}
@@
 
-The TKFST model, like the original TKF model, probably needs refinements in order to be useful.
-For example, it fails to model certain phenomena observed in natural
+The TKFST model, like the original TKF model, probably needs
+refinements in order to accurately model many structural RNAs.  For
+example, it fails to model certain phenomena observed in natural
 RNA structures (such as base-stacking or tetraloops) and in alignments
@@
 The richer grammar, as described in \cite{Holmes2005},
-is richer than TKFST:
+is much more complex than TKFST:
 excluding the substitution model, it has 14 free parameters
@@
 should be sufficient for ergodic sampling of reconstructions on any larger phylogeny,
-using an MCMC kernel similar to that of \cite{JensenHein2002}.
+using a MCMC kernel similar to that of \cite{JensenHein2002}.
 
@@
 The naive exponential upper-bound gives a maximal state space of size $O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.
-Using our uninformed search algorithm, we determine that there are 287 accessible states
+Using our uninformed search algorithm, we determined that there are 287 accessible states
 and 686 possible transitions between these states (compare with the $287^2 \simeq 8 \cdot 10^4$ transitions
@@
 $O(a \cdot b^2 \cdot N) = O(7 \cdot 21^2 \cdot 3) \Rightarrow 9,261$ transitions
-given in ``The composition algorithm.''
+suggested in ``The composition algorithm.''
 
 % to do: dump output of tripletscfg.pl --tkfst into supplement
-Text S? contains the complete model specification, including states and transitions,
+Text S1 contains the complete model specification, including states and transitions,
 produced by our model composition algorithm.  The extreme complexity of the
@@
 We implemented constrained maximum-likelihood inference of the structural alignment 
-and ancestral structure of three extant sequences (Algorithms \ref{alg:cyk}-\ref{alg:cyktraceback})
-in a C++ program (\indiegram).
+and ancestral structure of three extant sequences in a C++ program (\indiegram).
 For tractability, \indiegram\ uses the concept of fold envelopes
@@
 structure of extant sequences.  While experimentally-determined
-structures of individual RNAs are rare, curated deep sequence
+structures of individual RNAs are relatively rare, curated deep sequence
 alignments, such as those constructed for ribosomal RNAs
-\cite{Gutell93}, are more frequently available.  By constraining the
+\cite{Gutell93}, are frequently available for characterized RNA families.  By constraining the
 inference algorithms with such sequence alignments, the memory and
@@
 interest will soon be feasible.
-We used the TKF Structure Tree model on a 3-taxon phylogeny to 
-reconstruct most-probable ancestral structures of three \emph{nanos} 3'
-translational control elements (TCEs) and three tRNAs, two small RNAs whose
-structures are well-known and which are accessible given
-current computational limitations. While the phylogenetic trees
-relating the tested sequences have short branch lengths, making the
-reconstruction problem easy by forcing the reconstructed structures to
-be essentially-identical to those of one of the extant RNAs
-(reconstructed structures not shown; see \indiegramurl), these
-are nonetheless useful proof-of-concept datasets.
-
 \tabref{complexities} shows estimates of the memory and time required
-to reconstruct biologically-interesting subunits of the two RNAs
-tested, the \emph{nanos} 3' translational control element and tRNAs,
+to reconstruct biologically-interesting subunits of 
+the \emph{nanos} 3' translational control element and tRNAs,
 as well as two small RNAs which show significant structural
@@
 interesting candidates for ancestral reconstruction.
+The reconstructed structures for three \emph{nanos} 3'
+translational control elements (TCEs) and three tRNAs, which could be
+analyzed given current computational limitations, can be found at
+\indiegramurl; however, the phylogenetic trees relating the tested
+sequences have short branch lengths, making the reconstruction problem easy by forcing the reconstructed structures to
+be essentially-identical to those of one of the extant RNAs.
 
@@
 alignment) and considered only alignments with at least two ancestral
-stems, sequences of length $\in [30, 70]$, loops of length $\in [3, 10]$ and
-stems of lengths $\in [1, 20]$.
+stems, sequences of length $\in [30, 70]$ bp, loops of length $\in [3,
+10]$ bp and stems of lengths $\in [1, 20]$ bp.
 
@@
 increases in proportion to the outgroup branch length, becoming
-near-impossible for very long outgroup branch lengths, the same need not
-true for structural reconstruction accuracy.  As the outgroup branch
-length increases, its structure becomes increasingly decoupled from
+near-impossible for very long outgroup branch lengths, the same need
+not be true for structural reconstruction accuracy.  As the branch
+length of the outgroup increases, the outgroup's structure becomes increasingly decoupled from
 that of its ancestor, allowing the algorithm to ``ignore'' the
@@
 may be found in \cite{KleinEddy2003}.
-A conditional framework similar to ours in some respects is described by Sakakibara {\em et al} \cite{Sakakibara2003}.
+A conditional framework similar to ours in some respects is described
+by Sakakibara et al \cite{Sakakibara2003}.
 The statistical inference algorithms for multiple-sequence SCFGs are closely related to the ``protosequence algorithm'' of Sankoff \cite{Sankoff85},
@@
 our model-construction algorithm is applicable to any model of the evolution of secondary
-structure which can be expressed as a parse-tree transducer.
+structure which can be expressed as a Pair SCFG.
 Realistic structural and thermodynamic effects---such as base-stacking
@@
 An implementation of inference algorithms for models on the three-taxon phylogeny
-is sufficient to construct an MCMC sampling algorithm
+is sufficient to construct a MCMC sampling algorithm
 over many sequences on an arbitrary phylogeny.
@@
 
-% to do: discuss the possibility of progressive reconstruction a la Ortheus
-% "Another heuristic that could be tried..."
-
 The evolutionary models we have described form a theoretical
@@
     An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
-     A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
+     A singlet transducer (the horizontal gray box) emits an ancestral sequence and structure
      and two branch transducers (the gray boxes labeled $T_X$ and $T_Y$)
@@
      An ancestral sequence $W$ evolves into three descendant sequences $X$, $Y$ and $Z$.
-     A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
+     A singlet transducer (the horizontal gray box) emits an ancestral sequence and structure
      and three branch transducers (the gray boxes labeled $T_X$, $T_Y$ and $T_Z$)
@@
      from which it follows that evaluating the likelihood function
-     on this star phylogeny is sufficient for a sampling algorithm on an arbitrary phylogeny.
+     on this star phylogeny is sufficient, in principle, for a sampling algorithm on an arbitrary phylogeny.
    }
@@
     into 25 bins and simulated 25 alignments for each bin.
-    Parameters used were $\lambda_1 = 0.025$ and $\mu_1 = 0.03$ for
-    loop sequence and $\lambda_2 = 0.007$ and $\mu_2 = 0.01$ for stem
+    Parameters used were $\lambda_l = 0.025$ and $\mu_l = 0.03$ for
+    loop sequence and $\lambda_l = 0.007$ and $\mu_l = 0.01$ for stem
     sequence; the probability of a stem insertion was $0.1$.  These
-    yielded a mean loop length of 5 and a mean stem length of 2.33,
+    yielded a mean loop length of 5 bp and a mean stem length of 2.33 bp,
     with $\sim 0.3$ substructure indels per alignment.
@@
     that there be at least two ancestral stems, loops of length $\in [3,
-    10]$ and stems of length $\in [1, 20]$; to reduce the complexity of
+    10]$ bp and stems of length $\in [1, 20]$ bp; to reduce the complexity of
     our algorithms we additionally required that the sequences have
-    lengths $\in [30, 70]$.
+    lengths $\in [30, 70]$ bp.
+    ``Structural reconstruction accuracy'' is the fraction of
+    base-pairs in the ancestral structure present in the
+    maximum-likelihood reconstruction thereof; ``Alignment accuracy''
+    is a measure of sequence alignment accuracy which takes both
+    aligned and unaligned (indels) nucleotides into account,
+    qualitatively corresponding to giving equal weight to sensitivity
+    and specificity (see
+    \cite{SchwartzMyersPachter2006} for a precise definition).
 }
@@
     & $|$ & $W_L$ & $1-\gamma_l(t)$ & & & &  \\ \hline
-    $W_L$ & $\rightarrow$ & $u \,\, w \,\, M_L$ & $\alpha_l(t) \cdot M_l(u,w)$ & $W_S$ & $\rightarrow$ & $u \,\, w \,\, M_S \,\, x \,\, v$ & $\alpha_s(t) \cdot M_s(uv,wx)$ \\ \hline
-    & $|$ & $u \,\, D_L$ & $1-\alpha_l(t)$ & & $|$ & $u \,\, D_S \,\, v$ & $1-\alpha_s(t)$ \\ \hline
+    $W_L$ & $\rightarrow$ & $w \,\, M_L$ & $\alpha_l(t) \cdot M_l(u \rightarrow w)$ & $W_S$ & $\rightarrow$ & $w \,\, M_S \,\, x$ & $\alpha_s(t) \cdot M_s(uv \rightarrow wx)$ \\ \hline
+    & $|$ & $D_L$ & $1-\alpha_l(t)$ & & $|$ & $D_S$ & $1-\alpha_s(t)$ \\ \hline
     & $|$ & $B$ & $\alpha_l(t)$ & & $|$ & $B_e$ & $1$ \\ \hline
@@
     nucleotide $w$ and descendant paired nucleotides $(w,x)$;
-    $M_l(u,w)$ and $M_l(uv,wx)$ are the conditional distributions
-    (i.e. match probabilities) of, respectively, descendant unpaired
+    $M_l(u \rightarrow w)$ and $M_l(uv \rightarrow wx)$ are the conditional distributions
+    (i.e., match probabilities) of a descendant unpaired
     nucleotide $w$ given an ancestral unpaired nucleotide $u$ and
     descendant paired nucleotides $(w,x)$ given ancestral nucleotides
-    $(u,v)$. The functions $\alpha_n(t)$, $\beta_n(t)$ and
-    $\gamma_n(t)$ are parametrized by the insertion and deletion rates
+    $(u,v)$. The functions $\alpha_{l,s}(t)$, $\beta_{l,s}(t)$ and
+    $\gamma_{l,s}(t)$ are parametrized by the insertion and deletion rates
     of the TKFST model and are defined in ``A simple model of RNA
@@
   These statistics are summed over all pairs of sequences in the multiple alignment;
-  therefore, ``Sensitivity'' for pairwise residue alignments is equivalent to the Sum of Pairs Score, or SPS \cite{ThompsonEtal99}.
+  therefore, ``Sensitivity'' for pairwise residue alignments is equivalent to the Sum of Pairs Score or SPS \cite{ThompsonEtal99}.
   ``g2intron'' is the RFAM entry {\tt Intron\_gpII}, containing domains V and VI of the Group II intron.
@@
     \cite{GardnerEtAl2005} (identifiers AB042432.1-14140\_14072,
-    Z82044.1-16031\_16103 and AC008670.6-83725\_83795) The Group II
+    Z82044.1-16031\_16103 and AC008670.6-83725\_83795). The group II
     intron sequences (identifiers Z00044.1-87253\_87177,

### 2009-05-21 response-to-reviews2.txt
Final edits.

@@
 incorporated into our inference algorithms with the alignment
-constraints used by e.g. QRNA and PFOLD, or the simultaneous fold and
-alignment constraints used by e.g. Stemloc and CONSAN.
+constraints used by, e.g., QRNA and PFOLD, or the simultaneous fold and
+alignment constraints used by, e.g., Stemloc and CONSAN.
 
@@
 Inside, Outside, and CYK traceback algorithms for three sequences,
-whether completely unconstrained, or constrained by a pre-determined
-set of folds for each sequence ("fold envelopes").  They are written
-in pseudocode which corresponds closely to our actual implementation
-in the program (INDIEGRAM) which we describe.
+constrained by a pre-determined set of folds for each sequence ("fold
+envelopes").  They are written in pseudocode which corresponds closely
+to our actual implementation in the program (INDIEGRAM) which we
+describe.
 
@@
 We thank the reviewer for pointing out errors in the original
-figures. These arose during manual conversion of indiegram's output
-parse trees into Stockholm-format alignments, and due to a bug in
-xrate's reconstruction code. We have now automated this process
-(i.e. indiegram now outputs Stockholm format) and fixed the bug in
-xrate, verifying that this has eliminated the errors (e.g. see the
-simulation results).
+figures. These were due to mistakes made during manual conversion of
+indiegram's output parse trees into Stockholm-format alignments and a
+bug in the code used for maximum-likelihood reconstruction of
+sequences (the xrate program). We have now automated this process and
+fixed the bug, verifying that this has eliminated the errors.
 
 We removed the tRNA and nanos reconstructions and posted (corrected)
-versions online, at the following URL (which is linked from the
-indiegram webpage):
+versions online at the following URL (which is linked from the
+indiegram webpage),
+http://biowiki.org/IndiegramExamples,
+although we note in the text that the reconstructions were too easy
+because of short branch lengths.
 
-http://biowiki.org/IndiegramExamples
-
-We hope that these may still be useful for illustrative purposes for
-anyone trying to use the software, but we have removed references from
-the paper (since, as the reviewer notes, the short branch lengths make
-the answers somewhat obvious).
-
-The paper still includes the (experimentally-based) runtime estimates for
-different length classes of RNA genes.
+The paper still includes the runtime estimates for different length
+classes of RNA genes.
 
@@
 We have accordingly changed the title to "Evolutionary triplet models of
-structured RNA".
+structured RNA"; we welcome title suggestions from the reviewer!
 
@@
 
-We have removed the references to Hopfield's hypothesis from the revised manuscript.
+We have removed the references to Hopfield's hypothesis from the
+revised manuscript.
 
@@
 
-[distinction between clearest representation of the grammar for didactic purposes, versus the best representation for implementation purposes; for implementation we used a Moore machine representation]
+We distinguished between these states to help clarify our method,
+where states are classified as "Start" or "Insert" to simply the
+statement of the composition algorithm.  For implementation purposes
+we certainly agree that it is a mistake to have "duplicate" states and
+when possible, we remove these from the resulting 3-sequence SCFG.
 
@@
 > I_L can produce an LS bifurcation directly, rather than invoking 
-> nonterminal B. The L and I_L productions don't appear to be correctly 
+> nonterminal B.
+
+As with L and I_L, we drew this distinction for clarity of approach,
+so that we could distinguish between states which emitted nonterminals
+and bifurcation states to make the composition algorithm more
+intuitive.  A transition S -> B_e -> L is classified as an "empty
+bifurcation" as discussed in the revised text and replaced with an
+"equivalent direct transition" of precisely the form which you suggest.
+
+> The L and I_L productions don't appear to be correctly 
 > normalized, unless the (undefined) term p_S is 0.0.
+
+The equilibrium distribution of unpaired nucleotides is co-normalized
+with the stem probability p_S (labeled as \pi_l (S) for clarity in the
+revised manuscript), so that the production rules for L and I_L are
+properly normalized.  We have added a note to this effect in the
+caption for Table 2 in the revised manuscript.
 
@@
 > side -- where are they being absorbed from?)
-
 > The productions also 
@@
 
-5' and 3' ends of structures are now labeled (e.g., Figures 3 and 4 of
-the revised manuscript).
+The 5' and 3' ends of structures are now labeled (e.g., Figures 3 and
+4 of the revised manuscript).
 

### 2009-05-21 paper.tex
Final edits.

@@
 The statistics of the TKFST model on the three-taxon phylogeny
-illustrate the advantages of our procedure for model
-construction.  While the singlet and branch transducers are relatively simple---the singlet transducer,
-shown in \tabref{tkfstsinglet}, has 7 total states and 2 bifurcation states and
-the branch transducer, shown in \tabref{tkfstbranch}, has 21 total states
-and 6 bifurcation states---the ensemble model of three extant sequences 
-is very complex.
-The naive exponential upper-bound gives a maximal state space of size $O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.
-Using our uninformed search algorithm, we determined that there are 287 accessible states
-and 686 possible transitions between these states (compare with the $287^2 \simeq 8 \cdot 10^4$ transitions
-estimated with the exponential calculation).  After performing the
+illustrate the advantages of our procedure for model construction.
+While the singlet and branch transducers are relatively simple---the
+singlet transducer, shown in \tabref{tkfstsinglet}, has 7 total states
+and 2 bifurcation states and the branch transducer, shown in
+\tabref{tkfstbranch}, has 21 total states and 6 bifurcation
+states---the ensemble model of three extant sequences is very complex.
+The naive exponential upper-bound gives a maximal state space of size
+$O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.  Using our
+uninformed search algorithm, we determined that there are 287
+accessible states and 686 possible transitions between these states
+(compare with the $287^2 \simeq 8 \cdot 10^4$ transitions estimated
+with the exponential calculation).  After performing the
 transformations described in ``Eliminating null cycles in SCFGs'' to
-eliminate useless windback states,
-the ensemble model has a reduced state space with 230 states,
-albeit at the cost of extra transitions, bring the total to 1,789 transitions
-(here we are trading reduced memory complexity, which is linear in the number
-of states, for increased time complexity, which is linear in the number of transitions).
-Note that both before and after the reduction in complexity,
-the total number of states and transitions are less than the approximate bounds
-of $O(a \cdot b \cdot N) = O(7 \cdot 21 \cdot 3) \Rightarrow 441$ states
-and
-$O(a \cdot b^2 \cdot N) = O(7 \cdot 21^2 \cdot 3) \Rightarrow 9,261$ transitions
-suggested in ``The composition algorithm.''
-
-% to do: dump output of tripletscfg.pl --tkfst into supplement
-Text S1 contains the complete model specification, including states and transitions,
-produced by our model composition algorithm.  The extreme complexity of the
+eliminate useless windback states, the ensemble model has a reduced
+state space with 230 states, albeit at the cost of extra transitions,
+bring the total to 1,789 transitions (here we are trading reduced
+memory complexity, which is linear in the number of states, for
+increased time complexity, which is linear in the number of
+transitions).  Note that both before and after the reduction in
+complexity, the total number of states and transitions are less than
+the approximate bounds of $O(a \cdot b \cdot N) = O(7 \cdot 21 \cdot
+3) \Rightarrow 441$ states and $O(a \cdot b^2 \cdot N) = O(7 \cdot
+21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in ``The
+composition algorithm.''  Nonetheless, the extreme complexity of the
 ensemble model, despite the simplicity of the underlying model of RNA
 structure, makes clear the necessity for automated procedures for
-model construction.
+model construction.  Dataset S1 gives the state space of the composed
+model constructed by the search algorithm and Dataset S2 the reduced
+model after eliminating ``windback'' states; both are in Graphviz format for
+visualization and show the state of the singlet transducer generating
+ancestral sequence as well as the states of the branch transducers
+generating observed sequences.
 
@@
   \begin{flushleft}
-    The state types for this model are shown in \tabref{tkfstsinglet-types}.
-    The singlet transducer generates ancestral RNA sequences and
-    structures.
-    We use the notation of formal grammars to represent state
-    transformation rules; for example, the rule $I_L \rightarrow
-    u\,\,I_L$ corresponds to (in a Pair HMM) an $\Sinsert$ state $I_L$
-    emitting a nucleotide $u$ and then making a self-transition.
-    Both loop ($L$ and $I_L$) and stem ($S$ and $I_S$) sequence evolve as TKF sequences
-    with length parameters $\kappa_l$ and $\kappa_s$ (defined in ``A simple model of RNA structural evolution'').
-    $\pi_l(u)$ and $\pi_s(uv)$ are the equilibrium distributions of unpaired nucleotides $x$ and
-    paired nucleotides $(u,v)$.
-    The bifurcation state $B_e$ is used to end stem sequences
-    (only loop sequences are allowed to transition to the empty string).
+    The state types for this model are shown in
+    \tabref{tkfstsinglet-types}.  The singlet transducer generates
+    ancestral RNA sequences and structures.  We use the notation of
+    formal grammars to represent state transformation rules; for
+    example, the rule $I_L \rightarrow u\,\,I_L$ corresponds to (in a
+    Pair HMM) an $\Sinsert$ state $I_L$ emitting a nucleotide $u$ and
+    then making a self-transition.  Both loop ($L$ and $I_L$) and stem
+    ($S$ and $I_S$) sequence evolve as TKF sequences with length
+    parameters $\kappa_l$ and $\kappa_s$ (defined in ``A simple model
+    of RNA structural evolution'').  $\pi_l(u)$ and $\pi_s(uv)$ are
+    the equilibrium distributions of unpaired nucleotides $u$ and
+    paired nucleotides $(u,v)$ and are normalized such that
+    $\pi_l(S) + \sum_u \pi_l(u) = 1$ and $\sum_{u,v} \pi_s(uv) = 1$.  The bifurcation state $B_e$ is used
+    to end stem sequences (only loop sequences are allowed to
+    transition to the empty string).
   \end{flushleft}

### 2009-05-21 paper.tex
comment for revision
small edits

@@
 algorithms fast enough for analysis of many biological datasets.
-As discussed below, the complexity of our inference algorithms is
+As discussed below, the complexity of general inference algorithms is
 prohibitively high for many problems of interest.
@@
 Similarly, we might consider only ancestral structures which are
-compatible with a curated multiple sequence alignment.  Such constraints are
-commonly used by programs for RNA sequence analysis \cite{RivasEddy2001,Holmes2005}.
+compatible with a given multiple sequence alignment, or a relatively small set of candidate alignments (as in the ORTHEUS program \cite{PatenHolmesBirney2008}). 
+Such constraints are
+commonly used by programs for SCFG-based RNA sequence analysis such as QRNA \cite{RivasEddy2001}, Stemloc \cite{Holmes2005} and CONSAN \cite{DowellEddy2006}.
+Alignment and structural constraints can be combined \cite{Holmes2005}.
 
@@
 
-The stochastic state machine which generates parse trees from the marginal distribution $P(X)$ is a {\bf singlet transducer}.
-The machine that absorbs parse trees $X$ and generates modified parse trees $Y$ from the conditional distribution $P(Y | X, T)$ is a {\bf branch transducer}.
-Singlet transducers only emit symbols to their output sequence;
-branch transducers, in contrast, can both emit symbols to their output
-and absorb symbols from their input sequence.
-
-Much as the states of Pair HMMs with no null states, other than the $\Sstart$
-and $\Send$ states, can be classified as $\Sinsert$ or $\Smatch$,
-transducers can have states of type $\Sstart$, $\Send$,
+Consider the stochastic grammar which generates parse trees from the marginal distribution $P(X)$.
+It is occasionally convenient to represent this grammar as a transducer whose input is constrained to be null,
+i.e. a machine that accepts a dummy (empty) input sequence, and outputs sequence $X$.
+We refer to this as the {\bf singlet transducer}.
+In contrast, the more general type of transducer that absorbs parse trees $X$ and generates modified parse trees $Y$ from the conditional distribution $P(Y | X, T)$ is a {\bf branch transducer}.
+By definition, singlet transducers only emit symbols to their output sequence, and use a restricted set of state types.
+Branch transducers, in contrast, can both emit symbols to their outputs and absorb symbols from their inputs, and so use the full range of state types.
+
+Transducers can have states of type $\Sstart$, $\Send$,
 $\Swait$, $\Sinsert$ and $\Smatch$.
 The first three state types, $\Sstart$, $\Send$ and $\Swait$, are null:
-They do not emit or absorb any symbols and are required solely for
+they do not emit or absorb any symbols and are required solely for
 ``organizational'' purposes (see following section).
@@
 
-The ensemble model contains empty bifurcations {\em and} empty paths.
+The ensemble grammar contains many rules that can be applied redundantly,
+together or in isolation, to generate subtrees of the parse tree that do
+not generate any terminals, and could be replaced with simpler rules.
+In fact, this is a general property of SCFGs with null states and/or bifurcations.
+In this subsection, we describe how to transform the SCFG so as to eliminate such redundant rules.
+We begin by identifying two types of redundant parse-structures: {\bf empty bifurcations} and {\bf empty paths}.
+
 An {\bf empty bifurcation} occurs when a child branch of a bifurcation state
@@
 N_3$ between null states $N_1$ and $N_3$ in place of the empty
-bifurcation $N_1 \rightarrow B \rightarrow (N_2 \,\, N_3) \rightarrow
+parse-subtree $N_1 \rightarrow B \rightarrow (N_2 \,\, N_3) \rightarrow
 (End \,\, N_3)$, where $B$ is a bifurcation state with children $(N_2
@@
 
-Similarly, our model construction procedure generically produces many
-{\bf empty paths} which do not emit any terminal symbols.
+In contrast, an {\bf empty path} is defined as any parse-subtree {\em without} bifurcations that does not emit terminal symbols.
 Most common are paths of ``windback'' null transitions via non-emitting placeholder states.
 If $\Semit$ states $E_1$ and $E_2$ are connected in the state
-graph by $\Snull$ states $N_1$ and $N_2$, then the transition $E_1
+graph via $\Snull$ states $N_1$ and $N_2$, then the path $E_1
 \rightarrow N_1 \rightarrow N_2 \rightarrow E_2$ with  probability
 $P(E_1 \rightarrow N_1) \cdot P(N_1 \rightarrow N_2) \cdot P(N_2
-\rightarrow E_2)$ can be replaced with a transition $E_1 \rightarrow
+\rightarrow E_2)$ can be replaced by a single direct transition $E_1 \rightarrow
 E_2$ with an identical probability.
@@
 
-We refer to empty paths and empty bifurcations as {\bf null cycles}.
-They occur (in simpler form) in Hidden Markov Models (which are special cases of SCFGs)
+Empty paths occur in Hidden Markov Models (which are special cases of SCFGs)
 and independent-sites models (which can be viewed as special cases of HMMs).
-In all such models, null cycles represent histories that are valid
+Conceptually, empty paths can represent histories that are valid
 according to the model but cannot be resolved by direct observation.
@@
 or they can be artefactual (e.g., transitions between placeholder null states of an HMM).
-Often, one can account for the likelihood of such events by exact marginalization.
-For example, eliminating missing ``empty'' columns from an independent-sites model of sequence involves a straightforward application of a correction factor
+
+We refer to empty paths and empty bifurcations collectively as {\bf null cycles}.
+Our goal is an algorithmic procedure to account for the likelihood of such cycles by exact marginalization.
+For simpler models, solutions to this problem are published.
+Eliminating missing (empty) columns from an independent-sites model of sequence involves a straightforward application of a correction factor
 to account for the proportion of columns that are unobserved \cite{RivasEddy2008}.
 The slightly more complicated situation of missing emissions in a HMM can be dealt with by
-summing over all null paths, yielding a geometric series that is solvable by matrix inversion
-\cite{Holmes2003,Holmes2007,Lunter2007}; the ``wing retraction'' algorithms used in HMMER can be viewed as a special case of matrix inversion by iterated elimination \cite{hmmer}.
-
-Algorithms for exact elimination of HMM null cycles by matrix inversion are implemented
+summing over all empty paths, yielding a geometric series that is solvable by matrix inversion
+\cite{Holmes2003,Holmes2007,Lunter2007}.
+
+Algorithms for exact elimination of empty paths from HMMs by matrix inversion are implemented
 by programs such as PhyloComposer \cite{Holmes2003,Holmes2007} and
@@
 same probability distribution over sequences.
-However, these solutions do not easily generalize to SCFGs, due to the
-possible presence of null bifurcations.
+The ``wing retraction'' algorithms used in the repetitively structured profile HMMs of HMMER \cite{hmmer} appear similar in operation.
+However, these solutions do not easily generalize to SCFGs, which contain bifurcations
+(and, therefore, may generate empty bifurcations as well as empty paths).
 
@@
 should be sufficient for ergodic sampling of reconstructions on any larger phylogeny,
-using a MCMC kernel similar to that of \cite{JensenHein2002}.
+using e.g. a Gibbs-sampling MCMC kernel \cite{JensenHein2002}
+or a progressive suboptimal-alignment sampling heuristic \cite{PatenHolmesBirney2008}.
 
@@
 We simulated the evolution of RNAs under the TKFST model along
-3-taxon phylogenies, where we kept the branch lengths of the two
+3-taxon phylogenies (with one internal node), where we kept the branch lengths of two
 sibling species constant and varied the branch length of the
@@
 10]$ bp and stems of lengths $\in [1, 20]$ bp.
+We then used \indiegram\ to reconstruct maximum likelihood reconstructions of the ancestral structure and the four-way multiple alignment (ancestor + three descendants).
 
@@
 have a perfectly-parametrized evolutionary model, recent results
-demonstrating that inference of sequence alignment is surprisingly
+demonstrating that inference of sequence alignment is
 robust to mis-parametrized probabilistic models \cite{Lunter2007}

### 2009-05-22 paper.tex
small edits

@@
 Due to this possibility of unobserved ancestral sequence,
-the standard CYK and Inside algorithms fail on the ensemble model.
-A potentially-unbounded number of residues in the ancestor can be
+the standard CYK and Inside recursions fail on the ensemble model:
+that is, they contain cyclic self-references which cannot be resolved.
+The intuitive reason for this is that a (potentially-unbounded) number of residues in the ancestor can be
 absent in all extant sequences.
@@
 are absent in any extant sequence, further complicates inference.
-While there exists a straightforward fix for maximum-likelihood
-inference with the CYK algorithm, the Inside algorithm requires
-further computational machinery.
-
+While there exists a straightforward fix for the CYK algorithm, the Inside algorithm requires more work.
+
+Consider the CYK algorithm first.
 In order to correctly perform maximum-likelihood inference, we only
@@
 linguistics requires only that the CYK parse be preserved \cite{GecsegSteinby97}.)
-However, posterior-decoding inference, which uses the Inside and
-Outside algorithms, required a modified DP algorithm which preserves
-the entire probability distribution over parses; the transformations
-just described to make CYK work properly are insufficient for the
+
+The transformations just described to make CYK work properly are insufficient for the
 Inside and Outside algorithms.
-
-There are two possible solutions to this difficulty: We can either
+For CYK, null cycles can be ignored, since they should never appear in the maximum likelihood parse.
+For Inside and Outside, null cycles cannot be ignored, since they will contribute to the likelihood sum over parse trees.
+
+There are two possible solutions to this difficulty: we can either
 create new versions of the standard inference algorithms for SCFGs
@@
 knowledge, efficient exact null-cycle elimination for SCFGs is an open
-mathematical problem (complicated by the presence of null cycles
+problem (complicated by the presence of null cycles
 involving bifurcations).
@@
 
-The ensemble grammar contains many rules that can be applied redundantly,
+As noted above, the ensemble grammar contains many rules that can be applied redundantly,
 together or in isolation, to generate subtrees of the parse tree that do

### 2009-05-26 paper.tex
edits

@@
 absent in all extant sequences.
-This manifests as cycles of $\Snull$ states within the multiple-sequence SCFG,
+This is manifested as cycles of $\Snull$ states within the multiple-sequence SCFG,
 states which emit to the ancestral sequence but not to any extant sequence.
@@
 are absent in any extant sequence, further complicates inference.
-While there exists a straightforward fix for the CYK algorithm, the Inside algorithm requires more work.
-
-Consider the CYK algorithm first.
+While there exists a straightforward fix for the CYK algorithm, the Inside algorithm is more complicated.
+
+We consider the CYK algorithm first.
 In order to correctly perform maximum-likelihood inference, we only
@@
 Inside and Outside algorithms.
-For CYK, null cycles can be ignored, since they should never appear in the maximum likelihood parse.
-For Inside and Outside, null cycles cannot be ignored, since they will contribute to the likelihood sum over parse trees.
+For CYK, parse trees containing null cycles can be ignored, since the maximum likelihood parse tree should never contain null cycles.
+For Inside and Outside, such null-cycle parses cannot be ignored, since they will contribute to the likelihood sum over parse trees.
 
@@
 and necessarily has worse time complexity than the standard inference
-algorithms due to the need to iterate over parts of the (large) DP
+algorithms due to the need to reiterate over (large) parts of the DP
 matrix.  The second approach, of modifying the structure of the SCFGs
-themselves, has a closed-form solution for the case of HMMs, but to our
-knowledge, efficient exact null-cycle elimination for SCFGs is an open
-problem (complicated by the presence of null cycles
-involving bifurcations).
-
-We therefore developed a method to remove null cycles from
-the ensemble model, described below, in order to use the standard inference algorithms
+themselves, is well-understood for the case of HMMs, but to our
+knowledge, still an open problem for SCFGs
+(complicated by the presence of null cycles involving bifurcations).
+
+We therefore developed a method to eliminate null cycles from
+the ensemble model, in order to use the standard inference algorithms
 for SCFGs without incurring the further computational complexity
 required by modifying the DP algorithms themselves.
+We will describe this elimination procedure before proceeding to the descriptions of the DP algorithms.
 
@@
 together or in isolation, to generate subtrees of the parse tree that do
-not generate any terminals, and could be replaced with simpler rules.
-In fact, this is a general property of SCFGs with null states and/or bifurcations.
-In this subsection, we describe how to transform the SCFG so as to eliminate such redundant rules.
-We begin by identifying two types of redundant parse-structures: {\bf empty bifurcations} and {\bf empty paths}.
+not generate any terminals.
+This generates cyclic dependencies in the standard DP recursions for inference.
+In this subsection, we describe how to transform the SCFG so as to eliminate such redundant rules, yielding strictly acyclic DP recursions.
+This transformation can be applied to any SCFG so as to remove null states and/or bifurcations:
+the procedure is not restricted to grammars that were generated using our transducer composition algorithm.
+
+We begin by identifying two distinct classes of redundant parse-subtree: {\bf empty bifurcations} and {\bf empty paths}.
+We will eliminate each of these in turn.
 
@@
 
-Here, we show how null cycles can be removed systematically from SCFGs by two steps:
+Null cycles can be removed systematically from SCFGs by two steps:
 \begin{description}
@@
 a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
-and a non-negative weight function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.
-
-(The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
-The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)
+and a probability function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.
+
+% (The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
+% The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)
 
@@
 Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
-then define the parse tree weight $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.
+then define the parse tree likelihood $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.
 
@@
 The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
-Of particular relevance to null state elimination are the probabilities $P(\epsilon|A)$,
+Of particular relevance to null state elimination is the probability $P(\epsilon|A)$,
 where $\epsilon$ is the empty sequence.
-These are the probabilities that a nonterminal will expire without generating any sequence.
+This is the probability that a nonterminal $A$ will expire without generating any sequence.
 
@@
 
-Furthermore, we say that the grammar $G$ has {\em RNA normal-form states} if
+Suppose that grammar $G$ has RNA normal-form rules.
+We further say that $G$ has {\em RNA normal-form states} if
 each nonterminal (state) $A \in \gnons$ takes one of the following forms:
@@
 
-The idea is that $N'$ is a null state that is equivalent to null state $N$ for non-empty sequence.
+The idea is that $N'$ (in $G'$) is a null state that is equivalent to null state $N$ (in $G$) for non-empty sequence only.
 That is, for any terminal sequence $S$,
@@
 dependence of reconstruction accuracy on outgroup branch length.
+(We here use the term ``outgroup'' loosely, to denote the variable-length branch in a three-branch study,
+where the other two branches are held at unit length.
+Technically, the term ``outgroup'' is only correct if the outgroup's branch is longer than the other two branches.)
+
 We simulated the evolution of RNAs under the TKFST model along
@@
 information which it contains.
-Similarly, while we can perfectly reconstruct the ancestral
-structure given an outgroup branch length of 0 (the trivial case), we
-cannot perfectly reconstruct the alignment.
-
-While these results correspond to a best-case scenario in which we
-have a perfectly-parametrized evolutionary model, recent results
+In the trivial case where the outgroup branch length is zero (so the ancestor is identical to the outgroup),
+we can perfectly reconstruct the ancestral structure (as expected),
+but we cannot perfectly reconstruct the three-way multiple alignment (since there are still two finite-length branches, on which some unreconstructable mutations inevitably occur).
+
+These results correspond to a best-case scenario in which we
+have a perfectly-parametrized evolutionary model and exactly-known descendant structures.
+Recent results
 demonstrating that inference of sequence alignment is
@@
 ancestral structures.
+The effect of imperfect knowledge of the ancestral structures has not been explored.
 
@@
 by Sakakibara et al \cite{Sakakibara2003}.
-The statistical inference algorithms for multiple-sequence SCFGs are closely related to the ``protosequence algorithm'' of Sankoff \cite{Sankoff85},
-a seminal work in both RNA structural alignment and ancestral reconstruction.
+The dynamic programming inference algorithms for multiple-sequence SCFGs are closely related to the ``protosequence algorithm'' of Sankoff \cite{Sankoff85}.
+%a seminal work in both RNA structural alignment and ancestral reconstruction.
 
@@
 using the three-way sampling kernel which we have described
-would require resources far in excess of what are currently available;
+would require resources far in excess of those currently available;
 barring the availability of supercomputers with terabytes of memory,

### 2009-05-26 paper.tex
edits
added SimulFold ref

@@
 
+The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
+
 \newpage
@@
 structural alignment.
+An alternative MCMC scheme for sampling RNA phylogeny, structure and alignment was implemented in the SimulFold program \cite{MeyerMiklos2007}.
+SimulFold does not use a strictly normalized probabilistic model, and it is unclear how structure and indels interact in the model
+(for example, it does not penalize deletion of one half of a basepair).
 

### 2009-05-26 paper.tex
added SimulFold ref

--- 2009-05-26 paper.tex
@@
 evolution, we simulated data under the model and tested the dependence
-of alignment inference and ancestral structure reconstruction accuracy on the evolutionary
-distance of the outgroup of the phylogeny.  Our results suggest that
+of alignment inference and ancestral structure reconstruction accuracy on phylogenetic branch length.
+Our results suggest that
 we may be able to accurately reconstruct ancestral structures despite
@@
 \item[A.] What is the likelihood for the observed sequence data?
-\item[B.] Can we sample from (or exactly compute) the posterior distribution of the unobserved sequence at the root node?
-\item[C.] Can we sample from the posterior of the unobserved sequences at the other internal nodes?
+\item[B.] Can we sample (find the mode, take moments, etc.) from the posterior distribution of the unobserved sequence at the root node?
+\item[C.] Can we sample (etc.) from the posterior of the unobserved sequences at the other internal nodes?
 \item[D.] Can we estimate summaries of the evolutionary history, such as
@@
 progressive reconstruction used by the \ortheus\ program for
-reconstructing ancestral genomes \cite{PatenEtAl2008}.
+reconstructing ancestral genomes \cite{PatenHolmesBirney2008}.
 Another promising direction is to consider variations on the three-way sampling kernel,
@@
 
-The evolutionary models we have described form a theoretical
+An alternative MCMC scheme for sampling RNA phylogeny, structure and alignment was developed for the SimulFold program \cite{MeyerMiklos2007}.
+SimulFold does not use a strictly normalized probabilistic model, resulting in some oddities in the ways that structure and indels interact
+(for example, it does not penalize deletion of one half of a basepair).
+Currently, it is not clear how appropriate SimulFold would be for ancestral reconstruction,
+although it has several advantages (e.g. explicit treatment of pseudoknots).
+Of course, MCMC kernels can often be adapted for use with different models, and the MCMC moves used by SimulFold may be useful for inference under our model as well.
+
+In summary, the evolutionary models we have described form a theoretical
 foundation for studying the structural evolution of RNA gene families

### 2009-05-27 response-to-reviews2.txt
cut swathes out of intro to elimination/DP supersection

@@
 supplementary material in this section and have attempted to make the
-main text a self-contained description of the composition algorithm.
+main text a self-contained description of the composition algorithm,
+including its key corollaries (elimination of null states, and DP inference).
 We have attempted to remove redundant text, including the two
-paragraphs which the reviewer notes.
+paragraphs which the reviewer notes, and other sentences.
 
@@
 is no more than a formalization of an intuitive concept, we believe
-that it is nonetheless substantial, due to the sheer complexity
+that it is nonetheless an essential reference point, due to the sheer complexity
 of the ensemble models so created.  The "TKFST" model which we use
@@
 sequences (the xrate program). We have now automated this process and
-fixed the bug, verifying that this has eliminated the errors.
+fixed the bug, verifying that this has eliminated the errors
+(in fact this verification is incorporated into our new benchmark).
 
@@
 
-We chose to call it a "singlet" transducer so that we used similar
-terminology for the model generating the marginal and conditional
-distributions.  A singlet transducer can be viewed as a special case
-of a transducer, where the input being transduced is empty.  Indeed,
-by "singlet transducer" we mean the same thing as is often referred to
-as a "single-sequence SCFG".  We apologize if it is poor terminology;
-we can change it if the reviewer feels that doing so would help
-clarify the methods.
+By "singlet transducer" we mean the same thing as is often referred to
+as a "single-sequence SCFG".  Instead of "a model that generates a
+parse tree X", we use the contrivance of saying "a model that inputs a
+(dummy) empty parse tree, and outputs a parse tree X".  It's annoying
+but it keeps the notation straight: it allows us to be consistent
+throughout the manuscript. Rather than always having to specify
+whether a model is generative (SCFG) or transductive (input-output),
+we can treat all models as transducers, if we just think of generative
+models as transducers with a (dummy) input that is always zero/empty.
+
+We apologize if it is poor terminology. We just think it's as clear as
+we can get. We can change it if the reviewer feels that doing so would
+help clarify the methods.
 

### 2009-05-27 paper.tex
cut swathes out of intro to elimination/DP supersection

@@
 of multiple related RNAs, using the multiple-sequence SCFG we have derived.
-The model construction algorithm described above creates an evolutionary model
+
+The model construction algorithm described above creates a phylogenetic SCFG
 which generates a joint distribution over both ancestral and extant
-sequences.  If we regard the ancestral sequences as hidden variables
-to be marginalized over, then this model corresponds to a
-jointly-normalized multiple-sequence SCFG (modeling the extant sequences).
-We can then use the Cocke-Younger-Kasami (CYK) algorithm to infer the
-maximum-likelihood alignment, secondary structure and common ancestor
-of the extant sequences or the Inside algorithm to sample from
-the joint posterior distribution of these unknowns.
-
-The possibility of fragments of ancestral sequence that are unobserved
-in any of the descendant sequences, which can be unbounded in length,
-leads to complications when performing inference on the multiple-sequence
-SCFG. The SCFG can contain null cycles, corresponding to generation of
-these unobserved fragments, which complicates inference.
-
-Due to this possibility of unobserved ancestral sequence,
-the standard CYK and Inside recursions fail on the ensemble model:
-that is, they contain cyclic self-references which cannot be resolved.
-The intuitive reason for this is that a (potentially-unbounded) number of residues in the ancestor can be
-absent in all extant sequences.
-This is manifested as cycles of $\Snull$ states within the multiple-sequence SCFG,
-states which emit to the ancestral sequence but not to any extant sequence.
-The presence of bifurcations with one or more possibly-empty child branches,
-corresponding to entire substructures in the ancestral sequence that
-are absent in any extant sequence, further complicates inference.
-While there exists a straightforward fix for the CYK algorithm, the Inside algorithm is more complicated.
-
-We consider the CYK algorithm first.
-In order to correctly perform maximum-likelihood inference, we only
-require that the maximum-likelihood parse tree corresponds to the parse
-returned by the CYK algorithm; the entire
-probability distribution over parses need not be preserved.
-Because a state path which includes a cycle of $\Snull$ states
-necessarily has lower probability than an equivalent one without, we
-can simply ignore such $\Snull$ cycles during CYK.
-Bifurcations with empty child branches cannot be similarly ignored,
-but this problem can be addressed by creating equivalent effective
-direct transitions between the parents of these bifurcation states and
-their non-empty children.  Once these transformations have been performed,
-maximum-likelihood inference can be performed using a standard
-three-sequence CYK algorithm on the ensemble model.
-(Similarly, the null cycle elimination discussed in computational
-linguistics requires only that the CYK parse be preserved \cite{GecsegSteinby97}.)
-
-The transformations just described to make CYK work properly are insufficient for the
-Inside and Outside algorithms.
-For CYK, parse trees containing null cycles can be ignored, since the maximum likelihood parse tree should never contain null cycles.
-For Inside and Outside, such null-cycle parses cannot be ignored, since they will contribute to the likelihood sum over parse trees.
-
-There are two possible solutions to this difficulty: we can either
-create new versions of the standard inference algorithms for SCFGs
-which can cope with SCFGs with null cycles or we can modify the
-structure of the SCFGs generated by our model-construction algorithm
-such that they are free of null cycles.  The first approach of
-creating new algorithms is simple in principle---we can simply
-iteratively re-fill entries of the DP matrix, corresponding to summing
-the probabilities for unobserved null cycles, until the approximation
-is sufficiently good---but it is difficult to implement efficiently
-and necessarily has worse time complexity than the standard inference
-algorithms due to the need to reiterate over (large) parts of the DP
-matrix.  The second approach, of modifying the structure of the SCFGs
-themselves, is well-understood for the case of HMMs, but to our
-knowledge, still an open problem for SCFGs
-(complicated by the presence of null cycles involving bifurcations).
-
-We therefore developed a method to eliminate null cycles from
-the ensemble model, in order to use the standard inference algorithms
-for SCFGs without incurring the further computational complexity
-required by modifying the DP algorithms themselves.
-We will describe this elimination procedure before proceeding to the descriptions of the DP algorithms.
+sequences. A parse tree for this SCFG represents a structural and evolutionary
+explanation of the extant sequences, including a complete ancestral reconstruction.
+Therefore, given a set of extant sequences, many of the questions of interest to us
+can be reduced to searches over, or summarizations of, the set of possible parse trees.
+
+Well-known algorithms exist for searching or summing over SCFG parse trees.
+The Cocke-Younger-Kasami (CYK) algorithm performs maximum-likelihood (ML) inference;
+the Inside algorithm can be used to sum over parse trees or sample them {\em a posteriori};
+and the Inside-Outside algorithm yields posterior probabilities for individual parse tree nodes.
+
+All of these algorithms are, however, complicated for phylogenetic SCFGs by
+the possibility of fragments of ancestral sequence that were universally deleted and therefore
+are unobserved in any of the extant sequences.
+These unobserved fragments can be unbounded in length.
+In the parse tree, these fragments correspond to cyclic detours through $\Snull$ states,
+along with bifurcations where one or both sides terminate without observable emissions
+(e.g. $B \to X\ Y$ followed by $X \to \epsilon$).
+In the CYK, Inside and Outside recursions, this causes cyclic dependencies which cannot be resolved.
+
+Therefore, before presenting equivalents of the CYK, Inside and Outside algorithms,
+we first give a method to eliminate ``null cycles'' from
+the ensemble model, by transforming any SCFG to an equivalent acyclic SCFG.
+Some such step is required, in order to deal with cyclic dependencies, but alternative approaches to ours are clearly possible.
+A CYK-only approach appears in the computational linguistics literature \cite{GecsegSteinby97}.
+We have also developed an independent heuristic that iteratively re-fills entries of the DP matrix
+until the approximation is sufficiently good.
 
@@
 
-We refer to empty paths and empty bifurcations collectively as {\bf null cycles}.
+Empty paths and empty bifurcations can be combined to give finite-probability sequences of rules
+that transform a nonterminal back into itself, with no observable emissions.
+We refer to such sequences of rules as {\bf null cycles}.
 Our goal is an algorithmic procedure to account for the likelihood of such cycles by exact marginalization.
@@
 null cycles from the multiple-sequence SCFGs generated by our
-model-construction algorithm, we can use the standard CYK, Inside and
-Outside algorithms for multiple-sequence SCFGs given by Sankoff \cite{Sankoff85}.
+model-construction algorithm, we can compute likelihoods and sample parse trees
+using the ``standard'' CYK, Inside and Outside algorithms for multiple-sequence SCFGs \cite{Sankoff85,Durbin98}.
 
 The asymptotic time and memory complexities of our inference algorithms
-are essentially the same as for the Sankoff algorithm \cite{Sankoff85}:
+are essentially the same as for Sankoff's algorithm \cite{Sankoff85}:
 the DP algorithms take memory $O(A \cdot L^{2N})$ and time $O(B \cdot L^{3N})$
@@
 described earlier, in practice we frequently have expert knowledge
-about the structures or evolutionary histories of the sequences of
-interest, such as a curated multiple alignment.  We can use these
-constraints to reduce the accessible volume of the DP matrix \cite{Holmes2005}.
+(such as a curated multiple alignment)
+about the structures and/or evolutionary histories of the sequences of
+interest.  We can use this knowledge as a constraint to reduce the accessible volume of the DP matrix \cite{Holmes2005}.
 Algorithms \ref{alg:inside}--\ref{alg:cyktraceback}, described below, give the Inside, CYK,

### 2009-05-27 paper.tex
more edits

@@
 Consider the stochastic grammar which generates parse trees from the marginal distribution $P(X)$.
-It is occasionally convenient to represent this grammar as a transducer whose input is constrained to be null,
+It is convenient to represent this grammar as a transducer whose input is constrained to be null,
 i.e. a machine that accepts a dummy (empty) input sequence, and outputs sequence $X$.
@@
 
-The model construction algorithm described above creates a phylogenetic SCFG
-which generates a joint distribution over both ancestral and extant
-sequences. A parse tree for this SCFG represents a structural and evolutionary
+The transducer composition algorithm described above constructs a phylogenetic SCFG
+for both ancestral and extant sequences.
+A parse tree for this SCFG represents a structural and evolutionary
 explanation of the extant sequences, including a complete ancestral reconstruction.
-Therefore, given a set of extant sequences, many of the questions of interest to us
+Consequently, given a set of extant sequences, many of the questions of interest to us
 can be reduced to searches over, or summarizations of, the set of possible parse trees.
 
-Well-known algorithms exist for searching or summing over SCFG parse trees.
+Well-known algorithms already exist for maxing or summing over SCFG parse tree likelihoods
+(indeed, this is the motivation for our entire approach).
 The Cocke-Younger-Kasami (CYK) algorithm performs maximum-likelihood (ML) inference;
 the Inside algorithm can be used to sum over parse trees or sample them {\em a posteriori};
-and the Inside-Outside algorithm yields posterior probabilities for individual parse tree nodes.
-
-All of these algorithms are, however, complicated for phylogenetic SCFGs by
-the possibility of fragments of ancestral sequence that were universally deleted and therefore
+and the Inside-Outside algorithm yields posterior probabilities for individual parse tree nodes \cite{Durbin98}.
+
+All of these algorithms are, however, complicated (at least in our models) by
+the existence of ``null cycles'' in the grammar.
+A null cycle is a parse tree fragment that is redundant and could be removed, such as a detour through $\Snull$ states ($A \to X \to Y \to X \to Y \to B$)
+that could be replaced by a direct transition ($A \to B$).
+Biologically, null cycles correspond to fragments of ancestral sequence that were universally deleted and therefore
 are unobserved in any of the extant sequences.
-These unobserved fragments can be unbounded in length.
-In the parse tree, these fragments correspond to cyclic detours through $\Snull$ states,
-along with bifurcations where one or both sides terminate without observable emissions
-(e.g. $B \to X\ Y$ followed by $X \to \epsilon$).
-In the CYK, Inside and Outside recursions, this causes cyclic dependencies which cannot be resolved.
-
-Therefore, before presenting equivalents of the CYK, Inside and Outside algorithms,
-we first give a method to eliminate ``null cycles'' from
+These unobserved fragments can be unbounded in length (and so, therefore, can the parse tree).
+Within the CYK, Inside and Outside recursions, this causes cyclic dependencies which cannot be resolved.
+
+Below we describe a method to eliminate null cycles from
 the ensemble model, by transforming any SCFG to an equivalent acyclic SCFG.
-Some such step is required, in order to deal with cyclic dependencies, but alternative approaches to ours are clearly possible.
-A CYK-only approach appears in the computational linguistics literature \cite{GecsegSteinby97}.
-We have also developed an independent heuristic that iteratively re-fills entries of the DP matrix
-until the approximation is sufficiently good.
-
-\subsubsection*{Eliminating null cycles in SCFGs}
+We then present multiple-sequence versions of the CYK, Inside and Outside algorithms.
+
+As an aside: while some sort of null-cycle elimination is often required in order to deal with cyclic dependencies, there are several ways to accomplish this other than the algorithm presented here.
+A simpler approach (that only works for the CYK algorithm) appears in the computational linguistics literature \cite{GecsegSteinby97}.
+We have also developed a heuristic for CYK that simply ignores null cycles, as well as an iterative approximation that loops several times over cyclically-dependent cells of the DP matrix
+until the estimate starts to converge. For conciseness, we have omitted descriptions of these methods, presenting only the exact elimination algorithm.
+
+\subsubsection*{Exact elimination of null cycles in SCFGs}
 
@@
 In contrast, an {\bf empty path} is defined as any parse-subtree {\em without} bifurcations that does not emit terminal symbols.
-Most common are paths of ``windback'' null transitions via non-emitting placeholder states.
 If $\Semit$ states $E_1$ and $E_2$ are connected in the state
@@
 
-Empty paths and empty bifurcations can be combined to give finite-probability sequences of rules
+In our composite model, empty paths occur whenever a series of branch transducers ``winds back'' into $\Swait$ states.
+Empty bifurcations occur when an entire substructure, present in an ancestor, is deleted in all that ancestor's extant descendants.
+
+Empty paths and empty bifurcations are problematic because they can be combined to give finite-probability sequences of rules
 that transform a nonterminal back into itself, with no observable emissions.
 We refer to such sequences of rules as {\bf null cycles}.
-Our goal is an algorithmic procedure to account for the likelihood of such cycles by exact marginalization.
+As noted, null cycles generate cyclic dependencies in the CYK, Inside \& Outside algorithms.
+Our goal is an algorithmic procedure to resolve these dependencies and account for the likelihood of such cycles by exact marginalization.
+
 For simpler models, solutions to this problem are published.
-Eliminating missing (empty) columns from an independent-sites model of sequence involves a straightforward application of a correction factor
-to account for the proportion of columns that are unobserved \cite{RivasEddy2008}.
+Missing (empty) columns in independent-sites models can be accounted for by applying a correction factor $(1-p)^{-1}$
+to account for the proportion of columns $p$ that are unobserved \cite{RivasEddy2008}.
 The slightly more complicated situation of missing emissions in a HMM can be dealt with by
@@
 \cite{Holmes2003,Holmes2007,Lunter2007}.
-
-Algorithms for exact elimination of empty paths from HMMs by matrix inversion are implemented
-by programs such as PhyloComposer \cite{Holmes2003,Holmes2007} and
-HMMoC \cite{Lunter2007}. These algorithms effectively replace the HMM
-with an equivalent HMM that contains no null cycles but generates the
+Such algorithms
+%, implemented by programs such as PhyloComposer \cite{Holmes2003,Holmes2007} and HMMoC \cite{Lunter2007},
+effectively replace the HMM
+with another HMM that contains no null cycles but is equivalent to the original, in that it models the
 same probability distribution over sequences.
-The ``wing retraction'' algorithms used in the repetitively structured profile HMMs of HMMER \cite{hmmer} appear similar in operation.
-However, these solutions do not easily generalize to SCFGs, which contain bifurcations
-(and, therefore, may generate empty bifurcations as well as empty paths).
+%The ``wing retraction'' algorithms used in the repetitively structured profile HMMs of HMMER \cite{hmmer} appear similar in operation.
+However, these solutions do not easily generalize to SCFGs
+(which may have empty bifurcations as well as empty paths).
 
@@
 
-We give a formal description of the null-cycle elimination procedure,
+Note that step (i) is unique to SCFGs; step (ii), in contrast, is very similar to the empty-path elimination algorithm for HMMs.
+
+We now give a formal description of the null-cycle elimination procedure,
 along with procedures for probabilistically ``restoring'' null cycles
-to sampled parse trees and Inside-Outside expectation counts, below.
+to sampled parse trees and Inside-Outside expectation counts.
 
@@
 Exact inference on a star phylogeny with $N$ extant sequences therefore
-has complexities $O(A \cdot L^{2N})$ and $O(B \cdot L^{3N})$ in space and
-time for a multiple-SCFG with $A$ states and $B$ bifurcations. As
+has complexities $O(A \cdot L^{2N})$ and $O(B \cdot L^{3N})$ in memory and
+time (respectively) for a multiple-SCFG with $A$ states and $B$ bifurcations. As
 described earlier, in practice we frequently have expert knowledge
@@
 about the structures and/or evolutionary histories of the sequences of
-interest.  We can use this knowledge as a constraint to reduce the accessible volume of the DP matrix \cite{Holmes2005}.
+interest.  We can use this knowledge as a constraint to reduce the accessible volume, and hence the storage requirements, of the DP matrix \cite{Holmes2005}.
 Algorithms \ref{alg:inside}--\ref{alg:cyktraceback}, described below, give the Inside, CYK,
@@
 using a simplified model which was introduced in previous work,
-the TKF Structure Tree model \cite{Holmes2004}, a simple probabilistic
+the TKF Structure Tree \cite{Holmes2004}, a simplified probabilistic
 model of the evolution of RNA structure.
@@
 The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
+Another script (\animateevolsayer) can be used to make animations of the evolving RNA structures, using the RNAplot program in the \viennarna\ package \cite{HofackerEtAl94}.
 
@@
 % Prior art
-The work reported in this paper builds on substantial prior art in the areas of evolutionary modeling and ancestral reconstruction.
+The work reported in this paper builds on extensive prior art in the areas of evolutionary modeling and ancestral reconstruction.
 The reconstruction of ancient sequences was first proposed in 1963 by Pauling and Zuckerkandl \cite{PaulingZuckerkandl63};
@@
 whether by parsimony \cite{Edwards63,HendyPenny82} or likelihood \cite{Felsenstein81}.
-There has been much effort to model the evolution of sequences related
-by a tree with models permitting indels
+There is a substantial body of work to model indels on phylogenies
 \cite{ThorneEtal91,HeinEtal2000,Hein2001,HolmesBruno2001,SteelHein2001,Holmes2003,KnudsenMiyamoto2003,LunterSongMiklosHein2003,MiklosLunterHolmes2004,LunterEtAl2004,BradleyHolmes2007,SatijaEtAl2008}.
@@
 particularly at the genomic scale \cite{MillerHausslerEtAl2006,PatenHolmesBirney2008}.
-There is substantial prior work in computational linguistics
+There is also prior work in computational linguistics
 on the theory of transducers for sequences \cite{MohriPereiraRiley2000}
@@
 %allowing fast training on structural alignments of pairs of sequences.
+Other phenomena of RNA evolution may prove more difficult:
+modeling helix slippage with a branch transducer is awkward,
+let alone more radical changes in structure;
+pseudoknots, too, are impossible with the models we have described here.
+Even so, variants of our models could be used for proposing candidate alignments
+for more accurate scoring by such models.
 
@@
 such algorithms will only be feasible for short RNAs (\tabref{complexities}).
-Thus, alternate strategies will have to be explored, such as the
-progressive reconstruction used by the \ortheus\ program for
-reconstructing ancestral genomes \cite{PatenHolmesBirney2008}.
-Another promising direction is to consider variations on the three-way sampling kernel,
+A promising direction is to consider variations on the three-way sampling kernel,
 such as the importance-sampling approach described for the TKF model
@@
 
-In summary, the evolutionary models we have described form a theoretical
-foundation for studying the structural evolution of RNA gene families
-in detail. By opening the door to ancestral sequence reconstruction
-and paleogenomics, these algorithms will allow us to test our
-knowledge of RNA evolution by direct experimental investigation
-of early ribonucleic machines.
+While MCMC provides the most information about the posterior distribution of evolutionary histories,
+in practice a maximum likelihood inference may be adequate (and typically faster).
+The progressive profiling used by the \ortheus\ program for
+reconstructing ancestral genomes is promising \cite{PatenHolmesBirney2008}.
+This approach is similar to a progressive multiple alignment algorithm, in that it proceeds via a single postorder (leaf-to-root) traversal of the phylogeny.
+As each node is visited, a profile is generated for that node, by aligning the profiles of its children to a composite transducer using DP, then sampling a finite number of traceback paths through the DP matrix.
+The profile is not linear: the sampled paths instead form a reticulate network, a.k.a. a partial order graph \cite{LeeGrassoSharlow2002}.
+An equivalent of \ortheus\ for RNA reconstruction should be possible,
+representing the intermediate profiles using transducers.
+
+In summary, the evolutionary models and algorithms we have described form a
+systematic theoretical platform on which we can test different optimization and sampling strategies
+for studying the structural evolution of RNA gene families in detail.
+Stochastic grammars are powerful tools for this task; although they will not be the only tools we need,
+particularly as we move towards modeling RNA evolution in greater detail,
+they are flexible and serve our current purposes well.
+Our hope is that these algorithms will allow us, and others, to test our
+knowledge of RNA evolution by computational reconstruction and (eventually)
+direct experimental investigation of early ribonucleic machines.
 

### 2009-05-29 response-to-reviews2.txt
more edits
Revised supplement; Ian's changes to Rob's changes

@@
 main text a self-contained description of the composition algorithm,
-including its key corollaries (elimination of null states, and DP inference).
+including its key corollaries (elimination of null states, and DP
+inference).  The presentation of the composition algorithm in the
+supplement now merely reprises the presentation in the main text using
+more formal notation. We hope this is an appropriate use of the
+supplementary information, but we welcome additional suggestions to
+improve the flow of the manuscript for the reader.
+
 We have attempted to remove redundant text, including the two
@@
 reviewer's recommendation, the full description of this algorithm is
-described in the main text ("Eliminating null cycles in SCFGs") of the
-revised manuscript.  This new algorithm avoids the increased time
-complexity associated with the iterative inference of "loopy DP."
+described in the main text ("Exact elimination of null cycles in
+SCFGs") of the revised manuscript.  This new algorithm avoids the
+increased time complexity associated with the iterative inference of
+"loopy DP."
 
@@
 indiegram webpage),
-http://biowiki.org/IndiegramExamples,
+
+http://biowiki.org/IndiegramExamples
+
 although we note in the text that the reconstructions were too easy
@@
 parse tree X", we use the contrivance of saying "a model that inputs a
-(dummy) empty parse tree, and outputs a parse tree X".  It's annoying
+(dummy) empty parse tree, and outputs a parse tree X".  It's awkward,
 but it keeps the notation straight: it allows us to be consistent
@@
 
-We completely agree and have included an extended discussion of this in
-the revised text.  The primary difficulty is that while this marginalization
-is reasonably straightforward for independent-sites models and HMMs,
-the possible presence of null bifurcations complicates matters for SCFGs.
+We completely agree. Such an algorithm was initially a goal of this
+work, but was abandoned. We thank the reviewer for challenging us to
+attack it afresh; a verbal presentation of the algorithm we came up
+with is given in the revised text, with a full formal description
+(including notation) in the supplement. (Since this is a standalone
+result that can be applied to any SCFG, we think it fits better as an
+appendix in the supplement; the obvious cost is omitting some detail
+from the main paper, but the PLoS Authors' Guidelines and the
+reviewer's preference for including more detail in the manuscript are
+potentially in conflict on the best approach to take here. We welcome
+suggestions for improving the flow generally.)
+
+As a technical aside, the primary difficulty with SCFG null cycle
+elimination is that while this marginalization is reasonably
+straightforward for independent-sites models and HMMs, the possible
+presence of null bifurcations complicates matters for SCFGs.  Indeed,
+no *closed form* elimination is possible (due to bifurcations, such an
+elimination would involve solving simultaneous nonlinear equations, as
+explained in the supplement) but an iterative approximation
+(e.g. Newton-Raphson) appears adequate.
 

### 2009-05-29 paper.tex
Revised supplement; Ian's changes to Rob's changes

@@
   {\Large
-    \textbf{Evolutionary triplet models of structured RNA}
+    \textbf{\titlestring}
   }
 \\
-Robert K. Bradley$^{1}$, 
-Ian Holmes$^{1,2,\ast}$
-\\
-\textbf{1} Biophysics Graduate Group, University of California, Berkeley, CA, USA
-\\
-\textbf{2} Department of Bioengineering, University of California, Berkeley, CA, USA
-\\
-$\ast$ E-mail: indiegram@postbox.biowiki.org
+\authorstring
 \end{flushleft}
@@
 draws on formal models of computational linguistics as well as the
-1985 ``protosequence'' algorithm of David Sankoff.  The output of the
+1985 protosequence algorithm of David Sankoff.  The output of the
 composition algorithm is a multiple-sequence stochastic context-free
@@
 \section*{Author Summary}
-% to do: write this
+
+A number of leading methods for bioinformatics analysis of noncoding RNA
+ use probabilistic grammars as models for pairwise alignments.
+We show that any such pairwise grammar can be extended to an entire phylogeny,
+ by considering the pairwise grammar as a machine (a ``transducer'') that locally transforms one RNA structure into another
+ (thereby modeling a single parent-child relationship in the tree).
+In addition to current applications such as genefinding, homology detection, alignment and secondary structure prediction,
+ this should permit statistical reconstruction of RNA genes ancestral to many present-day sequences.
+We describe algorithms, software implementations, and an elementary benchmark wherein
+ the complexity and accuracy of three-way RNA reconstruction is measured as the length of one branch is varied.
+In the Discussion we consider how the three-way RNA alignment-reconstruction-folding algorithm, which is currently very expensive computationally,
+ might be made more efficient so that larger phylogenies could be considered.
 
@@
 structure \cite{SmithEtAl2008}.  Stochastic context-free grammars
-(SCFGs), which can efficiently model the long-range correlations
-generated by RNA secondary structures, are natural probabilistic models of such
+(SCFGs), which can efficiently detect the long-range correlations
+of RNA base-pairing structures, are natural probabilistic models of such
 phenomena and have been used for ncRNA homology detection
@@
 % Two parts: pairwise & multiple
-By analogy with substitution processes, which are well-understood \cite{Felsenstein2003},
-we may split the ancestral reconstruction problem into two halves.
+By analogy with models of substitution processes, which are well-understood \cite{Felsenstein2003},
+we may take the problem of building phylogenetic models of RNA evolution and split it into two halves.
 The first half is the development of a {\bf pairwise model},
@@
 In substitution processes, the pairwise model is a conditional substitution matrix.
-Often (but not always) the pairwise model is derived from an instantaneous model of change,
-i.e., a continuous-time Markov chain (parametrized by a rate matrix);
-obtaining the transition probabilities of this chain (via exponentiation of the rate matrix)
-yields a pairwise model whose parameters are smoothly-varying functions of a finite time parameter $T$.
+Often (but not always) the pairwise model, representing a finite evolutionary time $T$,
+is derived from an instantaneous model of change over an infinitesimal time interval,
+i.e., a continuous-time Markov chain (parametrized by a rate matrix).
+Obtaining the transition probabilities of this chain (via exponentiation of the rate matrix)
+yields a pairwise model whose parameters are smoothly-varying functions of $T$.
 A pairwise model represents an individual branch of a phylogenetic tree, with $T$ representing the length of that branch.
@@
 % (second part: multiple)
-The second half of the problem involves extending the model (and related inference algorithms) from a single branch to a complete phylogeny,
+The second half of the phylogenetic modeling problem involves extending the model (and related inference algorithms) from a single branch to a complete phylogeny,
 i.e., from a pairwise model of two sequences to a {\bf multiple-sequence model} of many sequences.
@@
 \cite{Nielsen2001} approaches.
-Many of these questions and answers can be unified by an
-appropriate framework;
-for example, another representation of answers A-C
-is that the ``pruning'' and ``peeling'' algorithms (combined) are just
+Another representation of answers A-C
+is that the pruning and peeling algorithms (combined) are just
 the sum-product algorithm on a directed graphical model \cite{Pearl82},
@@
 Felsenstein's pruning algorithm, for example, is essentially blind to the parametric form of the pairwise substitution model;
-it just assumes that a substitution model is provided for every branch.
+it just assumes that a substitution matrix is provided for every branch.
 Subsequent models developed by other researchers can be plugged into the pruning algorithm without modification \cite{Yang94b,WhelanGoldman2001}.
@@
 joint distribution, $P (X, Y | T)$.
-This joint distribution can be factored as $P (X) \cdot P (Y | X, T)$, 
+This joint distribution can be factored, $P (X) \cdot P (Y | X, T)$, 
 where $P(X)$ is the marginal distribution over ancestral sequences
@@
 The conditional distribution $P(Y | X, T)$ then corresponds to a conditional Pair HMM,
-a finite-state machine which transforms one sequence (the input, $X$) into another (the output, $Y$).
+a discrete-state machine which transforms one sequence (the input, $X$) into another (the output, $Y$).
 Following computational linguists, we call this conditionally-normalized state machine a
@@
 they do not emit or absorb any symbols and are required solely for
-``organizational'' purposes (see following section).
+organizational purposes (see following section).
 Two types of states can emit and/or absorb symbols, $\Sinsert$ and $\Smatch$.
@@
 
-As stated above, the Pair SCFG must be ``conditionally normalized''
+As stated above, the Pair SCFG must be conditionally normalized
 so that models can be chained together,
@@
 
-We therefore need an algorithm to efficiently construct the ``state graph'' of the
+We therefore need an algorithm to efficiently construct the state graph of the
 ensemble model, consisting of a list of accessible states and the
@@
 The allowed transitions of the ensemble can be categorized as follows:
-\begin{enumerate}
-\item {\bf Null Transition}:
+\begin{description}
+\item[{\bf Null Transition}:]
   A branch transducer makes a transition into a $\Swait$ state, with no terminal emission or bifurcation.
-\item {\bf Terminal Emission}:
+\item[{\bf Terminal Emission}:]
   A singlet or branch transducer makes a transition into a state of type $\Sinsert$,
@@
   or when the cascade reaches the leaves of the tree.
-\item {\bf Bifurcation}:
+\item[{\bf Bifurcation}:]
   A singlet or branch transducer makes a transition into a state of type $\Sinsert$ that spawns
@@
   which may then be conserved or deleted in the descendant structures.
-\item {\bf End Transition}:
+\item[{\bf End Transition}:]
   The singlet transducer at the root makes a transition to the $\Send$ state,
@@
   terminating the current branch of the parse tree.
-\end{enumerate}
+\end{description}
 
 Co-ordination between the various branch machines is achieved by
-specifying an ordering on the nodes and by having branch transducers ``pause'' in $\Swait$
+specifying an ordering on the nodes and by having branch transducers pause in $\Swait$
 states while waiting to absorb a symbol from the node above.  
@@
 as follows.
-Let the total order on the nodes correspond to a preorder traversal of
-the tree, from which it follows that $m \prec n$ if $m$ is ancestral to $n$.
+Let the total order on the nodes correspond to any preorder traversal of
+the tree; thus, ``$m$ is ancestral to $n$'' is sufficient-but-not-necessary for ``$m \prec n$''.
 Let ${\cal T}_m$ denote the singlet or branch transducer which emits symbols to node $m$.
@@
   states (where $m \prec n$).  ${\cal T}_m$ is free to make any transition.
-\item{Type 2:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer emits a symbol, forcing ${\cal T}_m$ into a $\Smatch$ state so it can absorb that symbol.
+\item{Type 2:} Transducer ${\cal T}_m$ is in a $\Swait$ state.
+Its parent transducer enters a $\Smatch$ or $\Sinsert$ state, emitting a symbol and forcing ${\cal T}_m$ into a $\Smatch$ state so it can absorb that symbol.
 \item{Type 3:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer enters the $\Send$ state, forcing ${\cal T}_m$ into the $\Send$ state as well.
 \end{description}
-These possible transitions of the ensemble generate multiple alignments as follows:
+
+A notational prescription for the allowed transitions may be found in \supptext.
+
+\subsubsection*{How the ensemble generates multiple alignments}
+
+The possible transitions of the ensemble generate multiple alignments as follows:
 \begin{enumerate}
@@
 \item \label{item:InitialWindback}
-  Before any residues can appear at the root, the branch transducers all ``wind back'' into $\Swait$ states,
+  Before any residues can appear at the root, the branch transducers all wind back into $\Swait$ states,
   via type-1 transitions. This occurs in reverse order (i.e.,a postorder traversal of the tree).
@@
   (if any did), followed by transducers on branches two steps away from the root, then three steps, and so on
-  (these can all be regarded as occurring simultaneously, in a single ``wave'' or ``cascade'' of emissions).
+  (these can all be regarded as occurring simultaneously, in a single cascading wave of emissions).
 \item \label{item:Windback}
@@
 
+\subsubsection*{Complexity of the transducer ensemble}
+
 The total sizes of both the state space and the transition matrix are,
@@
 omitting inaccessible state configurations and (2) eliminating
-``windback'' states as described in the following section (which we
+null windback states as described in the following section (which we
 believe will prevent affine gap penalties from generating exponential
@@
 
-Well-known algorithms already exist for maxing or summing over SCFG parse tree likelihoods
-(indeed, this is the motivation for our entire approach).
+Well-known algorithms already exist for maxing or summing over SCFG parse tree likelihoods.
 The Cocke-Younger-Kasami (CYK) algorithm performs maximum-likelihood (ML) inference;
@@
 
-In our composite model, empty paths occur whenever a series of branch transducers ``winds back'' into $\Swait$ states.
+In our composite model, empty paths occur whenever a series of branch transducers winds back into $\Swait$ states.
 Empty bifurcations occur when an entire substructure, present in an ancestor, is deleted in all that ancestor's extant descendants.
@@
 
-Null cycles can be removed systematically from SCFGs by two steps:
+\supptext\ includes a complete formal algorithm for exact null-cycle elimination in SCFGs,
+along with procedures for probabilistically restoring null cycles
+to sampled parse trees and Inside-Outside expectation counts.
+Informally, the essence of the algorithm is contained within the following two steps:
 \begin{description}
@@
 
-We now give a formal description of the null-cycle elimination procedure,
-along with procedures for probabilistically ``restoring'' null cycles
-to sampled parse trees and Inside-Outside expectation counts.
-
-\paragraph{Definitions.}
-
-Let $G = ( \Omega, \gnons, \grules, P )$ be a stochastic context-free grammar (SCFG)
-consisting of
-a set of terminal symbols $\Omega$,
-a set of nonterminal symbols (a.k.a. ``states'') $\gnons$,
-a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
-and a probability function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.
-
-% (The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
-% The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)
-
-Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e., rules in which $A$ appears on the left),
-and let $\rightside{A}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right, including bifurcations which also generate another nonterminal as well as $A$).
-Let $\allrules{A} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
-Define these also on sets of nonterminals, e.g., $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.
-
-Let $\gtrees$ be the set of all parse trees for $G$.
-Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
-then define the parse tree likelihood $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.
-
-If $\sum_{T \in \gtrees} P(T) = 1$, we say that $G$ is {\em probabilistically normalized by parse tree}.
-If $\sum_{\rho \in \leftside{A}} P(\rho) = 1$ for all $A \in \gnons$, we say that $G$ is {\em probabilistically normalized by production rule}.
-Note that normalization by production rule $\Rightarrow$ normalization by parse tree.
-
-Let $S \in \Omega^\ast$ denote a terminal sequence.
-Let $\mathrm{seq}:\gtrees \to \Omega^\ast$ be the function mapping a parse tree to its terminal sequence.
-and let $\mathrm{root}:\gtrees \to \gnons$ be the function returning the root nonterminal of a parse tree.
-The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
-Of particular relevance to null state elimination is the probability $P(\epsilon|A)$,
-where $\epsilon$ is the empty sequence.
-This is the probability that a nonterminal $A$ will expire without generating any sequence.
-
-Following earlier formalisms \cite{Durbin98,HolmesRubin2002a,Holmes2005},
-we say that the grammar $G$ has {\em RNA normal-form rules} if
-each rule in $\grules$ takes one of the following four forms:
-
-\noindent \hrulefill
-\begin{description}
-\item[Termination rules] with one nonterminal on the left and the empty string on the right
-\begin{eqnarray*}
-A & \to & \epsilon
-\end{eqnarray*}
-\item[Transition rules] with one nonterminal on the left and the right
-\begin{eqnarray*}
-A & \to & B
-\end{eqnarray*}
-\item[Bifurcation rules] with one nonterminal on the left and two on the right
-\begin{eqnarray*}
-A & \to & B\ C
-\end{eqnarray*}
-\item[Emission rules] with one nonterminal on the left and right and at least one terminal on the right
-\begin{eqnarray*}
-A & \to & x\ B \\
-A & \to & B\ y \\
-A & \to & x\ B\ y
-\end{eqnarray*}
-\end{description}
-\noindent \hrulefill
-
-\noindent
-Here $A,B,C \in \gnons$ and $x,y \in \Omega$.
-
-Suppose that grammar $G$ has RNA normal-form rules.
-We further say that $G$ has {\em RNA normal-form states} if
-each nonterminal (state) $A \in \gnons$ takes one of the following forms:
-
-\noindent \hrulefill
-\begin{description}
-\item[Null states]: $\leftside{A}$ contains only transition and termination rules.
-\item[Bifurcation states]: $\leftside{A}$ contains exactly one bifurcation rule, $A \to X\ Y$, where $X \neq Y$ and $X,Y$ are both null states.
-\item[Emit states]: $\leftside{A}$ contains only emission rules. Further, $\leftside{A} = \rightside{B}$ for some null state $B$.
-This null state $B$ is called $A$'s {\em post-emit} state.
-\end{description}
-\noindent \hrulefill
-
-Define a {\em null cycle} to be a nonterminal $A$
-and a sequence of rules $\rho_1, \rho_2 \ldots \rho_k$ that, when applied consecutively to $A$, leave $A$ unchanged;
-that is, $\rho_k(\rho_{k-1} \ldots \rho_2(\rho_1(A))) = A$.
-Define a {\em null subtree} to be a null cycle using at least one bifurcation rule.
-
-Suppose that $G = ( \Omega, \gnons, \grules, P )$ and $G' = ( \Omega, \gnons', \grules', P' )$ are two grammars.
-We say that $G$ and $G'$ are {\em equivalent in sequence} if there is a mapping between nonterminals, $f:\gnons \to \gnons'$,
-such that $P(S|A) = P(S|f(A))$.
-We say that $G$ and $G'$ are {\em equivalent in parse} if there is a mapping between parse trees, $g:\gtrees \to \gtrees'$,
-such that $P'(T') = \sum_{T:T'=g(T)} P(T)$ and we can define
-\[
-P(T|T') = P(T|g(T)=T') = \frac{P(T)}{P'(T')}
-\]
-
-Note that equivalence in parse $\Rightarrow$ equivalence in sequence.
-
-Suppose that a grammar $G$ contains null cycles.
-We seek to transform $G$ into a grammar $G''$ that is equivalent in parse and sequence, but has no null cycles.
-If $G$ is normalized by parse tree, then $G''$ will be too; but $G''$ is not (necessarily) normalized by production rule.
-
-We do the transformation in two steps, $G \to G' \to G''$.
-We also provide a stochastic ``null cycle restoration'' algorithm for sampling from $P(T|T'')$.
-
-\paragraph{Eliminating null bifurcations.}
-
-Note first that any SCFG can be transformed into an equivalent one with RNA normal-form states by adding null states.
-Without loss of generality, we therefore consider SCFGs with RNA normal-form states.
-
-Let $G = ( \Omega, \gnons, \grules, P )$ be such an SCFG.
-Let $\gnulls \subseteq \gnons$ be the set of null states in $\gnons$, {\bf excluding} post-emit states.
-Let $\gemits \subset \gnons$ be the set of post-emit states.
-Let $\gbifs \subseteq \gnons$ be the set of bifurcation states.
-
-We here define the grammar $G'$,
-\[
-G' = \left( \Omega, \gnons', \grules', P' \right)
-\]
-which is equivalent to $G$ in sequence but has no null subtrees.
-
-The new states $\gnons'$ are defined as follows.
-Start with $\gnons$; remove bifurcation states;
-then introduce a new state $N'$ for every null $N \in \gnulls$ and three new states $B^0, B', B^2$ for every bifurcation $B \in \gbifs$.
-
-\[
-\gnons' = 
-\left( \bigcup_{N \in \gnulls} \{ N' \} \right) \cup \left( \bigcup_{B \in \gbifs} \{ B^0, B', B^2 \} \right) \cup \gnons \setminus \gbifs
-\]
-
-The idea is that $N'$ (in $G'$) is a null state that is equivalent to null state $N$ (in $G$) for non-empty sequence only.
-That is, for any terminal sequence $S$,
-
-\[
-P'(S|N') = \left\{ \begin{array}{ll} P(S|N) & \mbox{if $S \neq \epsilon$} \\ 0 & \mbox{if $S = \epsilon$} \end{array} \right.
-\]
-
-Similarly, $B'$ is a null state that is equivalent to bifurcation state $B$ for non-empty sequence,
-while $B^2$ is a null state that is is equivalent to $B$ for parse trees where both children of $B$ are non-empty.
-
-In contrast, $B^0$ is {\bf exactly} equivalent to $B$ in sequence.
-However, null subtrees in $B^0$ are explicitly accounted for,
-their probabilities factored into the transitions $B' \to L'$ and $B' \to R'$ and the termination $B^0 \to \epsilon$,
-using the inside probabilities for empty sequences, $P(\epsilon|L)$ and $P(\epsilon|R)$.
-
-The probabilities $P(\epsilon|X)$ are related by the following system of equations
-
-\[
-P(\epsilon|X) = P(X \to \epsilon) + \sum_Y P(X \to Y) P(\epsilon|Y) + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R)
-\]
-
-which are nonlinearly coupled but may be solved numerically, e.g., by the Newton-Raphson method,
-or by iterated approximation starting from a lower bound $P(\epsilon|X) \geq 0$.
-
-The new rules and their probabilities are
-\[
-\begin{array}{rrcl}
-\forall \rho \in (\grules \setminus \allrules{\gbifs}): & P'(\rho)             & = & P(\rho) \\
-                                \forall N \in \gnulls: & P'(N' \to \epsilon)  & = & 0  \\
-                          \ldots \forall B \in \gbifs: & P'(N' \to B')        & = & P(N \to B) \\
-                         \ldots \forall M \in \gnulls: & P'(N' \to M')        & = & P(N \to M) \\
-                         \ldots \forall E \in \gemits: & P'(N' \to E)         & = & P(N \to E) \\
-               \forall (A \to B) \in \rightside{\gbifs}: & P'(A \to B^0)        & = & P(A \to B) \\
-              \forall (B \to LR) \in \leftside{\gbifs}: & P'(B^0 \to \epsilon) & = & P(B \to LR)\ P(\epsilon | L)\ P(\epsilon | R) \\
-                                                       & P'(B^0 \to B')       & = & 1 \\
-                                                       & P'(B' \to \epsilon)  & = & 0 \\
-                                                       & P'(B' \to L')        & = & P(B \to LR)\ P(\epsilon | R) \\
-                                                       & P'(B' \to R')        & = & P(B \to LR)\ P(\epsilon | L) \\
-                                                       & P'(B' \to B^2)       & = & 1 \\
-                                                       & P'(B^2 \to L'\ R')   & = & P(B \to LR) \\
-\end{array}
-\]
-
-Note that grammar $G'$, like $G$, has RNA normal-form states.
-Note also that $G'$ is not, in general, normalized by production rule;
-however, $G'$ is equivalent to $G$ in parse and is therefore normalized by parse tree (if $G$ is).
-
-\paragraph{Eliminating null states.}
-
-We now proceed to eliminate null cycles from $G'$.
-Since null subtrees have been eliminated, the remaining null cycles use only transition rules.
-
-We create the grammar
-\[
-G'' = \left( \Omega, \gnons', \grules'', P'' \right)
-\]
-which has the same nonterminals as $G'$, but different rules and rule probabilities.
-
-Let $\gnulls' \subseteq \gnons'$ be the set of null states in $\gnons'$, {\bf including} post-emit states.
-
-Define ${\bf t}$, the {\em transition matrix} of $G'$, as $t_{XY} = P'(X \to Y)$ for all $X,Y \in \gnons'$.
-The effective transition probability $q_{XY}$ between two states $X,Y$
-sums over all paths through null states (in the summand, $n$ is the length of the null path):
-\[
-{\bf q} = \sum_{n=0}^\infty {\bf t}^n = ({\bf 1}-{\bf t})^{-1}
-\]
-Here ${\bf 1}$ is the identity matrix.
-The matrix inverse may be computed in the usual ways (Gauss-Jordan elimination, LU decomposition, etc.)
-
-Since the bifurcation states of $G'$ explicitly generate non-empty sequence,
-the system of equations relating the probabilities $P'(\epsilon|X)$ is completely linear
-and may be solved in the same way.
-Let $u_X = P'(\epsilon|X)$ and $v_X = P'(X \to \epsilon)$.
-Then
-
-\[
-{\bf u} = {\bf v} + {\bf tu}
-\]
-
-whose solution is ${\bf u} = {\bf qv}$.
-Thus $P'(\epsilon|X) = \sum_Y q_{XY} P'(Y \to \epsilon)$.
-
-We now define $P''$ as follows.
-\begin{itemize}
-\item For all rules $\rho \in \grules'$ that are {\bf not} transitions or terminations, set $P''(\rho) = P'(\rho)$.
-\item For terminations from null states $X \in \gnulls'$, set $P''(X \to \epsilon) = u_X$.
-(NB this may create some terminations $X \to \epsilon$ that were not present in $G'$.)
-\item For transitions from null states $X \in \gnulls'$ to non-null states $Y \notin \gnulls'$, set $P''(X \to Y) = q_{XY}$.
-(NB this may create some transitions $X \to Y$ that were not present in $G'$.)
-\item For transitions {\bf to} null states $X \in \gnulls'$, set $P''(A \to X) = 0$.
-\end{itemize}
-
-Although we have left null states in $G''$, they now have no incoming transitions
-and are inaccessible unless they are post-bifurcation states (i.e.,states which appear on the right-hand side of bifurcation rules), post-emit states, or the start (root) state.
-All other null states can therefore be dropped from $\gnons''$.
-
-\paragraph{Restoring null states.}
-
-Suppose that $\rho''$ is a rule in $G''$.
-Algorithm~\ref{alg:restoreTransitions} samples from the distribution of equivalent parse subtrees in $G'$ (possibly containing null cycles).
-In order to sample from $P(T'|T'')$ we need simply map Algorithm~\ref{alg:restoreTransitions} to each rule in $T''$.
-
-For parameter estimation by Expectation Maximization and some other applications, it is useful to know the expected number
-of times that a transition was used, summed over the posterior distribution of parse trees (including those with null cycles).
-If $n''(\rho'')$ is the expected number of times that transition $\rho''$ was used according to an Inside-Outside computation on $G''$,
-then the corresponding expectations $n'(\rho')$ are given by
-
-\begin{eqnarray*}
-       n'(X \to Y) & = & n''(X \to Y) \frac{P''(X \to Y)}{q_{XY}} + \sum_Z n''(X \to Z) \frac{P''(X \to Y) q_{YZ}}{q_{XZ}} \\
-n'(X \to \epsilon) & = & n''(X \to \epsilon) \frac{P''(X \to \epsilon)}{P''(\epsilon | X)}
- + \sum_W n''(W \to \epsilon) \frac{q_{WX} P''(X \to \epsilon)}{P''(\epsilon | W)}
-\end{eqnarray*}
-
-Expectations for other rules (bifurcations and emissions) are the same for $G'$ as $G''$.
-
-\paragraph{Restoring null bifurcations.}
-
-Suppose that $\rho'$ is a rule in $G'$.
-Algorithm~\ref{alg:restoreBifurcations} samples from the distribution of equivalent parse subtrees in $G$ (possibly containing null subtrees).
-This algorithm also calls Algorithm~\ref{alg:sampleNullSubtree}, which samples from the distribution of empty subtrees rooted at a particular nonterminal.
-In order to sample from $P(T|T')$ we need simply map Algorithm~\ref{alg:restoreBifurcations} to each rule in $T'$.
-
-If $n'(\rho')$ is the expected number of times that rule $\rho'$ was used by $G'$,
-then the corresponding expectations $n(\rho)$ are given by
-
-\begin{eqnarray*}
-    n(B \to L\ R) & = & n'(B' \to L') + n'(B' \to R') + n'(B^0 \to L'\ R') + d(B \to L\ R) \\
-       n(X \to Y) & = & n'(X \to Y) + d(X \to Y) \\
-n(X \to \epsilon) & = & n'(X \to \epsilon) + d(X \to \epsilon)
-\end{eqnarray*}
-Expectations for emissions are the same for $G$ as $G'$.
-In the above expressions $d(\rho)$ is the expected usage of rule $\rho$ by null subtrees:
-
-\[
-d(\rho) = \sum_{(B \to LR) \in \grules} \left[
-n'(B' \to L') c_R(\rho)
-+ n'(B' \to R') c_L(\rho)
-+ n'(B^0 \to \epsilon) (c_L(\rho) + c_R(\rho))
-\right]
-\]
-
-where $c_W(\rho)$ is the expected usage of rule $\rho$ by an empty parse tree rooted at $W$,
-given by
-
-\begin{eqnarray*}
-c_X(\rho) P(\epsilon | X) & = & P(X \to \epsilon) \delta_{\rho=(X \to \epsilon)}
- + \sum_Y P(X \to Y) P(\epsilon|Y) \left( c_Y(\rho) + \delta_{\rho=(X \to Y)} \right)
-\\ & &
- + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R) \left( c_L(\rho) + c_R(\rho) + \delta_{\rho=(X \to LR)} \right)
-\end{eqnarray*}
-
-where $\delta_U$ is the Kronecker delta (1 if condition $U$ is true, 0 if it is false).
-Note that in contrast to the system of equations for $P(\epsilon|X)$, this is a linear system of equations
-of the form ${\bf c} = {\bf Mc} + {\bf k}$, which may be solved by matrix inversion:
-${\bf c} = ({\bf 1}-{\bf M})^{-1} {\bf k}$.
-
 \subsubsection*{Dynamic programming algorithms for inference}
@@
 model-construction algorithm, we can compute likelihoods and sample parse trees
-using the ``standard'' CYK, Inside and Outside algorithms for multiple-sequence SCFGs \cite{Sankoff85,Durbin98}.
+using the standard CYK, Inside and Outside algorithms for multiple-sequence SCFGs \cite{Sankoff85,Durbin98}.
 
@@
 bifurcations.
-Note that $A$ and $B$ are also dependent on $N$ (see ``The TKFST model
-on a three-taxon phylogeny'').
+Note that $A$ and $B$ are also dependent on $N$ (see \secname{The TKFST model on a three-taxon phylogeny}).
 
@@
 \beqn \eqnlabel{foldenv}
-\foldenv^{(X)} \subset \left\{ (i,j):0 \le i \le j \le L^{(X)} \right\} \, ;
+\foldenv^{(X)} \subseteq \left\{ (i,j):0 \le i \le j \le L^{(X)} \right\} \, ;
 \quad L^{(X)} = |X| \, .
@@
 
-The Outside algorithm uses an outside-inside ordering, where subsequences starting at a fixed $i$ are reverse-sorted 
-with respect to the inside-outside ordering described above.  Subsequences in $\foldenv^{(X)}$ are then sorted 
+The Outside algorithm uses an outside-inside ordering, which is the exact reverse of the inside-outside ordering described above.
+Subsequences in $\foldenv^{(X)}$ are sorted 
 in the same order as coordinate pairs $(i,j)$ are generated by the 
 iteration $\{\,\mathrm{for}\, i = 0 \,\mathrm{to}\, L^{(X)} \,\,\{ \mathrm{for}\, j = L^{(X)}\, \mathrm{to}\, i \}\,\}$.
+
 We frequently refer to subsequences by their index in the fold envelope.
@@
 
-Note that we are assuming that the transformations described in ``Eliminating null states'' have been performed, such that there are no cycles of $\Snull$ states as well as no empty bifurcations.
+Note that we are assuming that the transformations described in \secname{Exact elimination of null cycles in SCFGs} have been performed, such that there are no cycles of $\Snull$ states as well as no empty bifurcations.
 
@@
 and (2) structure prediction of homologous structured RNAs and
-comparing its performance to one of the best-performing pairwise
-SCFGs for RNA multiple alignment \cite{BradleyPachterHolmes2008}.  The
+comparing its performance to \stemloc\ (one of the better-performing pairwise
+SCFGs used for RNA multiple alignment \cite{BradleyPachterHolmes2008}).  The
 results of these benchmarks, reported in \tabref{bralibaseII-align}
@@
 fundamental to our approach and can be replaced by a different
-and more realistic pairwise model, such as the pairwise SCFG used in 
+and more realistic pairwise model, such as the \stemloc\ pairwise SCFG used in 
 these comparisons \cite{BradleyPachterHolmes2008}.
@@
 other comparisons of SCFGs at structure prediction,
-such as the study by \cite{DowellEddy2004}.
+such as the study of \cite{DowellEddy2004}.
 
@@
 with the exponential calculation).  After performing the
-transformations described in ``Eliminating null cycles in SCFGs'' to
+transformations described in \secname{Exact elimination of null cycles in SCFGs} to
 eliminate useless windback states, the ensemble model has a reduced
@@
 3) \Rightarrow 441$ states and $O(a \cdot b^2 \cdot N) = O(7 \cdot
-21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in ``The
-composition algorithm.''  Nonetheless, the extreme complexity of the
+21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in \secname{The composition algorithm}.
+ Nonetheless, the extreme complexity of the
 ensemble model, despite the simplicity of the underlying model of RNA
@@
 model constructed by the search algorithm and Dataset S2 the reduced
-model after eliminating ``windback'' states; both are in Graphviz format for
+model after eliminating windback states; both are in Graphviz format for
 visualization and show the state of the singlet transducer generating
@@
 The inference algorithms in \indiegram\ could be further constrained
-to enforce, for example, a fixed multiple alignment or consensus
-structure of extant sequences.  While experimentally-determined
+to enforce, for example, a fixed multiple alignment or a consensus
+structure for extant sequences.  While experimentally-determined
 structures of individual RNAs are relatively rare, curated deep sequence
@@
 time complexity of the algorithms could be dramatically reduced.  Such
-constraints can be naturally expressed with ``alignment envelopes,''
-the sequence-level counterpart of fold envelopes \cite{Holmes2005}.
+constraints can be naturally expressed with {\em alignment envelopes},'
+the alignment-space analogue of fold envelopes \cite{Holmes2005}.
 However, in this paper we focus on model construction and inference
@@
 length of the outgroup increases, the outgroup's structure becomes increasingly decoupled from
-that of its ancestor, allowing the algorithm to ``ignore'' the
+that of its ancestor, allowing the scoring scheme to ``ignore'' the
 information which it contains.
@@
 we can perfectly reconstruct the ancestral structure (as expected),
-but we cannot perfectly reconstruct the three-way multiple alignment (since there are still two finite-length branches, on which some unreconstructable mutations inevitably occur).
+but we cannot perfectly reconstruct the three-way multiple alignment
+ (since there are still two finite-length branches,
+ which on average contribute a greater-than-zero rate of unreconstructable mutations).
 
@@
 
-The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
-Another script (\animateevolsayer) can be used to make animations of the evolving RNA structures, using the RNAplot program in the \viennarna\ package \cite{HofackerEtAl94}.
-
 \newpage
@@
 Following the conception of paleogenetics \cite{PaulingZuckerkandl63},
-a large number of synthetic reconstructions of ancient proteins have been
+a large number of synthetic reconstructions of ancestral protein sequences have been
 reported in the literature
 \cite{MalcolmEtAl90,StackhouseEtAl90,ZhangRosenberg2002,GaucherEtAl2003,ThomsonEtAl2005,ChangEtAl2002,SunEtAl2002,OrtlundEtAl2007}.
-There are also the beginnings of a movement to reconstruct ancient DNA
+There is also scientific interest in reconstructing DNA sequences
 \cite{IvicsEtAl91,AdeyEtAl94,BlanchetteEtAl2004,NoonanEtAl2006,Gaucher2007b,EliasTuller2007,PatenHolmesBirney2008}.
@@
 The work reported in this paper builds on extensive prior art in the areas of evolutionary modeling and ancestral reconstruction.
+Reviewing all of this would take several books, but we can note some key references.
 The reconstruction of ancient sequences was first proposed in 1963 by Pauling and Zuckerkandl \cite{PaulingZuckerkandl63};
@@
 In particular, an early example of a pairwise conditional model $P(Y|X)$ for structure-dependent RNA evolution
-may be found in \cite{KleinEddy2003}.
+was given by Eddy {\em et al} \cite{KleinEddy2003}.
 A conditional framework similar to ours in some respects is described
-by Sakakibara et al \cite{Sakakibara2003}.
-The dynamic programming inference algorithms for multiple-sequence SCFGs are closely related to the ``protosequence algorithm'' of Sankoff \cite{Sankoff85}.
+by Sakakibara {\em et al} \cite{Sakakibara2003}.
+The dynamic programming inference algorithms for multiple-sequence SCFGs are closely related to the protosequence algorithm of Sankoff \cite{Sankoff85}.
 %a seminal work in both RNA structural alignment and ancestral reconstruction.
@@
 although it has several advantages (e.g. explicit treatment of pseudoknots).
-Of course, MCMC kernels can often be adapted for use with different models, and the MCMC moves used by SimulFold may be useful for inference under our model as well.
+Of course, MCMC kernels are inherently adaptable to other purposes: the MCMC moves developed for SimulFold may be useful for inference under different models.
 
 While MCMC provides the most information about the posterior distribution of evolutionary histories,
-in practice a maximum likelihood inference may be adequate (and typically faster).
+in practice a maximum likelihood inference may be adequate (and typically much faster).
 The progressive profiling used by the \ortheus\ program for
@@
 for studying the structural evolution of RNA gene families in detail.
-Stochastic grammars are powerful tools for this task; although they will not be the only tools we need,
-particularly as we move towards modeling RNA evolution in greater detail,
-they are flexible and serve our current purposes well.
-Our hope is that these algorithms will allow us, and others, to test our
-knowledge of RNA evolution by computational reconstruction and (eventually)
+Stochastic grammars are powerful tools for this task, although they will not be the only tools we need,
+particularly as we move towards modeling RNA evolution in greater detail.
+Our hope is that these algorithms will allow us to test and refine our
+understanding of RNA evolution by computational reconstruction and (eventually)
 direct experimental investigation of early ribonucleic machines.
@@
     ($S$ and $I_S$) sequence evolve as TKF sequences with length
-    parameters $\kappa_l$ and $\kappa_s$ (defined in ``A simple model
-    of RNA structural evolution'').  $\pi_l(u)$ and $\pi_s(uv)$ are
+    parameters $\kappa_l$ and $\kappa_s$ (defined in \secname{A simple model
+    of RNA structural evolution}).  $\pi_l(u)$ and $\pi_s(uv)$ are
     the equilibrium distributions of unpaired nucleotides $u$ and
@@
     $\gamma_{l,s}(t)$ are parametrized by the insertion and deletion rates
-    of the TKFST model and are defined in ``A simple model of RNA
-    structural evolution.''
+    of the TKFST model and are defined in \secname{A simple model of RNA
+    structural evolution}.
   \end{flushleft}
@@
   We compared the performance of the TKFST model for progressive multiple alignment
-  of RNAs against the performance of a grammar with a richer model of RNA structure (``\stemloc''; \cite{Holmes2005}).
+  of RNAs against the performance of a grammar with a richer model of RNA structure (\stemloc\ \cite{Holmes2005}).
   Sensitivity is defined as $\mbox{TP}/(\mbox{TP} + \mbox{FN})$ and PPV is defined as $\mbox{TP}/(\mbox{TP} + \mbox{FP})$,
@@
     progressive multiple alignment of RNAs against the performance of a grammar with 
-    a richer model of RNA structure (``\stemloc''; \cite{Holmes2005}).
+    a richer model of RNA structure (\stemloc\ \cite{Holmes2005}).
     ``g2intron'' is the RFAM entry {\tt Intron\_gpII}, containing domains V and VI of the Group II intron.

### 2009-05-29 supplementary.tex
Revised supplement; Ian's changes to Rob's changes

@@
 \input{defs.tex}
+% no idea why \outside was redefined with a "\right," instead of a normal comma in defs.tex, but am redefining it back here - IH, 5/29/09
+\renewcommand{\outside}[7]{\beta_{#1}\left(#2,#3,#4\right)}
 %% END MACROS SECTION
@@
   {\Large
-    \textbf{Evolutionary models of structured RNA: Text S1}
+    \textbf{\titlestring: \supptext}
   }
 \\
-Robert K. Bradley$^{1}$, 
-Ian Holmes$^{1,2,\ast}$
-\\
-\textbf{1} Biophysics Graduate Group, University of California, Berkeley, CA, USA
-\\
-\textbf{2} Department of Bioengineering, University of California, Berkeley, CA, USA
-\\
-$\ast$ E-mail: indiegram@postbox.biowiki.org
+\authorstring
 \end{flushleft}
@@
 \tableofcontents
+
+\newpage
+\section{Exact elimination of SCFG null cycles}
+
+The following section describes how to transform any SCFG so as to remove null cycles while preserving access to the full posterior probability distribution over parses, including parses with null cycles.
+
+\subsection{Definitions}
+
+Let $G = ( \Omega, \gnons, \grules, P )$ be a stochastic context-free grammar (SCFG)
+consisting of
+a set of terminal symbols $\Omega$,
+a set of nonterminal symbols (a.k.a. ``states'') $\gnons$,
+a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
+and a probability function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.
+
+% (The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
+% The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)
+
+Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e., rules in which $A$ appears on the left),
+and let $\rightside{A}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right, including bifurcations which also generate another nonterminal as well as $A$).
+Let $\allrules{A} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
+Define these also on sets of nonterminals, e.g., $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.
+
+Let $\gtrees$ be the set of all parse trees for $G$.
+Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
+then define the parse tree likelihood $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.
+
+If $\sum_{T \in \gtrees} P(T) = 1$, we say that $G$ is {\em probabilistically normalized by parse tree}.
+If $\sum_{\rho \in \leftside{A}} P(\rho) = 1$ for all $A \in \gnons$, we say that $G$ is {\em probabilistically normalized by production rule}.
+Note that normalization by production rule $\Rightarrow$ normalization by parse tree.
+
+Let $S \in \Omega^\ast$ denote a terminal sequence.
+Let $\mathrm{seq}:\gtrees \to \Omega^\ast$ be the function mapping a parse tree to its terminal sequence.
+and let $\mathrm{root}:\gtrees \to \gnons$ be the function returning the root nonterminal of a parse tree.
+The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
+Of particular relevance to null state elimination is the probability $P(\epsilon|A)$,
+where $\epsilon$ is the empty sequence.
+This is the probability that a nonterminal $A$ will expire without generating any sequence.
+
+Following earlier formalisms \cite{Durbin98,HolmesRubin2002a,Holmes2005},
+we say that the grammar $G$ has {\em RNA normal-form rules} if
+each rule in $\grules$ takes one of the following four forms:
+
+\noindent \hrulefill
+\begin{description}
+\item[Termination rules] with one nonterminal on the left and the empty string on the right
+\begin{eqnarray*}
+A & \to & \epsilon
+\end{eqnarray*}
+\item[Transition rules] with one nonterminal on the left and the right
+\begin{eqnarray*}
+A & \to & B
+\end{eqnarray*}
+\item[Bifurcation rules] with one nonterminal on the left and two on the right
+\begin{eqnarray*}
+A & \to & B\ C
+\end{eqnarray*}
+\item[Emission rules] with one nonterminal on the left and right and at least one terminal on the right
+\begin{eqnarray*}
+A & \to & x\ B \\
+A & \to & B\ y \\
+A & \to & x\ B\ y
+\end{eqnarray*}
+\end{description}
+\noindent \hrulefill
+
+\noindent
+Here $A,B,C \in \gnons$ and $x,y \in \Omega$.
+
+Suppose that grammar $G$ has RNA normal-form rules.
+We further say that $G$ has {\em RNA normal-form states} if
+each nonterminal (state) $A \in \gnons$ takes one of the following forms:
+
+\noindent \hrulefill
+\begin{description}
+\item[Null states]: $\leftside{A}$ contains only transition and termination rules.
+\item[Bifurcation states]: $\leftside{A}$ contains exactly one bifurcation rule, $A \to X\ Y$, where $X \neq Y$ and $X,Y$ are both null states.
+\item[Emit states]: $\leftside{A}$ contains only emission rules. Further, $\leftside{A} = \rightside{B}$ for some null state $B$.
+This null state $B$ is called $A$'s {\em post-emit} state.
+\end{description}
+\noindent \hrulefill
+
+Define a {\em null cycle} to be a nonterminal $A$
+and a sequence of rules $\rho_1, \rho_2 \ldots \rho_k$ that, when applied consecutively to $A$, leave $A$ unchanged;
+that is, $\rho_k(\rho_{k-1} \ldots \rho_2(\rho_1(A))) = A$.
+Define a {\em null subtree} to be a null cycle using at least one bifurcation rule.
+
+Suppose that $G = ( \Omega, \gnons, \grules, P )$ and $G' = ( \Omega, \gnons', \grules', P' )$ are two grammars.
+We say that $G$ and $G'$ are {\em equivalent in sequence} if there is a mapping between nonterminals, $f:\gnons \to \gnons'$,
+such that $P(S|A) = P(S|f(A))$.
+We say that $G$ and $G'$ are {\em equivalent in parse} if there is a mapping between parse trees, $g:\gtrees \to \gtrees'$,
+such that $P'(T') = \sum_{T:T'=g(T)} P(T)$ and we can define
+\[
+P(T|T') = P(T|g(T)=T') = \frac{P(T)}{P'(T')}
+\]
+
+Note that equivalence in parse $\Rightarrow$ equivalence in sequence.
+
+Suppose that a grammar $G$ contains null cycles.
+We seek to transform $G$ into a grammar $G''$ that is equivalent in parse and sequence, but has no null cycles.
+If $G$ is normalized by parse tree, then $G''$ will be too; but $G''$ is not (necessarily) normalized by production rule.
+
+We do the transformation in two steps, $G \to G' \to G''$.
+We also provide a stochastic ``null cycle restoration'' algorithm for sampling from $P(T|T'')$.
+
+\subsection{Eliminating null bifurcations}
+
+Note first that any SCFG can be transformed into an equivalent one with RNA normal-form states by adding null states.
+Without loss of generality, we therefore consider SCFGs with RNA normal-form states.
+
+Let $G = ( \Omega, \gnons, \grules, P )$ be such an SCFG.
+Let $\gnulls \subseteq \gnons$ be the set of null states in $\gnons$, {\bf excluding} post-emit states.
+Let $\gemits \subset \gnons$ be the set of post-emit states.
+Let $\gbifs \subseteq \gnons$ be the set of bifurcation states.
+
+We here define the grammar $G'$,
+\[
+G' = \left( \Omega, \gnons', \grules', P' \right)
+\]
+which is equivalent to $G$ in sequence but has no null subtrees.
+
+The new states $\gnons'$ are defined as follows.
+Start with $\gnons$; remove bifurcation states;
+then introduce a new state $N'$ for every null $N \in \gnulls$ and three new states $B^0, B', B^2$ for every bifurcation $B \in \gbifs$.
+
+\[
+\gnons' = 
+\left( \bigcup_{N \in \gnulls} \{ N' \} \right) \cup \left( \bigcup_{B \in \gbifs} \{ B^0, B', B^2 \} \right) \cup \gnons \setminus \gbifs
+\]
+
+The idea is that $N'$ (in $G'$) is a null state that is equivalent to null state $N$ (in $G$) for non-empty sequence only.
+That is, for any terminal sequence $S$,
+
+\[
+P'(S|N') = \left\{ \begin{array}{ll} P(S|N) & \mbox{if $S \neq \epsilon$} \\ 0 & \mbox{if $S = \epsilon$} \end{array} \right.
+\]
+
+Similarly, $B'$ is a null state that is equivalent to bifurcation state $B$ for non-empty sequence,
+while $B^2$ is a null state that is is equivalent to $B$ for parse trees where both children of $B$ are non-empty.
+
+In contrast, $B^0$ is {\bf exactly} equivalent to $B$ in sequence.
+However, null subtrees in $B^0$ are explicitly accounted for,
+their probabilities factored into the transitions $B' \to L'$ and $B' \to R'$ and the termination $B^0 \to \epsilon$,
+using the inside probabilities for empty sequences, $P(\epsilon|L)$ and $P(\epsilon|R)$.
+
+The probabilities $P(\epsilon|X)$ are related by the following system of equations
+
+\[
+P(\epsilon|X) = P(X \to \epsilon) + \sum_Y P(X \to Y) P(\epsilon|Y) + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R)
+\]
+
+which are nonlinearly coupled (via bifurcations) but may be solved numerically, e.g., by the Newton-Raphson method,
+or by iterated approximation starting from a lower bound $P(\epsilon|X) \geq 0$.
+
+The new rules and their probabilities are
+\[
+\begin{array}{rrcl}
+\forall \rho \in (\grules \setminus \allrules{\gbifs}): & P'(\rho)             & = & P(\rho) \\
+                                \forall N \in \gnulls: & P'(N' \to \epsilon)  & = & 0  \\
+                          \ldots \forall B \in \gbifs: & P'(N' \to B')        & = & P(N \to B) \\
+                         \ldots \forall M \in \gnulls: & P'(N' \to M')        & = & P(N \to M) \\
+                         \ldots \forall E \in \gemits: & P'(N' \to E)         & = & P(N \to E) \\
+               \forall (A \to B) \in \rightside{\gbifs}: & P'(A \to B^0)        & = & P(A \to B) \\
+              \forall (B \to LR) \in \leftside{\gbifs}: & P'(B^0 \to \epsilon) & = & P(B \to LR)\ P(\epsilon | L)\ P(\epsilon | R) \\
+                                                       & P'(B^0 \to B')       & = & 1 \\
+                                                       & P'(B' \to \epsilon)  & = & 0 \\
+                                                       & P'(B' \to L')        & = & P(B \to LR)\ P(\epsilon | R) \\
+                                                       & P'(B' \to R')        & = & P(B \to LR)\ P(\epsilon | L) \\
+                                                       & P'(B' \to B^2)       & = & 1 \\
+                                                       & P'(B^2 \to L'\ R')   & = & P(B \to LR) \\
+\end{array}
+\]
+
+Note that grammar $G'$, like $G$, has RNA normal-form states.
+Note also that $G'$ is not, in general, normalized by production rule;
+however, $G'$ is equivalent to $G$ in parse and is therefore normalized by parse tree (if $G$ is).
+
+\subsection{Eliminating null states}
+
+We now proceed to eliminate null cycles from $G'$.
+Since null subtrees have been eliminated, the remaining null cycles use only transition rules.
+
+We create the grammar
+\[
+G'' = \left( \Omega, \gnons', \grules'', P'' \right)
+\]
+which has the same nonterminals as $G'$, but different rules and rule probabilities.
+
+Let $\gnulls' \subseteq \gnons'$ be the set of null states in $\gnons'$, {\bf including} post-emit states.
+
+Define ${\bf t}$, the {\em transition matrix} of $G'$, as $t_{XY} = P'(X \to Y)$ for all $X,Y \in \gnons'$.
+The effective transition probability $q_{XY}$ between two states $X,Y$
+sums over all paths through null states (in the summand, $n$ is the length of the null path):
+\[
+{\bf q} = \sum_{n=0}^\infty {\bf t}^n = ({\bf 1}-{\bf t})^{-1}
+\]
+Here ${\bf 1}$ is the identity matrix.
+The matrix inverse may be computed in the usual ways (Gauss-Jordan elimination, LU decomposition, etc.)
+
+Since the bifurcation states of $G'$ explicitly generate non-empty sequence,
+the system of equations relating the probabilities $P'(\epsilon|X)$ is completely linear
+and may be solved in the same way.
+Let $u_X = P'(\epsilon|X)$ and $v_X = P'(X \to \epsilon)$.
+Then
+
+\[
+{\bf u} = {\bf v} + {\bf tu}
+\]
+
+whose solution is ${\bf u} = {\bf qv}$.
+Thus $P'(\epsilon|X) = \sum_Y q_{XY} P'(Y \to \epsilon)$.
+
+We now define $P''$ as follows.
+\begin{itemize}
+\item For all rules $\rho \in \grules'$ that are {\bf not} transitions or terminations, set $P''(\rho) = P'(\rho)$.
+\item For terminations from null states $X \in \gnulls'$, set $P''(X \to \epsilon) = u_X$.
+(NB this may create some terminations $X \to \epsilon$ that were not present in $G'$.)
+\item For transitions from null states $X \in \gnulls'$ to non-null states $Y \notin \gnulls'$, set $P''(X \to Y) = q_{XY}$.
+(NB this may create some transitions $X \to Y$ that were not present in $G'$.)
+\item For transitions {\bf to} null states $X \in \gnulls'$, set $P''(A \to X) = 0$.
+\end{itemize}
+
+Although we have left null states in $G''$, they now have no incoming transitions
+and are inaccessible unless they are post-bifurcation states (i.e.,states which appear on the right-hand side of bifurcation rules), post-emit states, or the start (root) state.
+All other null states can therefore be dropped from $\gnons''$.
+
+\subsection{Restoring null states}
+
+Suppose that $\rho''$ is a rule in $G''$.
+Algorithm~\ref{alg:restoreTransitions} samples from the distribution of equivalent parse subtrees in $G'$ (possibly containing null cycles).
+In order to sample from $P(T'|T'')$ we need simply map Algorithm~\ref{alg:restoreTransitions} to each rule in $T''$.
+
+For parameter estimation by Expectation Maximization and some other applications, it is useful to know the expected number
+of times that a transition was used, summed over the posterior distribution of parse trees (including those with null cycles).
+If $n''(\rho'')$ is the expected number of times that transition $\rho''$ was used according to an Inside-Outside computation on $G''$,
+then the corresponding expectations $n'(\rho')$ are given by
+
+\begin{eqnarray*}
+       n'(X \to Y) & = & n''(X \to Y) \frac{P''(X \to Y)}{q_{XY}} + \sum_Z n''(X \to Z) \frac{P''(X \to Y) q_{YZ}}{q_{XZ}} \\
+n'(X \to \epsilon) & = & n''(X \to \epsilon) \frac{P''(X \to \epsilon)}{P''(\epsilon | X)}
+ + \sum_W n''(W \to \epsilon) \frac{q_{WX} P''(X \to \epsilon)}{P''(\epsilon | W)}
+\end{eqnarray*}
+
+Expectations for other rules (bifurcations and emissions) are the same for $G'$ as $G''$.
+
+\subsection{Restoring null bifurcations}
+
+Suppose that $\rho'$ is a rule in $G'$.
+Algorithm~\ref{alg:restoreBifurcations} samples from the distribution of equivalent parse subtrees in $G$ (possibly containing null subtrees).
+This algorithm also calls Algorithm~\ref{alg:sampleNullSubtree}, which samples from the distribution of empty subtrees rooted at a particular nonterminal.
+In order to sample from $P(T|T')$ we need simply map Algorithm~\ref{alg:restoreBifurcations} to each rule in $T'$.
+
+If $n'(\rho')$ is the expected number of times that rule $\rho'$ was used by $G'$,
+then the corresponding expectations $n(\rho)$ are given by
+
+\begin{eqnarray*}
+    n(B \to L\ R) & = & n'(B' \to L') + n'(B' \to R') + n'(B^0 \to L'\ R') + d(B \to L\ R) \\
+       n(X \to Y) & = & n'(X \to Y) + d(X \to Y) \\
+n(X \to \epsilon) & = & n'(X \to \epsilon) + d(X \to \epsilon)
+\end{eqnarray*}
+Expectations for emissions are the same for $G$ as $G'$.
+In the above expressions $d(\rho)$ is the expected usage of rule $\rho$ by null subtrees:
+
+\[
+d(\rho) = \sum_{(B \to LR) \in \grules} \left[
+n'(B' \to L') c_R(\rho)
++ n'(B' \to R') c_L(\rho)
++ n'(B^0 \to \epsilon) (c_L(\rho) + c_R(\rho))
+\right]
+\]
+
+where $c_W(\rho)$ is the expected usage of rule $\rho$ by an empty parse tree rooted at $W$,
+given by
+
+\begin{eqnarray*}
+c_X(\rho) P(\epsilon | X) & = & P(X \to \epsilon) \delta_{\rho=(X \to \epsilon)}
+ + \sum_Y P(X \to Y) P(\epsilon|Y) \left( c_Y(\rho) + \delta_{\rho=(X \to Y)} \right)
+\\ & &
+ + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R) \left( c_L(\rho) + c_R(\rho) + \delta_{\rho=(X \to LR)} \right)
+\end{eqnarray*}
+
+where $\delta_U$ is the Kronecker delta (1 if condition $U$ is true, 0 if it is false).
+Note that in contrast to the system of equations for $P(\epsilon|X)$, this is a linear system of equations
+of the form ${\bf c} = {\bf Mc} + {\bf k}$, which may be solved by matrix inversion:
+${\bf c} = ({\bf 1}-{\bf M})^{-1} {\bf k}$.
 
@@
 
-\subsection{Formal Grammars} \seclabel{grammars}
+\subsection{Formal grammars} \seclabel{grammars}
 
@@
 but this is generically not true for a more general grammar
-(for example, see the TKF91 Structure Tree model).
+(for example, see the TKF Structure Tree model).
 
@@
 \newpage
-\section{Inference with multi-sequence SCFGs}
-
-The model construction algorithm described above creates a composed state machine
-which generates the joint distribution $P(X, Y, Z, W)$.
-As described in \secref{multigrammars}, there exists a corresponding multi-sequence SCFG
-which generates the same probability distribution.
-It follows that we can perform inference on our composed model with the well-known Sankoff
-algorithms.
-
-As described in the paper, we can use the CYK algorithm directly, but
-the Inside and Outside algorithms fail for generic multi-sequence models
-due to the presence of looping $\Snull$ transitions.  We here give algorithms for inference 
-in the case of no looping $\Snull$ transitions and explain how to modify them to 
-create {\bf loopy DP} iterative algorithms for our models.
-
-We here give versions for the model of three extant sequences
-show in \figref{threeway}; the generalization to $N$ sequences is straightforward.
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics [scale=0.4] {figs/threeway.pdf}
-  \caption{\figlabel{threeway}
-    A star phylogeny with three (extant) leaf sequences.
-    An ancestral sequence $W$ evolves into three descendant sequences $X$, $Y$ and $Z$.
-    A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
-    and three branch transducers (the gray boxes labeled $\Delta T_X$, $\Delta T_Y$ and $\Delta T_Z$)
-    mutate it according to the specified multi-sequence model.
-    Gray nodes correspond to observed data and white nodes unobserved data.
-    If the branch transducers are time-reversible, then this star phylogeny with three leaves is
-    the neighborhood of any interior node in a (binary) phylogenetic tree, 
-    from which it follows that evaluating the likelihood function
-    on this star phylogeny is sufficient for a sampling algorithm on an arbitrary phylogeny.
-  }
-\end{figure}
-
-\subsection{Notation}
-The 3 (ungapped) sequences to be aligned are $x^{(1)}$, $x^{(2)}$ and $x^{(3)}$
-(we refer to them by an index rather than as $X$, $Y$ and $Z$ for notational convenience).
-The $\mathrm{i}^{\mathrm{th}}$ symbol of sequence $x^{(1)}$ is written $x^{(1)}_i$ 
-and the length of the same sequence is $L_1 = |x^{(1)}|$.
-
-\subsection{Fold Envelopes} \seclabel{foldenv}
-We use the fold envelope concept \cite{HolmesRubin2002a,Holmes2004} to constrain the set of structures our algorithms consider.
-A fold envelope $\foldenv$ for a sequence $x$ is a set of coordinate pairs satisfying
-\beqn \label{eqn:cabbage}
-\foldenv \subset \left\{ (i,j):0 \le i \le j \le L \right\} \, .
-\eeqn
-We consider a subsequence $x_{i+1} \dots x_{j}$ only if the corresponding coordinate pair $(i,j) \in \foldenv$.
-The unconstrained fold envelope has set equality in \eqref{eqn:cabbage}.
-
-We use two orderings for sequences in the fold envelopes.  An inside-outside ordering is used for the iteration 
-in the Inside algorithm: Subsequences are ordered such that each successive subsequence 
-contains all previous subsequences in the fold envelope.  More precisely, subsequences in $\mathscr{F}$ are sorted  
-in the same order as coordinate pairs $(i,j)$ are generated by the iteration $\{\,\mathrm{for}\, i = L \,\mathrm{to}\, 0 \,\,\{ \mathrm{for}\, j = i \,\mathrm{to}\, L \}\,\}$.
-
-The Outside algorithm uses an outside-inside ordering, where subsequences starting at a fixed $i$ are reverse-sorted 
-with respect to the inside-outside ordering described above.  Subsequences in $\mathscr{F}$ are then sorted 
-in the same order as coordinate pairs $(i,j)$ are generated by the 
-iteration $\{\,\mathrm{for}\, i = 0 \,\mathrm{to}\, L \,\,\{ \mathrm{for}\, j = L\, \mathrm{to}\, i \}\,\}$.
-
-We frequently refer to subsequences by their index in the fold envelope.
-The $\mth$ subsequence in $\foldenv$ is labeled $m$ and corresponds to the coordinate pair $(i_{m}, j_{m})$.  The 
-index of a pair $(i,j) \in \foldenv$ is $n[i,j]$.
-
-Inward and outward emission connections, specifying which subsequences are reachable from 
-a given subsequence and state, are defined as
-\begin{align}
-c_{in}\left(\bvec{b}; m\right) &= n\left[ i_{m} + \Delta_{\bvec{b}}^{L}, j_{m} - \Delta_{\bvec{b}}^{R} \right] \\
-c_{out}\left(\bvec{b}; m\right) &= n\left[ i_{m} - \Delta_{\bvec{b}}^{L}, j_{m} + \Delta_{\bvec{b}}^{R} \right]
-\end{align}
-The emission connection is undefined if the corresponding subsequence is not in the fold envelope.\footnote{We defined emission connections slightly differently than do \cite{HolmesRubin2002a}.}  Inward, outward-left and outward-right bifurcation connections are defined as
-\begin{align}
-b_{in}(n) &= \left\{ \left( n_L,n_R \right): i_{n_L} = i_{n},\, j_{n_L} = i_{n_R},\, j_{n_R} = j_{n} \right\} \\
-b_{out,L}(n) &= \left\{ \left( n_O,n_L \right): i_{n_L} = i_{n_O},\, j_{n_L} = i_{n},\, j_{n} = j_{n_O} \right\} \\
-b_{out,R}(n) &= \left\{ \left( n_O,n_R \right): i_{n} = i_{n_O},\, j_{n} = i_{n_R},\, j_{n_R} = j_{n_O} \right\}
-\end{align}
-
-We generally write out explicit subsequence coordinate pairs $(i,j)$ when their usage will make mathematical formulae clearer and 
-fold-envelope labels $n$ when writing pseudocode.  An implementation will rely on such an iterator over subsequences in the fold envelope.
-
-\subsection{Inside}
-The inside probability $\inside{\bvec{a}}{i_1}{j_1}{i_2}{j_2}{i_3}{j_3}$ is the summed probability of subsequences 
-$x^{(1)}_{i_1+1} \dots x^{(1)}_{j_1}$, $x^{(2)}_{i_2+1} \dots x^{(2)}_{j_2}$, $x^{(3)}_{i_3+1} \dots x^{(3)}_{j_3}$
-under all paths through the model which are rooted in state $\bvec{a}$.  
-\algref{inside} gives pseudocode for the fold-envelope version of the Inside algorithm.  The subroutines 
-$\mathrm{calcTransEmitProb}()$, $\mathrm{calcLBifurcProb}()$ and $\mathrm{calcRBifurcProb}()$
-used in the Inside algorithm are defined below.
-
-\begin{algorithm}[!ht]
-
-  \SetKwData{bifurcProb}{bifurcProb}
-  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
-  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
-  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
-  \SetLine
-  \KwIn{sequences $x^{(1)},x^{(2)},x^{(3)}$}
-
-  initialization\;
-  \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(1)} \in \foldenv^{(1)}$} {
-    \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(2)} \in \foldenv^{(2)}$} {
-      \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(3)} \in \foldenv^{(3)}$} {
-        \BlankLine
-        \ForEach{state $\bvec{a}$}{
-          \BlankLine
-          \bifurcProb $\leftarrow 0$\;
-          \ForEach{$\left( n^{(1)}_L,n^{(1)}_R \right) \in b_{in}\left( n^{(1)} \right)$} {
-            \ForEach{$\left( n^{(2)}_L,n^{(2)}_R \right) \in b_{in}\left( n^{(2)} \right)$} {
-              \ForEach{$\left( n^{(3)}_L,n^{(3)}_R \right) \in b_{in}\left( n^{(3)} \right)$} {
-                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
-                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
-              }
-            }
-          }
-          \BlankLine
-          $\alpha_{\bvec{a}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$ $\leftarrow$ \calcTransEmitProb$\left( \bvec{a}; n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
-          $\alpha_{\bvec{a}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$ += \bifurcProb\;
-          store $\alpha_{\bvec{a}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
-        }
-        \BlankLine
-      }
-    }
-  }
-  \KwRet{$\alpha_{\bvec{a}}\left( n[0,L_1],n[0,L_2],n[0,L_3] \right)$}\;
-
-  \caption{\alglabel{inside}
-    The constrained Inside algorithm for three sequences $x^{(1)},x^{(2)},x^{(3)}$.
-    States $\bvec{a}$ in the iteration over states are sorted in Inside fill order
-    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
-  }
-\end{algorithm}
-
-The transition and emission probability $\mathrm{calcTransEmitProb}(\bvec{a}; \cdot)$ can be calculated recursively 
-in the unconstrained case as
-\begin{align} \label{eqn:insidetransemit}
-  & \sum_{y_1\in\{x^{(1)}_{i_1+1},\Tnull\}}\sum_{z_1\in\{x^{(1)}_{j_1},\Tnull\}} \sum_{y_2\in\{x^{(2)}_{i_2+1},\Tnull\}}\sum_{z_2\in\{x^{(2)}_{j_2},\Tnull\}} \sum_{y_3\in\{x^{(3)}_{i_3+1},\Tnull\}}\sum_{z_3\in\{x^{(3)}_{j_3},\Tnull\}} \nonumber \\
-  &\qquad \left[ \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{y}\bvec{b}\bvec{z}} \weight (\bvec{a}\to\bvec{y}\,\bvec{b}\,\bvec{z}) \inside{\bvec{b}}{i_1+|y_1|}{j_1-|z_1|}{i_2+|y_2|}{j_2-|z_2|}{i_3+|y_3|}{j_3-|z_3|} \right] .
-\nonumber
-\end{align}
-Null transitions are caught in the sum when the terminals $\bvec{y} = \bvec{z} = \Tnull$.
-The constrained case is handled differently: The multiple sums are replaced by a single iteration over states $\bvec{b}$
-which connect subsequences $n^{(1)}, n^{(2)}, n^{(3)}$ to others in the fold envelopes.
-Pseudocode for the constrained calculation is given in \algref{insidetransemit}.
-
-\begin{algorithm}[!ht]
-  \SetKwData{emitProb}{emitProb}
-  \SetKwFunction{Weight}{Weight}
-  \SetLine
-  \KwIn{state $\bvec{a}, n^{(1)}, n^{(2)}, n^{(3)}$, intermediate Inside matrix $\alpha$}
-
-  \emitProb $\leftarrow 0$\;
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
-    \emitProb += \Weight$\left( \bvec{a} \to \bvec{b} \right)\, \alpha_{\bvec{b}}\left( n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
-  }
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{y}\,\bvec{b}\,\bvec{z}$} {
-    \lIf{$c_{in}\left(\bvec{b}; n^{(1)}\right) \notin \foldenv^{(1)}$ or $c_{in}\left(\bvec{b}; n^{(2)}\right) \notin \foldenv^{(2)}$ or $c_{in}\left(\bvec{b}; n^{(3)}\right) \notin \foldenv^{(3)}$}{next\;}
-    \emitProb += \Weight$\left( \bvec{a} \to \bvec{y}\,\bvec{b}\,\bvec{z} \right)\, \alpha_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(1)}\right), c_{in}\left(\bvec{b}; n^{(2)}\right), c_{in}\left(\bvec{b}; n^{(3)}\right) \right)$\;
-  }
-  \KwRet{\emitProb}\;
-  \caption{\alglabel{insidetransemit}
-    Subroutine $\mathrm{calcTransEmitProb}()$ for the Inside algorithm.
-  }
-\end{algorithm}
-
-The left-bifurcation probability $\mathrm{calcLBifurcProb}\left( \bvec{a}; n^{(1)}_L,n^{(1)}_R,n^{(2)}_L,n^{(2)}_R,n^{(3)}_L,n^{(3)}_R \right)$ is
-\[ \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} \weight (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \alpha_{\bvec{c}}\left( n^{(1)}_L,n^{(2)}_L,n^{(3)}_L\right) \, \alpha_{\bvec{b}}\left( n^{(1)}_R,n^{(2)}_R,n^{(3)}_R\right) \]
-and the right-bifurcation probability $\mathrm{calcRBifurcProb}\left( \bvec{a}; n^{(1)}_L,n^{(1)}_R,n^{(2)}_L,n^{(2)}_R,n^{(3)}_L,n^{(3)}_R \right)$ is
-\[ \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} \weight (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \alpha_{\bvec{b}}\left( n^{(1)}_L,n^{(2)}_L,n^{(3)}_L\right) \, \alpha_{\bvec{d}}\left( n^{(1)}_R,n^{(2)}_R,n^{(3)}_R\right) .
-\]
-
-The boundary condition gives the probability of a subsequence of length 0,
-\beqn \label{eqn:insidetoend}
-\inside{\bvec{a}}{j_1}{j_1}{j_2}{j_2}{j_3}{j_3} = t(\bvec{a},\Send) + \sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}} \weight (\bvec{a}\to\bvec{b}) \, \inside{\bvec{a}}{j_1}{j_1}{j_2}{j_2}{j_3}{j_3} \, ,
-\eeqn
-for $0 \le j_1 \le L_1, 0 \le j_2 \le L_2, 0 \le j_3 \le L_3$.
-We are assuming that there are no cycles of $\Snull$ states as well as no bifurcations which can result in no emissions.
-The termination condition is 
-\[ P\left(x^{(1)},x^{(2)},x^{(3)}\right) = \alpha_{\mathrm{Start}}(0,L_1,0,L_2,0,L_3) \, , \]
-where $\mathrm{Start}$ is the unique start state of the model.
-
-We write the DP algorithms using the $\weight (\cdot)$ notation in order to preserve generality: In our formalism, a transition $a \to x\,b$ has $\emit(b) = x$, but the more common convention is to have $\emit(a) = x$.
-The difference will show up only in the value assigned to $\weight (a \to x\,b)$, so our DP algorithms can be used in both cases.
-
-\subsection{CYK}
-The CYK algorithm can be obtained from the Inside algorithm by replacing sums over paths through the model (or equivalently, parses)
-with the $\max()$ operation (e.g., in \eqref{eqn:insidetoend}, $\sum_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}}$ will be replaced by $ \max_{\bvec{b}|\,\exists\, \bvec{a}\to\bvec{b}}$).
-The CYK probabilities for indices $(i_1,j_1,i_2,j_2,i_3,j_3)$ then represent 
-the probability of the most likely path through the model for 
-subsequences $x^{(1)}_{i_1+1} \dots x^{(1)}_{j_1}$, $x^{(2)}_{i_2+1} \dots x^{(2)}_{j_2}$, $x^{(3)}_{i_3+1} \dots x^{(3)}_{j_3}$.
-
-\subsection{Outside}
-The outside probability $\outside{\bvec{b}}{i_1}{j_1}{i_2}{j_2}{i_3}{j_3}$ is the summed probability of the sequences $x^{(1)}$, $x^{(2)}$, $x^{(3)}$ 
-under all paths through the model which are rooted in the start state of the model, excluding all paths for the subsequences 
-$x^{(1)}_{i_1+1} \dots x^{(1)}_{j_1}$, $x^{(2)}_{i_2+1} \dots x^{(2)}_{j_2}$, $x^{(3)}_{i_3+1} \dots x^{(3)}_{j_3}$ 
-which are rooted in the state $\bvec{b}$.
-\algref{outside} gives pseudocode for the fold-envelope version of the Outside algorithm.
-The subroutines $\mathrm{calcTransEmitProb}()$, $\mathrm{calcLBifurcProb}()$ and $\mathrm{calcRBifurcProb}()$
-used in the Outside algorithm are defined below.
-
-\begin{algorithm}[!ht]
-  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
-  \SetKwData{bifurcProb}{bifurcProb}
-  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
-  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
-  \SetLine
-  \KwIn{sequences $x^{(1)},x^{(2)},x^{(3)}$, CYK matrix $\alpha$}
-
-  initialization\;
-  \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(1)} \in \foldenv^{(1)}$} {
-    \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(2)} \in \foldenv^{(2)}$} {
-      \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(3)} \in \foldenv^{(3)}$} {
-        \BlankLine
-        \ForEach{state $\bvec{b}$}{
-          \BlankLine
-          \bifurcProb $\leftarrow 0$\;
-          \ForEach{$\left( n^{(1)}_O,n^{(1)}_L \right) \in b_{out,L}\left( n^{(1)} \right)$}{
-            \ForEach{$\left( n^{(2)}_O,n^{(2)}_L \right) \in b_{out,L}\left( n^{(2)} \right)$}{
-              \ForEach{$\left( n^{(3)}_O,n^{(3)}_L \right) \in b_{out,L}\left( n^{(3)} \right)$}{
-                \bifurcProb += \calcLBifurcProb($\bvec{b}; \cdot$)\;
-              }
-            }
-          }
-          \ForEach{$\left( n^{(1)}_O,n^{(1)}_R \right) \in b_{out,R}\left( n^{(1)} \right)$}{
-            \ForEach{$\left( n^{(2)}_O,n^{(2)}_R \right) \in b_{out,R}\left( n^{(2)} \right)$}{
-              \ForEach{$\left( n^{(3)}_O,n^{(3)}_R \right) \in b_{out,R}\left( n^{(3)} \right)$}{
-                \bifurcProb += \calcRBifurcProb($\bvec{b}; \cdot$)\;
-              }
-            }
-          }
-          \BlankLine
-          $\beta_{\bvec{b}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right) \leftarrow$ \calcTransEmitProb$\left( \bvec{b}; n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
-          $\beta_{\bvec{b}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$ += \bifurcProb\;
-          store $\beta_{\bvec{b}} \left( n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
-        }
-        \BlankLine
-      }
-    }
-  }
-  \caption{\alglabel{outside}
-    The constrained Outside algorithm for three sequences $x^{(1)},x^{(2)},x^{(3)}$.
-    States $\bvec{a}$ in the iteration over states are sorted in Outside fill order
-    with $\Semit$ states first, then $\Snull$ states in topological order.
-  }
-\end{algorithm}
-
-The transition and emission probability $\mathrm{calcTransEmitProb}()$ can be calculated recursively in the unconstrained case as
-\begin{align}
-  & \sum_{y_1\in\{x^{(1)}_{i_1},\Tnull\}}\sum_{z_1\in\{x^{(1)}_{j_1+1},\Tnull\}} \sum_{y_2\in\{x^{(2)}_{i_2},\Tnull\}}\sum_{z_2\in\{x^{(2)}_{j_2+1},\Tnull\}} \sum_{y_3\in\{x^{(3)}_{i_3},\Tnull\}}\sum_{z_3\in\{x^{(3)}_{j_3+1},\Tnull\}} \nonumber \\
-  & \qquad \left[ \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{y}\bvec{b}\bvec{z}} \weight (\bvec{a}\to\bvec{y}\,\bvec{b}\,\bvec{z}) \, \outside{\bvec{a}}{i_1-|y_1|}{j_1+|z_1|}{i_2-|y_2|}{j_2+|z_2|}{i_3-|y_3|}{j_3+|z_3|}  \right] .
-\nonumber
-\end{align}
-As with the Inside algorithm, an efficient implementation of the fold-envelope constraints requires a different treatment, given in \algref{outsidetransemit}.
-
-The left-bifurcation probability $\mathrm{calcLBifurcProb}\left( \bvec{b}; n^{(1)}_O,n^{(1)}_L,n^{(2)}_O,n^{(2)}_L,n^{(3)}_O,n^{(3)}_L \right)$ is
-\[ \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{c}\bvec{b}} \weight (\bvec{a}\to\bvec{c}\,\bvec{b}) \, \beta_{\bvec{a}}\left( n^{(1)}_O,n^{(2)}_O,n^{(3)}_O\right) \, \alpha_{\bvec{c}}\left( n^{(1)}_L,n^{(2)}_L,n^{(3)}_L\right) \]
-and the right-bifurcation probability $\mathrm{calcRBifurcProb}\left( \bvec{b}; n^{(1)}_O,n^{(1)}_R,n^{(2)}_O,n^{(2)}_R,n^{(3)}_O,n^{(3)}_R \right)$ is
-\[ \sum_{\bvec{a}|\,\exists\, \bvec{a}\to\bvec{b}\bvec{d}} \weight (\bvec{a}\to\bvec{b}\,\bvec{d}) \, \beta_{\bvec{a}}\left( n^{(1)}_O,n^{(2)}_O,n^{(3)}_O\right) \, \alpha_{\bvec{d}}\left( n^{(1)}_R,n^{(2)}_R,n^{(3)}_R\right) \]
-
-\begin{algorithm}[!ht]
-  \SetKwData{emitProb}{emitProb}
-  \SetKwFunction{Weight}{Weight}
-  \SetLine
-  \KwIn{state $\bvec{b}, n^{(1)}, n^{(2)}, n^{(3)}$, intermediate Outside matrix $\beta$}
-
-  \emitProb $\leftarrow 0$\;
-  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{b}$} {
-    \emitProb += \Weight$\left( \bvec{a} \to \bvec{b} \right)\, \beta_{\bvec{a}}\left( n^{(1)}, n^{(2)}, n^{(3)} \right)$\;
-  }
-  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{y}\,\bvec{b}\,\bvec{z}$} {
-    \lIf{$c_{out}\left(\bvec{b}; n^{(1)}\right) \notin \foldenv^{(1)}$ or $c_{out}\left(\bvec{b}; n^{(2)}\right) \notin \foldenv^{(2)}$ or $c_{out}\left(\bvec{b}; n^{(3)}\right) \notin \foldenv^{(3)}$}{next\;}
-    \emitProb += \Weight$\left( \bvec{a} \to \bvec{y}\,\bvec{b}\,\bvec{z} \right)\, \beta_{\bvec{a}}\left( c_{out}\left(\bvec{b}; n^{(1)}\right), c_{out}\left(\bvec{b}; n^{(2)}\right), c_{out}\left(\bvec{b}; n^{(3)}\right) \right)$\;
-  }
-  \KwRet{\emitProb}\;
-  \caption{\alglabel{outsidetransemit}
-    Subroutine $\mathrm{calcTransEmitProb}()$ for the Outside algorithm.
-  }
-\end{algorithm}
-
-The boundary condition is just
-\beqn
-\outside{\mathrm{Start}}{0}{L_1}{0}{L_2}{0}{L_3} = 1 \, .
-\eeqn
-
-\subsection{Loopy DP}
-
-The standard Inside and Outside algorithms fail on grammars with 1) cycles of $\Snull$ states, 
-or 2) bifurcations with empty children
-because they are incapable of performing a probabilistic summation over $\Snull$ cycles
-(empty children of bifurcations can give rise to effective cycles of $\Snull$ states).
-Phrased qualitatively, the standard Inside and Outside fail because the iterations over
-states $\bvec{a}$ in \algref{inside} and \algref{outside}
-can correspond to, at most, single $\Snull$ cycles rather than the infinitely many which are possible.
-
-While we believe that an efficient analytic summation is impossible,
-we can use an iterative approach to approximate an exact result to high precision.
-A brute-force approach is to just repeatedly iterate over all states $\bvec{a}$ until
-convergence is reached, but this is computationally-costly.
-A much more efficient algorithm exists.
-The key insight is that only $\Snull$ states which are strongly connected
-(reachable from each other in the state graph by either other $\Snull$ states
-or bifurcations with empty children) can contribute to any possible $\Snull$ cycle.
-It follows that instead of repeatedly iterating over all states $\bvec{a}$,
-it suffices to decompose the state (sub) graph of $\Snull$ states and bifurcation
-states with possibly-empty children into its strongly-connected components
-and repeatedly iterate over these connected components rather than the entire graph.
-Decomposition of a graph into its strongly-connected components can
-be done in linear time \cite{Tarjan1972},
-so the overall complexity of inference is still determined by the DP iterations themselves.
-If we iterate over each connected component 5 times, then upper bounds on 
-the complexity of inference are
-$O(5 \cdot A \cdot L^{2n})$ and $O(5 \cdot A \cdot L^{3n})$ in space and time for a model 
-with $A$ states of $n$ extant sequences.
-For the star phylogeny of \figref{threeway},
-
-\newpage
 \section{The TKF Structure Tree model as a transducer} \seclabel{tkfst}
 
-We can cast the TKF91 Structure Tree model as a transducer.
-{\bf Why is this important/useful? e.g. ``This means we can automatically deduce rules like those shown in Tables~\tabnum{abc} through~\tabref{agapbc}...''}
+We can cast the TKF Structure Tree (TKFST) model as a transducer.
+This means we can automatically deduce rules like those shown for the TKFST model in the main paper.
 
@@
     Singlet transducers can only have states of type $\Sstart$ or $\Sinsert$.
-    {\bf This is the indiegram-style transducer equivalent of the SCFG in \tabref{a} ...}
+    This is the indiegram-style transducer equivalent of the singlet TKFST SCFG in the main paper.
   }
@@
     in structure to the singlet transducer.  They are used to insert a new stem-loop structure.
-    {\bf This is the indiegram-style transducer equivalent of the SCFG in \tabref{ab} ...}
+    This is the indiegram-style transducer equivalent of the conditional pair TKFST SCFG in the main paper.
   }
@@
 
+\subsection{Simulation of RNA family evolution}
+The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
+Another script (\animateevolsayer) can be used to make animations of the evolving RNA structures, in combination with the RNAplot program in the \viennarna\ package \cite{HofackerEtAl94}.
+
 \newpage
-\section{Reconstruction of covariant substitution histories}
+\section{Computational experiments}
+
+\subsection{Reconstruction of covariant substitution histories}
 
@@
 
-\bibliography{../latex-inputs/alignment,../latex-inputs/duplication,../latex-inputs/genomics,../latex-inputs/ncrna}
+\bibliography{../latex-inputs/alignment,../latex-inputs/reconstruction,../latex-inputs/duplication,../latex-inputs/genomics,../latex-inputs/ncrna,../latex-inputs/url}
 

### 2009-05-29 response-to-reviews2.txt
Revised supplement; Ian's changes to Rob's changes

@@
 We completely agree. Such an algorithm was initially a goal of this
-work, but was abandoned. We thank the reviewer for challenging us to
-attack it afresh; a verbal presentation of the algorithm we came up
-with is given in the revised text, with a full formal description
-(including notation) in the supplement. (Since this is a standalone
-result that can be applied to any SCFG, we think it fits better as an
-appendix in the supplement; the obvious cost is omitting some detail
-from the main paper, but the PLoS Authors' Guidelines and the
-reviewer's preference for including more detail in the manuscript are
-potentially in conflict on the best approach to take here. We welcome
-suggestions for improving the flow generally.)
+work, but was dropped. We thank the reviewer for challenging us to
+try again; a verbal presentation of the algorithm we came up with is
+given in the revised text, with a full formal description (including
+notation) in the supplement. (Since this is a standalone result that
+can be applied to any SCFG, we think it fits better as a technical
+appendix in the supplement, with a verbal description in the main
+paper; the obvious cost is omitting some detail from the main paper.
+The PLoS Authors' Guidelines (which recommend using the supplement for
+algorithmic details) and the reviewer's preference for including more
+detail in the manuscript are potentially in conflict on the best
+approach to take here. We welcome advice on how to improve the flow
+generally.)
 
@@
 elimination is that while this marginalization is reasonably
-straightforward for independent-sites models and HMMs, the possible
-presence of null bifurcations complicates matters for SCFGs.  Indeed,
-no *closed form* elimination is possible (due to bifurcations, such an
-elimination would involve solving simultaneous nonlinear equations, as
-explained in the supplement) but an iterative approximation
-(e.g. Newton-Raphson) appears adequate.
+straightforward for independent-sites models and HMMs, essentially
+involving summing a geometric series to obtain a correction factor,
+the possible presence of null bifurcations complicates matters for
+SCFGs.  Indeed, no *closed form* elimination is possible (due to
+bifurcations, such an elimination would involve solving simultaneous
+nonlinear equations, as explained in the supplement) but an iterative
+approximation (e.g. Newton-Raphson) appears adequate.
 
@@
 
+Additionally, we have used our simulation script to prepare animations
+showing illustrative evolutionary trajectories of RNA structures under
+our model.  The software to produce these animations is available and
+briefly mentioned in the supplement.  (If there were some way to
+include these animations in the online version, we would like to do
+that, though this may have to wait for the <video> tag of HTML 5...)
+
 > Figure 4 and Figure 5 seem unnecessary. Both are too schematic to 

### 2009-05-29 paper.tex
Revised supplement; Ian's changes to Rob's changes

@@
 %\vspace{.5\baselineskip}
-We assume that the phylogeny, including both the tree topology and
-branch lengths, is known.
+
+We assume here that the phylogeny, including both the tree topology and branch lengths, is given.
+This constraint somewhat arbitrarily confines the problem domain to the rather restricted case where we already have a phylogeny (and potentially a consensus structure),
+and we seek only an alignment plus ancestral reconstruction.
+It is important to note, however, that as far as the {\em model} is concerned, none of the constraints (phylogeny, structure or alignment) are final.
+For example, the probabilistic framework readily allows us to compare likelihoods of two different phylogenetic trees, by constructing a composite transducer for each tree.
+Thus, MCMC samplers for alignments could in principal be extended to phylogenies (albeit at a computational cost).
+We revisit such MCMC samplers in the Discussion.
 
 This paper focuses on model construction and inference algorithms
-rather than the heuristics and constraints which will be necessary to make these
+rather than the heuristics which will be necessary to make these
 algorithms fast enough for analysis of many biological datasets.

### 2009-06-01 s1.tex
Text S1 (composition algorithm).

--- 2009-05-29 supplementary.tex
@@
   {\Large
-    \textbf{\titlestring: \supptext}
+    \textbf{\titlestring: \supptext{1}}
   }
@@
 \tableofcontents
-
-\newpage
-\section{Exact elimination of SCFG null cycles}
-
-The following section describes how to transform any SCFG so as to remove null cycles while preserving access to the full posterior probability distribution over parses, including parses with null cycles.
-
-\subsection{Definitions}
-
-Let $G = ( \Omega, \gnons, \grules, P )$ be a stochastic context-free grammar (SCFG)
-consisting of
-a set of terminal symbols $\Omega$,
-a set of nonterminal symbols (a.k.a. ``states'') $\gnons$,
-a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
-and a probability function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.
-
-% (The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
-% The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)
-
-Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e., rules in which $A$ appears on the left),
-and let $\rightside{A}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right, including bifurcations which also generate another nonterminal as well as $A$).
-Let $\allrules{A} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
-Define these also on sets of nonterminals, e.g., $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.
-
-Let $\gtrees$ be the set of all parse trees for $G$.
-Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
-then define the parse tree likelihood $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.
-
-If $\sum_{T \in \gtrees} P(T) = 1$, we say that $G$ is {\em probabilistically normalized by parse tree}.
-If $\sum_{\rho \in \leftside{A}} P(\rho) = 1$ for all $A \in \gnons$, we say that $G$ is {\em probabilistically normalized by production rule}.
-Note that normalization by production rule $\Rightarrow$ normalization by parse tree.
-
-Let $S \in \Omega^\ast$ denote a terminal sequence.
-Let $\mathrm{seq}:\gtrees \to \Omega^\ast$ be the function mapping a parse tree to its terminal sequence.
-and let $\mathrm{root}:\gtrees \to \gnons$ be the function returning the root nonterminal of a parse tree.
-The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
-Of particular relevance to null state elimination is the probability $P(\epsilon|A)$,
-where $\epsilon$ is the empty sequence.
-This is the probability that a nonterminal $A$ will expire without generating any sequence.
-
-Following earlier formalisms \cite{Durbin98,HolmesRubin2002a,Holmes2005},
-we say that the grammar $G$ has {\em RNA normal-form rules} if
-each rule in $\grules$ takes one of the following four forms:
-
-\noindent \hrulefill
-\begin{description}
-\item[Termination rules] with one nonterminal on the left and the empty string on the right
-\begin{eqnarray*}
-A & \to & \epsilon
-\end{eqnarray*}
-\item[Transition rules] with one nonterminal on the left and the right
-\begin{eqnarray*}
-A & \to & B
-\end{eqnarray*}
-\item[Bifurcation rules] with one nonterminal on the left and two on the right
-\begin{eqnarray*}
-A & \to & B\ C
-\end{eqnarray*}
-\item[Emission rules] with one nonterminal on the left and right and at least one terminal on the right
-\begin{eqnarray*}
-A & \to & x\ B \\
-A & \to & B\ y \\
-A & \to & x\ B\ y
-\end{eqnarray*}
-\end{description}
-\noindent \hrulefill
-
-\noindent
-Here $A,B,C \in \gnons$ and $x,y \in \Omega$.
-
-Suppose that grammar $G$ has RNA normal-form rules.
-We further say that $G$ has {\em RNA normal-form states} if
-each nonterminal (state) $A \in \gnons$ takes one of the following forms:
-
-\noindent \hrulefill
-\begin{description}
-\item[Null states]: $\leftside{A}$ contains only transition and termination rules.
-\item[Bifurcation states]: $\leftside{A}$ contains exactly one bifurcation rule, $A \to X\ Y$, where $X \neq Y$ and $X,Y$ are both null states.
-\item[Emit states]: $\leftside{A}$ contains only emission rules. Further, $\leftside{A} = \rightside{B}$ for some null state $B$.
-This null state $B$ is called $A$'s {\em post-emit} state.
-\end{description}
-\noindent \hrulefill
-
-Define a {\em null cycle} to be a nonterminal $A$
-and a sequence of rules $\rho_1, \rho_2 \ldots \rho_k$ that, when applied consecutively to $A$, leave $A$ unchanged;
-that is, $\rho_k(\rho_{k-1} \ldots \rho_2(\rho_1(A))) = A$.
-Define a {\em null subtree} to be a null cycle using at least one bifurcation rule.
-
-Suppose that $G = ( \Omega, \gnons, \grules, P )$ and $G' = ( \Omega, \gnons', \grules', P' )$ are two grammars.
-We say that $G$ and $G'$ are {\em equivalent in sequence} if there is a mapping between nonterminals, $f:\gnons \to \gnons'$,
-such that $P(S|A) = P(S|f(A))$.
-We say that $G$ and $G'$ are {\em equivalent in parse} if there is a mapping between parse trees, $g:\gtrees \to \gtrees'$,
-such that $P'(T') = \sum_{T:T'=g(T)} P(T)$ and we can define
-\[
-P(T|T') = P(T|g(T)=T') = \frac{P(T)}{P'(T')}
-\]
-
-Note that equivalence in parse $\Rightarrow$ equivalence in sequence.
-
-Suppose that a grammar $G$ contains null cycles.
-We seek to transform $G$ into a grammar $G''$ that is equivalent in parse and sequence, but has no null cycles.
-If $G$ is normalized by parse tree, then $G''$ will be too; but $G''$ is not (necessarily) normalized by production rule.
-
-We do the transformation in two steps, $G \to G' \to G''$.
-We also provide a stochastic ``null cycle restoration'' algorithm for sampling from $P(T|T'')$.
-
-\subsection{Eliminating null bifurcations}
-
-Note first that any SCFG can be transformed into an equivalent one with RNA normal-form states by adding null states.
-Without loss of generality, we therefore consider SCFGs with RNA normal-form states.
-
-Let $G = ( \Omega, \gnons, \grules, P )$ be such an SCFG.
-Let $\gnulls \subseteq \gnons$ be the set of null states in $\gnons$, {\bf excluding} post-emit states.
-Let $\gemits \subset \gnons$ be the set of post-emit states.
-Let $\gbifs \subseteq \gnons$ be the set of bifurcation states.
-
-We here define the grammar $G'$,
-\[
-G' = \left( \Omega, \gnons', \grules', P' \right)
-\]
-which is equivalent to $G$ in sequence but has no null subtrees.
-
-The new states $\gnons'$ are defined as follows.
-Start with $\gnons$; remove bifurcation states;
-then introduce a new state $N'$ for every null $N \in \gnulls$ and three new states $B^0, B', B^2$ for every bifurcation $B \in \gbifs$.
-
-\[
-\gnons' = 
-\left( \bigcup_{N \in \gnulls} \{ N' \} \right) \cup \left( \bigcup_{B \in \gbifs} \{ B^0, B', B^2 \} \right) \cup \gnons \setminus \gbifs
-\]
-
-The idea is that $N'$ (in $G'$) is a null state that is equivalent to null state $N$ (in $G$) for non-empty sequence only.
-That is, for any terminal sequence $S$,
-
-\[
-P'(S|N') = \left\{ \begin{array}{ll} P(S|N) & \mbox{if $S \neq \epsilon$} \\ 0 & \mbox{if $S = \epsilon$} \end{array} \right.
-\]
-
-Similarly, $B'$ is a null state that is equivalent to bifurcation state $B$ for non-empty sequence,
-while $B^2$ is a null state that is is equivalent to $B$ for parse trees where both children of $B$ are non-empty.
-
-In contrast, $B^0$ is {\bf exactly} equivalent to $B$ in sequence.
-However, null subtrees in $B^0$ are explicitly accounted for,
-their probabilities factored into the transitions $B' \to L'$ and $B' \to R'$ and the termination $B^0 \to \epsilon$,
-using the inside probabilities for empty sequences, $P(\epsilon|L)$ and $P(\epsilon|R)$.
-
-The probabilities $P(\epsilon|X)$ are related by the following system of equations
-
-\[
-P(\epsilon|X) = P(X \to \epsilon) + \sum_Y P(X \to Y) P(\epsilon|Y) + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R)
-\]
-
-which are nonlinearly coupled (via bifurcations) but may be solved numerically, e.g., by the Newton-Raphson method,
-or by iterated approximation starting from a lower bound $P(\epsilon|X) \geq 0$.
-
-The new rules and their probabilities are
-\[
-\begin{array}{rrcl}
-\forall \rho \in (\grules \setminus \allrules{\gbifs}): & P'(\rho)             & = & P(\rho) \\
-                                \forall N \in \gnulls: & P'(N' \to \epsilon)  & = & 0  \\
-                          \ldots \forall B \in \gbifs: & P'(N' \to B')        & = & P(N \to B) \\
-                         \ldots \forall M \in \gnulls: & P'(N' \to M')        & = & P(N \to M) \\
-                         \ldots \forall E \in \gemits: & P'(N' \to E)         & = & P(N \to E) \\
-               \forall (A \to B) \in \rightside{\gbifs}: & P'(A \to B^0)        & = & P(A \to B) \\
-              \forall (B \to LR) \in \leftside{\gbifs}: & P'(B^0 \to \epsilon) & = & P(B \to LR)\ P(\epsilon | L)\ P(\epsilon | R) \\
-                                                       & P'(B^0 \to B')       & = & 1 \\
-                                                       & P'(B' \to \epsilon)  & = & 0 \\
-                                                       & P'(B' \to L')        & = & P(B \to LR)\ P(\epsilon | R) \\
-                                                       & P'(B' \to R')        & = & P(B \to LR)\ P(\epsilon | L) \\
-                                                       & P'(B' \to B^2)       & = & 1 \\
-                                                       & P'(B^2 \to L'\ R')   & = & P(B \to LR) \\
-\end{array}
-\]
-
-Note that grammar $G'$, like $G$, has RNA normal-form states.
-Note also that $G'$ is not, in general, normalized by production rule;
-however, $G'$ is equivalent to $G$ in parse and is therefore normalized by parse tree (if $G$ is).
-
-\subsection{Eliminating null states}
-
-We now proceed to eliminate null cycles from $G'$.
-Since null subtrees have been eliminated, the remaining null cycles use only transition rules.
-
-We create the grammar
-\[
-G'' = \left( \Omega, \gnons', \grules'', P'' \right)
-\]
-which has the same nonterminals as $G'$, but different rules and rule probabilities.
-
-Let $\gnulls' \subseteq \gnons'$ be the set of null states in $\gnons'$, {\bf including} post-emit states.
-
-Define ${\bf t}$, the {\em transition matrix} of $G'$, as $t_{XY} = P'(X \to Y)$ for all $X,Y \in \gnons'$.
-The effective transition probability $q_{XY}$ between two states $X,Y$
-sums over all paths through null states (in the summand, $n$ is the length of the null path):
-\[
-{\bf q} = \sum_{n=0}^\infty {\bf t}^n = ({\bf 1}-{\bf t})^{-1}
-\]
-Here ${\bf 1}$ is the identity matrix.
-The matrix inverse may be computed in the usual ways (Gauss-Jordan elimination, LU decomposition, etc.)
-
-Since the bifurcation states of $G'$ explicitly generate non-empty sequence,
-the system of equations relating the probabilities $P'(\epsilon|X)$ is completely linear
-and may be solved in the same way.
-Let $u_X = P'(\epsilon|X)$ and $v_X = P'(X \to \epsilon)$.
-Then
-
-\[
-{\bf u} = {\bf v} + {\bf tu}
-\]
-
-whose solution is ${\bf u} = {\bf qv}$.
-Thus $P'(\epsilon|X) = \sum_Y q_{XY} P'(Y \to \epsilon)$.
-
-We now define $P''$ as follows.
-\begin{itemize}
-\item For all rules $\rho \in \grules'$ that are {\bf not} transitions or terminations, set $P''(\rho) = P'(\rho)$.
-\item For terminations from null states $X \in \gnulls'$, set $P''(X \to \epsilon) = u_X$.
-(NB this may create some terminations $X \to \epsilon$ that were not present in $G'$.)
-\item For transitions from null states $X \in \gnulls'$ to non-null states $Y \notin \gnulls'$, set $P''(X \to Y) = q_{XY}$.
-(NB this may create some transitions $X \to Y$ that were not present in $G'$.)
-\item For transitions {\bf to} null states $X \in \gnulls'$, set $P''(A \to X) = 0$.
-\end{itemize}
-
-Although we have left null states in $G''$, they now have no incoming transitions
-and are inaccessible unless they are post-bifurcation states (i.e.,states which appear on the right-hand side of bifurcation rules), post-emit states, or the start (root) state.
-All other null states can therefore be dropped from $\gnons''$.
-
-\subsection{Restoring null states}
-
-Suppose that $\rho''$ is a rule in $G''$.
-Algorithm~\ref{alg:restoreTransitions} samples from the distribution of equivalent parse subtrees in $G'$ (possibly containing null cycles).
-In order to sample from $P(T'|T'')$ we need simply map Algorithm~\ref{alg:restoreTransitions} to each rule in $T''$.
-
-For parameter estimation by Expectation Maximization and some other applications, it is useful to know the expected number
-of times that a transition was used, summed over the posterior distribution of parse trees (including those with null cycles).
-If $n''(\rho'')$ is the expected number of times that transition $\rho''$ was used according to an Inside-Outside computation on $G''$,
-then the corresponding expectations $n'(\rho')$ are given by
-
-\begin{eqnarray*}
-       n'(X \to Y) & = & n''(X \to Y) \frac{P''(X \to Y)}{q_{XY}} + \sum_Z n''(X \to Z) \frac{P''(X \to Y) q_{YZ}}{q_{XZ}} \\
-n'(X \to \epsilon) & = & n''(X \to \epsilon) \frac{P''(X \to \epsilon)}{P''(\epsilon | X)}
- + \sum_W n''(W \to \epsilon) \frac{q_{WX} P''(X \to \epsilon)}{P''(\epsilon | W)}
-\end{eqnarray*}
-
-Expectations for other rules (bifurcations and emissions) are the same for $G'$ as $G''$.
-
-\subsection{Restoring null bifurcations}
-
-Suppose that $\rho'$ is a rule in $G'$.
-Algorithm~\ref{alg:restoreBifurcations} samples from the distribution of equivalent parse subtrees in $G$ (possibly containing null subtrees).
-This algorithm also calls Algorithm~\ref{alg:sampleNullSubtree}, which samples from the distribution of empty subtrees rooted at a particular nonterminal.
-In order to sample from $P(T|T')$ we need simply map Algorithm~\ref{alg:restoreBifurcations} to each rule in $T'$.
-
-If $n'(\rho')$ is the expected number of times that rule $\rho'$ was used by $G'$,
-then the corresponding expectations $n(\rho)$ are given by
-
-\begin{eqnarray*}
-    n(B \to L\ R) & = & n'(B' \to L') + n'(B' \to R') + n'(B^0 \to L'\ R') + d(B \to L\ R) \\
-       n(X \to Y) & = & n'(X \to Y) + d(X \to Y) \\
-n(X \to \epsilon) & = & n'(X \to \epsilon) + d(X \to \epsilon)
-\end{eqnarray*}
-Expectations for emissions are the same for $G$ as $G'$.
-In the above expressions $d(\rho)$ is the expected usage of rule $\rho$ by null subtrees:
-
-\[
-d(\rho) = \sum_{(B \to LR) \in \grules} \left[
-n'(B' \to L') c_R(\rho)
-+ n'(B' \to R') c_L(\rho)
-+ n'(B^0 \to \epsilon) (c_L(\rho) + c_R(\rho))
-\right]
-\]
-
-where $c_W(\rho)$ is the expected usage of rule $\rho$ by an empty parse tree rooted at $W$,
-given by
-
-\begin{eqnarray*}
-c_X(\rho) P(\epsilon | X) & = & P(X \to \epsilon) \delta_{\rho=(X \to \epsilon)}
- + \sum_Y P(X \to Y) P(\epsilon|Y) \left( c_Y(\rho) + \delta_{\rho=(X \to Y)} \right)
-\\ & &
- + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R) \left( c_L(\rho) + c_R(\rho) + \delta_{\rho=(X \to LR)} \right)
-\end{eqnarray*}
-
-where $\delta_U$ is the Kronecker delta (1 if condition $U$ is true, 0 if it is false).
-Note that in contrast to the system of equations for $P(\epsilon|X)$, this is a linear system of equations
-of the form ${\bf c} = {\bf Mc} + {\bf k}$, which may be solved by matrix inversion:
-${\bf c} = ({\bf 1}-{\bf M})^{-1} {\bf k}$.
 
@@
 By labeling the nonterminals of a regular or stochastic context-free grammar (SCFG) as states of an abstract machine, allowing these states to absorb and emit 
-appropriate terminal symbols,\footnote{The relationship between formal grammars and abstract state machines is more transparent
-in the Mealy machine view, but the Moore machine view turns out to be more useful for our purposes.},
-and carefully assigning transition and emission weights,
+appropriate terminal symbols, and carefully assigning transition and emission weights,
 we can create a state machine which generates the same language, with the same distribution of weights, as that produced by the original grammar.\footnote{We speak of
@@
 then so can all machines at $\{m|m \trianglerighteq \hat{n}\}$.
-The TKF Structure Tree model (\secref{tkfst}) satisfies this condition.
+The TKF Structure Tree model satisfies this condition.
 A more general approach is probably possible, but it involves summing over paths $a': a_m \to a' \to \Send$,
 handling possible bifurcations in these paths, etc.
-
-\newpage
-\section{The TKF Structure Tree model as a transducer} \seclabel{tkfst}
-
-We can cast the TKF Structure Tree (TKFST) model as a transducer.
-This means we can automatically deduce rules like those shown for the TKFST model in the main paper.
-
-\subsection{The single-sequence TKFST model as a singlet transducer}
-
-The states of the singlet transducer are shown in \tabref{tkfstsinglet}.
-Allowed transitions between these states are shown in the paper.
-
-\begin{table}[!ht]
-  \centering
-  \begin{tabular}{lrrrr}
-    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
-    $L$ & $\Sstart$ \\
-    $I_L$ & $\Sinsert$ & & $(x,\Tnull)$ & $p(x)$ \\
-    \\
-    $S$ & $\Sstart$ \\
-    $I_S$ & $\Sinsert$ & & $(x,y)$ & $p(x,y)$ \\
-    \\
-    $B$ & $\Sinsert$ & & $(L\,S)$ & 1 \\
-  \end{tabular}
-  \caption{
-    \tablabel{tkfstsinglet}
-    States of the singlet transducer of the TKF Structure Tree model \cite{Holmes2004}.
-    Singlet transducers can only have states of type $\Sstart$ or $\Sinsert$.
-    This is the indiegram-style transducer equivalent of the singlet TKFST SCFG in the main paper.
-  }
-\end{table}
-
-\subsection{The two-sequence TKFST model as a branch transducer}
-
-The states of the branch transducer are shown in \tabref{tkfstbranch}.
-Allowed transitions between these states are shown in the paper.
-
-\begin{table}[!ht]
-  \centering
-  \begin{tabular}{lrrrr}
-    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
-    $L$ & $\Sstart$ \\
-    $I_L$ & $\Sinsert$ & & $(u,\Tnull)$ & $p(u)$ \\
-    $M_L$ & $\Smatch$ & $(x,\Tnull)$ & $(u,\Tnull)$ & $p(u|x)$ \\
-    $D_L$ & $\Smatch$ & $(x,\Tnull)$ & $(\Tgap,\Tnull)$ & 1 \\
-    $W_L$ & $\Swait$ \\
-    \\
-    $S$ & $\Sstart$ \\
-    $I_S$ & $\Sinsert$ & & $(u,v)$ & $p(u,v)$ \\
-    $M_S$ & $\Smatch$ & $(x,y)$ & $(u,v)$ & $p(u,v|x,y)$ \\
-    $D_S$ & $\Smatch$ & $(x,y)$ & $(\Tgap,\Tgap)$ & 1 \\
-    $W_S$ & $\Swait$ \\
-    \\
-    $B_i$ & $\Sinsert$ & & $(L_i\,S_i)$ & 1 \\
-    $B$ & $\Smatch$ & $(L\,S)$ & $(L\,S)$ & 1 \\
-    $B_p$ & $\Smatch$ & $(L\,S)$ & $(L\,\Send)$ & 1 \\
-    $B_e$ & $\Smatch$ & $(L\,\Send)$ & $(L\,\Send)$ & 1 \\
-  \end{tabular}
-  \caption{\tablabel{tkfstbranch}
-    States of the branch transducer of the TKF Structure Tree model \cite{Holmes2004}.
-    States which have the same names as states of the singlet transducer in \tabref{tkfstsinglet}
-    are the branch equivalent of the corresponding singlet states
-    (e.g. a $\Smatch$ state might be the branch equivalent of an $\Sinsert$ state).
-    States $L_i$ and $S_i$ are the $\Sstart$ states of a sub-model (not shown) identical
-    in structure to the singlet transducer.  They are used to insert a new stem-loop structure.
-    This is the indiegram-style transducer equivalent of the conditional pair TKFST SCFG in the main paper.
-  }
-\end{table}
-
-\subsection{The multi-sequence TKFST model as a composite transducer}
-
-We can use the state graph construction algorithm described 
-in the paper and detailed in \secref{stategraph} to create a model of the simultaneous evolution
-of several sequences.
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics [scale=0.4] {figs/parent.pdf}
-  \caption{\figlabel{parent}
-    A simple example of transducer composition to build a multi-sequence model of two extant sequences.
-    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
-    A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
-    and two branch transducers (the gray boxes labeled $\Delta T_X$ and $\Delta T_Y$)
-    mutate it according to the specified multi-sequence model.
-    Gray nodes correspond to observed data and white nodes unobserved data.
-  }
-\end{figure}
-
-Consider the simple model shown in \figref{parent}.
-The state of the multi-sequence model describing this model is a 3-vector $\bvec{a} = \left( a_1,\,a_2,\,a_3 \right)$,
-where $a_1$ is the state of the singlet transducer generating the ancestral sequence $W$
-and $a_2$ and $a_3$ are the states of the branch transducers evolving $W$ into
-extant sequences $X$ and $Y$.
-
-We can show some of the allowed transitions of this multi-sequence model.
-The state of the branch transducer associated with the active node $n$ is shown in bold.
-
-\paragraph{Stem creation.}
-
-A stem is created at the root node $W$ (corresponding to a bifurcation in the singlet transducer $a_1$)
-and survives in the sequence $X$ at node 2 but is deleted in the sequence $Y$ at node 3.
-
-\begin{align}
-  \nakedthreevec{1:}{2:}{3:} \quad
-  \threevec{L}{L}{\mathbf{L}} &
-  \to \threevec{\mathbf{L}}{W_L}{W_L}
-  \to \threevec{B}{B}{B_p}
-  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{S}{\mathbf{S}}{\Send}
-  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\mathbf{S}}{W_S}{\Send} \\
-  & \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\mathbf{S}}{W_S}{\Send}
-\end{align}
-
-\paragraph{Stem insertion.}
-
-A stem sequence is inserted in sequence $X$ at node 2.
-
-\begin{align}
-  \nakedthreevec{1:}{2:}{3:} \quad
-  \threevec{L}{L}{\mathbf{L}} \to \threevec{L}{\mathbf{L}}{W_L}
-  \to \threevec{L}{B_i}{B}
-  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\Send}{S}{\mathbf{S}}
-  \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\Send}{\mathbf{S}}{W_S}
-\end{align}
-
-\paragraph{Stem termination.}
-
-All stem sequences are ended by (possibly empty) loop sequences.
-
-\begin{align}
-  \nakedthreevec{1:}{2:}{3:} \quad
-  \threevec{\mathbf{S}}{W_S}{W_S}
-  \to \threevec{B_e}{B_e}{B_e}
-  \to \threevec{\Send}{\Send}{\Send} \, \threevec{L}{L}{\mathbf{L}}
-  \to \threevec{\Send}{\Send}{\Send} \, \threevec{\Send}{\Send}{\Send}
-\end{align}
-
-The functions $\alpha(t)$, $\beta(t)$ and $\gamma(t)$ are parametrized by the insertion and deletion rates of the Structure Tree model.
-They are defined for loop sequences as
-\begin{align}
-  \kappa_1 &= \lambda_1 / \mu_1 \\
-  \alpha_1 &= \exp \left( -\mu_1 t \right) \\
-  \beta_1 &= \frac{\lambda_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) } \\
-  \gamma_1 &= 1 - \frac{\mu_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\left( 1 - \exp (- \mu_1 t) \right) \left(\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) \right) }
-\end{align}
-and similarly for stem sequences \cite{Holmes2004}.
-
-\newpage
-\input{tkfst.tex}
-
-\newpage
-\section{Software}
-
-We have written software tools in Perl and C++ implementing much of the theory described here.
-All tools are available from \darturl\ as part of the \dart\ software package for sequence analysis.
-
-\subsection{Automated grammar construction}
-We implemented our model construction algorithm on the star phylogeny (\figref{threeway}) 
-in a set of Perl scripts.
-Given a singlet transducer modeling ancestral sequences and structures and
-a branch transducer modeling structural evolution,
-the scripts generate C++ code to create the corresponding jointly-normalized SCFG.
-All possible models of structural evolution which can be represented by a Pair SCFG
-are permitted as input to the scripts,
-allowing for flexible and automated model design.
-
-Given files describing the singlet and branch transducers,
-including weights of all transitions which may be functions of evolutionary distance,
-the package ComposedTreeTransducer::FourWayComposedTT can automatically generate
-the state graph and transition matrix of a multi-sequence model of three extant sequences
-(\figref{threeway}).
-It removes the useless windback $\Snull$ states described in the paper and
-introduces effective direct transitions caused by bifurcations with possibly-empty children.
-The package ComposedTreeTransducer::TripletSCFG transforms a multi-sequence
-model created by ComposedTreeTransducer::FourWayComposedTT into the 
-corresponding jointly-normalized three-sequence SCFG (\secref{multigrammars})
-and generates C++ code to build the model.
-
-Example singlet and branch transducers files are provided for a simple 
-Pair HMM model, a simple Pair SCFG model and the full TKF Structure Tree model.
-
-\subsection{Reconstruction of ancestral structures}
-The program \indiegram\ can perform maximum-likelihood inference on the
-three-sequence SCFGs automatically generated by the ComposedTreeTransducer::TripletSCFG package.
-Complete or no structural information for the three extant sequences can be supplied
-as input.
-
-\subsection{Simulation of RNA family evolution}
-The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
-Another script (\animateevolsayer) can be used to make animations of the evolving RNA structures, in combination with the RNAplot program in the \viennarna\ package \cite{HofackerEtAl94}.
-
-\newpage
-\section{Computational experiments}
-
-\subsection{Reconstruction of covariant substitution histories}
-
-Our \xrate\ program is a multiple-alignment analysis tool
-that estimates rate parameters for a broad class of models, including covariant RNA substitution models \cite{KlostermanEtAl2006}.
-It can also be used to predict consensus secondary structures for alignments.
-We implemented ancestral sequence reconstruction in \xrate, including posterior probabilities that the reconstructions are correct.
-
-Given a multiple sequence alignment and a phylogenetic tree,
-\xrate\ estimates maximum-likelihood values of rate and probability parameters by Expectation Maximization.
-During parameter estimation, any unspecified ancestral sequences are summed out using Felsenstein's peeling algorithm.
-The new feature is that the program can now find the ancestral sequence that has the highest posterior probability,
-contingent on the maximum-likelihood estimates of the model parameters.
-During the parameter estimation and ancestral reconstruction steps, the secondary structure may be specified, or it may be summed out as a latent variable.
-
-For the ancestral reconstruction experiments described here,
-we assumed that alignment, phylogeny and secondary structure were known,
-but that ancestral sequences were unknown.
-
-A simple computational experiment demonstrates the need for covariant substitution models.
-We first used \xrate\ to estimate maximum-likelihood parameters for a covariant model of RNA base-pair substitution.
-These rates were estimated from ribosomal RNA alignments
-derived by \cite{DowellEddy2006} from the European rRNA database \cite{WuytsEtAl2004}.
-We also fit a ``naive'' general reversible point-substitution model to these alignments, again using \xrate.
-All subsequent results are implicitly conditioned on these maximum-likelihood rate estimates.
-Next, we used the companion program SIMGRAM to generate 5000 random alignments (including ancestral sequence),
-simulating on a 75-taxon phylogeny from an RNA gene family in the \RFAM\ database \cite{GriffithsJonesEtAl2003}.
-(Specifically, we chose the type-I Hammerhead ribozyme, one of the 5\% most populous \RFAM\ families.
- We repeated this experiment with other \RFAM\ phylogenies, but the general trends reported here were not dependent on the choice of tree.)
-We stripped the ancestral sequence out of the simulated alignments (leaving the true structure annotation intact),
-then re-estimated the ancestral sequence with \xrate, using both the (correct) covariant substitution model and the (naive) non-covariant point-substitution model.
-The imputed ancestral sequences were compared to the true sequences (known from the simulation),
-and the model-derived posterior probabilities were compared to the empirical accuracy of the corresponding reconstructed sequence.
-
-\begin{figure}[ht]
-  \begin{center}
-     \includegraphics [width=0.45\textwidth] {figs/hammer_paired.pdf}
-     \hspace*{\fill}
-     \includegraphics [width=0.45\textwidth] {figs/xval_paired.pdf}
-  \end{center}
-  \caption{
-    \figlabel{simulation}
-    Accuracy of confidence estimates for base-pair reconstruction in simulations (left) and cross-validations on real data (right).
-    Reconstructions were performed using \xrate\ \cite{KlostermanEtAl2006},
-    using both covariant base-pair substitution models
-    and naive, non-covariant, single-base point-substitution models.
-    The plots show the relationship between the posterior probability that a model-based base-pair reconstruction is correct (vertical axis),
-    and the empirical probability that reconstructions with that posterior probability are actually correct (horizontal axis).
-    If the model's posterior probability estimates are correct, these points should lie on a straight diagonal line.
-    Left: simulation test on the hammerhead ribozyme phylogeny.
-    A covariant model (estimated from rRNA alignments) was used to generate the simulated alignments on the given phylogeny; the same model was then used to reconstruct ancestral sequence and obtain confidence estimates.
-    As expected, the model accurately estimates the probability that the reconstructed base-pair is correct.
-    A naive, non-covariant model (the general reversible model, estimated from the same alignments) was also used for reconstruction and confidence estimates.
-    This model {\em under}-estimates the probability of correct reconstructions,
-    since its equilibrium frequency has a higher entropy than the covariant model:
-    i.e. it assigns probability more evenly across the sixteen possible base-pairs (as opposed to the covariant model, which strongly favors the six canonical base-pairs).
-    Right: cross-validation (``hold-one-out'') test using \RFAM\ alignments with published (verified) structures.
-    Here, the covariant model slightly {\em over}-estimates the probability of correctness.
-    The rRNA alignments (from which the model rates were estimated) have fewer non-canonical base-pairs than the \RFAM\ alignments,
-    so that the covariant model entropy is slightly {\em lower} than the empirical base-pair distribution;
-    this may explain the discrepancy.
-    The naive model's confidence estimates are correct
-    for high-accuracy reconstructions,
-    over-estimates for medium-accuracy reconstructions,
-    and under-estimates for low-accuracy reconstructions.
-  }
-\end{figure}
-
-\figref{simulation} (left) shows the results of these comparisons.
-When using the same (covariant) model for simulation and reconstruction, the posterior probabilities are an excellent unbiased estimate of the frequency with which the model reconstructions are correct.
-However, the naive point-substitution model, which does not include covariant substitutions at base-paired sites, systematically under-estimates these probabilities.
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | wc -l` / `wc -l $f`,"\n"}' Hammerhead_1.bptest
-In our simulation, the ancestral error rate for the naive model (4.9\%) was significantly higher than that of the covariant model (1.7\%).
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Hammerhead_1.bptest
-Furthermore, 74\% of incorrect base-pairs predicted by the naive model were non-canonical (i.e. not AU, CG, GC, UA, GU or UG),
-compared to only 2\% of incorrect predictions by the covariant model.
-
-\begin{figure}[p]
-  \begin{center}
-     \includegraphics [width=0.45\textwidth] {figs/hammer_hist.pdf}
-     \hspace*{\fill}
-     \includegraphics [width=0.45\textwidth] {figs/xval_hist.pdf}
-  \end{center}
-  \caption{
-    \figlabel{histograms}
-    Proportions of reconstructed base-pairs that are incorrect (and the subset of those that are non-canonical)
-    in simulations (left) and cross-validations on real data (right).
-    Base-pair reconstructions that are incorrect AND non-canonical
-    might potentially disrupt a synthetic reconstructed structure.
-    While the covariant model does not always get more base-pairs correct than the naive model,
-    it predicts fewer non-canonical base-pairs.
-    The sharpest difference between the models are in the lowest posterior-probability categories,
-    corresponding to low-confidence predictions (e.g. sequences on deep branches, or base-pairs with few homologues).
-    In this category, the covariant model also makes far fewer errors than the naive model.
-    See main text and caption to \figref{simulation} for details of simulation and cross-validation procedures.
-    \RFAM\ families with verified structures tended to be less phylogenetically diverse,
-    with roughly half the total branch length (on average), than the tree used for our simulations;
-    this may explain the relative paucity of low posterior-probability errors in the right-hand plot, compared to the left.
-% perl -e 'use Newick;use Stockholm::Database;$db=Stockholm::Database->from_file(shift);for$a(@$db){next if$a->columns>500 || @{$a->seqname}>400;$tree=Newick->from_string(@{$a->gf_NH});$b=0;for$l(@{$tree->branch_length}){$b+=$l}print"$b\n"}' ~/rfam/Training/Published.stock | meansd.pl
-% perl -e 'use Newick;use Stockholm::Database;$db=Stockholm::Database->from_file(shift);for$a(@$db){$tree=Newick->from_string(@{$a->gf_NH});$b=0;for$l(@{$tree->branch_length}){$b+=$l}print"$b\n"}' ~/rfam/Training/Hammerhead_1.stock
-  }
-\end{figure}
-
-We also performed cross-validation (hold-one-out) experiments. Starting from \RFAM\ alignments whose secondary structure is labeled ``Published'' (i.e. experimentally validated),
-we estimated phylogenetic trees using the neighbor-joining algorithm
-with the Jukes-Cantor model,
-then removed individual sequences
- and attempted to reconstruct them using \xrate.
-(This is possible because the substitution models are time-reversible, so we can treat any node as ancestral.)
-In this experiment, the error rates for base-pair reconstructions again differed by a few percent, though both error rates were higher than in the simulation test
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | wc -l` / `wc -l $f`,"\n"}' Published.xval-bptest
-(covariant model: 11\%, non-covariant model: 16\%).
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Published.xval-bptest
-Of the incorrectly-reconstructed base-pairs, the non-covariant model again predicted more non-canonical pairs (40\%) than the covariant model (17\%),
-especially at lower confidence levels (\figref{histograms}, right).
-
-Posterior probability estimates for both models were
-systematically higher in the cross-validation test than they were in the simulation (\figref{simulation}, right).
-The presence of this trend in both models
-suggests that it may be due to differences between the European rRNA alignments
-(from which the rates were estimated)
-and the \RFAM\ alignments (which were used in the cross-validation test).
-This is confirmed by our empirical observations of
-covariant rate matrices estimated from the two alignment datasets:
-base-pairs in the \RFAM\ alignments are more frequently
-non-canonical, and appear to evolve faster (relative to single-stranded regions),
-than base-pairs in the rRNA alignments
-(for details, see {\tt biowiki.org/RnaModels}).
-Repeating the cross-validation experiment with rates estimated from the \RFAM\ alignments,
-we find the covariant model predicted slightly fewer non-canonical base-pairs
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Published.pubxval-bptest
-(14\% rather than 17\%)
-and produced slightly more accurate posterior probability estimates (data not shown).
-Although this is no longer a strict cross-validation (since parameters were estimated from the test dataset),
-it nevertheless emphasizes the value of correct models.
-%Another possible source of errors in the cross-validation (compared to the simulation)
-%is inaccurately estimated phylogenetic topologies and branch lengths;
-%this might be improvable by using a covariant model in the phylogenetic estimation step (rather than Jukes-Cantor),
-%and/or by using Bayesian phylogenetic estimation tools (rather than neighbor-joining).
-
-Accurate estimates of posterior probabilities are likely to be important:
-sub-optimal predictions are a significant piece of the reconstruction puzzle \cite{WilliamsEtAl2006,PollockChang2007,ChangEtAl2007}.
-Ancestral reconstruction by probabilistic modeling does not yield a single definitive sequence, but rather a probability distribution over sequences.
-Combinatorial synthesis by degenerate primer assembly has been advocated as a means of sampling from this distribution \cite{Gaucher2007}.
-Nucleotide-perfect reconstructions may not always be possible:
-it may be more productive to aim for accurate confidence estimates
-(including sub-optimal predictions)
-than to focus exclusively on getting everything right the first time.
-
-Based on these tests, we conclude that deep phylogenetic reconstructions of ribosomes, and other RNAs, will require covariant substitution models that take account of RNA secondary structure.
-We may reasonably deduce that indel reconstructions will, similarly, need to take account of RNA structure.
-These results, therefore, strongly motivate the development of multi-sequence models for RNA structure, such as the TKF Structure Tree.
-
-\section{Terminological asides}
-
-This subsection contains some asides about terminology,
-and can be omitted on a first reading of this paper.
-The intent of this subsection is to highlight connections to related fields
-in molecular evolution and computer science.
-
-\subsection{Phylogenetic grammars}
-
-We here describe some relationships to grammar models in current use in molecular evolution.
-
-% Removing this sentence since I do not think that describing the relationship between transducers and SCFGs as an "injective mapping" will make it any clearer for most readers -- it is a bit like describing a string as a "concatenative monoid" -- IH
-%There exists an injective mapping from the evolutionary models generated by our model-construction algorithm to multi-sequence SCFGs. In other words, 
-Given a singlet transducer, a phylogenetic tree, and a mapping from tree to branch transducers,
-the transducer composition algorithm returns a corresponding multi-sequence SCFG
-which models the joint probability distribution $P(X_1 , ... , X_{n+m})$.
-
-Several related bioinformatics grammars can be represented within this multi-SCFG framework,
-allowing for substantial re-use of prior modeling theory:
-\begin{itemize}
-\item String transducers (which are similar to HMMs) can be viewed as special cases of parse-tree transducers,
-without the facility to model long-distance correlations (base-pairs) or bifurcated structures.
-\item Phylo-HMMs and phylo-SCFGs \cite{KlostermanEtAl2006}
-can be viewed as special cases of (respectively) string transducers and parse-tree transducers.
-(Phylo-HMMs and phylo-SCFGs typically model substitutions but not indels,
-and so can be viewed as compositions of a richly-featured singleton transducer with a restricted class of branch transducers.)
-\item Sakakibara \cite{Sakakibara2003} has described ``Pair HMMs on tree structures'' that are closely related to our parse-tree transducers
-(though without the composition algorithms for scoring multiple alignments).
-\item Sakakibara has also described pair stochastic tree-adjoining grammars, or TAGs \cite{Sakakibara2004}.
-TAGs are a formal generalization of SCFGs to
-mildly context-sensitive grammars \cite{MamitsukaAbe94,JoshiSchabes97,RivasEddy2000}.
-We speculate that the conditionally-normalized transducer analogues of such Pair-TAGs may encompass the
-phylogenetic model of covariant RNA substitutions and indels on pseudoknotted structures
-described by \cite{MeyerMiklos2007},
-just as parse-tree transducers encompass phylo-SCFGs.
-\end{itemize}
-
-\subsection{Statistical Alignment}
-
-The term {\bf Statistical Alignment} was introduced by Hein \cite{HeinEtal2000}
-to describe DP algorithms for phylogenetic grammars.
-Although this term has been adopted by others in molecular evolution \cite{Miklos2002,Metzler2003,DrummondRambaut2007,SatijaEtAl2008},
-we've avoided it (except in this section) for the following four reasons:
-(i) Several RNA alignment methods already use another probability distribution
-that's completely different to (though equally as valid as) the phylogenetic probability distribution over alignments:
-namely, the Boltzmann probability distribution over secondary structures
-\cite{GorodkinEtAl97,MathewsTurner02,HofackerEtAl2004}.
-(ii) Many alignment tools make use of prior and posterior distributions over pairwise alignments
-\cite{Miyazawa94,BucherHofmann96,Holmes98,BrayPachter2004,DoEtAl2004,Holmes2005,LoytynojaGoldman2005,SchwartzPachter2007,BradleyPachterHolmes2008};
-all of these are arguably ``statistical'' alignment tools as well.
-(iii) Some of the above-cited methods use sum-over-pairs alignment metrics, taking the expectation of such metrics over pairwise alignment posteriors,
-and finding the multiple alignment that maximizes this expectation \cite{DoEtAl2004,SchwartzPachter2007,BradleyPachterHolmes2008};
-others do progressive multiple alignment using pair grammars \cite{BrayPachter2004,Holmes2005,LoytynojaGoldman2005}.
-Speculatively, such pair-based heuristics may (in some sense) approximate probabilistic alignment inference with true phylogenetic likelihoods;
-similar convergences have, at least, been demonstrated between other pair-based heuristics \cite{EickmeyerEtAl2008}.
-Thus, even using Hein's strict definition,
-it's possible that a broad range of methods systematically {\em approximate} ``Statistical Alignment'';
-this cautions against using the term too exclusively.
-(iv) Even more broadly, many alignment tools make use of statistics to estimate alignment ``significance''.
-In this frequentist approach, probability distributions for alignment scores (particularly Gumbel distributions)
-are estimated for random sequences.
-The tail integrals of these distributions then yield P-values for high-scoring database search results
-\cite{KarlinAltschul93,Mott2000,Eddy2008}.
-Thus, while the term ``Statistical Alignment'' has a quite specific technical meaning in molecular evolution
-(a meaning which certainly encompasses the algorithms we describe here),
-the broader bioinformatics literature may have competing claims to the term.
-In any case, the goal of this work is statistical {\em reconstruction} of RNA protosequences,
-rather than alignment {\em per se}.
-
-\subsection{Parse trees and stacks}
-
-Grammars are abstract models for languages;
-finite-state automata are the corresponding abstract models for those languages' parsers.
-Most of this paper uses the grammatical view of RNA structure evolution;
-in this short section we digress briefly to mention the automata-theoretic view of these models.
-
-The two non-local features that set SCFGs apart from HMMs are bifurcations and paired emissions \cite{Durbin98}.
-In RNA structure analysis, these features are used to model (respectively) branched-loops and base-pairs.
-
-A stochastic pushdown automaton can also model these features.
-Such an automaton is essentially an HMM with a LIFO stack (LIFO = ``Last In, First Out'').
-
-The pushdown automaton can parse two sequences simultaneously, aligning them and their parse trees.
-It handles bifurcations and paired emissions by pushing symbols onto its stack:
-a bifurcation by pushing a nonterminal, and a paired emission by pushing a terminal.
-The pushed terminals and nonterminals are popped off the stack when the machine reaches its $\Send$ state.
-
-The conditionally-normalized version of a pushdown automaton is a pushdown transducer:
-a machine that absorbs {\em and} parses an input sequence (or absorbs an already-parsed input sequence)
-and emits a (parsed) output sequence.
-This is the machine that we describe as a parse-tree transducer.
 

### 2009-06-01 s2.tex
Text S2 (null-cycle elimination)

--- 2009-05-29 supplementary.tex
@@
   {\Large
-    \textbf{\titlestring: \supptext}
+    \textbf{\titlestring: \supptext{2}}
   }
@@
 
-\newpage
-\section{Two-sequence transducer models}
-
-We here give formal definitions of two-sequence models and the state machines which
-generate the factored probability distribution $P (X, Y | \Delta T) = P (X) \cdot P (Y | X, \Delta T)$,
-where the marginal $P(X)$ is generated by a {\bf singlet transducer}
-and the conditional $P (Y | X, \Delta T)$ by a {\bf branch transducer}.
-We refer to the branch transducer as $\theta$, so the conditional distribution
-is more precisely $P (Y | X, \Delta T, \theta)$.
-
-\subsection{Formal grammars} \seclabel{grammars}
-
-There is a close relationship between formal grammars and the singlet and branch transducer abstract state machines.
-By labeling the nonterminals of a regular or stochastic context-free grammar (SCFG) as states of an abstract machine, allowing these states to absorb and emit 
-appropriate terminal symbols,\footnote{The relationship between formal grammars and abstract state machines is more transparent
-in the Mealy machine view, but the Moore machine view turns out to be more useful for our purposes.},
-and carefully assigning transition and emission weights,
-we can create a state machine which generates the same language, with the same distribution of weights, as that produced by the original grammar.\footnote{We speak of
-weights associated with a transition rather than probabilities in order to allow for more general models.}
-We therefore use the terms ``nonterminal'' and ``state'' interchangeably
-and refer to the ``parse tree'' generated by singlet and branch transducers.
-
-Phrased more precisely, there is an isomorphism between the singlet and branch transducers of two-sequence models
-and Pair SCFGs.  Practically speaking, this means that for every joint distribution $P(X, Y | \Delta T)$ 
-generated by a singlet and branch transducers, there exists a Pair SCFG which generates the same distribution.
-
-\subsection{States, transitions and emissions}
-A singlet transducer has states of type $\Sstart$ and $\Sinsert$.
-Each state $\phi\in\Phi$ of the branch transducer has type
-\[ \type(\phi) \in \lbrace\Sstart,\,\Sinsert,\,\Smatch,\,\Swait,\,\Send\rbrace \, . \]
-State typing is as follows:
-\begin{itemize}
-  \item A $\Sstart$ state begins a branch of the parse tree.
-  \item An $\Sinsert$ state emits, but does not absorb, symbols.
-  \item A $\Smatch$ state absorbs and (possibly) emits symbols.
-  \item A $\Swait$ state is a null state which allows a branch transducer to ``pause''
-    while it waits for an input symbol from another transducer.
-    In the two-sequence case, the input symbols is emitted by the singlet transducer generating
-    the ancestral sequence.
-  \item An $\Send$ state ends a branch of the parse tree.
-\end{itemize}
-The names of state types are similar to the $\{\Sstart,\,\Sinsert,\,\Smatch,\,\Send\}$ states of the familiar Pair HMM.
-Deletions are handled as a special case by states of type $\Smatch$ which absorb but do not emit symbols pairs.
-Only states of type $\Smatch$ can absorb symbol pairs $(x,y)$.
-
-Bifurcations in the grammar, when considered as emission of nonterminals, can be handled analogously
-to terminal emission.  States $\phi: \type(\phi) \in \{\Sinsert,\,\Smatch\}$ can emit nonterminal pairs $(c\,d)$,
-where $c$ or $d$ can be the null symbol, and the emitted pairs $(c\,d)$ can be absorbed by states of type $\Smatch$.
-
-The emission weight of a nonterminal pair $(c\,d)$ from the state $b: \type(b) = \Smatch$, 
-conditional on absorption of a nonterminal pair $(l\,m)$, is 
-\[ e_b ( c\,d|l\,m,\theta ) \, . \]
-
-The functions $\emit()$ and $\absorb()$ are defined to return the emission or absorption of a particular state,
-\begin{align*}
-  \emit(\phi) &:= \left\{\begin{array}{ll} (x,y) & x,y = \mathrm{terminal\,\,or\,\,null} \\(c\,d) & c,d = \mathrm{nonterminal\,\,or\,\,null}. \end{array} \right. \\
-  \absorb(\phi) &:= \left\{\begin{array}{ll} (x,y) & x,y = \mathrm{terminal\,\,or\,\,null} \\(c\,d) & c,d = \mathrm{nonterminal\,\,or\,\,null}. \end{array} \right.
-\end{align*}
-
-We frequently use the notation ${}^{uv}\phi^{xy}$ to indicate that a state of type $\Smatch$ absorbs a symbol pair $(u,v)$
-and emits a pair $(x,y)$.
-The notation for bifurcations is slightly different:
-A bifurcation state which left-emits a nonterminal $d$ and makes a transition to a state $\phi$ with weight 1 is written as $\bifurc{B}{d}{\phi}$.
-
-A transition between states $a$ and $b$ of the branch transducer $\theta$ has a weight
-\[ t(a,b|\theta) = t(a \to b|\theta). \]
-Terminal emission is handled by states $\phi: \type(\phi) \in \{\Smatch,\,\Sinsert\}$, which emit symbol pairs $(x,y)$,
-where $x$ or $y$ can be the null symbol.
-In this paired-emission perspective, left single-terminal emissions $x$ are represented as $(x\,\Tnull)$;
-right single-terminal emissions are handled similarly.
-The weight of an emission of a terminal pair $(x,y)$ from a state $b: \type(b) = \Smatch$, 
-conditioned on absorption of a terminal pair $(u,v)$, is 
-\[ e_b ( x,y|u,v,\theta ) \, . \]
-Recall that the emission weights of states of type $\Smatch$ are defined conditioned upon the absorbed symbols.
-
-\newpage
-\section{Multiple-sequence transducer models} \seclabel{evolutionarymodels}
-
-We can use our two-sequence models to construct a model of many sequence related by a
-guide phylogenetic tree.
-The guide tree specifies the (conjectured) phylogenetic relationship of all sequences.
-A singlet transducer, which emits, but does not absorb, symbols, lies at the root of the guide tree and 
-serves as a generative model of the ancestral sequence.
-A branch transducer represents the evolution of an ancestral sequence into a single descendant 
-sequence, that is, the action of evolution along the single-branch tree (Ancestor $\to$ Descendant).
-To represent the evolution of an ancestral sequence into many descendant sequences
-(whose phylogenetic relationship is specified by the guide tree),
-we place a branch transducer on each branch of the guide tree to create a multiple-sequence model.
-
-If the branch grammar has no bifurcations and only left or right emissions are allowed, then
-the language generated is a regular string language and the corresponding jointly normalized abstract state machine is 
-an HMM.  In this simplest case our formalism for creating a multiple-sequence model reduces to that given 
-by \cite{Holmes2003} for combining HMMs on a guide tree.
-
-\subsection{The guide tree}
-The nodes of the tree are labeled $1,\dots,N$ in the order reached by any preorder depth-first traversal of the tree.
-The length of each branch $(\parent(m) \to m)$ is given by the evolutionary time $t_m$.
-To specify ancestor-descendant relationships, we introduce notation:
-$\mDn$ ($\mNDn$) means node $m$ is descended from (not descended from) node $n$, and
-$\mDEn$ ($\mNDEn$) means node $m$ is descended from or identical to (not descended from and not identical to) node $n$.
-
-\subsection{States, transitions and emissions}
-The multiple-sequence model is formed by the composition/intersection of $(N-1)$ branch transducers 
-such that there is a branch transducer on each branch $(\parent(m) \to m);\, m = 2,\dots,N$ of the guide tree and a singlet transducer at the root node.
-Our framework allows for the placement of different branch transducers, with a unique set of nonterminals (state space) $\Phi$ and
-allowed transitions between states and corresponding weights, on each branch.  $\theta^{(m)}$ denotes 
-the branch transducer governing evolution along the branch $(\parent(m) \to m)$ of the guide tree.
-
-States of the multiple-sequence model are represented by as N-dimensional vectors $\bvec{a}$,
-\[ \bvec{a} = \statevec{a} \, . \]
-These states are typed as
-\[ \type(\bvec{a}) \in \lbrace\Sstart,\,\Semit,\,\Sbifurc,\,\Snull,\,\Send\rbrace \, . \]
-State typing is as follows:
-\begin{itemize}
-  \item A $\Sstart$ state begins a branch of the parse tree of one or more of the $N$ sequences on the phylogenetic tree.
-  \item An $\Semit$ state emits symbols (terminals) to one or more of the $N$ sequences.
-  \item A $\Sbifurc$ state corresponds to a bifurcation in the parse tree of
-    one or more of the $N$ sequences.
-  \item A $\Snull$ state corresponds to any non-$\Send$ state which represents neither
-    an emission or bifurcation.
-  \item An $\Send$ state ends a branch of the parse tree.
-\end{itemize}
-States are typed according to the transition by which they are reachable
-(details of the typing are given in \secref{transitions}).
-
-Transitions and emissions of the multiple-sequence model are defined in terms of the transitions and emissions 
-of the branch transducers at each node as well as the singlet transducer at the root node.
-The weight of a transition $\bvec{a} \to \bvec{b}$ is therefore
-\beqn
-t(\bvec{a},\bvec{b}) = \prod_{m|a_m \ne b_m} \transweight{m}{a_m}{b_m} \, ,
-\eeqn
-and the weight of an emission $(\bvec{x}\,\bvec{y})$ from a state $\bvec{b}$ is
-\beqn
-\totalemissionweight{\bvec{b}}{\bvec{x}\,\bvec{y}} = \prod_m e_{b_m}\left(x_m\,y_m|x_{\parent(m)}\,y_{\parent(m)},\theta^{(m)}\right) \, .
-\eeqn
-We frequently will not explicitly write out the conditional dependence on the absorbed terminals $(x_{\parent(m)}\,y_{\parent(m)})$,
-but the reader should keep in mind that in general emission weights will depend on the absorbed symbols.
-
-\subsection{Formal grammars} \seclabel{multigrammars}
-
-Analogously to the case with two-sequence models (\secref{grammars}),
-there exists a one-to-one mapping between the multiple-sequence models generated by our model-construction algorithm 
-and multi-sequence SCFGs.
-In other words, given a singlet and branch transducers of a two-sequence model, as well
-as a guide tree relating the extant sequences, there exists a corresponding multi-sequence SCFG
-which generates the same joint probability distribution $P(X_1 , ... , X_n)$.
-
-\subsection{Constructing the state graph} \seclabel{stategraph}
-
-As described in the paper,
-we need a way to efficiently construct the state graph of the multi-sequence model,
-where the state graph consists of a list of accessible states and the possible transitions between them.
-This state graph can be constructed by an uninformed depth-first search,
-where at each step of the search we obtain the possible child nodes 
-by applying one of the following possible transitions of the multi-sequence model:
-\begin{enumerate}
-\item {\bf Null transition}:
-  The state of a single branch transducer is updated, with no terminal emission or bifurcation.
-\item {\bf Terminal emission}:
-  A state makes a transition to a $\Sinsert$ state.  The emitted symbol is passed down the guide
-  tree to all descendant branch transducers, which transition to states of type $\Smatch$.
-\item {\bf Bifurcation}:
-  A state makes a transition to a special $\Sinsert$ state which emits a new branch of the parse tree.
-  The emitted symbol is passed down the guide tree to all descendant branch transducers, 
-  which transition to states of type $\Smatch$.
-\item {\bf End transition}:
-  The singlet transducer associated with the root sequence can transition to the $\Send$ state,
-  signaling that this parse tree is finished.
-\end{enumerate}
-
-Each of these transitions is explained in detail in the following section.
-
-\subsection{Allowed transitions} \seclabel{transitions}
-
-Following \cite{Holmes2003},  we let transitions of the multi-sequence model begin at the active node and cascade down the guide tree
-as appropriate.  Unless defined otherwise, node $n$ is the active node of the multi-sequence model with state $\bvec{a}$, 
-\begin{align}
-  n &= \activenode{a} \\
-  &= \mathrm{argmax}_m \left\{\type(a_m) \notin \{\Swait,\,\Send\}\right\} \, . \label{eqn:activenode} 
-\end{align}
-Each possible allowed transition is obtained by making a valid change (updating) the 
-state of the singlet or one or more of the branch transducers of the multi-sequence model.
-
-\subsubsection*{Null Transitions}
-
-\[ \statevec{a} \to \statevec{b} \]
-
-\begin{align} \label{eqn:nulltransition}
-  \type(\bvec{b}) &= \Snull  \\
-  \weight(\bvec{a}\to\bvec{b}) &= t(\bvec{a},\bvec{b}) \\
-  &= \transweight{n}{a_n}{b_n} .
-\end{align}
-
-This transition updates the state of the branch transducer at the active node $n$ of the guide tree, leaving 
-the rest unchanged, with no corresponding terminal emission or bifurcation in the grammar.
-Nodes other than the active node do not change state, $a_m = b_m \,\forall\, m \ne n$, and the only 
-allowable transitions of this form are to states $b_n:\,\type(b_n) \in \{\Sstart,\,\Swait\}$.  Transitions
-to states of type $\Sinsert$ or $\Smatch$ result in emissions, and transitions to the end state $\Send$
-are handled as a special case.
-
-\subsubsection*{Terminal Emission} \seclabel{terminalemission}
-\[ \statevec{a} \to \statevec{x}\,\statevec{b}\,\statevec{y} \]
-
-\begin{align} \label{eqn:terminalemission}
-  \type(\bvec{b}) &= \Semit \\
-  \weight(\bvec{a}\to\bvec{x}\,\bvec{b}\,\bvec{y}) &=  t(\bvec{a},\bvec{b}) \cdot \totalemissionweight{\bvec{b}}{\bvec{x}\,\bvec{y}} \\
-  &= \left[ \prod_{m|a_m \ne b_m} \transweight{m}{a_m}{b_m} \right] \cdot \left[ \prod_m \emissionweight{m}{b_m}{x_m\,y_m} \right] ,
-\end{align}
-where we are defining states with no emissions to emit $(\Tnull\,\Tnull)$ with weight 1, 
-$\emissionweight{m}{b_m}{x_m\,y_m} = 1$ if $(x_m\,y_m) = (\Tnull\,\Tnull)$ and $\emit(b_m) = \Tnull$.
-
-The active node $n$ makes a transition to a state $b_n:\,\type(b_n) = \Sinsert$, emitting
-a terminal symbol pair $\emit(b_n) = (x_n\,y_n)$.  This symbol pair is passed down the guide tree to descendant nodes $\{m|\mDn, \type(a_m) \ne \Send\}$, 
-forcing them to make a transition from states of type $\Swait$ to states of type $\Smatch$, which can absorb
-terminal pairs $(x,y)$.  Qualitatively, this transition and emission could represent the evolution
-of two paired nucleotides along the subtree rooted at node $n$ of the complete guide tree.
-If one of $\bvec{x}$ or $\bvec{y}$ is null, then \eqref{eqn:terminalemission} could represent the evolution of
-a single unpaired nucleotide along the subtree rooted at node $n$.
-
-\paragraph{Left emission.}
-All terminals $\bvec{y}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission} are null.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(x_m\,y_m) = (\Tgap\,\Tnull)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(x_n\,y_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (x_{\parent(m)}\,\Tnull)$. \\
-  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tgap\,\Tnull) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull \end{array} \right. $
-\end{tabular}
-
-\paragraph{Right emission.}
-All terminals $\bvec{x}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission} are null.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(x_m\,y_m) = (\Tnull\,\Tgap)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(x_n\,y_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (\Tnull\,y_{\parent(m)})$. \\
-  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tnull\,\Tgap) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull . \end{array} \right. $
-\end{tabular}
-
-\paragraph{Paired emission.}
-
-There is at least one non-null terminal in both $\bvec{x}$ and $\bvec{y}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission}.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(x_m\,y_m) = (\Tgap\,\Tgap)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(x_n\,y_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (x_{\parent(m)}\,y_{\parent(m)})$. \\
-  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tgap\,\Tgap) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull . \end{array} \right. $
-\end{tabular}
-
-\subsubsection*{Bifurcations}
-\[ \statevec{a} \to \statevec{c}\,\statevec{b}\,\statevec{d} \]
-
-\begin{align} \label{eqn:nonterminalemission}
-  \type(\bvec{b}) &= \Sbifurc \\
-  \weight(\bvec{a}\to\bvec{c}\,\bvec{b}\,\bvec{d}) &=  t(\bvec{a},\bvec{b}) \cdot \totalemissionweight{\bvec{b}}{\bvec{c}\,\bvec{d}} \\
-  &= \left[ \prod_{m|\,a_m \ne b_m} \transweight{m}{a_m}{b_m} \right] \cdot \left[ \prod_m \emissionweight{m}{b_m}{c_m\,d_m} \right] .
-\end{align}
-where we are defining the emission weight of the $\Send$ nonterminal to be 1, $\emissionweight{m}{b_m}{c_m\,d_m} = 1$ if $c_m = \Send$ or $d_m = \Send$.
-
-Bifurcations are handled similarly to terminal emission.  The active node $n$ can undergo a bifurcation by making a transition 
-$a_n \to b_n,\, b_n:\,\type(b_n) = \Sinsert$, emitting a pair of nonterminals $\emit(b_n) = (c_n\,d_n)$.  Descendant nodes $\{m|\mDn,\type(a_m) \ne \Send\}$
-are forced to make a transition from states of type $\Swait$ to states of type $\Smatch$ which can absorb nonterminal pairs $(c\,d)$.
-All emissions are pairwise, so left and right bifurcations are represented as pairs $(\bvec{c}\,\bvec{d})$ where either $\bvec{c}$ or $\bvec{d}$ is null.
-If $\bvec{d}$ is null, then \eqref{eqn:nonterminalemission} could represent the insertion and subsequent evolution of a new RNA stem-loop structure.
-
-\paragraph{Left bifurcation.}
-All nonterminals $\bvec{d}$ in the transition $\bvec{a} \to \bvec{c}\,\bvec{b}\,\bvec{d}$ \eqref{eqn:nonterminalemission} are null.
-The nonterminals $\bvec{c}$ are the ``new'' states (for example, corresponding to a newly formed stem);
-the nonterminals $\bvec{b}$ are the states which will generate the (evolved) ancestral sequence.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(c_m\,d_m) = (\Send\,\Tnull)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(c_n\,d_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m)$ \\
-  & $(c_m\,d_m) = \emit(b_m) . $
-\end{tabular}
-
-\paragraph{Right bifurcation.}
-All nonterminals $\bvec{c}$ in the transition $\bvec{a} \to \bvec{c}\,\bvec{b}\,\bvec{d}$ \eqref{eqn:nonterminalemission} are null.
-The nonterminals $\bvec{d}$ are the ``new'' states;
-the nonterminals $\bvec{b}$ are the states which will generate the (evolved) ancestral sequence.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(c_m\,d_m) = (\Tnull\,\Send)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(c_n\,d_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m)$ \\
-  & $(c_m\,d_m) = \emit(b_m) . $
-\end{tabular}
-
-\paragraph{Paired bifurcation}
-There is at least one non-null nonterminal in both $\bvec{c}$ and $\bvec{d}$ in the transition $\bvec{a} \to \bvec{c}\,\bvec{b}\,\bvec{d}$ \eqref{eqn:nonterminalemission}.
-The nonterminals $\bvec{c}$ and $\bvec{d}$ are both ``new'' states;
-the nonterminals $\bvec{b}$ are the states which will generate the (evolved) ancestral sequence.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(c_m\,d_m) = (\Send\,\Send)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(c_n\,d_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m)$ \\
-  & $(c_m\,d_m) = \emit(b_m) . $
-\end{tabular}
-
-This paired-bifurcation is included for completeness--for example, it could be used to model symmetric loops--but it
-increases the complexity of grammar parsing.
-
-\subsubsection*{Transition to $\Send$}
-\[ \statevec{a} \to \Send \]
-
-\begin{align} \label{eqn:transtoend}
-  \weight(\bvec{a}\to\Send) &=  t(\bvec{a},\Send) \\
-  &= \prod_{m|\type(a_m)\ne\Send} \transweight{m}{a_m}{\Send} .
-\end{align}
-
-The singlet transducer associated with the highest active ancestral node,
-\beqn
-\hat{n} = \argmin_m \left\{ \type(a_m) \in \{\Sstart,\,\Sinsert\} \right\}
-\eeqn
-can make a transition to the state $\Send$,
-forcing the entire multi-sequence model to transition to $\Send$.\footnote{The node $\hat{n}$ which initiates the transition to $\Send$ 
-is the root of the greatest active subtree of the whole guide tree (called such because $a_m = \Send \, \forall \, m \ntrianglerighteq \hat{n}$).}
-If the branch transducer does not permit inserted bifurcations then $\hat{n} = 1$ always,
-but this is generically not true for a more general grammar
-(for example, see the TKF Structure Tree model).
-
-We require:
-
-\begin{tabular}{rl}
-  Nodes $m \ntrianglerighteq \hat{n}$: & $a_m = \Send$. \\
-  Node $\hat{n}$: & $\type(a_{\hat{n}}) \in \{\Sstart,\,\Sinsert\}$ \\
-  & $\exists$ a transition $a_{\hat{n}} \to \Send$.  \\
-  Nodes $m \vartriangleright \hat{n}$: & $\type(a_m) = \Swait$ \\
-  & $\exists$ a transition $a_m \to \Send$. \\
-\end{tabular}
-
-In many probabilistic models, the transition from a state of type $\Swait$ to the $\Send$ state has weight 1 conditional on absorbing 
-the $\Send$ symbol (called $\epsilon$ in formal grammar theory),
-but we here allow for a more general contribution to the total weight $\mathbb{F}$ of the transition.
-
-$\hat{n}$ and $\weight(\bvec{a}\to\Send)$ are so defined in order to ensure that there exists a direct path
-along which the end symbol can be passed down the tree.
-The grammar should be designed such that if the singlet transducer at node $\hat{n}$ can make a transition to $\Send$,
-then so can all machines at $\{m|m \trianglerighteq \hat{n}\}$.
-The TKF Structure Tree model (\secref{tkfst}) satisfies this condition.
-A more general approach is probably possible, but it involves summing over paths $a': a_m \to a' \to \Send$,
-handling possible bifurcations in these paths, etc.
-
-\newpage
-\section{The TKF Structure Tree model as a transducer} \seclabel{tkfst}
-
-We can cast the TKF Structure Tree (TKFST) model as a transducer.
-This means we can automatically deduce rules like those shown for the TKFST model in the main paper.
-
-\subsection{The single-sequence TKFST model as a singlet transducer}
-
-The states of the singlet transducer are shown in \tabref{tkfstsinglet}.
-Allowed transitions between these states are shown in the paper.
-
-\begin{table}[!ht]
-  \centering
-  \begin{tabular}{lrrrr}
-    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
-    $L$ & $\Sstart$ \\
-    $I_L$ & $\Sinsert$ & & $(x,\Tnull)$ & $p(x)$ \\
-    \\
-    $S$ & $\Sstart$ \\
-    $I_S$ & $\Sinsert$ & & $(x,y)$ & $p(x,y)$ \\
-    \\
-    $B$ & $\Sinsert$ & & $(L\,S)$ & 1 \\
-  \end{tabular}
-  \caption{
-    \tablabel{tkfstsinglet}
-    States of the singlet transducer of the TKF Structure Tree model \cite{Holmes2004}.
-    Singlet transducers can only have states of type $\Sstart$ or $\Sinsert$.
-    This is the indiegram-style transducer equivalent of the singlet TKFST SCFG in the main paper.
-  }
-\end{table}
-
-\subsection{The two-sequence TKFST model as a branch transducer}
-
-The states of the branch transducer are shown in \tabref{tkfstbranch}.
-Allowed transitions between these states are shown in the paper.
-
-\begin{table}[!ht]
-  \centering
-  \begin{tabular}{lrrrr}
-    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
-    $L$ & $\Sstart$ \\
-    $I_L$ & $\Sinsert$ & & $(u,\Tnull)$ & $p(u)$ \\
-    $M_L$ & $\Smatch$ & $(x,\Tnull)$ & $(u,\Tnull)$ & $p(u|x)$ \\
-    $D_L$ & $\Smatch$ & $(x,\Tnull)$ & $(\Tgap,\Tnull)$ & 1 \\
-    $W_L$ & $\Swait$ \\
-    \\
-    $S$ & $\Sstart$ \\
-    $I_S$ & $\Sinsert$ & & $(u,v)$ & $p(u,v)$ \\
-    $M_S$ & $\Smatch$ & $(x,y)$ & $(u,v)$ & $p(u,v|x,y)$ \\
-    $D_S$ & $\Smatch$ & $(x,y)$ & $(\Tgap,\Tgap)$ & 1 \\
-    $W_S$ & $\Swait$ \\
-    \\
-    $B_i$ & $\Sinsert$ & & $(L_i\,S_i)$ & 1 \\
-    $B$ & $\Smatch$ & $(L\,S)$ & $(L\,S)$ & 1 \\
-    $B_p$ & $\Smatch$ & $(L\,S)$ & $(L\,\Send)$ & 1 \\
-    $B_e$ & $\Smatch$ & $(L\,\Send)$ & $(L\,\Send)$ & 1 \\
-  \end{tabular}
-  \caption{\tablabel{tkfstbranch}
-    States of the branch transducer of the TKF Structure Tree model \cite{Holmes2004}.
-    States which have the same names as states of the singlet transducer in \tabref{tkfstsinglet}
-    are the branch equivalent of the corresponding singlet states
-    (e.g. a $\Smatch$ state might be the branch equivalent of an $\Sinsert$ state).
-    States $L_i$ and $S_i$ are the $\Sstart$ states of a sub-model (not shown) identical
-    in structure to the singlet transducer.  They are used to insert a new stem-loop structure.
-    This is the indiegram-style transducer equivalent of the conditional pair TKFST SCFG in the main paper.
-  }
-\end{table}
-
-\subsection{The multi-sequence TKFST model as a composite transducer}
-
-We can use the state graph construction algorithm described 
-in the paper and detailed in \secref{stategraph} to create a model of the simultaneous evolution
-of several sequences.
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics [scale=0.4] {figs/parent.pdf}
-  \caption{\figlabel{parent}
-    A simple example of transducer composition to build a multi-sequence model of two extant sequences.
-    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
-    A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
-    and two branch transducers (the gray boxes labeled $\Delta T_X$ and $\Delta T_Y$)
-    mutate it according to the specified multi-sequence model.
-    Gray nodes correspond to observed data and white nodes unobserved data.
-  }
-\end{figure}
-
-Consider the simple model shown in \figref{parent}.
-The state of the multi-sequence model describing this model is a 3-vector $\bvec{a} = \left( a_1,\,a_2,\,a_3 \right)$,
-where $a_1$ is the state of the singlet transducer generating the ancestral sequence $W$
-and $a_2$ and $a_3$ are the states of the branch transducers evolving $W$ into
-extant sequences $X$ and $Y$.
-
-We can show some of the allowed transitions of this multi-sequence model.
-The state of the branch transducer associated with the active node $n$ is shown in bold.
-
-\paragraph{Stem creation.}
-
-A stem is created at the root node $W$ (corresponding to a bifurcation in the singlet transducer $a_1$)
-and survives in the sequence $X$ at node 2 but is deleted in the sequence $Y$ at node 3.
-
-\begin{align}
-  \nakedthreevec{1:}{2:}{3:} \quad
-  \threevec{L}{L}{\mathbf{L}} &
-  \to \threevec{\mathbf{L}}{W_L}{W_L}
-  \to \threevec{B}{B}{B_p}
-  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{S}{\mathbf{S}}{\Send}
-  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\mathbf{S}}{W_S}{\Send} \\
-  & \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\mathbf{S}}{W_S}{\Send}
-\end{align}
-
-\paragraph{Stem insertion.}
-
-A stem sequence is inserted in sequence $X$ at node 2.
-
-\begin{align}
-  \nakedthreevec{1:}{2:}{3:} \quad
-  \threevec{L}{L}{\mathbf{L}} \to \threevec{L}{\mathbf{L}}{W_L}
-  \to \threevec{L}{B_i}{B}
-  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\Send}{S}{\mathbf{S}}
-  \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\Send}{\mathbf{S}}{W_S}
-\end{align}
-
-\paragraph{Stem termination.}
-
-All stem sequences are ended by (possibly empty) loop sequences.
-
-\begin{align}
-  \nakedthreevec{1:}{2:}{3:} \quad
-  \threevec{\mathbf{S}}{W_S}{W_S}
-  \to \threevec{B_e}{B_e}{B_e}
-  \to \threevec{\Send}{\Send}{\Send} \, \threevec{L}{L}{\mathbf{L}}
-  \to \threevec{\Send}{\Send}{\Send} \, \threevec{\Send}{\Send}{\Send}
-\end{align}
-
-The functions $\alpha(t)$, $\beta(t)$ and $\gamma(t)$ are parametrized by the insertion and deletion rates of the Structure Tree model.
-They are defined for loop sequences as
-\begin{align}
-  \kappa_1 &= \lambda_1 / \mu_1 \\
-  \alpha_1 &= \exp \left( -\mu_1 t \right) \\
-  \beta_1 &= \frac{\lambda_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) } \\
-  \gamma_1 &= 1 - \frac{\mu_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\left( 1 - \exp (- \mu_1 t) \right) \left(\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) \right) }
-\end{align}
-and similarly for stem sequences \cite{Holmes2004}.
-
-\newpage
-\input{tkfst.tex}
-
-\newpage
-\section{Software}
-
-We have written software tools in Perl and C++ implementing much of the theory described here.
-All tools are available from \darturl\ as part of the \dart\ software package for sequence analysis.
-
-\subsection{Automated grammar construction}
-We implemented our model construction algorithm on the star phylogeny (\figref{threeway}) 
-in a set of Perl scripts.
-Given a singlet transducer modeling ancestral sequences and structures and
-a branch transducer modeling structural evolution,
-the scripts generate C++ code to create the corresponding jointly-normalized SCFG.
-All possible models of structural evolution which can be represented by a Pair SCFG
-are permitted as input to the scripts,
-allowing for flexible and automated model design.
-
-Given files describing the singlet and branch transducers,
-including weights of all transitions which may be functions of evolutionary distance,
-the package ComposedTreeTransducer::FourWayComposedTT can automatically generate
-the state graph and transition matrix of a multi-sequence model of three extant sequences
-(\figref{threeway}).
-It removes the useless windback $\Snull$ states described in the paper and
-introduces effective direct transitions caused by bifurcations with possibly-empty children.
-The package ComposedTreeTransducer::TripletSCFG transforms a multi-sequence
-model created by ComposedTreeTransducer::FourWayComposedTT into the 
-corresponding jointly-normalized three-sequence SCFG (\secref{multigrammars})
-and generates C++ code to build the model.
-
-Example singlet and branch transducers files are provided for a simple 
-Pair HMM model, a simple Pair SCFG model and the full TKF Structure Tree model.
-
-\subsection{Reconstruction of ancestral structures}
-The program \indiegram\ can perform maximum-likelihood inference on the
-three-sequence SCFGs automatically generated by the ComposedTreeTransducer::TripletSCFG package.
-Complete or no structural information for the three extant sequences can be supplied
-as input.
-
-\subsection{Simulation of RNA family evolution}
-The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
-Another script (\animateevolsayer) can be used to make animations of the evolving RNA structures, in combination with the RNAplot program in the \viennarna\ package \cite{HofackerEtAl94}.
-
-\newpage
-\section{Computational experiments}
-
-\subsection{Reconstruction of covariant substitution histories}
-
-Our \xrate\ program is a multiple-alignment analysis tool
-that estimates rate parameters for a broad class of models, including covariant RNA substitution models \cite{KlostermanEtAl2006}.
-It can also be used to predict consensus secondary structures for alignments.
-We implemented ancestral sequence reconstruction in \xrate, including posterior probabilities that the reconstructions are correct.
-
-Given a multiple sequence alignment and a phylogenetic tree,
-\xrate\ estimates maximum-likelihood values of rate and probability parameters by Expectation Maximization.
-During parameter estimation, any unspecified ancestral sequences are summed out using Felsenstein's peeling algorithm.
-The new feature is that the program can now find the ancestral sequence that has the highest posterior probability,
-contingent on the maximum-likelihood estimates of the model parameters.
-During the parameter estimation and ancestral reconstruction steps, the secondary structure may be specified, or it may be summed out as a latent variable.
-
-For the ancestral reconstruction experiments described here,
-we assumed that alignment, phylogeny and secondary structure were known,
-but that ancestral sequences were unknown.
-
-A simple computational experiment demonstrates the need for covariant substitution models.
-We first used \xrate\ to estimate maximum-likelihood parameters for a covariant model of RNA base-pair substitution.
-These rates were estimated from ribosomal RNA alignments
-derived by \cite{DowellEddy2006} from the European rRNA database \cite{WuytsEtAl2004}.
-We also fit a ``naive'' general reversible point-substitution model to these alignments, again using \xrate.
-All subsequent results are implicitly conditioned on these maximum-likelihood rate estimates.
-Next, we used the companion program SIMGRAM to generate 5000 random alignments (including ancestral sequence),
-simulating on a 75-taxon phylogeny from an RNA gene family in the \RFAM\ database \cite{GriffithsJonesEtAl2003}.
-(Specifically, we chose the type-I Hammerhead ribozyme, one of the 5\% most populous \RFAM\ families.
- We repeated this experiment with other \RFAM\ phylogenies, but the general trends reported here were not dependent on the choice of tree.)
-We stripped the ancestral sequence out of the simulated alignments (leaving the true structure annotation intact),
-then re-estimated the ancestral sequence with \xrate, using both the (correct) covariant substitution model and the (naive) non-covariant point-substitution model.
-The imputed ancestral sequences were compared to the true sequences (known from the simulation),
-and the model-derived posterior probabilities were compared to the empirical accuracy of the corresponding reconstructed sequence.
-
-\begin{figure}[ht]
-  \begin{center}
-     \includegraphics [width=0.45\textwidth] {figs/hammer_paired.pdf}
-     \hspace*{\fill}
-     \includegraphics [width=0.45\textwidth] {figs/xval_paired.pdf}
-  \end{center}
-  \caption{
-    \figlabel{simulation}
-    Accuracy of confidence estimates for base-pair reconstruction in simulations (left) and cross-validations on real data (right).
-    Reconstructions were performed using \xrate\ \cite{KlostermanEtAl2006},
-    using both covariant base-pair substitution models
-    and naive, non-covariant, single-base point-substitution models.
-    The plots show the relationship between the posterior probability that a model-based base-pair reconstruction is correct (vertical axis),
-    and the empirical probability that reconstructions with that posterior probability are actually correct (horizontal axis).
-    If the model's posterior probability estimates are correct, these points should lie on a straight diagonal line.
-    Left: simulation test on the hammerhead ribozyme phylogeny.
-    A covariant model (estimated from rRNA alignments) was used to generate the simulated alignments on the given phylogeny; the same model was then used to reconstruct ancestral sequence and obtain confidence estimates.
-    As expected, the model accurately estimates the probability that the reconstructed base-pair is correct.
-    A naive, non-covariant model (the general reversible model, estimated from the same alignments) was also used for reconstruction and confidence estimates.
-    This model {\em under}-estimates the probability of correct reconstructions,
-    since its equilibrium frequency has a higher entropy than the covariant model:
-    i.e. it assigns probability more evenly across the sixteen possible base-pairs (as opposed to the covariant model, which strongly favors the six canonical base-pairs).
-    Right: cross-validation (``hold-one-out'') test using \RFAM\ alignments with published (verified) structures.
-    Here, the covariant model slightly {\em over}-estimates the probability of correctness.
-    The rRNA alignments (from which the model rates were estimated) have fewer non-canonical base-pairs than the \RFAM\ alignments,
-    so that the covariant model entropy is slightly {\em lower} than the empirical base-pair distribution;
-    this may explain the discrepancy.
-    The naive model's confidence estimates are correct
-    for high-accuracy reconstructions,
-    over-estimates for medium-accuracy reconstructions,
-    and under-estimates for low-accuracy reconstructions.
-  }
-\end{figure}
-
-\figref{simulation} (left) shows the results of these comparisons.
-When using the same (covariant) model for simulation and reconstruction, the posterior probabilities are an excellent unbiased estimate of the frequency with which the model reconstructions are correct.
-However, the naive point-substitution model, which does not include covariant substitutions at base-paired sites, systematically under-estimates these probabilities.
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | wc -l` / `wc -l $f`,"\n"}' Hammerhead_1.bptest
-In our simulation, the ancestral error rate for the naive model (4.9\%) was significantly higher than that of the covariant model (1.7\%).
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Hammerhead_1.bptest
-Furthermore, 74\% of incorrect base-pairs predicted by the naive model were non-canonical (i.e. not AU, CG, GC, UA, GU or UG),
-compared to only 2\% of incorrect predictions by the covariant model.
-
-\begin{figure}[p]
-  \begin{center}
-     \includegraphics [width=0.45\textwidth] {figs/hammer_hist.pdf}
-     \hspace*{\fill}
-     \includegraphics [width=0.45\textwidth] {figs/xval_hist.pdf}
-  \end{center}
-  \caption{
-    \figlabel{histograms}
-    Proportions of reconstructed base-pairs that are incorrect (and the subset of those that are non-canonical)
-    in simulations (left) and cross-validations on real data (right).
-    Base-pair reconstructions that are incorrect AND non-canonical
-    might potentially disrupt a synthetic reconstructed structure.
-    While the covariant model does not always get more base-pairs correct than the naive model,
-    it predicts fewer non-canonical base-pairs.
-    The sharpest difference between the models are in the lowest posterior-probability categories,
-    corresponding to low-confidence predictions (e.g. sequences on deep branches, or base-pairs with few homologues).
-    In this category, the covariant model also makes far fewer errors than the naive model.
-    See main text and caption to \figref{simulation} for details of simulation and cross-validation procedures.
-    \RFAM\ families with verified structures tended to be less phylogenetically diverse,
-    with roughly half the total branch length (on average), than the tree used for our simulations;
-    this may explain the relative paucity of low posterior-probability errors in the right-hand plot, compared to the left.
-% perl -e 'use Newick;use Stockholm::Database;$db=Stockholm::Database->from_file(shift);for$a(@$db){next if$a->columns>500 || @{$a->seqname}>400;$tree=Newick->from_string(@{$a->gf_NH});$b=0;for$l(@{$tree->branch_length}){$b+=$l}print"$b\n"}' ~/rfam/Training/Published.stock | meansd.pl
-% perl -e 'use Newick;use Stockholm::Database;$db=Stockholm::Database->from_file(shift);for$a(@$db){$tree=Newick->from_string(@{$a->gf_NH});$b=0;for$l(@{$tree->branch_length}){$b+=$l}print"$b\n"}' ~/rfam/Training/Hammerhead_1.stock
-  }
-\end{figure}
-
-We also performed cross-validation (hold-one-out) experiments. Starting from \RFAM\ alignments whose secondary structure is labeled ``Published'' (i.e. experimentally validated),
-we estimated phylogenetic trees using the neighbor-joining algorithm
-with the Jukes-Cantor model,
-then removed individual sequences
- and attempted to reconstruct them using \xrate.
-(This is possible because the substitution models are time-reversible, so we can treat any node as ancestral.)
-In this experiment, the error rates for base-pair reconstructions again differed by a few percent, though both error rates were higher than in the simulation test
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | wc -l` / `wc -l $f`,"\n"}' Published.xval-bptest
-(covariant model: 11\%, non-covariant model: 16\%).
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Published.xval-bptest
-Of the incorrectly-reconstructed base-pairs, the non-covariant model again predicted more non-canonical pairs (40\%) than the covariant model (17\%),
-especially at lower confidence levels (\figref{histograms}, right).
-
-Posterior probability estimates for both models were
-systematically higher in the cross-validation test than they were in the simulation (\figref{simulation}, right).
-The presence of this trend in both models
-suggests that it may be due to differences between the European rRNA alignments
-(from which the rates were estimated)
-and the \RFAM\ alignments (which were used in the cross-validation test).
-This is confirmed by our empirical observations of
-covariant rate matrices estimated from the two alignment datasets:
-base-pairs in the \RFAM\ alignments are more frequently
-non-canonical, and appear to evolve faster (relative to single-stranded regions),
-than base-pairs in the rRNA alignments
-(for details, see {\tt biowiki.org/RnaModels}).
-Repeating the cross-validation experiment with rates estimated from the \RFAM\ alignments,
-we find the covariant model predicted slightly fewer non-canonical base-pairs
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Published.pubxval-bptest
-(14\% rather than 17\%)
-and produced slightly more accurate posterior probability estimates (data not shown).
-Although this is no longer a strict cross-validation (since parameters were estimated from the test dataset),
-it nevertheless emphasizes the value of correct models.
-%Another possible source of errors in the cross-validation (compared to the simulation)
-%is inaccurately estimated phylogenetic topologies and branch lengths;
-%this might be improvable by using a covariant model in the phylogenetic estimation step (rather than Jukes-Cantor),
-%and/or by using Bayesian phylogenetic estimation tools (rather than neighbor-joining).
-
-Accurate estimates of posterior probabilities are likely to be important:
-sub-optimal predictions are a significant piece of the reconstruction puzzle \cite{WilliamsEtAl2006,PollockChang2007,ChangEtAl2007}.
-Ancestral reconstruction by probabilistic modeling does not yield a single definitive sequence, but rather a probability distribution over sequences.
-Combinatorial synthesis by degenerate primer assembly has been advocated as a means of sampling from this distribution \cite{Gaucher2007}.
-Nucleotide-perfect reconstructions may not always be possible:
-it may be more productive to aim for accurate confidence estimates
-(including sub-optimal predictions)
-than to focus exclusively on getting everything right the first time.
-
-Based on these tests, we conclude that deep phylogenetic reconstructions of ribosomes, and other RNAs, will require covariant substitution models that take account of RNA secondary structure.
-We may reasonably deduce that indel reconstructions will, similarly, need to take account of RNA structure.
-These results, therefore, strongly motivate the development of multi-sequence models for RNA structure, such as the TKF Structure Tree.
-
-\section{Terminological asides}
-
-This subsection contains some asides about terminology,
-and can be omitted on a first reading of this paper.
-The intent of this subsection is to highlight connections to related fields
-in molecular evolution and computer science.
-
-\subsection{Phylogenetic grammars}
-
-We here describe some relationships to grammar models in current use in molecular evolution.
-
-% Removing this sentence since I do not think that describing the relationship between transducers and SCFGs as an "injective mapping" will make it any clearer for most readers -- it is a bit like describing a string as a "concatenative monoid" -- IH
-%There exists an injective mapping from the evolutionary models generated by our model-construction algorithm to multi-sequence SCFGs. In other words, 
-Given a singlet transducer, a phylogenetic tree, and a mapping from tree to branch transducers,
-the transducer composition algorithm returns a corresponding multi-sequence SCFG
-which models the joint probability distribution $P(X_1 , ... , X_{n+m})$.
-
-Several related bioinformatics grammars can be represented within this multi-SCFG framework,
-allowing for substantial re-use of prior modeling theory:
-\begin{itemize}
-\item String transducers (which are similar to HMMs) can be viewed as special cases of parse-tree transducers,
-without the facility to model long-distance correlations (base-pairs) or bifurcated structures.
-\item Phylo-HMMs and phylo-SCFGs \cite{KlostermanEtAl2006}
-can be viewed as special cases of (respectively) string transducers and parse-tree transducers.
-(Phylo-HMMs and phylo-SCFGs typically model substitutions but not indels,
-and so can be viewed as compositions of a richly-featured singleton transducer with a restricted class of branch transducers.)
-\item Sakakibara \cite{Sakakibara2003} has described ``Pair HMMs on tree structures'' that are closely related to our parse-tree transducers
-(though without the composition algorithms for scoring multiple alignments).
-\item Sakakibara has also described pair stochastic tree-adjoining grammars, or TAGs \cite{Sakakibara2004}.
-TAGs are a formal generalization of SCFGs to
-mildly context-sensitive grammars \cite{MamitsukaAbe94,JoshiSchabes97,RivasEddy2000}.
-We speculate that the conditionally-normalized transducer analogues of such Pair-TAGs may encompass the
-phylogenetic model of covariant RNA substitutions and indels on pseudoknotted structures
-described by \cite{MeyerMiklos2007},
-just as parse-tree transducers encompass phylo-SCFGs.
-\end{itemize}
-
-\subsection{Statistical Alignment}
-
-The term {\bf Statistical Alignment} was introduced by Hein \cite{HeinEtal2000}
-to describe DP algorithms for phylogenetic grammars.
-Although this term has been adopted by others in molecular evolution \cite{Miklos2002,Metzler2003,DrummondRambaut2007,SatijaEtAl2008},
-we've avoided it (except in this section) for the following four reasons:
-(i) Several RNA alignment methods already use another probability distribution
-that's completely different to (though equally as valid as) the phylogenetic probability distribution over alignments:
-namely, the Boltzmann probability distribution over secondary structures
-\cite{GorodkinEtAl97,MathewsTurner02,HofackerEtAl2004}.
-(ii) Many alignment tools make use of prior and posterior distributions over pairwise alignments
-\cite{Miyazawa94,BucherHofmann96,Holmes98,BrayPachter2004,DoEtAl2004,Holmes2005,LoytynojaGoldman2005,SchwartzPachter2007,BradleyPachterHolmes2008};
-all of these are arguably ``statistical'' alignment tools as well.
-(iii) Some of the above-cited methods use sum-over-pairs alignment metrics, taking the expectation of such metrics over pairwise alignment posteriors,
-and finding the multiple alignment that maximizes this expectation \cite{DoEtAl2004,SchwartzPachter2007,BradleyPachterHolmes2008};
-others do progressive multiple alignment using pair grammars \cite{BrayPachter2004,Holmes2005,LoytynojaGoldman2005}.
-Speculatively, such pair-based heuristics may (in some sense) approximate probabilistic alignment inference with true phylogenetic likelihoods;
-similar convergences have, at least, been demonstrated between other pair-based heuristics \cite{EickmeyerEtAl2008}.
-Thus, even using Hein's strict definition,
-it's possible that a broad range of methods systematically {\em approximate} ``Statistical Alignment'';
-this cautions against using the term too exclusively.
-(iv) Even more broadly, many alignment tools make use of statistics to estimate alignment ``significance''.
-In this frequentist approach, probability distributions for alignment scores (particularly Gumbel distributions)
-are estimated for random sequences.
-The tail integrals of these distributions then yield P-values for high-scoring database search results
-\cite{KarlinAltschul93,Mott2000,Eddy2008}.
-Thus, while the term ``Statistical Alignment'' has a quite specific technical meaning in molecular evolution
-(a meaning which certainly encompasses the algorithms we describe here),
-the broader bioinformatics literature may have competing claims to the term.
-In any case, the goal of this work is statistical {\em reconstruction} of RNA protosequences,
-rather than alignment {\em per se}.
-
-\subsection{Parse trees and stacks}
-
-Grammars are abstract models for languages;
-finite-state automata are the corresponding abstract models for those languages' parsers.
-Most of this paper uses the grammatical view of RNA structure evolution;
-in this short section we digress briefly to mention the automata-theoretic view of these models.
-
-The two non-local features that set SCFGs apart from HMMs are bifurcations and paired emissions \cite{Durbin98}.
-In RNA structure analysis, these features are used to model (respectively) branched-loops and base-pairs.
-
-A stochastic pushdown automaton can also model these features.
-Such an automaton is essentially an HMM with a LIFO stack (LIFO = ``Last In, First Out'').
-
-The pushdown automaton can parse two sequences simultaneously, aligning them and their parse trees.
-It handles bifurcations and paired emissions by pushing symbols onto its stack:
-a bifurcation by pushing a nonterminal, and a paired emission by pushing a terminal.
-The pushed terminals and nonterminals are popped off the stack when the machine reaches its $\Send$ state.
-
-The conditionally-normalized version of a pushdown automaton is a pushdown transducer:
-a machine that absorbs {\em and} parses an input sequence (or absorbs an already-parsed input sequence)
-and emits a (parsed) output sequence.
-This is the machine that we describe as a parse-tree transducer.
-
 \bibliography{../latex-inputs/alignment,../latex-inputs/reconstruction,../latex-inputs/duplication,../latex-inputs/genomics,../latex-inputs/ncrna,../latex-inputs/url}
 
+\clearpage
+\section*{Algorithms}
+
+\begin{algorithm}[!ht]
+  \SetLine
+  \SetKwFunction{rnd}{Random[0,1]}
+  \KwIn{Rule $\rho'' \in \grules''$}
+  \KwOut{Parse (sub)tree $T' \in \gtrees'$}
+
+  \Switch{$\rho''$}{
+    \Case{$X \to Y$}{
+      \eIf{\rnd $< P(X \to Y) / q_{XY}$}{
+        \KwRet{$(X \to Y)$} \;
+      }{
+        Let $z = \sum_W q_{XW} P'(W \to Y)$ \;
+        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to Y) / z$ \;
+        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to Y)$} \;
+      }
+    }
+    \Case{$X \to \epsilon$}{
+      \eIf{\rnd $< P'(X \to \epsilon) / P'(\epsilon | X)$}{
+        \KwRet{$(X \to \epsilon)$} \;
+      }{
+        Let $z = \sum_W q_{XW} P'(W \to \epsilon)$ \;
+        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to \epsilon) / z$ \;
+        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to \epsilon)$} \;
+      }
+    }
+    \Other{
+      \KwRet{$(\rho'')$} \;
+    }
+  }
+  \caption{\label{alg:restoreTransitions}
+    Subroutine $\mathrm{restoreTransitions}(\rho'')$ for the null cycle elimination procedure.
+  }
+\end{algorithm}
+
+\begin{algorithm}[!ht]
+  \SetLine
+  \SetKwFunction{rnd}{Random[0,1]}
+  \KwIn{Rule $\rho' \in \grules'$}
+  \KwOut{Parse (sub)tree $T \in \gtrees$}
+
+  \Switch{$\rho'$}{
+    \Case{$B^0 \to \epsilon$}{
+      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
+      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
+      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
+      \KwRet{$(B \to T_L\ T_R)$} \;
+    }
+    \Case{$B' \to L'$}{
+      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
+      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
+      \KwRet{$(B \to L\ T_R)$} \;
+    }
+    \Case{$B' \to R'$}{
+      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
+      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
+      \KwRet{$(B \to T_L\ R)$} \;
+    }
+    \Case{$B^2 \to L'\ R'$}{
+      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
+      \KwRet{$(B \to L\ R)$} \;
+    }
+    \lCase{$B^0 \to B'$}{
+      \KwRet{$()$} \;
+    }
+    \lCase{$B' \to B^2$}{
+      \KwRet{$()$} \;
+    }
+    \Other{
+      Let $\rho$ be the original rule in $\grules$ \;
+      \KwRet{$(\rho)$} \;
+    }
+  }
+
+  \caption{\label{alg:restoreBifurcations}
+    Subroutine $\mathrm{restoreBifurcations}(\rho')$ for the null cycle elimination procedure.
+  }
+\end{algorithm}
+
+\begin{algorithm}[!ht]
+  \SetLine
+  \SetKwFunction{rnd}{Random[0,1]}
+  \KwIn{Nonterminal $A \in \gnons$}
+  \KwOut{Parse tree $T \in \gtrees: \mathrm{seq}(T) = \epsilon, \mathrm{root}(T) = A$}
+
+  \eIf{$A$ is a bifurcation state, $A \to X\ Y$,}{
+    $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
+    $T_Y \leftarrow \mathrm{sampleNullSubtree}(Y)$ \;
+    \KwRet{$(A \rightarrow T_X\ T_Y)$} \;
+  }{
+    \eIf{\rnd $< P(A \to \epsilon) / P(\epsilon|A)$}{
+      \KwRet{$(A \to \epsilon)$} \;
+    }{
+      Let $z = \sum_Y P(A \to Y) P(\epsilon | Y)$ \;
+      Sample state $X$ from probability distribution $P(X) = P(A \to X) P(\epsilon | X) / z$ \;
+      $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
+      \KwRet{$(A \rightarrow T_X)$} \;
+    }
+  }
+
+  \caption{\label{alg:sampleNullSubtree}
+    Subroutine $\mathrm{sampleNullSubtree}(A)$ for the null cycle elimination procedure.
+  }
+\end{algorithm}
+
 \end{document}

### 2009-06-01 s3.tex
Text S3 (software)

--- 2009-05-29 supplementary.tex
@@
   {\Large
-    \textbf{\titlestring: \supptext}
+    \textbf{\titlestring: \supptext{3}}
   }
@@
 \newpage
-\tableofcontents
-
-\newpage
-\section{Exact elimination of SCFG null cycles}
-
-The following section describes how to transform any SCFG so as to remove null cycles while preserving access to the full posterior probability distribution over parses, including parses with null cycles.
-
-\subsection{Definitions}
-
-Let $G = ( \Omega, \gnons, \grules, P )$ be a stochastic context-free grammar (SCFG)
-consisting of
-a set of terminal symbols $\Omega$,
-a set of nonterminal symbols (a.k.a. ``states'') $\gnons$,
-a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
-and a probability function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.
-
-% (The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
-% The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)
-
-Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e., rules in which $A$ appears on the left),
-and let $\rightside{A}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right, including bifurcations which also generate another nonterminal as well as $A$).
-Let $\allrules{A} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
-Define these also on sets of nonterminals, e.g., $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.
-
-Let $\gtrees$ be the set of all parse trees for $G$.
-Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
-then define the parse tree likelihood $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.
-
-If $\sum_{T \in \gtrees} P(T) = 1$, we say that $G$ is {\em probabilistically normalized by parse tree}.
-If $\sum_{\rho \in \leftside{A}} P(\rho) = 1$ for all $A \in \gnons$, we say that $G$ is {\em probabilistically normalized by production rule}.
-Note that normalization by production rule $\Rightarrow$ normalization by parse tree.
-
-Let $S \in \Omega^\ast$ denote a terminal sequence.
-Let $\mathrm{seq}:\gtrees \to \Omega^\ast$ be the function mapping a parse tree to its terminal sequence.
-and let $\mathrm{root}:\gtrees \to \gnons$ be the function returning the root nonterminal of a parse tree.
-The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
-Of particular relevance to null state elimination is the probability $P(\epsilon|A)$,
-where $\epsilon$ is the empty sequence.
-This is the probability that a nonterminal $A$ will expire without generating any sequence.
-
-Following earlier formalisms \cite{Durbin98,HolmesRubin2002a,Holmes2005},
-we say that the grammar $G$ has {\em RNA normal-form rules} if
-each rule in $\grules$ takes one of the following four forms:
-
-\noindent \hrulefill
-\begin{description}
-\item[Termination rules] with one nonterminal on the left and the empty string on the right
-\begin{eqnarray*}
-A & \to & \epsilon
-\end{eqnarray*}
-\item[Transition rules] with one nonterminal on the left and the right
-\begin{eqnarray*}
-A & \to & B
-\end{eqnarray*}
-\item[Bifurcation rules] with one nonterminal on the left and two on the right
-\begin{eqnarray*}
-A & \to & B\ C
-\end{eqnarray*}
-\item[Emission rules] with one nonterminal on the left and right and at least one terminal on the right
-\begin{eqnarray*}
-A & \to & x\ B \\
-A & \to & B\ y \\
-A & \to & x\ B\ y
-\end{eqnarray*}
-\end{description}
-\noindent \hrulefill
-
-\noindent
-Here $A,B,C \in \gnons$ and $x,y \in \Omega$.
-
-Suppose that grammar $G$ has RNA normal-form rules.
-We further say that $G$ has {\em RNA normal-form states} if
-each nonterminal (state) $A \in \gnons$ takes one of the following forms:
-
-\noindent \hrulefill
-\begin{description}
-\item[Null states]: $\leftside{A}$ contains only transition and termination rules.
-\item[Bifurcation states]: $\leftside{A}$ contains exactly one bifurcation rule, $A \to X\ Y$, where $X \neq Y$ and $X,Y$ are both null states.
-\item[Emit states]: $\leftside{A}$ contains only emission rules. Further, $\leftside{A} = \rightside{B}$ for some null state $B$.
-This null state $B$ is called $A$'s {\em post-emit} state.
-\end{description}
-\noindent \hrulefill
-
-Define a {\em null cycle} to be a nonterminal $A$
-and a sequence of rules $\rho_1, \rho_2 \ldots \rho_k$ that, when applied consecutively to $A$, leave $A$ unchanged;
-that is, $\rho_k(\rho_{k-1} \ldots \rho_2(\rho_1(A))) = A$.
-Define a {\em null subtree} to be a null cycle using at least one bifurcation rule.
-
-Suppose that $G = ( \Omega, \gnons, \grules, P )$ and $G' = ( \Omega, \gnons', \grules', P' )$ are two grammars.
-We say that $G$ and $G'$ are {\em equivalent in sequence} if there is a mapping between nonterminals, $f:\gnons \to \gnons'$,
-such that $P(S|A) = P(S|f(A))$.
-We say that $G$ and $G'$ are {\em equivalent in parse} if there is a mapping between parse trees, $g:\gtrees \to \gtrees'$,
-such that $P'(T') = \sum_{T:T'=g(T)} P(T)$ and we can define
-\[
-P(T|T') = P(T|g(T)=T') = \frac{P(T)}{P'(T')}
-\]
-
-Note that equivalence in parse $\Rightarrow$ equivalence in sequence.
-
-Suppose that a grammar $G$ contains null cycles.
-We seek to transform $G$ into a grammar $G''$ that is equivalent in parse and sequence, but has no null cycles.
-If $G$ is normalized by parse tree, then $G''$ will be too; but $G''$ is not (necessarily) normalized by production rule.
-
-We do the transformation in two steps, $G \to G' \to G''$.
-We also provide a stochastic ``null cycle restoration'' algorithm for sampling from $P(T|T'')$.
-
-\subsection{Eliminating null bifurcations}
-
-Note first that any SCFG can be transformed into an equivalent one with RNA normal-form states by adding null states.
-Without loss of generality, we therefore consider SCFGs with RNA normal-form states.
-
-Let $G = ( \Omega, \gnons, \grules, P )$ be such an SCFG.
-Let $\gnulls \subseteq \gnons$ be the set of null states in $\gnons$, {\bf excluding} post-emit states.
-Let $\gemits \subset \gnons$ be the set of post-emit states.
-Let $\gbifs \subseteq \gnons$ be the set of bifurcation states.
-
-We here define the grammar $G'$,
-\[
-G' = \left( \Omega, \gnons', \grules', P' \right)
-\]
-which is equivalent to $G$ in sequence but has no null subtrees.
-
-The new states $\gnons'$ are defined as follows.
-Start with $\gnons$; remove bifurcation states;
-then introduce a new state $N'$ for every null $N \in \gnulls$ and three new states $B^0, B', B^2$ for every bifurcation $B \in \gbifs$.
-
-\[
-\gnons' = 
-\left( \bigcup_{N \in \gnulls} \{ N' \} \right) \cup \left( \bigcup_{B \in \gbifs} \{ B^0, B', B^2 \} \right) \cup \gnons \setminus \gbifs
-\]
-
-The idea is that $N'$ (in $G'$) is a null state that is equivalent to null state $N$ (in $G$) for non-empty sequence only.
-That is, for any terminal sequence $S$,
-
-\[
-P'(S|N') = \left\{ \begin{array}{ll} P(S|N) & \mbox{if $S \neq \epsilon$} \\ 0 & \mbox{if $S = \epsilon$} \end{array} \right.
-\]
-
-Similarly, $B'$ is a null state that is equivalent to bifurcation state $B$ for non-empty sequence,
-while $B^2$ is a null state that is is equivalent to $B$ for parse trees where both children of $B$ are non-empty.
-
-In contrast, $B^0$ is {\bf exactly} equivalent to $B$ in sequence.
-However, null subtrees in $B^0$ are explicitly accounted for,
-their probabilities factored into the transitions $B' \to L'$ and $B' \to R'$ and the termination $B^0 \to \epsilon$,
-using the inside probabilities for empty sequences, $P(\epsilon|L)$ and $P(\epsilon|R)$.
-
-The probabilities $P(\epsilon|X)$ are related by the following system of equations
-
-\[
-P(\epsilon|X) = P(X \to \epsilon) + \sum_Y P(X \to Y) P(\epsilon|Y) + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R)
-\]
-
-which are nonlinearly coupled (via bifurcations) but may be solved numerically, e.g., by the Newton-Raphson method,
-or by iterated approximation starting from a lower bound $P(\epsilon|X) \geq 0$.
-
-The new rules and their probabilities are
-\[
-\begin{array}{rrcl}
-\forall \rho \in (\grules \setminus \allrules{\gbifs}): & P'(\rho)             & = & P(\rho) \\
-                                \forall N \in \gnulls: & P'(N' \to \epsilon)  & = & 0  \\
-                          \ldots \forall B \in \gbifs: & P'(N' \to B')        & = & P(N \to B) \\
-                         \ldots \forall M \in \gnulls: & P'(N' \to M')        & = & P(N \to M) \\
-                         \ldots \forall E \in \gemits: & P'(N' \to E)         & = & P(N \to E) \\
-               \forall (A \to B) \in \rightside{\gbifs}: & P'(A \to B^0)        & = & P(A \to B) \\
-              \forall (B \to LR) \in \leftside{\gbifs}: & P'(B^0 \to \epsilon) & = & P(B \to LR)\ P(\epsilon | L)\ P(\epsilon | R) \\
-                                                       & P'(B^0 \to B')       & = & 1 \\
-                                                       & P'(B' \to \epsilon)  & = & 0 \\
-                                                       & P'(B' \to L')        & = & P(B \to LR)\ P(\epsilon | R) \\
-                                                       & P'(B' \to R')        & = & P(B \to LR)\ P(\epsilon | L) \\
-                                                       & P'(B' \to B^2)       & = & 1 \\
-                                                       & P'(B^2 \to L'\ R')   & = & P(B \to LR) \\
-\end{array}
-\]
-
-Note that grammar $G'$, like $G$, has RNA normal-form states.
-Note also that $G'$ is not, in general, normalized by production rule;
-however, $G'$ is equivalent to $G$ in parse and is therefore normalized by parse tree (if $G$ is).
-
-\subsection{Eliminating null states}
-
-We now proceed to eliminate null cycles from $G'$.
-Since null subtrees have been eliminated, the remaining null cycles use only transition rules.
-
-We create the grammar
-\[
-G'' = \left( \Omega, \gnons', \grules'', P'' \right)
-\]
-which has the same nonterminals as $G'$, but different rules and rule probabilities.
-
-Let $\gnulls' \subseteq \gnons'$ be the set of null states in $\gnons'$, {\bf including} post-emit states.
-
-Define ${\bf t}$, the {\em transition matrix} of $G'$, as $t_{XY} = P'(X \to Y)$ for all $X,Y \in \gnons'$.
-The effective transition probability $q_{XY}$ between two states $X,Y$
-sums over all paths through null states (in the summand, $n$ is the length of the null path):
-\[
-{\bf q} = \sum_{n=0}^\infty {\bf t}^n = ({\bf 1}-{\bf t})^{-1}
-\]
-Here ${\bf 1}$ is the identity matrix.
-The matrix inverse may be computed in the usual ways (Gauss-Jordan elimination, LU decomposition, etc.)
-
-Since the bifurcation states of $G'$ explicitly generate non-empty sequence,
-the system of equations relating the probabilities $P'(\epsilon|X)$ is completely linear
-and may be solved in the same way.
-Let $u_X = P'(\epsilon|X)$ and $v_X = P'(X \to \epsilon)$.
-Then
-
-\[
-{\bf u} = {\bf v} + {\bf tu}
-\]
-
-whose solution is ${\bf u} = {\bf qv}$.
-Thus $P'(\epsilon|X) = \sum_Y q_{XY} P'(Y \to \epsilon)$.
-
-We now define $P''$ as follows.
-\begin{itemize}
-\item For all rules $\rho \in \grules'$ that are {\bf not} transitions or terminations, set $P''(\rho) = P'(\rho)$.
-\item For terminations from null states $X \in \gnulls'$, set $P''(X \to \epsilon) = u_X$.
-(NB this may create some terminations $X \to \epsilon$ that were not present in $G'$.)
-\item For transitions from null states $X \in \gnulls'$ to non-null states $Y \notin \gnulls'$, set $P''(X \to Y) = q_{XY}$.
-(NB this may create some transitions $X \to Y$ that were not present in $G'$.)
-\item For transitions {\bf to} null states $X \in \gnulls'$, set $P''(A \to X) = 0$.
-\end{itemize}
-
-Although we have left null states in $G''$, they now have no incoming transitions
-and are inaccessible unless they are post-bifurcation states (i.e.,states which appear on the right-hand side of bifurcation rules), post-emit states, or the start (root) state.
-All other null states can therefore be dropped from $\gnons''$.
-
-\subsection{Restoring null states}
-
-Suppose that $\rho''$ is a rule in $G''$.
-Algorithm~\ref{alg:restoreTransitions} samples from the distribution of equivalent parse subtrees in $G'$ (possibly containing null cycles).
-In order to sample from $P(T'|T'')$ we need simply map Algorithm~\ref{alg:restoreTransitions} to each rule in $T''$.
-
-For parameter estimation by Expectation Maximization and some other applications, it is useful to know the expected number
-of times that a transition was used, summed over the posterior distribution of parse trees (including those with null cycles).
-If $n''(\rho'')$ is the expected number of times that transition $\rho''$ was used according to an Inside-Outside computation on $G''$,
-then the corresponding expectations $n'(\rho')$ are given by
-
-\begin{eqnarray*}
-       n'(X \to Y) & = & n''(X \to Y) \frac{P''(X \to Y)}{q_{XY}} + \sum_Z n''(X \to Z) \frac{P''(X \to Y) q_{YZ}}{q_{XZ}} \\
-n'(X \to \epsilon) & = & n''(X \to \epsilon) \frac{P''(X \to \epsilon)}{P''(\epsilon | X)}
- + \sum_W n''(W \to \epsilon) \frac{q_{WX} P''(X \to \epsilon)}{P''(\epsilon | W)}
-\end{eqnarray*}
-
-Expectations for other rules (bifurcations and emissions) are the same for $G'$ as $G''$.
-
-\subsection{Restoring null bifurcations}
-
-Suppose that $\rho'$ is a rule in $G'$.
-Algorithm~\ref{alg:restoreBifurcations} samples from the distribution of equivalent parse subtrees in $G$ (possibly containing null subtrees).
-This algorithm also calls Algorithm~\ref{alg:sampleNullSubtree}, which samples from the distribution of empty subtrees rooted at a particular nonterminal.
-In order to sample from $P(T|T')$ we need simply map Algorithm~\ref{alg:restoreBifurcations} to each rule in $T'$.
-
-If $n'(\rho')$ is the expected number of times that rule $\rho'$ was used by $G'$,
-then the corresponding expectations $n(\rho)$ are given by
-
-\begin{eqnarray*}
-    n(B \to L\ R) & = & n'(B' \to L') + n'(B' \to R') + n'(B^0 \to L'\ R') + d(B \to L\ R) \\
-       n(X \to Y) & = & n'(X \to Y) + d(X \to Y) \\
-n(X \to \epsilon) & = & n'(X \to \epsilon) + d(X \to \epsilon)
-\end{eqnarray*}
-Expectations for emissions are the same for $G$ as $G'$.
-In the above expressions $d(\rho)$ is the expected usage of rule $\rho$ by null subtrees:
-
-\[
-d(\rho) = \sum_{(B \to LR) \in \grules} \left[
-n'(B' \to L') c_R(\rho)
-+ n'(B' \to R') c_L(\rho)
-+ n'(B^0 \to \epsilon) (c_L(\rho) + c_R(\rho))
-\right]
-\]
-
-where $c_W(\rho)$ is the expected usage of rule $\rho$ by an empty parse tree rooted at $W$,
-given by
-
-\begin{eqnarray*}
-c_X(\rho) P(\epsilon | X) & = & P(X \to \epsilon) \delta_{\rho=(X \to \epsilon)}
- + \sum_Y P(X \to Y) P(\epsilon|Y) \left( c_Y(\rho) + \delta_{\rho=(X \to Y)} \right)
-\\ & &
- + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R) \left( c_L(\rho) + c_R(\rho) + \delta_{\rho=(X \to LR)} \right)
-\end{eqnarray*}
-
-where $\delta_U$ is the Kronecker delta (1 if condition $U$ is true, 0 if it is false).
-Note that in contrast to the system of equations for $P(\epsilon|X)$, this is a linear system of equations
-of the form ${\bf c} = {\bf Mc} + {\bf k}$, which may be solved by matrix inversion:
-${\bf c} = ({\bf 1}-{\bf M})^{-1} {\bf k}$.
-
-\newpage
-\section{Two-sequence transducer models}
-
-We here give formal definitions of two-sequence models and the state machines which
-generate the factored probability distribution $P (X, Y | \Delta T) = P (X) \cdot P (Y | X, \Delta T)$,
-where the marginal $P(X)$ is generated by a {\bf singlet transducer}
-and the conditional $P (Y | X, \Delta T)$ by a {\bf branch transducer}.
-We refer to the branch transducer as $\theta$, so the conditional distribution
-is more precisely $P (Y | X, \Delta T, \theta)$.
-
-\subsection{Formal grammars} \seclabel{grammars}
-
-There is a close relationship between formal grammars and the singlet and branch transducer abstract state machines.
-By labeling the nonterminals of a regular or stochastic context-free grammar (SCFG) as states of an abstract machine, allowing these states to absorb and emit 
-appropriate terminal symbols,\footnote{The relationship between formal grammars and abstract state machines is more transparent
-in the Mealy machine view, but the Moore machine view turns out to be more useful for our purposes.},
-and carefully assigning transition and emission weights,
-we can create a state machine which generates the same language, with the same distribution of weights, as that produced by the original grammar.\footnote{We speak of
-weights associated with a transition rather than probabilities in order to allow for more general models.}
-We therefore use the terms ``nonterminal'' and ``state'' interchangeably
-and refer to the ``parse tree'' generated by singlet and branch transducers.
-
-Phrased more precisely, there is an isomorphism between the singlet and branch transducers of two-sequence models
-and Pair SCFGs.  Practically speaking, this means that for every joint distribution $P(X, Y | \Delta T)$ 
-generated by a singlet and branch transducers, there exists a Pair SCFG which generates the same distribution.
-
-\subsection{States, transitions and emissions}
-A singlet transducer has states of type $\Sstart$ and $\Sinsert$.
-Each state $\phi\in\Phi$ of the branch transducer has type
-\[ \type(\phi) \in \lbrace\Sstart,\,\Sinsert,\,\Smatch,\,\Swait,\,\Send\rbrace \, . \]
-State typing is as follows:
-\begin{itemize}
-  \item A $\Sstart$ state begins a branch of the parse tree.
-  \item An $\Sinsert$ state emits, but does not absorb, symbols.
-  \item A $\Smatch$ state absorbs and (possibly) emits symbols.
-  \item A $\Swait$ state is a null state which allows a branch transducer to ``pause''
-    while it waits for an input symbol from another transducer.
-    In the two-sequence case, the input symbols is emitted by the singlet transducer generating
-    the ancestral sequence.
-  \item An $\Send$ state ends a branch of the parse tree.
-\end{itemize}
-The names of state types are similar to the $\{\Sstart,\,\Sinsert,\,\Smatch,\,\Send\}$ states of the familiar Pair HMM.
-Deletions are handled as a special case by states of type $\Smatch$ which absorb but do not emit symbols pairs.
-Only states of type $\Smatch$ can absorb symbol pairs $(x,y)$.
-
-Bifurcations in the grammar, when considered as emission of nonterminals, can be handled analogously
-to terminal emission.  States $\phi: \type(\phi) \in \{\Sinsert,\,\Smatch\}$ can emit nonterminal pairs $(c\,d)$,
-where $c$ or $d$ can be the null symbol, and the emitted pairs $(c\,d)$ can be absorbed by states of type $\Smatch$.
-
-The emission weight of a nonterminal pair $(c\,d)$ from the state $b: \type(b) = \Smatch$, 
-conditional on absorption of a nonterminal pair $(l\,m)$, is 
-\[ e_b ( c\,d|l\,m,\theta ) \, . \]
-
-The functions $\emit()$ and $\absorb()$ are defined to return the emission or absorption of a particular state,
-\begin{align*}
-  \emit(\phi) &:= \left\{\begin{array}{ll} (x,y) & x,y = \mathrm{terminal\,\,or\,\,null} \\(c\,d) & c,d = \mathrm{nonterminal\,\,or\,\,null}. \end{array} \right. \\
-  \absorb(\phi) &:= \left\{\begin{array}{ll} (x,y) & x,y = \mathrm{terminal\,\,or\,\,null} \\(c\,d) & c,d = \mathrm{nonterminal\,\,or\,\,null}. \end{array} \right.
-\end{align*}
-
-We frequently use the notation ${}^{uv}\phi^{xy}$ to indicate that a state of type $\Smatch$ absorbs a symbol pair $(u,v)$
-and emits a pair $(x,y)$.
-The notation for bifurcations is slightly different:
-A bifurcation state which left-emits a nonterminal $d$ and makes a transition to a state $\phi$ with weight 1 is written as $\bifurc{B}{d}{\phi}$.
-
-A transition between states $a$ and $b$ of the branch transducer $\theta$ has a weight
-\[ t(a,b|\theta) = t(a \to b|\theta). \]
-Terminal emission is handled by states $\phi: \type(\phi) \in \{\Smatch,\,\Sinsert\}$, which emit symbol pairs $(x,y)$,
-where $x$ or $y$ can be the null symbol.
-In this paired-emission perspective, left single-terminal emissions $x$ are represented as $(x\,\Tnull)$;
-right single-terminal emissions are handled similarly.
-The weight of an emission of a terminal pair $(x,y)$ from a state $b: \type(b) = \Smatch$, 
-conditioned on absorption of a terminal pair $(u,v)$, is 
-\[ e_b ( x,y|u,v,\theta ) \, . \]
-Recall that the emission weights of states of type $\Smatch$ are defined conditioned upon the absorbed symbols.
-
-\newpage
-\section{Multiple-sequence transducer models} \seclabel{evolutionarymodels}
-
-We can use our two-sequence models to construct a model of many sequence related by a
-guide phylogenetic tree.
-The guide tree specifies the (conjectured) phylogenetic relationship of all sequences.
-A singlet transducer, which emits, but does not absorb, symbols, lies at the root of the guide tree and 
-serves as a generative model of the ancestral sequence.
-A branch transducer represents the evolution of an ancestral sequence into a single descendant 
-sequence, that is, the action of evolution along the single-branch tree (Ancestor $\to$ Descendant).
-To represent the evolution of an ancestral sequence into many descendant sequences
-(whose phylogenetic relationship is specified by the guide tree),
-we place a branch transducer on each branch of the guide tree to create a multiple-sequence model.
-
-If the branch grammar has no bifurcations and only left or right emissions are allowed, then
-the language generated is a regular string language and the corresponding jointly normalized abstract state machine is 
-an HMM.  In this simplest case our formalism for creating a multiple-sequence model reduces to that given 
-by \cite{Holmes2003} for combining HMMs on a guide tree.
-
-\subsection{The guide tree}
-The nodes of the tree are labeled $1,\dots,N$ in the order reached by any preorder depth-first traversal of the tree.
-The length of each branch $(\parent(m) \to m)$ is given by the evolutionary time $t_m$.
-To specify ancestor-descendant relationships, we introduce notation:
-$\mDn$ ($\mNDn$) means node $m$ is descended from (not descended from) node $n$, and
-$\mDEn$ ($\mNDEn$) means node $m$ is descended from or identical to (not descended from and not identical to) node $n$.
-
-\subsection{States, transitions and emissions}
-The multiple-sequence model is formed by the composition/intersection of $(N-1)$ branch transducers 
-such that there is a branch transducer on each branch $(\parent(m) \to m);\, m = 2,\dots,N$ of the guide tree and a singlet transducer at the root node.
-Our framework allows for the placement of different branch transducers, with a unique set of nonterminals (state space) $\Phi$ and
-allowed transitions between states and corresponding weights, on each branch.  $\theta^{(m)}$ denotes 
-the branch transducer governing evolution along the branch $(\parent(m) \to m)$ of the guide tree.
-
-States of the multiple-sequence model are represented by as N-dimensional vectors $\bvec{a}$,
-\[ \bvec{a} = \statevec{a} \, . \]
-These states are typed as
-\[ \type(\bvec{a}) \in \lbrace\Sstart,\,\Semit,\,\Sbifurc,\,\Snull,\,\Send\rbrace \, . \]
-State typing is as follows:
-\begin{itemize}
-  \item A $\Sstart$ state begins a branch of the parse tree of one or more of the $N$ sequences on the phylogenetic tree.
-  \item An $\Semit$ state emits symbols (terminals) to one or more of the $N$ sequences.
-  \item A $\Sbifurc$ state corresponds to a bifurcation in the parse tree of
-    one or more of the $N$ sequences.
-  \item A $\Snull$ state corresponds to any non-$\Send$ state which represents neither
-    an emission or bifurcation.
-  \item An $\Send$ state ends a branch of the parse tree.
-\end{itemize}
-States are typed according to the transition by which they are reachable
-(details of the typing are given in \secref{transitions}).
-
-Transitions and emissions of the multiple-sequence model are defined in terms of the transitions and emissions 
-of the branch transducers at each node as well as the singlet transducer at the root node.
-The weight of a transition $\bvec{a} \to \bvec{b}$ is therefore
-\beqn
-t(\bvec{a},\bvec{b}) = \prod_{m|a_m \ne b_m} \transweight{m}{a_m}{b_m} \, ,
-\eeqn
-and the weight of an emission $(\bvec{x}\,\bvec{y})$ from a state $\bvec{b}$ is
-\beqn
-\totalemissionweight{\bvec{b}}{\bvec{x}\,\bvec{y}} = \prod_m e_{b_m}\left(x_m\,y_m|x_{\parent(m)}\,y_{\parent(m)},\theta^{(m)}\right) \, .
-\eeqn
-We frequently will not explicitly write out the conditional dependence on the absorbed terminals $(x_{\parent(m)}\,y_{\parent(m)})$,
-but the reader should keep in mind that in general emission weights will depend on the absorbed symbols.
-
-\subsection{Formal grammars} \seclabel{multigrammars}
-
-Analogously to the case with two-sequence models (\secref{grammars}),
-there exists a one-to-one mapping between the multiple-sequence models generated by our model-construction algorithm 
-and multi-sequence SCFGs.
-In other words, given a singlet and branch transducers of a two-sequence model, as well
-as a guide tree relating the extant sequences, there exists a corresponding multi-sequence SCFG
-which generates the same joint probability distribution $P(X_1 , ... , X_n)$.
-
-\subsection{Constructing the state graph} \seclabel{stategraph}
-
-As described in the paper,
-we need a way to efficiently construct the state graph of the multi-sequence model,
-where the state graph consists of a list of accessible states and the possible transitions between them.
-This state graph can be constructed by an uninformed depth-first search,
-where at each step of the search we obtain the possible child nodes 
-by applying one of the following possible transitions of the multi-sequence model:
-\begin{enumerate}
-\item {\bf Null transition}:
-  The state of a single branch transducer is updated, with no terminal emission or bifurcation.
-\item {\bf Terminal emission}:
-  A state makes a transition to a $\Sinsert$ state.  The emitted symbol is passed down the guide
-  tree to all descendant branch transducers, which transition to states of type $\Smatch$.
-\item {\bf Bifurcation}:
-  A state makes a transition to a special $\Sinsert$ state which emits a new branch of the parse tree.
-  The emitted symbol is passed down the guide tree to all descendant branch transducers, 
-  which transition to states of type $\Smatch$.
-\item {\bf End transition}:
-  The singlet transducer associated with the root sequence can transition to the $\Send$ state,
-  signaling that this parse tree is finished.
-\end{enumerate}
-
-Each of these transitions is explained in detail in the following section.
-
-\subsection{Allowed transitions} \seclabel{transitions}
-
-Following \cite{Holmes2003},  we let transitions of the multi-sequence model begin at the active node and cascade down the guide tree
-as appropriate.  Unless defined otherwise, node $n$ is the active node of the multi-sequence model with state $\bvec{a}$, 
-\begin{align}
-  n &= \activenode{a} \\
-  &= \mathrm{argmax}_m \left\{\type(a_m) \notin \{\Swait,\,\Send\}\right\} \, . \label{eqn:activenode} 
-\end{align}
-Each possible allowed transition is obtained by making a valid change (updating) the 
-state of the singlet or one or more of the branch transducers of the multi-sequence model.
-
-\subsubsection*{Null Transitions}
-
-\[ \statevec{a} \to \statevec{b} \]
-
-\begin{align} \label{eqn:nulltransition}
-  \type(\bvec{b}) &= \Snull  \\
-  \weight(\bvec{a}\to\bvec{b}) &= t(\bvec{a},\bvec{b}) \\
-  &= \transweight{n}{a_n}{b_n} .
-\end{align}
-
-This transition updates the state of the branch transducer at the active node $n$ of the guide tree, leaving 
-the rest unchanged, with no corresponding terminal emission or bifurcation in the grammar.
-Nodes other than the active node do not change state, $a_m = b_m \,\forall\, m \ne n$, and the only 
-allowable transitions of this form are to states $b_n:\,\type(b_n) \in \{\Sstart,\,\Swait\}$.  Transitions
-to states of type $\Sinsert$ or $\Smatch$ result in emissions, and transitions to the end state $\Send$
-are handled as a special case.
-
-\subsubsection*{Terminal Emission} \seclabel{terminalemission}
-\[ \statevec{a} \to \statevec{x}\,\statevec{b}\,\statevec{y} \]
-
-\begin{align} \label{eqn:terminalemission}
-  \type(\bvec{b}) &= \Semit \\
-  \weight(\bvec{a}\to\bvec{x}\,\bvec{b}\,\bvec{y}) &=  t(\bvec{a},\bvec{b}) \cdot \totalemissionweight{\bvec{b}}{\bvec{x}\,\bvec{y}} \\
-  &= \left[ \prod_{m|a_m \ne b_m} \transweight{m}{a_m}{b_m} \right] \cdot \left[ \prod_m \emissionweight{m}{b_m}{x_m\,y_m} \right] ,
-\end{align}
-where we are defining states with no emissions to emit $(\Tnull\,\Tnull)$ with weight 1, 
-$\emissionweight{m}{b_m}{x_m\,y_m} = 1$ if $(x_m\,y_m) = (\Tnull\,\Tnull)$ and $\emit(b_m) = \Tnull$.
-
-The active node $n$ makes a transition to a state $b_n:\,\type(b_n) = \Sinsert$, emitting
-a terminal symbol pair $\emit(b_n) = (x_n\,y_n)$.  This symbol pair is passed down the guide tree to descendant nodes $\{m|\mDn, \type(a_m) \ne \Send\}$, 
-forcing them to make a transition from states of type $\Swait$ to states of type $\Smatch$, which can absorb
-terminal pairs $(x,y)$.  Qualitatively, this transition and emission could represent the evolution
-of two paired nucleotides along the subtree rooted at node $n$ of the complete guide tree.
-If one of $\bvec{x}$ or $\bvec{y}$ is null, then \eqref{eqn:terminalemission} could represent the evolution of
-a single unpaired nucleotide along the subtree rooted at node $n$.
-
-\paragraph{Left emission.}
-All terminals $\bvec{y}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission} are null.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(x_m\,y_m) = (\Tgap\,\Tnull)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(x_n\,y_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (x_{\parent(m)}\,\Tnull)$. \\
-  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tgap\,\Tnull) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull \end{array} \right. $
-\end{tabular}
-
-\paragraph{Right emission.}
-All terminals $\bvec{x}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission} are null.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(x_m\,y_m) = (\Tnull\,\Tgap)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(x_n\,y_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (\Tnull\,y_{\parent(m)})$. \\
-  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tnull\,\Tgap) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull . \end{array} \right. $
-\end{tabular}
-
-\paragraph{Paired emission.}
-
-There is at least one non-null terminal in both $\bvec{x}$ and $\bvec{y}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission}.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(x_m\,y_m) = (\Tgap\,\Tgap)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(x_n\,y_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (x_{\parent(m)}\,y_{\parent(m)})$. \\
-  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tgap\,\Tgap) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull . \end{array} \right. $
-\end{tabular}
-
-\subsubsection*{Bifurcations}
-\[ \statevec{a} \to \statevec{c}\,\statevec{b}\,\statevec{d} \]
-
-\begin{align} \label{eqn:nonterminalemission}
-  \type(\bvec{b}) &= \Sbifurc \\
-  \weight(\bvec{a}\to\bvec{c}\,\bvec{b}\,\bvec{d}) &=  t(\bvec{a},\bvec{b}) \cdot \totalemissionweight{\bvec{b}}{\bvec{c}\,\bvec{d}} \\
-  &= \left[ \prod_{m|\,a_m \ne b_m} \transweight{m}{a_m}{b_m} \right] \cdot \left[ \prod_m \emissionweight{m}{b_m}{c_m\,d_m} \right] .
-\end{align}
-where we are defining the emission weight of the $\Send$ nonterminal to be 1, $\emissionweight{m}{b_m}{c_m\,d_m} = 1$ if $c_m = \Send$ or $d_m = \Send$.
-
-Bifurcations are handled similarly to terminal emission.  The active node $n$ can undergo a bifurcation by making a transition 
-$a_n \to b_n,\, b_n:\,\type(b_n) = \Sinsert$, emitting a pair of nonterminals $\emit(b_n) = (c_n\,d_n)$.  Descendant nodes $\{m|\mDn,\type(a_m) \ne \Send\}$
-are forced to make a transition from states of type $\Swait$ to states of type $\Smatch$ which can absorb nonterminal pairs $(c\,d)$.
-All emissions are pairwise, so left and right bifurcations are represented as pairs $(\bvec{c}\,\bvec{d})$ where either $\bvec{c}$ or $\bvec{d}$ is null.
-If $\bvec{d}$ is null, then \eqref{eqn:nonterminalemission} could represent the insertion and subsequent evolution of a new RNA stem-loop structure.
-
-\paragraph{Left bifurcation.}
-All nonterminals $\bvec{d}$ in the transition $\bvec{a} \to \bvec{c}\,\bvec{b}\,\bvec{d}$ \eqref{eqn:nonterminalemission} are null.
-The nonterminals $\bvec{c}$ are the ``new'' states (for example, corresponding to a newly formed stem);
-the nonterminals $\bvec{b}$ are the states which will generate the (evolved) ancestral sequence.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(c_m\,d_m) = (\Send\,\Tnull)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(c_n\,d_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m)$ \\
-  & $(c_m\,d_m) = \emit(b_m) . $
-\end{tabular}
-
-\paragraph{Right bifurcation.}
-All nonterminals $\bvec{c}$ in the transition $\bvec{a} \to \bvec{c}\,\bvec{b}\,\bvec{d}$ \eqref{eqn:nonterminalemission} are null.
-The nonterminals $\bvec{d}$ are the ``new'' states;
-the nonterminals $\bvec{b}$ are the states which will generate the (evolved) ancestral sequence.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(c_m\,d_m) = (\Tnull\,\Send)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(c_n\,d_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m)$ \\
-  & $(c_m\,d_m) = \emit(b_m) . $
-\end{tabular}
-
-\paragraph{Paired bifurcation}
-There is at least one non-null nonterminal in both $\bvec{c}$ and $\bvec{d}$ in the transition $\bvec{a} \to \bvec{c}\,\bvec{b}\,\bvec{d}$ \eqref{eqn:nonterminalemission}.
-The nonterminals $\bvec{c}$ and $\bvec{d}$ are both ``new'' states;
-the nonterminals $\bvec{b}$ are the states which will generate the (evolved) ancestral sequence.
-
-\begin{tabular}{rl}
-  Nodes $\mNDEn$: & $a_m = b_m$. \\
-  & $(c_m\,d_m) = (\Send\,\Send)$. \\
-  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
-  & $(c_n\,d_n) = \emit(b_n)$. \\
-  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
-  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m)$ \\
-  & $(c_m\,d_m) = \emit(b_m) . $
-\end{tabular}
-
-This paired-bifurcation is included for completeness--for example, it could be used to model symmetric loops--but it
-increases the complexity of grammar parsing.
-
-\subsubsection*{Transition to $\Send$}
-\[ \statevec{a} \to \Send \]
-
-\begin{align} \label{eqn:transtoend}
-  \weight(\bvec{a}\to\Send) &=  t(\bvec{a},\Send) \\
-  &= \prod_{m|\type(a_m)\ne\Send} \transweight{m}{a_m}{\Send} .
-\end{align}
-
-The singlet transducer associated with the highest active ancestral node,
-\beqn
-\hat{n} = \argmin_m \left\{ \type(a_m) \in \{\Sstart,\,\Sinsert\} \right\}
-\eeqn
-can make a transition to the state $\Send$,
-forcing the entire multi-sequence model to transition to $\Send$.\footnote{The node $\hat{n}$ which initiates the transition to $\Send$ 
-is the root of the greatest active subtree of the whole guide tree (called such because $a_m = \Send \, \forall \, m \ntrianglerighteq \hat{n}$).}
-If the branch transducer does not permit inserted bifurcations then $\hat{n} = 1$ always,
-but this is generically not true for a more general grammar
-(for example, see the TKF Structure Tree model).
-
-We require:
-
-\begin{tabular}{rl}
-  Nodes $m \ntrianglerighteq \hat{n}$: & $a_m = \Send$. \\
-  Node $\hat{n}$: & $\type(a_{\hat{n}}) \in \{\Sstart,\,\Sinsert\}$ \\
-  & $\exists$ a transition $a_{\hat{n}} \to \Send$.  \\
-  Nodes $m \vartriangleright \hat{n}$: & $\type(a_m) = \Swait$ \\
-  & $\exists$ a transition $a_m \to \Send$. \\
-\end{tabular}
-
-In many probabilistic models, the transition from a state of type $\Swait$ to the $\Send$ state has weight 1 conditional on absorbing 
-the $\Send$ symbol (called $\epsilon$ in formal grammar theory),
-but we here allow for a more general contribution to the total weight $\mathbb{F}$ of the transition.
-
-$\hat{n}$ and $\weight(\bvec{a}\to\Send)$ are so defined in order to ensure that there exists a direct path
-along which the end symbol can be passed down the tree.
-The grammar should be designed such that if the singlet transducer at node $\hat{n}$ can make a transition to $\Send$,
-then so can all machines at $\{m|m \trianglerighteq \hat{n}\}$.
-The TKF Structure Tree model (\secref{tkfst}) satisfies this condition.
-A more general approach is probably possible, but it involves summing over paths $a': a_m \to a' \to \Send$,
-handling possible bifurcations in these paths, etc.
-
-\newpage
-\section{The TKF Structure Tree model as a transducer} \seclabel{tkfst}
-
-We can cast the TKF Structure Tree (TKFST) model as a transducer.
-This means we can automatically deduce rules like those shown for the TKFST model in the main paper.
-
-\subsection{The single-sequence TKFST model as a singlet transducer}
-
-The states of the singlet transducer are shown in \tabref{tkfstsinglet}.
-Allowed transitions between these states are shown in the paper.
-
-\begin{table}[!ht]
-  \centering
-  \begin{tabular}{lrrrr}
-    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
-    $L$ & $\Sstart$ \\
-    $I_L$ & $\Sinsert$ & & $(x,\Tnull)$ & $p(x)$ \\
-    \\
-    $S$ & $\Sstart$ \\
-    $I_S$ & $\Sinsert$ & & $(x,y)$ & $p(x,y)$ \\
-    \\
-    $B$ & $\Sinsert$ & & $(L\,S)$ & 1 \\
-  \end{tabular}
-  \caption{
-    \tablabel{tkfstsinglet}
-    States of the singlet transducer of the TKF Structure Tree model \cite{Holmes2004}.
-    Singlet transducers can only have states of type $\Sstart$ or $\Sinsert$.
-    This is the indiegram-style transducer equivalent of the singlet TKFST SCFG in the main paper.
-  }
-\end{table}
-
-\subsection{The two-sequence TKFST model as a branch transducer}
-
-The states of the branch transducer are shown in \tabref{tkfstbranch}.
-Allowed transitions between these states are shown in the paper.
-
-\begin{table}[!ht]
-  \centering
-  \begin{tabular}{lrrrr}
-    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
-    $L$ & $\Sstart$ \\
-    $I_L$ & $\Sinsert$ & & $(u,\Tnull)$ & $p(u)$ \\
-    $M_L$ & $\Smatch$ & $(x,\Tnull)$ & $(u,\Tnull)$ & $p(u|x)$ \\
-    $D_L$ & $\Smatch$ & $(x,\Tnull)$ & $(\Tgap,\Tnull)$ & 1 \\
-    $W_L$ & $\Swait$ \\
-    \\
-    $S$ & $\Sstart$ \\
-    $I_S$ & $\Sinsert$ & & $(u,v)$ & $p(u,v)$ \\
-    $M_S$ & $\Smatch$ & $(x,y)$ & $(u,v)$ & $p(u,v|x,y)$ \\
-    $D_S$ & $\Smatch$ & $(x,y)$ & $(\Tgap,\Tgap)$ & 1 \\
-    $W_S$ & $\Swait$ \\
-    \\
-    $B_i$ & $\Sinsert$ & & $(L_i\,S_i)$ & 1 \\
-    $B$ & $\Smatch$ & $(L\,S)$ & $(L\,S)$ & 1 \\
-    $B_p$ & $\Smatch$ & $(L\,S)$ & $(L\,\Send)$ & 1 \\
-    $B_e$ & $\Smatch$ & $(L\,\Send)$ & $(L\,\Send)$ & 1 \\
-  \end{tabular}
-  \caption{\tablabel{tkfstbranch}
-    States of the branch transducer of the TKF Structure Tree model \cite{Holmes2004}.
-    States which have the same names as states of the singlet transducer in \tabref{tkfstsinglet}
-    are the branch equivalent of the corresponding singlet states
-    (e.g. a $\Smatch$ state might be the branch equivalent of an $\Sinsert$ state).
-    States $L_i$ and $S_i$ are the $\Sstart$ states of a sub-model (not shown) identical
-    in structure to the singlet transducer.  They are used to insert a new stem-loop structure.
-    This is the indiegram-style transducer equivalent of the conditional pair TKFST SCFG in the main paper.
-  }
-\end{table}
-
-\subsection{The multi-sequence TKFST model as a composite transducer}
-
-We can use the state graph construction algorithm described 
-in the paper and detailed in \secref{stategraph} to create a model of the simultaneous evolution
-of several sequences.
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics [scale=0.4] {figs/parent.pdf}
-  \caption{\figlabel{parent}
-    A simple example of transducer composition to build a multi-sequence model of two extant sequences.
-    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
-    A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
-    and two branch transducers (the gray boxes labeled $\Delta T_X$ and $\Delta T_Y$)
-    mutate it according to the specified multi-sequence model.
-    Gray nodes correspond to observed data and white nodes unobserved data.
-  }
-\end{figure}
-
-Consider the simple model shown in \figref{parent}.
-The state of the multi-sequence model describing this model is a 3-vector $\bvec{a} = \left( a_1,\,a_2,\,a_3 \right)$,
-where $a_1$ is the state of the singlet transducer generating the ancestral sequence $W$
-and $a_2$ and $a_3$ are the states of the branch transducers evolving $W$ into
-extant sequences $X$ and $Y$.
-
-We can show some of the allowed transitions of this multi-sequence model.
-The state of the branch transducer associated with the active node $n$ is shown in bold.
-
-\paragraph{Stem creation.}
-
-A stem is created at the root node $W$ (corresponding to a bifurcation in the singlet transducer $a_1$)
-and survives in the sequence $X$ at node 2 but is deleted in the sequence $Y$ at node 3.
-
-\begin{align}
-  \nakedthreevec{1:}{2:}{3:} \quad
-  \threevec{L}{L}{\mathbf{L}} &
-  \to \threevec{\mathbf{L}}{W_L}{W_L}
-  \to \threevec{B}{B}{B_p}
-  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{S}{\mathbf{S}}{\Send}
-  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\mathbf{S}}{W_S}{\Send} \\
-  & \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\mathbf{S}}{W_S}{\Send}
-\end{align}
-
-\paragraph{Stem insertion.}
-
-A stem sequence is inserted in sequence $X$ at node 2.
-
-\begin{align}
-  \nakedthreevec{1:}{2:}{3:} \quad
-  \threevec{L}{L}{\mathbf{L}} \to \threevec{L}{\mathbf{L}}{W_L}
-  \to \threevec{L}{B_i}{B}
-  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\Send}{S}{\mathbf{S}}
-  \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\Send}{\mathbf{S}}{W_S}
-\end{align}
-
-\paragraph{Stem termination.}
-
-All stem sequences are ended by (possibly empty) loop sequences.
-
-\begin{align}
-  \nakedthreevec{1:}{2:}{3:} \quad
-  \threevec{\mathbf{S}}{W_S}{W_S}
-  \to \threevec{B_e}{B_e}{B_e}
-  \to \threevec{\Send}{\Send}{\Send} \, \threevec{L}{L}{\mathbf{L}}
-  \to \threevec{\Send}{\Send}{\Send} \, \threevec{\Send}{\Send}{\Send}
-\end{align}
-
-The functions $\alpha(t)$, $\beta(t)$ and $\gamma(t)$ are parametrized by the insertion and deletion rates of the Structure Tree model.
-They are defined for loop sequences as
-\begin{align}
-  \kappa_1 &= \lambda_1 / \mu_1 \\
-  \alpha_1 &= \exp \left( -\mu_1 t \right) \\
-  \beta_1 &= \frac{\lambda_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) } \\
-  \gamma_1 &= 1 - \frac{\mu_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\left( 1 - \exp (- \mu_1 t) \right) \left(\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) \right) }
-\end{align}
-and similarly for stem sequences \cite{Holmes2004}.
-
-\newpage
-\input{tkfst.tex}
-
-\newpage
 \section{Software}
 
-We have written software tools in Perl and C++ implementing much of the theory described here.
 All tools are available from \darturl\ as part of the \dart\ software package for sequence analysis.
@@
 \subsection{Automated grammar construction}
-We implemented our model construction algorithm on the star phylogeny (\figref{threeway}) 
+We implemented our model construction algorithm on the three-taxon star phylogeny
 in a set of Perl scripts.
@@
 the package ComposedTreeTransducer::FourWayComposedTT can automatically generate
-the state graph and transition matrix of a multi-sequence model of three extant sequences
-(\figref{threeway}).
+the state graph and transition matrix of a multi-sequence model of three extant sequences.
 It removes the useless windback $\Snull$ states described in the paper and
@@
 model created by ComposedTreeTransducer::FourWayComposedTT into the 
-corresponding jointly-normalized three-sequence SCFG (\secref{multigrammars})
+corresponding jointly-normalized three-sequence SCFG
 and generates C++ code to build the model.
@@
 
-\newpage
-\section{Computational experiments}
-
-\subsection{Reconstruction of covariant substitution histories}
-
-Our \xrate\ program is a multiple-alignment analysis tool
-that estimates rate parameters for a broad class of models, including covariant RNA substitution models \cite{KlostermanEtAl2006}.
-It can also be used to predict consensus secondary structures for alignments.
-We implemented ancestral sequence reconstruction in \xrate, including posterior probabilities that the reconstructions are correct.
-
-Given a multiple sequence alignment and a phylogenetic tree,
-\xrate\ estimates maximum-likelihood values of rate and probability parameters by Expectation Maximization.
-During parameter estimation, any unspecified ancestral sequences are summed out using Felsenstein's peeling algorithm.
-The new feature is that the program can now find the ancestral sequence that has the highest posterior probability,
-contingent on the maximum-likelihood estimates of the model parameters.
-During the parameter estimation and ancestral reconstruction steps, the secondary structure may be specified, or it may be summed out as a latent variable.
-
-For the ancestral reconstruction experiments described here,
-we assumed that alignment, phylogeny and secondary structure were known,
-but that ancestral sequences were unknown.
-
-A simple computational experiment demonstrates the need for covariant substitution models.
-We first used \xrate\ to estimate maximum-likelihood parameters for a covariant model of RNA base-pair substitution.
-These rates were estimated from ribosomal RNA alignments
-derived by \cite{DowellEddy2006} from the European rRNA database \cite{WuytsEtAl2004}.
-We also fit a ``naive'' general reversible point-substitution model to these alignments, again using \xrate.
-All subsequent results are implicitly conditioned on these maximum-likelihood rate estimates.
-Next, we used the companion program SIMGRAM to generate 5000 random alignments (including ancestral sequence),
-simulating on a 75-taxon phylogeny from an RNA gene family in the \RFAM\ database \cite{GriffithsJonesEtAl2003}.
-(Specifically, we chose the type-I Hammerhead ribozyme, one of the 5\% most populous \RFAM\ families.
- We repeated this experiment with other \RFAM\ phylogenies, but the general trends reported here were not dependent on the choice of tree.)
-We stripped the ancestral sequence out of the simulated alignments (leaving the true structure annotation intact),
-then re-estimated the ancestral sequence with \xrate, using both the (correct) covariant substitution model and the (naive) non-covariant point-substitution model.
-The imputed ancestral sequences were compared to the true sequences (known from the simulation),
-and the model-derived posterior probabilities were compared to the empirical accuracy of the corresponding reconstructed sequence.
-
-\begin{figure}[ht]
-  \begin{center}
-     \includegraphics [width=0.45\textwidth] {figs/hammer_paired.pdf}
-     \hspace*{\fill}
-     \includegraphics [width=0.45\textwidth] {figs/xval_paired.pdf}
-  \end{center}
-  \caption{
-    \figlabel{simulation}
-    Accuracy of confidence estimates for base-pair reconstruction in simulations (left) and cross-validations on real data (right).
-    Reconstructions were performed using \xrate\ \cite{KlostermanEtAl2006},
-    using both covariant base-pair substitution models
-    and naive, non-covariant, single-base point-substitution models.
-    The plots show the relationship between the posterior probability that a model-based base-pair reconstruction is correct (vertical axis),
-    and the empirical probability that reconstructions with that posterior probability are actually correct (horizontal axis).
-    If the model's posterior probability estimates are correct, these points should lie on a straight diagonal line.
-    Left: simulation test on the hammerhead ribozyme phylogeny.
-    A covariant model (estimated from rRNA alignments) was used to generate the simulated alignments on the given phylogeny; the same model was then used to reconstruct ancestral sequence and obtain confidence estimates.
-    As expected, the model accurately estimates the probability that the reconstructed base-pair is correct.
-    A naive, non-covariant model (the general reversible model, estimated from the same alignments) was also used for reconstruction and confidence estimates.
-    This model {\em under}-estimates the probability of correct reconstructions,
-    since its equilibrium frequency has a higher entropy than the covariant model:
-    i.e. it assigns probability more evenly across the sixteen possible base-pairs (as opposed to the covariant model, which strongly favors the six canonical base-pairs).
-    Right: cross-validation (``hold-one-out'') test using \RFAM\ alignments with published (verified) structures.
-    Here, the covariant model slightly {\em over}-estimates the probability of correctness.
-    The rRNA alignments (from which the model rates were estimated) have fewer non-canonical base-pairs than the \RFAM\ alignments,
-    so that the covariant model entropy is slightly {\em lower} than the empirical base-pair distribution;
-    this may explain the discrepancy.
-    The naive model's confidence estimates are correct
-    for high-accuracy reconstructions,
-    over-estimates for medium-accuracy reconstructions,
-    and under-estimates for low-accuracy reconstructions.
-  }
-\end{figure}
-
-\figref{simulation} (left) shows the results of these comparisons.
-When using the same (covariant) model for simulation and reconstruction, the posterior probabilities are an excellent unbiased estimate of the frequency with which the model reconstructions are correct.
-However, the naive point-substitution model, which does not include covariant substitutions at base-paired sites, systematically under-estimates these probabilities.
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | wc -l` / `wc -l $f`,"\n"}' Hammerhead_1.bptest
-In our simulation, the ancestral error rate for the naive model (4.9\%) was significantly higher than that of the covariant model (1.7\%).
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Hammerhead_1.bptest
-Furthermore, 74\% of incorrect base-pairs predicted by the naive model were non-canonical (i.e. not AU, CG, GC, UA, GU or UG),
-compared to only 2\% of incorrect predictions by the covariant model.
-
-\begin{figure}[p]
-  \begin{center}
-     \includegraphics [width=0.45\textwidth] {figs/hammer_hist.pdf}
-     \hspace*{\fill}
-     \includegraphics [width=0.45\textwidth] {figs/xval_hist.pdf}
-  \end{center}
-  \caption{
-    \figlabel{histograms}
-    Proportions of reconstructed base-pairs that are incorrect (and the subset of those that are non-canonical)
-    in simulations (left) and cross-validations on real data (right).
-    Base-pair reconstructions that are incorrect AND non-canonical
-    might potentially disrupt a synthetic reconstructed structure.
-    While the covariant model does not always get more base-pairs correct than the naive model,
-    it predicts fewer non-canonical base-pairs.
-    The sharpest difference between the models are in the lowest posterior-probability categories,
-    corresponding to low-confidence predictions (e.g. sequences on deep branches, or base-pairs with few homologues).
-    In this category, the covariant model also makes far fewer errors than the naive model.
-    See main text and caption to \figref{simulation} for details of simulation and cross-validation procedures.
-    \RFAM\ families with verified structures tended to be less phylogenetically diverse,
-    with roughly half the total branch length (on average), than the tree used for our simulations;
-    this may explain the relative paucity of low posterior-probability errors in the right-hand plot, compared to the left.
-% perl -e 'use Newick;use Stockholm::Database;$db=Stockholm::Database->from_file(shift);for$a(@$db){next if$a->columns>500 || @{$a->seqname}>400;$tree=Newick->from_string(@{$a->gf_NH});$b=0;for$l(@{$tree->branch_length}){$b+=$l}print"$b\n"}' ~/rfam/Training/Published.stock | meansd.pl
-% perl -e 'use Newick;use Stockholm::Database;$db=Stockholm::Database->from_file(shift);for$a(@$db){$tree=Newick->from_string(@{$a->gf_NH});$b=0;for$l(@{$tree->branch_length}){$b+=$l}print"$b\n"}' ~/rfam/Training/Hammerhead_1.stock
-  }
-\end{figure}
-
-We also performed cross-validation (hold-one-out) experiments. Starting from \RFAM\ alignments whose secondary structure is labeled ``Published'' (i.e. experimentally validated),
-we estimated phylogenetic trees using the neighbor-joining algorithm
-with the Jukes-Cantor model,
-then removed individual sequences
- and attempted to reconstruct them using \xrate.
-(This is possible because the substitution models are time-reversible, so we can treat any node as ancestral.)
-In this experiment, the error rates for base-pair reconstructions again differed by a few percent, though both error rates were higher than in the simulation test
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | wc -l` / `wc -l $f`,"\n"}' Published.xval-bptest
-(covariant model: 11\%, non-covariant model: 16\%).
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Published.xval-bptest
-Of the incorrectly-reconstructed base-pairs, the non-covariant model again predicted more non-canonical pairs (40\%) than the covariant model (17\%),
-especially at lower confidence levels (\figref{histograms}, right).
-
-Posterior probability estimates for both models were
-systematically higher in the cross-validation test than they were in the simulation (\figref{simulation}, right).
-The presence of this trend in both models
-suggests that it may be due to differences between the European rRNA alignments
-(from which the rates were estimated)
-and the \RFAM\ alignments (which were used in the cross-validation test).
-This is confirmed by our empirical observations of
-covariant rate matrices estimated from the two alignment datasets:
-base-pairs in the \RFAM\ alignments are more frequently
-non-canonical, and appear to evolve faster (relative to single-stranded regions),
-than base-pairs in the rRNA alignments
-(for details, see {\tt biowiki.org/RnaModels}).
-Repeating the cross-validation experiment with rates estimated from the \RFAM\ alignments,
-we find the covariant model predicted slightly fewer non-canonical base-pairs
-% perl -e '$f=shift;for$m(qw(g n)){print "$m: ",100 * `grep "$m= 0" $f | grep "${m}can= 0" | wc -l` / `grep "$m= 0" $f | wc -l`,"\n"}' Published.pubxval-bptest
-(14\% rather than 17\%)
-and produced slightly more accurate posterior probability estimates (data not shown).
-Although this is no longer a strict cross-validation (since parameters were estimated from the test dataset),
-it nevertheless emphasizes the value of correct models.
-%Another possible source of errors in the cross-validation (compared to the simulation)
-%is inaccurately estimated phylogenetic topologies and branch lengths;
-%this might be improvable by using a covariant model in the phylogenetic estimation step (rather than Jukes-Cantor),
-%and/or by using Bayesian phylogenetic estimation tools (rather than neighbor-joining).
-
-Accurate estimates of posterior probabilities are likely to be important:
-sub-optimal predictions are a significant piece of the reconstruction puzzle \cite{WilliamsEtAl2006,PollockChang2007,ChangEtAl2007}.
-Ancestral reconstruction by probabilistic modeling does not yield a single definitive sequence, but rather a probability distribution over sequences.
-Combinatorial synthesis by degenerate primer assembly has been advocated as a means of sampling from this distribution \cite{Gaucher2007}.
-Nucleotide-perfect reconstructions may not always be possible:
-it may be more productive to aim for accurate confidence estimates
-(including sub-optimal predictions)
-than to focus exclusively on getting everything right the first time.
-
-Based on these tests, we conclude that deep phylogenetic reconstructions of ribosomes, and other RNAs, will require covariant substitution models that take account of RNA secondary structure.
-We may reasonably deduce that indel reconstructions will, similarly, need to take account of RNA structure.
-These results, therefore, strongly motivate the development of multi-sequence models for RNA structure, such as the TKF Structure Tree.
-
-\section{Terminological asides}
-
-This subsection contains some asides about terminology,
-and can be omitted on a first reading of this paper.
-The intent of this subsection is to highlight connections to related fields
-in molecular evolution and computer science.
-
-\subsection{Phylogenetic grammars}
-
-We here describe some relationships to grammar models in current use in molecular evolution.
-
-% Removing this sentence since I do not think that describing the relationship between transducers and SCFGs as an "injective mapping" will make it any clearer for most readers -- it is a bit like describing a string as a "concatenative monoid" -- IH
-%There exists an injective mapping from the evolutionary models generated by our model-construction algorithm to multi-sequence SCFGs. In other words, 
-Given a singlet transducer, a phylogenetic tree, and a mapping from tree to branch transducers,
-the transducer composition algorithm returns a corresponding multi-sequence SCFG
-which models the joint probability distribution $P(X_1 , ... , X_{n+m})$.
-
-Several related bioinformatics grammars can be represented within this multi-SCFG framework,
-allowing for substantial re-use of prior modeling theory:
-\begin{itemize}
-\item String transducers (which are similar to HMMs) can be viewed as special cases of parse-tree transducers,
-without the facility to model long-distance correlations (base-pairs) or bifurcated structures.
-\item Phylo-HMMs and phylo-SCFGs \cite{KlostermanEtAl2006}
-can be viewed as special cases of (respectively) string transducers and parse-tree transducers.
-(Phylo-HMMs and phylo-SCFGs typically model substitutions but not indels,
-and so can be viewed as compositions of a richly-featured singleton transducer with a restricted class of branch transducers.)
-\item Sakakibara \cite{Sakakibara2003} has described ``Pair HMMs on tree structures'' that are closely related to our parse-tree transducers
-(though without the composition algorithms for scoring multiple alignments).
-\item Sakakibara has also described pair stochastic tree-adjoining grammars, or TAGs \cite{Sakakibara2004}.
-TAGs are a formal generalization of SCFGs to
-mildly context-sensitive grammars \cite{MamitsukaAbe94,JoshiSchabes97,RivasEddy2000}.
-We speculate that the conditionally-normalized transducer analogues of such Pair-TAGs may encompass the
-phylogenetic model of covariant RNA substitutions and indels on pseudoknotted structures
-described by \cite{MeyerMiklos2007},
-just as parse-tree transducers encompass phylo-SCFGs.
-\end{itemize}
-
-\subsection{Statistical Alignment}
-
-The term {\bf Statistical Alignment} was introduced by Hein \cite{HeinEtal2000}
-to describe DP algorithms for phylogenetic grammars.
-Although this term has been adopted by others in molecular evolution \cite{Miklos2002,Metzler2003,DrummondRambaut2007,SatijaEtAl2008},
-we've avoided it (except in this section) for the following four reasons:
-(i) Several RNA alignment methods already use another probability distribution
-that's completely different to (though equally as valid as) the phylogenetic probability distribution over alignments:
-namely, the Boltzmann probability distribution over secondary structures
-\cite{GorodkinEtAl97,MathewsTurner02,HofackerEtAl2004}.
-(ii) Many alignment tools make use of prior and posterior distributions over pairwise alignments
-\cite{Miyazawa94,BucherHofmann96,Holmes98,BrayPachter2004,DoEtAl2004,Holmes2005,LoytynojaGoldman2005,SchwartzPachter2007,BradleyPachterHolmes2008};
-all of these are arguably ``statistical'' alignment tools as well.
-(iii) Some of the above-cited methods use sum-over-pairs alignment metrics, taking the expectation of such metrics over pairwise alignment posteriors,
-and finding the multiple alignment that maximizes this expectation \cite{DoEtAl2004,SchwartzPachter2007,BradleyPachterHolmes2008};
-others do progressive multiple alignment using pair grammars \cite{BrayPachter2004,Holmes2005,LoytynojaGoldman2005}.
-Speculatively, such pair-based heuristics may (in some sense) approximate probabilistic alignment inference with true phylogenetic likelihoods;
-similar convergences have, at least, been demonstrated between other pair-based heuristics \cite{EickmeyerEtAl2008}.
-Thus, even using Hein's strict definition,
-it's possible that a broad range of methods systematically {\em approximate} ``Statistical Alignment'';
-this cautions against using the term too exclusively.
-(iv) Even more broadly, many alignment tools make use of statistics to estimate alignment ``significance''.
-In this frequentist approach, probability distributions for alignment scores (particularly Gumbel distributions)
-are estimated for random sequences.
-The tail integrals of these distributions then yield P-values for high-scoring database search results
-\cite{KarlinAltschul93,Mott2000,Eddy2008}.
-Thus, while the term ``Statistical Alignment'' has a quite specific technical meaning in molecular evolution
-(a meaning which certainly encompasses the algorithms we describe here),
-the broader bioinformatics literature may have competing claims to the term.
-In any case, the goal of this work is statistical {\em reconstruction} of RNA protosequences,
-rather than alignment {\em per se}.
-
-\subsection{Parse trees and stacks}
-
-Grammars are abstract models for languages;
-finite-state automata are the corresponding abstract models for those languages' parsers.
-Most of this paper uses the grammatical view of RNA structure evolution;
-in this short section we digress briefly to mention the automata-theoretic view of these models.
-
-The two non-local features that set SCFGs apart from HMMs are bifurcations and paired emissions \cite{Durbin98}.
-In RNA structure analysis, these features are used to model (respectively) branched-loops and base-pairs.
-
-A stochastic pushdown automaton can also model these features.
-Such an automaton is essentially an HMM with a LIFO stack (LIFO = ``Last In, First Out'').
-
-The pushdown automaton can parse two sequences simultaneously, aligning them and their parse trees.
-It handles bifurcations and paired emissions by pushing symbols onto its stack:
-a bifurcation by pushing a nonterminal, and a paired emission by pushing a terminal.
-The pushed terminals and nonterminals are popped off the stack when the machine reaches its $\Send$ state.
-
-The conditionally-normalized version of a pushdown automaton is a pushdown transducer:
-a machine that absorbs {\em and} parses an input sequence (or absorbs an already-parsed input sequence)
-and emits a (parsed) output sequence.
-This is the machine that we describe as a parse-tree transducer.
-
 \bibliography{../latex-inputs/alignment,../latex-inputs/reconstruction,../latex-inputs/duplication,../latex-inputs/genomics,../latex-inputs/ncrna,../latex-inputs/url}

### 2009-06-01 response-to-reviews2.txt
Revised supplement; Ian's changes to Rob's changes
Edits per changes to supplementary material.

@@
 basepair is likely to be perfectly aligned.) This sort of flexible
-fold and alignment constraint is, of course, implemented in CONSAN and
+fold and alignment constraint is implemented in both CONSAN and
 Stemloc. However, we note that the Pair SCFGs used in both those
@@
 
-We do hope that the sophistication of the constraint schemes used with
-these "tree transducer" models will increase over time. Eventually we
-hope to develop alignment constraints, simultaneous alignment-and-fold
-constraints, and other constraints (and/or sampling schemes) to enable
-efficient analysis of multiple phylogenetically-related sequences. The
-present work represents a first attempt at using these sorts of models
-for more than two sequences, but we hope it will not be the last
-attempt, and we acknowledge room for improvement.
+We hope & anticipate that the sophistication of the constraint schemes
+used with these sorts of model will steadily increase over
+time. Eventually we hope to develop alignment constraints,
+simultaneous alignment-and-fold constraints, and other constraints
+(and/or sampling schemes) to enable efficient analysis of multiple
+phylogenetically-related sequences. The present work represents a
+first attempt at using these sorts of models for more than two
+sequences, but we hope it will not be the last attempt, and we
+acknowledge room for improvement.
 
@@
 
-Thank you for pointing this out.  We have removed the reference to the
-supplementary material in this section and have attempted to make the
+Thank you for pointing this out.  We have attempted to make the
 main text a self-contained description of the composition algorithm,
-including its key corollaries (elimination of null states, and DP
+including its key corollaries (elimination of null states and DP
 inference).  The presentation of the composition algorithm in the
@@
 figures. These were due to mistakes made during manual conversion of
-indiegram's output parse trees into Stockholm-format alignments and a
+indiegram's output parse trees into structural alignments and a
 bug in the code used for maximum-likelihood reconstruction of
@@
 paper; the obvious cost is omitting some detail from the main paper.
-The PLoS Authors' Guidelines (which recommend using the supplement for
-algorithmic details) and the reviewer's preference for including more
-detail in the manuscript are potentially in conflict on the best
-approach to take here. We welcome advice on how to improve the flow
-generally.)
+We welcome advice on how to improve the flow generally.)
 

### 2009-06-01 paper.tex
Edits per changes to supplementary material.

@@
 
-A notational prescription for the allowed transitions may be found in \supptext.
+A notational prescription for the allowed transitions may be found in \supptext{1}.
 
@@
 
-\supptext\ includes a complete formal algorithm for exact null-cycle elimination in SCFGs,
+\supptext{2} includes a complete formal algorithm for exact null-cycle elimination in SCFGs,
 along with procedures for probabilistically restoring null cycles
@@
 the packages, allowing for flexible, automated model design.
+
+The available software is described in \supptext{3}.
 
@@
 \begin{algorithm}[!ht]
-  \SetLine
-  \SetKwFunction{rnd}{Random[0,1]}
-  \KwIn{Rule $\rho'' \in \grules''$}
-  \KwOut{Parse (sub)tree $T' \in \gtrees'$}
-
-  \Switch{$\rho''$}{
-    \Case{$X \to Y$}{
-      \eIf{\rnd $< P(X \to Y) / q_{XY}$}{
-        \KwRet{$(X \to Y)$} \;
-      }{
-        Let $z = \sum_W q_{XW} P'(W \to Y)$ \;
-        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to Y) / z$ \;
-        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to Y)$} \;
-      }
-    }
-    \Case{$X \to \epsilon$}{
-      \eIf{\rnd $< P'(X \to \epsilon) / P'(\epsilon | X)$}{
-        \KwRet{$(X \to \epsilon)$} \;
-      }{
-        Let $z = \sum_W q_{XW} P'(W \to \epsilon)$ \;
-        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to \epsilon) / z$ \;
-        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to \epsilon)$} \;
-      }
-    }
-    \Other{
-      \KwRet{$(\rho'')$} \;
-    }
-  }
-  \caption{\label{alg:restoreTransitions}
-    Subroutine $\mathrm{restoreTransitions}(\rho'')$ for the null cycle elimination procedure.
-  }
-\end{algorithm}
-
-\begin{algorithm}[!ht]
-  \SetLine
-  \SetKwFunction{rnd}{Random[0,1]}
-  \KwIn{Rule $\rho' \in \grules'$}
-  \KwOut{Parse (sub)tree $T \in \gtrees$}
-
-  \Switch{$\rho'$}{
-    \Case{$B^0 \to \epsilon$}{
-      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
-      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
-      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
-      \KwRet{$(B \to T_L\ T_R)$} \;
-    }
-    \Case{$B' \to L'$}{
-      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
-      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
-      \KwRet{$(B \to L\ T_R)$} \;
-    }
-    \Case{$B' \to R'$}{
-      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
-      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
-      \KwRet{$(B \to T_L\ R)$} \;
-    }
-    \Case{$B^2 \to L'\ R'$}{
-      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
-      \KwRet{$(B \to L\ R)$} \;
-    }
-    \lCase{$B^0 \to B'$}{
-      \KwRet{$()$} \;
-    }
-    \lCase{$B' \to B^2$}{
-      \KwRet{$()$} \;
-    }
-    \Other{
-      Let $\rho$ be the original rule in $\grules$ \;
-      \KwRet{$(\rho)$} \;
-    }
-  }
-
-  \caption{\label{alg:restoreBifurcations}
-    Subroutine $\mathrm{restoreBifurcations}(\rho')$ for the null cycle elimination procedure.
-  }
-\end{algorithm}
-
-\begin{algorithm}[!ht]
-  \SetLine
-  \SetKwFunction{rnd}{Random[0,1]}
-  \KwIn{Nonterminal $A \in \gnons$}
-  \KwOut{Parse tree $T \in \gtrees: \mathrm{seq}(T) = \epsilon, \mathrm{root}(T) = A$}
-
-  \eIf{$A$ is a bifurcation state, $A \to X\ Y$,}{
-    $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
-    $T_Y \leftarrow \mathrm{sampleNullSubtree}(Y)$ \;
-    \KwRet{$(A \rightarrow T_X\ T_Y)$} \;
-  }{
-    \eIf{\rnd $< P(A \to \epsilon) / P(\epsilon|A)$}{
-      \KwRet{$(A \to \epsilon)$} \;
-    }{
-      Let $z = \sum_Y P(A \to Y) P(\epsilon | Y)$ \;
-      Sample state $X$ from probability distribution $P(X) = P(A \to X) P(\epsilon | X) / z$ \;
-      $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
-      \KwRet{$(A \rightarrow T_X)$} \;
-    }
-  }
-
-  \caption{\label{alg:sampleNullSubtree}
-    Subroutine $\mathrm{sampleNullSubtree}(A)$ for the null cycle elimination procedure.
-  }
-\end{algorithm}
-
-\begin{algorithm}[!ht]
 

### 2009-06-01 paper.tex
Edits....

@@
 We describe a ``transducer composition'' algorithm for extending
-pairwise statistical models of RNA structural evolution to models of
+pairwise probabilistic models of RNA structural evolution to models of
 multiple sequences related by a phylogenetic tree.  This algorithm
@@
 
-A number of leading methods for bioinformatics analysis of noncoding RNA
- use probabilistic grammars as models for pairwise alignments.
-We show that any such pairwise grammar can be extended to an entire phylogeny,
- by considering the pairwise grammar as a machine (a ``transducer'') that locally transforms one RNA structure into another
- (thereby modeling a single parent-child relationship in the tree).
-In addition to current applications such as genefinding, homology detection, alignment and secondary structure prediction,
- this should permit statistical reconstruction of RNA genes ancestral to many present-day sequences.
-We describe algorithms, software implementations, and an elementary benchmark wherein
- the complexity and accuracy of three-way RNA reconstruction is measured as the length of one branch is varied.
-In the Discussion we consider how the three-way RNA alignment-reconstruction-folding algorithm, which is currently very expensive computationally,
- might be made more efficient so that larger phylogenies could be considered.
+A number of leading methods for bioinformatics analysis of structural
+RNAs use probabilistic grammars as models for pairs of homologous
+RNAs.  We show that any such pairwise grammar can be extended to an
+entire phylogeny by treating the pairwise grammar as a machine (a
+``transducer'') that models a single ancestor-descendant relationship
+in the tree, transforming one RNA structure into another.  In addition
+to current applications such as RNA genefinding, homology detection,
+alignment and secondary structure prediction, this should permit
+probabilistic reconstruction of RNA genes ancestral to many
+present-day sequences.  We describe algorithms, software
+implementations, and an elementary benchmark of the dependence of the
+accuracy of three-taxon RNA reconstruction on the outgroup branch
+length.  In the Discussion we consider how the three-taxon RNA
+alignment-reconstruction-folding algorithm, which is currently very
+computationally-expensive, might be made more efficient so that larger
+phylogenies could be considered.
 
@@
 i.e., from a pairwise model of two sequences to a {\bf multiple-sequence model} of many sequences.
-In a typical situation, the sequences at the leaves of the tree are observed, but those at internal nodes are unobserved.
+In a typical situation, the sequences at the leaves of the tree are
+observed but those at internal nodes are not.
 Questions of interest then include:
@@
 \item[B.] Can we sample (find the mode, take moments, etc.) from the posterior distribution of the unobserved sequence at the root node?
-\item[C.] Can we sample (etc.) from the posterior of the unobserved sequences at the other internal nodes?
+\item[C.] Can we sample from the posterior of the unobserved sequences at the other internal nodes?
 \item[D.] Can we estimate summaries of the evolutionary history, such as
@@
 (3) freely-available software implementing algorithms (1) and (2) for
-the simplified case of a three-branch star-topology tree.
+the simplified case of a three-taxon star-topology tree.
 While the idea of composing conditionally-normalized models on trees
@@
 Studies of related indel models have suggested that an implementation
-of dynamic programming (DP) algorithms on the three-branch tree topology is sufficient to
+of dynamic programming (DP) algorithms on a three-taxon tree is sufficient to
 draw samples from the posterior distribution of ancestral sequences on
@@
 time and memory required to reconstruct the structures of several RNA
-families for the case of a 3-taxon phylogeny, which we have
+families for the case of a three-taxon phylogeny, which we have
 implemented in the program \indiegram. While by these estimates only
@@
 %\vspace{.5\baselineskip}
-
-We assume here that the phylogeny, including both the tree topology and branch lengths, is given.
-This constraint somewhat arbitrarily confines the problem domain to the rather restricted case where we already have a phylogeny (and potentially a consensus structure),
-and we seek only an alignment plus ancestral reconstruction.
-It is important to note, however, that as far as the {\em model} is concerned, none of the constraints (phylogeny, structure or alignment) are final.
-For example, the probabilistic framework readily allows us to compare likelihoods of two different phylogenetic trees, by constructing a composite transducer for each tree.
-Thus, MCMC samplers for alignments could in principal be extended to phylogenies (albeit at a computational cost).
-We revisit such MCMC samplers in the Discussion.
+We assume here that the phylogeny, including both the tree topology
+and branch lengths, is given.
+% Ian, I moved the bits which you added here to the discussion; please
+% edit per your discretion.
 
@@
 dependence of reconstruction accuracy on outgroup branch length.
-(We here use the term ``outgroup'' loosely, to denote the variable-length branch in a three-branch study,
-where the other two branches are held at unit length.
-Technically, the term ``outgroup'' is only correct if the outgroup's branch is longer than the other two branches.)
+(We here use the term ``outgroup'' loosely to denote the
+variable-length branch in our three-taxon study,
+where the other two branches are held at unit length.)
 
 We simulated the evolution of RNAs under the TKFST model along
-3-taxon phylogenies (with one internal node), where we kept the branch lengths of two
+three-taxon phylogenies (with one internal node), where we kept the branch lengths of two
 sibling species constant and varied the branch length of the
@@
 
-An alternative MCMC scheme for sampling RNA phylogeny, structure and alignment was developed for the SimulFold program \cite{MeyerMiklos2007}.
-SimulFold does not use a strictly normalized probabilistic model, resulting in some oddities in the ways that structure and indels interact
-(for example, it does not penalize deletion of one half of a basepair).
-Currently, it is not clear how appropriate SimulFold would be for ancestral reconstruction,
-although it has several advantages (e.g. explicit treatment of pseudoknots).
-Of course, MCMC kernels are inherently adaptable to other purposes: the MCMC moves developed for SimulFold may be useful for inference under different models.
+An alternative MCMC scheme for sampling RNA phylogeny, structure and
+alignment was developed for the SimulFold program
+\cite{MeyerMiklos2007}.  SimulFold does not use a strictly normalized
+probabilistic model, resulting in some oddities in the ways that
+structure and indels interact (for example, it does not penalize
+deletion of one half of a basepair).  Currently, it is not clear how
+appropriate SimulFold would be for ancestral reconstruction, although
+it has several advantages (e.g. explicit treatment of pseudoknots).
+Of course, MCMC kernels are inherently adaptable to other purposes:
+the MCMC moves developed for SimulFold may be useful for inference
+under different models.
+
+This paper focuses on the case where the tree topology is known, but
+many of the methods which we describe below can be extended to the
+more general case where none of the possible constraints (phylogeny,
+structure or alignment) are final.  For example, the probabilistic
+framework readily allows us to compare likelihoods of two different
+phylogenetic trees by constructing a composite transducer for each
+tree.  Thus, the MCMC samplers described above for alignments could,
+in principal, be extended to phylogenies (albeit at a computational
+cost).
 

### 2009-06-01 paper.tex
Edits

@@
 Below we describe a method to eliminate null cycles from
-the ensemble model, by transforming any SCFG to an equivalent acyclic SCFG.
+the ensemble model by transforming any SCFG to an equivalent acyclic SCFG.
 We then present multiple-sequence versions of the CYK, Inside and Outside algorithms.
 
-As an aside: while some sort of null-cycle elimination is often required in order to deal with cyclic dependencies, there are several ways to accomplish this other than the algorithm presented here.
-A simpler approach (that only works for the CYK algorithm) appears in the computational linguistics literature \cite{GecsegSteinby97}.
-We have also developed a heuristic for CYK that simply ignores null cycles, as well as an iterative approximation that loops several times over cyclically-dependent cells of the DP matrix
-until the estimate starts to converge. For conciseness, we have omitted descriptions of these methods, presenting only the exact elimination algorithm.
+While some sort of null-cycle elimination is often required in order
+to deal with cyclic dependencies, there are several ways to accomplish
+this other than the algorithm presented below.  A simpler approach
+(that only works for the CYK algorithm) appears in the computational
+linguistics literature \cite{GecsegSteinby97}.  We have also developed
+a heuristic for CYK that simply ignores null cycles as well as an
+iterative approximation that loops several times over
+cyclically-dependent cells of the DP matrix until the estimate starts
+to converge. For conciseness, we have omitted descriptions of these
+methods, presenting only the exact elimination algorithm.
 
@@
 structure, makes clear the necessity for automated procedures for
-model construction.  Dataset S1 gives the state space of the composed
+model construction.  Dataset S1 gives the state space of the ensemble
 model constructed by the search algorithm and Dataset S2 the reduced

### 2009-06-02 paper.tex
More edits

@@
 Let the total order on the nodes correspond to any preorder traversal of
-the tree; thus, ``$m$ is ancestral to $n$'' is sufficient-but-not-necessary for ``$m \prec n$''.
+the tree; thus, ``$m$ is ancestral to $n$'' is sufficient-but-not-necessary for ``$m \prec n$.''
 Let ${\cal T}_m$ denote the singlet or branch transducer which emits symbols to node $m$.
@@
 the packages, allowing for flexible, automated model design.
-
 The available software is described in \supptext{3}.
@@
 We illustrated our method for building models of structured sequences
-using a simplified model which was introduced in previous work,
+using a model which was introduced in previous work,
 the TKF Structure Tree \cite{Holmes2004}, a simplified probabilistic
@@
 using estimates reported by a phylo-grammar for RNA secondary
-structure prediction, \pfold\ \cite{KnudsenHein2003} and an
+structure prediction, \pfold\ \cite{KnudsenHein2003}, and an
 implementation of pairwise alignment for the TKF Structure Tree model,
@@
 should be sufficient for ergodic sampling of reconstructions on any larger phylogeny,
-using e.g. a Gibbs-sampling MCMC kernel \cite{JensenHein2002}
+using, e.g., a Gibbs-sampling MCMC kernel \cite{JensenHein2002}
 or a progressive suboptimal-alignment sampling heuristic \cite{PatenHolmesBirney2008}.
@@
 3) \Rightarrow 441$ states and $O(a \cdot b^2 \cdot N) = O(7 \cdot
-21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in \secname{The composition algorithm}.
+21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in \secname{The composition algorithm.}
  Nonetheless, the extreme complexity of the
@@
 time complexity of the algorithms could be dramatically reduced.  Such
-constraints can be naturally expressed with {\em alignment envelopes},'
+constraints can be naturally expressed with ``alignment envelopes,''
 the alignment-space analogue of fold envelopes \cite{Holmes2005}.
@@
 
-\newpage
 \section*{Discussion}
@@
 appropriate SimulFold would be for ancestral reconstruction, although
-it has several advantages (e.g. explicit treatment of pseudoknots).
+it has several advantages (e.g., explicit treatment of pseudoknots).
 Of course, MCMC kernels are inherently adaptable to other purposes:
@@
 This paper focuses on the case where the tree topology is known, but
-many of the methods which we describe below can be extended to the
+many of the methods which we have described can be extended to the
 more general case where none of the possible constraints (phylogeny,

### 2009-06-02 simulations/run-handalign.pl
Added code to get TKF91 Viterbi alignment for evolsayer simulations

@@ -1,0 +1,79 @@
+#!/usr/bin/perl -w
+
+# imports
+use Stockholm;
+use FindBin qw($Bin);
+
+# read command-line args
+my ($true_alignment) = @ARGV;
+
+# create temp filenames
+my $temp_alignment = "$true_alignment.handalign-input";
+
+# define constants
+my $tiny = .00001;  # default length for dummy branch
+my $handalign = "handalign";  # path to handalign binary
+my $loopmodel = "$Bin/loopmodel.eg";  # substitution model for handalign (should be equivalent to loop-submodel of pfold model used by evolsayer.pl to generate alignments)
+
+# load alignment
+my $true = Stockholm->from_file ($true_alignment);
+warn "True alignment\n";
+print $true->to_string;
+
+# read stuff from alignment
+my %seq = %{$true->seqdata};
+my $treestr = join "", @{$true->gf_NH};
+my $args = join "", @{$true->gf_PARAM};
+
+# extract branch lengths & sequences
+my (%branchlen, %seqlen);
+for my $branch (qw(X Y Z)) {
+    if ($treestr =~ /$branch:([0-9\.eE\+\-]+)/) {
+	$branchlen{$branch} = $1;
+    } else {
+	die "Couldn't find $branch in $treestr";
+    }
+    die "Couldn't find row $branch in alignment" unless exists $seq{$branch};
+    $seq{$branch} =~ s/[\.\-_]//g;
+    $seqlen{$branch} = length $seq{$branch};
+}
+
+# create rows of new alignment
+my %newrow;
+$newrow{'X'} = $seq{'X'} . ("-" x ($seqlen{'Y'} + $seqlen{'Z'}));
+$newrow{'Y'} = ("-" x $seqlen{'X'}) . $seq{'Y'} . ("-" x $seqlen{'Z'});
+$newrow{'Z'} = ("-" x ($seqlen{'X'} + $seqlen{'Y'})) . $seq{'Z'};
+
+$newrow{'subroot'} = $newrow{'Z'};
+$newrow{'subroot'} =~ s/[^\.\-_]/*/g;
+
+$newrow{'root'} = $branchlen{'X'} < $branchlen{'Y'} ? $newrow{'X'} : $newrow{'Y'};
+$newrow{'root'} =~ s/[^\.\-_]/*/g;
+
+# create new alignment
+my $new = Stockholm->new;
+
+$new->seqname ([qw(root subroot X Y Z)]);
+$new->seqdata (\%newrow);
[…]

### 2009-06-02 simulations/compare.pl
Added rules for comparing TKF & indiegram alignments

@@ -5,9 +5,12 @@ use Stockholm;
 
 my $ref = Stockholm->from_file (shift);
 my $cmp = Stockholm->from_file (shift);
+my $tkf = Stockholm->from_file (shift);
 
-@{$ref->seqname} = ("X", "Y", "Z");
-@{$cmp->seqname} = ("X", "Y", "Z");
+# indiegram adds cruft to the sequence names and prints non-Stockholm stuff that gets read back incorrectly by Stockholm.pm
+# handalign and indiegram both add "root" and "subroot" lines
+# so we need to tidy up the sequence name fields...
+@{$ref->seqname} = @{$cmp->seqname} = @{$tkf->seqname} = ("X", "Y", "Z");
 
 $ref->gc->{"ancestral_SS"} = $ref->gr->{"SS"}->{"root"};
 delete $ref->gr->{"SS"}->{"root"};
@@ -23,18 +26,10 @@ while (my ($name, $data) = each %{$cmp->seqdata}) {
 
 $ref->to_file ("/tmp/ref.stock");
 $cmp->to_file ("/tmp/cmp.stock");
+$tkf->to_file ("/tmp/tkf.stock");
 
-my $cmd = "cmpalign.pl /tmp/ref.stock /tmp/cmp.stock >/tmp/cmpalign.out";
-system ($cmd) == 0 or die "Couldn't run command: $cmd\n";
-
-open CMPALIGN, "</tmp/cmpalign.out" or die "Couldn't open '/tmp/cmpalign.out'\n";
-my ($acc, $sn, $ppv);
-while (<CMPALIGN>) {
-  if (/^Acc\s+(\S+)$/) { $acc = $1; }
-  if (/^Sn\s+(\S+)$/) { $sn = $1; }
-  if (/^PPV\s+(\S+)$/) { $ppv = $1; }  
-}
-close CMPALIGN;
+my ($acc, $sn, $ppv) = cmpalign ("/tmp/ref.stock", "/tmp/cmp.stock");
+my ($acc_tkf, $sn_tkf, $ppv_tkf) = cmpalign ("/tmp/ref.stock", "/tmp/tkf.stock");
 
 my $refanc = $ref->gc->{"ancestral_SS"};
 my $refseq = $ref->seqdata->{"root"};
@@ -82,7 +77,7 @@ $refancstock->to_file ("/tmp/refanc.stock");
[…]

### 2009-06-02 simulations/run-handalign.pl

@@ -18,7 +18,7 @@ my $loopmodel = "$Bin/loopmodel.eg";  # substitution model for handalign (should
 # load alignment
 my $true = Stockholm->from_file ($true_alignment);
 warn "True alignment\n";
-print $true->to_string;
+warn $true->to_string;
 
 # read stuff from alignment
 my %seq = %{$true->seqdata};
@@ -74,6 +74,8 @@ print TEMP $new->to_string;
 close TEMP or die "Couldn't write $temp_alignment: $!";
 
 # run handalign
-my $command = "$handalign -tkf91 -l $loop_len -d $mu -m $loopmodel -s 0 -r -ub $temp_alignment -ha -hca '-lstdc++'";
+my $command = "$handalign -tkf91 -l $loop_len -d $mu -m $loopmodel -s 0 -r -ub $temp_alignment -ha -hca '-lstdc++' -hc /tmp/indiegram-benchmark-cache";
 warn "Running $command\n";
-system $command;
+my @output = `$command`;
+#print grep (!/root/, @output);
+print @output;

### 2009-06-02 simulations/compare.pl
Added Makefile rules for comparing TKF & indiegram alignments

@@ -93,19 +93,19 @@ print join ("\t", ($acc, $sn, $ppv, $ancoverlap, $acc_tkf, $sn_tkf, $ppv_tkf)), 
 
 sub cmpalign {
     my ($ref, $cmp) = @_;
-    my $cmd = "cmpalign $ref $cmp >/tmp/cmpalign.out";
+    my $cmd = "cmpalign -s $ref $cmp >/tmp/cmpalign.out";
     warn "Running $cmd\n";
     system ($cmd) == 0 or die "Couldn't run command: $cmd\n";
 
     local *CMPALIGN;
     open CMPALIGN, "</tmp/cmpalign.out" or die "Couldn't open '/tmp/cmpalign.out'\n";
-    my ($acc, $sn, $ppv);
+    my ($acc, $sn, $ppv, $tcs);
     while (<CMPALIGN>) {
-	if (/^Acc\s+(\S+)$/) { $acc = $1; }
-	if (/^Sn\s+(\S+)$/) { $sn = $1; }
-	if (/^PPV\s+(\S+)$/) { $ppv = $1; }  
+	if (/\S/) {
+	    ($acc, $sn, $ppv, $tcs) = split;
+	}
     }
     close CMPALIGN;
 
-    return ($acc, $sn, $ppv);
+    return ($acc, $sn, $ppv, $tcs);
 }

### 2009-06-02 simulations/run-handalign.pl
turned off hmmoc cache in handalign experiments

@@ -40,23 +40,23 @@ for my $branch (qw(X Y Z)) {
 
 # create rows of new alignment
 my %newrow;
-$newrow{'X'} = $seq{'X'} . ("-" x ($seqlen{'Y'} + $seqlen{'Z'}));
-$newrow{'Y'} = ("-" x $seqlen{'X'}) . $seq{'Y'} . ("-" x $seqlen{'Z'});
-$newrow{'Z'} = ("-" x ($seqlen{'X'} + $seqlen{'Y'})) . $seq{'Z'};
+$newrow{'Z'} = $seq{'Z'} . ("-" x ($seqlen{'Y'} + $seqlen{'X'}));
+$newrow{'Y'} = ("-" x $seqlen{'Z'}) . $seq{'Y'} . ("-" x $seqlen{'X'});
+$newrow{'X'} = ("-" x ($seqlen{'Z'} + $seqlen{'Y'})) . $seq{'X'};
 
-$newrow{'subroot'} = $newrow{'Z'};
-$newrow{'subroot'} =~ s/[^\.\-_]/*/g;
+$newrow{'xyzroot'} = $newrow{'Z'};
+$newrow{'xyzroot'} =~ s/[^\.\-_]/*/g;
 
-$newrow{'root'} = $branchlen{'X'} < $branchlen{'Y'} ? $newrow{'X'} : $newrow{'Y'};
-$newrow{'root'} =~ s/[^\.\-_]/*/g;
+$newrow{'xyroot'} = $branchlen{'X'} < $branchlen{'Y'} ? $newrow{'X'} : $newrow{'Y'};
+$newrow{'xyroot'} =~ s/[^\.\-_]/*/g;
 
 # create new alignment
 my $new = Stockholm->new;
 
-$new->seqname ([qw(root subroot X Y Z)]);
+$new->seqname ([qw(xyzroot xyroot X Y Z)]);
 $new->seqdata (\%newrow);
 
-$new->add_gf ("NH", "((X:$branchlen{X},Y:$branchlen{Y})root:$branchlen{Z},Z:$tiny)subroot;");
+$new->add_gf ("NH", "((X:$branchlen{X},Y:$branchlen{Y})xyroot:$branchlen{Z},Z:$tiny)xyzroot;");
 
 # extract loop TKF parameters from alignment
 my ($lambda, $mu);
@@ -74,7 +74,8 @@ print TEMP $new->to_string;
 close TEMP or die "Couldn't write $temp_alignment: $!";
 
 # run handalign
[…]

### 2009-06-03 simulations/compare.pl
Added stemloc & stemlocAMA comparisons to Makefile (but not yet plots.r)

@@ -3,9 +3,13 @@
 use strict;
 use Stockholm;
 
+my $tmpdir = "/tmp";
+
 my $ref = Stockholm->from_file (shift);
 my $cmp = Stockholm->from_file (shift);
 my $tkf = Stockholm->from_file (shift);
+my $stemloc = Stockholm->from_file (shift);
+my $stemlocama = Stockholm->from_file (shift);
 
 # indiegram adds cruft to the sequence names and prints non-Stockholm stuff that gets read back incorrectly by Stockholm.pm
 # handalign and indiegram both add "root" and "subroot" lines
@@ -24,12 +28,12 @@ while (my ($name, $data) = each %{$cmp->seqdata}) {
   }
 }
 
-$ref->to_file ("/tmp/ref.stock");
-$cmp->to_file ("/tmp/cmp.stock");
-$tkf->to_file ("/tmp/tkf.stock");
+for my $file (qw(ref cmp tkf stemloc stemlocama)) {
+    eval('$'.$file)->to_file ("$tmpdir/$file.stock");
+}
 
-my ($acc, $sn, $ppv) = cmpalign ("/tmp/ref.stock", "/tmp/cmp.stock");
-my ($acc_tkf, $sn_tkf, $ppv_tkf) = cmpalign ("/tmp/ref.stock", "/tmp/tkf.stock");
+my ($acc, $sn, $ppv, $tcs) = cmpalign ("$tmpdir/ref.stock", "$tmpdir/cmp.stock");
+my @acc_sn_ppv_tcs = map ([cmpalign ("$tmpdir/ref.stock", "$tmpdir/$_.stock")], qw(tkf stemloc stemlocama));
 
 my $refanc = $ref->gc->{"ancestral_SS"};
 my $refseq = $ref->seqdata->{"root"};
@@ -73,14 +77,14 @@ $cmpancstock->seqdata->{"cmp"} = $cmpseq;
 $refancstock->gr->{"SS"}->{"ref"} = $refanc;
 $cmpancstock->gr->{"SS"}->{"cmp"} = $cmpanc;
 
-$refancstock->to_file ("/tmp/refanc.stock");
-$cmpancstock->to_file ("/tmp/cmpanc.stock");
+$refancstock->to_file ("$tmpdir/refanc.stock");
+$cmpancstock->to_file ("$tmpdir/cmpanc.stock");
 
[…]

### 2009-06-03 simulations/plots.r
Updated plots.r so simulations.pdf now includes TKF accuracy

@@ -6,9 +6,11 @@ require (stats)
 toPDF = TRUE
 
 alignColor = "blue"
+alignTKFColor = "cyan"
 structureColor = "red"
 
 alignShape = 16
+alignTKFShape = 16
 structureShape = 17
 
 ## default plot filename
@@ -33,12 +35,17 @@ data = read.table ("inferred/indiegram.analysis", header = TRUE, sep = "\t")
 
 ## average datapoints for each outgroup length
 align = array()
+alignTKF = array()
 structure = array()
 bins = unique (data$z)
 for (bin in bins) {
   align = rbind (
     align,
     c (bin, mean (subset (data$AlignAcc,
+                          data$z == bin))))
+  alignTKF = rbind (
+    alignTKF,
+    c (bin, mean (subset (data$AlignAccTKF,
                           data$z == bin))))
   structure = rbind (
     structure,
@@ -48,11 +55,12 @@ for (bin in bins) {
 
 ## remove NA rows
 align = align[-1,]
+alignTKF = alignTKF[-1,]
 structure = structure[-1,]
 
 plot.new()
 plot.window (xlim = c (0, 2.5),
-             ylim = c (0.4, 1))
+             ylim = c (0, 1))
 axis (1,
       c (0, 0.5, 1, 1.5, 2, 2.5))
 axis (2)
@@ -63,13 +71,18 @@ title (xlab = "Outgroup branch length",
 points (align,
         col = alignColor, pch = alignShape, cex = 1)
 lines (lowess (align), col = alignColor, lwd = 1.5)
+
+points (alignTKF,
+        col = alignTKFColor, pch = alignTKFShape, cex = 1)
+lines (lowess (alignTKF), col = alignTKFColor, lwd = 1.5)
+
 points (structure,
         col = structureColor, pch = structureShape, cex = 1)
 lines (lowess (structure), col = structureColor, lwd = 1.5)
 
-legend ("bottomleft", legend = c ("Structural reconstruction accuracy", "Alignment accuracy"),
-        col = c (structureColor, alignColor),
-        pch = c (structureShape, alignShape),
+legend ("bottomleft", legend = c ("Structural reconstruction accuracy (TKFST ML)", "Alignment accuracy (TKFST ML)", "Alignment accuracy (TKF91 ML)"),
+        col = c (structureColor, alignColor, alignTKFColor),
+        pch = c (structureShape, alignShape, alignTKFShape),
         xjust = 1,
         inset = 0.02)
 

### 2009-06-03 paper.tex
minor edit

@@
     with $\sim 0.3$ substructure indels per alignment.
-    We selected challenging alignments to reconstruct by requiring
+    We selected alignments to reconstruct by requiring
     that there be at least two ancestral stems, loops of length $\in [3,

### 2009-06-04 response-to-reviews2.txt
renamed software appendix (Text S3) to Text S4; added new Text S3 for TKFST grammars; added a bunch of other stuff

@@
 
-Thank you for pointing this out.  We have attempted to make the
-main text a self-contained description of the composition algorithm,
+Thank you for pointing this out.  We have attempted to make the main
+text a self-contained description of the composition algorithm,
 including its key corollaries (elimination of null states and DP
 inference).  The presentation of the composition algorithm in the
-supplement now merely reprises the presentation in the main text using
-more formal notation. We hope this is an appropriate use of the
-supplementary information, but we welcome additional suggestions to
-improve the flow of the manuscript for the reader.
+supplement (text S1) now merely reprises the presentation in the main
+text using more formal notation. Likewise, the null cycle elimination
+algorithm is discussed verbally in the main paper, and described again
+using formal notation in text S2. Texts S3 and S4 describe aspects of
+the software that are also probably too detailed for the main paper.
+
+We hope this is an appropriate use of the supplementary information,
+but we welcome additional suggestions to improve the flow of the
+manuscript for the reader.
 
@@
 given in the revised text, with a full formal description (including
-notation) in the supplement. (Since this is a standalone result that
+notation) in supplement S2. (Since this is a standalone result that
 can be applied to any SCFG, we think it fits better as a technical
@@
 the possible presence of null bifurcations complicates matters for
-SCFGs.  Indeed, no *closed form* elimination is possible (due to
-bifurcations, such an elimination would involve solving simultaneous
-nonlinear equations, as explained in the supplement) but an iterative
-approximation (e.g. Newton-Raphson) appears adequate.
+SCFGs.  Indeed, no *closed form* elimination is possible (because of
+bifurcations: such an elimination would involve solving simultaneous
+nonlinear equations, as shown in supplement S2) but an iterative
+approximation (e.g. Newton-Raphson) appears to be adequate.
 
@@
 > I_S; these nonterminals have identical production rules.
+
+This is a case of shoe-horning the grammar into the input format
+accepted by indiegram (and by our composition algorithm).
 

### 2009-06-04 paper.tex
renamed software appendix (Text S3) to Text S4; added new Text S3 for TKFST grammars; added a bunch of other stuff

@@
 
-\supptext{2} includes a complete formal algorithm for exact null-cycle elimination in SCFGs,
+\supptext{2}\ includes a complete formal algorithm for exact null-cycle elimination in SCFGs,
 along with procedures for probabilistically restoring null cycles
@@
 original TKF model, sequence evolves under a time-homogeneous linear
-birth-death-immigration process \cite{Kendall1948}.  Single characters
+birth-death-immigration process \cite{Kendall1948}.  Single characters (``links'')
 are inserted with rate $\lambda$ and deleted with rate $\mu$.  At
@@
 structure.  In this model, loop and stem regions are
-mutually nested (\figref{tkfst}).  Both loops and stems have geometric length
-distributions with parameters $\kappa_l$ and $\kappa_s$.  Single bases
-are inserted and deleted in loops with rates $\lambda_l$ and $\mu_l$;
+mutually nested (\figref{tkfst}):
+the parameter $\pi_l(S)$ determines the proportion of links within loop sequences that are nested stems,
+and every stem sequence has a nested loop at the end.
+Single bases are inserted and deleted in loops with rates $\lambda_l$ and $\mu_l$;
 similarly, base-pairs are inserted and deleted in stems with rates
-$\lambda_s$ and $\mu_s$.  Insertions of a new stem into an existing
+$\lambda_s$ and $\mu_s$.
+Both loops and stems have geometric length
+distributions with parameters $\kappa_l = \lambda_l / \mu_l$ and $\kappa_s = \lambda_s / \mu_s$.  
+Insertions of a new stem into an existing
 loop sequence (or deletions of an existing stem) occur at the same
@@
 and similarly for stem sequences \cite{Holmes2004}.
+
+A few other useful statistics for the TKFST model:
+the expected number of links in a loop sequence is $K_l = \frac{\lambda_l}{\mu_l - \lambda_l}$
+and in a stem sequence $K_s = \frac{\lambda_s}{\mu_s - \lambda_s}$.
+Since $\pi_l(S)$ of the links in a loop sequence are nested stems,
+and since each stem has twice as many nucleotides as it has links (since each link is a base {\em pair}),
+the expected number of bases in a loop sequence is
+\[
+B_l = \frac{K_l (1-\pi_l(S)) + 2K_l \pi_l(S)} {1 - \pi_l(S) K_l}
+\]
+The expected number of bases in a stem sequence is
+\[
+B_s = 2K_s + B_l
+\]
+The expected number of bases that are created/removed when a loop-sequence link is inserted/deleted is
+\[
+D_l = (1-\pi_l(S)) + \pi_l(S) B_s
+\]
+The expected number of stems directly rooted in a given loop sequence is
+$U = \pi_l(S) K_l$
+and the expected number of stems directly rooted in, {\bf or indirectly descended from}, a given loop sequence is
+$V = U / (1 - U)$
+(note that this is also the expected total number of loop sequences indirectly descended from a given loop sequence).
+Therefore, in the equilibrium structure,
+the expected number of stems is $V$;
+of loops, $V+1$;
+of unpaired bases, $K_l (1 - \pi_l(S)) (V+1)$;
+and of base-pairs, $K_s V$.
+In a tree with total branch length $T$,
+the expected number of single-base deletions is $\mu_l T K_l (1 - \pi_l(S)) (V+1)$;
+of base-pair deletions, $\mu_s T K_s V$;
+and of substructure deletions, $\mu_l T V$.
 
@@
 outgroup, we conducted a simulation study to investigate the
-dependence of reconstruction accuracy on outgroup branch length.
+dependence of reconstruction accuracy on outgroup branch length,
+and to investigate the performance of our reconstruction method (when simulating directly from the model)
+compared to simpler reconstruction methods that ignore either structure or phylogeny.
 (We here use the term ``outgroup'' loosely to denote the
@@
 ancestral structures.
-The effect of imperfect knowledge of the ancestral structures has not been explored.
+%The effect of imperfect knowledge of the ancestral structures has not been explored.
+% [commented out because the following tests DO address this in a roundabout way; what we mean is that indiegram has not been tested when leaf structures are unknown]
+
+To compare \indiegram\ to simpler reconstruction methods, we also measured the alignment accuracy using various simpler statistical reconstruction techniques.
+We evaluated three maximum likelihood inference algorithms that use simpler gap-aware phylogenetic models,
+and one posterior decoding algorithm that uses a non-phylogenetic sum-of-pairs scoring scheme.
+These results are plotted on \figref{simulations} alongside the \indiegram\ accuracies.
+In detail, the reconstruction methods used were
+
+\begin{description}
+\item{TKF91 ML:} with the TKF91 model \cite{}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
+The correct phylogenetic tree and branch lengths were supplied (as they were for the \indiegram\ benchmark).
+The insertion, deletion and substitution rates for the TKF91 model were set equal to those of the loop submodel of TKFST.
+\item{Long Indel ML:} with a single-event trajectory approximation to the long indel model \cite{MiklosLunterHolmes2004}, the \handel\ package was used to align the three extant sequences and reconstruct the ancestor.
+The correct phylogenetic tree and branch lengths were supplied.
+The deletion and substitution rates were set equal to those of the loop submodel of TKFST.
+The mean indel length was set equal to $D_l$, the mean number of bases that are created/removed by an insertion/deletion in the loop submodel of TKFST;
+the mean equilibrium sequence length (and thereby the insertion rate) was equal to $B_l$, the mean number of bases in TKFST at equilibrium
+(\secname{A simple model of RNA structural evolution} has formulae for these quantities in terms of the TKFST rate parameters).
+\item{Stemloc ML:} the \stemloc\ program was used to align the three sequences via single-linkage clustering with a Pair SCFG \cite{Holmes2005}.
+The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
+\item{Stemloc AMA:} the \stemloc\ program was used to align the three sequences in AMA mode, using posterior decoding to optimize a sum-over-pairs expected alignment accuracy \cite{BradleyPachterHolmes2008}.
+The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
+\end{description}
+
+We did not measure the accuracy of the reconstructed ancestral structure in these tests (as we did in the \indiegram\ test),
+since \handel\ does not consider structures, whereas \stemloc\ and \stemlocama\ do not provide ancestral reconstructions.
+
+{\bf SUMMARY OF THESE RESULTS}
+% TKF is less accurate than indiegram, although for short branch lengths it is almost identical; the TKF and indiegram accuracies diverge at longer branch lengths
+% Long indel?
+% Stemloc?
+% StemlocAMA?
 

### 2009-06-04 s3.tex
renamed software appendix (Text S3) to Text S4; added new Text S3 for TKFST grammars; added a bunch of other stuff

@@
 \newpage
-\section{Software}
+\section{Inference grammars for the TKFST model}
+\seclabel{TKFST}
 
-All tools are available from \darturl\ as part of the \dart\ software package for sequence analysis.
+This section describes multi-sequence grammars that can be used to infer alignments, structures and ancestors under the
+TKF Structure Tree (TKFST) model.
 
-\subsection{Automated grammar construction}
-We implemented our model construction algorithm on the three-taxon star phylogeny
-in a set of Perl scripts.
-Given a singlet transducer modeling ancestral sequences and structures and
-a branch transducer modeling structural evolution,
-the scripts generate C++ code to create the corresponding jointly-normalized SCFG.
-All possible models of structural evolution which can be represented by a Pair SCFG
-are permitted as input to the scripts,
-allowing for flexible and automated model design.
+\input{tkfst.tex}
 
-Given files describing the singlet and branch transducers,
-including weights of all transitions which may be functions of evolutionary distance,
-the package ComposedTreeTransducer::FourWayComposedTT can automatically generate
-the state graph and transition matrix of a multi-sequence model of three extant sequences.
-It removes the useless windback $\Snull$ states described in the paper and
-introduces effective direct transitions caused by bifurcations with possibly-empty children.
-The package ComposedTreeTransducer::TripletSCFG transforms a multi-sequence
-model created by ComposedTreeTransducer::FourWayComposedTT into the 
-corresponding jointly-normalized three-sequence SCFG
-and generates C++ code to build the model.
+\newpage
+\section{The TKF Structure Tree model as a transducer} \seclabel{tkfst}
 
-Example singlet and branch transducers files are provided for a simple 
-Pair HMM model, a simple Pair SCFG model and the full TKF Structure Tree model.
+We can cast the TKF Structure Tree (TKFST) model as a transducer.
+This means we can automatically deduce rules like those shown for the TKFST model in \secref{TKFST}.
 
-\subsection{Reconstruction of ancestral structures}
-The program \indiegram\ can perform maximum-likelihood inference on the
-three-sequence SCFGs automatically generated by the ComposedTreeTransducer::TripletSCFG package.
-Complete or no structural information for the three extant sequences can be supplied
-as input.
+\subsection{The single-sequence TKFST model as a singlet transducer}
 
-\subsection{Simulation of RNA family evolution}
-The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
-Another script (\animateevolsayer) can be used to make animations of the evolving RNA structures, in combination with the RNAplot program in the \viennarna\ package \cite{HofackerEtAl94}.
+The states of the singlet transducer are shown in \tabref{tkfstsinglet}.
+Allowed transitions between these states are shown in \secref{TKFST}.
 
-\bibliography{../latex-inputs/alignment,../latex-inputs/reconstruction,../latex-inputs/duplication,../latex-inputs/genomics,../latex-inputs/ncrna,../latex-inputs/url}
+\begin{table}[!ht]
+  \centering
+  \begin{tabular}{lrrrr}
+    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
+    $L$ & $\Sstart$ \\
+    $I_L$ & $\Sinsert$ & & $(x,\Tnull)$ & $p(x)$ \\
+    \\
+    $S$ & $\Sstart$ \\
+    $I_S$ & $\Sinsert$ & & $(x,y)$ & $p(x,y)$ \\
+    \\
+    $B$ & $\Sinsert$ & & $(L\,S)$ & 1 \\
+  \end{tabular}
+  \caption{
+    \tablabel{tkfstsinglet}
+    States of the singlet transducer of the TKF Structure Tree model \cite{Holmes2004}.
+    Singlet transducers can only have states of type $\Sstart$ or $\Sinsert$.
+    This is the indiegram-style transducer equivalent of the singlet TKFST SCFG in \secref{TKFST}.
+  }
+\end{table}
 
-\end{document}
+\subsection{The two-sequence TKFST model as a branch transducer}
+
+The states of the branch transducer are shown in \tabref{tkfstbranch}.
+Allowed transitions between these states are shown in \secref{TKFST}.
+
+\begin{table}[!ht]
+  \centering
+  \begin{tabular}{lrrrr}
+    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
+    $L$ & $\Sstart$ \\
+    $I_L$ & $\Sinsert$ & & $(u,\Tnull)$ & $p(u)$ \\
+    $M_L$ & $\Smatch$ & $(x,\Tnull)$ & $(u,\Tnull)$ & $p(u|x)$ \\
+    $D_L$ & $\Smatch$ & $(x,\Tnull)$ & $(\Tgap,\Tnull)$ & 1 \\
+    $W_L$ & $\Swait$ \\
+    \\
+    $S$ & $\Sstart$ \\
+    $I_S$ & $\Sinsert$ & & $(u,v)$ & $p(u,v)$ \\
+    $M_S$ & $\Smatch$ & $(x,y)$ & $(u,v)$ & $p(u,v|x,y)$ \\
+    $D_S$ & $\Smatch$ & $(x,y)$ & $(\Tgap,\Tgap)$ & 1 \\
+    $W_S$ & $\Swait$ \\
+    \\
+    $B_i$ & $\Sinsert$ & & $(L_i\,S_i)$ & 1 \\
+    $B$ & $\Smatch$ & $(L\,S)$ & $(L\,S)$ & 1 \\
+    $B_p$ & $\Smatch$ & $(L\,S)$ & $(L\,\Send)$ & 1 \\
+    $B_e$ & $\Smatch$ & $(L\,\Send)$ & $(L\,\Send)$ & 1 \\
+  \end{tabular}
+  \caption{\tablabel{tkfstbranch}
+    States of the branch transducer of the TKF Structure Tree model \cite{Holmes2004}.
+    States which have the same names as states of the singlet transducer in \tabref{tkfstsinglet}
+    are the branch equivalent of the corresponding singlet states
+    (e.g. a $\Smatch$ state might be the branch equivalent of an $\Sinsert$ state).
+    States $L_i$ and $S_i$ are the $\Sstart$ states of a sub-model (not shown) identical
+    in structure to the singlet transducer.  They are used to insert a new stem-loop structure.
+    This is the indiegram-style transducer equivalent of the conditional pair TKFST SCFG in \secref{TKFST}.
+  }
+\end{table}
+
+\subsection{The multi-sequence TKFST model as a composite transducer}
+
+We can use the state graph construction algorithm described 
+in the paper and detailed in Text S1 to create a model of the simultaneous evolution
+of several sequences.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics [scale=0.4] {figs/parent.pdf}
+  \caption{\figlabel{parent}
+    A simple example of transducer composition to build a multi-sequence model of two extant sequences.
+    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
+    A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
+    and two branch transducers (the gray boxes labeled $\Delta T_X$ and $\Delta T_Y$)
+    mutate it according to the specified multi-sequence model.
+    Gray nodes correspond to observed data and white nodes unobserved data.
+  }
+\end{figure}
+
+Consider the simple model shown in \figref{parent}.
+The state of the multi-sequence model describing this model is a 3-vector $\bvec{a} = \left( a_1,\,a_2,\,a_3 \right)$,
+where $a_1$ is the state of the singlet transducer generating the ancestral sequence $W$
+and $a_2$ and $a_3$ are the states of the branch transducers evolving $W$ into
+extant sequences $X$ and $Y$.
+
+We can show some of the allowed transitions of this multi-sequence model.
+The state of the branch transducer associated with the active node $n$ is shown in bold.
+
+\paragraph{Stem creation.}
+
+A stem is created at the root node $W$ (corresponding to a bifurcation in the singlet transducer $a_1$)
+and survives in the sequence $X$ at node 2 but is deleted in the sequence $Y$ at node 3.
+
+\begin{align}
+  \nakedthreevec{1:}{2:}{3:} \quad
+  \threevec{L}{L}{\mathbf{L}} &
+  \to \threevec{\mathbf{L}}{W_L}{W_L}
+  \to \threevec{B}{B}{B_p}
+  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{S}{\mathbf{S}}{\Send}
+  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\mathbf{S}}{W_S}{\Send} \\
+  & \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\mathbf{S}}{W_S}{\Send}
+\end{align}
+
+\paragraph{Stem insertion.}
+
+A stem sequence is inserted in sequence $X$ at node 2.
+
+\begin{align}
+  \nakedthreevec{1:}{2:}{3:} \quad
+  \threevec{L}{L}{\mathbf{L}} \to \threevec{L}{\mathbf{L}}{W_L}
+  \to \threevec{L}{B_i}{B}
+  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\Send}{S}{\mathbf{S}}
+  \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\Send}{\mathbf{S}}{W_S}
+\end{align}
+
+\paragraph{Stem termination.}
+
+All stem sequences are ended by (possibly empty) loop sequences.
+
+\begin{align}
+  \nakedthreevec{1:}{2:}{3:} \quad
+  \threevec{\mathbf{S}}{W_S}{W_S}
+  \to \threevec{B_e}{B_e}{B_e}
+  \to \threevec{\Send}{\Send}{\Send} \, \threevec{L}{L}{\mathbf{L}}
+  \to \threevec{\Send}{\Send}{\Send} \, \threevec{\Send}{\Send}{\Send}
+\end{align}
+
+The functions $\alpha(t)$, $\beta(t)$ and $\gamma(t)$ are parametrized by the insertion and deletion rates of the Structure Tree model.
+They are defined for loop sequences as
+\begin{align}
+  \kappa_1 &= \lambda_1 / \mu_1 \\
+  \alpha_1 &= \exp \left( -\mu_1 t \right) \\
+  \beta_1 &= \frac{\lambda_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) } \\
+  \gamma_1 &= 1 - \frac{\mu_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\left( 1 - \exp (- \mu_1 t) \right) \left(\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) \right) }
+\end{align}
+and similarly for stem sequences \cite{Holmes2004}.

### 2009-06-04 s4.tex
renamed software appendix (Text S3) to Text S4; added new Text S3 for TKFST grammars; added a bunch of other stuff

--- 2009-05-29 supplementary.tex
@@
   {\Large
-    \textbf{\titlestring: \supptext}
+    \textbf{\titlestring: \supptext{4}}
   }

### 2009-06-04 simulations/plots.r

@@ -7,10 +7,14 @@ toPDF = TRUE
 
 alignColor = "blue"
 alignTKFColor = "cyan"
+alignStemlocColor = "skyblue"
+alignStemlocAmaColor = "lightskyblue"
 structureColor = "red"
 
 alignShape = 16
 alignTKFShape = 16
+alignStemlocShape = 16
+alignStemlocAmaShape = 16
 structureShape = 17
 
 ## default plot filename
@@ -36,6 +40,8 @@ data = read.table ("inferred/indiegram.analysis", header = TRUE, sep = "\t")
 ## average datapoints for each outgroup length
 align = array()
 alignTKF = array()
+alignStemloc = array()
+alignStemlocAma = array()
 structure = array()
 bins = unique (data$z)
 for (bin in bins) {
@@ -47,6 +53,14 @@ for (bin in bins) {
     alignTKF,
     c (bin, mean (subset (data$AlignAccTKF,
                           data$z == bin))))
+  alignStemloc = rbind (
+    alignStemloc,
+    c (bin, mean (subset (data$AlignAccStemloc,
+                          data$z == bin))))
+  alignStemlocAma = rbind (
+    alignStemlocAma,
+    c (bin, mean (subset (data$AlignAccStemlocAma,
+                          data$z == bin))))
   structure = rbind (
     structure,
     c (bin, mean (subset (data$StructureReconstruction,
@@ -56,6 +70,8 @@ for (bin in bins) {
 ## remove NA rows
 align = align[-1,]
 alignTKF = alignTKF[-1,]
+alignStemloc = alignStemloc[-1,]
+alignStemlocAma = alignStemlocAma[-1,]
 structure = structure[-1,]
 
 plot.new()
@@ -72,17 +88,30 @@ points (align,
         col = alignColor, pch = alignShape, cex = 1)
 lines (lowess (align), col = alignColor, lwd = 1.5)
 
+
 points (alignTKF,
         col = alignTKFColor, pch = alignTKFShape, cex = 1)
 lines (lowess (alignTKF), col = alignTKFColor, lwd = 1.5)
+
+
+points (alignStemloc,
+        col = alignStemlocColor, pch = alignStemlocShape, cex = 1)
+lines (lowess (alignStemloc), col = alignStemlocColor, lwd = 1.5)
+
+
+points (alignStemlocAma,
+        col = alignStemlocAmaColor, pch = alignStemlocAmaShape, cex = 1)
[…]

### 2009-06-04 simulations/run-long.pl

@@ -1,0 +1,87 @@
+#!/usr/bin/perl -w
+
+# imports
+use Stockholm;
+use FindBin qw($Bin);
+
+# read command-line args
+my ($true_alignment) = @ARGV;
+
+# create temp filenames
+my $temp_alignment = "$true_alignment.handalign-input";
+
+# define constants
+my $tiny = .00001;  # default length for dummy branch
+my $handalign = "handalign";  # path to handalign binary
+my $loopmodel = "$Bin/loopmodel.eg";  # substitution model for handalign (should be equivalent to loop-submodel of pfold model used by evolsayer.pl to generate alignments)
+
+# load alignment
+my $true = Stockholm->from_file ($true_alignment);
+warn "True alignment\n";
+warn $true->to_string;
+
+# read stuff from alignment
+my %seq = %{$true->seqdata};
+my $treestr = join "", @{$true->gf_NH};
+my $args = join "", @{$true->gf_PARAM};
+
+# extract branch lengths & sequences
+my (%branchlen, %seqlen);
+for my $branch (qw(X Y Z)) {
+    if ($treestr =~ /$branch:([0-9\.eE\+\-]+)/) {
+	$branchlen{$branch} = $1;
+    } else {
+	die "Couldn't find $branch in $treestr";
+    }
+    die "Couldn't find row $branch in alignment" unless exists $seq{$branch};
+    $seq{$branch} =~ s/[\.\-_]//g;
+    $seqlen{$branch} = length $seq{$branch};
+}
+
+# create rows of new alignment
+my %newrow;
+$newrow{'Z'} = $seq{'Z'} . ("-" x ($seqlen{'Y'} + $seqlen{'X'}));
+$newrow{'Y'} = ("-" x $seqlen{'Z'}) . $seq{'Y'} . ("-" x $seqlen{'X'});
+$newrow{'X'} = ("-" x ($seqlen{'Z'} + $seqlen{'Y'})) . $seq{'X'};
+
+$newrow{'xyzroot'} = $newrow{'Z'};
+$newrow{'xyzroot'} =~ s/[^\.\-_]/*/g;
+
+$newrow{'xyroot'} = $branchlen{'X'} < $branchlen{'Y'} ? $newrow{'X'} : $newrow{'Y'};
+$newrow{'xyroot'} =~ s/[^\.\-_]/*/g;
+
+# create new alignment
+my $new = Stockholm->new;
+
+$new->seqname ([qw(xyzroot xyroot X Y Z)]);
[…]

### 2009-06-04 paper.tex
renamed software appendix (Text S3) to Text S4; added new Text S3 for TKFST grammars; added a bunch of other stuff, including long indel model tests

@@
 \begin{description}
-\item{TKF91 ML:} with the TKF91 model \cite{}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
+\item{TKF91 ML:} with the TKF91 model \cite{ThorneEtal91}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
 The correct phylogenetic tree and branch lengths were supplied (as they were for the \indiegram\ benchmark).
 The insertion, deletion and substitution rates for the TKF91 model were set equal to those of the loop submodel of TKFST.
+{\bf This may be understood as a naive reconstruction that completely ignores the stem sub-model of TKFST.}
 \item{Long Indel ML:} with a single-event trajectory approximation to the long indel model \cite{MiklosLunterHolmes2004}, the \handel\ package was used to align the three extant sequences and reconstruct the ancestor.
@@
 (\secname{A simple model of RNA structural evolution} has formulae for these quantities in terms of the TKFST rate parameters).
+{\bf This model improves on the previous model (TKF91) by introducing affine gap penalties.}
 \item{Stemloc ML:} the \stemloc\ program was used to align the three sequences via single-linkage clustering with a Pair SCFG \cite{Holmes2005}.
 The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
-\item{Stemloc AMA:} the \stemloc\ program was used to align the three sequences in AMA mode, using posterior decoding to optimize a sum-over-pairs expected alignment accuracy \cite{BradleyPachterHolmes2008}.
+{\bf This model improves on the previous model (Long Indel) by introducing RNA secondary structure, though it no longer models phylogeny as rigorously as the Long Indel model.}
+\item{Stemloc AMA:} the \stemloc\ program was used to align the three sequences in ``sequence annealing'' mode,
+ a posterior decoding method that attempts to optimize AMA, a sum-over-pairs alignment accuracy metric \cite{BradleyPachterHolmes2008}.
 The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
+{\bf This program uses the previous model but abandons phylogeny altogether, instead attempting to maximize a non-phylogenetic alignment accuracy metric.}
 \end{description}
@@
     base-pairs in the ancestral structure present in the
-    maximum-likelihood reconstruction thereof; ``Alignment accuracy''
-    is a measure of sequence alignment accuracy which takes both
-    aligned and unaligned (indels) nucleotides into account,
-    qualitatively corresponding to giving equal weight to sensitivity
-    and specificity (see
-    \cite{SchwartzMyersPachter2006} for a precise definition).
+    maximum-likelihood reconstruction thereof, estimated by aligning the two structures using \stemloc\ \cite{Holmes2005}.
+    ``Alignment accuracy'' is the Alignment Metric Accuracy (AMA),
+    effectively the probability that any given residue is correctly
+    aligned, averaged over all pairs of sequences
+    (see \cite{SchwartzMyersPachter2006} for a precise definition).
 }

### 2009-06-04 simulations/compare.pl
updated indiegram.analysis rules to include TCS for all tests, and added long indel test

@@ -8,17 +8,19 @@ my $tmpdir = "/tmp";
 my $ref = Stockholm->from_file (shift);
 my $cmp = Stockholm->from_file (shift);
 my $tkf = Stockholm->from_file (shift);
+my $long = Stockholm->from_file (shift);
 my $stemloc = Stockholm->from_file (shift);
 my $stemlocama = Stockholm->from_file (shift);
 
 # indiegram adds cruft to the sequence names and prints non-Stockholm stuff that gets read back incorrectly by Stockholm.pm
 # handalign and indiegram both add "root" and "subroot" lines
 # so we need to tidy up the sequence name fields...
-@{$ref->seqname} = @{$cmp->seqname} = @{$tkf->seqname} = ("X", "Y", "Z");
+for my $align ($ref, $cmp, $tkf, $long, $stemloc, $stemlocama) { @{$align->seqname} = ("X", "Y", "Z") }
 
 $ref->gc->{"ancestral_SS"} = $ref->gr->{"SS"}->{"root"};
 delete $ref->gr->{"SS"}->{"root"};
 
+# IH (6/4/09)... the following code scares the bejaysus out of me... can you really mix "delete" and "each" like this? wow
 while (my ($name, $data) = each %{$cmp->seqdata}) {
   if ($name =~ /([XYZ])\/.+/) {
     delete $cmp->seqdata->{$name};
@@ -28,12 +30,12 @@ while (my ($name, $data) = each %{$cmp->seqdata}) {
   }
 }
 
-for my $file (qw(ref cmp tkf stemloc stemlocama)) {
+for my $file (qw(ref cmp tkf long stemloc stemlocama)) {
     eval('$'.$file)->to_file ("$tmpdir/$file.stock");
 }
 
-my ($acc, $sn, $ppv, $tcs) = cmpalign ("$tmpdir/ref.stock", "$tmpdir/cmp.stock");
-my @acc_sn_ppv_tcs = map ([cmpalign ("$tmpdir/ref.stock", "$tmpdir/$_.stock")], qw(tkf stemloc stemlocama));
+my @tests = qw(cmp tkf long stemloc stemlocama);
+my @acc_sn_ppv_tcs = map ([cmpalign ("$tmpdir/ref.stock", "$tmpdir/$_.stock")], @tests);
 
[…]

### 2009-06-04 simulations/plots.r
long indel model test

@@ -7,14 +7,16 @@ toPDF = TRUE
 
 alignColor = "blue"
 alignTKFColor = "cyan"
+alignLongColor = "darkblue"
 alignStemlocColor = "skyblue"
-alignStemlocAmaColor = "lightskyblue"
+alignStemlocAmaColor = "darkblue"
 structureColor = "red"
 
 alignShape = 16
-alignTKFShape = 16
-alignStemlocShape = 16
-alignStemlocAmaShape = 16
+alignTKFShape = 1
+alignLongShape = 2
+alignStemlocShape = 3
+alignStemlocAmaShape = 4
 structureShape = 17
 
 ## default plot filename
@@ -40,6 +42,7 @@ data = read.table ("inferred/indiegram.analysis", header = TRUE, sep = "\t")
 ## average datapoints for each outgroup length
 align = array()
 alignTKF = array()
+alignLong = array()
 alignStemloc = array()
 alignStemlocAma = array()
 structure = array()
@@ -52,6 +55,10 @@ for (bin in bins) {
   alignTKF = rbind (
     alignTKF,
     c (bin, mean (subset (data$AlignAccTKF,
+                          data$z == bin))))
+  alignLong = rbind (
+    alignTKF,
+    c (bin, mean (subset (data$AlignAccLong,
                           data$z == bin))))
   alignStemloc = rbind (
     alignStemloc,
@@ -70,6 +77,7 @@ for (bin in bins) {
 ## remove NA rows
 align = align[-1,]
 alignTKF = alignTKF[-1,]
+alignLong = alignLong[-1,]
 alignStemloc = alignStemloc[-1,]
 alignStemlocAma = alignStemlocAma[-1,]
 structure = structure[-1,]
@@ -94,6 +102,11 @@ points (alignTKF,
 lines (lowess (alignTKF), col = alignTKFColor, lwd = 1.5)
 
 
+points (alignLong,
+        col = alignLongColor, pch = alignLongShape, cex = 1)
+lines (lowess (alignLong), col = alignLongColor, lwd = 1.5)
+
+
 points (alignStemloc,
         col = alignStemlocColor, pch = alignStemlocShape, cex = 1)
 lines (lowess (alignStemloc), col = alignStemlocColor, lwd = 1.5)
@@ -109,9 +122,9 @@ points (structure,
 lines (lowess (structure), col = structureColor, lwd = 1.5)
 
 
[…]

### 2009-06-05 simulations/debug-0.1-13.pl
debug script to fetch initialization lines for relevant transitions and bifurcations in test-case parse tree

@@ -1,0 +1,11 @@
+#!/usr/bin/perl
+$tkfst = $ENV{'DARTDIR'} . "/src/indiegram/tkfst.cc";
+$parse = "alignments/0.1/13.reference.parse";
+open IN, "<$parse" or die;
+open OUT, ">$parse.inits";
+print OUT "// $parse.inits: generated by $0 from $parse and $tkfst\n\n";
+while(<IN>){next unless /^\s*[^;\s]/;chomp;if(/\(?(\S+).*\->\s+(\S+)\s+(\S+)/){mysys("egrep \"init_bifurc.*$1.*$2.*$3\" $tkfst");$x=undef}else{s/^\s*\(?([^\(\)\s]+).*/$1/; if ( defined$x ) {($a,$b)=/end/?($_,$x):($x,$_);mysys("egrep '^[^M]*transition_scores\[^M]*$a\[^M]*$b' $tkfst")}$x=$_ eq"end"?undef:$_}}
+sub mysys { my$c=shift; # print OUT "// $c\n",`$c`;
+print OUT `$c`}
+close IN;
+close OUT or die;

### 2009-06-05 simulations/debug-0.1-13.pl
debug script to fetch initialization lines for relevant transitions and bifurcations in test-case parse tree

@@ -4,8 +4,12 @@ $parse = "alignments/0.1/13.reference.parse";
 open IN, "<$parse" or die;
 open OUT, ">$parse.inits";
 print OUT "// $parse.inits: generated by $0 from $parse and $tkfst\n\n";
-while(<IN>){next unless /^\s*[^;\s]/;chomp;if(/\(?(\S+).*\->\s+(\S+)\s+(\S+)/){mysys("egrep \"init_bifurc.*$1.*$2.*$3\" $tkfst");$x=undef}else{s/^\s*\(?([^\(\)\s]+).*/$1/; if ( defined$x ) {($a,$b)=/end/?($_,$x):($x,$_);mysys("egrep '^[^M]*transition_scores\[^M]*$a\[^M]*$b' $tkfst")}$x=$_ eq"end"?undef:$_}}
-sub mysys { my$c=shift; # print OUT "// $c\n",`$c`;
-print OUT `$c`}
+while(<IN>){chomp;if(/^\s*\(?([^\(\);\s]+)/){$y=$1; if ( defined$x ) {($a,$b)=($y=~/end/)?($y,$x):($x,$y);mysys("egrep '^[^M]*transition_scores".(($a=~/start|end/)?"[^M]*$a\[^M]*$b":"\\.transition \\($a, $b\\)")."' $tkfst")}
+if(/\(?(\S+).*\->\s+(\S+)\s+(\S+)/){mysys("egrep \"init_bifurc.*$1.*$2.*$3\" $tkfst");$x=undef}else{$x=$y eq"end"?undef:$y}
+}}
+sub mysys { my$c=shift;
+print OUT "// $c\n",`$c`,"\n";
+#print OUT `$c`;
+}
 close IN;
 close OUT or die;

### 2009-06-12 simulations/plots.r
Bugfix for legend.

@@ -122,9 +122,10 @@ points (structure,
 lines (lowess (structure), col = structureColor, lwd = 1.5)
 
 
-legend ("bottomleft", legend = c ("Structural reconstruction accuracy (TKFST ML)", "Alignment accuracy (TKFST ML)", "Alignment accuracy (TKF91 ML)", "Alignment accuracy (Long Indel ML)", "Alignment accuracy (Stemloc ML)", "Alignment accuracy (Stemloc AMA)"),
+legend (x = "bottomleft",
+        legend = c ("Structural reconstruction accuracy (TKFST ML)", "Alignment accuracy (TKFST ML)", "Alignment accuracy (TKF91 ML)", "Alignment accuracy (Long Indel ML)", "Alignment accuracy (Stemloc ML)", "Alignment accuracy (Stemloc AMA)"),
         col = c (structureColor, alignColor, alignTKFColor, alignLongColor, alignStemlocColor, alignStemlocAmaColor),
-        pch = c (structureShape, alignShape, alignTKFShape, alignLongShape, alignStemlocColor, alignStemlocAmaColor),
+        pch = c (structureShape, alignShape, alignTKFShape, alignLongShape, alignStemlocShape, alignStemlocAmaShape),
         xjust = 1,
         inset = 0.02)
 

### 2009-06-12 simulations/compare.pl
fix

@@ -12,6 +12,8 @@ my $long = Stockholm->from_file (shift);
 my $stemloc = Stockholm->from_file (shift);
 my $stemlocama = Stockholm->from_file (shift);
 
+my $cmpalignexec = "$ENV{'DARTDIR'}/bin/cmpalign";
+
 # indiegram adds cruft to the sequence names and prints non-Stockholm stuff that gets read back incorrectly by Stockholm.pm
 # handalign and indiegram both add "root" and "subroot" lines
 # so we need to tidy up the sequence name fields...
@@ -99,7 +101,7 @@ print join ("\t", $ancoverlap, map (@$_, @acc_sn_ppv_tcs)), "\n";
 
 sub cmpalign {
     my ($ref, $cmp) = @_;
-    my $cmd = "cmpalign -s $ref $cmp";
+    my $cmd = "$cmpalignexec -s $ref $cmp";
 #    warn "Running $cmd\n";
     my $line = `$cmd`;
     warn "$cmd  ==>  $line";

### 2009-06-12 paper.tex
Minor changes.

@@
 The results of our simulations, illustrated in \figref{simulations},
-demonstrate that the accuracy of both structural
-reconstruction and alignment accuracy depend on the outgroup
-branch length.  However, while the difficulty of alignment inference
-increases in proportion to the outgroup branch length, becoming
-near-impossible for very long outgroup branch lengths, the same need
-not be true for structural reconstruction accuracy.  As the branch
-length of the outgroup increases, the outgroup's structure becomes increasingly decoupled from
-that of its ancestor, allowing the scoring scheme to ``ignore'' the
-information which it contains.
-In the trivial case where the outgroup branch length is zero (so the ancestor is identical to the outgroup),
-we can perfectly reconstruct the ancestral structure (as expected),
-but we cannot perfectly reconstruct the three-way multiple alignment
- (since there are still two finite-length branches,
- which on average contribute a greater-than-zero rate of unreconstructable mutations).
-
-These results correspond to a best-case scenario in which we
-have a perfectly-parametrized evolutionary model and exactly-known descendant structures.
-Recent results
-demonstrating that inference of sequence alignment is
-robust to mis-parametrized probabilistic models \cite{Lunter2007}
-suggest that the same may be true for probabilistic inference of
-ancestral structures.
+demonstrate that while the difficulty of alignment inference increases
+in proportion to the outgroup branch length, becoming very challenging
+for long outgroup branch lengths, the same need not be true for
+structural reconstruction accuracy.  As the branch length of the
+outgroup increases, the outgroup's structure becomes increasingly
+decoupled from that of its ancestor, allowing the scoring scheme to
+``ignore'' the information which it contains.  In the trivial case
+where the outgroup branch length is zero (so the ancestor is identical
+to the outgroup), we can perfectly reconstruct the ancestral structure
+(as expected), but we cannot perfectly reconstruct the three-way
+multiple alignment (since there are still two finite-length branches,
+which on average contribute a greater-than-zero rate of
+unreconstructable mutations).
+
+These results correspond to a best-case scenario in which we have a
+perfectly-parametrized evolutionary model and exactly-known descendant
+structures.  Recent results demonstrating that inference of sequence
+alignment is robust to mis-parametrized probabilistic models
+\cite{Lunter2007} suggest that the same may be true for probabilistic
+inference of ancestral structures.
 %The effect of imperfect knowledge of the ancestral structures has not been explored.
@@
 To compare \indiegram\ to simpler reconstruction methods, we also measured the alignment accuracy using various simpler statistical reconstruction techniques.
-We evaluated three maximum likelihood inference algorithms that use simpler gap-aware phylogenetic models,
+We evaluated three maximum likelihood inference algorithms that use simpler gap-aware phylogenetic models
 and one posterior decoding algorithm that uses a non-phylogenetic sum-of-pairs scoring scheme.

### 2009-06-15 simulations/plots.r
Separated simulation plots

@@ -1,26 +1,28 @@
 
 require (graphics)
-require (CarbonEL)
+## require (CarbonEL)
 require (stats)
 
 toPDF = TRUE
 
-alignColor = "blue"
-alignTKFColor = "cyan"
-alignLongColor = "darkblue"
-alignStemlocColor = "skyblue"
-alignStemlocAmaColor = "darkblue"
+alignColor = "red"
+alignTKFColor = "green"
+alignLongColor = "blue"
+alignStemlocColor = "yellow"
+alignStemlocAmaColor = "purple"
+
 structureColor = "red"
 
 alignShape = 16
-alignTKFShape = 1
-alignLongShape = 2
-alignStemlocShape = 3
-alignStemlocAmaShape = 4
-structureShape = 17
+alignTKFShape = 15
+alignLongShape = 1
+alignStemlocShape = 2
+alignStemlocAmaShape = 20
+
+structureShape = 16
 
 ## default plot filename
-plotfile = "simulations.pdf"
+plotfile = "ama.pdf"
 plotdim = c (6, 4)
 
 ## open PDF
@@ -40,14 +42,24 @@ par (oma = c (0, 0, 0, 0))
 data = read.table ("inferred/indiegram.analysis", header = TRUE, sep = "\t")
 
 ## average datapoints for each outgroup length
+
 align = array()
 alignTKF = array()
 alignLong = array()
 alignStemloc = array()
 alignStemlocAma = array()
+
+alignTCS = array()
+alignTCSTKF = array()
+alignTCSLong = array()
+alignTCSStemloc = array()
+alignTCSStemlocAma = array()
+
 structure = array()
+
 bins = unique (data$z)
 for (bin in bins) {
+
   align = rbind (
     align,
     c (bin, mean (subset (data$AlignAcc,
@@ -68,6 +80,28 @@ for (bin in bins) {
     alignStemlocAma,
     c (bin, mean (subset (data$AlignAccStemlocAma,
                           data$z == bin))))
+
+  alignTCS = rbind (
+    alignTCS,
+    c (bin, mean (subset (data$AlignTCS,
+                          data$z == bin))))
+  alignTCSTKF = rbind (
+    alignTCSTKF,
+    c (bin, mean (subset (data$AlignTCSTKF,
+                          data$z == bin))))
+  alignTCSLong = rbind (
+    alignTCSTKF,
+    c (bin, mean (subset (data$AlignTCSLong,
[…]

### 2009-06-15 paper.tex
paper updates

--- 2009-06-04 paper.tex
@@
 % 4: stree-mutations-with-structures.pdf
-% 5: simulations.pdf
+% 5: perfect.pdf
+% 6: ama.pdf
 
@@
 
-We implemented these algorithms for a three-taxon phylogeny.  Using
-our implementation on a simple model of pairwise RNA structural
-evolution, we simulated data under the model and tested the dependence
-of alignment inference and ancestral structure reconstruction accuracy on phylogenetic branch length.
-Our results suggest that
-we may be able to accurately reconstruct ancestral structures despite
-sequence and structural divergence if the evolutionary process is
-well-understood.
-We discuss the implications of our results for RNA algorithm design
-and phylogenetic reconstruction of RNA evolutionary history.
-
-The programs described are available
-as part of the \dart\ software package for sequence analysis,
-released under the GNU Public License at
-\darturl.
+We implemented the above algorithms for a simple model of pairwise RNA structural evolution;
+in particular the algorithms for maximum likelihood (ML) alignment of three known RNA structures an a known phylogeny
+and inference of the common ancestral structure.
+We compared this ML algorithm to a variety of related, but simpler, techniques,
+including ML alignment algorithms for simpler models that omitted various aspects of the full model
+and also a heuristic (sum-over-pairs) posterior-decoding alignment algorithm for one of the simpler models.
+
+Our results suggest that ML algorithms incorporating basepair structure are the most likely to estimate the alignment perfectly.
+However, the heuristic posterior-decoding algorithm scores higher on finer-grained alignment accuracy metrics.
+Since the posterior-decoding heuristic is substantially faster than exact phylogenetic ML,
+this further motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics
+for paleogenetics and other alignment problems.
 
@@
 Such constraints are
-commonly used by programs for SCFG-based RNA sequence analysis such as QRNA \cite{RivasEddy2001}, Stemloc \cite{Holmes2005} and CONSAN \cite{DowellEddy2006}.
+commonly used by programs for SCFG-based RNA sequence analysis such as QRNA \cite{RivasEddy2001}, \stemloc\ \cite{Holmes2005} and CONSAN \cite{DowellEddy2006}.
 Alignment and structural constraints can be combined \cite{Holmes2005}.
@@
 
-\subsection*{The dependence of reconstruction accuracy on outgroup
-  branch length}
+\subsection*{Comparison of alignment methods}
 
@@
 the reconstruction problem was made easy by the presence of a close
-outgroup, we conducted a simulation study to investigate the
-dependence of reconstruction accuracy on outgroup branch length,
-and to investigate the performance of our reconstruction method (when simulating directly from the model)
-compared to simpler reconstruction methods that ignore either structure or phylogeny.
+outgroup, we conducted a simulation study of the dependence of reconstruction accuracy on outgroup branch length,
+with the further goal of comparing the performance of our reconstruction method (when simulating directly from the model)
+to simpler reconstruction methods that ignore either structure or phylogeny.
 (We here use the term ``outgroup'' loosely to denote the
@@
 sibling species constant and varied the branch length of the
-outgroup.  We chose the parameters of the TKFST model to
-generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
-alignment) and considered only alignments with at least two ancestral
-stems, sequences of length $\in [30, 70]$ bp, loops of length $\in [3,
-10]$ bp and stems of lengths $\in [1, 20]$ bp.
-We then used \indiegram\ to reconstruct maximum likelihood reconstructions of the ancestral structure and the four-way multiple alignment (ancestor + three descendants).
-
-The results of our simulations, illustrated in \figref{simulations},
-demonstrate that the accuracy of both structural
-reconstruction and alignment accuracy depend on the outgroup
-branch length.  However, while the difficulty of alignment inference
-increases in proportion to the outgroup branch length, becoming
-near-impossible for very long outgroup branch lengths, the same need
-not be true for structural reconstruction accuracy.  As the branch
-length of the outgroup increases, the outgroup's structure becomes increasingly decoupled from
-that of its ancestor, allowing the scoring scheme to ``ignore'' the
-information which it contains.
-In the trivial case where the outgroup branch length is zero (so the ancestor is identical to the outgroup),
-we can perfectly reconstruct the ancestral structure (as expected),
-but we cannot perfectly reconstruct the three-way multiple alignment
- (since there are still two finite-length branches,
- which on average contribute a greater-than-zero rate of unreconstructable mutations).
-
-These results correspond to a best-case scenario in which we
-have a perfectly-parametrized evolutionary model and exactly-known descendant structures.
-Recent results
-demonstrating that inference of sequence alignment is
-robust to mis-parametrized probabilistic models \cite{Lunter2007}
-suggest that the same may be true for probabilistic inference of
-ancestral structures.
-%The effect of imperfect knowledge of the ancestral structures has not been explored.
-% [commented out because the following tests DO address this in a roundabout way; what we mean is that indiegram has not been tested when leaf structures are unknown]
-
-To compare \indiegram\ to simpler reconstruction methods, we also measured the alignment accuracy using various simpler statistical reconstruction techniques.
-We evaluated three maximum likelihood inference algorithms that use simpler gap-aware phylogenetic models,
-and one posterior decoding algorithm that uses a non-phylogenetic sum-of-pairs scoring scheme.
-These results are plotted on \figref{simulations} alongside the \indiegram\ accuracies.
-In detail, the reconstruction methods used were
+outgroup between $[0, 2.5]$ at steps of size $0.1$.
+    Parameters used in the simulation were $\lambda_l = 0.025$ and $\mu_l = 0.03$ for
+    loop sequence and $\lambda_l = 0.007$ and $\mu_l = 0.01$ for stem
+    sequence; the probability of a stem insertion was $0.1$.  These
+    yielded a mean loop length of 5 bp and a mean stem length of 2.33 bp,
+    with $\sim 0.3$ substructure indels per alignment.
+    We selected alignments to reconstruct by requiring
+    that there be at least two ancestral stems, loops of length $\in [3,
+    10]$ bp and stems of length $\in [1, 20]$ bp; to reduce the complexity of
+    our algorithms we additionally required that the sequences have
+    lengths $\in [30, 70]$ bp.
+
+We then attempted three-way multiple alignment (and, in some cases, reconstruction of the ancestor)
+using a variety of statistical inference algorithms.
+We sought insight as to the relative importance of the following factors in reconstructing ancestral RNA:
+(i) modeling the secondary structure;
+(ii) modeling the phylogenetic topology \& branch lengths;
+(iii) using posterior-decoding algorithms to maximize the expected alignment {\em accuracy},
+rather than picking the single {\em most likely} alignment \cite{Holmes98,SchwartzPachter2007,BradleyPachterHolmes2008}.
+
+The alignment programs we used in this benchmark were
+\indiegram\ (exact ML inference of alignment and ancestral structure, given phylogeny, descendant structures and correct model);
+\stemloc\ (a greedy ML heuristic, ignoring phylogeny in favor of a single-linkage clustering of the descendant structures);
+\stemlocama\ (a posterior-decoding heuristic, maximizing the alignment's {\em expected accuracy} rather than its {\em likelihood});
+and
+\handel\ (ML alignment under various indel models that ignore secondary structure completely).
+In detail, the reconstruction methods were
 
 \begin{description}
-\item{TKF91 ML:} with the TKF91 model \cite{}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
+\item{\stemloc:} the \stemloc\ program was used to align the three sequences via single-linkage clustering with a Pair SCFG \cite{Holmes2005}.
+The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
+{\bf This model improves on the previous model (Long Indel) by introducing RNA secondary structure, though it no longer models phylogeny as rigorously as the Long Indel model} (it uses single-linkage clustering instead of a true phylogeny).
+\item{\stemlocama:} the \stemloc\ program was used to align the three sequences in ``sequence annealing'' mode,
+ a posterior decoding method that attempts to optimize AMA, a sum-over-pairs alignment accuracy metric \cite{BradleyPachterHolmes2008}.
+The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
+{\bf This program uses the previous model but instead of maximizing likelihood, it attempts to maximize an alignment accuracy metric.}
+\item{TKF91:} with the TKF91 model \cite{ThorneEtal91}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
 The correct phylogenetic tree and branch lengths were supplied (as they were for the \indiegram\ benchmark).
 The insertion, deletion and substitution rates for the TKF91 model were set equal to those of the loop submodel of TKFST.
-\item{Long Indel ML:} with a single-event trajectory approximation to the long indel model \cite{MiklosLunterHolmes2004}, the \handel\ package was used to align the three extant sequences and reconstruct the ancestor.
+{\bf This may be understood as a naive reconstruction that completely ignores the stem sub-model of TKFST.}
+\item{Long Indel:} with a single-event trajectory approximation to the long indel model \cite{MiklosLunterHolmes2004}, the \handel\ package was used to align the three extant sequences and reconstruct the ancestor.
 The correct phylogenetic tree and branch lengths were supplied.
@@
 (\secname{A simple model of RNA structural evolution} has formulae for these quantities in terms of the TKFST rate parameters).
-\item{Stemloc ML:} the \stemloc\ program was used to align the three sequences via single-linkage clustering with a Pair SCFG \cite{Holmes2005}.
-The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
-\item{Stemloc AMA:} the \stemloc\ program was used to align the three sequences in AMA mode, using posterior decoding to optimize a sum-over-pairs expected alignment accuracy \cite{BradleyPachterHolmes2008}.
-The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
+{\bf This model improves on the previous model (TKF91) by introducing affine gap penalties.}
 \end{description}
 
-We did not measure the accuracy of the reconstructed ancestral structure in these tests (as we did in the \indiegram\ test),
-since \handel\ does not consider structures, whereas \stemloc\ and \stemlocama\ do not provide ancestral reconstructions.
-
-{\bf SUMMARY OF THESE RESULTS}
-% TKF is less accurate than indiegram, although for short branch lengths it is almost identical; the TKF and indiegram accuracies diverge at longer branch lengths
-% Long indel?
-% Stemloc?
-% StemlocAMA?
+We focused primarily on alignment accuracy, under the simplifying assumption that this will be a good indicator for
+(and perhaps the primary determinant of) the accuracy of ancestral reconstruction.
+
+We first consider the {\em perfect alignment rate}; that is, the number of times each method gets the alignment exactly correct.
+Theory predicts that ML inference, using the correct model and parameters, should maximize the perfect alignment rate.
+Inspecting \figref{perfect}, we find this to be the case, except when the outgroup is very distant
+and the best perfect alignment rate drops below 0.05 (at which point 125 trials may not be sampling it sufficiently).
+We also note that the ML version of \stemloc\ is near-optimal, and slightly better than the posterior-decoding \stemloc-AMA.
+Finally, we note that the structure-blind models (TKF91 and Long Indel) perform consistently worse than the structure-aware methods,
+with no decisive advantage shown by either TKF91 (linear gap penalty) or Long Indel (affine gap penalty).
+
+A subtly different trend is apparent when we look at the accuracy of the returned alignment,
+measured not by the all-or-nothing metric of whether the alignment is {\em perfect},
+but rather by a metric that count the number of correct alignment columns.
+\figref{ama} shows the {\em Alignment Metric Accuracy} (AMA),
+a generalization of the {\em Sum-of Pairs Score} (SPS)
+which measures the proportion of residues which are correctly aligned, averaged over all pairs of sequences \cite{SchwartzMyersPachter2006}.
+Here, we see that \stemlocama---which attempts to find the alignment with maximum expected AMA---edges out both
+\indiegram\ and the ML version of \stemloc.
+The decision-theoretic scoring scheme of \stemlocama\ appears to offset the minor incorrectness of its model.
+
+Again, when looking at the AMA statistics, the most important factor is clearly the incorporation of some form of basepair structure
+into the model: \handel, whether using TKF91 or the Long Indel model, performs much worse than the SCFG-based methods.
+This is to be expected: whenever a basepair-aware method aligns one half of a basepair, it gets the other nucleotide
+correctly aligned for free.
 
@@
 
+Given the excellent performance of \stemlocama's sequence annealing, particularly when measured using its own scoring metric (AMA),
+such posterior-decoding methods should also be considered for reconstruction.
+
 In summary, the evolutionary models and algorithms we have described form a
@@
   \centering
-  % \includegraphics[width=5in]{figs_plos/simulations.pdf}
+  % \includegraphics[width=5in]{figs_plos/perfect.pdf}
+  \caption{
+    \textbf{Dependence of perfect alignment rate on outgroup branch
+      length.}
+    We investigated the perfect alignment rate
+    (the number of times that a given alignment program estimates the alignment 100\% correctly, with no errors)
+    as a function of outgroup branch length, simulating the
+    evolution of three structural RNAs under our model.  The two
+    sister species were equidistant from the ancestral sequence, with
+    a branch length of 1, corresponding to one expected
+    substitution per site in loop sequence.  We allowed the outgroup
+    branch length $t$ to vary between $[0, 2.5]$.  We selected 25 equally spaced values of $t$ in this range, spaced $0.1$ apart,
+    and simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
+    Since the perfect alignment rate is rather low, we further aggregated the $t$-values for this plot into bins of five;
+    thus, for example, the bin named ``$0 to 0.5$'' includes $t \in \{ 0, 0.1, 0.2, 0.3, 0.4 \}$.
+    The vertical axis counts the number of perfect alignments in each bin, out of a total of 125 trials per bin.
+}
+  \figlabel{perfect}
+\end{figure}
+
+% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
+\begin{figure}[!ht]
+  \centering
+  % \includegraphics[width=5in]{figs_plos/ama.pdf}
   \caption{
@@
     substitution per site in loop sequence.  We allowed the outgroup
-    branch length to vary between $[0, 2.5]$.  We divided this interval
-    into 25 bins and simulated 25 alignments for each bin.
-    Parameters used were $\lambda_l = 0.025$ and $\mu_l = 0.03$ for
-    loop sequence and $\lambda_l = 0.007$ and $\mu_l = 0.01$ for stem
-    sequence; the probability of a stem insertion was $0.1$.  These
-    yielded a mean loop length of 5 bp and a mean stem length of 2.33 bp,
-    with $\sim 0.3$ substructure indels per alignment.
-    We selected alignments to reconstruct by requiring
-    that there be at least two ancestral stems, loops of length $\in [3,
-    10]$ bp and stems of length $\in [1, 20]$ bp; to reduce the complexity of
-    our algorithms we additionally required that the sequences have
-    lengths $\in [30, 70]$ bp.
-    ``Structural reconstruction accuracy'' is the fraction of
-    base-pairs in the ancestral structure present in the
-    maximum-likelihood reconstruction thereof; ``Alignment accuracy''
-    is a measure of sequence alignment accuracy which takes both
-    aligned and unaligned (indels) nucleotides into account,
-    qualitatively corresponding to giving equal weight to sensitivity
-    and specificity (see
-    \cite{SchwartzMyersPachter2006} for a precise definition).
-}
-  \figlabel{simulations}
+    branch length $t$ to vary between $[0, 2.5]$.  We selected 25 equally spaced values of $t$ in this range, spaced $0.1$ apart,
+    and simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
+    The Alignment Metric Accuracy (AMA) is, roughly, the proportion of residues that are correctly
+    aligned, averaged over all pairs of sequences
+    (see \cite{SchwartzMyersPachter2006} for a precise definition; we used AMA with Gap Factor 1).
+}
+  \figlabel{ama}
 \end{figure}

### 2009-06-15 simulations/plots.r

@@ -291,7 +291,7 @@ bigbins = unique (round ((bins / delta) / chunks)) * binwidth + binwidth / 2
 bigbins = bigbins[-6]
 
 # get bin names
-#binnamefun <- function(x) cat (x - binwidth/2, " to ", x + binwidth/2)
+#binnamefun <- function(x) cat (x - binwidth/2, " to ", x + binwidth/2 - delta)
 #binname = apply(bigbins, 1, binnamefun)
 binname = c("0 to 0.5", "0.5 to 1", "1 to 1.5", "1.5 to 2", "2 to 2.5");
 

### 2009-06-15 paper.tex
paper updates

@@
 Theory predicts that ML inference, using the correct model and parameters, should maximize the perfect alignment rate.
-Inspecting \figref{perfect}, we find this to be the case, except when the outgroup is very distant
-and the best perfect alignment rate drops below 0.05 (at which point 125 trials may not be sampling it sufficiently).
+Inspecting \figref{perfect}, we find this to be the case, except where the outgroup is very distant and the bins are probably undersampled
+(when the perfect alignment rate drops below $\sim 5/125$, 125 trials may be too few).
 We also note that the ML version of \stemloc\ is near-optimal, and slightly better than the posterior-decoding \stemloc-AMA.
@@
     Since the perfect alignment rate is rather low, we further aggregated the $t$-values for this plot into bins of five;
-    thus, for example, the bin named ``$0 to 0.5$'' includes $t \in \{ 0, 0.1, 0.2, 0.3, 0.4 \}$.
+    thus, for example, the bin named ``$0 to 0.4$'' includes $t \in \{ 0, 0.1, 0.2, 0.3, 0.4 \}$.
     The vertical axis counts the number of perfect alignments in each bin, out of a total of 125 trials per bin.

### 2009-06-15 simulations/plots.r

@@ -293,7 +293,7 @@ bigbins = bigbins[-6]
 # get bin names
 #binnamefun <- function(x) cat (x - binwidth/2, " to ", x + binwidth/2 - delta)
 #binname = apply(bigbins, 1, binnamefun)
-binname = c("0 to 0.5", "0.5 to 1", "1 to 1.5", "1.5 to 2", "2 to 2.5");
+binname = c("0 to 0.4", "0.5 to 0.9", "1 to 1.4", "1.5 to 1.9", "2 to 2.4");
 
 # do binning
 exact = array()

### 2009-06-16 paper.tex
paper updates

@@
 (when the perfect alignment rate drops below $\sim 5/125$, 125 trials may be too few).
-We also note that the ML version of \stemloc\ is near-optimal, and slightly better than the posterior-decoding \stemloc-AMA.
+We also note that the ML version of \stemloc\ is near-optimal, despite the \stemloc\ pair-SCFG being
+slightly different from the TKFST pair-SCFG in parameterization and structure
+ (e.g. \stemloc's grammar does not allow insertion/deletion of entire substructures).
+The ML version of \stemloc\ is also observed to have a slightly higher perfect alignment rate than the posterior-decoding version (\stemloc-AMA).
 Finally, we note that the structure-blind models (TKF91 and Long Indel) perform consistently worse than the structure-aware methods,
@@
 
-A subtly different trend is apparent when we look at the accuracy of the returned alignment,
-measured not by the all-or-nothing metric of whether the alignment is {\em perfect},
-but rather by a metric that count the number of correct alignment columns.
-\figref{ama} shows the {\em Alignment Metric Accuracy} (AMA),
-a generalization of the {\em Sum-of Pairs Score} (SPS)
-which measures the proportion of residues which are correctly aligned, averaged over all pairs of sequences \cite{SchwartzMyersPachter2006}.
-Here, we see that \stemlocama---which attempts to find the alignment with maximum expected AMA---edges out both
-\indiegram\ and the ML version of \stemloc.
-The decision-theoretic scoring scheme of \stemlocama\ appears to offset the minor incorrectness of its model.
-
-Again, when looking at the AMA statistics, the most important factor is clearly the incorporation of some form of basepair structure
-into the model: \handel, whether using TKF91 or the Long Indel model, performs much worse than the SCFG-based methods.
-This is to be expected: whenever a basepair-aware method aligns one half of a basepair, it gets the other nucleotide
+A subtly different ranking emerges from consideration of the {\em alignment accuracy}.
+In \figref{ama}, we abandon the all-or-nothing metric of counting only perfect alignments,
+instead using a metric that shows what proportion of the alignment is correct.
+Specifically, we plot the {\em Alignment Metric Accuracy} (AMA) as a function of outgroup branch length.
+AMA measures the proportion of residues which are correctly aligned, averaged over all pairs of sequences \cite{SchwartzMyersPachter2006}.
+\figref{ama} reveals that \stemlocama\ (which attempts to find the alignment with the maximum expected AMA)
+edges out both \indiegram\ and the ML version of \stemloc\ (both of which attempt to find the alignment with the maximum likelihood).
+These results, compared to the subtly different story told by the perfect alignment rate,
+underscore the point that benchmark results for alignment methods can depend exquisitely on the choice of accuracy metric.
+The superiority of ML methods is only assured in terms of perfect alignment rate, and not necessarily other accuracy metrics.
+
+With the AMA statistic, as with the perfect alignment rate, the most important factor distinguishing the various models
+is clearly the incorporation of some form of basepair structure:
+structure-less \handel\  (regardless of linear {\em vs} affine gap penalty) performs much worse than the structure-aware SCFG methods.
+Intuitively, this is to be expected: whenever a basepair-aware method aligns one half of a basepair, it gets the other nucleotide
 correctly aligned for free.

### 2009-06-16 paper.tex
ranking of factors affecting accuracy

@@
 \paragraph{The Outside algorithm.}
-The Outside algorithm is primarily used an an intermediary for posterior decoding on the model.  It is analogous to the Backward algorithm for HMMs.
+The Outside algorithm is primarily used an an intermediary for calculating nucleotide-level posterior probabilities, e.g. for posterior decoding on the model.
+It is analogous to the Backward algorithm for HMMs.
 
@@
 
-With the AMA statistic, as with the perfect alignment rate, the most important factor distinguishing the various models
-is clearly the incorporation of some form of basepair structure:
+Taken together, these results show that the most important factor distinguishing the various models
+is the incorporation of some form of basepair structure:
 structure-less \handel\  (regardless of linear {\em vs} affine gap penalty) performs much worse than the structure-aware SCFG methods.
@@
 correctly aligned for free.
+
+The second most important, amongst the factors we have considered in this experiment, is use of the correct posterior-decoding algorithm for the objective function at hand
+(c.f. \stemloc\ {\em vs} \stemlocama).
+An obvious corollorary is the importance of choosing an appropriate objective function in the first place
+(c.f. perfect alignment rate {\em vs} AMA).
+
+Lastly, the fine details of the scoring scheme---including branch lengths, substitution scores, gap penalties and so forth---appear to
+be the least important of the factors we considered, yielding observable differences only when all other aspects of the inference procedure were more-or-less equal.
 

### 2009-06-16 paper.tex
last edits

@@
 
+The above-described TKFST SCFGs must be transformed slightly before they can be loaded into \indiegram.
+The grammars are presented in \indiegram\ format in \supptext{4}.
+
 A few other useful statistics for the TKFST model:
@@
 The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
-{\bf This model improves on the previous model (Long Indel) by introducing RNA secondary structure, though it no longer models phylogeny as rigorously as the Long Indel model} (it uses single-linkage clustering instead of a true phylogeny).
+Instead of modeling a true phylogeny by introducing unobserved ancestral sequences, it just does single-linkage clustering of the observed sequences.
 \item{\stemlocama:} the \stemloc\ program was used to align the three sequences in ``sequence annealing'' mode,
@@
 The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
-{\bf This program uses the previous model but instead of maximizing likelihood, it attempts to maximize an alignment accuracy metric.}
+This program uses the same underlying pair SCFG as \stemloc, but instead of maximizing likelihood, it attempts to maximize an alignment accuracy metric.
 \item{TKF91:} with the TKF91 model \cite{ThorneEtal91}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
@@
 The insertion, deletion and substitution rates for the TKF91 model were set equal to those of the loop submodel of TKFST.
-{\bf This may be understood as a naive reconstruction that completely ignores the stem sub-model of TKFST.}
+This may be understood as a naive sequence-only reconstruction that completely ignores basepair structure (i.e. the stem sub-model of TKFST).
 \item{Long Indel:} with a single-event trajectory approximation to the long indel model \cite{MiklosLunterHolmes2004}, the \handel\ package was used to align the three extant sequences and reconstruct the ancestor.
@@
 (\secname{A simple model of RNA structural evolution} has formulae for these quantities in terms of the TKFST rate parameters).
-{\bf This model improves on the previous model (TKF91) by introducing affine gap penalties.}
+This model improves on the previous model (TKF91) by introducing affine gap penalties.
 \end{description}
 
-We focused primarily on alignment accuracy, under the simplifying assumption that this will be a good indicator for
-(and perhaps the primary determinant of) the accuracy of ancestral reconstruction.
+We measured alignment accuracy, under the simplifying assumption that this correlates well with ancestral reconstruction accuracy.
 
 We first consider the {\em perfect alignment rate}; that is, the number of times each method gets the alignment exactly correct.
-Theory predicts that ML inference, using the correct model and parameters, should maximize the perfect alignment rate.
+Theory predicts that Maximum Likelihood inference, using the correct model and parameters, should be the algorithm that scores the most perfect estimates.
 Inspecting \figref{perfect}, we find this to be the case, except where the outgroup is very distant and the bins are probably undersampled
-(when the perfect alignment rate drops below $\sim 5/125$, 125 trials may be too few).
+(i.e. the departure from prediction that we observe for low-identity alignments is not statistically significant:
+when the perfect alignment rate drops below $\sim 5/125$, then 125 trials are probably too few).
 We also note that the ML version of \stemloc\ is near-optimal, despite the \stemloc\ pair-SCFG being
@@
 The ML version of \stemloc\ is also observed to have a slightly higher perfect alignment rate than the posterior-decoding version (\stemloc-AMA).
-Finally, we note that the structure-blind models (TKF91 and Long Indel) perform consistently worse than the structure-aware methods,
-with no decisive advantage shown by either TKF91 (linear gap penalty) or Long Indel (affine gap penalty).
+Finally, we note that the structure-blind models (TKF91 and Long Indel) perform consistently worse than the structure-aware methods;
+furthermore, both linear (TKF91) and affine (Long Indel) gap-penalties perform equally bad in this test
+(note that the TKFST model, from which the true alignments were simulated, does not allow long-indel events,
+which may partly explain why affine gap-penalties do not help in this benchmark).
 
@@
 
-Taken together, these results show that the most important factor distinguishing the various models
+Taken together, these results suggest that the most important factor distinguishing the various models we have examined
 is the incorporation of some form of basepair structure:
-structure-less \handel\  (regardless of linear {\em vs} affine gap penalty) performs much worse than the structure-aware SCFG methods.
+structure-blind \handel\  (regardless of linear {\em vs} affine gap penalty) performs much worse than the structure-aware SCFG methods.
 Intuitively, this is to be expected: whenever a basepair-aware method aligns one half of a basepair, it gets the other nucleotide
 correctly aligned for free.
-
-The second most important, amongst the factors we have considered in this experiment, is use of the correct posterior-decoding algorithm for the objective function at hand
+In benchmarks of RNA multiple alignment programs, structure-aware scoring schemes routinely outperform structure-blind scoring schemes \cite{GardnerEtAl2005,WilmEtAl2006}.
+Since we know that modeling structure is very important, it's not too surprising that it turns out to be the most important of the factors we considered.
+
+The second most important, amongst the factors we have considered in this experiment, is
+selection of the most appropriate objective function for the task at hand
+(c.f. perfect alignment rate {\em vs} AMA),
+followed by use of the correct posterior-decoding algorithm for the chosen objective function
 (c.f. \stemloc\ {\em vs} \stemlocama).
-An obvious corollorary is the importance of choosing an appropriate objective function in the first place
-(c.f. perfect alignment rate {\em vs} AMA).
+This is a subtle but important point: before deciding exactly what inference algorithm we're going to use to reconstruct ancestral sequences,
+we need to decide whether we want to maximize
+(a) the probability that our reconstructed sequence is 100\% correct,
+(b) the expected number of nucleotides that are correctly reconstructed,
+(c) the expected number of base-pairs that are correctly reconstructed,
+(d) the expected number of stems that are correctly reconstructed,
+(e) some other metric.
+Each of these metrics would require a slightly different inference algorithm.
 
@@
 be the least important of the factors we considered, yielding observable differences only when all other aspects of the inference procedure were more-or-less equal.
+While such details of the model may affect reconstruction quality,
+they appear to have very minor influence on alignment quality.
 
@@
   \centering
-%  \includegraphics [scale=0.4] {figs/parent.pdf}
-   \caption{
-     \textbf{A simple example of transducer composition to build an
-       evolutionary model of two extant sequences.}
-    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
-     A singlet transducer (the horizontal gray box) emits an ancestral sequence and structure
-     and two branch transducers (the gray boxes labeled $T_X$ and $T_Y$)
-     mutate it according to the specified evolutionary model.
-     Gray nodes correspond to observed data and white nodes unobserved data.
-   }
-   \figlabel{parent}
- \end{figure}
-
- \begin{figure}[!ht]
-   \centering
- %  \includegraphics [scale=0.4] {figs/threeway.pdf}
-   \caption{
-     \textbf{Transducer composition on a star phylogeny with three
-       (extant) leaf sequences.}
-     An ancestral sequence $W$ evolves into three descendant sequences $X$, $Y$ and $Z$.
-     A singlet transducer (the horizontal gray box) emits an ancestral sequence and structure
-     and three branch transducers (the gray boxes labeled $T_X$, $T_Y$ and $T_Z$)
-     mutate it according to the specified evolutionary model.
-     Gray nodes correspond to observed data and white nodes unobserved data.
-     If the branch transducers are time-reversible, then this star phylogeny with three leaves is
-     the neighborhood of any interior node in a (binary) phylogenetic tree, 
-     from which it follows that evaluating the likelihood function
-     on this star phylogeny is sufficient, in principle, for a sampling algorithm on an arbitrary phylogeny.
-   }
-\figlabel{threeway}
- \end{figure}
-
-\begin{figure}[!ht]
-  \centering
 %  \includegraphics [width=0.5\textwidth] {figs_plos/stree-with-structure.pdf}
@@
   \caption{
-    \textbf{Dependence of perfect alignment rate on outgroup branch
+    \textbf{Dependence of perfect alignment rate on alignment method and outgroup branch
       length.}
-    We investigated the perfect alignment rate
-    (the number of times that a given alignment program estimates the alignment 100\% correctly, with no errors)
-    as a function of outgroup branch length, simulating the
-    evolution of three structural RNAs under our model.  The two
-    sister species were equidistant from the ancestral sequence, with
-    a branch length of 1, corresponding to one expected
-    substitution per site in loop sequence.  We allowed the outgroup
-    branch length $t$ to vary between $[0, 2.5]$.  We selected 25 equally spaced values of $t$ in this range, spaced $0.1$ apart,
-    and simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
-    Since the perfect alignment rate is rather low, we further aggregated the $t$-values for this plot into bins of five;
-    thus, for example, the bin named ``$0 to 0.4$'' includes $t \in \{ 0, 0.1, 0.2, 0.3, 0.4 \}$.
-    The vertical axis counts the number of perfect alignments in each bin, out of a total of 125 trials per bin.
+    Perfect alignment is when a given alignment program estimates the alignment 100\% correctly, with no errors.
+    Simulating the evolution of three structural RNAs under the TKFST model,
+    we investigated the dependency of perfect alignment rate on outgroup branch length.
+    The simulation included two sister species at unit distance from the ancestral sequence, plus one ``outgroup''
+    whose branch length $t$ was varied between $[0, 2.5]$ by selecting 25 equally spaced values of $t$ in this range, spaced $0.1$ apart.
+    (A unit-length branch here corresponds to one expected substitution per site in loop sequence.)
+    We simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
+    Since the perfect alignment rate is rather low, we further aggregated the $t$-values into bins of five;
+    thus, for example, the bin named ``$0 to 0.4$'' includes $t \in \{ 0, 0.1, 0.2, 0.3, 0.4 \}$ and represents $25 \times 5 = 125$ trials in total.
+    The perfect alignment rate was measured for various statistical alignment inference procedures.
+    These procedures are described in the text, but may be summarized very briefly as
+    ML under the true model (``\indiegram'');
+    greedy approximate-ML progressive alignment by single-linkage clustering with pair SCFGs (``\stemloc'');
+    sequence annealing, a form of posterior decoding to maximize a sum-over-pairs accuracy metric, using pair SCFGs to get the posterior probabilities (``\stemlocama'');
+    statistical alignment using the TKF91 model, i.e. linear gap-penalties (``TKF91''); and
+    statistical alignment using a long-indel model, i.e. affine gap-penalties (``Long Indel'').
 }
@@
   \caption{
-    \textbf{Dependence of reconstruction accuracy on outgroup branch
+    \textbf{Dependence of alignment metric accuracy on alignment method and outgroup branch
       length.}
-    We investigated the dependence of structural reconstruction and
-    alignment accuracy on outgroup branch length by simulating the
-    evolution of three structural RNAs under our model.  The two
-    sister species were equidistant from the ancestral sequence, with
-    a branch length of 1, corresponding to one expected
-    substitution per site in loop sequence.  We allowed the outgroup
-    branch length $t$ to vary between $[0, 2.5]$.  We selected 25 equally spaced values of $t$ in this range, spaced $0.1$ apart,
-    and simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
+    We simulated the evolution of three structural RNAs under the TKFST model.
+    The simulation included two sister species at unit distance from the ancestral sequence, plus one ``outgroup''
+    whose branch length $t$ was varied between $[0, 2.5]$ by selecting 25 equally spaced values of $t$ in this range, spaced $0.1$ apart.
+    We then simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
     The Alignment Metric Accuracy (AMA) is, roughly, the proportion of residues that are correctly
     aligned, averaged over all pairs of sequences
-    (see \cite{SchwartzMyersPachter2006} for a precise definition; we used AMA with Gap Factor 1).
+    (see \cite{SchwartzMyersPachter2006} for a precise definition; we set the AMA Gap Factor to 1).
+    The AMA between the true alignment and the inferred alignment was measured for various statistical alignment inference procedures.
+    These procedures are described in the text, but may be summarized very briefly as
+    ML under the true model (``\indiegram'');
+    greedy approximate-ML progressive alignment by single-linkage clustering with pair SCFGs (``\stemloc'');
+    sequence annealing, a form of posterior decoding to maximize a sum-over-pairs accuracy metric, using pair SCFGs to get the posterior probabilities (``\stemlocama'');
+    statistical alignment using the TKF91 model, i.e. linear gap-penalties (``TKF91''); and
+    statistical alignment using a long-indel model, i.e. affine gap-penalties (``Long Indel'').
 }

### 2009-06-16 s3.tex
last edits

--- 2009-06-01 s3.tex
@@
 
+ \begin{figure}[!ht]
+   \centering
+  \includegraphics [scale=0.4] {figs/threeway.pdf}
+   \caption{
+     \textbf{Transducer composition on a star phylogeny with three
+       (extant) leaf sequences.}
+     An ancestral sequence $W$ evolves into three descendant sequences $X$, $Y$ and $Z$.
+     A singlet transducer (the horizontal gray box) emits an ancestral sequence and structure
+     and three branch transducers (the gray boxes labeled $T_X$, $T_Y$ and $T_Z$)
+     mutate it according to the specified evolutionary model.
+     Gray nodes correspond to observed data and white nodes unobserved data.
+     If the branch transducers are time-reversible, then this star phylogeny with three leaves is
+     the neighborhood of any interior node in a (binary) phylogenetic tree, 
+     from which it follows that evaluating the likelihood function
+     on this star phylogeny is sufficient, in principle, for a sampling algorithm on an arbitrary phylogeny.
+   }
+\figlabel{threeway}
+ \end{figure}
+
 \subsection{Automated grammar construction}
-We implemented our model construction algorithm on the three-taxon star phylogeny
+We implemented our model construction algorithm on the three-taxon star phylogeny (\figref{threeway}) 
 in a set of Perl scripts.

### 2009-06-16 s4.tex
last edits

@@
 \newpage
-\section{Software}
+\section{Inference grammars for the TKFST model}
+\seclabel{TKFST}
 
-All tools are available from \darturl\ as part of the \dart\ software package for sequence analysis.
+This section describes multi-sequence grammars that can be used to infer alignments, structures and ancestors under the
+TKF Structure Tree (TKFST) model.
 
-\subsection{Automated grammar construction}
-We implemented our model construction algorithm on the three-taxon star phylogeny
-in a set of Perl scripts.
-Given a singlet transducer modeling ancestral sequences and structures and
-a branch transducer modeling structural evolution,
-the scripts generate C++ code to create the corresponding jointly-normalized SCFG.
-All possible models of structural evolution which can be represented by a Pair SCFG
-are permitted as input to the scripts,
-allowing for flexible and automated model design.
+\input{tkfst.tex}
 
-Given files describing the singlet and branch transducers,
-including weights of all transitions which may be functions of evolutionary distance,
-the package ComposedTreeTransducer::FourWayComposedTT can automatically generate
-the state graph and transition matrix of a multi-sequence model of three extant sequences.
-It removes the useless windback $\Snull$ states described in the paper and
-introduces effective direct transitions caused by bifurcations with possibly-empty children.
-The package ComposedTreeTransducer::TripletSCFG transforms a multi-sequence
-model created by ComposedTreeTransducer::FourWayComposedTT into the 
-corresponding jointly-normalized three-sequence SCFG
-and generates C++ code to build the model.
+\newpage
+\section{The TKF Structure Tree model as a transducer} \seclabel{tkfst}
 
-Example singlet and branch transducers files are provided for a simple 
-Pair HMM model, a simple Pair SCFG model and the full TKF Structure Tree model.
+We can cast the TKF Structure Tree (TKFST) model as a transducer.
+This means we can automatically deduce rules like those shown for the TKFST model in \secref{TKFST}.
 
-\subsection{Reconstruction of ancestral structures}
-The program \indiegram\ can perform maximum-likelihood inference on the
-three-sequence SCFGs automatically generated by the ComposedTreeTransducer::TripletSCFG package.
-Complete or no structural information for the three extant sequences can be supplied
-as input.
+\subsection{The single-sequence TKFST model as a singlet transducer}
 
-\subsection{Simulation of RNA family evolution}
-The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
-Another script (\animateevolsayer) can be used to make animations of the evolving RNA structures, in combination with the RNAplot program in the \viennarna\ package \cite{HofackerEtAl94}.
+The states of the singlet transducer are shown in \tabref{tkfstsinglet}.
+Allowed transitions between these states are shown in \secref{TKFST}.
+
+\begin{table}[!ht]
+  \centering
+  \begin{tabular}{lrrrr}
+    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
+    $L$ & $\Sstart$ \\
+    $I_L$ & $\Sinsert$ & & $(x,\Tnull)$ & $p(x)$ \\
+    \\
+    $S$ & $\Sstart$ \\
+    $I_S$ & $\Sinsert$ & & $(x,y)$ & $p(x,y)$ \\
+    \\
+    $B$ & $\Sinsert$ & & $(L\,S)$ & 1 \\
+  \end{tabular}
+  \caption{
+    \tablabel{tkfstsinglet}
+    States of the singlet transducer of the TKF Structure Tree model \cite{Holmes2004}.
+    Singlet transducers can only have states of type $\Sstart$ or $\Sinsert$.
+    This is the indiegram-style transducer equivalent of the singlet TKFST SCFG in \secref{TKFST}.
+  }
+\end{table}
+
+\subsection{The two-sequence TKFST model as a branch transducer}
+
+The states of the branch transducer are shown in \tabref{tkfstbranch}.
+Allowed transitions between these states are shown in \secref{TKFST}.
+
+\begin{table}[!ht]
+  \centering
+  \begin{tabular}{lrrrr}
+    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
+    $L$ & $\Sstart$ \\
+    $I_L$ & $\Sinsert$ & & $(u,\Tnull)$ & $p(u)$ \\
+    $M_L$ & $\Smatch$ & $(x,\Tnull)$ & $(u,\Tnull)$ & $p(u|x)$ \\
+    $D_L$ & $\Smatch$ & $(x,\Tnull)$ & $(\Tgap,\Tnull)$ & 1 \\
+    $W_L$ & $\Swait$ \\
+    \\
+    $S$ & $\Sstart$ \\
+    $I_S$ & $\Sinsert$ & & $(u,v)$ & $p(u,v)$ \\
+    $M_S$ & $\Smatch$ & $(x,y)$ & $(u,v)$ & $p(u,v|x,y)$ \\
+    $D_S$ & $\Smatch$ & $(x,y)$ & $(\Tgap,\Tgap)$ & 1 \\
+    $W_S$ & $\Swait$ \\
+    \\
+    $B_i$ & $\Sinsert$ & & $(L_i\,S_i)$ & 1 \\
+    $B$ & $\Smatch$ & $(L\,S)$ & $(L\,S)$ & 1 \\
+    $B_p$ & $\Smatch$ & $(L\,S)$ & $(L\,\Send)$ & 1 \\
+    $B_e$ & $\Smatch$ & $(L\,\Send)$ & $(L\,\Send)$ & 1 \\
+  \end{tabular}
+  \caption{\tablabel{tkfstbranch}
+    States of the branch transducer of the TKF Structure Tree model \cite{Holmes2004}.
+    States which have the same names as states of the singlet transducer in \tabref{tkfstsinglet}
+    are the branch equivalent of the corresponding singlet states
+    (e.g. a $\Smatch$ state might be the branch equivalent of an $\Sinsert$ state).
+    States $L_i$ and $S_i$ are the $\Sstart$ states of a sub-model (not shown) identical
+    in structure to the singlet transducer.  They are used to insert a new stem-loop structure.
+    This is the indiegram-style transducer equivalent of the conditional pair TKFST SCFG in \secref{TKFST}.
+  }
+\end{table}
+
+\subsection{The multi-sequence TKFST model as a composite transducer}
+
+We can use the state graph construction algorithm described 
+in the paper and detailed in Text S1 to create a model of the simultaneous evolution
+of several sequences.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics [scale=0.4] {figs/parent.pdf}
+  \caption{\figlabel{parent}
+    A simple example of transducer composition to build a multi-sequence model of two extant sequences.
+    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
+    A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
+    and two branch transducers (the gray boxes labeled $\Delta T_X$ and $\Delta T_Y$)
+    mutate it according to the specified multi-sequence model.
+    Gray nodes correspond to observed data and white nodes unobserved data.
+  }
+\end{figure}
+
+Consider the simple model shown in \figref{parent}.
+The state of the multi-sequence model describing this model is a 3-vector $\bvec{a} = \left( a_1,\,a_2,\,a_3 \right)$,
+where $a_1$ is the state of the singlet transducer generating the ancestral sequence $W$
+and $a_2$ and $a_3$ are the states of the branch transducers evolving $W$ into
+extant sequences $X$ and $Y$.
+
+We can show some of the allowed transitions of this multi-sequence model.
+The state of the branch transducer associated with the active node $n$ is shown in bold.
+
+\paragraph{Stem creation.}
+
+A stem is created at the root node $W$ (corresponding to a bifurcation in the singlet transducer $a_1$)
+and survives in the sequence $X$ at node 2 but is deleted in the sequence $Y$ at node 3.
+
+\begin{align}
+  \nakedthreevec{1:}{2:}{3:} \quad
+  \threevec{L}{L}{\mathbf{L}} &
+  \to \threevec{\mathbf{L}}{W_L}{W_L}
+  \to \threevec{B}{B}{B_p}
+  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{S}{\mathbf{S}}{\Send}
+  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\mathbf{S}}{W_S}{\Send} \\
+  & \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\mathbf{S}}{W_S}{\Send}
+\end{align}
+
+\paragraph{Stem insertion.}
+
+A stem sequence is inserted in sequence $X$ at node 2.
+
+\begin{align}
+  \nakedthreevec{1:}{2:}{3:} \quad
+  \threevec{L}{L}{\mathbf{L}} \to \threevec{L}{\mathbf{L}}{W_L}
+  \to \threevec{L}{B_i}{B}
+  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\Send}{S}{\mathbf{S}}
+  \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\Send}{\mathbf{S}}{W_S}
+\end{align}
+
+\paragraph{Stem termination.}
+
+All stem sequences are ended by (possibly empty) loop sequences.
+
+\begin{align}
+  \nakedthreevec{1:}{2:}{3:} \quad
+  \threevec{\mathbf{S}}{W_S}{W_S}
+  \to \threevec{B_e}{B_e}{B_e}
+  \to \threevec{\Send}{\Send}{\Send} \, \threevec{L}{L}{\mathbf{L}}
+  \to \threevec{\Send}{\Send}{\Send} \, \threevec{\Send}{\Send}{\Send}
+\end{align}
+
+The functions $\alpha(t)$, $\beta(t)$ and $\gamma(t)$ are parametrized by the insertion and deletion rates of the Structure Tree model.
+They are defined for loop sequences as
+\begin{align}
+  \kappa_1 &= \lambda_1 / \mu_1 \\
+  \alpha_1 &= \exp \left( -\mu_1 t \right) \\
+  \beta_1 &= \frac{\lambda_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) } \\
+  \gamma_1 &= 1 - \frac{\mu_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\left( 1 - \exp (- \mu_1 t) \right) \left(\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) \right) }
+\end{align}
+and similarly for stem sequences \cite{Holmes2004}.
 

### 2009-06-16 paper.tex
last edits

@@
 in the tree, transforming one RNA structure into another.  In addition
-to current applications such as RNA genefinding, homology detection,
-alignment and secondary structure prediction, this should permit
-probabilistic reconstruction of RNA genes ancestral to many
-present-day sequences.  We describe algorithms, software
-implementations, and an elementary benchmark of the dependence of the
-accuracy of three-taxon RNA reconstruction on the outgroup branch
-length.  In the Discussion we consider how the three-taxon RNA
+to phylogenetic enhancement of current applications such as RNA genefinding, homology detection,
+alignment and secondary structure prediction, this should enable
+probabilistic phylogenetic reconstruction of RNA sequences that are ancestral to
+present-day genes.  We describe statistical inference algorithms, software
+implementations, and a simulation-based comparison
+of three-taxon maximum likelihood alignment to several other methods
+for aligning three sibling RNAs.
+In the Discussion we consider how the three-taxon RNA
 alignment-reconstruction-folding algorithm, which is currently very

### 2009-06-16 paper.tex
last edits

@@
 structure, makes clear the necessity for automated procedures for
-model construction.  Dataset S1 gives the state space of the ensemble
-model constructed by the search algorithm and Dataset S2 the reduced
+model construction.  \dataset{1} gives the state space of the ensemble
+model constructed by the search algorithm and \dataset{2} the reduced
 model after eliminating windback states; both are in Graphviz format for

### 2009-06-16 paper.tex
last edits

@@
 including ML alignment algorithms for simpler models that omitted various aspects of the full model
-and also a heuristic (sum-over-pairs) posterior-decoding alignment algorithm for one of the simpler models.
-
-Our results suggest that ML algorithms incorporating basepair structure are the most likely to estimate the alignment perfectly.
-However, the heuristic posterior-decoding algorithm scores higher on finer-grained alignment accuracy metrics.
-Since the posterior-decoding heuristic is substantially faster than exact phylogenetic ML,
-this further motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics
+and also a posterior-decoding alignment algorithm for one of the simpler models.
+
+In our tests, incorporation of basepair structure was the most important factor for alignment accuracy;
+appropriate use of posterior-decoding was the next most important factor;
+and fine details of the model were least important.
+Since posterior-decoding heuristics can be substantially faster than exact phylogenetic ML,
+this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics
 for paleogenetics and other alignment problems.

### 2009-06-16 paper.tex
last edits
post-submission change to abstract

@@
 % figures in figs_plos:
-% 1: parent.pdf
-% 2: threeway.pdf
-% 3: stree-with-structure.pdf
-% 4: stree-mutations-with-structures.pdf
-% 5: perfect.pdf
-% 6: ama.pdf
+% 1: stree-with-structure.pdf
+% 2: stree-mutations-with-structures.pdf
+% 3: perfect.pdf
+% 4: ama.pdf
 
@@
 and fine details of the model were least important.
-Since posterior-decoding heuristics can be substantially faster than exact phylogenetic ML,
-this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics
-for paleogenetics and other alignment problems.
+Since posterior-decoding heuristics can be substantially faster than exact phylogenetic inference,
+this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics where possible,
+with transducer composition reserved for exact ML and MCMC inference on phylogenies.
+We discuss possible ways to make these phylogenetic inferences tractable.
 

### 2009-06-16 paper.tex
post-submission change to abstract

@@
 this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics where possible,
-with transducer composition reserved for exact ML and MCMC inference on phylogenies.
+with transducer composition reserved for exact ML (or MCMC) inference on phylogenies.
 We discuss possible ways to make these phylogenetic inferences tractable.

### 2009-06-16 paper.tex
post-submission change to abstract

@@
 
-In our tests, incorporation of basepair structure was the most important factor for alignment accuracy;
+In our tests, incorporation of basepair structure was the most important factor for accurate inference of alignment;
 appropriate use of posterior-decoding was the next most important factor;

### 2009-06-16 paper.tex
Moved algorithms into main text

@@
 and fine details of the model were least important.
-Since posterior-decoding heuristics can be substantially faster than exact phylogenetic inference,
-this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics where possible,
-with transducer composition reserved for exact ML (or MCMC) inference on phylogenies.
-We discuss possible ways to make these phylogenetic inferences tractable.
+Posterior-decoding heuristics can be substantially faster than exact phylogenetic inference,
+so this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics where possible.
+We discuss the use of transducer composition for ML (or MCMC) inference on phylogenies,
+including possible ways to make these phylogenetic inferences tractable.
 
@@
 interest.  We can use this knowledge as a constraint to reduce the accessible volume, and hence the storage requirements, of the DP matrix \cite{Holmes2005}.
-Algorithms \ref{alg:inside}--\ref{alg:cyktraceback}, described below, give the Inside, CYK,
-Outside and CYK traceback algorithms for a three-taxon star phylogeny,
-constrained using the ``fold envelope'' concept.
+The Inside, Outside, and CYK+traceback algorithms for a three-taxon star phylogeny
+can be constrained using the ``fold envelope'' concept, which will now be described.
 
@@
 
-We use two orderings for sequences in the fold envelopes.  An inside-outside ordering is used for the iteration 
+An inside$\to$outside ordering is used for the iteration 
 in the Inside algorithm: Subsequences are ordered such that each successive subsequence 
@@
 
-The Outside algorithm uses an outside-inside ordering, which is the exact reverse of the inside-outside ordering described above.
+The Outside algorithm uses the exact reverse of the inside$\to$outside ordering described above; we call this the outside$\to$inside ordering.
 Subsequences in $\foldenv^{(X)}$ are sorted 
@@
 
+Using the fold envelope formalism, the main iteration over cells in the Inside and CYK matrices can be expressed as three nested loops:
+one for each sequence, traversing the fold envelope subsequences in inside$\to$outside order.
+Conversely, the main iteration of the Outside algorithm consists of three nested outside$\to$inside loops.
+
 \paragraph{The Inside algorithm.}
@@
 
+\begin{algorithm}[!ht]
+
+  \SetKwData{bifurcProb}{bifurcProb}
+  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
+  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
+  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
+  \SetLine
+  \KwIn{sequences $X, Y, Z$}
+
+  \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
+    \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
+      \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
+        \BlankLine
+        \ForEach{state $\bvec{a}$}{
+          \BlankLine
+          \bifurcProb $\leftarrow 0$\;
+          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
+            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
+              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
+                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
+                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
+              }
+            }
+          }
+          \BlankLine
+          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ $\leftarrow$ \calcTransEmitProb$\left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
+          store $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+        }
+        \BlankLine
+      }
+    }
+  }
+  \KwRet{$\alpha_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;
+
+  \caption{\alglabel{inside}
+    The constrained Inside algorithm for three sequences $X, Y, Z$.
+    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
+    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
+  }
+\end{algorithm}
+
 The transition and emission probability $\mathrm{calcTransEmitProb}(\bvec{a}; \cdot)$ can be calculated 
@@
 Pseudocode for the constrained calculation is given in \algref{insidetransemit}.
+
+\begin{algorithm}[!ht]
+  \SetKwData{emitProb}{emitProb}
+  \SetLine
+  \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Inside matrix $\alpha$}
+
+  \emitProb $\leftarrow 0$\;
+  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
+    \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \alpha_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+  }
+  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
+    \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
+    \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \alpha_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
+  }
+  \KwRet{\emitProb}\;
+  \caption{\alglabel{insidetransemit}
+    Subroutine $\mathrm{calcTransEmitProb}()$ for the Inside algorithm.
+    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
+  }
+\end{algorithm}
 
@@
 
+\begin{algorithm}[!ht]
+
+  \SetKwData{maxProb}{maxProb}
+  \SetKwData{bifurcProb}{bifurcProb}
+  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
+  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
+  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
+  \SetLine
+  \KwIn{sequences $X, Y, Z$}
+
+  \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
+    \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
+      \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
+        \BlankLine
+        \ForEach{state $\bvec{a}$}{
+          \BlankLine
+          \bifurcProb $\leftarrow 0$\;
+          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
+            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
+              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
+                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
+                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
+              }
+            }
+          }
+          \BlankLine
+          $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ \\
+          $\qquad \leftarrow \max \left( \calcTransEmitProb \left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right), \bifurcProb \right)$ \;
+          store $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+        }
+        \BlankLine
+      }
+    }
+  }
+  \KwRet{$\gamma_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;
+
+  \caption{\alglabel{cyk}
+    The constrained CYK algorithm for three sequences $X, Y, Z$.
+    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
+    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
+  }
+\end{algorithm}
+
 The resulting CYK algorithm is shown in \algref{cyk}.
@@
 
+\begin{algorithm}[!ht]
+  \SetKwData{emitProb}{emitProb}
+  \SetLine
+  \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate CYK matrix $\gamma$}
+
+  \emitProb $\leftarrow 0$\;
+  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
+    \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{b} \right)\, \gamma_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \right)$\;
+  }
+  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
+    \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
+    \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \gamma_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right) \right)$\;
+  }
+  \KwRet{\emitProb}\;
+  \caption{\alglabel{cyktransemit}
+    Subroutine $\mathrm{calcTransEmitProb}()$ for the CYK algorithm.
+    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
+  }
+\end{algorithm}
+
+\paragraph{The CYK traceback algorithm.}
+The CYK traceback algorithm, in combination with the CYK algorithm, is
+used to find the most-likely state path generating the extant sequences
+(in other words, the maximum-likelihood parse generating the observed data).
+It is analogous to the Viterbi traceback algorithm for HMMs.
+\algref{cyktraceback} gives the constrained CYK traceback algorithm.
+
+\begin{algorithm}[!ht]
+
+  \SetKw{Select}{select}
+  \SetKw{Pop}{pop}
+  \SetKw{Push}{push}
+  \SetKw{Goto}{goto}
+  \SetKw{Mainloop}{main loop}
+  \SetKwData{coordinateStack}{coordinateStack}
+  \SetKwFunction{goto}{goto}
+  \SetKwBlock{Begin}{begin main loop:}{end}
+  \SetLine
+  \KwIn{sequences $X, Y, Z$}
+
+  pushdown stack coordinateStack; \tcc*[f]{hold (state, subsequence triplet) pairs} \\
+  $\bvec{a} \leftarrow \Sstart$; \tcc*[f]{current ensemble state} \\
+  $n^{(X)} \leftarrow N^{(X)}$; \tcc*[f]{current $X$ subsequence; $N^{(X)}$ is the outermost subsequence} \\
+  $n^{(Y)} \leftarrow N^{(Y)}$; \tcc*[f]{current $Y$ subsequence; $N^{(Y)}$ is the outermost subsequence} \\
+  $n^{(Z)} \leftarrow N^{(Z)}$; \tcc*[f]{current $Z$ subsequence; $N^{(Z)}$ is the outermost subsequence} \\
+  clear \coordinateStack;\
+
+  \Begin{
+    output current state $\bvec{a}$ and subsequence triplet $\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+    \SetLine
+    \uIf(\tcc*[f]{end of a parse subtree}){$\bvec{a}$ is the $\Send$ state}{
+      \lIf(\tcc*[f]{end of the parse tree}){\coordinateStack is empty}{\Return}
+      \Pop $\left( \bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ from \coordinateStack\;
+      \Goto \Mainloop\;
+    }
+    \uElseIf(\tcc*[f]{bifurcation $\bvec{a} \to \bvec{c} \bvec{b}$}){$\bvec{a}$ is a bifurcation state}{
+      \Select $\left( n^{(X)}_L, n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$,
+      $\left( n^{(Y)}_L, n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$,
+      $\left( n^{(Z)}_L, n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$ \\
+      such that \\
+      $\gamma_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)} \right) = \gamma_{\bvec{a}} \left(n^{(X)}_L, n^{(Y)}_L, n^{(Z)}_L \right)  \gamma_{\bvec{a}} \left(n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ \;
+      \Push $\left( \bvec{c}, n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ onto \coordinateStack\;
+      $\bvec{a} \leftarrow \bvec{b}$\;
+      $n^{(X)} \leftarrow n^{(X)}_L$\;
+      $n^{(Y)} \leftarrow n^{(Y)}_L$\;
+      $n^{(Z)} \leftarrow n^{(Z)}_L$\;
+      \Goto \Mainloop\;
+    }
+    \Else(\tcc*[f]{$\Semit$ or $\Snull$ state}){
+      $m^{(X)} \leftarrow c_{in}\left(\bvec{b}; n^{(X)}\right)$ \;
+      $m^{(Y)} \leftarrow c_{in}\left(\bvec{b}; n^{(Y)}\right)$ \;
+      $m^{(Z)} \leftarrow c_{in}\left(\bvec{b}; n^{(Z)}\right)$ \;
+      \Select $\bvec{b} \in \left\{ b : \exists \,\bvec{a} \to \bvec{l}\bvec{b}\bvec{r} \right\}$ \\
+      such that \\
+      $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) = P(\bvec{a} \to \bvec{l}\bvec{b}\bvec{r}) \gamma_{\bvec{b}} \left( m^{(X)}, m^{(Y)}, m^{(Z)} \right)$\;
+      $\bvec{a} \leftarrow \bvec{b}$\;
+      $n^{(X)} \leftarrow m^{(X)}$\;
+      $n^{(Y)} \leftarrow m^{(Y)}$\;
+      $n^{(Z)} \leftarrow m^{(Z)}$\;
+      \Goto \Mainloop\;
+    }
+  }
+  \caption{\alglabel{cyktraceback}
+    The constrained CYK traceback algorithm for three sequences $X, Y, Z$.
+  }
+\end{algorithm}
+
 \paragraph{The Outside algorithm.}
@@
 used in the Outside algorithm are defined below.
+
+\begin{algorithm}[!ht]
+  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
+  \SetKwData{bifurcProb}{bifurcProb}
+  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
+  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
+  \SetLine
+  \KwIn{sequences $X, Y, Z$, Inside matrix $\alpha$}
+
+  \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
+    \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
+      \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
+        \BlankLine
+        \ForEach{state $\bvec{b}$}{
+          \BlankLine
+          \bifurcProb $\leftarrow 0$\;
+          \ForEach{$\left( n^{(X)}_O,n^{(X)}_L \right) \in b_{out,L}\left( n^{(X)} \right)$}{
+            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_L \right) \in b_{out,L}\left( n^{(Y)} \right)$}{
+              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_L \right) \in b_{out,L}\left( n^{(Z)} \right)$}{
+                \bifurcProb += \calcLBifurcProb($\bvec{b}; \cdot$)\;
+              }
+            }
+          }
+          \ForEach{$\left( n^{(X)}_O,n^{(X)}_R \right) \in b_{out,R}\left( n^{(X)} \right)$}{
+            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_R \right) \in b_{out,R}\left( n^{(Y)} \right)$}{
+              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_R \right) \in b_{out,R}\left( n^{(Z)} \right)$}{
+                \bifurcProb += \calcRBifurcProb($\bvec{b}; \cdot$)\;
+              }
+            }
+          }
+          \BlankLine
+          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \leftarrow$ \calcTransEmitProb$\left( \bvec{b}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
+          store $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+        }
+        \BlankLine
+      }
+    }
+  }
+  \caption{\alglabel{outside}
+    The constrained Outside algorithm for three sequences $X, Y, Z$.
+    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Outside fill order
+    with $\Semit$ states first, then $\Snull$ states in topological order.
+  }
+\end{algorithm}
 
@@
 
-\paragraph{The CYK traceback algorithm.}
-The CYK traceback algorithm, in combination with the CYK algorithm, is
-used to find the most-likely state path generating the extant sequences
-(in other words, the maximum-likelihood parse generating the observed data).
-It is analogous to the Viterbi traceback algorithm for HMMs.
-\algref{cyktraceback} gives the constrained CYK traceback algorithm.
+\begin{algorithm}[!ht]
+  \SetKwData{emitProb}{emitProb}
+  \SetLine
+  \KwIn{state $\bvec{b}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Outside matrix $\beta$}
+
+  \emitProb $\leftarrow 0$\;
+  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{b}$} {
+    \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \beta_{\bvec{a}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+  }
+  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
+    \lIf{$c_{out}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{out}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{out}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
+    \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \beta_{\bvec{a}}\left( c_{out}\left(\bvec{b}; n^{(X)}\right), c_{out}\left(\bvec{b}; n^{(Y)}\right), c_{out}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
+  }
+  \KwRet{\emitProb}\;
+  \caption{\alglabel{outsidetransemit}
+    Subroutine $\mathrm{calcTransEmitProb}()$ for the Outside algorithm.
+    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
+  }
+\end{algorithm}
 
@@
 
-\clearpage
-\section*{Algorithms}
-
-\begin{algorithm}[!ht]
-
-  \SetKwData{bifurcProb}{bifurcProb}
-  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
-  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
-  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
-  \SetLine
-  \KwIn{sequences $X, Y, Z$}
-
-  \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
-    \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
-      \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
-        \BlankLine
-        \ForEach{state $\bvec{a}$}{
-          \BlankLine
-          \bifurcProb $\leftarrow 0$\;
-          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
-            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
-              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
-                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
-                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
-              }
-            }
-          }
-          \BlankLine
-          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ $\leftarrow$ \calcTransEmitProb$\left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
-          store $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-        }
-        \BlankLine
-      }
-    }
-  }
-  \KwRet{$\alpha_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;
-
-  \caption{\alglabel{inside}
-    The constrained Inside algorithm for three sequences $X, Y, Z$.
-    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
-    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
-  }
-\end{algorithm}
-
-\begin{algorithm}[!ht]
-  \SetKwData{emitProb}{emitProb}
-  \SetLine
-  \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Inside matrix $\alpha$}
-
-  \emitProb $\leftarrow 0$\;
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
-    \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \alpha_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-  }
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
-    \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
-    \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \alpha_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
-  }
-  \KwRet{\emitProb}\;
-  \caption{\alglabel{insidetransemit}
-    Subroutine $\mathrm{calcTransEmitProb}()$ for the Inside algorithm.
-    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
-  }
-\end{algorithm}
-
-\begin{algorithm}[!ht]
-
-  \SetKwData{maxProb}{maxProb}
-  \SetKwData{bifurcProb}{bifurcProb}
-  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
-  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
-  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
-  \SetLine
-  \KwIn{sequences $X, Y, Z$}
-
-  \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
-    \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
-      \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
-        \BlankLine
-        \ForEach{state $\bvec{a}$}{
-          \BlankLine
-          \bifurcProb $\leftarrow 0$\;
-          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
-            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
-              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
-                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
-                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
-              }
-            }
-          }
-          \BlankLine
-          $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ \\
-          $\qquad \leftarrow \max \left( \calcTransEmitProb \left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right), \bifurcProb \right)$ \;
-          store $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-        }
-        \BlankLine
-      }
-    }
-  }
-  \KwRet{$\gamma_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;
-
-  \caption{\alglabel{cyk}
-    The constrained CYK algorithm for three sequences $X, Y, Z$.
-    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
-    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
-  }
-\end{algorithm}
-
-\begin{algorithm}[!ht]
-  \SetKwData{emitProb}{emitProb}
-  \SetLine
-  \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate CYK matrix $\gamma$}
-
-  \emitProb $\leftarrow 0$\;
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
-    \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{b} \right)\, \gamma_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \right)$\;
-  }
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
-    \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
-    \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \gamma_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right) \right)$\;
-  }
-  \KwRet{\emitProb}\;
-  \caption{\alglabel{cyktransemit}
-    Subroutine $\mathrm{calcTransEmitProb}()$ for the CYK algorithm.
-    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
-  }
-\end{algorithm}
-
-\begin{algorithm}[!ht]
-  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
-  \SetKwData{bifurcProb}{bifurcProb}
-  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
-  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
-  \SetLine
-  \KwIn{sequences $X, Y, Z$, Inside matrix $\alpha$}
-
-  \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
-    \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
-      \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
-        \BlankLine
-        \ForEach{state $\bvec{b}$}{
-          \BlankLine
-          \bifurcProb $\leftarrow 0$\;
-          \ForEach{$\left( n^{(X)}_O,n^{(X)}_L \right) \in b_{out,L}\left( n^{(X)} \right)$}{
-            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_L \right) \in b_{out,L}\left( n^{(Y)} \right)$}{
-              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_L \right) \in b_{out,L}\left( n^{(Z)} \right)$}{
-                \bifurcProb += \calcLBifurcProb($\bvec{b}; \cdot$)\;
-              }
-            }
-          }
-          \ForEach{$\left( n^{(X)}_O,n^{(X)}_R \right) \in b_{out,R}\left( n^{(X)} \right)$}{
-            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_R \right) \in b_{out,R}\left( n^{(Y)} \right)$}{
-              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_R \right) \in b_{out,R}\left( n^{(Z)} \right)$}{
-                \bifurcProb += \calcRBifurcProb($\bvec{b}; \cdot$)\;
-              }
-            }
-          }
-          \BlankLine
-          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \leftarrow$ \calcTransEmitProb$\left( \bvec{b}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
-          store $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-        }
-        \BlankLine
-      }
-    }
-  }
-  \caption{\alglabel{outside}
-    The constrained Outside algorithm for three sequences $X, Y, Z$.
-    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Outside fill order
-    with $\Semit$ states first, then $\Snull$ states in topological order.
-  }
-\end{algorithm}
-
-\begin{algorithm}[!ht]
-  \SetKwData{emitProb}{emitProb}
-  \SetLine
-  \KwIn{state $\bvec{b}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Outside matrix $\beta$}
-
-  \emitProb $\leftarrow 0$\;
-  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{b}$} {
-    \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \beta_{\bvec{a}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-  }
-  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
-    \lIf{$c_{out}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{out}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{out}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
-    \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \beta_{\bvec{a}}\left( c_{out}\left(\bvec{b}; n^{(X)}\right), c_{out}\left(\bvec{b}; n^{(Y)}\right), c_{out}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
-  }
-  \KwRet{\emitProb}\;
-  \caption{\alglabel{outsidetransemit}
-    Subroutine $\mathrm{calcTransEmitProb}()$ for the Outside algorithm.
-    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
-  }
-\end{algorithm}
-
-\begin{algorithm}[!ht]
-
-  \SetKw{Select}{select}
-  \SetKw{Pop}{pop}
-  \SetKw{Push}{push}
-  \SetKw{Goto}{goto}
-  \SetKw{Mainloop}{main loop}
-  \SetKwData{coordinateStack}{coordinateStack}
-  \SetKwFunction{goto}{goto}
-  \SetKwBlock{Begin}{begin main loop:}{end}
-  \SetLine
-  \KwIn{sequences $X, Y, Z$}
-
-  pushdown stack coordinateStack; \tcc*[f]{hold (state, subsequence triplet) pairs} \\
-  $\bvec{a} \leftarrow \Sstart$; \tcc*[f]{current ensemble state} \\
-  $n^{(X)} \leftarrow N^{(X)}$; \tcc*[f]{current $X$ subsequence; $N^{(X)}$ is the outermost subsequence} \\
-  $n^{(Y)} \leftarrow N^{(Y)}$; \tcc*[f]{current $Y$ subsequence; $N^{(Y)}$ is the outermost subsequence} \\
-  $n^{(Z)} \leftarrow N^{(Z)}$; \tcc*[f]{current $Z$ subsequence; $N^{(Z)}$ is the outermost subsequence} \\
-  clear \coordinateStack;\
-
-  \Begin{
-    output current state $\bvec{a}$ and subsequence triplet $\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-    \SetLine
-    \uIf(\tcc*[f]{end of a parse subtree}){$\bvec{a}$ is the $\Send$ state}{
-      \lIf(\tcc*[f]{end of the parse tree}){\coordinateStack is empty}{\Return}
-      \Pop $\left( \bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ from \coordinateStack\;
-      \Goto \Mainloop\;
-    }
-    \uElseIf(\tcc*[f]{bifurcation $\bvec{a} \to \bvec{c} \bvec{b}$}){$\bvec{a}$ is a bifurcation state}{
-      \Select $\left( n^{(X)}_L, n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$,
-      $\left( n^{(Y)}_L, n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$,
-      $\left( n^{(Z)}_L, n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$ \\
-      such that \\
-      $\gamma_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)} \right) = \gamma_{\bvec{a}} \left(n^{(X)}_L, n^{(Y)}_L, n^{(Z)}_L \right)  \gamma_{\bvec{a}} \left(n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ \;
-      \Push $\left( \bvec{c}, n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ onto \coordinateStack\;
-      $\bvec{a} \leftarrow \bvec{b}$\;
-      $n^{(X)} \leftarrow n^{(X)}_L$\;
-      $n^{(Y)} \leftarrow n^{(Y)}_L$\;
-      $n^{(Z)} \leftarrow n^{(Z)}_L$\;
-      \Goto \Mainloop\;
-    }
-    \Else(\tcc*[f]{$\Semit$ or $\Snull$ state}){
-      $m^{(X)} \leftarrow c_{in}\left(\bvec{b}; n^{(X)}\right)$ \;
-      $m^{(Y)} \leftarrow c_{in}\left(\bvec{b}; n^{(Y)}\right)$ \;
-      $m^{(Z)} \leftarrow c_{in}\left(\bvec{b}; n^{(Z)}\right)$ \;
-      \Select $\bvec{b} \in \left\{ b : \exists \,\bvec{a} \to \bvec{l}\bvec{b}\bvec{r} \right\}$ \\
-      such that \\
-      $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) = P(\bvec{a} \to \bvec{l}\bvec{b}\bvec{r}) \gamma_{\bvec{b}} \left( m^{(X)}, m^{(Y)}, m^{(Z)} \right)$\;
-      $\bvec{a} \leftarrow \bvec{b}$\;
-      $n^{(X)} \leftarrow m^{(X)}$\;
-      $n^{(Y)} \leftarrow m^{(Y)}$\;
-      $n^{(Z)} \leftarrow m^{(Z)}$\;
-      \Goto \Mainloop\;
-    }
-  }
-  \caption{\alglabel{cyktraceback}
-    The constrained CYK traceback algorithm for three sequences $X, Y, Z$.
-  }
-\end{algorithm}
-
 \end{document}

### 2009-06-16 paper.tex
Moved algorithms into main text

@@
 so this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics where possible.
-We discuss the use of transducer composition for ML (or MCMC) inference on phylogenies,
-including possible ways to make these phylogenetic inferences tractable.
+For more exact probabilistic inference, we discuss the use of transducer composition for ML (or MCMC) inference on phylogenies,
+including possible ways to make the core operations tractable.
 

### 2009-07-02 paper.tex
fixed some typos

@@
 we place a branch transducer on each branch of the phylogeny.
-Figures \ref{fig:parent} and \ref{fig:threeway} show simple examples
-of building a model to represent such an evolution.
+% Figures \ref{fig:parent} and \ref{fig:threeway} show simple examples
+% of building a model to represent such an evolution.
 
@@
 assumed to be the (unobserved) ancestral sequence and the sequences
-$X$, $Y$, and $Z$ the (observed) extant sequences (as in Figures
-\ref{fig:parent} and \ref{fig:threeway}).
+$X$, $Y$, and $Z$ the (observed) extant sequences.
+% (as in Figures \ref{fig:parent} and \ref{fig:threeway}).
 
@@
 We first consider the {\em perfect alignment rate}; that is, the number of times each method gets the alignment exactly correct.
-Theory predicts that Maximum Likelihood inference, using the correct model and parameters, should be the algorithm that scores the most perfect estimates.
-Inspecting \figref{perfect}, we find this to be the case, except where the outgroup is very distant and the bins are probably undersampled
-(i.e. the departure from prediction that we observe for low-identity alignments is not statistically significant:
-when the perfect alignment rate drops below $\sim 5/125$, then 125 trials are probably too few).
+Theory predicts that Maximum Likelihood inference, using the correct model and parameters, should be asymptotically optimal (if one only counts perfect guesses).
+Inspecting \figref{perfect}, we find this to be almost the case; the exception is when the outgroup is very distant and the bins may be undersampled
+(the departure from prediction that we observe for low-identity alignments is not statistically significant:
+when the optimal success rate drops below $\sim 5/125$, then 125 trials are probably insufficient to compare two near-optimal methods).
 We also note that the ML version of \stemloc\ is near-optimal, despite the \stemloc\ pair-SCFG being

### 2009-07-02 simulations/debug-0.1-13.pl

@@ -1,6 +1,6 @@
 #!/usr/bin/perl
 $tkfst = $ENV{'DARTDIR'} . "/src/indiegram/tkfst.cc";
-$parse = "alignments/0.1/13.reference.parse";
+for $parse ("alignments/0.1/13.reference.parse", "inferred/0.1/13.indiegram.parse") {
 open IN, "<$parse" or die;
 open OUT, ">$parse.inits";
 print OUT "// $parse.inits: generated by $0 from $parse and $tkfst\n\n";
@@ -13,3 +13,4 @@ print OUT "// $c\n",`$c`,"\n";
 }
 close IN;
 close OUT or die;
+}

### 2009-07-13 paper.tex
more

@@
 %% PLEASE INCLUDE ALL MACROS BELOW
-\input{defs.tex}
+
+\usepackage{setspace}
+\doublespacing
+
+\input{parsetree.sty}
+
+\newcommand\titlestring{Evolutionary triplet models of structured RNA}
+\newcommand\authorstring{
+Robert K. Bradley$^{1}$, 
+Ian Holmes$^{1,2,\ast}$
+\\
+\textbf{1} Biophysics Graduate Group, University of California, Berkeley, CA, USA
+\\
+\textbf{2} Department of Bioengineering, University of California, Berkeley, CA, USA
+\\
+$\ast$ E-mail: indiegram@postbox.biowiki.org
+}
+
+\usepackage{array}
+
+\usepackage[boxed]{algorithm2e}
+\usepackage{mathrsfs}
+
+% Labels & references for sections, figures and tables
+% Commented out \secref definition & replaced \seclabel with a dummy, because PLoS doesn't like numbered section references -- IH, 11/11/08
+% (This edit appears to have been reversed as of 5/29/09 - presumably so that the supplementary information compiles OK using these same definitions - IH, 5/29/09)
+\newcommand{\secref}[1]{Section~\ref{sec:#1}}
+\newcommand{\seclabel}[1]{\label{sec:#1}}
+
+\newcommand{\secname}[1]{``#1''}  % PLoS-style section names
+
+% "Text S1", "Text S2", etc.
+\newcommand{\supptext}[1]{Text S#1}
+
+% "Dataset S1", "Dataset S2", etc.
+\newcommand{\dataset}[1]{Dataset S#1}
+
+\newcommand{\appref}[1]{Appendix~\ref{app:#1}}
+\newcommand{\applabel}[1]{\label{app:#1}}
+
+\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
+\newcommand{\figlabel}[1]{\label{fig:#1}}
+
+\newcommand{\tabnum}[1]{\ref{tab:#1}}
+\newcommand{\tabref}[1]{Table~\tabnum{#1}}
+\newcommand{\tablabel}[1]{\label{tab:#1}}
+
+\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
+\newcommand{\alglabel}[1]{\label{alg:#1}}
+
+\newcommand{\eqnref}[1]{Equation~\ref{eqn:#1}}
+\newcommand{\eqnlabel}[1]{\label{eqn:#1}}
+
+% Frequently used proper nouns that should have consistent typeface, capitalization, etc.
+\newcommand{\pfold}{PFOLD}
+\newcommand\darturl{{\tt http://biowiki.org/dart}}
+\newcommand\indiegramurl{{\tt http://biowiki.org/IndieGram}}
+\newcommand{\dart}{DART}
+
+\newcommand{\bralibaseII}{BRalibaseII}
+\newcommand{\balibase}{BAliBASE}
+
+\newcommand{\stemloc}{Stemloc}
+\newcommand{\stemlocama}{Stemloc-AMA}
+\newcommand{\xrate}{XRate}
+\newcommand{\evoldoer}{Evoldoer}
+\newcommand{\evolsayer}{{\tt evolsayer.pl}}
+\newcommand{\animateevolsayer}{{\tt animate-evolsayer.pl}}
+\newcommand{\indiegram}{Indiegram}
+\newcommand{\handel}{Handel}
+
+\newcommand{\contrafold}{CONTRAFOLD}
+\newcommand{\consan}{CONSAN}
+\newcommand{\clustalw}{ClustalW}
+\newcommand{\foldalign}{Foldalign}
+\newcommand{\foldalignm}{FoldalignM}
+\newcommand{\mastr}{MASTR}
+\newcommand{\rnasampler}{RNASampler}
+\newcommand{\murlet}{Murlet}
+\newcommand{\ortheus}{Ortheus}
+\newcommand{\rnaalifold}{RNAAlifold}
+\newcommand{\viennarna}{ViennaRNA}
+
+\newcommand{\RFAM}{RFAM}
+
+\newcommand{\denovo}{{\em de novo}}
+
+\newcommand{\nth}{\mathrm{n}^{\mathrm{th}}}
+\newcommand{\mth}{\mathrm{m}^{\mathrm{th}}}
+
+\newcommand{\type}[0]{\operatorname{type}}
+\newcommand{\emit}[0]{\operatorname{emit}}
+\newcommand{\absorb}[0]{\operatorname{absorb}}
+\newcommand{\argmax}[0]{\operatorname{argmax}}
+\newcommand{\argmin}[0]{\operatorname{argmin}}
+\newcommand{\parent}[0]{\operatorname{parent}}
+
+\newcommand{\weight}[0]{\operatorname{Weight}}
+\newcommand{\transweight}[3]{t\left(#2,#3|\theta^{(#1)}\right)}  % {m}{from}{to}
+\newcommand{\emissionweight}[3]{e_{#2}\left(#3|\theta^{(#1)}\right)}  % {m}{emitstate}{emission}
+\newcommand{\totalemissionweight}[2]{\mathbb{e}_{#1}(#2)}
+
+%\newcommand{\cyk}[7]{\gamma_{#1}(#2,#3,#4,#5,#6,#7)}
+\newcommand{\cyk}[7]{\gamma_{#1}\left(#2,#3,#4\right)}
+%\newcommand{\cyk}[7]{\gamma_{#1}(#2,#3,#4)}
+%\newcommand{\inside}[7]{\alpha_{#1}(#2,#3,#4,#5,#6,#7)}
+\newcommand{\inside}[7]{\alpha_{#1}\left(#2,#3,#4\right)}
+%\newcommand{\outside}[7]{\beta_{#1}(#2,#3,#4,#5,#6,#7)}
+\newcommand{\outside}[7]{\beta_{#1}\left(#2,#3\right,#4\right)}
+
+\newcommand{\Tnull}{\mathrm{null}}
+\newcommand{\Tgap}{\mathrm{GAP}}
+
+\newcommand{\statetype}[1]{\mathtt{#1}}
+\newcommand{\state}[3]{{}^{#2}\mathtt{#1}^{#3}}
+% Note that this will only work for left-bifurcations!
+%\newcommand{\bifurc}[3]{{\phantom{]}}^{#2}\mathtt{B[#3\,#1]}}}
+\newcommand{\bifurc}[3]{\mathtt{#1[#2\,#3]}}
+
+% state types for branch transducers
+\newcommand{\Sstart}[0]{\statetype{Start}}
+\newcommand{\Smatch}[0]{\statetype{Match}}
+\newcommand{\Sinsert}[0]{\statetype{Insert}}
+\newcommand{\Swait}[0]{\statetype{Wait}}
+\newcommand{\Send}[0]{\statetype{End}}
+
+% state types for evolutionary model
+\newcommand{\Snull}[0]{\statetype{Null}}
+\newcommand{\Sbifurc}[0]{\statetype{Bifurcation}}
+\newcommand{\Semit}[0]{\statetype{Emit}}
+
+\newcommand{\statevec}[1]{\left(\begin{array}{c}#1_1\\\vdots\\#1_N\end{array}\right)}
+\newcommand{\fourvec}[4]{\left(\begin{array}{c}#1\\#2\\#3\\#4\end{array}\right)}
+\newcommand{\nakedfourvec}[4]{\begin{array}{c}#1\\#2\\#3\\#4\end{array}}
+\newcommand{\threevec}[3]{\left(\begin{array}{c}#1\\#2\\#3\end{array}\right)}
+\newcommand{\nakedthreevec}[3]{\begin{array}{c}#1\\#2\\#3\end{array}}
+\newcommand{\bvec}[1]{\boldsymbol{#1}}
+\newcommand{\activenode}[1]{n(\bvec{#1})}
+
+\newcommand{\foldenv}{\mathscr{F}}
+
+\newcommand{\mDn}[0]{m \vartriangleright n}
+\newcommand{\mNDn}[0]{m \ntriangleright n}
+\newcommand{\mDEn}[0]{m \trianglerighteq n}
+\newcommand{\mNDEn}[0]{m \ntrianglerighteq n}
+
+\newcommand{\beqn}{\begin{equation}}
+\newcommand{\eeqn}{\end{equation}}
+
+\newcommand\gnons{{\cal N}}
+\newcommand\gnulls{{\bf N}}
+\newcommand\gbifs{{\bf B}}
+\newcommand\gemits{{\bf E}}
+\newcommand\gterms{\Omega}
+\newcommand\grules{{\bf R}}
+\newcommand\grhs{( \gnons \cup \Omega )^\ast}
+\newcommand\gtrees{{\cal T}}
+
+\newcommand\leftside[1]{{\cal L}(#1)}
+\newcommand\rightside[1]{{\cal R}(#1)}
+\newcommand\allrules[1]{{\cal A}(#1)}
+
+\newcommand\revisionnotes
+{\noindent \hrulefill{} \\ {\bf Revision Notes:}
+\begin{enumerate}
+\item These comments are not a complete summary of the review, just a partial strategy for addressing some of the points.
+The review is the more authoritative guide...
+He is giving us the benefit of the doubt but acceptance is still very much in the balance, I think, so we need to stick at it,
+and take it that extra step beyond minimal compliance with his suggestions.
+%
+\item Obviously we have to do simulation tests and hopefully EvolSayer will be of some use there.
+One thing to be careful of is that there are certain regimes of rate/branch-length parameter space where the ``trivial'' outcomes identified by the reviewer
+(ancestral sequence is always identical to one of the descendants) are the norm.
+Oscar and I have been looking at this quite a bit recently (see also 2003 paper by Mossel).
+The regime we are probably most interested in is where the branch lengths are of comparable length
+and so the two slightly-longer branches can ``out-vote'' the shorter branch under certain conditions.
+This should probably be commented on in the paper: the reviewer says he would be happy with ``a small, noncomprehensive set of anecdotes''
+but if we can rationalize why some parameter settings are more interesting than others, then that's all to the better.
+%
+\item I would actually be very wary of taking that ``small noncomprehensive set of anecdotes'' at face value.
+This is one area where I sense we're being given enough rope to hang ourselves... an unsatisfactory response to the simulation issue will be unequivocal grounds for rejection.
+Obviously there are limits to the number of simulations we can run, but with EvolSayer's output statistics (allowing us to filter out ``interesting'' cases), we can automate it somewhat.
+%
+\item I wrote out the TKF structure tree for the case $a \to b \to c$,
+which can be used to estimate the intermediate sequence $b$
+(e.g. for Redelings-Suchard).
+I think this could be useful in explaining tree transducer composition more concretely, and why is important:
+by inspecting these tables, the reader can see that the structure of this grammar is quite complicated,
+and should then be prepared for the idea that the ``composition algorithm'' is simply generating this table
+(or other tables for other phylogenies).
+The latex for this is in the file ``tkfst.tex'', and is currently include'd at the start of the supplementary information (it should go in the main paper though, I think).
+%
+\item Can we dump out the grammar for the three-branch TKFST and include it in the Supplement?
+(Or the main paper?) I realize it's huge, but this very hugeness does motivate the transducer composition algorithm,
+in the same way that the two-branch model (which is now included in a more human-readable form) does.
+%
+\item If we are moving the description of loopy DP to the main paper, I think it needs to be more precise.
+Currently it just says ``we repeatedly iterate over strongly-connected components of the state sub graph'',
+which falls short of the level of precision/detail with which the other DP algorithms are presented.
+The description should encompass alternate strategies (i.e. orders of visiting the states) that you have considered:
+randomized, Tarjan's algorithm, etc.
+The reader should be left understanding the pros-and-cons of each,
+ e.g. worst-case performance or difficulty of implementation.
+No need to agonize about this, but be helpful to potential future developers.
+  (Example reader: a grad student in our group who might try to implement loopy DP themselves.)
+%
+\item The descriptions of the DP algorithms do not currently include any traceback algorithms (CYK or Inside);
+therefore, we haven't technically described how to do ancestral reconstruction!
+Obviously these algorithms are almost trivial, but they're quite important.
+Implementing stochastic loopy-Inside traceback in tripletscfgdp.\{h,cc\} is a key step in transforming it into a useful tool
+(for example, with this extension, Indiegram would be pretty much ready-to-go for Redelings-Suchard sampling).
+The deterministic CYK traceback algorithm is {\em even more} important to present here, because implementing this algorithm is currently our main result!
+The paper should describe {\em both} these traceback algorithms.
+This shouldn't be that hard to do in a comprehensible, rigorous style
+ (loopy traceback is much simpler than the algorithms for restoring eliminated states to traceback paths, which are painfully described in my 2003 paper).
+%
+\item All algorithms should be described in full, rather than giving recipes for deriving some algorithms from others.
+In particular, the CYK algorithm is currently summarized by statements along the lines of
+``replace sums over paths through the model (or equivalently, parses) with the $\max()$ operation (e.g., in equation X, sum-over-complicated-expression will be replaced with max-over-complicated-expression)''.
+It shouldn't be too hard to write a latex command/macro that can be used (with ``$\sum$'' or ``$\max$'' as an argument) to generate complete versions of {\em both} algorithms.
+%
+\item The paper the reviewer cites as an example of the simplicity of missing data estimation (Rivas and Eddy (2008) PLoS CB 4:e1000172) is an IID-columns model for a multiple alignment.
+When he says that this issue ``may simply go away if you view the problem as a missing data problem'',
+that is precisely what we are talking about when we discuss null cycle elimination.
+I've tried to explain in ``tkfst.tex'' why this is harder than it superficially appears:
+\begin{enumerate}
+\item eliminating missing ``empty'' columns from an IID-columns model (such as in Rivas and Eddy, 2008) is easy.
+\item eliminating missing emissions from an HMM is slightly harder (c.f. ``wing retraction'' algorithm in HMMer source code)
+but general algorithms do exist.
+\item eliminating null states from an SCFG is much harder, due to null bifurcations.
+\end{enumerate}
+%
+\item The reviewer asks for early discussion of constraints in the context of the ``rate-limiting steps of paleogenetics''.
+We should take this opportunity to clarify that our approach is (1) formulate the model,  (2) apply appropriate constraints.
+Our emphasis here is on (1).
+Cite the literature: QRNA represents an alignment constraint on a pair SCFG, Holmes-Rubin 2002 uses fold constraints on pair SCFGs,
+stemloc uses fold+alignment constraints, Ortheus uses an alignment constraint, etc, but none of these things can be done
+(in a probabilistic modeling framework at least) until you can formulate the model.
+For this paper, we chose to condition on structures essentially because they are easy to constrain
+(fold envelopes can be applied independently to each sequence).
+But we could equally easily have conditioned on alignments. Constraints are easy to tweak, once you have a model.
+%
+\item The reviewer also comments on the phylogeny being a given.
+We can note that even though we constrain the phylogeny, more realistic phylogenetic applications are possible.
+Given a structural phylo-alignment (tree+alignment+structure), we can calculate the likelihood; therefore, given two such structural phylo-alignments,
+we can say which one is more probable (describe their relative posterior probabilities, etc.).
+From this it is possible to make a crude MCMC sampler, just by proposing/accepting/rejecting alignment/tree/structure changes.
+An {\em efficient} MCMC sampler needs to be able to make local topology changes and to resample the multiple alignment around a topology change;
+we can do this too, in principle, but it's beyond this paper.
+%
+\item As I understand it, pruning \& peeling are not exactly equivalent as the reviewer claims, {\em except} in the special case of a reversible model.
+Specifically, pruning is a {\em restricted case} of peeling, giving you the posterior probabilities at the root node, but not any other node.
+Of course, if you have a reversible model, you can re-root the tree anywhere, and so the two are equivalent for reversible models only.
+When Felsenstein says the two approaches are the same (on p253 of his book), it is quite a vague statement:
+e.g. later in the same paragraph, he says that both peeling and pruning are variants of Horner's rule, and particular cases of dynamic programming.
+All of this is somewhat academic; basically I'm thinking the pruning/peeling distinction might yet have some pedagogic value, and we might not need to abandon it entirely.
+%
+\item We should address all the reviewer's comments about the figures if possible.
+I have added xfig files ``stree.fig'' and ``stree-mutations.fig'' to the repository; the latter shows an evolutionary trajectory of the TKFST, as requested by the reviewer.
+%
+\item Finally, a very minor/trivial point: would you consider a name-change for IndieGram to something beginning with ``Evol''?
+The EvolDeeds package is EvolDoer (previously released as ``tkfstalign''), EvolSayer and IndieGram.
+I put a few random suggestions for Evol-names at {\bf biowiki.org/EvolDeeds}.
+Don't take this too seriously, of course!
+I'll understand if you want to stick with IndieGram, and this is obviously negligible compared to the other revisions.
+%
+\end{enumerate}
+\hrulefill{}}
+
 %% END MACROS SECTION
@@
 \section*{Abstract}
-The existence of an all-RNA primordial ribosome was proposed by
-Francis Crick in 1968.  Recently, a domain-level model for the
-original ribosome was proposed by Smith et al.  Given the abundance of
-ribosomal RNA sequence data and the many successful reconstructions of
-ancestral proteins, the reconstruction and synthesis of ancestral
-ribosomes (and other RNAs) is a feasible goal for paleogenetics.  This
-will require new bioinformatics tools and methods, in particular a
-robust theoretical framework for reconstructing histories of
-substitutions, indels and changes in RNA structure.
+% The existence of an all-RNA primordial ribosome was proposed by
+% Francis Crick in 1968.  Recently, a domain-level model for the
+% original ribosome was proposed by Smith et al.  Given the abundance of
+% ribosomal RNA sequence data and the many successful reconstructions of
+% ancestral proteins, the reconstruction and synthesis of ancestral
+% ribosomes (and other RNAs) is a feasible goal for paleogenetics.  This
+% will require new bioinformatics tools and methods, in particular a
+% robust theoretical framework for reconstructing histories of
+% substitutions, indels and changes in RNA structure.
 
@@
 
-We implemented the above algorithms for a simple model of pairwise RNA structural evolution;
-in particular the algorithms for maximum likelihood (ML) alignment of three known RNA structures an a known phylogeny
-and inference of the common ancestral structure.
-We compared this ML algorithm to a variety of related, but simpler, techniques,
-including ML alignment algorithms for simpler models that omitted various aspects of the full model
-and also a posterior-decoding alignment algorithm for one of the simpler models.
-
-In our tests, incorporation of basepair structure was the most important factor for accurate inference of alignment;
-appropriate use of posterior-decoding was the next most important factor;
-and fine details of the model were least important.
-Posterior-decoding heuristics can be substantially faster than exact phylogenetic inference,
-so this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics where possible.
-For more exact probabilistic inference, we discuss the use of transducer composition for ML (or MCMC) inference on phylogenies,
-including possible ways to make the core operations tractable.
+We implemented the above algorithms for a simple model of pairwise RNA
+structural evolution; in particular the algorithms for maximum
+likelihood (ML) alignment of three known RNA structures an a known
+phylogeny and inference of the common ancestral structure.  We
+compared this ML algorithm to a variety of related, but simpler,
+techniques, including ML alignment algorithms for simpler models that
+omitted various aspects of the full model and also a
+posterior-decoding alignment algorithm for one of the simpler models.
+
+In our tests, incorporation of basepair structure was the most
+important factor for accurate inference of alignment; appropriate use
+of posterior-decoding was the next most important factor; and fine
+details of the model were least important.  Posterior-decoding
+heuristics can be substantially faster than exact phylogenetic
+inference, so this motivates the use of sum-over-pairs (and
+approximate sum-over-pairs) heuristics where possible.  For more exact
+probabilistic inference, we discuss the use of transducer composition
+for ML (or MCMC) inference on phylogenies, including possible ways to
+make the core operations tractable.
 
@@
 
-\paragraph{Terminology and normalization.}
+\subsubsection*{Terminology and normalization}
 
@@
 
-\paragraph{Fold Envelopes.}
 We use the fold envelope concept \cite{HolmesRubin2002a,Holmes2004}
@@
 
-\paragraph{The Inside algorithm.}
 The Inside algorithm is used to calculate the likelihood of sequences under an ensemble model.  It is analogous to the Forward algorithm for HMMs.
@@
 
-\paragraph{The CYK algorithm.}
 The CYK algorithm is used to calculate the probability of the
@@
 
-\paragraph{The CYK traceback algorithm.}
 The CYK traceback algorithm, in combination with the CYK algorithm, is
@@
 
-\paragraph{The Outside algorithm.}
 The Outside algorithm is primarily used an an intermediary for calculating nucleotide-level posterior probabilities, e.g. for posterior decoding on the model.
@@
 3) \Rightarrow 441$ states and $O(a \cdot b^2 \cdot N) = O(7 \cdot
-21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in \secname{The composition algorithm.}
+21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in \secname{The composition algorithm}.
  Nonetheless, the extreme complexity of the

### 2009-07-14 paper.tex
Fixed tables per plos request.

@@
       SCFG) of the TKF Structure Tree model.}}
-  \begin{tabular}{|l|rrrl|}
+  \begin{tabular}{|l|r|r|r|l|}
     \hline
@@
     \textbf{Singlet transducer (single-sequence SCFG) of the TKF Structure Tree model.}}
-  \begin{tabular}{|rcl|r||rcl|r|}
+  \begin{tabular}{|l|r||l|r|}
     \hline
-    Source & $\rightarrow$ & Destination & probability & Source & $\rightarrow$ & Destination & probability \\ \hline
-    $L$ & $\rightarrow$ & $u \,\, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $S$ & $\rightarrow$ & $u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$ \\ \hline
-    & $|$ & $B$ & $\kappa_l \cdot \pi_l(S)$ & & $|$ & $B_e$ & $1-\kappa_s$ \\ \hline
-    & $|$ & $\Send$ & $1-\kappa_l$ & & & & \\ \hline
-    $I_L$ & $\rightarrow$ & $u \, \, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $I_S$ & $\rightarrow$ & $u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$\\ \hline
-    & $|$ & $B$ & $\kappa_l \cdot \pi_l(S)$ & & $|$ & $B_e$ & $1-\kappa_s$\\ \hline
-    & $|$ & $\Send$ & $1-\kappa_l$ & & & & \\ \hline \hline
-    $B$ & $\rightarrow$ & $(L \,\, S)$ & 1 & & & & \\ \hline
-    $B_e$ & $\rightarrow$ & $(L \,\, \Send)$ & 1 & & & & \\ \hline
+    Source $\rightarrow$ Destination & probability & Source $\rightarrow$ Destination & probability \\ \hline
+    $L \rightarrow u \,\, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $S \rightarrow u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$ \\ \hline
+    $\phantom{L} \rightarrow B$ & $\kappa_l \cdot \pi_l(S)$ & $\phantom{S} \rightarrow B_e$ & $1-\kappa_s$ \\ \hline
+    $\phantom{L} \rightarrow \Send$ & $1-\kappa_l$ & & \\ \hline
+    $I_L \rightarrow u \, \, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $I_S \rightarrow u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$\\ \hline
+    $\phantom{I_L} \rightarrow B$ & $\kappa_l \cdot \pi_l(S)$ & $\phantom{I_S} \rightarrow B_e$ & $1-\kappa_s$\\ \hline
+    $\phantom{I_L} \rightarrow \Send$ & $1-\kappa_l$ & & \\ \hline \hline
+    $B \rightarrow (L \,\, S)$ & 1 & & \\ \hline
+    $B_e \rightarrow (L \,\, \Send)$ & 1 & & \\ \hline
   \end{tabular}
@@
       model.}}
-  \begin{tabular}{|l|rrrl|} \hline
+  \begin{tabular}{|l|r|r|r|l|} \hline
     State & $\type$ & $\absorb$ & $\emit$ & description \\ \hline
@@
       the TKF Structure Tree model.}}
-  \begin{tabular}{|rcl|r||rcl|r|} \hline
-    Source & $\rightarrow$ & Destination & probability & Source & $\rightarrow$ & Destination & probability \\ \hline
-    $L$ & $\rightarrow$ & $w \,\, I_L$ & $\beta_l(t) \cdot \pi_l(w)$ & $S$ & $\rightarrow$ & $w \,\, I_S \,\, x$ & $\beta_s(t) \cdot \pi_s(wx)$ \\ \hline
-    & $|$ & $B_i$ & $\beta_l(t) \cdot \pi_l(S)$ & & $|$ & $W_S$ & $1-\beta_s(t)$ \\ \hline
-    & $|$ & $W_L$ & $1-\beta_l(t)$ & & & &  \\ \hline
-    $I_L$ & $\rightarrow$ & $w \,\, I_L$ & $\beta_l(t) \cdot \pi_l(w)$ & $I_S$ & $\rightarrow$ & $w \,\, I_S \,\, x$ & $\beta_s(t) \cdot \pi_s(wx)$ \\ \hline
-    & $|$ & $B_i$ & $\beta_l(t) \cdot \pi_l(S)$ & & $|$ & $W_S$ & $1-\beta_s(t)$ \\ \hline
-    & $|$ & $W_L$ & $1-\beta_l(t)$ & & & & \\ \hline
-    $M_L$ & $\rightarrow$ & $w \,\, I_L$ & $\beta_l(t) \cdot \pi_l(w)$ & $M_S$ & $\rightarrow$ & $w \,\, I_S \,\, x$ & $\beta_s(t) \cdot \pi_s(wx)$ \\ \hline
-    & $|$ & $B_i$ & $\beta_l(t) \cdot \pi_l(S)$ & & $|$ & $W_S$ & $1-\beta_s(t)$ \\ \hline
-    & $|$ & $W_L$ & $1-\beta_l(t)$ & & & &  \\ \hline
-    $D_L$ & $\rightarrow$ & $w \,\, I_L$ & $\gamma_l(t) \cdot \pi_l(w)$ & $D_S$ & $\rightarrow$ & $w \,\, I_S \,\, x$ & $\gamma_s(t) \cdot \pi_s(wx)$ \\ \hline
-    & $|$ & $B_i$ & $\gamma_l(t) \cdot \pi_l(S)$ & & $|$ & $W_S$ & $1-\gamma_s(t)$ \\ \hline
-    & $|$ & $W_L$ & $1-\gamma_l(t)$ & & & &  \\ \hline
-    $W_L$ & $\rightarrow$ & $w \,\, M_L$ & $\alpha_l(t) \cdot M_l(u \rightarrow w)$ & $W_S$ & $\rightarrow$ & $w \,\, M_S \,\, x$ & $\alpha_s(t) \cdot M_s(uv \rightarrow wx)$ \\ \hline
-    & $|$ & $D_L$ & $1-\alpha_l(t)$ & & $|$ & $D_S$ & $1-\alpha_s(t)$ \\ \hline
-    & $|$ & $B$ & $\alpha_l(t)$ & & $|$ & $B_e$ & $1$ \\ \hline
-    & $|$ & $B_p$ & $1-\alpha_l(t)$ & & & &  \\ \hline
-    & $|$ & $\Send$ & $1$ & & & &  \\ \hline \hline
-    $B$ & $\rightarrow$ & $L \,\, S$ & 1 & $B_p$ & $\rightarrow$ & $L \, \, \Send$ & 1 \\ \hline
-    $B_i$ & $\rightarrow$ & $L_i \,\, S_i$ & 1 & $B_e$ & $\rightarrow$ & $L \, \, \Send$ & 1 \\ \hline \hline
+  \begin{tabular}{|l|r||l|r|} \hline
+    Source $\rightarrow$ Destination & probability & Source $\rightarrow$ Destination & probability \\ \hline
+    $L \rightarrow w \,\, I_L$ & $\beta_l(t) \cdot \pi_l(w)$ & $S \rightarrow w \,\, I_S \,\, x$ & $\beta_s(t) \cdot \pi_s(wx)$ \\ \hline
+    $\phantom{L} \rightarrow B_i$ & $\beta_l(t) \cdot \pi_l(S)$ & $\phantom{S} \rightarrow W_S$ & $1-\beta_s(t)$ \\ \hline
+    $\phantom{L} \rightarrow W_L$ & $1-\beta_l(t)$ & & \\ \hline
+    $I_L \rightarrow w \,\, I_L$ & $\beta_l(t) \cdot \pi_l(w)$ & $I_S \rightarrow w \,\, I_S \,\, x$ & $\beta_s(t) \cdot \pi_s(wx)$ \\ \hline
+    $\phantom{I_L} \rightarrow B_i$ & $\beta_l(t) \cdot \pi_l(S)$ & $\phantom{I_S} \rightarrow W_S$ & $1-\beta_s(t)$ \\ \hline
+    $\phantom{I_L} \rightarrow W_L$ & $1-\beta_l(t)$ & & \\ \hline
+    $M_L \rightarrow w \,\, I_L$ & $\beta_l(t) \cdot \pi_l(w)$ & $M_S \rightarrow w \,\, I_S \,\, x$ & $\beta_s(t) \cdot \pi_s(wx)$ \\ \hline
+    $\phantom{M_L} \rightarrow B_i$ & $\beta_l(t) \cdot \pi_l(S)$ & $\phantom{M_S} \rightarrow W_S$ & $1-\beta_s(t)$ \\ \hline
+    $\phantom{M_L} \rightarrow W_L$ & $1-\beta_l(t)$ & & \\ \hline
+    $D_L \rightarrow w \,\, I_L$ & $\gamma_l(t) \cdot \pi_l(w)$ & $D_S \rightarrow w \,\, I_S \,\, x$ & $\gamma_s(t) \cdot \pi_s(wx)$ \\ \hline
+    $\phantom{D_L} \rightarrow B_i$ & $\gamma_l(t) \cdot \pi_l(S)$ & $\phantom{D_S} \rightarrow W_S$ & $1-\gamma_s(t)$ \\ \hline
+    $\phantom{D_L} \rightarrow W_L$ & $1-\gamma_l(t)$ & & \\ \hline
+    $W_L \rightarrow w \,\, M_L$ & $\alpha_l(t) \cdot M_l(u \rightarrow w)$ & $W_S \rightarrow w \,\, M_S \,\, x$ & $\alpha_s(t) \cdot M_s(uv \rightarrow wx)$ \\ \hline
+    $\phantom{W_L} \rightarrow D_L$ & $1-\alpha_l(t)$ & $\phantom{W_S} \rightarrow D_S$ & $1-\alpha_s(t)$ \\ \hline
+    $\phantom{W_L} \rightarrow B$ & $\alpha_l(t)$ & $\phantom{W_S} \rightarrow B_e$ & $1$ \\ \hline
+    $\phantom{W_L} \rightarrow B_p$ & $1-\alpha_l(t)$ & & \\ \hline
+    $\phantom{W_L} \rightarrow \Send$ & $1$ & & \\ \hline \hline
+    $B \rightarrow (L \,\, S)$ & 1 & $B_p \rightarrow (L \, \, \Send)$ & 1 \\ \hline
+    $B_i \rightarrow (L_i \,\, S_i)$ & 1 & $B_e \rightarrow (L \, \, \Send)$ & 1 \\ \hline \hline
   \end{tabular}
@@
     \textbf{Percentage sensitivity and positive predictive value (Sensitivity / PPV) for pairwise nucleotide-level alignments in the \bralibaseII\ benchmark.}}
-  \begin{tabular}{|r|rrrr|}
+  \begin{tabular}{|r|r|r|r|r|}
     \hline
@@
     \textbf{Percentage sensitivity and positive predictive value (Sensitivity / PPV) for predicted base-pairs in the \bralibaseII\ benchmark.}}
-  \begin{tabular}{|r|rrrr|}
+  \begin{tabular}{|r|r|r|r|r|}
     \hline
@@
      (as reported by \indiegram).}}
- \begin{tabular}{|l|rrr|} \hline
+ \begin{tabular}{|l|r|r|r|} \hline
     Family & Sequence lengths & Memory & Time \\ \hline

### 2009-07-14 paper.tex
more

@@
 \newcommand\allrules[1]{{\cal A}(#1)}
-
-\newcommand\revisionnotes
-{\noindent \hrulefill{} \\ {\bf Revision Notes:}
-\begin{enumerate}
-\item These comments are not a complete summary of the review, just a partial strategy for addressing some of the points.
-The review is the more authoritative guide...
-He is giving us the benefit of the doubt but acceptance is still very much in the balance, I think, so we need to stick at it,
-and take it that extra step beyond minimal compliance with his suggestions.
-%
-\item Obviously we have to do simulation tests and hopefully EvolSayer will be of some use there.
-One thing to be careful of is that there are certain regimes of rate/branch-length parameter space where the ``trivial'' outcomes identified by the reviewer
-(ancestral sequence is always identical to one of the descendants) are the norm.
-Oscar and I have been looking at this quite a bit recently (see also 2003 paper by Mossel).
-The regime we are probably most interested in is where the branch lengths are of comparable length
-and so the two slightly-longer branches can ``out-vote'' the shorter branch under certain conditions.
-This should probably be commented on in the paper: the reviewer says he would be happy with ``a small, noncomprehensive set of anecdotes''
-but if we can rationalize why some parameter settings are more interesting than others, then that's all to the better.
-%
-\item I would actually be very wary of taking that ``small noncomprehensive set of anecdotes'' at face value.
-This is one area where I sense we're being given enough rope to hang ourselves... an unsatisfactory response to the simulation issue will be unequivocal grounds for rejection.
-Obviously there are limits to the number of simulations we can run, but with EvolSayer's output statistics (allowing us to filter out ``interesting'' cases), we can automate it somewhat.
-%
-\item I wrote out the TKF structure tree for the case $a \to b \to c$,
-which can be used to estimate the intermediate sequence $b$
-(e.g. for Redelings-Suchard).
-I think this could be useful in explaining tree transducer composition more concretely, and why is important:
-by inspecting these tables, the reader can see that the structure of this grammar is quite complicated,
-and should then be prepared for the idea that the ``composition algorithm'' is simply generating this table
-(or other tables for other phylogenies).
-The latex for this is in the file ``tkfst.tex'', and is currently include'd at the start of the supplementary information (it should go in the main paper though, I think).
-%
-\item Can we dump out the grammar for the three-branch TKFST and include it in the Supplement?
-(Or the main paper?) I realize it's huge, but this very hugeness does motivate the transducer composition algorithm,
-in the same way that the two-branch model (which is now included in a more human-readable form) does.
-%
-\item If we are moving the description of loopy DP to the main paper, I think it needs to be more precise.
-Currently it just says ``we repeatedly iterate over strongly-connected components of the state sub graph'',
-which falls short of the level of precision/detail with which the other DP algorithms are presented.
-The description should encompass alternate strategies (i.e. orders of visiting the states) that you have considered:
-randomized, Tarjan's algorithm, etc.
-The reader should be left understanding the pros-and-cons of each,
- e.g. worst-case performance or difficulty of implementation.
-No need to agonize about this, but be helpful to potential future developers.
-  (Example reader: a grad student in our group who might try to implement loopy DP themselves.)
-%
-\item The descriptions of the DP algorithms do not currently include any traceback algorithms (CYK or Inside);
-therefore, we haven't technically described how to do ancestral reconstruction!
-Obviously these algorithms are almost trivial, but they're quite important.
-Implementing stochastic loopy-Inside traceback in tripletscfgdp.\{h,cc\} is a key step in transforming it into a useful tool
-(for example, with this extension, Indiegram would be pretty much ready-to-go for Redelings-Suchard sampling).
-The deterministic CYK traceback algorithm is {\em even more} important to present here, because implementing this algorithm is currently our main result!
-The paper should describe {\em both} these traceback algorithms.
-This shouldn't be that hard to do in a comprehensible, rigorous style
- (loopy traceback is much simpler than the algorithms for restoring eliminated states to traceback paths, which are painfully described in my 2003 paper).
-%
-\item All algorithms should be described in full, rather than giving recipes for deriving some algorithms from others.
-In particular, the CYK algorithm is currently summarized by statements along the lines of
-``replace sums over paths through the model (or equivalently, parses) with the $\max()$ operation (e.g., in equation X, sum-over-complicated-expression will be replaced with max-over-complicated-expression)''.
-It shouldn't be too hard to write a latex command/macro that can be used (with ``$\sum$'' or ``$\max$'' as an argument) to generate complete versions of {\em both} algorithms.
-%
-\item The paper the reviewer cites as an example of the simplicity of missing data estimation (Rivas and Eddy (2008) PLoS CB 4:e1000172) is an IID-columns model for a multiple alignment.
-When he says that this issue ``may simply go away if you view the problem as a missing data problem'',
-that is precisely what we are talking about when we discuss null cycle elimination.
-I've tried to explain in ``tkfst.tex'' why this is harder than it superficially appears:
-\begin{enumerate}
-\item eliminating missing ``empty'' columns from an IID-columns model (such as in Rivas and Eddy, 2008) is easy.
-\item eliminating missing emissions from an HMM is slightly harder (c.f. ``wing retraction'' algorithm in HMMer source code)
-but general algorithms do exist.
-\item eliminating null states from an SCFG is much harder, due to null bifurcations.
-\end{enumerate}
-%
-\item The reviewer asks for early discussion of constraints in the context of the ``rate-limiting steps of paleogenetics''.
-We should take this opportunity to clarify that our approach is (1) formulate the model,  (2) apply appropriate constraints.
-Our emphasis here is on (1).
-Cite the literature: QRNA represents an alignment constraint on a pair SCFG, Holmes-Rubin 2002 uses fold constraints on pair SCFGs,
-stemloc uses fold+alignment constraints, Ortheus uses an alignment constraint, etc, but none of these things can be done
-(in a probabilistic modeling framework at least) until you can formulate the model.
-For this paper, we chose to condition on structures essentially because they are easy to constrain
-(fold envelopes can be applied independently to each sequence).
-But we could equally easily have conditioned on alignments. Constraints are easy to tweak, once you have a model.
-%
-\item The reviewer also comments on the phylogeny being a given.
-We can note that even though we constrain the phylogeny, more realistic phylogenetic applications are possible.
-Given a structural phylo-alignment (tree+alignment+structure), we can calculate the likelihood; therefore, given two such structural phylo-alignments,
-we can say which one is more probable (describe their relative posterior probabilities, etc.).
-From this it is possible to make a crude MCMC sampler, just by proposing/accepting/rejecting alignment/tree/structure changes.
-An {\em efficient} MCMC sampler needs to be able to make local topology changes and to resample the multiple alignment around a topology change;
-we can do this too, in principle, but it's beyond this paper.
-%
-\item As I understand it, pruning \& peeling are not exactly equivalent as the reviewer claims, {\em except} in the special case of a reversible model.
-Specifically, pruning is a {\em restricted case} of peeling, giving you the posterior probabilities at the root node, but not any other node.
-Of course, if you have a reversible model, you can re-root the tree anywhere, and so the two are equivalent for reversible models only.
-When Felsenstein says the two approaches are the same (on p253 of his book), it is quite a vague statement:
-e.g. later in the same paragraph, he says that both peeling and pruning are variants of Horner's rule, and particular cases of dynamic programming.
-All of this is somewhat academic; basically I'm thinking the pruning/peeling distinction might yet have some pedagogic value, and we might not need to abandon it entirely.
-%
-\item We should address all the reviewer's comments about the figures if possible.
-I have added xfig files ``stree.fig'' and ``stree-mutations.fig'' to the repository; the latter shows an evolutionary trajectory of the TKFST, as requested by the reviewer.
-%
-\item Finally, a very minor/trivial point: would you consider a name-change for IndieGram to something beginning with ``Evol''?
-The EvolDeeds package is EvolDoer (previously released as ``tkfstalign''), EvolSayer and IndieGram.
-I put a few random suggestions for Evol-names at {\bf biowiki.org/EvolDeeds}.
-Don't take this too seriously, of course!
-I'll understand if you want to stick with IndieGram, and this is obviously negligible compared to the other revisions.
-%
-\end{enumerate}
-\hrulefill{}}
 

### 2009-07-15 paper.tex
Production edits

@@
 % ribosomal RNA sequence data and the many successful reconstructions of
-% ancestral proteins, the reconstruction and synthesis of ancestral
-% ribosomes (and other RNAs) is a feasible goal for paleogenetics.  This
-% will require new bioinformatics tools and methods, in particular a
-% robust theoretical framework for reconstructing histories of
-% substitutions, indels and changes in RNA structure.
+% ancestral proteins,
+
+The reconstruction and synthesis of ancestral RNAs is a feasible goal for paleogenetics.
+This will require new bioinformatics methods, including a
+robust statistical framework for reconstructing histories of
+substitutions, indels and structural changes.
 
@@
 In our tests, incorporation of basepair structure was the most
-important factor for accurate inference of alignment; appropriate use
-of posterior-decoding was the next most important factor; and fine
+important factor for accurate alignment inference; appropriate use
+of posterior-decoding was next; and fine
 details of the model were least important.  Posterior-decoding
 heuristics can be substantially faster than exact phylogenetic
-inference, so this motivates the use of sum-over-pairs (and
-approximate sum-over-pairs) heuristics where possible.  For more exact
+inference, so this motivates the use of sum-over-pairs heuristics where possible
+(and approximate sum-over-pairs).  For more exact
 probabilistic inference, we discuss the use of transducer composition

### 2009-07-17 paper.tex
Moved algorithms to separate latex files

@@
 
-\usepackage[boxed]{algorithm2e}
+% Commented out inclusion of algorithm2e environment because these are now separate figures in figs_plos subdir - IH, 7/17/09
+%\usepackage[boxed]{algorithm2e}
 \usepackage{mathrsfs}
@@
 
-\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
+% Changed algorithms to external figures - IH, 7/17/09
+%\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
+%\newcommand{\alglabel}[1]{\label{alg:#1}}
+\newcommand{\algref}[1]{Figure~\ref{alg:#1}}
 \newcommand{\alglabel}[1]{\label{alg:#1}}
@@
 % figures in figs_plos:
-% 1: stree-with-structure.pdf
-% 2: stree-mutations-with-structures.pdf
-% 3: perfect.pdf
-% 4: ama.pdf
+%  1: inside.pdf
+%  2: insidetransemit.pdf
+%  3: cyk.pdf
+%  4: cyktransemit.pdf
+%  5: cyktraceback.pdf
+%  6: outside.pdf
+%  7: outsidetransemit.pdf
+%  8: stree-with-structure.pdf
+%  9: stree-mutations-with-structures.pdf
+% 10: perfect.pdf
+% 11: ama.pdf
 
@@
 
-\begin{algorithm}[!ht]
-
-  \SetKwData{bifurcProb}{bifurcProb}
-  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
-  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
-  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
-  \SetLine
-  \KwIn{sequences $X, Y, Z$}
-
-  \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
-    \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
-      \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
-        \BlankLine
-        \ForEach{state $\bvec{a}$}{
-          \BlankLine
-          \bifurcProb $\leftarrow 0$\;
-          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
-            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
-              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
-                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
-                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
-              }
-            }
-          }
-          \BlankLine
-          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ $\leftarrow$ \calcTransEmitProb$\left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
-          store $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-        }
-        \BlankLine
-      }
-    }
-  }
-  \KwRet{$\alpha_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;
-
-  \caption{\alglabel{inside}
-    The constrained Inside algorithm for three sequences $X, Y, Z$.
-    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
-    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
-  }
-\end{algorithm}
-
 The transition and emission probability $\mathrm{calcTransEmitProb}(\bvec{a}; \cdot)$ can be calculated 
@@
 Pseudocode for the constrained calculation is given in \algref{insidetransemit}.
-
-\begin{algorithm}[!ht]
-  \SetKwData{emitProb}{emitProb}
-  \SetLine
-  \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Inside matrix $\alpha$}
-
-  \emitProb $\leftarrow 0$\;
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
-    \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \alpha_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-  }
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
-    \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
-    \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \alpha_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
-  }
-  \KwRet{\emitProb}\;
-  \caption{\alglabel{insidetransemit}
-    Subroutine $\mathrm{calcTransEmitProb}()$ for the Inside algorithm.
-    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
-  }
-\end{algorithm}
 
@@
 
-\begin{algorithm}[!ht]
-
-  \SetKwData{maxProb}{maxProb}
-  \SetKwData{bifurcProb}{bifurcProb}
-  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
-  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
-  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
-  \SetLine
-  \KwIn{sequences $X, Y, Z$}
-
-  \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
-    \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
-      \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
-        \BlankLine
-        \ForEach{state $\bvec{a}$}{
-          \BlankLine
-          \bifurcProb $\leftarrow 0$\;
-          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
-            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
-              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
-                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
-                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
-              }
-            }
-          }
-          \BlankLine
-          $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ \\
-          $\qquad \leftarrow \max \left( \calcTransEmitProb \left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right), \bifurcProb \right)$ \;
-          store $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-        }
-        \BlankLine
-      }
-    }
-  }
-  \KwRet{$\gamma_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;
-
-  \caption{\alglabel{cyk}
-    The constrained CYK algorithm for three sequences $X, Y, Z$.
-    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
-    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
-  }
-\end{algorithm}
-
 The resulting CYK algorithm is shown in \algref{cyk}.
@@
 
-\begin{algorithm}[!ht]
-  \SetKwData{emitProb}{emitProb}
-  \SetLine
-  \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate CYK matrix $\gamma$}
-
-  \emitProb $\leftarrow 0$\;
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
-    \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{b} \right)\, \gamma_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \right)$\;
-  }
-  \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
-    \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
-    \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \gamma_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right) \right)$\;
-  }
-  \KwRet{\emitProb}\;
-  \caption{\alglabel{cyktransemit}
-    Subroutine $\mathrm{calcTransEmitProb}()$ for the CYK algorithm.
-    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
-  }
-\end{algorithm}
-
 The CYK traceback algorithm, in combination with the CYK algorithm, is
@@
 \algref{cyktraceback} gives the constrained CYK traceback algorithm.
-
-\begin{algorithm}[!ht]
-
-  \SetKw{Select}{select}
-  \SetKw{Pop}{pop}
-  \SetKw{Push}{push}
-  \SetKw{Goto}{goto}
-  \SetKw{Mainloop}{main loop}
-  \SetKwData{coordinateStack}{coordinateStack}
-  \SetKwFunction{goto}{goto}
-  \SetKwBlock{Begin}{begin main loop:}{end}
-  \SetLine
-  \KwIn{sequences $X, Y, Z$}
-
-  pushdown stack coordinateStack; \tcc*[f]{hold (state, subsequence triplet) pairs} \\
-  $\bvec{a} \leftarrow \Sstart$; \tcc*[f]{current ensemble state} \\
-  $n^{(X)} \leftarrow N^{(X)}$; \tcc*[f]{current $X$ subsequence; $N^{(X)}$ is the outermost subsequence} \\
-  $n^{(Y)} \leftarrow N^{(Y)}$; \tcc*[f]{current $Y$ subsequence; $N^{(Y)}$ is the outermost subsequence} \\
-  $n^{(Z)} \leftarrow N^{(Z)}$; \tcc*[f]{current $Z$ subsequence; $N^{(Z)}$ is the outermost subsequence} \\
-  clear \coordinateStack;\
-
-  \Begin{
-    output current state $\bvec{a}$ and subsequence triplet $\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-    \SetLine
-    \uIf(\tcc*[f]{end of a parse subtree}){$\bvec{a}$ is the $\Send$ state}{
-      \lIf(\tcc*[f]{end of the parse tree}){\coordinateStack is empty}{\Return}
-      \Pop $\left( \bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ from \coordinateStack\;
-      \Goto \Mainloop\;
-    }
-    \uElseIf(\tcc*[f]{bifurcation $\bvec{a} \to \bvec{c} \bvec{b}$}){$\bvec{a}$ is a bifurcation state}{
-      \Select $\left( n^{(X)}_L, n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$,
-      $\left( n^{(Y)}_L, n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$,
-      $\left( n^{(Z)}_L, n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$ \\
-      such that \\
-      $\gamma_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)} \right) = \gamma_{\bvec{a}} \left(n^{(X)}_L, n^{(Y)}_L, n^{(Z)}_L \right)  \gamma_{\bvec{a}} \left(n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ \;
-      \Push $\left( \bvec{c}, n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ onto \coordinateStack\;
-      $\bvec{a} \leftarrow \bvec{b}$\;
-      $n^{(X)} \leftarrow n^{(X)}_L$\;
-      $n^{(Y)} \leftarrow n^{(Y)}_L$\;
-      $n^{(Z)} \leftarrow n^{(Z)}_L$\;
-      \Goto \Mainloop\;
-    }
-    \Else(\tcc*[f]{$\Semit$ or $\Snull$ state}){
-      $m^{(X)} \leftarrow c_{in}\left(\bvec{b}; n^{(X)}\right)$ \;
-      $m^{(Y)} \leftarrow c_{in}\left(\bvec{b}; n^{(Y)}\right)$ \;
-      $m^{(Z)} \leftarrow c_{in}\left(\bvec{b}; n^{(Z)}\right)$ \;
-      \Select $\bvec{b} \in \left\{ b : \exists \,\bvec{a} \to \bvec{l}\bvec{b}\bvec{r} \right\}$ \\
-      such that \\
-      $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) = P(\bvec{a} \to \bvec{l}\bvec{b}\bvec{r}) \gamma_{\bvec{b}} \left( m^{(X)}, m^{(Y)}, m^{(Z)} \right)$\;
-      $\bvec{a} \leftarrow \bvec{b}$\;
-      $n^{(X)} \leftarrow m^{(X)}$\;
-      $n^{(Y)} \leftarrow m^{(Y)}$\;
-      $n^{(Z)} \leftarrow m^{(Z)}$\;
-      \Goto \Mainloop\;
-    }
-  }
-  \caption{\alglabel{cyktraceback}
-    The constrained CYK traceback algorithm for three sequences $X, Y, Z$.
-  }
-\end{algorithm}
 
@@
 used in the Outside algorithm are defined below.
-
-\begin{algorithm}[!ht]
-  \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
-  \SetKwData{bifurcProb}{bifurcProb}
-  \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
-  \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
-  \SetLine
-  \KwIn{sequences $X, Y, Z$, Inside matrix $\alpha$}
-
-  \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
-    \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
-      \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
-        \BlankLine
-        \ForEach{state $\bvec{b}$}{
-          \BlankLine
-          \bifurcProb $\leftarrow 0$\;
-          \ForEach{$\left( n^{(X)}_O,n^{(X)}_L \right) \in b_{out,L}\left( n^{(X)} \right)$}{
-            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_L \right) \in b_{out,L}\left( n^{(Y)} \right)$}{
-              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_L \right) \in b_{out,L}\left( n^{(Z)} \right)$}{
-                \bifurcProb += \calcLBifurcProb($\bvec{b}; \cdot$)\;
-              }
-            }
-          }
-          \ForEach{$\left( n^{(X)}_O,n^{(X)}_R \right) \in b_{out,R}\left( n^{(X)} \right)$}{
-            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_R \right) \in b_{out,R}\left( n^{(Y)} \right)$}{
-              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_R \right) \in b_{out,R}\left( n^{(Z)} \right)$}{
-                \bifurcProb += \calcRBifurcProb($\bvec{b}; \cdot$)\;
-              }
-            }
-          }
-          \BlankLine
-          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \leftarrow$ \calcTransEmitProb$\left( \bvec{b}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
-          store $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-        }
-        \BlankLine
-      }
-    }
-  }
-  \caption{\alglabel{outside}
-    The constrained Outside algorithm for three sequences $X, Y, Z$.
-    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Outside fill order
-    with $\Semit$ states first, then $\Snull$ states in topological order.
-  }
-\end{algorithm}
 
@@
 where $N^{(X)}$ is the outermost subsequence for sequence $X$, etc.
-
-\begin{algorithm}[!ht]
-  \SetKwData{emitProb}{emitProb}
-  \SetLine
-  \KwIn{state $\bvec{b}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Outside matrix $\beta$}
-
-  \emitProb $\leftarrow 0$\;
-  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{b}$} {
-    \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \beta_{\bvec{a}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
-  }
-  \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
-    \lIf{$c_{out}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{out}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{out}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
-    \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \beta_{\bvec{a}}\left( c_{out}\left(\bvec{b}; n^{(X)}\right), c_{out}\left(\bvec{b}; n^{(Y)}\right), c_{out}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
-  }
-  \KwRet{\emitProb}\;
-  \caption{\alglabel{outsidetransemit}
-    Subroutine $\mathrm{calcTransEmitProb}()$ for the Outside algorithm.
-    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
-  }
-\end{algorithm}
 
@@
 \section*{Figure Legends}
+
+% The following one-liner was used to convert all algorithms into standalone tex files for use as figures
+% cat paper.tex | perl -e '@x=<>;$x=join("",@x);$x=~s/\r\n/\n/g;while($x=~/.begin.algorithm.(.*?).caption.*?.alglabel.([^\}]+)/sg){($a,$f)=($1,$2);open F,">figs_plos/$f.tex";print F "\\documentclass[10pt]{article}\n\\input{algdefs.tex}\n\\begin{document}\n\\begin{algorithm}$a\n\\end{algorithm}\n\\end{document}\n";close F}'
+
+\begin{figure}[!ht]
+% 
+%   \SetKwData{bifurcProb}{bifurcProb}
+%   \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
+%   \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
+%   \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
+%   \SetLine
+%   \KwIn{sequences $X, Y, Z$}
+% 
+%   \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
+%     \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
+%       \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
+%         \BlankLine
+%         \ForEach{state $\bvec{a}$}{
+%           \BlankLine
+%           \bifurcProb $\leftarrow 0$\;
+%           \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
+%             \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
+%               \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
+%                 \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
+%                 \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
+%               }
+%             }
+%           }
+%           \BlankLine
+%           $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ $\leftarrow$ \calcTransEmitProb$\left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+%           $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
+%           store $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+%         }
+%         \BlankLine
+%       }
+%     }
+%   }
+%   \KwRet{$\alpha_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;
+
+  \caption{\alglabel{inside}
+{\bf    The constrained Inside algorithm for three sequences $X, Y, Z$. }
+    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
+    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
+The subroutine $\mathrm{calcTransEmitProb}()$ is given in \algref{insidetransemit}.
+  }
+\end{figure}
+
+\begin{figure}[!ht]
+%   \SetKwData{emitProb}{emitProb}
+%   \SetLine
+%   \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Inside matrix $\alpha$}
+% 
+%   \emitProb $\leftarrow 0$\;
+%   \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
+%     \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \alpha_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+%   }
+%   \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
+%     \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
+%     \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \alpha_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
+%   }
+%   \KwRet{\emitProb}\;
+  \caption{\alglabel{insidetransemit}
+{\bf    Subroutine $\mathrm{calcTransEmitProb}()$ for the Inside algorithm. }
+    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
+The Inside algorithm itself is given in \algref{inside}.
+  }
+\end{figure}
+
+\begin{figure}[!ht]
+% 
+%   \SetKwData{maxProb}{maxProb}
+%   \SetKwData{bifurcProb}{bifurcProb}
+%   \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
+%   \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
+%   \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
+%   \SetLine
+%   \KwIn{sequences $X, Y, Z$}
+% 
+%   \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
+%     \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
+%       \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
+%         \BlankLine
+%         \ForEach{state $\bvec{a}$}{
+%           \BlankLine
+%           \bifurcProb $\leftarrow 0$\;
+%           \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
+%             \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
+%               \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
+%                 \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
+%                 \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
+%               }
+%             }
+%           }
+%           \BlankLine
+%           $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ \\
+%           $\qquad \leftarrow \max \left( \calcTransEmitProb \left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right), \bifurcProb \right)$ \;
+%           store $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+%         }
+%         \BlankLine
+%       }
+%     }
+%   }
+%   \KwRet{$\gamma_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;
+%   
+  \caption{\alglabel{cyk}
+{\bf    The constrained CYK algorithm for three sequences $X, Y, Z$. }
+    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Inside fill order
+    with $\Semit$ states first, then $\Snull$ states in reverse topological order.
+The subroutine $\mathrm{calcTransEmitProb}()$ is given in \algref{cyktransemit}.
+The algorithm to find the Maximum Likelihood parse tree by traceback through the CYK matrix is given in \algref{cyktraceback}.
+  }
+\end{figure}
+
+\begin{figure}[!ht]
+%   \SetKwData{emitProb}{emitProb}
+%   \SetLine
+%   \KwIn{state $\bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate CYK matrix $\gamma$}
+% 
+%   \emitProb $\leftarrow 0$\;
+%   \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{b}$} {
+%     \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{b} \right)\, \gamma_{\bvec{b}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \right)$\;
+%   }
+%   \ForEach{$\bvec{b} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
+%     \lIf{$c_{in}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{in}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{in}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
+%     \emitProb = $\max \left( \emitProb, P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \gamma_{\bvec{b}}\left( c_{in}\left(\bvec{b}; n^{(X)}\right), c_{in}\left(\bvec{b}; n^{(Y)}\right), c_{in}\left(\bvec{b}; n^{(Z)}\right) \right) \right)$\;
+%   }
+%   \KwRet{\emitProb}\;
+  \caption{\alglabel{cyktransemit}
+{\bf    Subroutine $\mathrm{calcTransEmitProb}()$ for the CYK algorithm. }
+    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
+The CYK algorithm itself is given in \algref{cyk}.
+  }
+\end{figure}
+
+\begin{figure}[!ht]
+% 
+%   \SetKw{Select}{select}
+%   \SetKw{Pop}{pop}
+%   \SetKw{Push}{push}
+%   \SetKw{Goto}{goto}
+%   \SetKw{Mainloop}{main loop}
+%   \SetKwData{coordinateStack}{coordinateStack}
+%   \SetKwFunction{goto}{goto}
+%   \SetKwBlock{Begin}{begin main loop:}{end}
+%   \SetLine
+%   \KwIn{sequences $X, Y, Z$}
+% 
+%   pushdown stack coordinateStack; \tcc*[f]{hold (state, subsequence triplet) pairs} \\
+%   $\bvec{a} \leftarrow \Sstart$; \tcc*[f]{current ensemble state} \\
+%   $n^{(X)} \leftarrow N^{(X)}$; \tcc*[f]{current $X$ subsequence; $N^{(X)}$ is the outermost subsequence} \\
+%   $n^{(Y)} \leftarrow N^{(Y)}$; \tcc*[f]{current $Y$ subsequence; $N^{(Y)}$ is the outermost subsequence} \\
+%   $n^{(Z)} \leftarrow N^{(Z)}$; \tcc*[f]{current $Z$ subsequence; $N^{(Z)}$ is the outermost subsequence} \\
+%   clear \coordinateStack;\
+% 
+%   \Begin{
+%     output current state $\bvec{a}$ and subsequence triplet $\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+%     \SetLine
+%     \uIf(\tcc*[f]{end of a parse subtree}){$\bvec{a}$ is the $\Send$ state}{
+%       \lIf(\tcc*[f]{end of the parse tree}){\coordinateStack is empty}{\Return}
+%       \Pop $\left( \bvec{a}, n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ from \coordinateStack\;
+%       \Goto \Mainloop\;
+%     }
+%     \uElseIf(\tcc*[f]{bifurcation $\bvec{a} \to \bvec{c} \bvec{b}$}){$\bvec{a}$ is a bifurcation state}{
+%       \Select $\left( n^{(X)}_L, n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$,
+%       $\left( n^{(Y)}_L, n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$,
+%       $\left( n^{(Z)}_L, n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$ \\
+%       such that \\
+%       $\gamma_{\bvec{a}} \left(n^{(X)}, n^{(Y)}, n^{(Z)} \right) = \gamma_{\bvec{a}} \left(n^{(X)}_L, n^{(Y)}_L, n^{(Z)}_L \right)  \gamma_{\bvec{a}} \left(n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ \;
+%       \Push $\left( \bvec{c}, n^{(X)}_R, n^{(Y)}_R, n^{(Z)}_R \right)$ onto \coordinateStack\;
+%       $\bvec{a} \leftarrow \bvec{b}$\;
+%       $n^{(X)} \leftarrow n^{(X)}_L$\;
+%       $n^{(Y)} \leftarrow n^{(Y)}_L$\;
+%       $n^{(Z)} \leftarrow n^{(Z)}_L$\;
+%       \Goto \Mainloop\;
+%     }
+%     \Else(\tcc*[f]{$\Semit$ or $\Snull$ state}){
+%       $m^{(X)} \leftarrow c_{in}\left(\bvec{b}; n^{(X)}\right)$ \;
+%       $m^{(Y)} \leftarrow c_{in}\left(\bvec{b}; n^{(Y)}\right)$ \;
+%       $m^{(Z)} \leftarrow c_{in}\left(\bvec{b}; n^{(Z)}\right)$ \;
+%       \Select $\bvec{b} \in \left\{ b : \exists \,\bvec{a} \to \bvec{l}\bvec{b}\bvec{r} \right\}$ \\
+%       such that \\
+%       $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) = P(\bvec{a} \to \bvec{l}\bvec{b}\bvec{r}) \gamma_{\bvec{b}} \left( m^{(X)}, m^{(Y)}, m^{(Z)} \right)$\;
+%       $\bvec{a} \leftarrow \bvec{b}$\;
+%       $n^{(X)} \leftarrow m^{(X)}$\;
+%       $n^{(Y)} \leftarrow m^{(Y)}$\;
+%       $n^{(Z)} \leftarrow m^{(Z)}$\;
+%       \Goto \Mainloop\;
+%     }
+%   }
+  \caption{\alglabel{cyktraceback}
+{\bf     The constrained CYK traceback algorithm for three sequences $X, Y, Z$.}
+The CYK algorithm itself is given in \algref{cyk}.
+  }
+\end{figure}
+
+\begin{figure}[!ht]
+%   \SetKwFunction{calcTransEmitProb}{calcTransEmitProb}
+%   \SetKwData{bifurcProb}{bifurcProb}
+%   \SetKwFunction{calcLBifurcProb}{calcLBifurcProb}
+%   \SetKwFunction{calcRBifurcProb}{calcRBifurcProb}
+%   \SetLine
+%   \KwIn{sequences $X, Y, Z$, Inside matrix $\alpha$}
+% 
+%   \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
+%     \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
+%       \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
+%         \BlankLine
+%         \ForEach{state $\bvec{b}$}{
+%           \BlankLine
+%           \bifurcProb $\leftarrow 0$\;
+%           \ForEach{$\left( n^{(X)}_O,n^{(X)}_L \right) \in b_{out,L}\left( n^{(X)} \right)$}{
+%             \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_L \right) \in b_{out,L}\left( n^{(Y)} \right)$}{
+%               \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_L \right) \in b_{out,L}\left( n^{(Z)} \right)$}{
+%                 \bifurcProb += \calcLBifurcProb($\bvec{b}; \cdot$)\;
+%               }
+%             }
+%           }
+%           \ForEach{$\left( n^{(X)}_O,n^{(X)}_R \right) \in b_{out,R}\left( n^{(X)} \right)$}{
+%             \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_R \right) \in b_{out,R}\left( n^{(Y)} \right)$}{
+%               \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_R \right) \in b_{out,R}\left( n^{(Z)} \right)$}{
+%                 \bifurcProb += \calcRBifurcProb($\bvec{b}; \cdot$)\;
+%               }
+%             }
+%           }
+%           \BlankLine
+%           $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \leftarrow$ \calcTransEmitProb$\left( \bvec{b}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+%           $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
+%           store $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+%         }
+%         \BlankLine
+%       }
+%     }
+%   }
+  \caption{\alglabel{outside}
+{\bf    The constrained Outside algorithm for three sequences $X, Y, Z$. }
+    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Outside fill order
+    with $\Semit$ states first, then $\Snull$ states in topological order.
+The subroutine $\mathrm{calcTransEmitProb}()$ is given in \algref{outsidetransemit}.
+  }
+\end{figure}
+
+\begin{figure}[!ht]
+%   \SetKwData{emitProb}{emitProb}
+%   \SetLine
+%   \KwIn{state $\bvec{b}, n^{(X)}, n^{(Y)}, n^{(Z)}$, intermediate Outside matrix $\beta$}
+% 
+%   \emitProb $\leftarrow 0$\;
+%   \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{b}$} {
+%     \emitProb += $P \left( \bvec{a} \to \bvec{b} \right)\, \beta_{\bvec{a}}\left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
+%   }
+%   \ForEach{$\bvec{a} : \exists \,\bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r}$} {
+%     \lIf{$c_{out}\left(\bvec{b}; n^{(X)}\right) \notin \foldenv^{(X)}$ or $c_{out}\left(\bvec{b}; n^{(Y)}\right) \notin \foldenv^{(Y)}$ or $c_{out}\left(\bvec{b}; n^{(Z)}\right) \notin \foldenv^{(Z)}$}{next\;}
+%     \emitProb += $P \left( \bvec{a} \to \bvec{l}\,\bvec{b}\,\bvec{r} \right)\, \beta_{\bvec{a}}\left( c_{out}\left(\bvec{b}; n^{(X)}\right), c_{out}\left(\bvec{b}; n^{(Y)}\right), c_{out}\left(\bvec{b}; n^{(Z)}\right) \right)$\;
+%   }
+%   \KwRet{\emitProb}\;
+  \caption{\alglabel{outsidetransemit}
+{\bf    Subroutine $\mathrm{calcTransEmitProb}()$ for the Outside algorithm. }
+    $\bvec{a}$ and $\bvec{b}$ are ensemble states; $\bvec{l}$ and $\bvec{r}$ are left and right terminal emissions.
+The Outside algorithm itself is given in \algref{outside}.
+  }
+\end{figure}
 

### 2009-07-17 paper.tex
converted figs to EPS
fixed CYK bug (bifurcations were using sum instead of max)

@@
 % figures in figs_plos:
-%  1: inside.pdf
-%  2: insidetransemit.pdf
-%  3: cyk.pdf
-%  4: cyktransemit.pdf
-%  5: cyktraceback.pdf
-%  6: outside.pdf
-%  7: outsidetransemit.pdf
-%  8: stree-with-structure.pdf
-%  9: stree-mutations-with-structures.pdf
-% 10: perfect.pdf
-% 11: ama.pdf
+%  1: inside.eps
+%  2: insidetransemit.eps
+%  3: cyk.eps
+%  4: cyktransemit.eps
+%  5: cyktraceback.eps
+%  6: outside.eps
+%  7: outsidetransemit.eps
+%  8: stree-with-structure.eps
+%  9: stree-mutations-with-structures.eps
+% 10: perfect.eps
+% 11: ama.eps
 
@@
 %               \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
-%                 \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
-%                 \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
+%                 $\bifurcProb \leftarrow \max \left( \bifurcProb, \calcLBifurcProb(\bvec{a}; \cdot) \right)$\;
+%                 $\bifurcProb \leftarrow \max \left( \bifurcProb, \calcRBifurcProb(\bvec{a}; \cdot) \right)$\;
 %               }
@@
 {\bf     The constrained CYK traceback algorithm for three sequences $X, Y, Z$.}
+For compactness we show only the ``if'' case for left-bifurcations ($\bvec{a} \to \bvec{c} \bvec{b}$);
+the case for right-bifurcations ($\bvec{a} \to \bvec{b} \bvec{d}$) is identical,
+with $(\bvec{c},\bvec{b})$ respectively replaced by $(\bvec{b},\bvec{d})$.
 The CYK algorithm itself is given in \algref{cyk}.
